### 2014-03-04 draft.tex
Initial Commit

\documentclass[journal]{style/vgtc}           % final (journal style)
%\documentclass[review,journal]{style/vgtc}         % review (journal style)
%\documentclass[widereview]{style/vgtc}             % wide-spaced review
%\documentclass[preprint,journal]{style/vgtc}       % preprint (journal style)
%\documentclass[electronic,journal]{style/vgtc}     % electronic version, journal

%% Uncomment one of the lines above depending on where your paper is
%% in the conference process. ``review'' and ``widereview'' are for review
%% submission, ``preprint'' is for pre-publication, and the final version
%% doesn't use a specific qualifier. Further, ``electronic'' includes
%% hyperreferences for more convenient online viewing.

%% Please use one of the ``review'' options in combination with the
%% assigned online id (see below) ONLY if your paper uses a double blind
%% review process. Some conferences, like IEEE Vis and InfoVis, have NOT
%% in the past.

%% Please note that the use of figures other than the optional teaser is not permitted on the first page
%% of the journal version.  Figures should begin on the second page and be
%% in CMYK or Grey scale format, otherwise, colour shifting may occur
%% during the printing process.  Papers submitted with figures other than the optional teaser on the
%% first page will be refused.

%% These three lines bring in essential packages: ``mathptmx'' for Type 1
%% typefaces, ``graphicx'' for inclusion of EPS figures. and ``times''
%% for proper handling of the times font family.

\usepackage{mathptmx}
\usepackage{graphicx}
\usepackage{times}

%% We encourage the use of mathptmx for consistent usage of times font
%% throughout the proceedings. However, if you encounter conflicts
%% with other math-related packages, you may want to disable it.

%% This turns references into clickable hyperlinks.
\usepackage[bookmarks,backref=true,linkcolor=black]{hyperref} %,colorlinks
\hypersetup{
  pdfauthor = {},
  pdftitle = {},
  pdfsubject = {},
  pdfkeywords = {},
  colorlinks=true,
  linkcolor= black,
  citecolor= black,
  pageanchor=true,
  urlcolor = black,
  plainpages = false,
  linktocpage
}

%% If you are submitting a paper to a conference for review with a double
%% blind reviewing process, please replace the value ``0'' below with your
%% OnlineID. Otherwise, you may safely leave it at ``0''.
\onlineid{0}

%% declare the category of your paper, only shown in review mode
\vgtccategory{Research}

%% allow for this line if you want the electronic option to work properly
\vgtcinsertpkg

%% In preprint mode you may define your own headline.
%\preprinttext{To appear in an IEEE VGTC sponsored conference.}

%% Paper title.

\title{Interactive Visual Analytics of Image-Centric Cohort Study Data}

%% This is how authors are specified in the journal style

%% indicate IEEE Member or Student Member in form indicated below
\author{Paul Klemm}
\authorfooter{
%% insert punctuation at end of each item
\item
 Otto-v.-Guericke-University Magdeburg
}

%other entries to be set up for journal
\shortauthortitle{Klemm \MakeLowercase{\textit{et al.}}: Interactive Visual Analytics of Image-Centric Cohort Study Data}
%\shortauthortitle{Firstauthor \MakeLowercase{\textit{et al.}}: Paper Title}

%% Abstract section.
\abstract{Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease specific risk factors.
%
Cohort studies comprise heterogenous data variables describing the medical condition as well as demographic and lifestyle factors of a subject.
%
Using well established statistical methods the data is hypothesis driven analyzed to find statistically significant variable correlations ('\emph{interactions}').
%
Modern cohort studies also incorporate medical image data.
%
Analyzing these data requires image segmentation, extraction of key figures and shape based subject grouping.
\\
We propose a Interactive Visual Analytics approach that enables epidemiologists to examine both image-based as well as sociodemographic and medical attribute data.
%We propose a Interactive Visual Analytics approach to provide epidemiologists with the tools necessary to analyze this data.
%
It allows for both classical hypothesis validation approaches as well as hypothesis generation by incorporating data mining methods.
%
Adaptive linked information visualization views and 3d-shape renderings are combined with epidemiological techniques.
%
Similarity measures between data variables are used to compute interesting changes in variable interactions for the current variable selection.
%
Shape based grouping of subjects is facilitated using hierarchical agglomerative clustering.
} % end of abstract

%% Keywords that describe your work. Will show as 'Index Terms' in journal
%% please capitalize first letter and insert punctuation after last keyword
\keywords{Interactive Visual Analytics, Epidemiology}

%% ACM Computing Classification System (CCS). 
%% See <http://www.acm.org/class/1998/> for details.
%% The ``\CCScat'' command takes four arguments.

\CCScatlist{ % not used in journal version
 \CCScat{K.6.1}{Management of Computing and Information Systems}%
{Project and People Management}{Life Cycle};
 \CCScat{K.7.m}{The Computing Profession}{Miscellaneous}{Ethics}
}

%% Uncomment below to include a teaser figure.
  % \teaser{
  % \centering
  % \includegraphics[width=16cm]{CypressView}
  % \caption{In the Clouds: Vancouver from Cypress Mountain.}
  % }

%% Uncomment below to disable the manuscript note
%\renewcommand{\manuscriptnotetxt}{}

%% Copyright space is enabled by default as required by guidelines.
%% It is disabled by the 'review' option or via the following command:
% \nocopyrightspace

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%% START OF THE PAPER %%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%% The ``\maketitle'' command must be the first command after the
%% ``begin{document}'' command. It prepares and prints the title block.

%% the only exception to this rule is the \firstsection command
\firstsection{Introduction}

\maketitle

%% \section{Introduction} %for journal use above \firstsection{..} instead

My Text

\section{Medical and Technical Background}

\section{Prior and Related Work}

\section{Interactive Visual Analytics in Cohort Study Data}

\section{Application}

\section{Summary and Conclusion}

%% if specified like this the section will be committed in review mode
\acknowledgments{SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no. 03ZIK012), the Ministry of Cultural Affairs as well as the Social Ministry of the Federal State of Mecklenburg-West Pomerania. Whole-body MR imaging was supported by a joint grant from Siemens Healthcare, Erlangen, Germany and the Federal State of Mecklenburg-Vorpommern. The University of Greifswald is a member of the ‘Centre of Knowledge Interchange’ program of the Siemens AG. This work was supported by the DFG Priority Program 1335: Scalable Visual Analytics.}

\bibliographystyle{abbrv}
%%use following if all content of bibtex file should be shown
%\nocite{*}
\bibliography{bibliography}
\end{document}

### 2014-03-04 draft.tex
improved Structure

@@
 %% indicate IEEE Member or Student Member in form indicated below
-\author{Paul Klemm}
+\author{TBA}
 \authorfooter{
@@
 
-My Text
+\section{Medical and Technical Background}
+\textbf{Image aquisition}\\
+- standardization of MR Scanners
 
-\section{Medical and Technical Background}
+\subsection{Epidemiological Workflow}
+Bernhards Paper
+
+\subsection{The Spine Dataset}
 
 \section{Prior and Related Work}
+- VMV Paper\\
+- Analysis of data in the Rotterdam study\\
+- VA in Epidemiology Part of Bernhards Paper
+	- Commercial visual Analytics Systems
 
 \section{Interactive Visual Analytics in Cohort Study Data}
+
+According to Steffens Terminology in Interactive Visual Analysis of Perfusion Data
+\\\\
+\textbf{Object Space}\\
+- Medical Image Data / Spine Segmentations\\
+\textbf{Attribute Space}\\
+- SHIP-Variables\\
+- Derived visualizations
+
+? Where to incorporate Levels of IVA
+
+\subsection{Feature Localization}
+- Projection from Image Space to Attribute data. This is more difficult in our application, because we currently only brush on derived features.\\
+- Clustering in Image Space yields Groups which can be analyzed using the attribute space
+
+\subsection{Local Investigation}
+- Selection of Information using Bar Charts, Scatterplots or Parallel Coordinates which projects the selection into the Object space\\
+- this selection is for categorical data already given implicitly bei projecting the 3D-View onto the Bar Charts/Mosaik Plots!\\
+- this aims to locale features of the data!\\
+- Gain Information about SHIP-Variables by putting them into the context of each other.
+The Pivot Table allows for direct numerical\\ analysis, while the information visualizations allow for better insight of the combination
+
+\subsection{Multivariate Analysis}
+- Selection of Elements in Scatterplots, Barcharts or Parallel Coordinates Views - observe how selection changes another view - this allows for multivariate analysis\\
+- Becker, R.A., Cleveland, W.S.: Brushing scatterplots. Technometrics 29(2) (1987)\\
+- Wang Baldonado, M.Q., Woodruff, A., Kuchinsky, A.: Guidelines for using multiple views in information visualization.
+In: AVI ’00: Proceedings of the working conference on Ad- vanced visual interfaces, pp.
+110–119.
+ACM Press, New York, NY, USA (2000).
+DOI\\ http://doi.acm.org/10.1145/345513.345271
+- By brushing individual parameters or create new binnings of parameter it is possible to see how they change in coordinated views. This is already implemented in the Cargo framework\\
+- by creating comparative 3d Visualizations it is possible to assess the influence of non-image parameters to the visual space.
 

### 2014-03-05 draft.tex
Created Structure and first chapters

@@
 
-\title{Interactive Visual Analytics of Image-Centric Cohort Study Data}
+\title{Interactive Visual Analysis of Image-Centric Cohort Study Data}
 
@@
 %
-Shape based grouping of subjects is facilitated using hierarchical agglomerative clustering.
+Shape based grouping of subjects is facilitated using hierarchical agglomerative clustering. (Remove Clustering reference?)
 } % end of abstract
@@
 %% \section{Introduction} %for journal use above \firstsection{..} instead
+Epidemiology aims to characterize health and disease by determining risk factors.
+%
+Clinical problems and questions answered using epidemiological methods comprise diagnosis accuracy, disease frequency, risk factors, disease prognosis, effectiveness of treatments or preventions and cause of diseases \cite{Fletcher2012}.
+%
+%Cohort studies are a epidemiological method to gather data about a group of subjects (a \emph{cohort)}) to make general statements about these problems.
+%
+Observations made by clinicians in the daily routine are translated into hypothesis.
+%
+These are used to determine environmental and lifestyle factors are as well as medical attributes which are believed to influence a condition of interest.
+%
+The data variables necessary are gathered using interviews and clinical examinations.
+%
+Statistical methods like regression analysis aim aim to check the attribute list for plausibility.
+%
+
+Longitudinal population-based studies like the Study of Health in Pomerania \cite{Volzke2011} aim to gather as much information as possible about a defined sample of people (a \emph{cohort}).
+%
+The sample is drawn randomized to avoid selection bias prohibit statements based on statistical correlations in the cohort to be inferred to the whole population.
+%
+Also a information bias needs to be avoided by strictly standardizing the data acquisition.
+%
+Statistical correlations are also prone to confounding, meaning that two factors influence each other and therefore should be normalized with respect to each other.
+%
+When for example one investigates risk factors for prostate cancer in male subjects the outcome is strongly dependent on the age.
+%
+Therefore results need to be age adjusted to be comparable.
+%
+Confounding variables, however, are often not obvious at all and characterizing them is already an epidemiological result.
+\\\\
+Modern cohort studies often include medical image data which introduces new problems.
+%
+Since it is unethical to expose people to radiation, non-harming imaging like Magnetic Resonance Imaging (MRI) or Ultrasound Imaging is used.
+%
+As MRI scans are expensive there exists a tradeoff between quality of the image data and their associated costs.
+%
+To quantify these data it is necessary to segment it.
+%
+Manual segmentation via radiological experts is possible but very costly and prone to inter- and intra observer variability.
+%
+Segmentation algorithms allow for (semi)-automated analysis of the data but require sophisticated methods due to high inter-subject variability caused by the subject diversity.
+%
+Analyzing spatial data with other epidemiological factors require techniques which reach beyond standard statistical methods.
+\\
+We propose a Interactive Visual Analysis approach to provide a way to analyze both image- and non-image data.
+%
+Visual queries and direct feedback of Visual Analytics systems allow for a fast exploration of the data space.
+%
+Intended as a extension to the well established epidemiological tools it provides a way to rapidly validate hypothesis as well as trigger hypothesis generation using Data Mining methods such as clustering.
+%
+
+%Interactive Visual Analysis (IVA) TODO QUELLE provides techniques which not only show promising potential for combining both 3d visualization with
+
+Our contributions are:
+\begin{itemize}
+	\item Applying the Interactive Visual Analysis approach to the epidemiological problem domain by characterizing special affordances of this context.
+	\item Provide an overview over the workflow for analyzing cohort study data to gain insigth into the large subject spaces.
+	\item Provide Visualization Techniques which combine both Information Visualization and 3D Rendering of Organ Shapes as well as combining them with well known epidemiological graphics and key figures.
+	\item Implement the presented methods in a Web Framework based on WebGL, D3js and Nodejs as backend.
+\end{itemize}
+
+%Statistical correlations derived from analyzing the data itself may be misleading since
+
+%Include Implementation details - web based; how are image data included; what technologies are used?
 
 \section{Medical and Technical Background}
-\textbf{Image aquisition}\\
-- standardization of MR Scanners
+In this section we want to give insight into the epidemiological workflow when analyzing cohort study data to identify the problems we address in this paper.
+%
+ToDo Define Epidemiological Outcome
+%
+\subsection{Epidemiological Workflow}
+Epidemiologists follow a strict workflow mainly driven by statistic tools to validate hypothesis.
+%
+Following Thew and colleagues publication on this matter, the workflow can be characterized as follows.
+%
+Hypothesis most commonly base on observations made by clinicians in their daily routine.
+%
+A set of attributes depicting conditions affected by the hypothesis is compiled accordingly.
+%
+Confounding variables need to be adjusted so that they do not affect the effect size of a attribute.
+%
+Statistical methods such as regression analysis are applied to measure the effect size of attributes to the outcome of interest.
+%
 
-\subsection{Epidemiological Workflow}
-Bernhards Paper
+Reproducibility of results is an epidemiological key requirement.
+%
+Longitudinal epidemiological studies require the acquired attributes to be comparable to evaluate them.
+%
+If the data acquisition process changes, a information bias is introduced to the data, disallowing inference between acquisition cycles.
+%
+This underlines the high quality standards to methods processing the data, whether to extract additional parameters or gain insight.
+%
+To determine, whether a subject is prone to be affected by a certain disease, relative risks a expressed through the evaluation of p values which indicate statistical significance.
+%
+Statistics tools such as SPSS and STATA play a major role for analyzing epidemiolgical data. 
+%
+Graphic data representation is largely used to show results rather than gaining insight.
 
-\subsection{The Spine Dataset}
+Bernhards Paper\\
+- Missing Data\\
+- Follow-Up Questions?\\
+- grouping essentiell\\
+- problems when analyzing image data\\
+	- what are Problems there\\
+	- how can shape be included?\\
+	- show image data of multiple subjects
+
+\subsection{Epidemiological Data}
+Benrhard Paper
+- sociodemographic data
+- medical data
+- pain indicators
+- DATA TYPES
+	- dichotomous data
+
+\paragraph{Image acquisition}: standardization of MR Scanners
+
+\subsection{The Study of Health in Pomerania (SHIP)}
 
@@
 - Analysis of data in the Rotterdam study\\
-- VA in Epidemiology Part of Bernhards Paper
-	- Commercial visual Analytics Systems
+- VA in Epidemiology Part of Bernhards Paper\\
+	- Commercial visual Analytics Systems\\
+- GPloms\\
+Unterlagen Wissenschaftliche Projekt
 
 \section{Interactive Visual Analytics in Cohort Study Data}
+\begin{figure}[htb]
+ \centering
+ %\includegraphics[width=1.5in]{figures/InteractionLoop}
+ \includegraphics[width=3.0in]{figures/InteractionLoop}
+ \caption{Interaction Overview}
+\end{figure}
 
@@
 - Projection from Image Space to Attribute data. This is more difficult in our application, because we currently only brush on derived features.\\
-- Clustering in Image Space yields Groups which can be analyzed using the attribute space
+- Clustering in Image Space yields Groups which can be analyzed using the attribute space\\
+- Create a Pipeline Overview over different Levels of different IVA Patterns and Stages
 
@@
 
+\subsection{Implementation}
+Diagram of used Technologies
+
 \section{Application}
+
+\subsection{The Spine Dataset}
+- Describe steps from gathering Information from the raw image files (segmentation, abstraction, visualization)\\
+- Input of Epidemiologists goes here!
 

### 2014-03-06 draft.tex
Added Parts of Data Chapter

@@
 %
+\textbf{Characterization of the healthy aging process! Which change indicates a unusual pathological change?}
 
@@
 Graphic data representation is largely used to show results rather than gaining insight.
+\\
+Group subjects using epidemiological factors is essential in order to make statements about their statitical power.
+%
+Grouping is carried out hypothesis driven.
+%
+Age for example is also divided into groups (\emph{binned}) when investigating its influence on a condition.
+%
+These groups depend strongly on the condition of interest and therefore there is no defined standard on how to categorize these values.
 
-Bernhards Paper\\
+\subsection{Epidemiological Data}
+Epidemiological data is highly heterogenous.
+%
+Information about medical history and examinations, genetic conditions, geographical data, questionaire answers, and image data yield complex data spaces for each subject.
+%
+Often data are derived from acquired variables to either group or threshold values or get information derived from reviewed data such as breast density data for women.
+%
+This underlines also the problem of missing data since for ethical, legal or medical reasons some data variables can not be gathered for each subject.
+%
+Follow-up examinations or -questions for conditions also produce variables only available for a small amount of subjects.
+%
+\\
+Indicators for medical conditions as well as questions about a subjects lifestyle are also often \emph{dichotomous}, meaning that they only have two manifestations (often \emph{Yes} or \emph{No}).
+%
+This allows for the calculation of \emph{odds ratios} which describe the relation of two \emph{dichotomous} variables, allowing for direct comparison of their influences.
+%
+Dichotomous data can also be derived by combining aggregating data variables to yield only two manifestations (e.g. subjects younger or older than 50).
+
+Bernhard Paper\\
+- sociodemographic data\\
+- medical data\\
+- pain indicators\\
+- DATA TYPES\\
+	- dichotomous data\\
+\\
 - Missing Data\\
 - Follow-Up Questions?\\
-- grouping essentiell\\
+- grouping essential\\
 - problems when analyzing image data\\
@@
 	- show image data of multiple subjects
-
-\subsection{Epidemiological Data}
-Benrhard Paper
-- sociodemographic data
-- medical data
-- pain indicators
-- DATA TYPES
-	- dichotomous data
 

### 2014-03-07 draft.tex
Added workflow image and wrote Epidemiological Data Chapter

@@
 \section{Medical and Technical Background}
+Wer ist an epidemiologischen Studien beteiligt?\\
+•  Ärzte (Facharzt für öffentliches Gesundheitswesen, Gene@ker)\\
+•  Medizinische Informa@ker mit Fokus auf Biometrie und Sta@s@k\\
+•  Bei klinischen Studien: Ärzte des entsprechenden Fachs
+
 In this section we want to give insight into the epidemiological workflow when analyzing cohort study data to identify the problems we address in this paper.
@@
 %
-\\
+
 Indicators for medical conditions as well as questions about a subjects lifestyle are also often \emph{dichotomous}, meaning that they only have two manifestations (often \emph{Yes} or \emph{No}).
@@
 Dichotomous data can also be derived by combining aggregating data variables to yield only two manifestations (e.g. subjects younger or older than 50).
+%
 
-Bernhard Paper\\
-- sociodemographic data\\
-- medical data\\
-- pain indicators\\
-- DATA TYPES\\
-	- dichotomous data\\
-\\
-- Missing Data\\
-- Follow-Up Questions?\\
-- grouping essential\\
-- problems when analyzing image data\\
-	- what are Problems there\\
-	- how can shape be included?\\
-	- show image data of multiple subjects
+\paragraph{Image acquisition.} Imaging techniques emitting an hazardous amount of radiation for the subject are not suited for ethical reasons.
+%
+MRI data is more expensive to obtain as CT data but does not affect the subjects health and is therefore the main method for collecting cohort study imaging data.
+%
+The quality of medical image data acquired for cohort studies is a tradeoff between accuracy and affordability \cite{Preim2014}.
+%
+This often yields image resolutions inferior to those of clinical day-to-day practice, which makes their analysis more challenging.
 
-\paragraph{Image acquisition}: standardization of MR Scanners
+\paragraph{Image analysis.} When analyzing medical image data there have decisions to be made on how they are \emph{compared} and \emph{quantified}.
+%
+Segmentations masks describing all parts of the shape of interest would be ideal since many different key figures can be derived from them.
+%
+Since these masks require sophisticated algorithms custom tailored to the data sets the epidemiologists are forced to measure the data by hand, which is a very tedious work with respect to the number of necessary landmarks and number of subjects.
+%
+Information derived by landmarks are also not nearly as expressive and versatile as segmentation masks.
+%
+They are also prone to a high inter-observer variability and hard to reproduce.
+%
+This gains even more momentum when analyzing multiple time steps!
+%
+Morphometric information from landmarks comprises thickness, diameter or length of a structure as well as grey-value distribution in an area (used for determine type of tissue).
+
+% Bernhard Paper\\
+% - sociodemographic data\\
+% - medical data\\
+% - pain indicators\\
+% - DATA TYPES\\
+% 	- dichotomous data\\
+% \\
+% - Missing Data\\
+% - Follow-Up Questions?\\
+% - grouping essential\\
+% - problems when analyzing image data\\
+% 	- what are Problems there\\
+% 	- how can shape be included?\\
+% 	- show image data of multiple subjects
 
 \subsection{The Study of Health in Pomerania (SHIP)}
+Starting 1997 with a cohort consisting of 4.308 subjects this cohort study located in northern germany aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
+%
+Data is collected independent of diseases in mind.
+%
+This allows the data set to be queried regarding many different diseases and conditions.
+%
+Subjects were examined in a 5-year time span, continuously adding new parameters including MRI scans in the last iteration of 2012.
+%
+The MRI protocol features a rich number of different sequences.
+%
+Also for women, breast MRI scans were acquired.
+%
+A second cohort \texttt{SHIP-Trend} was established in 2008 to acquire data about a younger population.
+%
+The protocols for analyzing the subjects between \texttt{SHIP} and \texttt{SHIP-Trend} remained the same, making them comparable.
+%
+The overall examination time for each person attending the study is two days.
 
@@
 \section{Interactive Visual Analytics in Cohort Study Data}
+\begin{figure}[htb]
+ \centering
+ %\includegraphics[width=1.5in]{figures/InteractionLoop}
+ \includegraphics[width=3.0in]{figures/workflow_comparison}
+ \caption{Workflow Comparison}
+\end{figure}
 \begin{figure}[htb]
@@
 
-\subsection{Feature Localization}
+\subsection{Feature Selection}
+- Automatic using Clustering\\
+- Automatic using Statistical Recommendations\\
+- Via User Input
+
+\subsection{Brushing the information space} 
+- derive new groups which serve as input for the feature selection
+
+\subsubsection{Feature Localization}
 - Projection from Image Space to Attribute data. This is more difficult in our application, because we currently only brush on derived features.\\
@@
 
-\subsection{Local Investigation}
+\subsubsection{Local Investigation}
 - Selection of Information using Bar Charts, Scatterplots or Parallel Coordinates which projects the selection into the Object space\\
@@
 
-\subsection{Multivariate Analysis}
+\subsubsection{Multivariate Analysis}
 - Selection of Elements in Scatterplots, Barcharts or Parallel Coordinates Views - observe how selection changes another view - this allows for multivariate analysis\\
@@
 - Describe steps from gathering Information from the raw image files (segmentation, abstraction, visualization)\\
-- Input of Epidemiologists goes here!
+- Input of Epidemiologists goes here!\\
+\\\\
+\textbf{From VMV'13 Paper}\\
+
+All whole-body MRI scans were acquired on a 1.5~Tesla scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany) by four trained technicians in a standardized way.
+Subjects were placed in the supine position.
+Five phased-array surface coils were placed to the head, neck, abdomen, pelvis, and lower extremities for whole-body imaging.
+The spine coil is embedded in the patient table.
+The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence (676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels).
+First, both sequences were placed over the cervical and upper thoracic spine.
+Then, they were placed over the lower thoracic and lumbar spine.
+The MRI software automatically composed a whole spine sequence from the two T1-weighted and T2-weighted sequences \cite{Hegenscheid2013}.
+We were provided with 490 data sets.
+
+The model is placed in the scene using an empirically chosen initialization point.
+The force acting on the model stems from aggregation of loads, which are derived from a potential field resulting from a weighted sum of the T1- and T2-weighted MRI images, see \cite{FEM2013}.
+After detecting all spines, we register the models because in a later clustering step we only want to capture the local deformation of the lumbar spine, not different locations in world space.
+The models are registered using the Kabsch Algorithm \cite{Kabsch1976}, which is designed to minimize the root mean squared deviation between paired sets of points.
+The model-based detection captures information about the spine canal curvature as well as the alignment of the vertebrae.
+It is not meant to capture information about vertebrae deformation and differences in spine canal extent.
 

### 2014-03-10 draft.tex
finished Related Work, wrote Introduction of IVA Chapter

@@
 %
-The sample is drawn randomized to avoid selection bias prohibit statements based on statistical correlations in the cohort to be inferred to the whole population.
+The sample is drawn randomized to avoid selection bias prohibit statements based on statistical correlations in the cohort to be inferred to the whole population (\emph{external validity}) \cite{Fletcher2012}.
 %
@@
 \\
-We propose a Interactive Visual Analysis approach to provide a way to analyze both image- and non-image data.
+We propose a Interactive Visual Analysis approach \cite{Thomas2005} to provide a way to analyze both image- and non-image data.
 %
@@
 %
-\textbf{Characterization of the healthy aging process! Which change indicates a unusual pathological change?}
+\textbf{ToDo Characterization of the healthy aging process! Which change indicates a unusual pathological change?}
 
@@
 
-\section{Medical and Technical Background}
+\section{Medical and Technical Background} \label{MedicalAndTechnicalBackground}
 Wer ist an epidemiologischen Studien beteiligt?\\
@@
 %
-\subsection{Epidemiological Workflow}
+\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
 Epidemiologists follow a strict workflow mainly driven by statistic tools to validate hypothesis.
@@
 
-\subsection{Epidemiological Data}
+\subsection{Epidemiological Data} \label{EpidemiologicalData}
 Epidemiological data is highly heterogenous.
@@
 
-\paragraph{Image acquisition.} Imaging techniques emitting an hazardous amount of radiation for the subject are not suited for ethical reasons.
+\paragraph{Image acquisition.} \label{ImageAcquisition} Imaging techniques emitting an hazardous amount of radiation for the subject are not suited for ethical reasons.
 %
@@
 
-\paragraph{Image analysis.} When analyzing medical image data there have decisions to be made on how they are \emph{compared} and \emph{quantified}.
+\paragraph{Image analysis.} \label{ImageAnalysis} When analyzing medical image data there have decisions to be made on how they are \emph{compared} and \emph{quantified}.
 %
@@
 \subsection{The Study of Health in Pomerania (SHIP)}
-Starting 1997 with a cohort consisting of 4.308 subjects this cohort study located in northern germany aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
+Starting 1997 with a cohort consisting of 4.308 subjects this cohort study located in northern Germany aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
 %
@@
 \section{Prior and Related Work}
-- VMV Paper\\
-- Analysis of data in the Rotterdam study\\
-- VA in Epidemiology Part of Bernhards Paper\\
-	- Commercial visual Analytics Systems\\
-- GPloms\\
-Unterlagen Wissenschaftliche Projekt
+Einfuehren von Helwigs Terminologie?
+
+Finding a visualization which fits all needs and communicates all aspects of the data equally is challenging.
+%
+Given the number of features of epidemiological data sets and their different manifestations, it is often a good solution to combine the strength of different visualization techniques in a unified system \cite{Buja91, Konyha2009}
+%
+Data mining tools like the Principal Component Analysis are able to reduce the dimension by extracting most expressive components, but making the influence of each variable hard to determine.
+\\\\
+The work of Turkay and colleagues is closest to ours albeit our focus on processing medical image data and variables with categorical manifestations \cite{Turkay2013}.
+%
+Investigating Data on an norwegian aging study their methods aim to amplify a hypothesis generation process when analyzing the data.
+%
+Statistical measures of metric variables such as mean, standard deviation, skewness, or inter-quartile range are used to create \emph{dimension plots}.
+%
+These transform dimensions into data points and make them comparable with respect to the derived descriptive measures.
+%
+This not only allows for comparing all continuous variables in a single plot but make their distribution change comprehensible.
+%
+This of course requires a good descriptive measure which captures the kind of change the user is interested in or which reflects unexpected data behavior.
+%
+The technique was applied to variables generated by segmenting the the brain into 45 parts and measure the number of voxel, volume and properties of the intensity values.
+%
+The method is strongly dependent on the descriptive derived measures of the epidemiological factors.
+%
+Hypothesis based on observations of changes in these plots may impose over-fitting to the data because the measure highlights certain statistical changes.
+%
+Our approach sticks more to the information extracted from the segmented image data and derive variable interaction with non-image epidemiological factors.
+% Disadvantage - Nur 2 abgeleitete Variablen sichtbar, schlecht mit anderen Variablen in Verbindung zu bekommen (müsste jedoch auch gehen) - viele Dinge gehen durch die Abstraktion aber auch flöten!
+
+Gresh and colleagues suggest with \texttt{WEAVE} one of the first systems which analyzed medical image and non-image data using linked views \cite{Gresh2000}.
+%
+Blaas and colleagues presented a similar system which analyzed medical image data and variables derived from them using views from the feature- and physical space \cite{Blaas2007}.
+%
+This approach already incorporated Data Mining approaches like dividing the data space using a k-nearest-neighbor technique and Principal Component Analysis.
+%
+Steenwijk and colleagues employ a relational database to organize the data to visualize subject data using linked views like parallel coordinates, scatterplots and time plots \cite{Steenwijk2010}.
+%
+Zhang and colleagues provide a web-based system for analyzing subject groups with linked views and batch-processing capabilities for categorizing new subject entries into the data set \cite{Zhang2012}.
+%
+Their understanding of a cohort differs from the understanding of the term in an epidemiolgical context.
+
+Commercial systems like Tableau or Spotfire provide a rich user interface that allows to apply Visual Analytics techniques without the need of writing any code.
+%
+With little efford linked views can be created using these tools, but the data procesinng possibilities like derivation of new variables or the volume rendering capabilities are very limited.
+%
+These systems share limitations in extensibility to a specific problem domain.
+%Spotfire offers the possibilities to interact with the \texttt{R} statistical programming language
+
+% *********************** Our own Work ***********************
+Klemm and colleagues used captured lumbar spine variabilities based on an semi-automatic shape-detection algorithm of 490 participants of the \texttt{SHIP-2} \cite{Klemm2013VMV}.
+%
+Hierarchical agglomerative clustering was used to divide the population into shape-related groups.
+%
+As proof of concept a relation between size of the segmented shape and measured size of the subjects was measured and behaved as expected.
+%
+This work focuses on incorporating these data as new features of the overall data set, making it possible to include it into the hypothesis validation and generation process.
+%
+When applying clustering techniques on the non-image data is was found that \texttt{k-Prototypes} and \texttt{DBSCAN} is appropriate in the epidemiological context but is strongly dependent on the chosen variables and distance measure \cite{Klemm2014BVM}.
+% *********************** / Our own Work ***********************
+
+Generalized Pairs Plot (\emph{GPLOMS}) is an information visualization technique that allows for heterogenous variables to be pairwise visualized using appropriate plots in a plot matrix grouped by type.
+%
+This technique is also useful to gain an overview over numerous variables and their distributions.
+%
+It uses histograms, bar charts, scatterplots and heat maps to visualize the different variable combinations with regard to their type.
+%
+Brushing capabilities allow for brushing and linking as well as filtering, but has limitations like making only one category brushable at a time.
+%
+We applied this technique to our data and it shows promising potential for simultaneously visualizing many different variable but does not fit in the scope of this paper.
+%
+A similar approach was taken by Dai and colleagues for risk factor exploration as they also incorporate choropleth maps of epidemiological factors (e.g.
+mortality rates in a region) with parallel coordinates, bar charts and scatterplots integrated regression lines \cite{Dai2005}.
+%
+From their findings regarding the interaction of cancer-related socio-demographic factors are drawn in a \emph{Concept Map} where related factors are connected via graph-edges.
+
+Chui and colleagues visualized interactions in time-dependant epidemiological data using time-series plots highlighting risk factors differences in age and gender \cite{Chui2011}.
+%
+\\\\
+% Image Data
+Comparing tissue between many subjects in an epidemiological context requires methods which allow for shape variance visualizations.
+%
+Caban and colleagues investigated the suitability of variance visualizations of shape distribution models and concluded in their user study that user favor spherical glyph representations over deformation grids and likelihood volumes \cite{Caban2011}.
+%
+The distribution of shapes in a space derived from a principal component analysis is plotted by Busking and Colleagues in a 2D-projected plane of the space \cite{Busking2010a}.
+%
+We incorporate the idea of combining 3D-Shape rendering with information visualization techniques.
+%
+Differences between structures are highlighted using color mapping of the difference to the mean shape, but is rather hard to recognize due to small renderings of each subject in the shape-space.
+%
+Via mesh morphing interpolated views can be created by the user in a separate view as well as comparisons in contour-view.
+%
+Applying our data sets to this technique yielded a cluttered shape space due to the many subjects.
+%
+The data needs to abstracted or summarized in order to work in this context.
+%
+In order to detect local deformation changes, Hermann and colleagues investigating shape related difference by letting the user specify a deformation of interest and showing corresponding changes in the shape using covariance tensors \cite{Hermann2014}.
+%
+This method allowed for rapid hypothesis validation and was able to reproduce textbook knowledge.
+%
+By plotting p-values in ventricle surfaces, Chou and colleagues were able to map disease-associated values directly on a 3D tissue representations \cite{Chou2009}.
+%
+This requires a geographic colocation of associated features.
+%
+
+% Interactive Visual Analysis - Steffens Paper
+The strength of the Interactive Visual Analysis approach described in the next section is it's versatility with respect to the application field \cite{Konyha2009}.
+%
+Oeltze and colleagues combined a linked view representation of results from a statistical analysis with feature localizations of the human blood flow with the goal of its evaluation \cite{Oeltze2007}.
+%
+%They aim to provide methods to analyze perfusion data with respect to 
+While we take similar steps when analyzing the data like employing statistical tests, our data is mostly independent from the medical image data and is not describing it--except the variables derived the segmentation model itself.
 
@@
  \centering
+ \label{fig:WorkflowComparison}
  %\includegraphics[width=1.5in]{figures/InteractionLoop}
  \includegraphics[width=3.0in]{figures/workflow_comparison}
- \caption{Workflow Comparison}
+ \caption{Visual Analytics systems are able to complement parts of the epidemiological workflow, not to replace it.
+The appropriate combination of statistical- and interactive driven analysis shows promising potential to unveil the information in the data.}
 \end{figure}
@@
 \end{figure}
+
+As described in subsection~\ref{EpidemiologicalWorkflow}, the epidemiolgical workflow is a strict sequence of steps taken by domain experts and needs to be reproducible and comprise statistical integrity.
+%
+Figure~\ref{fig:WorkflowComparison} (a) describes this workflow as steps consecutive series of steps.
+%
+The workflow we propose by introducing the \emph{IVA} principle into the epidemiological problem domain does not aim to replace the existing workflow but to complement its weaknesses 
+%
+In the current state the workflow handles the data as a black box.
+%
+A list of features describing the hypothesis is compiled and analyzed using statistical tests. 
+%
+The resulting value decides whether the data supports the hypothesis or not.
+%
+It would be possible that there actually are features of the data set which support the hypothesis by discriminating the population in the expected way, but with this approach it is not able to find those.
+%
+This becomes even more imminent when the workflow is adapted to the analysis of the medical image data.
+%
+Domain experts annotate tediously landmarks which allow to derive metrics like distances which are then handled like other features and analyzed using the same set of statistical tools.
+%
+Not only does this leave out the majority of the information in the medical image data by abstracting it to metrics, it is easily possible that information left out would heavily influence the result.
+%
+Considering more complex parts of the data would make those results more trustworthy and also could identify possible anatomical confounders--an epidemiolgical research result in itself.
+%
+Statistical tests check for validity of the number but not for their completeness or plausibility!
+\\\\
+IVA tries to illuminate the data black box by making the domain experts part of the feature list selection process.
+%
+Figure~\ref{fig:WorkflowComparison} (b) highlights the iterative process as part of the epidemiolgical workflow.
+%
+Note that the process also aims to project back into the hypothesis formulation step to amplify hypothesis generation 
+%
+This has to be handled with care since overfitting of expectations to the available data is an imminent danger as described by Turkay and colleagues \cite{Turkay2013}.
 
@@
 
-The model is placed in the scene using an empirically chosen initialization point.
-The force acting on the model stems from aggregation of loads, which are derived from a potential field resulting from a weighted sum of the T1- and T2-weighted MRI images, see \cite{FEM2013}.
+%The model is placed in the scene using an empirically chosen initialization point.
+The force acting on the model stems from aggregation of loads, which are derived from a potential field resulting from a weighted sum of the T1- and T2-weighted MRI images, see \cite{Rak2013}.
 After detecting all spines, we register the models because in a later clustering step we only want to capture the local deformation of the lumbar spine, not different locations in world space.
 The models are registered using the Kabsch Algorithm \cite{Kabsch1976}, which is designed to minimize the root mean squared deviation between paired sets of points.
+The model is placed in the scene using an empirically chosen initialization point.
+The force acting on the model stems from aggregation of loads, which are derived from a potential field resulting from a weighted sum of the T1- and T2-weighted MRI images, see \cite{Rak2013}.
+After detecting all spines, we register the models because in a later clustering step we only want to capture the local deformation of the lumbar spine, not different locations in world space.
+The models are registered using the Kabsch Algorithm, which is designed to minimize the root mean squared deviation between paired sets of points.
 The model-based detection captures information about the spine canal curvature as well as the alignment of the vertebrae.
@@
 %% if specified like this the section will be committed in review mode
-\acknowledgments{SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no.
+%\acknowledgments{SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no.
 03ZIK012), the Ministry of Cultural Affairs as well as the Social Ministry of the Federal State of Mecklenburg-West Pomerania.

### 2014-03-11 draft.tex
wrote IVA chapter and made ugly preview of implementation figure

@@
 %% insert punctuation at end of each item
-\item
- Otto-v.-Guericke-University Magdeburg
+% \item
+%  Otto-v.-Guericke-University Magdeburg
 }
@@
  \caption{Visual Analytics systems are able to complement parts of the epidemiological workflow, not to replace it.
-The appropriate combination of statistical- and interactive driven analysis shows promising potential to unveil the information in the data.}
+The appropriate combination of statistical- and interactive driven analysis shows promising potential to unveil the information in the data.
+(a) shows the standard epidemiological workflow, (b) the IVA supported one.
+The iterative red highlighted part is called the \emph{IVA Loop}.}
 \end{figure}
@@
  \includegraphics[width=3.0in]{figures/InteractionLoop}
- \caption{Interaction Overview}
+ \caption{Detailed Version of the \emph{IVA Loop}.
+Usually starting with a selection of Variable of interest (user-driven or via data mining techniques), the data is mapped using a visualization technique appropriate for the selected data types.
+The data is visualized in the independent and dependent domain space, which can then be brushed, yielding new groups which can be investigated using further variables.
+Note that adjacent steps are directly connected via feedback loops allowing for iterative refinement and give as much freedom to the user as possible.}
 \end{figure}
@@
 \\\\
-\textbf{Object Space}\\
-- Medical Image Data / Spine Segmentations\\
-\textbf{Attribute Space}\\
-- SHIP-Variables\\
-- Derived visualizations
 
-? Where to incorporate Levels of IVA
+\subsection{Image Centric Cohort Study Data in Interactive Visual Analytics Context}
+In the IVA context, data is divided to belong into two major view types.
+%
+The human body exposing shape information for the \emph{physical view}, also referred as the \emph{independent variables} \cite{Oeltze2013}.
+%
+This information space is usually displayed via volume rendering techniques \cite{Oeltze2007}.
+%
+%Spatiotemporal views include data from the \emph{Object Space Domain} renders data centered to the focused domain, in our example the medical image data space.
+%
+\emph{Dependent variables} in the epidemiological context can be divided twofold:
+\begin{itemize}
+	\item Variables derived from the image data. 
+	%
+	These measures abstracts shape information as quantification to allow for comparison. 
+	%
+	These variables describe image data and can also be used to brush in the image space.
+	\item Epidemiological socio-demographic or medical attribute data.
+	%
+	These values belong to every subject which is represented in the image space, but does not describe shape information.
+	%
+	This is the data epidemiologists usually want to correlate with image data.
+\end{itemize}
+%Other data associated to elements in the object space are considered to be the \emph{Attribute Space Domain} and are displayed using various range based views using statistical graphics and information visualization techniques like scatterplots, bar charts or parallel coordinates \cite{Oeltze2013, Oeltze2007}.
+%
+%Specific for the epidemiological application domain is that both Object- and Feature Space Domain influence the outcome of interest.
+% \textbf{Object Space}\\
+% - Medical Image Data / Spine Segmentations\\
+% \textbf{Attribute Space}\\
+% - SHIP-Variables\\
+% - Derived visualizations
+% 
+% ? Where to incorporate Levels of IVA
 
-\subsection{Feature Selection}
-- Automatic using Clustering\\
-- Automatic using Statistical Recommendations\\
-- Via User Input
+\subsection{Data Preprocessing}
+To include heterogenous epidemiological data in an IVA framework it is necessary to process it to obtain standardized views to the available features.
+%
+Due to the different acquisition modalities there have to be different techniques incorporated.
+%
+Data obtained using questionnaires or medical tests are often stored using statistical packages like \texttt{SPSS} or \texttt{Stata} which have a proprietary data format with limited export capabilities.
+%
+We found the best and easiest solution was to simply export the data in the respective tool to a character separated text file and then convert it to data types which are easier manageable like JSON or XML using our own classes.
+%
+In order to verify that the conversion worked as expected and the data is valid it is good practice to use data wrangling tools like \texttt{OpenRefine} to validate the data.
+%
+Exporting the data dictionary which stores information about each manifestation of a feature is also an important step to get a detailed description of data variables and the meaning and unit of measurement of their values.
+%
+Since the reasons for missing data have a wide range from ethical to medical and personal reasons, these are also included as error codes which have to marked as such in the data dictionary.
+\\\\
+Processing the image data associated to each subject is for the most part the extraction of information about a structure of interest.
+%
+This is either done manually by experts setting land marks (sometimes supported by algorithms connecting the land marks like Graph cuts) or by a (semi)-automatic detection, registration and segmentation.
+%
+Algorithms applied to the data do not only have to deal with a large inter-subject variability of the structure of interest but also needs to be reproducible \cite{Preim2014}.
+%
+Model based approaches have shown to be effective for this task \cite{Gloger2010, Gloger2012, Rak2013}.
+%
+If a segmentation yields only binary masks separating the structures, algorithms like the growing and adaptive shape can be applied creating a surface grid where each point is comparable throughout the population \cite{Ferrarini2007}.
+%
+Intensity based comparison can be achieved using rigid image registration, but model based results however are preferable \cite{Klemm2012}.
+%
+Comparison based on grey values is usually carried out to to measuring the quantity of fat, water or iron (liver) or distribution of grey and white brain tissue.
 
-\subsection{Brushing the information space} 
-- derive new groups which serve as input for the feature selection
+Morphometric variables are derived to allow for statistical comparison of the tissue which incorporate mostly position, volume and relative distances and alignment to other structures.
+% 
+% - Data Format\\
+% - Missing Data\\
+% - Registration of Image Data (VMV'12)\\
+% - Extraction of Image Parameter
+
+\subsection{IVA Patterns}
+%ToDo Groups are Variables as well!\\
+
+The explorative procedures when analyzing data using IVA can be divided into three different patterns, handling interaction between domain and range variables.
+
+\subsubsection{Local Investigation}
+This pattern projects information from image space to the range perspective.
+%
+As opposing to other IVA application domains this step is more complicated in the epidemiological context.
+%
+Shape information can not be brushed by incorporating ROI-selections but rather has to employ techniques that specify local deformation changes \cite{Hermann2014} or subjects that belong to a shape class.
+%
+Methods available for \emph{feature selection} strongly depend on the type of registration that was applied to extract the tissue of interest.
+%
+Model based segmentations or masks yield data structures capable of calculating mean shapes and distances between individuals or subject groups.
+%
+Feature selection is also possible by applying clustering algorithms in order to get shape-groups \cite{Klemm2013}.
+%
+These algorithms can be used to investigate interactions between shape-groups and other non-image based variables.
+%
+Another application is the outlier analysis.
+%
+Outliers can indicate segmentation errors or an outstanding group of individuals who may share a pathology.
+
+% Local Investigation
+% - Projection from Image Space to Attribute data. This is more difficult in our application, because we currently only brush on derived features.\\
+% - Clustering in Image Space yields Groups which can be analyzed using the attribute space\\
+% - Create a Pipeline Overview over different Levels of different IVA Patterns and Stages
+
+% \subsection{Feature Selection}
+
+% \subsection{Brushing the information space}
+% - derive new groups which serve as input for the feature selection
 
 \subsubsection{Feature Localization}
-- Projection from Image Space to Attribute data. This is more difficult in our application, because we currently only brush on derived features.\\
-- Clustering in Image Space yields Groups which can be analyzed using the attribute space\\
-- Create a Pipeline Overview over different Levels of different IVA Patterns and Stages
+% Subset selection of range variables yields features
+As described before, the vast majority of data points are considered to be dependent with respect to the image domain in the IVA context.
+%
+Selecting subjects based on image derived data can be seen as additional possibility of shape-related grouping.
+%
+For the most part the epidemiologist is interested in the shape of subjects within a range of a set of variables that describe the current hypothesis.
+%
+Epidemiologists are used to categorize data into groups that fit their hypothesis formulation.
+%
+Continuous variables like age are for example often divided in categories like young, aged and elderly.
+%
+Categorization is strongly dependent on the hypothesis and therefore requires suitable brushing techniques as described in section TODO.
+%
+%To gain insight into interactions between SHIP-
 
-\subsubsection{Local Investigation}
-- Selection of Information using Bar Charts, Scatterplots or Parallel Coordinates which projects the selection into the Object space\\
-- this selection is for categorical data already given implicitly bei projecting the 3D-View onto the Bar Charts/Mosaik Plots!\\
-- this aims to locale features of the data!\\
-- Gain Information about SHIP-Variables by putting them into the context of each other.
+% - Selection of Information using Bar Charts, Scatterplots or Parallel Coordinates which projects the selection into the Object space\\
+% - this selection is for categorical data already given implicitly bei projecting the 3D-View onto the Bar Charts/Mosaik Plots!\\
+% - this aims to locale features of the data!\\
+% - Gain Information about SHIP-Variables by putting them into the context of each other.
 The Pivot Table allows for direct numerical\\ analysis, while the information visualizations allow for better insight of the combination
@@
 \subsubsection{Multivariate Analysis}
-- Selection of Elements in Scatterplots, Barcharts or Parallel Coordinates Views - observe how selection changes another view - this allows for multivariate analysis\\
-- Becker, R.A., Cleveland, W.S.: Brushing scatterplots. Technometrics 29(2) (1987)\\
-- Wang Baldonado, M.Q., Woodruff, A., Kuchinsky, A.: Guidelines for using multiple views in information visualization.
+Introduced in the information visualization, the mutlivariate analysis means selecting subjects in like in the feature localization step and get have the result highlighted in another linked view displaying other non-image parameter.
+%
+This represents for the most part the usual approach which essentially is the analysis of how variables interact with each other, only in a visual analysis context.
+%
+Special for the application domain is the need of statistic measures which describe how variables correlate with each other given the selected groups.
+%
+These association also summarized using Pivot Tables which are popular in epidemiology and which are described in the following section.
+% - Selection of Elements in Scatterplots, Barcharts or Parallel Coordinates Views - observe how selection changes another view - this allows for multivariate analysis\\
+% - Becker, R.A., Cleveland, W.S.: Brushing scatterplots. Technometrics 29(2) (1987)\\
+% - Wang Baldonado, M.Q., Woodruff, A., Kuchinsky, A.: Guidelines for using multiple views in information visualization.
 In: AVI ’00: Proceedings of the working conference on Ad- vanced visual interfaces, pp.
@@
 DOI\\ http://doi.acm.org/10.1145/345513.345271
-- By brushing individual parameters or create new binnings of parameter it is possible to see how they change in coordinated views. This is already implemented in the Cargo framework\\
-- by creating comparative 3d Visualizations it is possible to assess the influence of non-image parameters to the visual space.
+% - By brushing individual parameters or create new binnings of parameter it is possible to see how they change in coordinated views. This is already implemented in the Cargo framework\\
+% - by creating comparative 3d Visualizations it is possible to assess the influence of non-image parameters to the visual space.
+
+\section{Visualization Techniques}
+\subsection{combined views}
+\subsection{Adaptive information visualization views}
+- Focus on intuitive Brushing!
+\subsection{Pivot Tables}
+\subsection{Interaction Workflow}
+- Automatic Suggestion of Variables
 
@@
 \section{Application}
+\begin{figure}[htb]
+ \centering
+ \label{fig:technologies}
+ %\includegraphics[width=1.5in]{figures/InteractionLoop}
+ \includegraphics[width=3.5in]{figures/technologies}
+ \caption{Bla}
+\end{figure}
 

### 2014-03-13 draft.tex
added application part, interaction part and new placeholder

@@
 
-Generalized Pairs Plot (\emph{GPLOMS}) is an information visualization technique that allows for heterogenous variables to be pairwise visualized using appropriate plots in a plot matrix grouped by type.
+Generalized Pairs Plot (\emph{GPLOMS}) is an information visualization technique that allows for heterogenous variables to be pairwise visualized using appropriate plots in a plot matrix grouped by type \cite{Francois2013}.
 %
@@
 %
-Feature selection is also possible by applying clustering algorithms in order to get shape-groups \cite{Klemm2013}.
+Feature selection is also possible by applying clustering algorithms in order to get shape-groups \cite{Klemm2013VMV}.
 %
@@
 
-\section{Visualization Techniques}
-\subsection{combined views}
-\subsection{Adaptive information visualization views}
-- Focus on intuitive Brushing!
+\section{Interaction- and Visualization Techniques}
+\begin{figure}[htb]
+ \centering
+ \label{fig:visualization}
+ %\includegraphics[width=1.5in]{figures/InteractionLoop}
+ \includegraphics[width=3.5in]{figures/visualization}
+ \caption{\emph{Image is not final} Screenshot from the front-end which is divided as follows: (a) Sidebar which contains all features as well as the groups defined in the analysis process; (b) Canvas area where features can be added via drag and drop and the visualization chosen automatically according to the data type; (c) Interactive Pivot Table which shows exact numbers for each displayed variable combination.
+The data displayed is used to analyze the lumbar spine.
+ }
+\end{figure}
+IVA levels defines different interaction levels. 
+%
+We employ the fourth and highest levels which includes next to brushing and linking, advanced brushing, extraction of attributes the development of visualizations custom tailored to the data sets.
+%
+The employed techniques needs to respect the epidemiological affordances.
+%
+Hypothesis generation bears the chance of over-fitting the data to expectations.
+%
+To avoid this, a timeline needs to be introduced which keeps track how many variable variations the user evaluated before coming to an conclusion.
+%
+Since epidemiologists are used to process groups based on table representations we decided to introduce an interactive solution in form of a Pivot Table.
+
+\subsection{Structure and Workflow}
+We divide the workspace into three major parts as seen in Figure~\ref{fig:visualization}.
+%
+The sidebar holds all available variables as well as groups either derived by user input or automatic clustering.
+
+From the sidebar, elements of interest can be put into the canvas area via drag and drop.
+%
+Doing so will automatically create a information visualization suitable for the current data type.
+%
+3D shape information about the investigated structure is displayed for each variable manifestation.
+%
+By dropping variables on existing visualizations the system creates a visualization that allows for comparison (e.g. mosaic plots for ordinal variables).
+%
+Elements can be brushed in the information visualizations and are linked to the other representations in the canvas.
+%
+Shape based clustering can either be applied to all subjects or subgroups.
+
+All elements in the canvas view are also represented in the interactive Pivot Table which gives detailed information how the subjects are distributed given the displayed variables.
+%
+Details to the different views are presented in the following sub-sections.
+
+\subsection{Sidebar}
+A overview of all variables is presented in a sidebar where they are categorized to different types like somatometric, disease- or lifestyle related, pain indicators and laboratory data.
+%
+It also contains subject groups either defined by user brushing or by automated clustering.
+%
+Groups are treated exactly like other variables since they work exactly the same way which is dividing the subject space into different labeled categories.
+% TODO Implement
+Bar charts show the distribution of manifestations of each variable in the sidebar.
+
+\subsection{Adaptive Feature Visualization}
+Inspired be the previously discussed \emph{GPloms} \cite{Francois2013} the visualization type is chosen dynamically based on the variable types and number of visualizations which need to be displayed.
+% TODO Implement MOSAIC PLOTS, Scatterplots, Parallel Coordinates
+If possible the medical image data is directly included into the plot as well by including mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
+%
+The 3D-plots can be navigated using standard mouse inputs and the camera is synced so that a direct comparison is given.
+%
+If a feature is dropped on an existing plot, the visualization changes dynamically to properly make them comparable. 
+%
+Each plot can be brushed using widgets.
+%
+It is able to duplicate brushes in order to create new groups which are evenly spread out.
+%
+A use case for this is when a continuous feature has to be divided into even groups.
+
 \subsection{Pivot Tables}
-\subsection{Interaction Workflow}
-- Automatic Suggestion of Variables
+When opening a random epidemiological paper the reader will in almost all cases find some sort of pivot table to present the data.
+%
+As seen in Figure~\ref{fig:visualization} (c) it is a good way to display how many subjects are in each group.
+%
+Pivot table quickly get confusing and cluttered when they are divided into to many subgroups.
+%
+We tackled this problem by making the order and number of displayed variables adaptable.
+%
+This also applies on the designation of Row or Column Variables.
+% TODO IMPLEMENT
+The mean shape for each resulting sub-group is also displayed for each subject.
+\\\\
+In the following sections we will discuss details on the implementation which relies on modern Web-Technologies.
 
 \subsection{Implementation}
-Diagram of used Technologies
-
-\section{Application}
 \begin{figure}[htb]
@@
  \includegraphics[width=3.5in]{figures/technologies}
- \caption{Bla}
+ \caption{\emph{Image is not final}. The front-end solution (left) uses state of the art web technologies like HTML5/CSS3, WebGL and SVG to display the data.
+ % 
+ The NodeJS based back-end (right) stores all image- and non-image data and transfers it to connected clients.
+ %
+ All computation heavy operations like calculation of mean-shapes or -distances as well as statistical processing is done by the server to keep hardware requirements of client systems low. 
+ %
+ Client-Server communication is accomplished via the WebSocket protocol.
+ }
 \end{figure}
+%
+In order to provide a fast communication loop between method development and expert input we decided to base all implementations on modern web technologies which employs various advantages.
+\begin{itemize}
+	\item No additional software needs to be installed, most people use decent state of the art web browser. If developed using proper standards, the applications even run on mobile devices. 
+	\item The Client-Server structure allows it to employ heavy computation on a server machine and transfer results to the client
+	\item Since Image data for several thousand subjects claims hundreds of Gigabytes disk space it can remain safely on the server and elements can be transferred on demand.
+	%
+	High confidentiality standards of the data can be met by restricting access via a account system
+	\item Recent developments in WebGL applications running in browsers with near native performance push the development into to the web which results in many open source libraries which are well documented, rich in examples and driven by active communities.
+\end{itemize}
+These advantages do not come without drawbacks.
+%
+Many methods which specialized libraries like the Visualization Toolkit (VTK) or R for statistics have build in need to be written from scratch in order to fit in the context.
+%
 
+The back-end is written using Node, which based on the Google V8 Javascript runtime environment.
+%
+Due to its event-driven non-blocking I/O model it is fast and does not freeze in case of heavy workload like mesh calculation.
+\\\\
+Non-image data for all Subjects including the data dictionary is stored on the Server in a JSON file.
+%
+Image data is available as raw MRI DICOM files as well as segmented Meshes which can be used to compare subjects.
+%
+On client connection the requested files are transferred.
+%
+The server processes calculation heavy statistical tasks like calculation of Odds Ratios or Chi Square tests for all variable combinations in order to keep the computation time on the client as low as possible.
+%
+
+The front-end is created using Twitter Bootstrap as foundation for the layout and basic UI elements using HTML5, CSS3 and Javascript.
+%
+Information Visualizations like Scatterplots and bar charts are created using the popular Data Driven Documents library which works well for attaching data to visible elements like vector graphics.
+%
+WebGL rendering is done using the Threejs which allows GPU Accelerated data rendering.
+%
+Communication between Client and Server runs through the WebSockets protocol.
+%
+Since our clustering algorithms are written in MatLab we hat to access them using the Node Server.
+%
+We accomplished this by converting them to parameterized standalone console applications which are spawned by node on client request and then reads the result from the console standard-out and returns it in a proper format to the client.
+%
+All parameter steered console applications can be incorporated in this context.
+
+\section{Application}
 \subsection{The Spine Dataset}
 - Describe steps from gathering Information from the raw image files (segmentation, abstraction, visualization)\\
+- Problem of sparse differences - Visualization has to be more abstract to emphasize the differences.\\
 - Input of Epidemiologists goes here!\\

### 2014-03-14 draft.tex
First correction

@@
 %
-Cohort studies comprise heterogenous data variables describing the medical condition as well as demographic and lifestyle factors of a subject.
+Cohort studies comprise heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors.
 %
-Using well established statistical methods the data is hypothesis driven analyzed to find statistically significant variable correlations ('\emph{interactions}').
+Using well established statistical methods the data is hypothesis driven analyzed to find statistically significant variable correlations (\emph{interactions}).
 %
@@
 Analyzing these data requires image segmentation, extraction of key figures and shape based subject grouping.
-\\
-We propose a Interactive Visual Analytics approach that enables epidemiologists to examine both image-based as well as sociodemographic and medical attribute data.
-%We propose a Interactive Visual Analytics approach to provide epidemiologists with the tools necessary to analyze this data.
+\\\\
+We propose a Interactive Visual Analysis approach that enables epidemiologists to examine both image-based as well as sociodemographic and medical attribute data.
 %
-It allows for both classical hypothesis validation approaches as well as hypothesis generation by incorporating data mining methods.
+It allows for both hypothesis validation approaches as well as hypothesis generation by incorporating data mining methods.
 %
@@
 %
-Shape based grouping of subjects is facilitated using hierarchical agglomerative clustering. (Remove Clustering reference?)
+Shape based grouping of subjects is facilitated using clustering techniques which operate on surface meshes extracted from the image segmentation.
 } % end of abstract
@@
 %
-These are used to determine environmental and lifestyle factors are as well as medical attributes which are believed to influence a condition of interest.
+These are used to determine environmental and lifestyle factors are as well as medical attributes which may influence a condition of interest.
 %
-The data variables necessary are gathered using interviews and clinical examinations.
+The data variables necessary are gathered using structured interviews and clinical examinations.
 %
@@
 %
-The sample is drawn randomized to avoid selection bias prohibit statements based on statistical correlations in the cohort to be inferred to the whole population (\emph{external validity}) \cite{Fletcher2012}.
+The sample is drawn randomized to avoid selection bias which prohibits statements based on statistical correlations in the cohort from being inferred to the whole population (\emph{external validity}) \cite{Fletcher2012}.
 %
@@
 %
-Statistical correlations are also prone to confounding, meaning that two factors influence each other and therefore should be normalized with respect to each other.
+Statistical correlations are also prone to \emph{confounding}, meaning that two factors are dependent and therefore should be normalized with respect to each other.
 %
-When for example one investigates risk factors for prostate cancer in male subjects the outcome is strongly dependent on the age.
+When for example one investigates risk factors for prostate cancer in male subjects, the outcome is strongly dependent on the age.
 %
@@
 %
-Confounding variables, however, are often not obvious at all and characterizing them is already an epidemiological result.
+Confounding variables, are often not obvious at all and characterizing them is already an epidemiological result.
 \\\\
@@
 %
-To quantify these data it is necessary to segment it.
+To quantify these data it is necessary to label each voxel regarding structure affiliation (\emph{segmentation}).
 %
-Manual segmentation via radiological experts is possible but very costly and prone to inter- and intra observer variability.
+Manual segmentation carried out by radiological experts is possible but very costly and prone to inter- and intra observer variability.
 %
@@
 %
-Analyzing spatial data with other epidemiological factors require techniques which reach beyond standard statistical methods.
-\\
+Analyzing spatial data with respect to other epidemiological factors requires techniques which reach beyond standard statistical methods.
+\\\\
 We propose a Interactive Visual Analysis approach \cite{Thomas2005} to provide a way to analyze both image- and non-image data.
@@
 %
-\textbf{ToDo Characterization of the healthy aging process! Which change indicates a unusual pathological change?}
-
-%Interactive Visual Analysis (IVA) TODO QUELLE provides techniques which not only show promising potential for combining both 3d visualization with
-
+In order to characterize the healthy aging process we aim to determine changes for subjects which indicate unusual pathological changes.
+\\\\
 Our contributions are:
 \begin{itemize}
-	\item Applying the Interactive Visual Analysis approach to the epidemiological problem domain by characterizing special affordances of this context.
-	\item Provide an overview over the workflow for analyzing cohort study data to gain insigth into the large subject spaces.
-	\item Provide Visualization Techniques which combine both Information Visualization and 3D Rendering of Organ Shapes as well as combining them with well known epidemiological graphics and key figures.
-	\item Implement the presented methods in a Web Framework based on WebGL, D3js and Nodejs as backend.
+	\item Applying the Interactive Visual Analysis technique set to the epidemiological problem domain by characterizing affordances of this context.
+	\item Provide an overview over the workflow for analyzing cohort study data to gain insight into the large diverse subject spaces.
+	\item Provide visualization techniques which combine both information visualization and 3D rendering of organ shapes as well as combining them with well known epidemiological graphics and key figures.
+	\item Implement the presented methods as a web framework based on WebGL, D3JS and NodeJS.
 \end{itemize}
@@
 \section{Medical and Technical Background} \label{MedicalAndTechnicalBackground}
-Wer ist an epidemiologischen Studien beteiligt?\\
-•  Ärzte (Facharzt für öffentliches Gesundheitswesen, Gene@ker)\\
-•  Medizinische Informa@ker mit Fokus auf Biometrie und Sta@s@k\\
-•  Bei klinischen Studien: Ärzte des entsprechenden Fachs
+% Wer ist an epidemiologischen Studien beteiligt?\\
+% •  Ärzte (Facharzt für öffentliches Gesundheitswesen, Gene@ker)\\
+% •  Medizinische Informa@ker mit Fokus auf Biometrie und Sta@s@k\\
+% •  Bei klinischen Studien: Ärzte des entsprechenden Fachs
 
@@
 %
-ToDo Define Epidemiological Outcome
+% TODO Define Epidemiological Outcome - 5Ds:
+% Death
+% Disease
+% Discomfort
+% Disability
+% Dissatisfaction
 %
 \subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
-Epidemiologists follow a strict workflow mainly driven by statistic tools to validate hypothesis.
+% TODO Define who participates in epidemiological studies ?
+Since it is a diverse science, many different experts work at epidemiological studies, ranging from specialized doctors, medical computer scientists with focus on biometrics to statisticians.
+%
+Epidemiologists follow a strict workflow mainly driven by statistic tools to validate hypothesis about disease specific risk factors.
 %
@@
 %
-Hypothesis most commonly base on observations made by clinicians in their daily routine.
+\begin{itemize}
+	\item Hypothesis most commonly base on observations made by clinicians in their daily routine.
 %
-A set of attributes depicting conditions affected by the hypothesis is compiled accordingly.
+	\item A set of attributes depicting conditions affected by the hypothesis is compiled accordingly.
 %
-Confounding variables need to be adjusted so that they do not affect the effect size of a attribute.
+	\item Confounding variables need to be adjusted so that they do not affect the effect size of a attribute.
 %
-Statistical methods such as regression analysis are applied to measure the effect size of attributes to the outcome of interest.
+	\item Statistical methods such as regression analysis are applied to measure the effect size of attributes to the outcome of interest.
+\end{itemize}
+The workflow is shown in Figure~\ref{fig:WorkflowComparison} (a).
+
+Reproducibility of results is an key requirement.
 %
-
-Reproducibility of results is an epidemiological key requirement.
-%
-Longitudinal epidemiological studies require the acquired attributes to be comparable to evaluate them.
+Longitudinal studies require the acquired attributes to be comparable to evaluate them.
 %
@@
 %
-To determine, whether a subject is prone to be affected by a certain disease, relative risks a expressed through the evaluation of p values which indicate statistical significance.
+To determine, whether a subject is prone to be affected by a certain disease, relative risks are expressed through the evaluation of p-values which indicate statistical significance.
 %
-Statistics tools such as SPSS and STATA play a major role for analyzing epidemiolgical data. 
+Statistics tools such as SPSS and STATA play a major role for analyzing epidemiological data. 
 %
-Graphic data representation is largely used to show results rather than gaining insight.
-\\
-Group subjects using epidemiological factors is essential in order to make statements about their statitical power.
+Graphic data representation is largely used to present results rather than gaining insight.
+
+Grouping subjects using epidemiological factors is essential in order to allow per-group risk determination.
 %
@@
 %
-Age for example is also divided into groups (\emph{binned}) when investigating its influence on a condition.
+Age for example is also divided into groups (e.g. in 20 year-steps) when investigating its influence on a condition.
 %
@@
 %
-Information about medical history and examinations, genetic conditions, geographical data, questionaire answers, and image data yield complex data spaces for each subject.
+Information about medical history and examinations, genetic conditions, geographical data, questionnaire results and image data yields a complex data space for each subject.
 %
@@
 %
-MRI data is more expensive to obtain as CT data but does not affect the subjects health and is therefore the main method for collecting cohort study imaging data.
+MRI data is more expensive to obtain than CT data but does not affect the subjects health and is therefore the main method for collecting cohort study imaging data.
 %
-The quality of medical image data acquired for cohort studies is a tradeoff between accuracy and affordability \cite{Preim2014}.
+The image quality is a tradeoff between accuracy and affordability \cite{Preim2014}.
 %
@@
 
-\paragraph{Image analysis.} \label{ImageAnalysis} When analyzing medical image data there have decisions to be made on how they are \emph{compared} and \emph{quantified}.
+\paragraph{Image analysis.} \label{ImageAnalysis} Decisions have to be made on how image data are \emph{compared} and \emph{quantified}.
 %
-Segmentations masks describing all parts of the shape of interest would be ideal since many different key figures can be derived from them.
+Segmentations masks labeling the voxel of a shape of interest would be ideal since many different key figures can be derived from them.
 %
@@
 %
-Data is collected independent of diseases in mind.
+Data is collected diseases-independent.
 %
@@
 
-Finding a visualization which fits all needs and communicates all aspects of the data equally is challenging.
+Designing a visualization which communicates all aspects of the data equally is challenging.
 %
-Given the number of features of epidemiological data sets and their different manifestations, it is often a good solution to combine the strength of different visualization techniques in a unified system \cite{Buja91, Konyha2009}
+Given the number of features of epidemiological data sets and their different manifestations, it is often a good solution to combine the strength of different visualization techniques in a unified system \cite{Buja91, Konyha2009}.
 %
-Data mining tools like the Principal Component Analysis are able to reduce the dimension by extracting most expressive components, but making the influence of each variable hard to determine.
+Data mining tools like the Principal Component Analysis are able to reduce the dimension by extracting most expressive components, but make the influence of each variable hard to determine.
 \\\\
@@
 %
-Investigating Data on an norwegian aging study their methods aim to amplify a hypothesis generation process when analyzing the data.
+Investigating Data on an norwegian aging study their methods aim to amplify a hypothesis generation process.
 %
@@
 %
-This of course requires a good descriptive measure which captures the kind of change the user is interested in or which reflects unexpected data behavior.
+This requires a good descriptive measure which captures the kind of change the user is interested in or which reflects unexpected data behavior.
 %
-The technique was applied to variables generated by segmenting the the brain into 45 parts and measure the number of voxel, volume and properties of the intensity values.
+The technique was applied to variables generated by segmenting the brain into 45 parts and measure the voxel number, volume and properties of the intensity values.
 %
-The method is strongly dependent on the descriptive derived measures of the epidemiological factors.
+The method is strongly dependent on the descriptive measures of the epidemiological factors.
 %
-Hypothesis based on observations of changes in these plots may impose over-fitting to the data because the measure highlights certain statistical changes.
+Hypothesis based on observations of changes in these plots may impose over-fitting to the data because the measure highlights only subsets of statistical changes.
 %
@@
 
-Gresh and colleagues suggest with \texttt{WEAVE} one of the first systems which analyzed medical image and non-image data using linked views \cite{Gresh2000}.
+Gresh and colleagues proposed \texttt{WEAVE}, one of the first systems which analyzed medical image and non-image data using linked views \cite{Gresh2000}.
 %
@@
 
-Commercial systems like Tableau or Spotfire provide a rich user interface that allows to apply Visual Analytics techniques without the need of writing any code.
+Commercial systems like \texttt{Tableau} or \texttt{Spotfire} provide a rich user interface that allow to apply Visual Analytics techniques without the need of writing any code.
 %
-With little efford linked views can be created using these tools, but the data procesinng possibilities like derivation of new variables or the volume rendering capabilities are very limited.
+With little effort, linked views can be created using these tools, but the data processing possibilities like derivation of new variables or the volume rendering capabilities are very limited.
 %
@@
 % *********************** Our own Work ***********************
-Klemm and colleagues used captured lumbar spine variabilities based on an semi-automatic shape-detection algorithm of 490 participants of the \texttt{SHIP-2} \cite{Klemm2013VMV}.
+Klemm and colleagues used lumbar spine variabilities based on an semi-automatic shape-detection algorithm of 490 participants of the \texttt{SHIP-2} \cite{Klemm2013VMV}.
 %
-Hierarchical agglomerative clustering was used to divide the population into shape-related groups.
+Hierarchical agglomerative clustering divided the population into shape-related groups.
 %
@@
 
-Generalized Pairs Plot (\emph{GPLOMS}) is an information visualization technique that allows for heterogenous variables to be pairwise visualized using appropriate plots in a plot matrix grouped by type \cite{Francois2013}.
+Generalized Pairs Plots (\texttt{GPLOM´S}) are an information visualization technique that allows for heterogenous variables to be pairwise compared using appropriate plots in a plot matrix grouped by type \cite{Francois2013}.
 %
@@
 %
+The inspiration on the chosen visualization techniques stems from this publication.
+%
 A similar approach was taken by Dai and colleagues for risk factor exploration as they also incorporate choropleth maps of epidemiological factors (e.g.
-mortality rates in a region) with parallel coordinates, bar charts and scatterplots integrated regression lines \cite{Dai2005}.
+mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines \cite{Dai2005}.
 %
@@
  \label{fig:WorkflowComparison}
- %\includegraphics[width=1.5in]{figures/InteractionLoop}
  \includegraphics[width=3.0in]{figures/workflow_comparison}
- \caption{Visual Analytics systems are able to complement parts of the epidemiological workflow, not to replace it.
+ \caption{ToDo Abbildung noch nicht final.
+Visual Analytics systems are able to complement parts of the epidemiological workflow, not replace it.
 The appropriate combination of statistical- and interactive driven analysis shows promising potential to unveil the information in the data.
@@
 \end{figure}
-\begin{figure}[htb]
- \centering
- %\includegraphics[width=1.5in]{figures/InteractionLoop}
- \includegraphics[width=3.0in]{figures/InteractionLoop}
- \caption{Detailed Version of the \emph{IVA Loop}.
-Usually starting with a selection of Variable of interest (user-driven or via data mining techniques), the data is mapped using a visualization technique appropriate for the selected data types.
-The data is visualized in the independent and dependent domain space, which can then be brushed, yielding new groups which can be investigated using further variables.
-Note that adjacent steps are directly connected via feedback loops allowing for iterative refinement and give as much freedom to the user as possible.}
-\end{figure}
-
+%
 As described in subsection~\ref{EpidemiologicalWorkflow}, the epidemiolgical workflow is a strict sequence of steps taken by domain experts and needs to be reproducible and comprise statistical integrity.
 %
-Figure~\ref{fig:WorkflowComparison} (a) describes this workflow as steps consecutive series of steps.
+Figure~\ref{fig:WorkflowComparison} (a) describes this workflow as consecutive series of steps.
 %
-The workflow we propose by introducing the \emph{IVA} principle into the epidemiological problem domain does not aim to replace the existing workflow but to complement its weaknesses 
+The workflow we propose by introducing the \emph{IVA} principle into the epidemiological application domain does not aim to replace the existing workflow but to complement its weaknesses.
 %
-In the current state the workflow handles the data as a black box.
+In the current state the workflow treats the data like a black box.
 %
@@
 %
-It would be possible that there actually are features of the data set which support the hypothesis by discriminating the population in the expected way, but with this approach it is not able to find those.
+It would be possible that there actually are features of the data set which support the hypothesis by discriminating the population in the expected way, but with this approach they are not highlighted in any way.
 %
@@
 %
-Not only does this leave out the majority of the information in the medical image data by abstracting it to metrics, it is easily possible that information left out would heavily influence the result.
+Not only does this leave out the majority of the information in the medical image data by abstracting it to single values, it is easily possible that information left out would heavily influence the result.
 %
@@
 \\\\
-IVA tries to illuminate the data black box by making the domain experts part of the feature list selection process.
+\emph{IVA} tries to illuminate the black box by making the domain experts part of the feature list selection process.
 %
@@
 %
-Note that the process also aims to project back into the hypothesis formulation step to amplify hypothesis generation 
+Note that it also aims to project back into the hypothesis formulation step to amplify hypothesis generation 
 %
@@
 
-According to Steffens Terminology in Interactive Visual Analysis of Perfusion Data
-\\\\
-
 \subsection{Image Centric Cohort Study Data in Interactive Visual Analytics Context}
-In the IVA context, data is divided to belong into two major view types.
+In the \emph{IVA} context, data is divided into two major view types.
 %
-The human body exposing shape information for the \emph{physical view}, also referred as the \emph{independent variables} \cite{Oeltze2013}.
+The human body exposing shape information for the \emph{physical view}, also referred as the \emph{independent variable} \cite{Oeltze2013}.
 %
@@
 \subsection{Data Preprocessing}
-To include heterogenous epidemiological data in an IVA framework it is necessary to process it to obtain standardized views to the available features.
+To include heterogenous epidemiological data in an \emph{IVA}-framework it is necessary to process it to obtain standardized views to the available features.
 %
@@
 %
-We found the best and easiest solution was to simply export the data in the respective tool to a character separated text file and then convert it to data types which are easier manageable like JSON or XML using our own classes.
+The best solution for us was to simply export the data in the respective tool to a character separated text file and then convert it to data types which are easier manageable like JSON or XML using our own classes.
 %
@@
 %
-Since the reasons for missing data have a wide range from ethical to medical and personal reasons, these are also included as error codes which have to marked as such in the data dictionary.
+Since the reasons for missing data have a wide range from ethical to medical and personal issues, these are also included as error codes which have to marked as such in the data dictionary.
 \\\\
-Processing the image data associated to each subject is for the most part the extraction of information about a structure of interest.
+Processing the image data associated to each subject consists for the most part of information extraction about a structure of interest.
 %
-This is either done manually by experts setting land marks (sometimes supported by algorithms connecting the land marks like Graph cuts) or by a (semi)-automatic detection, registration and segmentation.
+This is either done manually by experts setting land marks (sometimes supported by algorithms connecting the land marks like Graph-Cuts) or by a (semi)-automatic detection, registration and segmentation.
 %
@@
 %
-If a segmentation yields only binary masks separating the structures, algorithms like the growing and adaptive shape can be applied creating a surface grid where each point is comparable throughout the population \cite{Ferrarini2007}.
+If a segmentation yields only binary masks separating the structures, algorithms like Growing and Adaptive Shapes can be applied creating a surface grid where each point is comparable throughout the population \cite{Ferrarini2007}.
 %
@@
 %
-Comparison based on grey values is usually carried out to to measuring the quantity of fat, water or iron (liver) or distribution of grey and white brain tissue.
+Comparison based on grey values is usually carried out to to measuring the quantity of fat, water, and--application specific--iron content (liver) or distribution of grey and white brain tissue.
 
@@
 \subsection{IVA Patterns}
-%ToDo Groups are Variables as well!\\
 
-The explorative procedures when analyzing data using IVA can be divided into three different patterns, handling interaction between domain and range variables.
+The explorative procedures when analyzing data using \emph{IVA} can be divided into three different patterns, handling interaction between domain and range variables.
 
@@
 %
-As opposing to other IVA application domains this step is more complicated in the epidemiological context.
+As opposing to other \emph{IVA} application domains this step is more complicated in the epidemiological context.
 %
@@
 % Subset selection of range variables yields features
-As described before, the vast majority of data points are considered to be dependent with respect to the image domain in the IVA context.
+As described before, the vast majority of data points are considered to be dependent with respect to the image domain in the \emph{IVA} context.
 %
@@
 %
-Categorization is strongly dependent on the hypothesis and therefore requires suitable brushing techniques as described in section TODO.
+Categorization is strongly dependent on the hypothesis and therefore requires suitable brushing techniques as described in section~\ref{sec:AdaptiveFeatureVisualization}.
 %
@@
 \subsubsection{Multivariate Analysis}
-Introduced in the information visualization, the mutlivariate analysis means selecting subjects in like in the feature localization step and get have the result highlighted in another linked view displaying other non-image parameter.
+%Introduced in the information visualization, the multivariate analysis means selecting subjects in like in the feature localization step and get have the result highlighted in another linked view displaying other non-image parameter.
+Introduced in the information visualization, the multivariate analysis incorporates brushing and linking of of views displaying non-image parameter.
 %
-This represents for the most part the usual approach which essentially is the analysis of how variables interact with each other, only in a visual analysis context.
+%This represents for the most part the usual approach which essentially is the analysis of how variables interact with each other, only in a visual analysis context.
 %
@@
 \section{Interaction- and Visualization Techniques}
+%
+We employ the fourth and highest level which includes next to brushing and linking, advanced brushing, extraction of attributes the development of visualizations custom tailored to the data sets.
 \begin{figure}[htb]
  \centering
- \label{fig:visualization}
  %\includegraphics[width=1.5in]{figures/InteractionLoop}
@@
  }
+ \label{fig:visualization}
 \end{figure}
-IVA levels defines different interaction levels. 
-%
-We employ the fourth and highest levels which includes next to brushing and linking, advanced brushing, extraction of attributes the development of visualizations custom tailored to the data sets.
+\emph{IVA} levels define different levels interaction. 
 %
@@
 We divide the workspace into three major parts as seen in Figure~\ref{fig:visualization}.
+%
+\begin{figure}[htb]
+ \centering
+ \label{fig:InteractionLoop}
+ \includegraphics[width=3.0in]{figures/InteractionLoop}
+ \caption{Detailed Version of the \emph{IVA Loop}.
+Usually starting with a selection of Variable of interest (user-driven or via data mining techniques), the data is mapped using a visualization technique appropriate for the selected data types.
+The data is visualized in the independent and dependent domain space, which can then be brushed, yielding new groups which can be investigated using further variables.
+Note that adjacent steps are directly connected via feedback loops allowing for iterative refinement and give as much freedom to the user as possible.}
+\end{figure}
+TODO auf Fig~\ref{fig:InteractionLoop} zurückkommen
 %
@@
 
+ToDo\\
+- wie werden Confounder gefunden?\\
+- welche statistischen Kennzahlen werden eingebunden?\\
+
 \subsection{Sidebar}
@@
 
-\subsection{Adaptive Feature Visualization}
+\subsection{Adaptive Feature Visualization} \label{sec:AdaptiveFeatureVisualization}
 Inspired be the previously discussed \emph{GPloms} \cite{Francois2013} the visualization type is chosen dynamically based on the variable types and number of visualizations which need to be displayed.
@@
 A use case for this is when a continuous feature has to be divided into even groups.
+\\
+ToDo\\
+- Dichotomous data\\
+- Time-Line\\
+- Statistical Analysis (Odds ratios)
 
@@
 The mean shape for each resulting sub-group is also displayed for each subject.
+
+\subsection{Automated Feature Suggestion}
+As discussed previously, highlighting potential interesting values in the data set is one major benefit of the \emph{IVA} powered approach.
+%
+Turkay and colleagues used the approach to calculate various key figures based in the distribution functions of each feature derived from the image data \cite{Turkay2013}.
+%
+Since the majority of our data are categorical features, we have to employ different solutions.
+%
+A solution for this problem are odds ratios, which are a standard statistical tool for stating relations between features.
+%
+Odds ratios can only be calculated for $2\times2$ contingency tables which usually represent the presence of a condition in a population divided by a characteristic (e.g.
+male/female subjects with presence or absence of back pain).
+%
+To calculate odds ratios for variables with more than two manifestations, we calculate local odds ratios for each possible combination, yielding a matrix for each feature combination \cite{Rudas1998}.
+%
+When the subjects are divided into groups and the calculation is carried out again for all feature combination, the difference in sum of the odds weighted with the number of manifestations per feature can be used to indicate if the feature combination yields a difference.
+%
+These difference are then highlighted in a separate tab of the side bar "Interactions".
+%
+Interesting interactions then can be assessed creating a linked view using the standard drag and drop workflow.
 \\\\
 In the following sections we will discuss details on the implementation which relies on modern Web-Technologies.
+\\\\
+ToDo\\
+- This can be improved--summing up the values possibly not the cleverest solution--calculation of variance etc. possible\\
+- Matrix Visualization?
 
@@
 %
-In order to provide a fast communication loop between method development and expert input we decided to base all implementations on modern web technologies which employs various advantages.
+In order to provide a fast communication loop between method development and expert input we decided to base all implementations on modern web technologies which benefits from various advantages:
 \begin{itemize}
@@
 %
-Image data is available as raw MRI DICOM files as well as segmented Meshes which can be used to compare subjects.
+Image data is available as raw DICOM files as well as segmented Meshes which can be used to compare subjects.
 %

### 2014-03-17 draft.tex
added Bernhards correction and improved Abstrackt with Steffens feedback

@@
 %
-Using well established statistical methods the data is hypothesis driven analyzed to find statistically significant variable correlations (\emph{interactions}).
+%Using well established statistical methods the data is hypothesis driven analyzed to find statistically significant variable correlations (\emph{interactions}).
+Driven by a priori defined hypotheses, the data is analyzed using well established statistical methods in order to find statistically significant correlations between variables (\emph{interactions}).
 %
@@
 %
-Analyzing these data requires image segmentation, extraction of key figures and shape based subject grouping.
+Analyzing these data requires image segmentation, extraction of key figures, and shape based subject grouping.
+%
+The statistically driven epidemiological workflow only allows to determine \emph{interactions} between image-derived metrics such as distances extracted from landmarks of the segmentation model.
 \\\\
-We propose a Interactive Visual Analysis approach that enables epidemiologists to examine both image-based as well as sociodemographic and medical attribute data.
+We propose an Interactive Visual Analysis approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic variables and attributes derived from the image data.
 %
-It allows for both hypothesis validation approaches as well as hypothesis generation by incorporating data mining methods.
+Using all reliable information from the image segmentation linked to non-image variables aims to unveil \emph{interactions} by applying an iterative analysis approach.
+%As extension to the epidemiological workflow it aims to unveil interactions based on all reliable shape information rather than abstracting it to a few key figures.
+%Iterative analysis between ...
+allows for extracting interactions which without the visiual analysis driven approach maybe would be overlooked.
+%Einsicht in Interaktionen zwischen nicht-bilddaten und bilddaten auf basis aller verfügbarer Informationen
+%Iterative Analyse von Organstrukturen im Bezug auf Bilddaten 
+%Iterative analysis of image-based epidemiological datasets aims to provide insight into interactions between organ shape and % in order to provide insights into \emph{interactions} by incorporating additional information
+%It extends the epidemiological workflow with an iterative Analysis process and 
+It allows for hypothesis validation and hypothesis generation by incorporating human pattern recognition as well as data mining methods.
 %
-Adaptive linked information visualization views and 3d-shape renderings are combined with epidemiological techniques.
+Multiple coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
 %
@@
 %
-Clinical problems and questions answered using epidemiological methods comprise diagnosis accuracy, disease frequency, risk factors, disease prognosis, effectiveness of treatments or preventions and cause of diseases \cite{Fletcher2012}.
+Clinical problems, such as the selection of diagnostic tools and efficient treatment, are tackled using results of epidemiological research.
 %
-%Cohort studies are a epidemiological method to gather data about a group of subjects (a \emph{cohort)}) to make general statements about these problems.
+Also, the introduction of preventive measures in medicine and beyond, are based on epidemiological research, where, for example, subgroups with increased risk are identified \cite{Fletcher2012}.
+%Clinical problems and questions answered using epidemiological methods comprise diagnosis accuracy, disease frequency, risk factors, disease prognosis, effectiveness of treatments or preventions and cause of diseases \cite{Fletcher2012}.
 %
-Observations made by clinicians in the daily routine are translated into hypothesis.
+On the other hand, observations made by clinicians in the daily routine are translated into hypothesis for epidemiological research.
 %
-These are used to determine environmental and lifestyle factors are as well as medical attributes which may influence a condition of interest.
+These are used to determine environmental and lifestyle factors as well as medical attributes which may influence a condition of interest.
 %
@@
 %
-Statistical methods like regression analysis aim aim to check the attribute list for plausibility.
+Statistical methods like regression analysis aim to check the attribute list for plausibility.
 %
 
-Longitudinal population-based studies like the Study of Health in Pomerania \cite{Volzke2011} aim to gather as much information as possible about a defined sample of people (a \emph{cohort}).
+Longitudinal population-based studies, such as the Study of Health in Pomerania \cite{Volzke2011}, aim to gather as much information as possible about a defined sample of people (a \emph{cohort}).
 %
-The sample is drawn randomized to avoid selection bias which prohibits statements based on statistical correlations in the cohort from being inferred to the whole population (\emph{external validity}) \cite{Fletcher2012}.
+The sample is drawn randomized to avoid selection \cite{Fletcher2012}.
+%bias which prohibits statements based on statistical correlations in the cohort from being inferred to the whole population (\emph{external validity}) \cite{Fletcher2012}.
 %
-Also a information bias needs to be avoided by strictly standardizing the data acquisition.
-%
-Statistical correlations are also prone to \emph{confounding}, meaning that two factors are dependent and therefore should be normalized with respect to each other.
+%Also a information bias needs to be avoided by strictly standardizing the data acquisition.
 %
@@
 %
-Therefore results need to be age adjusted to be comparable.
+Therefore results need to be age-adjusted to be comparable.
 %
@@
 \\\\
-Modern cohort studies often include medical image data which introduces new problems.
+Modern cohort studies often include medical image data.
 %
-Since it is unethical to expose people to radiation, non-harming imaging like Magnetic Resonance Imaging (MRI) or Ultrasound Imaging is used.
+Since it is unethical to expose people to radiation, non-harming imaging such as Magnetic Resonance Imaging (MRI) or Ultrasound Imaging is used.
 %
-As MRI scans are expensive there exists a tradeoff between quality of the image data and their associated costs.
+%As MRI scans are time-consuming and expensive there exists a tradeoff between quality of the image data and their associated costs.
 %
@@
 \\\\
-We propose a Interactive Visual Analysis approach \cite{Thomas2005} to provide a way to analyze both image- and non-image data.
+We propose an Interactive Visual Analysis approach \cite{Thomas2005} to provide a way to analyze both image- and non-image data.
 %
@@
 %
-Intended as a extension to the well established epidemiological tools it provides a way to rapidly validate hypothesis as well as trigger hypothesis generation using Data Mining methods such as clustering.
+Intended as an extension to the well established epidemiological tools it provides a way to rapidly validate hypothesis as well as trigger \emph{hypothesis generation} using Data Mining methods such as clustering.
 %
@@
 \begin{itemize}
-	\item Applying the Interactive Visual Analysis technique set to the epidemiological problem domain by characterizing affordances of this context.
+	\item Applying the Interactive Visual Analysis technique set to the epidemiological problem domain by characterizing requirements of this context.
 	\item Provide an overview over the workflow for analyzing cohort study data to gain insight into the large diverse subject spaces.
@@
 \subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
-% TODO Define who participates in epidemiological studies ?
-Since it is a diverse science, many different experts work at epidemiological studies, ranging from specialized doctors, medical computer scientists with focus on biometrics to statisticians.
+Many different experts work at epidemiological studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
 %
-Epidemiologists follow a strict workflow mainly driven by statistic tools to validate hypothesis about disease specific risk factors.
+Epidemiologists follow a workflow mainly driven by statistic tools to validate hypothesis about disease specific risk factors.
 %
-Following Thew and colleagues publication on this matter, the workflow can be characterized as follows.
+Following Thew and colleagues, the workflow can be characterized as follows \cite{Thew2009}.
 %
@@
 	\item A set of attributes depicting conditions affected by the hypothesis is compiled accordingly.
-%
+% TODO: Bernhard: Effect size of Attribute ?
 	\item Confounding variables need to be adjusted so that they do not affect the effect size of a attribute.
@@
 
-Reproducibility of results is an key requirement.
+Reproducibility of results is an key requirement in epidemiology.
 %
@@
 %
-If the data acquisition process changes, a information bias is introduced to the data, disallowing inference between acquisition cycles.
+If the data acquisition process changes, a information bias is introduced to the data, hampering inference in acquisition cycles.
 %
-This underlines the high quality standards to methods processing the data, whether to extract additional parameters or gain insight.
+%This underlines the high quality standards to methods processing the data, whether to extract additional parameters or gain insight.
 %
-To determine, whether a subject is prone to be affected by a certain disease, relative risks are expressed through the evaluation of p-values which indicate statistical significance.
+To determine, whether a subject is prone to be affected by a certain disease, \emph{relative risks} are expressed through the evaluation of p-values which indicate statistical significance.
+%
+Statistical correlations are prone to \emph{confounding}, meaning that two factors are dependent and therefore should be normalized with respect to each other.
 %
@@
 \subsection{Epidemiological Data} \label{EpidemiologicalData}
-Epidemiological data is highly heterogenous.
+Epidemiological data is highly heterogenous and incomplete.
 %
@@
 
-Indicators for medical conditions as well as questions about a subjects lifestyle are also often \emph{dichotomous}, meaning that they only have two manifestations (often \emph{Yes} or \emph{No}).
+Indicators for medical conditions as well as questions about a subjects' lifestyle are also often \emph{dichotomous}--they have two manifestations (often \emph{Yes} or \emph{No}).
 %
@@
 
-\paragraph{Image acquisition.} \label{ImageAcquisition} Imaging techniques emitting an hazardous amount of radiation for the subject are not suited for ethical reasons.
+\paragraph{Image acquisition.} \label{ImageAcquisition} Imaging techniques emitting hazardous amounts of radiation for the subject are not suited for ethical reasons.
 %
@@
 %
-Segmentations masks labeling the voxel of a shape of interest would be ideal since many different key figures can be derived from them.
+Segmentation masks labeling the voxel of a shape of interest would be ideal since many different key figures can be derived from them.
 %
@@
 %
-Data is collected diseases-independent.
+Data is collected without focus on a specific disease.
 %
@@
 %
-Data mining tools like the Principal Component Analysis are able to reduce the dimension by extracting most expressive components, but make the influence of each variable hard to determine.
+The Principal Component Analysis (PCA) and similar techniques are able to reduce the dimension by extracting most expressive components, but make the influence of each variable hard to determine.
 \\\\
@@
 %
-This approach already incorporated Data Mining approaches like dividing the data space using a k-nearest-neighbor technique and Principal Component Analysis.
+This approach already incorporated Data Mining approaches such as dividing the data space using a k-nearest-neighbor technique and Principal Component Analysis.
 %
-Steenwijk and colleagues employ a relational database to organize the data to visualize subject data using linked views like parallel coordinates, scatterplots and time plots \cite{Steenwijk2010}.
+Steenwijk and colleagues employ a relational database to organize the data to visualize subject data using linked views such as parallel coordinates, scatterplots and time plots \cite{Steenwijk2010}.
 %
@@
 
-Commercial systems like \texttt{Tableau} or \texttt{Spotfire} provide a rich user interface that allow to apply Visual Analytics techniques without the need of writing any code.
+Commercial systems such as \texttt{Tableau} or \texttt{Spotfire} provide a rich user interface that allows to apply Visual Analytics techniques without the need of writing any code.
 %
-With little effort, linked views can be created using these tools, but the data processing possibilities like derivation of new variables or the volume rendering capabilities are very limited.
+With little effort, linked views can be created using these tools, but the data processing possibilities such as derivation of new variables or the volume rendering capabilities are very limited.
 %
@@
 %
-As proof of concept a relation between size of the segmented shape and measured size of the subjects was measured and behaved as expected.
+As proof of concept a relation between size of the segmented shape and measured size of the subjects was measured.
 %
-This work focuses on incorporating these data as new features of the overall data set, making it possible to include it into the hypothesis validation and generation process.
+This work focuses on incorporating these derived data as new features of the overall data set, making it possible to include it into the hypothesis validation and generation process.
 %
-When applying clustering techniques on the non-image data is was found that \texttt{k-Prototypes} and \texttt{DBSCAN} is appropriate in the epidemiological context but is strongly dependent on the chosen variables and distance measure \cite{Klemm2014BVM}.
+When applying clustering techniques on the non-image data it was found that \texttt{k-Prototypes} and \texttt{DBSCAN} is appropriate in the epidemiological context but is strongly dependent on the chosen variables and distance measure \cite{Klemm2014BVM}.
 % *********************** / Our own Work ***********************
@@
 %
-Brushing capabilities allow for brushing and linking as well as filtering, but has limitations like making only one category brushable at a time.
+Brushing, linking and filtering are feasible with \emph{GPLOMS}, but have limitations such as making only one category brushable at a time.
 %
-We applied this technique to our data and it shows promising potential for simultaneously visualizing many different variable but does not fit in the scope of this paper.
+%We applied this technique to our data and it shows promising potential for simultaneously visualizing many different variables. %but does not fit in the scope of this paper.
 %
-The inspiration on the chosen visualization techniques stems from this publication.
+%The inspiration on the chosen visualization techniques stems from this publication.
 %
@@
 %
-Caban and colleagues investigated the suitability of variance visualizations of shape distribution models and concluded in their user study that user favor spherical glyph representations over deformation grids and likelihood volumes \cite{Caban2011}.
+Caban and colleagues investigated the suitability of variance visualizations of shape distribution models and concluded in their user study that users favor spherical glyph representations over deformation grids and likelihood volumes \cite{Caban2011}.
 %
-The distribution of shapes in a space derived from a principal component analysis is plotted by Busking and Colleagues in a 2D-projected plane of the space \cite{Busking2010a}.
+The distribution of shapes in a space derived from a PCA is plotted by Busking and Colleagues in a 2D-projected plane of the space \cite{Busking2010a}.
 %
@@
 %
-Via mesh morphing interpolated views can be created by the user in a separate view as well as comparisons in contour-view.
+Via mesh morphing interpolated views can be created by the user in a separate view as well as comparisons in a contour view.
 %
-Applying our data sets to this technique yielded a cluttered shape space due to the many subjects.
+%Applying our data sets to this technique yielded a cluttered shape space due to the many subjects.
 %
@@
 % Interactive Visual Analysis - Steffens Paper
-The strength of the Interactive Visual Analysis approach described in the next section is it's versatility with respect to the application field \cite{Konyha2009}.
+The strength of the Interactive Visual Analysis approach described in the next section is its versatility with respect to the application field \cite{Konyha2009}.
 %
@@
 %They aim to provide methods to analyze perfusion data with respect to 
-While we take similar steps when analyzing the data like employing statistical tests, our data is mostly independent from the medical image data and is not describing it--except the variables derived the segmentation model itself.
+While we take similar steps when analyzing the data such as employing statistical tests, our data is mostly independent from the medical image data and is not describing it--except the variables derived the segmentation model itself.
 
@@
 %
-As described in subsection~\ref{EpidemiologicalWorkflow}, the epidemiolgical workflow is a strict sequence of steps taken by domain experts and needs to be reproducible and comprise statistical integrity.
+As described in subsection~\ref{EpidemiologicalWorkflow}, the epidemiolgical workflow is a sequence of steps taken by domain experts and needs to be reproducible and comprise statistical integrity.
 %
@@
 %
-It would be possible that there actually are features of the data set which support the hypothesis by discriminating the population in the expected way, but with this approach they are not highlighted in any way.
+It would be possible that there are actually features of the data set which support the hypothesis by discriminating the population in the expected way, but with this approach they are not highlighted in any way.
 %
-This becomes even more imminent when the workflow is adapted to the analysis of the medical image data.
+This becomes even more important when the workflow is adapted to the analysis of the medical image data.
 %
-Domain experts annotate tediously landmarks which allow to derive metrics like distances which are then handled like other features and analyzed using the same set of statistical tools.
+Domain experts would have to annotate landmarks tediously to derive metrics such as distances. %which are then handled like other features and analyzed using the same set of statistical tools.
 %
@@
 %
-The human body exposing shape information for the \emph{physical view}, also referred as the \emph{independent variable} \cite{Oeltze2013}.
+The human body exposing shape information for the \emph{physical view} \cite{Oeltze2013}.
 %
@@
 %
+These variables are also referred to as \emph{independent variables}.
+%
 \emph{Dependent variables} in the epidemiological context can be divided twofold:
+%
 \begin{itemize}
@@
 %
-Data obtained using questionnaires or medical tests are often stored using statistical packages like \texttt{SPSS} or \texttt{Stata} which have a proprietary data format with limited export capabilities.
+Data obtained using questionnaires or medical tests are often stored using statistical packages such as \texttt{SPSS} or \texttt{Stata} which have a proprietary data format with limited export capabilities.
 %
-The best solution for us was to simply export the data in the respective tool to a character separated text file and then convert it to data types which are easier manageable like JSON or XML using our own classes.
+The best solution for us was to simply export the data in the respective tool to a character separated text file and then convert it to data types which are easier manageable such as JSON or XML using our own classes.
 %
-In order to verify that the conversion worked as expected and the data is valid it is good practice to use data wrangling tools like \texttt{OpenRefine} to validate the data.
+In order to verify that the conversion worked as expected and the data is valid, it is good practice to use data wrangling tools such as \texttt{OpenRefine} to validate the data (find missing data, clean up bad formatting).
 %
-Exporting the data dictionary which stores information about each manifestation of a feature is also an important step to get a detailed description of data variables and the meaning and unit of measurement of their values.
+Exporting the data dictionary, which stores information about each manifestation of a feature is also an important step to get a detailed description of data variables and the meaning and unit of measurement of their values.
 %
-Since the reasons for missing data have a wide range from ethical to medical and personal issues, these are also included as error codes which have to marked as such in the data dictionary.
+The reasons for missing data have a wide range from ethical to medical and personal issues.
+%
+Therefore, these are also included as error codes which have to marked as such in the data dictionary.
 \\\\
@@
 %
-This is either done manually by experts setting land marks (sometimes supported by algorithms connecting the land marks like Graph-Cuts) or by a (semi)-automatic detection, registration and segmentation.
+This is either done manually by experts setting landmarks (sometimes supported by algorithms connecting the land marks such as Graph-Cuts TODO: REFERENCE) or by a (semi)-automatic detection, registration and segmentation.
 %
-Algorithms applied to the data do not only have to deal with a large inter-subject variability of the structure of interest but also needs to be reproducible \cite{Preim2014}.
+Algorithms, applied to the data, do not only have to deal with a large inter-subject variability of the structure of interest but also need to be reproducible \cite{Preim2014}.
 %
-Model based approaches have shown to be effective for this task \cite{Gloger2010, Gloger2012, Rak2013}.
+Model-based approaches have shown in principle to be effective for segmentation \cite{Gloger2010, Gloger2012} and detection \cite{Rak2013}.
 %
-If a segmentation yields only binary masks separating the structures, algorithms like Growing and Adaptive Shapes can be applied creating a surface grid where each point is comparable throughout the population \cite{Ferrarini2007}.
+If a segmentation yields only binary masks separating the structures, algorithms such as Growing and Adaptive Shapes can be applied creating a surface grid where each point is comparable throughout the population \cite{Ferrarini2007}.
 %
-Intensity based comparison can be achieved using rigid image registration, but model based results however are preferable \cite{Klemm2012}.
+Intensity-based comparisons can be achieved using rigid image registration, but model based results however are preferable \cite{Klemm2012}.
 %
@@
 %
-As opposing to other \emph{IVA} application domains this step is more complicated in the epidemiological context.
+As opposing to other \emph{IVA} application domains, this step is more complicated in the epidemiological context.
 %
@@
 %
-Model based segmentations or masks yield data structures capable of calculating mean shapes and distances between individuals or subject groups.
+Model-based segmentations or masks yield data structures capable of calculating mean shapes and distances between individuals or subject groups.
 %
@@
 %
-For the most part the epidemiologist is interested in the shape of subjects within a range of a set of variables that describe the current hypothesis.
+The epidemiologist is primarily interested in the shape of subjects within a range of a set of variables that describe the current hypothesis.
 %
@@
 %
-Continuous variables like age are for example often divided in categories like young, aged and elderly.
+Continuous variables such as age are for example often divided in categories like young, aged and elderly.
 %
-Categorization is strongly dependent on the hypothesis and therefore requires suitable brushing techniques as described in section~\ref{sec:AdaptiveFeatureVisualization}.
+Categorization is strongly dependent on the hypothesis and therefore requires suitable brushing techniques as described in Section~\ref{sec:AdaptiveFeatureVisualization}.
 %
@@
 %
-Special for the application domain is the need of statistic measures which describe how variables correlate with each other given the selected groups.
+The need of statistic measures which describe how variables correlate with each other given the selected groups is special for the application domain.
 %
-These association also summarized using Pivot Tables which are popular in epidemiology and which are described in the following section.
+These associations also summarized using Pivot Tables which are popular in epidemiology and which are described in the following section.
 % - Selection of Elements in Scatterplots, Barcharts or Parallel Coordinates Views - observe how selection changes another view - this allows for multivariate analysis\\
@@
 \section{Interaction- and Visualization Techniques}
-%
+% TODO: Bernhard: Section anders beginnen!
 We employ the fourth and highest level which includes next to brushing and linking, advanced brushing, extraction of attributes the development of visualizations custom tailored to the data sets.
@@
 %
-The employed techniques needs to respect the epidemiological affordances.
+The employed techniques needs to respect the epidemiological requirements.
 %
@@
 %
-Doing so will automatically create a information visualization suitable for the current data type.
+Doing so will automatically create an information visualization suitable for the current data type.
 %
@@
 \subsection{Sidebar}
-A overview of all variables is presented in a sidebar where they are categorized to different types like somatometric, disease- or lifestyle related, pain indicators and laboratory data.
+An overview of all variables is presented in a sidebar where they are categorized to different types such as somatometric, disease- or lifestyle related, pain indicators and laboratory data.
 %
@@
 %
-The 3D-plots can be navigated using standard mouse inputs and the camera is synced so that a direct comparison is given.
+The 3D-plots can be navigated using standard mouse inputs and the camera is synchronized so that a direct comparison is given.
 %
@@
 \subsection{Pivot Tables}
-When opening a random epidemiological paper the reader will in almost all cases find some sort of pivot table to present the data.
+Pivot tables are frequently used to present the data in epidemiological paper.
+%When opening a random epidemiological paper the reader will in almost all cases find some sort of pivot table to present the data.
 %
@@
  \includegraphics[width=3.5in]{figures/technologies}
- \caption{\emph{Image is not final}. The front-end solution (left) uses state of the art web technologies like HTML5/CSS3, WebGL and SVG to display the data.
+ \caption{\emph{Image is not final}. The front-end solution (left) uses state of the art web technologies such as HTML5/CSS3, WebGL and SVG to display the data.
  % 
@@
 %
-In order to provide a fast communication loop between method development and expert input we decided to base all implementations on modern web technologies which benefits from various advantages:
+In order to provide a fast communication loop between method development and expert input, we decided to base all implementations on modern web technologies which benefits from various advantages:
 \begin{itemize}
-	\item No additional software needs to be installed, most people use decent state of the art web browser. If developed using proper standards, the applications even run on mobile devices. 
+	\item No additional software needs to be installed, most people use decent state-of-the-art web browser, even on mobile devices.
 	\item The Client-Server structure allows it to employ heavy computation on a server machine and transfer results to the client
-	\item Since Image data for several thousand subjects claims hundreds of Gigabytes disk space it can remain safely on the server and elements can be transferred on demand.
+	\item Since image data for several thousand subjects claims hundreds of Gigabytes disk space it can remain safely on the server and elements can be transferred on demand.
 	%
@@
 \\\\
-Non-image data for all Subjects including the data dictionary is stored on the Server in a JSON file.
+Non-image data for all subjects including the data dictionary is stored on the Server in a JSON file.
 %
@@
 %
-The server processes calculation heavy statistical tasks like calculation of Odds Ratios or Chi Square tests for all variable combinations in order to keep the computation time on the client as low as possible.
+The server processes calculation heavy statistical tasks such as calculation of Odds Ratios or Chi Square tests for all variable combinations in order to keep the computation time on the client as low as possible.
 %
@@
 %
-Information Visualizations like Scatterplots and bar charts are created using the popular Data Driven Documents library which works well for attaching data to visible elements like vector graphics.
+Information Visualizations such as Scatterplots and bar charts are created using the popular Data Driven Documents library which works well for attaching data to visible elements like vector graphics.
 %

### 2014-03-19 draft.tex
changed abstract with help of Kai

@@
 %% indicate IEEE Member or Student Member in form indicated below
-\author{TBA}
+\author{--blind--}
 \authorfooter{
@@
 %other entries to be set up for journal
-\shortauthortitle{Klemm \MakeLowercase{\textit{et al.}}: Interactive Visual Analytics of Image-Centric Cohort Study Data}
+\shortauthortitle{Klemm \MakeLowercase{\textit{et al.}}: Interactive Visual Analysis of Image-Centric Cohort Study Data}
 %\shortauthortitle{Firstauthor \MakeLowercase{\textit{et al.}}: Paper Title}
@@
 %% Abstract section.
-\abstract{Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease specific risk factors.
+\abstract{%
+%(The problem). 
+Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease specific risk factors.
 %
@@
 %
-%Using well established statistical methods the data is hypothesis driven analyzed to find statistically significant variable correlations (\emph{interactions}).
-Driven by a priori defined hypotheses, the data is analyzed using well established statistical methods in order to find statistically significant correlations between variables (\emph{interactions}).
+The data is analyzed using a priori defined hypotheses to find statistically significant correlations between variables (\emph{interactions}).
 %
@@
 %
-Analyzing these data requires image segmentation, extraction of key figures, and shape based subject grouping.
+%Analyzing these data requires image segmentation, extraction of key figures, and shape based subject grouping.
 %
@@
 \\\\
+%(New Solution).
 We propose an Interactive Visual Analysis approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic variables and attributes derived from the image data.
 %
+This is achieved by combining brushing and linking enabled coordinated information visualization views and interactive 3D-shape renderings with epidemiological data representations such as pivot tables and key figures as odds ratios.
+%incorporating coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
+%
+%Multiple coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
+%
+%(Validation).
+The presented concepts are applied expert-guided to gather and evaluate hypothesis about the aging process of the lumbar spine.
+%
+%(Results). 
+It shows to be a more flexible comparison between image-and non-image data.
+%
+%(Implications).
+The new framework allows for hypothesis validation and hypothesis generation by incorporating human pattern recognition as well as data mining methods.
+%
 Using all reliable information from the image segmentation linked to non-image variables aims to unveil \emph{interactions} by applying an iterative analysis approach.
-%As extension to the epidemiological workflow it aims to unveil interactions based on all reliable shape information rather than abstracting it to a few key figures.
-%Iterative analysis between ...
+
+%
+%Similarity measures between data variables are used to compute interesting changes in variable interactions for the current variable selection.
+%
+%Shape based grouping of subjects is facilitated using clustering techniques which operate on surface meshes extracted from the image segmentation.
+} % end of abstract
+
+% \abstract{Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease specific risk factors.
+% %
+% Cohort studies comprise heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors.
+% %
+% %Using well established statistical methods the data is hypothesis driven analyzed to find statistically significant variable correlations (\emph{interactions}).
+% Driven by a priori defined hypotheses, the data is analyzed using well established statistical methods in order to find statistically significant correlations between variables (\emph{interactions}).
+% %
+% Modern cohort studies also incorporate medical image data.
+% %
+% Analyzing these data requires image segmentation, extraction of key figures, and shape based subject grouping.
+% %
+% The statistically driven epidemiological workflow only allows to determine \emph{interactions} between image-derived metrics such as distances extracted from landmarks of the segmentation model.
+% \\\\
+% We propose an Interactive Visual Analysis approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic variables and attributes derived from the image data.
+% %
+% Using all reliable information from the image segmentation linked to non-image variables aims to unveil \emph{interactions} by applying an iterative analysis approach.
+% %As extension to the epidemiological workflow it aims to unveil interactions based on all reliable shape information rather than abstracting it to a few key figures.
+% %Iterative analysis between ...
 allows for extracting interactions which without the visiual analysis driven approach maybe would be overlooked.
 %Einsicht in Interaktionen zwischen nicht-bilddaten und bilddaten auf basis aller verfügbarer Informationen
-%Iterative Analyse von Organstrukturen im Bezug auf Bilddaten 
-%Iterative analysis of image-based epidemiological datasets aims to provide insight into interactions between organ shape and % in order to provide insights into \emph{interactions} by incorporating additional information
-%It extends the epidemiological workflow with an iterative Analysis process and 
-It allows for hypothesis validation and hypothesis generation by incorporating human pattern recognition as well as data mining methods.
-%
-Multiple coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
-%
-Similarity measures between data variables are used to compute interesting changes in variable interactions for the current variable selection.
-%
-Shape based grouping of subjects is facilitated using clustering techniques which operate on surface meshes extracted from the image segmentation.
-} % end of abstract
+% %Iterative Analyse von Organstrukturen im Bezug auf Bilddaten 
+% %Iterative analysis of image-based epidemiological datasets aims to provide insight into interactions between organ shape and % in order to provide insights into \emph{interactions} by incorporating additional information
+% %It extends the epidemiological workflow with an iterative Analysis process and 
+% It allows for hypothesis validation and hypothesis generation by incorporating human pattern recognition as well as data mining methods.
+% %
+% Multiple coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
+% %
+% Similarity measures between data variables are used to compute interesting changes in variable interactions for the current variable selection.
+% %
+% Shape based grouping of subjects is facilitated using clustering techniques which operate on surface meshes extracted from the image segmentation.
+% } % end of abstract
 
@@
 %% please capitalize first letter and insert punctuation after last keyword
-\keywords{Interactive Visual Analytics, Epidemiology}
+\keywords{Interactive Visual Analysis, Epidemiology}
 
@@
 %
-This is either done manually by experts setting landmarks (sometimes supported by algorithms connecting the land marks such as Graph-Cuts TODO: REFERENCE) or by a (semi)-automatic detection, registration and segmentation.
+This is either done manually by experts setting landmarks (sometimes supported by algorithms connecting the land marks such as graph cuts \cite{GraphCut}) or by a (semi)-automatic detection, registration and segmentation.
 %
@@
 \section{Application}
+- Acceptance plays a big role - you have to pick up the user by presenting information they are used to in a new way
 \subsection{The Spine Dataset}
@@
 \section{Summary and Conclusion}
+- Future Work: Matrix-View of differences - more intuitive\\
+- UI more flexible by hiding UI-panes
 

### 2014-03-20 draft.tex
Abstract ready for mandatory submission

@@
 %(The problem). 
-Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease specific risk factors.
+Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease-specific risk factors.
 %
@@
 %
-This is achieved by combining brushing and linking enabled coordinated information visualization views and interactive 3D-shape renderings with epidemiological data representations such as pivot tables and key figures as odds ratios.
+This is achieved by combining brushing and linking enabled coordinated information visualization views and interactive 3D shape renderings with epidemiological data representations such as pivot tables and key figures as odds ratios.
 %incorporating coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
@@
 %(Validation).
-The presented concepts are applied expert-guided to gather and evaluate hypothesis about the aging process of the lumbar spine.
+The presented concepts are applied expert-guided to gather and evaluate hypotheses about the aging process of the lumbar spine.
 %
 %(Results). 
-It shows to be a more flexible comparison between image-and non-image data.
+It shows to be a more flexible comparison between image and non-image data.
 %

### 2014-03-21 draft.tex
first part of the Application section

@@
 %
-This is achieved by combining brushing and linking enabled coordinated information visualization views and interactive 3D shape renderings with epidemiological data representations such as pivot tables and key figures as odds ratios.
+This is achieved by combining brushing and linking enabled coordinated information visualization views and interactive 3D shape renderings with epidemiological data representations such as pivot tables and key figures as association measures.
 %incorporating coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
@@
 
-\section{Interactive Visual Analytics in Cohort Study Data}
+\section{Interactive Visual Analyis in Cohort Study Data}
 \begin{figure}[htb]
@@
 \end{figure}
+- Acceptance plays a big role - you have to pick up the user by presenting information they are used to in a new way\\
 %
@@
 \section{Application}
-- Acceptance plays a big role - you have to pick up the user by presenting information they are used to in a new way
+We apply the presented set of techniques to a data set which is compiled to analyze lower back pain which is one of the most common reasons for an adult to see a physicians \cite{Backpain}.
+%
+Data confidentiality and ethical reasons prohibit us from accessing the complete data space of the SHIP feature space.
+%
+The epidemiologists compiled feature list which is a tradeoff between complexity and ability to pass the ethics committee.
+%
+%The goal is to find features which \emph{interact} with lower back pain based on image and non-image data.
+%
 \subsection{The Spine Dataset}
-- Describe steps from gathering Information from the raw image files (segmentation, abstraction, visualization)\\
-- Problem of sparse differences - Visualization has to be more abstract to emphasize the differences.\\
-- Input of Epidemiologists goes here!\\
-\\\\
-\textbf{From VMV'13 Paper}\\
+We divide the data set in image and non-image data.
+%
+There are 136 features describing diagnosed diseases, lifestyle factors, women specific factors, pain indicators, laboratory values somatometric variables and are ordered accordingly.
+%
+The image data was acquired on a 1.5~Tesla scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany) by four trained technicians in a standardized way.
+%
+The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence (676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
 
-All whole-body MRI scans were acquired on a 1.5~Tesla scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany) by four trained technicians in a standardized way.
+\subsection{Processing the Image data}
+The lumbar spine was detected in the image data using a hierarchical finite element method according to \cite{Rak2013}.
+%
+This semi-automatic method requires the user to initialize the Tetrahedron-based finite element models (FEM) with a click on the L2-vertebra.
+%
+It uses a weighted sum of T1- and T2-weighted MRI images to detect the lumbar spine shape.
+%
+The registered models capture resilient information about shape of the lumbar spine canal as well as the position of the L1 -L5 vertebrae.
+%
+For clustering purposes we extracted the centerline of the spine canal of the lumbar spine canal which captures information about lordosis and scoleoses which is the medical term for spine curvature \cite{Klemm2013VMV}.
+%
+%This information is a major feature for diagnosing lumbar back pain.
+\subsection{Processing the Non-Image data}
+- SPSS Export? JSON-Export? Automatisches Anfuegen des Dictionaries?
+
+%\subsection{The Spine Prototype}
+
+- Clustering using Steffens Algorithm
+% - Describe steps from gathering Information from the raw image files (segmentation, abstraction, visualization)\\
+% - Problem of sparse differences - Visualization has to be more abstract to emphasize the differences.\\
+% - Input of Epidemiologists goes here!\\
+% \\\\
+% \textbf{From VMV'13 Paper}\\
+% 
+% All whole-body MRI scans were acquired on a 1.5~Tesla scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany) by four trained technicians in a standardized way.
 Subjects were placed in the supine position.
@@
 
-%The model is placed in the scene using an empirically chosen initialization point.
-The force acting on the model stems from aggregation of loads, which are derived from a potential field resulting from a weighted sum of the T1- and T2-weighted MRI images, see \cite{Rak2013}.
-After detecting all spines, we register the models because in a later clustering step we only want to capture the local deformation of the lumbar spine, not different locations in world space.
-The models are registered using the Kabsch Algorithm \cite{Kabsch1976}, which is designed to minimize the root mean squared deviation between paired sets of points.
-The model is placed in the scene using an empirically chosen initialization point.
+% The model is placed in the scene using an empirically chosen initialization point.
 The force acting on the model stems from aggregation of loads, which are derived from a potential field resulting from a weighted sum of the T1- and T2-weighted MRI images, see \cite{Rak2013}.
@@
 The models are registered using the Kabsch Algorithm, which is designed to minimize the root mean squared deviation between paired sets of points.
-The model-based detection captures information about the spine canal curvature as well as the alignment of the vertebrae.
+% The model-based detection captures information about the spine canal curvature as well as the alignment of the vertebrae.
 It is not meant to capture information about vertebrae deformation and differences in spine canal extent.

### 2014-03-22 draft.tex
Wrote on Application Section

@@
 %% please capitalize first letter and insert punctuation after last keyword
-\keywords{Interactive Visual Analysis, Epidemiology}
+\keywords{Interactive Visual Analysis, Epidemiology, Spine}
 
@@
 %
-Statistical measures of metric variables such as mean, standard deviation, skewness, or inter-quartile range are used to create \emph{dimension plots}.
+Statistical measures of continuous variables such as mean, standard deviation, skewness, or inter-quartile range are used to create \emph{dimension plots}.
 %
@@
 
-\section{Interaction- and Visualization Techniques}
+\section{Interaction- and Visualization Techniques} \label{Interaction- and Visualization Techniques}
 % TODO: Bernhard: Section anders beginnen!
@@
 \subsection{Automated Feature Suggestion}
+% https://de.wikipedia.org/wiki/Kontingenzkoeffizient#Cram.C3.A9rs_V
+TODO REWRITE USING CRAMERS V
+
 As discussed previously, highlighting potential interesting values in the data set is one major benefit of the \emph{IVA} powered approach.
@@
 %
+Epidemiological analysis of lumbar back pain such the work of Harreby and colleagues \cite{Harreby1996} is largely focused on non-image information.
+%
+If at all, only a few shape related features are included in comparable studies \cite{Lang2011}.
+%
+To our knowledge, this is the first approach on analyzing shape related information of the whole lumbar spine with other epidemiological features.
+
+%
+Determining risk factors can lead to \cite{Fletcher2012}:
+\begin{itemize}
+	\item a better understanding of effects of preventive measures such as occupational health and safety regulations
+	\item prognostic features for diagnosis and treatment of lumbar back pain
+	\item determination of particularly effected risk groups
+\end{itemize}
+%
+Characterizing the healthy aging process of the spine is a large stretch goal that allows to determine age-normalized probabilities for diseases using individual risk factors.
+
 Data confidentiality and ethical reasons prohibit us from accessing the complete data space of the SHIP feature space.
@@
 %
-\subsection{The Spine Dataset}
+\subsection{The Lumbar Spine Dataset}
 We divide the data set in image and non-image data.
@@
 
-\subsection{Processing the Image data}
+\subsection{Data Preprocessing} \label{Data Preprocessing}
+
+\paragraph{Image-Data.} \label{Image-Data}
 The lumbar spine was detected in the image data using a hierarchical finite element method according to \cite{Rak2013}.
@@
 %
-The registered models capture resilient information about shape of the lumbar spine canal as well as the position of the L1 -L5 vertebrae.
-%
-For clustering purposes we extracted the centerline of the spine canal of the lumbar spine canal which captures information about lordosis and scoleoses which is the medical term for spine curvature \cite{Klemm2013VMV}.
-%
-%This information is a major feature for diagnosing lumbar back pain.
-\subsection{Processing the Non-Image data}
-- SPSS Export? JSON-Export? Automatisches Anfuegen des Dictionaries?
-
-%\subsection{The Spine Prototype}
-
-- Clustering using Steffens Algorithm
+The registered models capture resilient information about shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae.
+%
+For clustering purposes we extracted the centerline of the spine canal of the lumbar spine canal which captures information about lordosis and scoliosis which are medical terms for spine curvature \cite{Klemm2013VMV}.
+
+TODO HOW MANY SUBJECTS?
+
+\paragraph{Non-Image Data.}
+
+In order to allow for fast and easy access of the data outside of tools like \texttt{SPSS} or \texttt{STATA} the data was exported to a JSON file format which can easily be loaded using Javascript.
+%
+Each variable is stored as object which contains: 
+\begin{itemize}
+	\item the data as array of values--categorical values and error codes are stored using IDs
+	\item the data type (continuous, nominal, ordinal, dichotomous)
+	\item a detailed description of the variable
+	\item the data dictionary which translates value- or error IDs to the actual values
+\end{itemize}
+%
+% Binning: https://www.researchgate.net/post/What_metric_best_captures_the_information_loss_when_binning_a_large_data-set_into_a_histogram_representation
+Continuous variables are discretized to allow for CramérsV contingency coefficient assessment.
+%
+In epidemiology, continuous data is usually categorized into ordinal groups of equal size.
+%
+Since the number of categories often strongly depends on the hypothesis, the discretization steps can be adapted dynamically.
+%
+To allow for hypothesis generation we set the number of groups to five if not specified otherwise.
+TODO Formula?
+
+\subsection{Shape Visualization and Clustering}
+The detection algorithm described in Section~\ref{Data Preprocessing} yields a model that determines voxel belonging to the lumbar spine.
+%
+The tetrahedron-based model has consists of corresponding grid points for each structure instance.
+%
+This allows for calculation of shape-distance and similarity.
+%
+We use this information to calculate mean-shapes for groups for plotting them as described in Section~\ref{Interaction- and Visualization Techniques}.
+%
+For dichotomous variables the color codes distances between mean shapes of the two groups, for variables with more than two manifestation the color encodes the distance to the mean shape of all variables.
+\\\\
+Shape based clustering is carried out via Agglomerative Hierarchical Clustering of the spine canal centerlines that are described in Section~\ref{Data Preprocessing} \cite{Klemm2013VMV}.
+%
+Since it is not possible to determine the number of clusters in a given group, it is determined automatically by computing a \emph{knee/elbow} point which is described as good tradeoff between number of cluster and a cluster evaluation metric \cite{Salvador2004}.
+%
+For details, see \cite{Klemm2013VMV}.
+%
+The method has proven to produce comprehensible results on a preliminary data set and was able to reproduce textbook knowledge before \cite{Klemm2013VMV}.
+
+\subsection{Exploratory Analysis of the Lumbar Spine Dataset}
+%
+We analyzed the given dataset expert guided to assess its suitability for both hypothesis validation and generation.
+%
+\subsubsection{Hypothesis free Analysis}
+\begin{figure}[htb]
+ \centering
+ \label{fig:hypopthesisfree}
+ %\includegraphics[width=1.5in]{figures/InteractionLoop}
+ \includegraphics[width=3.5in]{figures/hypothesisfree}
+ \caption{\emph{Image is not final}.
+(a) Clustering result of all subjects.
+The bar chart height indicates the number of subjects in the cluster.
+The difference to the mean-shape is color-coded where blue represents no difference and red a large difference.}
+\end{figure}
+%
+Analyzing the data set without a prior hypothesis requires a starting point which gives a overview over the data first \cite{Shneiderman1996}.
+%
+Shape based hypothesis free exploration starts with a shape-grouping step achieved by image-based clustering.
+%
+The results for this step can be seen in Figure~\ref{fig:hypopthesisfree}.
+%
+Cluster 9 represents the subjects with average shape.
+%
+Other shapes differ with respect to size, such as Cluster 2, 8, 10 and 3 where the last one also is more straight, which is expected for larger subjects.
+%
+Cluster 4, 7 and 5 contain outliers, characterized by their unusual shape and small number.
+%
+Cluster 8 - Viele nicht erwerbstätige - weiter hier
+
+TODO: While this approach does not assume any hypothesis, the data is met with a selection bias when compiling the list of related features by the domain experts.
+
+TODO Implement Color coding only in a certain direction!
+
+\subsubsection{Hypothesis based Analysis}
+%
+
+%
+
+\subsubsection{Concluding Domain-Expert Feedback}
+Our clinical partners were interested in the MRI data for the subjects in the outlier clusters because they are highly likely to exhibit pathologies.
+%
+We plan to include a \texttt{DICOM}-viewer to meet this wish.
+
 % - Describe steps from gathering Information from the raw image files (segmentation, abstraction, visualization)\\
@@
 - UI more flexible by hiding UI-panes
+- Dynamic Discretization to better fit the data distributions
+- More Shape based filtering - reorder by size, curvature, etc. ...
+- Color code different Aspects - Only difference in x or y direction
+
+Discretization of continous variables into bins of equal size may bias the underlying data by not considering the variable distribution function \cite{Fletcher2012}.
+% Fletcher 2012 - S. 257
+To reduce the number of false positive findings, the data space can be randomly be cut in half the hypothesis then can be cross-validated for statistical soundness.
+%
+This requires a large number of subjects, especially if the investigated features are rare and are only presented by a few subjects.
+%
+As suggested by Fletcher and Fletcher, we plan to validate the hypothesis using the \texttt{SHIP-TREND} cohort \cite{Fletcher2012}.
 

### 2014-03-23 draft.tex
Done with Application Part, images missing, still need feedback on two parts

@@
 - Time-Line\\
-- Statistical Analysis (Odds ratios)
+- Statistical Analysis (Odds ratios)\\
+
+TODO 2\\
+- 3D Plots are small multiples - reference Tufte!
 
@@
 \section{Application}
-We apply the presented set of techniques to a data set which is compiled to analyze lower back pain which is one of the most common reasons for an adult to see a physicians \cite{Backpain}.
+We applied the presented set of techniques to a data set which is compiled to analyze lower back pain. 
+%
+It is one of the most common reasons for an adult to see a physicians in the western civilization \cite{Backpain}.
 %
@@
 %
-If at all, only a few shape related features are included in comparable studies \cite{Lang2011}.
+If at all, only a few shape related features are included in comparable studies, for example by Lang and colleagues \cite{Lang2011}.
 %
 To our knowledge, this is the first approach on analyzing shape related information of the whole lumbar spine with other epidemiological features.
-
-%
-Determining risk factors can lead to \cite{Fletcher2012}:
+%
+Determining risk factors in this area can lead to \cite{Fletcher2012} (evtl. weglassen, Dopplung bei Epidemiological Background?):
 \begin{itemize}
@@
 \end{itemize}
-%
-Characterizing the healthy aging process of the spine is a large stretch goal that allows to determine age-normalized probabilities for diseases using individual risk factors.
+% Determining risk factors in this area can lead to \cite{Fletcher2012} (evtl. weglassen, Dopplung bei Epidemiological Background?):
+% \begin{itemize}
+% 	\item a better understanding of effects of preventive measures such as occupational health and safety regulations
+% 	\item prognostic features for diagnosis and treatment of lumbar back pain
+% 	\item determination of particularly effected risk groups
+% \end{itemize}
+%
+Characterizing the healthy aging process of the spine is a large stretch goal that allows to determine age-normalized probabilities for spine-related diseases by incorporating individual risk factors.
 
@@
 %
-The epidemiologists compiled feature list which is a tradeoff between complexity and ability to pass the ethics committee.
+Our clinical partners compiled a feature list which is a tradeoff between complexity and limitations of the responsible ethics committee. (vllt. etwas hart formuliert).
 %
@@
 %
-There are 136 features describing diagnosed diseases, lifestyle factors, women specific factors, pain indicators, laboratory values somatometric variables and are ordered accordingly.
+There are 136 features describing diagnosed diseases, lifestyle factors, women specific factors, pain indicators, laboratory values, somatometric variables and are ordered accordingly.
 %
@@
 \subsection{Data Preprocessing} \label{Data Preprocessing}
-
+Transformation operations on the data to prepare it for the presented prototype are denoted as data preprocessing.
+%We denote all data transformation steps necessary outside of our prototype for it to be prepared to work in our prototype as data preprocessing.
+%
 \paragraph{Image-Data.} \label{Image-Data}
-The lumbar spine was detected in the image data using a hierarchical finite element method according to \cite{Rak2013}.
-%
-This semi-automatic method requires the user to initialize the Tetrahedron-based finite element models (FEM) with a click on the L2-vertebra.
+The lumbar spine was detected in the image data using a hierarchical finite element method according to Rak and colleagues \cite{Rak2013}.
+%
+This semi-automatic method requires the user to initialize the Tetrahedron-based finite element models (FEM) with a click on the L3-vertebra.
+%
+Two user defined landmarks on the top and bottom of the L3-vertebra are used to obtain an initial height estimation of the model.
+%Two more user- defined landmarks are the top and bottom of the L3-vertebra, which transforms size of the model accordingly to match the L3-vertebra .
 %
@@
 %
-The registered models capture resilient information about shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae.
+The registered models capture resilient information about shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae \cite{Klemm2013VMV}.
+%
+Due to TODO ANTWORT VON MARKO HIER, 983 models are obtained.
 %
@@
 
-TODO HOW MANY SUBJECTS?
-
 \paragraph{Non-Image Data.}
 
-In order to allow for fast and easy access of the data outside of tools like \texttt{SPSS} or \texttt{STATA} the data was exported to a JSON file format which can easily be loaded using Javascript.
-%
-Each variable is stored as object which contains: 
+To ensure fast and easy data access outside of statistical processors like \texttt{SPSS} or \texttt{STATA}, the data was exported to the JSON file format which can easily be parsed by modern programming languages.
+%
+Each feature is stored as object which contains: 
 \begin{itemize}
@@
 % Binning: https://www.researchgate.net/post/What_metric_best_captures_the_information_loss_when_binning_a_large_data-set_into_a_histogram_representation
-Continuous variables are discretized to allow for CramérsV contingency coefficient assessment.
+Continuous variables are discretized to allow for \emph{Cram\'{e}rs' V} contingency coefficient assessment.
 %
@@
 To allow for hypothesis generation we set the number of groups to five if not specified otherwise.
-TODO Formula?
+% TODO MIT KAI FORMEL ERARBEITEN?
 
 \subsection{Shape Visualization and Clustering}
-The detection algorithm described in Section~\ref{Data Preprocessing} yields a model that determines voxel belonging to the lumbar spine.
-%
-The tetrahedron-based model has consists of corresponding grid points for each structure instance.
+%The detection algorithm described in Section~\ref{Data Preprocessing} yields a model that determines voxel belonging to the lumbar spine.
+%
+The tetrahedron-based model detection model described in Section~\ref{Data Preprocessing} consists of corresponding grid points for each structure instance.
 %
@@
 %
-We use this information to calculate mean-shapes for groups for plotting them as described in Section~\ref{Interaction- and Visualization Techniques}.
-%
-For dichotomous variables the color codes distances between mean shapes of the two groups, for variables with more than two manifestation the color encodes the distance to the mean shape of all variables.
-\\\\
-Shape based clustering is carried out via Agglomerative Hierarchical Clustering of the spine canal centerlines that are described in Section~\ref{Data Preprocessing} \cite{Klemm2013VMV}.
-%
-Since it is not possible to determine the number of clusters in a given group, it is determined automatically by computing a \emph{knee/elbow} point which is described as good tradeoff between number of cluster and a cluster evaluation metric \cite{Salvador2004}.
+This information is used to calculate mean-shapes as described in Section~\ref{Interaction- and Visualization Techniques}.
+
+Shape distance is mapped onto color.
+%
+For dichotomous variables, the color codes distances between mean shapes of the two groups, for variables with more than two manifestation it encodes the distance to the global mean shape of all subjects (Figure~\ref{fig:hypothesisfree}).
+\\\\
+Shape based clustering is carried out via Agglomerative Hierarchical clustering of the spine canal ,centerlines which are described in Section~\ref{Data Preprocessing} \cite{Klemm2013VMV}.
+%
+Since it is not possible to determine the number of clusters in a given group, it is automatically computed using a \emph{knee/elbow} point which is described as tradeoff between number of cluster and a cluster evaluation metric \cite{Salvador2004}.
 %
@@
 %
-The method has proven to produce comprehensible results on a preliminary data set and was able to reproduce textbook knowledge before \cite{Klemm2013VMV}.
+The method has proven to produce comprehensible results on a preliminary data set and was able to reproduce textbook knowledge \cite{Klemm2013VMV}.
 
@@
 %
-We analyzed the given dataset expert guided to assess its suitability for both hypothesis validation and generation.
-%
-\subsubsection{Hypothesis free Analysis}
+%We analyzed the given dataset expert guided to assess its suitability for supporting both hypothesis-free analysis as well as generation.
+Expert guided analysis assessed the suitability of our approach for supporting both hypothesis-free analysis as well as hypothesis generation.
+%
+\subsubsection{Hypothesis free Analysis} \label{Hypothesis free Analysis}
 \begin{figure}[htb]
  \centering
- \label{fig:hypopthesisfree}
  %\includegraphics[width=1.5in]{figures/InteractionLoop}
@@
 The difference to the mean-shape is color-coded where blue represents no difference and red a large difference.}
+ \label{fig:hypothesisfree}
 \end{figure}
@@
 %
-The results for this step can be seen in Figure~\ref{fig:hypopthesisfree}.
-%
+The results for this step can be seen in Figure~\ref{fig:hypothesisfree} (a).
+
 Cluster 9 represents the subjects with average shape.
 %
-Other shapes differ with respect to size, such as Cluster 2, 8, 10 and 3 where the last one also is more straight, which is expected for larger subjects.
+Other shapes differ with respect to size, such as Cluster 2, 8, 10 and 3 where the last one also is more straight, which is usual for subjects with larger body height.
 %
@@
 %
-Cluster 8 - Viele nicht erwerbstätige - weiter hier
-
-TODO: While this approach does not assume any hypothesis, the data is met with a selection bias when compiling the list of related features by the domain experts.
-
-TODO Implement Color coding only in a certain direction!
-
-\subsubsection{Hypothesis based Analysis}
-%
-
-%
-
-\subsubsection{Concluding Domain-Expert Feedback}
-Our clinical partners were interested in the MRI data for the subjects in the outlier clusters because they are highly likely to exhibit pathologies.
+Cluster 8 was of special interest because of its large distance to the mean shape while still exposing the second highest subject count.
+%
+Looking at the \emph{Cram\'{e}rs' V} contingency values of the group reveals interactions of this group with employment status, body size, age, thyroid nodules and blood-fat value.
+%
+TODO Feedback einholen von Frau Hegenscheid, TODO Implement Color coding only in a certain direction!\\
+TODO Selection of the variables and display them using the pivot table!
+\\\\
+While this approach does not assume any hypothesis, the data is met with a selection bias when compiling the list of related features by the domain experts.
+%
+It is arguable if this first filtering step, which is purely based on expert experience, is really a disadvantage, since it rules out many parameters which may interact with the presented features but the value of this knowledge would be small \cite{Wiley2008}.
+
+\subsubsection{Hypothesis based Analysis} \label{Hypothesis based Analysis}
+\begin{figure}[htb]
+ \centering
+ %\includegraphics[width=1.5in]{figures/InteractionLoop}
+ \includegraphics[width=3.5in]{figures/placeholder}
+ \caption{\emph{Image is not final}. (a) "Did you experience back pain in the past three month" Yes No; (b) Clustering of Yes}
+ \label{fig:hypopthesisbased}
+\end{figure}
+%
+If the user has already a hypothesis about a relation between a non-image feature regarding shape the workflow slightly differs from the hypothesis free analysis.
+%
+The starting point of the analysis is the selection of a feature of interest by dragging it into the canvas area and view the subjects distribution as well as their shape differences.
+%
+In our use case, the epidemiologist was interested in the questionnaire answer "Did you experience back pain in the past three month".
+%
+The mean shapes of the resulting visualization as seen in Figure~\ref{fig:hypopthesisbased} (a) show no difference between the two groups.
+%
+Either there are no differences or the variance information was lost in the mean-shape calculation.
+%
+Since the focus is on subjects which suffer from back pain, the clustering result of these subjects are then drawn into the canvas area, yielding six cluster as seen in Figure~\ref{fig:hypopthesisbased} (b).
+%
+Cluster 5 stood out for having a so-called hyperlordosis, a strong curvature of the lumbar spine which is a indicator for back pain.
+
+\emph{Cram\'{e}rs' V} contingency values highlighted relationships of this cluster with joint degeneration, meat eating habits, preoccupation, back pain, neck or shoulder pain and waist circumference.
+%
+Since the prior selection only yields subjects that report back pain, the pain indicators specify the pain localization for the subjects.
+
+It is well known that overweight is a indicator for back pain.
+%
+While the Body Mass Index (BMI) is a key figure for assessing height and weight of a subject, it does not tell us anything on how the weight is distributed in the body.
+%
+Our clinical partner were interested in the fact that this group presented a correlation with waist circumference.
+%
+Our finding follows the recent trends that indicate that BMI is not a good measure for assessing body-shape since healthy weight is dependent on many other measures \cite{Ahima2013}.
+%
+It indicates that waist circumference rather than the BMI interacts with unusual shaped spines for subjects with lumbar back pain.
+%
+The influence of the parameter is now in the focus of further analysis.
+
+\subsubsection{Follow-Up Tasks and Concluding Domain-Expert Feedback}
+Defining a causal relationship solely based on observed correlations of two features is \emph{cum hoc ergo propter hoc}--correlation does not automatically imply causation \cite{Tufte2003}.
+%
+The observed correlations need to be carefully checked for confounder and medical soundness!
+%
+Statisticians validate causal inferences of the drawn conclusions.
+
+Features that potentially interact with a disease related condition need to be validated.
+%
+To increase the probability of the observation not to be random, our clinical partners cross check for interactions in another cohort, the \texttt{SHIP-TREND}.
+%If features are identified as interesting for a disease, the epidemiologists look for interactions in another cohort, the \texttt{SHIP-TREND}.
+%
+%If the correlation is observed there as well, it is likely not random.
+\\\\
+Our clinical partners pointed out the usability of the presented methods to guide their attention to features which where not in the focus of their attention and expectance.
+%
+The explorative nature of the methods work well for gathering interactions which may act as confounder, as outcome of a disease or as an actual cause or risk factor.
+%"\emph{Correlation does not imply causation}" is a famous saying in statistics, meaning that correlation does not necessarily imply causal relationships \cite{}.
+%
+This distinction is hard to make and requires a lot of clinical experience.
+%
+The combination of multiple views with shape information help to connect many different information sources to make the large information spaces cognitive feasible.
+%
+
+On the feature level, our clinical partners were interested in the MRI scans for the subjects in the outlier clusters because they are highly likely to exhibit pathologies.
 %
@@
 % - Input of Epidemiologists goes here!\\
-% \\\\
-% \textbf{From VMV'13 Paper}\\
-% 
-% All whole-body MRI scans were acquired on a 1.5~Tesla scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany) by four trained technicians in a standardized way.
-Subjects were placed in the supine position.
-Five phased-array surface coils were placed to the head, neck, abdomen, pelvis, and lower extremities for whole-body imaging.
-The spine coil is embedded in the patient table.
-The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence (676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels).
-First, both sequences were placed over the cervical and upper thoracic spine.
-Then, they were placed over the lower thoracic and lumbar spine.
-The MRI software automatically composed a whole spine sequence from the two T1-weighted and T2-weighted sequences \cite{Hegenscheid2013}.
-We were provided with 490 data sets.
-
-% The model is placed in the scene using an empirically chosen initialization point.
-The force acting on the model stems from aggregation of loads, which are derived from a potential field resulting from a weighted sum of the T1- and T2-weighted MRI images, see \cite{Rak2013}.
-After detecting all spines, we register the models because in a later clustering step we only want to capture the local deformation of the lumbar spine, not different locations in world space.
-The models are registered using the Kabsch Algorithm, which is designed to minimize the root mean squared deviation between paired sets of points.
-% The model-based detection captures information about the spine canal curvature as well as the alignment of the vertebrae.
-It is not meant to capture information about vertebrae deformation and differences in spine canal extent.
 
@@
 - Color code different Aspects - Only difference in x or y direction
-
-Discretization of continous variables into bins of equal size may bias the underlying data by not considering the variable distribution function \cite{Fletcher2012}.
+- Calculation of contingency based on the contingency differences - how does the current selection differ compared to the whole data set?
+
+Discretization of continuos variables into bins of equal size may bias the underlying data by not considering the variable distribution function \cite{Fletcher2012}.
 % Fletcher 2012 - S. 257

### 2014-03-24 draft.tex
Reworked Introduction, Background and Related Work

@@
 %% \section{Introduction} %for journal use above \firstsection{..} instead
-Epidemiology aims to characterize health and disease by determining risk factors.
+\textbf{(Background)} Epidemiology aims at characterizing health and disease by determining risk factors.
 %
@@
 Also, the introduction of preventive measures in medicine and beyond, are based on epidemiological research, where, for example, subgroups with increased risk are identified \cite{Fletcher2012}.
-%Clinical problems and questions answered using epidemiological methods comprise diagnosis accuracy, disease frequency, risk factors, disease prognosis, effectiveness of treatments or preventions and cause of diseases \cite{Fletcher2012}.
 %
@@
 %
-
-Longitudinal population-based studies, such as the Study of Health in Pomerania \cite{Volzke2011}, aim to gather as much information as possible about a defined sample of people (a \emph{cohort}).
-%
-The sample is drawn randomized to avoid selection \cite{Fletcher2012}.
-%bias which prohibits statements based on statistical correlations in the cohort from being inferred to the whole population (\emph{external validity}) \cite{Fletcher2012}.
-%
-%Also a information bias needs to be avoided by strictly standardizing the data acquisition.
-%
-When for example one investigates risk factors for prostate cancer in male subjects, the outcome is strongly dependent on the age.
-%
-Therefore results need to be age-adjusted to be comparable.
-%
-Confounding variables, are often not obvious at all and characterizing them is already an epidemiological result.
-\\\\
+\\\\
+\textbf{(Problem)}
+Longitudinal population-based studies, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}, aim to gather as much information as possible about a defined sample of people (a \emph{cohort}).
+%
+% The sample is drawn randomized to avoid selection \cite{Fletcher2012}.
+% %
+% When for example one investigates risk factors for prostate cancer in male subjects, the outcome is strongly dependent on the age.
+% %
+% Therefore results need to be age-adjusted to be comparable.
+% %
+% Confounding variables, are often not obvious at all and characterizing them is already an epidemiological result.
+%
 Modern cohort studies often include medical image data.
 %
-Since it is unethical to expose people to radiation, non-harming imaging such as Magnetic Resonance Imaging (MRI) or Ultrasound Imaging is used.
-%
-%As MRI scans are time-consuming and expensive there exists a tradeoff between quality of the image data and their associated costs.
-%
-To quantify these data it is necessary to label each voxel regarding structure affiliation (\emph{segmentation}).
-%
-Manual segmentation carried out by radiological experts is possible but very costly and prone to inter- and intra observer variability.
-%
-Segmentation algorithms allow for (semi)-automated analysis of the data but require sophisticated methods due to high inter-subject variability caused by the subject diversity.
+These data needs to be segmented to analyze its features for correlations with non-image data.
+%
+Manual segmentation is not feasible since it leads to a high variability which is at odds with the high reproducibility required in epidemiology research.
+%
+Semi-automatic techniques are more promising but challenging since the used methods as MRI and ultrasound are subject to inhomogeneity and noise.
 %
 Analyzing spatial data with respect to other epidemiological factors requires techniques which reach beyond standard statistical methods.
-\\\\
-We propose an Interactive Visual Analysis approach \cite{Thomas2005} to provide a way to analyze both image- and non-image data.
+
+Compiling a list of features for tests of statistical resilience based on experience-driven hypotheses leaves out other features in the data which potentially interact with a disease.
+%
+This also applies to the chosen landmarks which are used to quantify medical image data information.
+%
+The standard workflow lacks methods which highlight features of interest the epidemiologists did not consider.
+%The hypothesis driven analysis approach of these data based on expectations of epidemiologists leaves out features which potentially interact with a disease outside.
+%The hypothesis driven analysis approach of these data leaves out features which potentially interact with a disease because they they are it is outside of the erwartungshaltung of the epidemiologists.
+%leaves out features which potentially interact with a disease because they are usually outside of the practical experience of the epidemiologists.
+%
+%Analyzing these data for correlating features requires preprocessing steps as segmentation of the structures.
+%
+%Manual segmentation is not feasible since it leads to a high variability which is att odds with the high reproducibility required in epidemiology research. 
+%
+%Semi-automatic techniques are more promising but challenging since MRI and ultrasound are subject to inhomogeneity and noise
+%Due to the lack of sophisticated segmentation algorithms this is often achieved by manually specifying landmarks by radiologists which is costly and prone to inter- and intra observer variability.
+%
+% Since it is unethical to expose people to radiation, non-harming imaging such as Magnetic Resonance Imaging (MRI) or Ultrasound Imaging is used.
+% %
+% To quantify these data it is necessary to label each voxel regarding structure affiliation (\emph{segmentation}).
+% %
+% Manual segmentation carried out by radiological experts is possible but very costly and prone to inter- and intra observer variability.
+% %
+% Segmentation algorithms allow for (semi)-automated analysis of the data but require sophisticated methods due to high inter-subject variability caused by the subject diversity.
+% %
+% Analyzing spatial data with respect to other epidemiological factors requires techniques which reach beyond standard statistical methods.
+\\\\
+\textbf{(Goals)}
+We propose an Interactive Visual Analysis approach \cite{Thomas2005} to provide a way to analyze image- and non-image data.
 %
@@
 %
-In order to characterize the healthy aging process we aim to determine changes for subjects which indicate unusual pathological changes.
-\\\\
+Easy publishing of developed methods driven by modern web technologies intends to trigger a fast feedback loop between us and the epidemiologists.
+% http://wiki.infowiss.net/Gulf_of_execution ?
+We applied our approach to a data set compiled to analyze diseases related to the lumbar spine and aim to determine features, which indicate pathological changes.
+\\\\
+\textbf{(Contributions)}
 Our contributions are:
 \begin{itemize}
-	\item Applying the Interactive Visual Analysis technique set to the epidemiological problem domain by characterizing requirements of this context.
-	\item Provide an overview over the workflow for analyzing cohort study data to gain insight into the large diverse subject spaces.
-	\item Provide visualization techniques which combine both information visualization and 3D rendering of organ shapes as well as combining them with well known epidemiological graphics and key figures.
-	\item Implement the presented methods as a web framework based on WebGL, D3JS and NodeJS.
+%	\item providing an overview over the workflow for analyzing cohort study data to gain insight into the large diverse subject spaces.
+	\item describing an Interactive Visual Analysis workflow for analyzing image-based epidemiological data including both hypothesis-driven and for hypothesis-generation based on a characterization of the standard epidemiological workflow
+	\item providing visualization techniques which combine both information visualization and 3D rendering of organ shapes as well as combining them with well known epidemiological graphics and key figures.
+	\item highlighting interesting subject groups and feature associations using shape-based clustering and statistical contingency measures
+	%Applying Interactive Visual Analysis toolset to image-based analysis in epidemiology with focus hypothesis generation
+	%\item Applying Interactive Visual Analysis toolset to image-based analysis in epidemiology with focus hypothesis generation
+	% \item Describing a Interactive Visual Analysis system aiming to support hypothesis generation .. of image-based cohort study analysis
+% 		\item Characterize the problems of epidemiological context which  %Applying the Interactive Visual Analysis technique set to the epidemiological problem domain by characterizing requirements of this context.
+% 	\item Applying the Interactive Visual Analysis technique set to the epidemiological problem domain by characterizing requirements of this context.
+% 	\item Provide an overview over the workflow for analyzing cohort study data to gain insight into the large diverse subject spaces.
+% 	\item Provide visualization techniques which combine both information visualization and 3D rendering of organ shapes as well as combining them with well known epidemiological graphics and key figures.
+	\item implementing the presented methods as a web framework based on WebGL, D3JS and NodeJS.
 \end{itemize}
@@
 
-\section{Medical and Technical Background} \label{MedicalAndTechnicalBackground}
+\section{Epidemiological and Technical Background} \label{MedicalAndTechnicalBackground}
 % Wer ist an epidemiologischen Studien beteiligt?\\
@@
 \subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
-Many different experts work at epidemiological studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
+%Epidemiology diversity is reflected in the different experts working at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
+The diversity of epidemiology is reflected in the different experts who work at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
+%Many different experts work at epidemiological studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
 %
@@
 \begin{itemize}
-	\item Hypothesis most commonly base on observations made by clinicians in their daily routine.
+	\item Hypotheses are based most commonly on observations made by clinicians in their daily routine.
 %
@@
 % TODO: Bernhard: Effect size of Attribute ?
-	\item Confounding variables need to be adjusted so that they do not affect the effect size of a attribute.
+	\item Confounding features are adjusted so that they do not affect the effect size of a attribute.
 %
@@
 
-Reproducibility of results is an key requirement in epidemiology.
+Reproducibility of results is an epidemiological key requirement.
 %
@@
 %This underlines the high quality standards to methods processing the data, whether to extract additional parameters or gain insight.
-%
+
+Grouping subjects using epidemiological features is essential in order to allow per-group risk determination.
+%
+%Grouping is carried out hypothesis driven.
+Grouping depends on the underlying hypothesis.
+%
+Age for example is divided into groups (e.g. in 20 year-steps) when investigating its influence on a condition.
+%
+These groups depend strongly on the condition of interest and therefore there is no defined standard on how to categorize these values.
+
 To determine, whether a subject is prone to be affected by a certain disease, \emph{relative risks} are expressed through the evaluation of p-values which indicate statistical significance.
 %
-Statistical correlations are prone to \emph{confounding}, meaning that two factors are dependent and therefore should be normalized with respect to each other.
-%
-Statistics tools such as SPSS and STATA play a major role for analyzing epidemiological data.
+Statistical correlations are prone to \emph{confounding}, meaning that the interaction of two features are influenced by a third feature which needs to isolated.
+% Statistical correlations are prone to \emph{confounding}, meaning that two features are dependent and therefore should be normalized with respect to each other.
+%
+Statistics tools such as \texttt{SPSS}\footnote{Product of IBM; \url{www.ibm.com/software/de/analytics/spss/}} and \texttt{Stata}\footnote{Product of Stata Corp.; \url{http://www.stata.com/}} play a major role for analyzing epidemiological data.
 %
 Graphic data representation is largely used to present results rather than gaining insight.
-
-Grouping subjects using epidemiological factors is essential in order to allow per-group risk determination.
-%
-Grouping is carried out hypothesis driven.
-%
-Age for example is also divided into groups (e.g. in 20 year-steps) when investigating its influence on a condition.
-%
-These groups depend strongly on the condition of interest and therefore there is no defined standard on how to categorize these values.
 
@@
 %
-This underlines also the problem of missing data since for ethical, legal or medical reasons some data variables can not be gathered for each subject.
-%
-Follow-up examinations or -questions for conditions also produce variables only available for a small amount of subjects.
+This underlines also the problem of missing data since for ethical, legal or medical reasons some features can not be gathered for each subject.
+%
+Follow-up examinations or -questions for conditions also yield features only available for a small amount of subjects.
 %
@@
 %
-Dichotomous data can also be derived by combining aggregating data variables to yield only two manifestations (e.g. subjects younger or older than 50).
+Dichotomous data can also be derived by aggregating features to yield only two manifestations (e.g. subjects younger or older than 50 years).
 %
@@
 This often yields image resolutions inferior to those of clinical day-to-day practice, which makes their analysis more challenging.
+%
+The equipment used to gather medical image data is kept on the initial software and hardware version to ensure comparability in and between acquisition cycles.
 
@@
 %
-Segmentation masks labeling the voxel of a shape of interest would be ideal since many different key figures can be derived from them.
-%
-Since these masks require sophisticated algorithms custom tailored to the data sets the epidemiologists are forced to measure the data by hand, which is a very tedious work with respect to the number of necessary landmarks and number of subjects.
+Segmentation masks labeling the voxels of an anatomical structure would be ideal since many different key figures, e.g. volume, largest diameter or aspect ratio, can be derived from them.
+%
+Since reliable and efficient segmentation techniques for these data are not available in general, the epidemiologists are forced to measure the data by hand, which is a very tedious work with respect to the number of necessary landmarks and number of subjects.
 %
@@
 %
-This gains even more momentum when analyzing multiple time steps!
+This gains even more momentum when analyzing multiple time steps.
 %
@@
 \subsection{The Study of Health in Pomerania (SHIP)}
-Starting 1997 with a cohort consisting of 4.308 subjects this cohort study located in northern Germany aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
-%
-Data is collected without focus on a specific disease.
+After the pioneering Rotterdam study (started in 1990) several image-centric cohort study initiatives have evolved, they slightly differ in clinical focus, acquired data and epidemiologic research questions.
+%
+Starting 1997 with a cohort consisting of 4.308 subjects, the SHIP, located in northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
+%
+Data is collected without focus on a group of diseases.
 %
@@
 %
-The MRI protocol features a rich number of different sequences.
-%
-Also for women, breast MRI scans were acquired.
+The MRI protocol features a rich number of sequences.
+%
+%In addition to other MRI data, breast MRI scans were acquired for women.
 %
@@
 \section{Prior and Related Work}
-Einfuehren von Helwigs Terminologie?
-
-Designing a visualization which communicates all aspects of the data equally is challenging.
-%
-Given the number of features of epidemiological data sets and their different manifestations, it is often a good solution to combine the strength of different visualization techniques in a unified system \cite{Buja91, Konyha2009}.
+%Einfuehren von Helwigs Terminologie?
+Designing a visualization which conveys all data aspects equally is challenging.
+%
+Given the number of features of epidemiological data sets and their different manifestations, the strength of different visualization techniques need to be combined \cite{Buja91, Konyha2009}.
 %
@@
 %
-Investigating Data on an norwegian aging study their methods aim to amplify a hypothesis generation process.
-%
-Statistical measures of continuous variables such as mean, standard deviation, skewness, or inter-quartile range are used to create \emph{dimension plots}.
-%
-These transform dimensions into data points and make them comparable with respect to the derived descriptive measures.
-%
-This not only allows for comparing all continuous variables in a single plot but make their distribution change comprehensible.
-%
-This requires a good descriptive measure which captures the kind of change the user is interested in or which reflects unexpected data behavior.
-%
-The technique was applied to variables generated by segmenting the brain into 45 parts and measure the voxel number, volume and properties of the intensity values.
+Their methods aim to amplify a hypothesis generation process for analyzing data of a Norwegian aging study.
+%
+Statistical measures of continuous variables such as mean, standard deviation, skewness, or inter-quartile range are used to create \emph{dimension plots} that make them comparable with respect to the derived descriptive measures.
+%
+%These transform dimensions into data points and make them comparable with respect to the derived descriptive measures, making them comparable.
+%
+%This not only allows for comparing all continuous variables in a single plot but make their distribution change comprehensible.
+%
+%This requires a good descriptive measure which captures the kind of change the user is interested in or which reflects unexpected data behavior.
+%
+%The technique was applied to variables generated by segmenting the brain into 45 parts and measure the voxel number, volume and properties of the intensity values.
 %
 The method is strongly dependent on the descriptive measures of the epidemiological factors.
-%
-Hypothesis based on observations of changes in these plots may impose over-fitting to the data because the measure highlights only subsets of statistical changes.
+
+%\emph{Over-fitting} of hypotheses to the data may be imposed  is a danger pointed out by the authors when hypotheses to the data based on observations of changes in these plots the measure highlights only subsets of statistical changes.
+Hypotheses based on observations of changes in these plots may impose \emph{over-fitting} to the data because the measure highlights only subsets of statistical changes.
 %
@@
 
+\paragraph{Visualizing Image and Non-Image Data. }
 Gresh and colleagues proposed \texttt{WEAVE}, one of the first systems which analyzed medical image and non-image data using linked views \cite{Gresh2000}.
@@
 %
-This approach already incorporated Data Mining approaches such as dividing the data space using a k-nearest-neighbor technique and Principal Component Analysis.
+They incorporated data mining methods such as dividing the data space using a k-nearest-neighbor technique and the PCA.
 %
@@
 %
-Their understanding of a cohort differs from the understanding of the term in an epidemiolgical context.
-
-Commercial systems such as \texttt{Tableau} or \texttt{Spotfire} provide a rich user interface that allows to apply Visual Analytics techniques without the need of writing any code.
-%
-With little effort, linked views can be created using these tools, but the data processing possibilities such as derivation of new variables or the volume rendering capabilities are very limited.
+Their understanding of a cohort differs from the understanding of the term in an epidemiological context.
+
+\paragraph{Visualizing Heterogenous Non-Image Data.}
+Generalized Pairs Plots (\texttt{GPLOM´S}) are an information visualization technique comparing heterogenous variables pairwise using a plot-matrix grouped by type \cite{Francois2013}.
+%
+It is useful to gain an overview over numerous variables and their distributions.
+%This technique is also useful to gain an overview over numerous variables and their distributions.
+%
+Histograms, bar charts, scatterplots and heat maps are used to visualize variable combinations with regard to their type.
+%
+%Brushing, linking and filtering are feasible with \texttt{GPLOMS}, but have limitations such as making only one category brushable at a time.
+%
+%We applied this technique to our data and it shows promising potential for simultaneously visualizing many different variables. %but does not fit in the scope of this paper.
+%
+%The inspiration on the chosen visualization techniques stems from this publication.
+%
+Dai and colleagues explored risk factors by incorporating choropleth maps of epidemiological features (e.g.
+mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines \cite{Dai2005}.
+%A similar approach was taken by Dai and colleagues for risk factor exploration as they also incorporate choropleth maps of epidemiological factors (e.g.
+mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines \cite{Dai2005}.
+%
+Their findings yielded a \emph{Concept Map} which linked cancer-related interactions via graph edges.
+%From their findings regarding the interaction of cancer-related socio-demographic factors are drawn in a \emph{Concept Map} where related factors are connected via graph-edges.
+Chui and colleagues visualized interactions in time-dependent epidemiological data using time-series plots highlighting risk factors differences in age and gender \cite{Chui2011}.
+
+\paragraph{Commercial Data Visualization.}
+Commercial systems such as \texttt{Tableau}\footnote{Owned by Tableau Software; \url{www.tableausoftware.com}} or \texttt{Spotfire}\footnote{Owned by TIBCO; \url{spotfire.tibco.com}} provide a rich user interface that allows to apply Visual Analytics techniques without the need of writing any code.
+%
+With little effort, linked views can be created, but the data processing possibilities such as derivation of new variables or the volume rendering capabilities are very limited.
 %
@@
 %Spotfire offers the possibilities to interact with the \texttt{R} statistical programming language 
+% Image Data
+\paragraph{Visualizing Image Data.}
+Comparing tissue between many subjects in an epidemiological context requires methods which allow for shape variance visualizations.
+%
+Caban and colleagues investigated the suitability of variance visualizations of shape distribution models and concluded in their user study that users favor spherical glyph representations over deformation grids and likelihood volumes \cite{Caban2011}.
+%
+The distribution of shapes in a space derived from a PCA is plotted by Busking and Colleagues in a 2D-projected plane of the space \cite{Busking2010a}.
+%
+Via mesh morphing interpolated views can be created by the user in a separate view as well as comparisons in a contour view.
+%
+Distance to the mean shape is color-coded.
+%Differences between structures are highlighted using color mapping of the difference to the mean shape, but is rather hard to recognize due to small renderings of each subject in the shape-space.
+%Via mesh morphing interpolated views can be created by the user in a separate view as well as comparisons in a contour view.
+%
+We incorporate the idea of combining 3D-Shape rendering with information visualization techniques.
+%
+%Applying our data sets to this technique yielded a cluttered shape space due to the many subjects.
+%
+%The data needs to abstracted or summarized in order to work in this context.
+%
+Hermann and colleagues identify local deformation changes by investigating shape related difference \cite{Hermann2014}. 
+%
+The user specifies a deformation of interest and showing corresponding changes in the shape using covariance tensors.
+%
+This method allowed for rapid hypotheses validation and was able to reproduce textbook knowledge.
+%
+%By plotting p-values in ventricle surfaces, Chou and colleagues were able to map disease-associated values directly on a 3D tissue representations \cite{Chou2009}.
+%
+%This requires a geographic colocation of associated features.
+%
 
 % *********************** Our own Work ***********************
-Klemm and colleagues used lumbar spine variabilities based on an semi-automatic shape-detection algorithm of 490 participants of the \texttt{SHIP-2} \cite{Klemm2013VMV}.
+% Prior Work
+\paragraph{SHIP-Data analysis.}
+Klemm and colleagues visualized lumbar spine variabilities based on an semi-automatic shape-detection algorithm of 490 participants of the \texttt{SHIP-2} cohort \cite{Klemm2013VMV}.
 %
@@
 %
-As proof of concept a relation between size of the segmented shape and measured size of the subjects was measured.
+As proof of concept a relation between size of the segmented shape and measured size of the subjects was shown.
 %
@@
 
-Generalized Pairs Plots (\texttt{GPLOM´S}) are an information visualization technique that allows for heterogenous variables to be pairwise compared using appropriate plots in a plot matrix grouped by type \cite{Francois2013}.
-%
-This technique is also useful to gain an overview over numerous variables and their distributions.
-%
-It uses histograms, bar charts, scatterplots and heat maps to visualize the different variable combinations with regard to their type.
-%
-Brushing, linking and filtering are feasible with \emph{GPLOMS}, but have limitations such as making only one category brushable at a time.
-%
-%We applied this technique to our data and it shows promising potential for simultaneously visualizing many different variables. %but does not fit in the scope of this paper.
-%
-%The inspiration on the chosen visualization techniques stems from this publication.
-%
-A similar approach was taken by Dai and colleagues for risk factor exploration as they also incorporate choropleth maps of epidemiological factors (e.g.
-mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines \cite{Dai2005}.
-%
-From their findings regarding the interaction of cancer-related socio-demographic factors are drawn in a \emph{Concept Map} where related factors are connected via graph-edges.
-
-Chui and colleagues visualized interactions in time-dependant epidemiological data using time-series plots highlighting risk factors differences in age and gender \cite{Chui2011}.
-%
-\\\\
-% Image Data
-Comparing tissue between many subjects in an epidemiological context requires methods which allow for shape variance visualizations.
-%
-Caban and colleagues investigated the suitability of variance visualizations of shape distribution models and concluded in their user study that users favor spherical glyph representations over deformation grids and likelihood volumes \cite{Caban2011}.
-%
-The distribution of shapes in a space derived from a PCA is plotted by Busking and Colleagues in a 2D-projected plane of the space \cite{Busking2010a}.
-%
-We incorporate the idea of combining 3D-Shape rendering with information visualization techniques.
-%
-Differences between structures are highlighted using color mapping of the difference to the mean shape, but is rather hard to recognize due to small renderings of each subject in the shape-space.
-%
-Via mesh morphing interpolated views can be created by the user in a separate view as well as comparisons in a contour view.
-%
-%Applying our data sets to this technique yielded a cluttered shape space due to the many subjects.
-%
-The data needs to abstracted or summarized in order to work in this context.
-%
-In order to detect local deformation changes, Hermann and colleagues investigating shape related difference by letting the user specify a deformation of interest and showing corresponding changes in the shape using covariance tensors \cite{Hermann2014}.
-%
-This method allowed for rapid hypothesis validation and was able to reproduce textbook knowledge.
-%
-By plotting p-values in ventricle surfaces, Chou and colleagues were able to map disease-associated values directly on a 3D tissue representations \cite{Chou2009}.
-%
-This requires a geographic colocation of associated features.
-%
-
+\paragraph{Interactive Visual Analysis}
 % Interactive Visual Analysis - Steffens Paper
@@
 %
-Oeltze and colleagues combined a linked view representation of results from a statistical analysis with feature localizations of the human blood flow with the goal of its evaluation \cite{Oeltze2007}.
-%
-%They aim to provide methods to analyze perfusion data with respect to 
-While we take similar steps when analyzing the data such as employing statistical tests, our data is mostly independent from the medical image data and is not describing it--except the variables derived the segmentation model itself.
+Oeltze and colleagues combined a linked view representation of results from a statistical analysis with feature localizations of the tissue perfusion flow with the goal of its evaluation \cite{Oeltze2007}.
+%
+
+While we take similar steps when analyzing the data such as employing statistical tests, our data is mostly independent from the medical image data and is not describing it.
+%--except the variables derived the segmentation model itself.
 
@@
 %
-Due to the different acquisition modalities there have to be different techniques incorporated.
+Multimodal features require different techniques.
+%Due to the different acquisition modalities there have to be different techniques incorporated.
 %
@@
 \\\\
-Processing the image data associated to each subject consists for the most part of information extraction about a structure of interest.
+Processing the image data associated to each subject consists for the most part of information extraction about a anatomical structure.
 %
@@
 %
-Algorithms, applied to the data, do not only have to deal with a large inter-subject variability of the structure of interest but also need to be reproducible \cite{Preim2014}.
+Algorithms, applied to the data, do not only have to deal with a large inter-subject variability of the anatomical structure, but also need to be reproducible \cite{Preim2014}.
 %

### 2014-03-25 draft.tex
Reworked Chapter 4

@@
 %(New Solution).
-We propose an Interactive Visual Analysis approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic variables and attributes derived from the image data.
+We propose an Interactive Visual Analysis (\emph{IVA}) approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic variables and attributes derived from the image data.
 %
@@
 \textbf{(Goals)}
-We propose an Interactive Visual Analysis approach \cite{Thomas2005} to provide a way to analyze image- and non-image data.
+We propose an Interactive Visual Analysis (\emph{IVA}) approach \cite{Thomas2005} to provide a way to analyze image- and non-image data.
 %
@@
 %\emph{Over-fitting} of hypotheses to the data may be imposed  is a danger pointed out by the authors when hypotheses to the data based on observations of changes in these plots the measure highlights only subsets of statistical changes.
-Hypotheses based on observations of changes in these plots may impose \emph{over-fitting} to the data because the measure highlights only subsets of statistical changes.
+Hypotheses based on observations of changes in these plots may impose \emph{overfitting} to the data because the measure highlights only subsets of statistical changes.
 %
@@
 % Interactive Visual Analysis - Steffens Paper
-The strength of the Interactive Visual Analysis approach described in the next section is its versatility with respect to the application field \cite{Konyha2009}.
+The strength of the \emph{IVA} approach described in the next section is its versatility with respect to the application field \cite{Konyha2009}.
 %
@@
 
-\section{Interactive Visual Analyis in Cohort Study Data}
+%\section{Interactive Visual Analyis in Cohort Study Data}
+\section{Image Centric Cohort Study Data in Interactive Visual Analysis Context} \label{Image Centric Cohort Study Data in Interactive Visual Analysis Context}
 \begin{figure}[htb]
  \centering
- \label{fig:WorkflowComparison}
  \includegraphics[width=3.0in]{figures/workflow_comparison}
- \caption{ToDo Abbildung noch nicht final.
-Visual Analytics systems are able to complement parts of the epidemiological workflow, not replace it.
+ \caption{\emph{IVA} tools are able to complement parts of the epidemiological workflow, not replace it.
 The appropriate combination of statistical- and interactive driven analysis shows promising potential to unveil the information in the data.
@@
 The iterative red highlighted part is called the \emph{IVA Loop}.}
+  \label{fig:WorkflowComparison}
 \end{figure}
-- Acceptance plays a big role - you have to pick up the user by presenting information they are used to in a new way\\
-%
-As described in subsection~\ref{EpidemiologicalWorkflow}, the epidemiolgical workflow is a sequence of steps taken by domain experts and needs to be reproducible and comprise statistical integrity.
-%
-Figure~\ref{fig:WorkflowComparison} (a) describes this workflow as consecutive series of steps.
-%
-The workflow we propose by introducing the \emph{IVA} principle into the epidemiological application domain does not aim to replace the existing workflow but to complement its weaknesses.
+%- Acceptance plays a big role - you have to pick up the user by presenting information they are used to in a new way\\
+%
+Subsection~\ref{EpidemiologicalWorkflow} described the epidemiological workflow as a sequence of steps taken by domain experts which need to comprise reproducibility and statistical integrity.
+%
+Figure~\ref{fig:WorkflowComparison} (a) shows this workflow as consecutive series of steps.
+%
+Introducing the \emph{IVA} principle to the epidemiological application domain does aim to complement its weaknesses rather than replacing the existing workflow.
+%The workflow we propose by introducing the \emph{IVA} principle into the epidemiological application domain does not aim to replace the existing workflow but to complement its weaknesses.
 %
@@
 %
-A list of features describing the hypothesis is compiled and analyzed using statistical tests. 
-%
-The resulting value decides whether the data supports the hypothesis or not.
-%
-It would be possible that there are actually features of the data set which support the hypothesis by discriminating the population in the expected way, but with this approach they are not highlighted in any way.
-%
-This becomes even more important when the workflow is adapted to the analysis of the medical image data.
-%
-Domain experts would have to annotate landmarks tediously to derive metrics such as distances. %which are then handled like other features and analyzed using the same set of statistical tools.
-%
-Not only does this leave out the majority of the information in the medical image data by abstracting it to single values, it is easily possible that information left out would heavily influence the result.
+Statistical tests on a hypothesis-describing feature list yields a value deciding whether the data supports the hypothesis or not.
+%A list of features describing the hypothesis is compiled and analyzed using statistical tests. 
+%
+%The resulting value decides whether the data supports the hypothesis or not.
+%
+Features not included in the analysis may potentially support the chosen hypothesis by discriminating the population in the expected way, but are not highlighted in any way.
+%It would be possible that there are actually features of the data set which support the hypothesis by discriminating the population in the expected way, but with this approach they are not highlighted in any way.
+%
+%Adapting this workflow to medical image data analysis also highlights this problem.
+This becomes even more important when the workflow is adapted to the analysis of the medical image data, where
+domain experts annotate landmarks tediously to derive metrics such as diameters. %which are then handled like other features and analyzed using the same set of statistical tools.
+%
+This leaves out the majority of information in the image data by abstracting it to single values.
+%
+It is easily possible that information left out would heavily influence the result.
+%Not only does this leave out the majority of the information in the medical image data by abstracting it to single values, it is easily possible that information left out would heavily influence the result.
 %
 Considering more complex parts of the data would make those results more trustworthy and also could identify possible anatomical confounders--an epidemiolgical research result in itself.
-%
+
 Statistical tests check for validity of the number but not for their completeness or plausibility!
 \\\\
-\emph{IVA} tries to illuminate the black box by making the domain experts part of the feature list selection process.
-%
-Figure~\ref{fig:WorkflowComparison} (b) highlights the iterative process as part of the epidemiolgical workflow.
-%
-Note that it also aims to project back into the hypothesis formulation step to amplify hypothesis generation 
-%
-This has to be handled with care since overfitting of expectations to the available data is an imminent danger as described by Turkay and colleagues \cite{Turkay2013}.
-
-\subsection{Image Centric Cohort Study Data in Interactive Visual Analytics Context}
+\emph{IVA} tries to illuminate the black box by making the domain experts part of an iterative feature list selection process, which is shown in Figure~\ref{fig:WorkflowComparison} (b) as part of the epidemiological workflow.
+%
+%Figure~\ref{fig:WorkflowComparison} (b) highlights the iterative process as part of the epidemiolgical workflow.
+%
+It also aims to project back into the hypothesis formulation step to amplify hypothesis generation.
+%Note that it also aims to project back into the hypothesis formulation step to amplify hypothesis generation 
+%
+This has to be handled with care since \emph{overfitting} of expectations to the data is an imminent danger as described by Turkay and colleagues \cite{Turkay2013}.
+
+\paragraph{Domain and Range Variables. }
 In the \emph{IVA} context, data is divided into two major view types.
@@
 %
-These variables are also referred to as \emph{independent variables}.
-%
-\emph{Dependent variables} in the epidemiological context can be divided twofold:
+These variables spanning the 3D-space are also referred to as \emph{domain variables}.
+%
+\emph{Range variables} in the epidemiological context can be divided twofold:
 %
@@
 	%
-	These measures abstracts shape information as quantification to allow for comparison. 
-	%
-	These variables describe image data and can also be used to brush in the image space.
+	These measures quantify shape information to allow for comparison and can also be used to brush in the image space.
+	%
+	%These variables describe image data and can also be used to brush in the image space.
 	\item Epidemiological socio-demographic or medical attribute data.
 	%
-	These values belong to every subject which is represented in the image space, but does not describe shape information.
+	These values are associated with every subject represented in the image space. 
+	%
+	They do not describe shape information.
 	%
@@
 
-\subsection{Data Preprocessing}
-To include heterogenous epidemiological data in an \emph{IVA}-framework it is necessary to process it to obtain standardized views to the available features.
-%
-Multimodal features require different techniques.
-%Due to the different acquisition modalities there have to be different techniques incorporated.
-%
-Data obtained using questionnaires or medical tests are often stored using statistical packages such as \texttt{SPSS} or \texttt{Stata} which have a proprietary data format with limited export capabilities.
-%
-The best solution for us was to simply export the data in the respective tool to a character separated text file and then convert it to data types which are easier manageable such as JSON or XML using our own classes.
-%
-In order to verify that the conversion worked as expected and the data is valid, it is good practice to use data wrangling tools such as \texttt{OpenRefine} to validate the data (find missing data, clean up bad formatting).
-%
-Exporting the data dictionary, which stores information about each manifestation of a feature is also an important step to get a detailed description of data variables and the meaning and unit of measurement of their values.
-%
-The reasons for missing data have a wide range from ethical to medical and personal issues.
-%
-Therefore, these are also included as error codes which have to marked as such in the data dictionary.
-\\\\
-Processing the image data associated to each subject consists for the most part of information extraction about a anatomical structure.
-%
-This is either done manually by experts setting landmarks (sometimes supported by algorithms connecting the land marks such as graph cuts \cite{GraphCut}) or by a (semi)-automatic detection, registration and segmentation.
-%
-Algorithms, applied to the data, do not only have to deal with a large inter-subject variability of the anatomical structure, but also need to be reproducible \cite{Preim2014}.
-%
-Model-based approaches have shown in principle to be effective for segmentation \cite{Gloger2010, Gloger2012} and detection \cite{Rak2013}.
-%
-If a segmentation yields only binary masks separating the structures, algorithms such as Growing and Adaptive Shapes can be applied creating a surface grid where each point is comparable throughout the population \cite{Ferrarini2007}.
-%
-Intensity-based comparisons can be achieved using rigid image registration, but model based results however are preferable \cite{Klemm2012}.
-%
-Comparison based on grey values is usually carried out to to measuring the quantity of fat, water, and--application specific--iron content (liver) or distribution of grey and white brain tissue.
-
-Morphometric variables are derived to allow for statistical comparison of the tissue which incorporate mostly position, volume and relative distances and alignment to other structures.
-% 
-% - Data Format\\
-% - Missing Data\\
-% - Registration of Image Data (VMV'12)\\
-% - Extraction of Image Parameter
-
 \subsection{IVA Patterns}
-
-The explorative procedures when analyzing data using \emph{IVA} can be divided into three different patterns, handling interaction between domain and range variables.
-
-\subsubsection{Local Investigation}
-This pattern projects information from image space to the range perspective.
-%
-As opposing to other \emph{IVA} application domains, this step is more complicated in the epidemiological context.
-%
-Shape information can not be brushed by incorporating ROI-selections but rather has to employ techniques that specify local deformation changes \cite{Hermann2014} or subjects that belong to a shape class.
-%
-Methods available for \emph{feature selection} strongly depend on the type of registration that was applied to extract the tissue of interest.
+%
+\emph{IVA} knows three different exploration patterns, handling interaction of domain and range variables.
+%The explorative procedures when analyzing data using \emph{IVA} can be divided into three different patterns, handling interaction between domain and range variables.
+
+\paragraph{Local Investigation}
+This pattern projects information from domain space to the range perspective.
+%
+This step is more complicated in epidemiological context compared to other \emph{IVA} application domains.
+%As opposing to other \emph{IVA} application domains, this step is more complicated in the epidemiological context.
+%
+Shape information can not be brushed using ROI-selections but rather has to employ techniques that specify local deformation changes \cite{Hermann2014} or subjects that belong to a shape class.
+%
+\emph{Feature selection} methods strongly depend on the type of segmentation used to extract the tissue of interest.
 %
@@
 %
-Another application is the outlier analysis.
-%
-Outliers can indicate segmentation errors or an outstanding group of individuals who may share a pathology.
+Analysis of outliers can indicate segmentation errors or an group of subjects sharing a pathology.
+%
+%Another application is the outlier analysis. 
+%
+%Outliers can indicate segmentation errors or an outstanding group of individuals who may share a pathology.
 
@@
 
-\subsubsection{Feature Localization}
+\paragraph{Feature Localization}
 % Subset selection of range variables yields features
-As described before, the vast majority of data points are considered to be dependent with respect to the image domain in the \emph{IVA} context.
+The vast majority of features are considered to be dependent with respect to the image domain in the \emph{IVA} context.
+%As described before, the vast majority of data points are considered to be dependent with respect to the image domain in the \emph{IVA} context.
 %
@@
 %
-Categorization is strongly dependent on the hypothesis and therefore requires suitable brushing techniques as described in Section~\ref{sec:AdaptiveFeatureVisualization}.
+%Categorization is strongly dependent on the hypothesis and therefore requires suitable brushing techniques as described in Section~\ref{sec:AdaptiveFeatureVisualization}.
 %
@@
 
-\subsubsection{Multivariate Analysis}
+\paragraph{Multivariate Analysis}
 %Introduced in the information visualization, the multivariate analysis means selecting subjects in like in the feature localization step and get have the result highlighted in another linked view displaying other non-image parameter.
-Introduced in the information visualization, the multivariate analysis incorporates brushing and linking of of views displaying non-image parameter.
+%Introduced in the information visualization, the multivariate analysis incorporates brushing and linking of of views displaying non-image parameter.
+Multivariate analysis incorporates brushing and linking of of views displaying non-image parameter.
+%
+The type of the parameters which are compared determines the information visualization technique used.
 %
@@
 %
-The need of statistic measures which describe how variables correlate with each other given the selected groups is special for the application domain.
-%
-These associations also summarized using Pivot Tables which are popular in epidemiology and which are described in the following section.
+Statistic measures are needed which describe how variables correlate given the selected groups.
+%The need of statistic measures which describe how variables correlate with each other given the selected groups is special for the application domain.
+%
+These associations are also summarized using pivot tables which are popular in epidemiology.
 % - Selection of Elements in Scatterplots, Barcharts or Parallel Coordinates Views - observe how selection changes another view - this allows for multivariate analysis\\
@@
 
-\section{Interaction- and Visualization Techniques} \label{Interaction- and Visualization Techniques}
-% TODO: Bernhard: Section anders beginnen!
-We employ the fourth and highest level which includes next to brushing and linking, advanced brushing, extraction of attributes the development of visualizations custom tailored to the data sets.
+\subsection{Data Preprocessing} \label{Data Preprocessing}
+Transformation operations on the data to prepare it for a \emph{IVA} system are denoted as data preprocessing.
+\paragraph{Non-Image data. }
+%To include heterogenous epidemiological data in an \emph{IVA}-framework it is necessary to process it to obtain standardized views to the available features.
+%
+Multimodal features require different techniques.
+%Due to the different acquisition modalities there have to be different techniques incorporated.
+%
+Data obtained using questionnaires or medical tests are often stored using statistical packages such as \texttt{SPSS} or \texttt{Stata}, which have a proprietary data format with limited export capabilities.
+%
+Exporting the data in the respective tool to a CSV file and then convert it to file types that are easily manageable such as JSON or XML makes it readable for all modern programming languages.
+%The best solution for us was to simply export the data in the respective tool to a CSV file and then convert it to data types which are easier manageable such as JSON or XML using our own classes.
+%
+This can be achieved using data wrangling tools such as \texttt{OpenRefine}\footnote{Developed by Google, Open Source; \url{http://openrefine.org/}}, which also validates the data (find missing data, clean up bad formatting, transform scales).
+%In order to verify that the conversion worked as expected and the data is valid, it is good practice to use data wrangling tools such as \texttt{OpenRefine} to validate the data (find missing data, clean up bad formatting).
+%
+Including the data dictionary, which stores information about each manifestation of a feature, helps to get a detailed description of data variables and the meaning and unit of measurement of their values.
+%Exporting the data dictionary, which stores information about each manifestation of a feature, is also an important step to get a detailed description of data variables and the meaning and unit of measurement of their values.
+%
+Missing data is denoted using Error codes indicating its cause ranging from ethical to medical and personal issues.
+%The reasons for missing data have a wide range from ethical to medical and personal issues.
+%
+%Therefore, these are also included as error codes which have to marked as such in the data dictionary.
+\paragraph{Image data. }
+Processing the image data associated to each subject consists for the most part of information extraction about a anatomical structure.
+%
+This is either done manually by experts setting landmarks (sometimes supported by algorithms connecting the land marks such as graph cuts \cite{GraphCut}) or by a (semi)-automatic detection, registration and segmentation.
+%
+These algorithms have to deal with a large inter-subject variability of the anatomical structure and need to produce reproducible results \cite{Preim2014}.
+%
+Model-based approaches have shown in principle to be effective for segmentation \cite{Gloger2010, Gloger2012} and detection \cite{Rak2013}.
+%
+If a segmentation yields only binary masks separating the structures, algorithms such as Growing and Adaptive Shapes can be applied creating a surface grid where each point is comparable throughout the population \cite{Ferrarini2007}.
+%
+Intensity-based comparisons can be achieved using rigid image registration, but model based results however are preferable \cite{Klemm2012}.
+%
+Grey value comparison is usually used to measure the quantity of fat, water, and--application specific--iron content (liver) or the distribution of grey and white brain tissue.
+
+Morphometric variables are derived to allow for statistical comparison of the tissue which incorporate mostly positions, diameters, volumes and relative distances and alignment to other structures.
+% 
+% - Data Format\\
+% - Missing Data\\
+% - Registration of Image Data (VMV'12)\\
+% - Extraction of Image Parameter
+\subsection{Analysis workflow}
+\begin{figure}[htb]
+ \centering
+ \includegraphics[width=3.0in]{figures/InteractionLoop}
+ % TODO: Recreate Image: Also include Cramér´s V which is part of the visualization
+ \caption{Detailed Version of the \emph{IVA Loop}.
+Usually starting with a selection of Feature of interest (user-driven or via data mining techniques), the data is mapped using a visualization technique appropriate for the selected data types.
+The data is visualized in the range and domain space, which can then be brushed, yielding new groups which can be investigated using further features.
+Note that adjacent steps are directly connected via feedback loops allowing for iterative refinement and give as much freedom to the user as possible.}
+ \label{fig:InteractionLoop}
+\end{figure}
+Our proposed \emph{IVA} workflow knows three major steps as illustrated in Figure~\ref{fig:InteractionLoop}: Feature selection, -visualization and -brushing.
+%
+An hypothesis-driven analysis usually starts with the selection of features.
+%
+A feature selection can also be derived from a shape-based clustering which creates shape groups.
+%
+Hypothesis-generation with focus on image-data starts with a shape based clustering.
+%
+The feature is mapped using a automatically chosen visualization appropriate for its data type.
+%
+The Visualization techniques have to combine both image-and non image data in order to set domain and range data in relation to each other.
+%
+Subsequently the visualization can be either brushed or new variables can be added to the analysis.
+%
+Brushes act as feature as they divide the subject space just like categorical variables and lead therefore to a new variable selection.
+%
+Selecting features also triggers a contingency analysis highlighting variables interacting with selected features, which is described in the following section.
+
+%The starting point of \emph{IVA Loop} is determined whether the analysis is hypothesis-driven or exploratory.
+%
+%An analysis with the aim of hypothesis generation of image-based data starts with a clustering of the structure 
+%
+A sample workflow using interaction- and visualization techniques described in the following section can be seed in Figure~\ref{fig:visualization}.
+
+\subsection{Interaction- and Visualization Techniques} \label{Interaction- and Visualization Techniques}
+%-------------------------------- Figure --------------------------------
+% TODO: VISUALIZATION Figure sollte umgebaut werden, sodass sie die Schritte bei der Analyse hervorhebt
 \begin{figure}[htb]
@@
 \end{figure}
-\emph{IVA} levels define different levels interaction. 
-%
-The employed techniques needs to respect the epidemiological requirements.
-%
-Hypothesis generation bears the chance of over-fitting the data to expectations.
-%
-To avoid this, a timeline needs to be introduced which keeps track how many variable variations the user evaluated before coming to an conclusion.
-%
-Since epidemiologists are used to process groups based on table representations we decided to introduce an interactive solution in form of a Pivot Table.
-
-\subsection{Structure and Workflow}
+%-------------------------------- /Figure -------------------------------
+%We employ the fourth and highest level which includes next to brushing and linking, advanced brushing, extraction of attributes the development of visualizations custom tailored to the data sets.
+%
+%\emph{IVA} levels define different levels interaction. 
+%
+Suitability of an interaction- and visualization technique for epidemiological data depends on its ability to intuitively compare multiple data features at once while highlighting new interesting interactions.
+%
+%Since epidemiologists come from different expertise, their different background has to be taken into account.
+%
+The chances of a method to be adapted by epidemiologists increases highly if their domain language is used and the methods reflect routines they are used to.
+%
+Visual evaluation of data is therefore as important as methods allowing for numerical data analysis.
+%
+In the following sections we present the different parts of our proposed \emph{IVA} system for image-based cohort study analysis.
+%Chosing or creating well suited visualization- and interaction techniques for epidemiological data
+%The employed techniques need to respect the epidemiological requirements.
+
+% TODO Einbauen
+% Hypothesis generation bears the chance of \emph{overfitting} the data to expectations.
+% %
+% To avoid this, a timeline needs to be introduced which keeps track how many variable variations the user evaluated before coming to an conclusion.
+% %
+% Since epidemiologists are used to process groups based on table representations we decided to introduce an interactive solution in form of a Pivot Table.
+
+\subsubsection{Structure of the system} \label{Structure and Workflow}
 We divide the workspace into three major parts as seen in Figure~\ref{fig:visualization}.
-%
-\begin{figure}[htb]
- \centering
- \label{fig:InteractionLoop}
- \includegraphics[width=3.0in]{figures/InteractionLoop}
- \caption{Detailed Version of the \emph{IVA Loop}.
-Usually starting with a selection of Variable of interest (user-driven or via data mining techniques), the data is mapped using a visualization technique appropriate for the selected data types.
-The data is visualized in the independent and dependent domain space, which can then be brushed, yielding new groups which can be investigated using further variables.
-Note that adjacent steps are directly connected via feedback loops allowing for iterative refinement and give as much freedom to the user as possible.}
-\end{figure}
-TODO auf Fig~\ref{fig:InteractionLoop} zurückkommen
+\begin{itemize}
+	\item The sidebar, which contains all epidemiological features. Cluster results are treated like features and are part of the sidebar as well.
+	\item The canvas which holds all visualizations. Elements can be added, arranged, resized and removed freely.
+	\item The interactive Pivot Table gives detailed numerical information of the features in the canvas view. This view on the data come natural to epidemiologists.
+	\item The contingency view which depicts relations for features in the canvas.
+\end{itemize}
 %
@@
 
-From the sidebar, elements of interest can be put into the canvas area via drag and drop.
+Features can be dragged from the sidebar into the canvas area.
 %
@@
 
-\subsection{Sidebar}
+\paragraph{Sidebar}
 An overview of all variables is presented in a sidebar where they are categorized to different types such as somatometric, disease- or lifestyle related, pain indicators and laboratory data.
@@
 
-\subsection{Adaptive Feature Visualization} \label{sec:AdaptiveFeatureVisualization}
+\paragraph{Adaptive Feature Visualization} \label{sec:AdaptiveFeatureVisualization}
 Inspired be the previously discussed \emph{GPloms} \cite{Francois2013} the visualization type is chosen dynamically based on the variable types and number of visualizations which need to be displayed.
@@
 
-\subsection{Pivot Tables}
+\paragraph{Pivot Tables}
 Pivot tables are frequently used to present the data in epidemiological paper.
@@
 
-\subsection{Automated Feature Suggestion}
+\subsubsection{Automated Feature Suggestion}
 % https://de.wikipedia.org/wiki/Kontingenzkoeffizient#Cram.C3.A9rs_V
@@
 
-\subsection{Data Preprocessing} \label{Data Preprocessing}
-Transformation operations on the data to prepare it for the presented prototype are denoted as data preprocessing.
-%We denote all data transformation steps necessary outside of our prototype for it to be prepared to work in our prototype as data preprocessing.
-%
-\paragraph{Image-Data.} \label{Image-Data}
-The lumbar spine was detected in the image data using a hierarchical finite element method according to Rak and colleagues \cite{Rak2013}.
-%
-This semi-automatic method requires the user to initialize the Tetrahedron-based finite element models (FEM) with a click on the L3-vertebra.
-%
-Two user defined landmarks on the top and bottom of the L3-vertebra are used to obtain an initial height estimation of the model.
-%Two more user- defined landmarks are the top and bottom of the L3-vertebra, which transforms size of the model accordingly to match the L3-vertebra .
-%
-It uses a weighted sum of T1- and T2-weighted MRI images to detect the lumbar spine shape.
-%
-The registered models capture resilient information about shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae \cite{Klemm2013VMV}.
-%
-Due to TODO ANTWORT VON MARKO HIER, 983 models are obtained.
-%
-For clustering purposes we extracted the centerline of the spine canal of the lumbar spine canal which captures information about lordosis and scoliosis which are medical terms for spine curvature \cite{Klemm2013VMV}.
-
+\subsection{Data Preprocessing} \label{application:Data Preprocessing}
+Following methods described in Section~\ref{Data Preprocessing}, the data is preprocessed as follows for the presented prototype.
+%Transformation operations on the data to prepare it for the presented prototype are denoted as data preprocessing.
+%
 \paragraph{Non-Image Data.}
@@
 To allow for hypothesis generation we set the number of groups to five if not specified otherwise.
-% TODO MIT KAI FORMEL ERARBEITEN?
+
+\paragraph{Image-Data.} \label{Image-Data}
+The lumbar spine was detected in the image data using a hierarchical finite element method according to Rak and colleagues \cite{Rak2013}.
+%
+This semi-automatic method requires the user to initialize the Tetrahedron-based finite element models (FEM) with a click on the L3-vertebra.
+%
+Two user defined landmarks on the top and bottom of the L3-vertebra are used to obtain an initial height estimation of the model.
+%Two more user- defined landmarks are the top and bottom of the L3-vertebra, which transforms size of the model accordingly to match the L3-vertebra .
+%
+It uses a weighted sum of T1- and T2-weighted MRI images to detect the lumbar spine shape.
+%
+The registered models capture resilient information about shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae \cite{Klemm2013VMV}.
+%
+Due to TODO ANTWORT VON MARKO HIER, 983 models are obtained.
+%
+For clustering purposes we extracted the centerline of the spine canal of the lumbar spine canal which captures information about lordosis and scoliosis which are medical terms for spine curvature \cite{Klemm2013VMV}.
 
@@
 %
-The tetrahedron-based model detection model described in Section~\ref{Data Preprocessing} consists of corresponding grid points for each structure instance.
+The tetrahedron-based model detection model described in Section~\ref{application:Data Preprocessing} consists of corresponding grid points for each structure instance.
 %
@@
 \\\\
-Shape based clustering is carried out via Agglomerative Hierarchical clustering of the spine canal ,centerlines which are described in Section~\ref{Data Preprocessing} \cite{Klemm2013VMV}.
+Shape based clustering is carried out via Agglomerative Hierarchical clustering of the spine canal, centerlines which are described in Section~\ref{application:Data Preprocessing} \cite{Klemm2013VMV}.
 %

### 2014-03-26 draft.tex
All sections are written, Images still not done

@@
 %% \section{Introduction} %for journal use above \firstsection{..} instead
-\textbf{(Background)} Epidemiology aims at characterizing health and disease by determining risk factors.
+Epidemiology aims at characterizing health and disease by determining risk factors.
 %
@@
 \\\\
-\textbf{(Problem)}
 Longitudinal population-based studies, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}, aim to gather as much information as possible about a defined sample of people (a \emph{cohort}).
@@
 \\\\
-\textbf{(Goals)}
 We propose an Interactive Visual Analysis (\emph{IVA}) approach \cite{Thomas2005} to provide a way to analyze image- and non-image data.
@@
 \\\\
-\textbf{(Contributions)}
 Our contributions are:
@@
 %
-The Visualization techniques have to combine both image-and non image data in order to set domain and range data in relation to each other.
+The visualization techniques have to combine both image-and non image data in order to set domain and range data in relation to each other.
 %
@@
 % TODO: VISUALIZATION Figure sollte umgebaut werden, sodass sie die Schritte bei der Analyse hervorhebt
-\begin{figure}[htb]
+\begin{figure*}[htb]
  \centering
@@
  \includegraphics[width=3.5in]{figures/visualization}
+ % \includegraphics[width=1\textwidth]{figures/visualization}
  \caption{\emph{Image is not final} Screenshot from the front-end which is divided as follows: (a) Sidebar which contains all features as well as the groups defined in the analysis process; (b) Canvas area where features can be added via drag and drop and the visualization chosen automatically according to the data type; (c) Interactive Pivot Table which shows exact numbers for each displayed variable combination.
@@
  \label{fig:visualization}
-\end{figure}
+\end{figure*}
 %-------------------------------- /Figure -------------------------------
@@
 
-\subsubsection{Structure of the system} \label{Structure and Workflow}
-We divide the workspace into three major parts as seen in Figure~\ref{fig:visualization}.
+\subsubsection{System Structure} \label{Structure and Workflow}
+We divide the workspace into four major parts as seen in Figure~\ref{fig:visualization}.
 \begin{itemize}
@@
 	\item The interactive Pivot Table gives detailed numerical information of the features in the canvas view. This view on the data come natural to epidemiologists.
-	\item The contingency view which depicts relations for features in the canvas.
+	\item The contingency view depicting relations for features in the canvas.
 \end{itemize}
 %
-The sidebar holds all available variables as well as groups either derived by user input or automatic clustering.
-
+% TODO Include:
+% The sidebar holds all available variables as well as groups either derived by user input or automatic clustering.
+% 
+% Features can be dragged from the sidebar into the canvas area.
+% %
+% Doing so will automatically create an information visualization suitable for the current data type.
+% %
+% 3D shape information about the investigated structure is displayed for each variable manifestation.
+% %
+% By dropping variables on existing visualizations the system creates a visualization that allows for comparison (e.g. mosaic plots for ordinal variables).
+% %
+% Elements can be brushed in the information visualizations and are linked to the other representations in the canvas.
+% %
+% Shape based clustering can either be applied to all subjects or subgroups.
+% 
+% All elements in the canvas view are also represented in the interactive Pivot Table which gives detailed information how the subjects are distributed given the displayed variables.
+% %
+% Details to the different views are presented in the following sub-sections.
+
+\paragraph{Sidebar.}
+An overview of all features is presented in a sidebar where they are categorized to different types such as somatometric, disease- or lifestyle related, pain indicators and laboratory data.
+%
+It also contains subject groups either defined by brushing or automated shape-clustering.
+%
+Groups are treated like features since they work exactly the same way by dividing the subject space into labeled categories.
+%
 Features can be dragged from the sidebar into the canvas area.
 %
-Doing so will automatically create an information visualization suitable for the current data type.
-%
-3D shape information about the investigated structure is displayed for each variable manifestation.
-%
-By dropping variables on existing visualizations the system creates a visualization that allows for comparison (e.g. mosaic plots for ordinal variables).
-%
-Elements can be brushed in the information visualizations and are linked to the other representations in the canvas.
-%
-Shape based clustering can either be applied to all subjects or subgroups.
-
-All elements in the canvas view are also represented in the interactive Pivot Table which gives detailed information how the subjects are distributed given the displayed variables.
-%
-Details to the different views are presented in the following sub-sections.
-
-ToDo\\
-- wie werden Confounder gefunden?\\
-- welche statistischen Kennzahlen werden eingebunden?\\
-
-\paragraph{Sidebar}
-An overview of all variables is presented in a sidebar where they are categorized to different types such as somatometric, disease- or lifestyle related, pain indicators and laboratory data.
-%
-It also contains subject groups either defined by user brushing or by automated clustering.
-%
-Groups are treated exactly like other variables since they work exactly the same way which is dividing the subject space into different labeled categories.
-% TODO Implement
-Bar charts show the distribution of manifestations of each variable in the sidebar.
-
-\paragraph{Adaptive Feature Visualization} \label{sec:AdaptiveFeatureVisualization}
-Inspired be the previously discussed \emph{GPloms} \cite{Francois2013} the visualization type is chosen dynamically based on the variable types and number of visualizations which need to be displayed.
+This triggers an adaptive feature visualization suitable for the current data type.
+%Bar charts show the distribution of manifestations of each variable in the sidebar.
+
+\paragraph{Adaptive Feature Visualization.} \label{sec:AdaptiveFeatureVisualization}
+Inspired be previously the discussed \texttt{GPloms} \cite{Francois2013} the visualization type is chosen dynamically based on the variable types and number of visualizations which need to be displayed.
+%
 % TODO Implement MOSAIC PLOTS, Scatterplots, Parallel Coordinates
-If possible the medical image data is directly included into the plot as well by including mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
-%
-The 3D-plots can be navigated using standard mouse inputs and the camera is synchronized so that a direct comparison is given.
-%
+Following Tuftes concept of \emph{small multiples} \cite{Tufte1983} the medical image data is directly included into the plot by including color coded mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
+%If possible the medical image data is directly included into the plot as well by including mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
+%
+The 3D-plots can be navigated using standard mouse inputs, the camera is synchronized between all views to enable direct comparison.
+%
+Mapping the distance from a group mean shape to the global mean to color allows to assess local shape changes.
+%%TODO Adaptive visualization!
 If a feature is dropped on an existing plot, the visualization changes dynamically to properly make them comparable. 
@@
 %
-It is able to duplicate brushes in order to create new groups which are evenly spread out.
-%
-A use case for this is when a continuous feature has to be divided into even groups.
-\\
-ToDo\\
-- Dichotomous data\\
-- Time-Line\\
-- Statistical Analysis (Odds ratios)\\
-
-TODO 2\\
-- 3D Plots are small multiples - reference Tufte!
-
-\paragraph{Pivot Tables}
-Pivot tables are frequently used to present the data in epidemiological paper.
+Brush selections are propagated to all visualizations allowing for fast feature querying.
+% TODO Timeline
+%
+%It is possible to duplicate brushes in order to create new groups which are evenly spread out.
+%
+%A use case for this is when a continuous feature has to be divided into even groups.
+
+\paragraph{Pivot Tables.}
+Pivot tables are frequently used to present the data in epidemiological publications.
+%
+Epidemiologists are used to process groups based on table representations so decided to introduce an interactive pivot table.
 %When opening a random epidemiological paper the reader will in almost all cases find some sort of pivot table to present the data.
@@
 This also applies on the designation of Row or Column Variables.
+%
+Another way to avoid clutter is the user-driven selection of displayed variables.
+%
+To allow better comparison with respect to features the values of each cell can also be displayed as percentage of the feature represented of either the row or the column.
 % TODO IMPLEMENT
-The mean shape for each resulting sub-group is also displayed for each subject.
-
-\subsubsection{Automated Feature Suggestion}
-% https://de.wikipedia.org/wiki/Kontingenzkoeffizient#Cram.C3.A9rs_V
-TODO REWRITE USING CRAMERS V
-
+%The mean shape for each resulting sub-group is also displayed for each subject.
+
+\paragraph{Automated Feature Suggestion using a Contingency Matrix.}
 As discussed previously, highlighting potential interesting values in the data set is one major benefit of the \emph{IVA} powered approach.
@@
 %
-A solution for this problem are odds ratios, which are a standard statistical tool for stating relations between features.
-%
-Odds ratios can only be calculated for $2\times2$ contingency tables which usually represent the presence of a condition in a population divided by a characteristic (e.g.
+The \emph{Cram\'{e}rs' V} contingency coefficient can be used to calculate coherences between categorical variables \cite{CramerV}.
+%
+It is based on \emph{Pearson's $X^2$} distribution test \cite{ChiSquare}, which uses contingency tables holding the counts of subjects for all possible manifestations of two variables.
+%
+\emph{Cram\'{e}rs' V} is defined as:
+\begin{equation}
+V = \sqrt{\frac{X^2}{N(k-1)}},
+\end{equation}
+where $X^2$ equals \emph{Pearson's chi squared}, $N$ is the total of observations and $k$ either the row or column count, depending on which one is lower.
+%
+$V$ assumes values between $0$, meaning two variables are completely independent and $1$ indicating they are the same.
+%
+\emph{Cram\'{e}rs' V} is always positive and does therefore not allow statements about dependency direction.
+
+It shares the same restrictions as \emph{Pearson's $X^2$} which include next to others expected count larger than 5 for $80\%$ of the contingency table entry and no expected value can be smaller than one.
+%
+Therefore are manifestations who are only exposed by small subject groups not assessable using this technique.
+%
+This makes them not assessable for epidemiological analysis as well since statistical validation needs a minimum count to be seen as valid.
+% Limitations and usual space
+The contingency matrix highlights contingencies of visualized features with features in the data set.
+%
+This aims to highlight features which may interact with the focused hypothesis as well as trigger new hypotheses.
+%
+Contingency is visualized using an interactive adjacency matrix with association power mapped to color hue.
+%
+The distinction whether a interaction is a confounder or and effect depends on the context defined by the hypothesis and is a decision to be made by the domain expert.
+\\\\
+In the following Section we discuss how we implemented the presented methods using free on open Web standards.
+% \subsubsection{Automated Feature Suggestion}
+% % https://de.wikipedia.org/wiki/Kontingenzkoeffizient#Cram.C3.A9rs_V
+% 
+% As discussed previously, highlighting potential interesting values in the data set is one major benefit of the \emph{IVA} powered approach.
+% %
+% Turkay and colleagues used the approach to calculate various key figures based in the distribution functions of each feature derived from the image data \cite{Turkay2013}.
+% %
+% Since the majority of our data are categorical features, we have to employ different solutions.
+% %
+% A solution for this problem are odds ratios, which are a standard statistical tool for stating relations between features.
+% %
+% Odds ratios can only be calculated for $2\times2$ contingency tables which usually represent the presence of a condition in a population divided by a characteristic (e.g.
 male/female subjects with presence or absence of back pain).
-%
-To calculate odds ratios for variables with more than two manifestations, we calculate local odds ratios for each possible combination, yielding a matrix for each feature combination \cite{Rudas1998}.
-%
-When the subjects are divided into groups and the calculation is carried out again for all feature combination, the difference in sum of the odds weighted with the number of manifestations per feature can be used to indicate if the feature combination yields a difference.
-%
-These difference are then highlighted in a separate tab of the side bar "Interactions".
-%
-Interesting interactions then can be assessed creating a linked view using the standard drag and drop workflow.
-\\\\
-In the following sections we will discuss details on the implementation which relies on modern Web-Technologies.
-\\\\
-ToDo\\
-- This can be improved--summing up the values possibly not the cleverest solution--calculation of variance etc. possible\\
-- Matrix Visualization?
+% %
+% To calculate odds ratios for variables with more than two manifestations, we calculate local odds ratios for each possible combination, yielding a matrix for each feature combination \cite{Rudas1998}.
+% %
+% When the subjects are divided into groups and the calculation is carried out again for all feature combination, the difference in sum of the odds weighted with the number of manifestations per feature can be used to indicate if the feature combination yields a difference.
+% %
+% These difference are then highlighted in a separate tab of the side bar "Interactions".
+% %
+% Interesting interactions then can be assessed creating a linked view using the standard drag and drop workflow.
+% \\\\
+% In the following sections we will discuss details on the implementation which relies on modern Web-Technologies.
+% \\\\
+% ToDo\\
+% - This can be improved--summing up the values possibly not the cleverest solution--calculation of variance etc. possible\\
+% - Matrix Visualization?
 
@@
 %
-In order to provide a fast communication loop between method development and expert input, we decided to base all implementations on modern web technologies which benefits from various advantages:
+To provide a fast communication loop between method development and expert input, we decided to rely on modern web technologies which benefits from various advantages:
 \begin{itemize}
 	\item No additional software needs to be installed, most people use decent state-of-the-art web browser, even on mobile devices.
-	\item The Client-Server structure allows it to employ heavy computation on a server machine and transfer results to the client
-	\item Since image data for several thousand subjects claims hundreds of Gigabytes disk space it can remain safely on the server and elements can be transferred on demand.
+	\item The client-server structure allows for employing heavy computation on a server machine and transfer results to the client.
+	\item Since image data for several thousand subjects claims hundreds of gigabytes disk space it can remain safely on the server and elements can be transferred on demand.
 	%
@@
 	\item Recent developments in WebGL applications running in browsers with near native performance push the development into to the web which results in many open source libraries which are well documented, rich in examples and driven by active communities.
+	\item Web technologies are free.
 \end{itemize}
@@
 %
-Many methods which specialized libraries like the Visualization Toolkit (VTK) or R for statistics have build in need to be written from scratch in order to fit in the context.
-%
-
-The back-end is written using Node, which based on the Google V8 Javascript runtime environment.
+Many methods which specialized libraries, like the Visualization Toolkit (VTK) or R for statistics have build in, need to be written from scratch in order to fit in the context.
+%
+
+The back-end is written using \texttt{NodeJS}\footnote{Developed by Joyent Inc, \url{nodejs.org}}, which based on the Google V8 Javascript runtime environment.
 %
@@
 \\\\
-Non-image data for all subjects including the data dictionary is stored on the Server in a JSON file.
-%
-Image data is available as raw DICOM files as well as segmented Meshes which can be used to compare subjects.
+Non-image data for all subjects including the data dictionary is stored in a JSON file on the Server.
+%
+Image data is available as raw DICOM files as well as segmented meshes, which can be used to compare subjects.
 %
@@
 %
-The server processes calculation heavy statistical tasks such as calculation of Odds Ratios or Chi Square tests for all variable combinations in order to keep the computation time on the client as low as possible.
-%
-
-The front-end is created using Twitter Bootstrap as foundation for the layout and basic UI elements using HTML5, CSS3 and Javascript.
-%
-Information Visualizations such as Scatterplots and bar charts are created using the popular Data Driven Documents library which works well for attaching data to visible elements like vector graphics.
-%
-WebGL rendering is done using the Threejs which allows GPU Accelerated data rendering.
+The server processes calculation heavy statistical tasks such as calculation of \emph{Cram\'{e}rs' V} values for all variable combinations in order to keep the computation time on the client as low as possible.
+%
+
+The front-end is created using \texttt{Twitter Bootstrap}\footnote{Developed by Twitter, \url{getbootstrap.com}} as foundation for the layout and basic UI elements using \texttt{HTML5}, \texttt{CSS3} and \texttt{Javascript}.
+%
+Information Visualizations such as scatterplots and bar charts are created using the popular \texttt{Data-Driven Documents (D3.js)} \cite{D3} library, which works well for attaching data to visible elements like vector graphics and provides powerful transformation and mapping tools.
+%
+\texttt{WebGL} rendering is done using the \texttt{Three.js}\footnote{Originally developed by Ricardo Cabello, \url{threejs.org}} and allows GPU Accelerated data rendering.
 %
@@
 %
-Since our clustering algorithms are written in MatLab we hat to access them using the Node Server.
-%
-We accomplished this by converting them to parameterized standalone console applications which are spawned by node on client request and then reads the result from the console standard-out and returns it in a proper format to the client.
+Since our clustering algorithms are written in \texttt{MatLab}\footnote{Owned by The MathWorks, \url{www.mathworks.com}} we hat to access them using the \texttt{NodeJS} Server.
+%
+We accomplished this by converting it to a parameterized standalone console application, that is spawned by \texttt{NodeJS} on client request and then reads the result from the console standard-out and returns it in a proper format to the client.
 %
 All parameter steered console applications can be incorporated in this context.
+
+The following section describes how our presented \emph{IVA} workflow works for a epidemiological use case.
 
@@
 %
-Due to TODO ANTWORT VON MARKO HIER, 983 models are obtained.
+Due to incorrect initialization, strongly deformed spines, contrast-differences and artifacts, the model was not able to detect lumbar spines for all subjects.
+%
+We obtained and work with 983 models.
 %
@@
 %
-TODO Feedback einholen von Frau Hegenscheid, TODO Implement Color coding only in a certain direction!\\
-TODO Selection of the variables and display them using the pivot table!
+TODO Feedback Greifswald, 
+%TODO Implement Color coding only in a certain direction!\\
+%TODO Selection of the variables and display them using the pivot table!
 \\\\
@@
 \section{Summary and Conclusion}
-- Future Work: Matrix-View of differences - more intuitive\\
-- UI more flexible by hiding UI-panes
-- Dynamic Discretization to better fit the data distributions
-- More Shape based filtering - reorder by size, curvature, etc. ...
-- Color code different Aspects - Only difference in x or y direction
-- Calculation of contingency based on the contingency differences - how does the current selection differ compared to the whole data set?
-
-Discretization of continuos variables into bins of equal size may bias the underlying data by not considering the variable distribution function \cite{Fletcher2012}.
-% Fletcher 2012 - S. 257
-To reduce the number of false positive findings, the data space can be randomly be cut in half the hypothesis then can be cross-validated for statistical soundness.
+%Seen as an extension to the existing statistics driven workflow, image-centric analysis of epidemiological data from a IVA approach by illuminating the data-black box.
+The presented are classified as extension to the existing statistics driven epidemiological.
+%
+Image-centric epidemiological analysis benefits from a IVA approach by illuminating the data-black box.
+%
+Visualization of multivariate data using connected views and different views allows to get fast visual feedback about subject groups.
+%
+Brushing and linking makes the data tangible and adaptable to formulated hypotheses.
+%
+The use of pivot tables is familiar to epidemiologist while embracing the power of interactive adjustment of the features shown.
+%
+Automatic suggestion of correlations using contingency methods like \emph{Cram\'{e}rs' V} trigger \emph{hypothesis generation} by highlighting features potentially overlooked by the experts.
+%
+Shape based clustering assesses variability of a anatomical structure in the context of non-image features such as disease indicators or lifestyle factors.
+%
+The iterative nature of the \emph{IVA} approach reflects versatility as both \emph{hypothesis driven analysis} and \emph{hypothesis generation} is allowed.
+\\\\\
+Our future work comprises a number of improvements:
+\begin{itemize}
+	\item dynamic discretization of continuous variables to better fit the data distributions
+	\item shape brushing methods to intuitively query subjects using image-data
+	\item incorporate more statistical methods and views that are familiar to the epidemiologists (odds ratios, box plots)
+	\item calculation of contingency differences for selected sub groups--highlight how feature correlation change for a given groups
+	\item more visualization techniques which incorporate both image and non-image data
+	\item support control-groups to check resilience of observations as suggested by Fletcher and Fletcher \cite{Fletcher2012}.
+\end{itemize}
+% These benefits 
+% - Future Work: Matrix-View of differences - more intuitive\\
+% - UI more flexible by hiding UI-panes
+% - More Shape based filtering - reorder by size, curvature, etc. ...
+% - Color code different Aspects - Only difference in x or y direction
+% - Adjacency Matrix is symmetrical, space can be saved
+% - Calculation of contingency based on the contingency differences - how does the current selection differ compared to the whole data set?
+To reduce the number of false positive findings, the data space can also be randomly be cut in half the hypothesis then can be cross-validated for statistical soundness.
 %
@@
 %
-As suggested by Fletcher and Fletcher, we plan to validate the hypothesis using the \texttt{SHIP-TREND} cohort \cite{Fletcher2012}.
+%As suggested by Fletcher and Fletcher, we plan to validate the hypothesis using the \texttt{SHIP-TREND} cohort \cite{Fletcher2012}.
+\\\\
+As the number of image-centric cohort studies, participating subjects, gathered features and imaging modalities rises and advances towards comparability between cohort studies are made, the gap between data complexity and assessability increases.
+%
+Our work focuses on closing this gap, allowing the domain experts to dig deep into the data and potentially obtain unexpected findings.
+%
+We believe web technologies pave the way to allow as many experts as possible to analyze this data and provide a fast exchange between users and developers employing many different devices.
+%
+Visual analysis shows to be a promising way to clear the view on complex epidemiological data to uncover its secrets.
+
+%With increasing number of image-centric cohort studies, >evalueted< subjects, features collected and major advantages in methods that allow for comparabality of findings in different studies the gap between
 

### 2014-03-27 draft.tex
Incorporated Henrys feedback, two new figures

@@
 %
-The data is analyzed using a priori defined hypotheses to find statistically significant correlations between variables (\emph{interactions}).
+The data is analyzed using a priori defined hypotheses to find statistically significant correlations between variables (\emph{associations}).
 %
@@
 %
-The statistically driven epidemiological workflow only allows to determine \emph{interactions} between image-derived metrics such as distances extracted from landmarks of the segmentation model.
+The statistically driven epidemiological workflow only allows to determine \emph{associations} between image-derived metrics such as distances extracted from landmarks of the segmentation model.
 \\\\
@@
 %
-Using all reliable information from the image segmentation linked to non-image variables aims to unveil \emph{interactions} by applying an iterative analysis approach.
-
-%
-%Similarity measures between data variables are used to compute interesting changes in variable interactions for the current variable selection.
+Using all reliable information from the image segmentation linked to non-image variables aims to unveil \emph{associations} by applying an iterative analysis approach.
+
+%
+%Similarity measures between data variables are used to compute interesting changes in variable associations for the current variable selection.
 %
@@
 % %
-% %Using well established statistical methods the data is hypothesis driven analyzed to find statistically significant variable correlations (\emph{interactions}).
-% Driven by a priori defined hypotheses, the data is analyzed using well established statistical methods in order to find statistically significant correlations between variables (\emph{interactions}).
+% %Using well established statistical methods the data is hypothesis driven analyzed to find statistically significant variable correlations (\emph{associations}).
+% Driven by a priori defined hypotheses, the data is analyzed using well established statistical methods in order to find statistically significant correlations between variables (\emph{associations}).
 % %
@@
 % %
-% The statistically driven epidemiological workflow only allows to determine \emph{interactions} between image-derived metrics such as distances extracted from landmarks of the segmentation model.
+% The statistically driven epidemiological workflow only allows to determine \emph{associations} between image-derived metrics such as distances extracted from landmarks of the segmentation model.
 % \\\\
@@
 % %
-% Using all reliable information from the image segmentation linked to non-image variables aims to unveil \emph{interactions} by applying an iterative analysis approach.
-% %As extension to the epidemiological workflow it aims to unveil interactions based on all reliable shape information rather than abstracting it to a few key figures.
+% Using all reliable information from the image segmentation linked to non-image variables aims to unveil \emph{associations} by applying an iterative analysis approach.
+% %As extension to the epidemiological workflow it aims to unveil associations based on all reliable shape information rather than abstracting it to a few key figures.
 % %Iterative analysis between ...
-allows for extracting interactions which without the visiual analysis driven approach maybe would be overlooked.
+allows for extracting associations which without the visiual analysis driven approach maybe would be overlooked.
 %Einsicht in Interaktionen zwischen nicht-bilddaten und bilddaten auf basis aller verfügbarer Informationen
 % %Iterative Analyse von Organstrukturen im Bezug auf Bilddaten 
-% %Iterative analysis of image-based epidemiological datasets aims to provide insight into interactions between organ shape and % in order to provide insights into \emph{interactions} by incorporating additional information
+% %Iterative analysis of image-based epidemiological datasets aims to provide insight into associations between organ shape and % in order to provide insights into \emph{associations} by incorporating additional information
 % %It extends the epidemiological workflow with an iterative Analysis process and 
@@
 % %
-% Similarity measures between data variables are used to compute interesting changes in variable interactions for the current variable selection.
+% Similarity measures between data variables are used to compute interesting changes in variable associations for the current variable selection.
 % %
@@
 %
-Statistical methods like regression analysis aim to check the attribute list for plausibility.
+Methods like regression analysis check the attribute list for statistical soundness.
 %
 \\\\
-Longitudinal population-based studies, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}, aim to gather as much information as possible about a defined sample of people (a \emph{cohort}).
+Longitudinal population-based studies, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}, gather as much information as possible about a defined sample of people (a \emph{cohort}).
 %
@@
 %
-Modern cohort studies often include medical image data.
+Cohort studies often include medical image data.
 %
@@
 %	\item providing an overview over the workflow for analyzing cohort study data to gain insight into the large diverse subject spaces.
-	\item describing an Interactive Visual Analysis workflow for analyzing image-based epidemiological data including both hypothesis-driven and for hypothesis-generation based on a characterization of the standard epidemiological workflow
-	\item providing visualization techniques which combine both information visualization and 3D rendering of organ shapes as well as combining them with well known epidemiological graphics and key figures.
+	\item an Interactive Visual Analysis workflow for analyzing image-based epidemiological data including both hypothesis-driven and for hypothesis-generation based on a characterization of the standard epidemiological workflow
+	\item visualization techniques, which combine both information visualization and 3D rendering of organ shapes as well as combining them with epidemiological graphics and key figures.
 	\item highlighting interesting subject groups and feature associations using shape-based clustering and statistical contingency measures
@@
 % 	\item Provide visualization techniques which combine both information visualization and 3D rendering of organ shapes as well as combining them with well known epidemiological graphics and key figures.
-	\item implementing the presented methods as a web framework based on WebGL, D3JS and NodeJS.
+	\item implementing the presented methods as a web framework based on \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}.
 \end{itemize}
@@
 
-In this section we want to give insight into the epidemiological workflow when analyzing cohort study data to identify the problems we address in this paper.
+In this section we give insight into the epidemiological workflow when analyzing cohort study data to identify the problems we address in this paper.
 %
@@
 
-Grouping subjects using epidemiological features is essential in order to allow per-group risk determination.
+In longitudinal cohort design, grouping subjects using epidemiological features is essential in order to allow per-group risk determination.
 %
@@
 %
-These groups depend strongly on the condition of interest and therefore there is no defined standard on how to categorize these values.
+These groups depend strongly on the condition of interest and therefore there is no standard on how to categorize these.
 
@@
 %
-Statistical correlations are prone to \emph{confounding}, meaning that the interaction of two features are influenced by a third feature which needs to isolated.
+Statistical correlations are prone to \emph{confounding}, meaning that the association of two features are influenced by a third feature which needs to isolated.
 % Statistical correlations are prone to \emph{confounding}, meaning that two features are dependent and therefore should be normalized with respect to each other.
 %
-Statistics tools such as \texttt{SPSS}\footnote{Product of IBM; \url{www.ibm.com/software/de/analytics/spss/}} and \texttt{Stata}\footnote{Product of Stata Corp.; \url{http://www.stata.com/}} play a major role for analyzing epidemiological data.
+%Statistics tools such as \texttt{SPSS}\footnote{Product of IBM; \url{www.ibm.com/software/de/analytics/spss/}} and \texttt{Stata}\footnote{Product of Stata Corp.; \url{http://www.stata.com/}} play a major role for analyzing epidemiological data.
+Statistics tools such as \texttt{SPSS}\footnote{Product of IBM; \url{www.ibm.com/software/de/analytics/spss/}} play a major role for analyzing epidemiological data.
 %
@@
 
-Indicators for medical conditions as well as questions about a subjects' lifestyle are also often \emph{dichotomous}--they have two manifestations (often \emph{Yes} or \emph{No}).
-%
-This allows for the calculation of \emph{odds ratios} which describe the relation of two \emph{dichotomous} variables, allowing for direct comparison of their influences.
+Indicators for medical conditions as well as questions about a subjects' lifestyle are also often \emph{dichotomous}--they have two manifestations (\emph{Yes} or \emph{No}).
+%
+%This allows for the calculation of \emph{odds ratios} which describe the relation of two \emph{dichotomous} variables, allowing for direct comparison of their influences.
 %
@@
 %
-MRI data is more expensive to obtain than CT data but does not affect the subjects health and is therefore the main method for collecting cohort study imaging data.
+%MRI data is more expensive to obtain than CT data but does not affect the subjects health and is therefore the main method for collecting cohort study imaging data.
+MRI the main method for collecting cohort study imaging data.
 %
@@
 %
-The equipment used to gather medical image data is kept on the initial software and hardware version to ensure comparability in and between acquisition cycles.
+The equipment used to gather medical image data is kept, if possible, on the initial software and hardware version to ensure comparability in and between acquisition cycles.
 
@@
 \subsection{The Study of Health in Pomerania (SHIP)}
-After the pioneering Rotterdam study (started in 1990) several image-centric cohort study initiatives have evolved, they slightly differ in clinical focus, acquired data and epidemiologic research questions.
+After the pioneering Rotterdam study (started in 1990) several MR imaging studies initiatives have evolved, they slightly differ in clinical focus, acquired data and epidemiologic research questions.
 %
@@
 %
-A second cohort \texttt{SHIP-Trend} was established in 2008 to acquire data about a younger population.
+%A second cohort \texttt{SHIP-Trend} was established in 2008 to acquire data about a younger population.
+A second cohort \texttt{SHIP-Trend} was established in 2008.
 %
@@
 %
-Our approach sticks more to the information extracted from the segmented image data and derive variable interaction with non-image epidemiological factors.
+Our approach sticks more to the information extracted from the segmented image data and derive variable associations with non-image epidemiological factors.
 % Disadvantage - Nur 2 abgeleitete Variablen sichtbar, schlecht mit anderen Variablen in Verbindung zu bekommen (müsste jedoch auch gehen) - viele Dinge gehen durch die Abstraktion aber auch flöten!
@@
 %
-Their findings yielded a \emph{Concept Map} which linked cancer-related interactions via graph edges.
-%From their findings regarding the interaction of cancer-related socio-demographic factors are drawn in a \emph{Concept Map} where related factors are connected via graph-edges.
-Chui and colleagues visualized interactions in time-dependent epidemiological data using time-series plots highlighting risk factors differences in age and gender \cite{Chui2011}.
+Their findings yielded a \emph{Concept Map} which linked cancer-related associations via graph edges.
+%From their findings regarding theassociationn of cancer-related socio-demographic factors are drawn in a \emph{Concept Map} where related factors are connected via graph-edges.
+Chui and colleagues visualized associations in time-dependent epidemiological data using time-series plots highlighting risk factors differences in age and gender \cite{Chui2011}.
 
@@
 Considering more complex parts of the data would make those results more trustworthy and also could identify possible anatomical confounders--an epidemiolgical research result in itself.
-
-Statistical tests check for validity of the number but not for their completeness or plausibility!
+%Statistical tests check for validity of the number but not for their completeness or plausibility!
 \\\\
@@
 %
-\emph{IVA} knows three different exploration patterns, handling interaction of domain and range variables.
-%The explorative procedures when analyzing data using \emph{IVA} can be divided into three different patterns, handling interaction between domain and range variables.
+\emph{IVA} knows three different exploration patterns, handling associations of domain and range variables.
+%The explorative procedures when analyzing data using \emph{IVA} can be divided into three different patterns, handlinassociationon between domain and range variables.
 
@@
 %
-These algorithms can be used to investigate interactions between shape-groups and other non-image based variables.
+These algorithms can be used to investigate associations between shape-groups and other non-image based variables.
 %
@@
 %
-The epidemiologist is primarily interested in the shape of subjects within a range of a set of variables that describe the current hypothesis.
-%
-Epidemiologists are used to categorize data into groups that fit their hypothesis formulation.
+The epidemiologist is interested in the shape of subjects within a range of a set of variables that describe the current hypothesis.
+%
+Epidemiologists may categorize data into groups that fit their hypothesis formulation.
 %
@@
 %
-%To gain insight into interactions between SHIP-
+%To gain insight intassociationons between SHIP-
 
@@
  %\includegraphics[width=1.5in]{figures/InteractionLoop}
- \includegraphics[width=3.5in]{figures/visualization}
- % \includegraphics[width=1\textwidth]{figures/visualization}
- \caption{\emph{Image is not final} Screenshot from the front-end which is divided as follows: (a) Sidebar which contains all features as well as the groups defined in the analysis process; (b) Canvas area where features can be added via drag and drop and the visualization chosen automatically according to the data type; (c) Interactive Pivot Table which shows exact numbers for each displayed variable combination.
+ % \includegraphics[width=3.5in]{figures/visualization}
+ %\includegraphics[width=1\textwidth]{figures/visualization}
+ \includegraphics[width=1\textwidth, resolution=300]{figures/visualization}
+ \caption{(Left) Screenshot from the front-end which is divided as follows: (a) Sidebar which contains all features as well as the groups defined in the analysis process; (b) Canvas area where features can be added via drag and drop and the visualization chosen automatically according to the data type; (c) Interactive Pivot Table which shows exact numbers for each displayed variable combination.
 The data displayed is used to analyze the lumbar spine.
+Features can be added freely on the canvas via drag and drop.
+Dropping the \emph{age} parameter on the already plotted \emph{body size} container creates a mosaic plot combining both features (right).
  }
@@
 %
-Suitability of an interaction- and visualization technique for epidemiological data depends on its ability to intuitively compare multiple data features at once while highlighting new interesting interactions.
+Suitability of an interaction- and visualization technique for epidemiological data depends on its ability to intuitively compare multiple data features at once while highlighting new interesting associations.
 %
@@
 In the following sections we present the different parts of our proposed \emph{IVA} system for image-based cohort study analysis.
-%Chosing or creating well suited visualization- and interaction techniques for epidemiological data
+%Chosing or creating well suited visualization- anassociationon techniques for epidemiological data
 %The employed techniques need to respect the epidemiological requirements.
@@
 % Since epidemiologists are used to process groups based on table representations we decided to introduce an interactive solution in form of a Pivot Table.
+
+\begin{figure}[htb]
+ \centering
+ \includegraphics[width=3.5in]{figures/similarity_matrix}
+ \caption{Adjacency matrix of 131 features with a grand total of 17.161 combinations.
+Similarity is calculated using the \emph{Cram\'{e}rs' V} contingency value.
+Color brightness encodes association strength.
+The enlarged excerpt shows associations for shape clusters of subjects with and without diagnosed spine attrition, which show associations between gender, weight, body height and smoking behavior.
+ }
+ \label{fig:similarity}
+\end{figure}
 
@@
 %
-The distinction whether a interaction is a confounder or and effect depends on the context defined by the hypothesis and is a decision to be made by the domain expert.
+The distinction whether an association is a confounder or and effect depends on the context defined by the hypothesis and is a decision to be made by the domain expert.
 \\\\
@@
 % %
-% Interesting interactions then can be assessed creating a linked view using the standard drag and drop workflow.
+% Interestinassociationons then can be assessed creating a linked view using the standard drag and drop workflow.
 % \\\\
@@
 %
-Looking at the \emph{Cram\'{e}rs' V} contingency values of the group reveals interactions of this group with employment status, body size, age, thyroid nodules and blood-fat value.
+Looking at the \emph{Cram\'{e}rs' V} contingency values of the group reveal associations of this group with employment status, body size, age, thyroid nodules and blood-fat value.
 %
 TODO Feedback Greifswald, 
+While all observed features seem as plausible associations related to back pain, the values indicate subjects with chronic back pain which radiates into the legs.
+%
+Metabolic parameters such as blood-fat and blood-sugar are also possibly associated features.
+%
+Employment status may be a confounder describing the lifestyle of a subject.
 %TODO Implement Color coding only in a certain direction!\\
@@
 %
-To increase the probability of the observation not to be random, our clinical partners cross check for interactions in another cohort, the \texttt{SHIP-TREND}.
-%If features are identified as interesting for a disease, the epidemiologists look for interactions in another cohort, the \texttt{SHIP-TREND}.
+To increase the probability of the observation not to be random, they are cross check for associations in \texttt{SHIP-TREND} as a second, independent population sample, which was examined with methods identical to SHIP.
+%If features are identified as interesting for a disease, the epidemiologists look for associations in another cohort, the \texttt{SHIP-TREND}.
 %
@@
 \\\\
-Our clinical partners pointed out the usability of the presented methods to guide their attention to features which where not in the focus of their attention and expectance.
-%
-The explorative nature of the methods work well for gathering interactions which may act as confounder, as outcome of a disease or as an actual cause or risk factor.
+The presented methods guide attention to features which where not in the focus of attention and expectance of clinical researchers.
+%
+The explorative nature of the methods work well for gathering associations which may act as confounder, as outcome of a disease or as an actual cause or risk factor.
 %"\emph{Correlation does not imply causation}" is a famous saying in statistics, meaning that correlation does not necessarily imply causal relationships \cite{}.
@@
 
-On the feature level, our clinical partners were interested in the MRI scans for the subjects in the outlier clusters because they are highly likely to exhibit pathologies.
-%
-We plan to include a \texttt{DICOM}-viewer to meet this wish.
+%On the feature level, displaying MRI scans is interesting for the subjects in the outlier clusters because they are highly likely to exhibit pathologies.
+Displaying MRI scans for the subjects in the outlier cluster is promising because they are highly likely to exhibit pathologies.
+%
+%We plan to include a \texttt{DICOM}-viewer to meet this wish.
 

### 2014-03-28 draft.tex
Added two cites, included half of Steffens Feedback

@@
 %
-These data needs to be segmented to analyze its features for correlations with non-image data.
-%
-Manual segmentation is not feasible since it leads to a high variability which is at odds with the high reproducibility required in epidemiology research.
+%These data needs to be segmented to analyze its features for correlations with non-image data.
+These data need to be concurrently analyzed with the non-image data.
+This requires their segmentation and the derivation of features such as kidney volume, curvature of the spine or tissue texture of the liver.
+%
+%Manual segmentation is not feasible since it leads to a high variability which is at odds with the high reproducibility required in epidemiology research.
 %
@@
 
-\section{Epidemiological and Technical Background} \label{MedicalAndTechnicalBackground}
+\section{Epidemiological Background} \label{MedicalAndTechnicalBackground}
 % Wer ist an epidemiologischen Studien beteiligt?\\
@@
 
-In this section we give insight into the epidemiological workflow when analyzing cohort study data to identify the problems we address in this paper.
+%In this section we give insight into the epidemiological workflow when analyzing cohort study data and derive several challenges. %study data to identify the problems we address in this paper.
+In this section we describe the epidemiological workflow and associated requirements.
+% give insight into the epidemiological workflow when analyzing cohort study data and derive several challenges
 %
@@
 %
-Epidemiologists follow a workflow mainly driven by statistic tools to validate hypothesis about disease specific risk factors.
-%
-Following Thew and colleagues, the workflow can be characterized as follows \cite{Thew2009}.
-%
-\begin{itemize}
-	\item Hypotheses are based most commonly on observations made by clinicians in their daily routine.
+Epidemiologists follow a workflow mainly driven by statistic tools to validate hypotheses about disease specific risk factors.
+%
+Following Thew and colleagues, the workflow, shown in Figure~\ref{fig:WorkflowComparison} (a), can be characterized as follows \cite{Thew2009}:
+%
+\begin{enumerate}
+%	\item Hypotheses are based most commonly on observations made by clinicians in their daily routine.
+	\item A Hypothesis is derived from observations made by clinicians in their daily routine.
 %
@@
 % TODO: Bernhard: Effect size of Attribute ?
-	\item Confounding features are adjusted so that they do not affect the effect size of a attribute.
-%
-	\item Statistical methods such as regression analysis are applied to measure the effect size of attributes to the outcome of interest.
-\end{itemize}
-The workflow is shown in Figure~\ref{fig:WorkflowComparison} (a).
+	\item Confounding features are adjusted so that they do not influence the effect size of a attribute.
+%
+	\item Statistical methods such as regression analysis assess the association of selected features with the investigated condition.%are applied to measure the effect size of attributes to the outcome of interest.
+	%\item Statistical methods such as regression analysis are applied to measure the effect size of attributes to the outcome of interest.
+\end{enumerate}
+%In the following we discuss the problems and challenges regarding the workflow.
 
@@
 Statistical correlations are prone to \emph{confounding}, meaning that the association of two features are influenced by a third feature which needs to isolated.
+%
+A famous example for a confounder is the correlation observable between shoe size and mortality where it can be observed that people with larger shoes have a smaller life expectation than people with small shoe size.
+%
+Shoe size actually is associated with gender where women have a smaller feet than men but have a longer life expectation.
 % Statistical correlations are prone to \emph{confounding}, meaning that two features are dependent and therefore should be normalized with respect to each other.
@@
 %Statistics tools such as \texttt{SPSS}\footnote{Product of IBM; \url{www.ibm.com/software/de/analytics/spss/}} and \texttt{Stata}\footnote{Product of Stata Corp.; \url{http://www.stata.com/}} play a major role for analyzing epidemiological data.
+
 Statistics tools such as \texttt{SPSS}\footnote{Product of IBM; \url{www.ibm.com/software/de/analytics/spss/}} play a major role for analyzing epidemiological data.
@@
 %
-Information about medical history and examinations, genetic conditions, geographical data, questionnaire results and image data yields a complex data space for each subject.
-%
-Often data are derived from acquired variables to either group or threshold values or get information derived from reviewed data such as breast density data for women.
-%
-This underlines also the problem of missing data since for ethical, legal or medical reasons some features can not be gathered for each subject.
-%
-Follow-up examinations or -questions for conditions also yield features only available for a small amount of subjects.
+Information about medical history and examinations, genetic conditions, geographical data, questionnaire results and image data yield a complex data space for each subject.
+%
+%Often data are derived from acquired variables to either group or threshold values or to get information derived from reviewed data such as breast density data for women.
+%
+For ethical, legal or medical reasons some features can not be gathered for each subject.
+%
+The most obvious example is women-specific questions about menstrual status or number of born children.
+%
+Follow-up examinations or -questions for conditions like medications taken after a diagnosed disease also yield features only available for a small amount of subjects.
 %
@@
 %
+Medical examinations mostly comprise categorical (e.g. levels of back pain) and continuous values (e.g. age or body size).
 
@@
 This often yields image resolutions inferior to those of clinical day-to-day practice, which makes their analysis more challenging.
-%
-The equipment used to gather medical image data is kept, if possible, on the initial software and hardware version to ensure comparability in and between acquisition cycles.
+%The equipment used to gather medical image data is kept, if possible, on the initial software and hardware version to ensure comparability in and between acquisition cycles.
+The equipment used to gather medical image data is kept, if possible, on the initial software and hardware version during a longitudinal study to ensure comparability in and between acquisition cycles.
 
@@
 %
-Information derived by landmarks are also not nearly as expressive and versatile as segmentation masks.
+Information derived by landmarks such as top and bottom point are also not nearly as expressive and versatile as segmentation masks describing the whole vertebra.
+%Information derived by landmarks such as top and bottom point of a vertebra are also not nearly as expressive and versatile as segmentation masks describing the whole vertebra.
 %
@@
 %
-Morphometric information from landmarks comprises thickness, diameter or length of a structure as well as grey-value distribution in an area (used for determine type of tissue).
+Morphometric information from landmarks comprises thickness, diameter or length of a structure as well as grey-value distribution in an area (used for determining the type of tisse).
 
@@
 \subsection{The Study of Health in Pomerania (SHIP)}
-After the pioneering Rotterdam study (started in 1990) several MR imaging studies initiatives have evolved, they slightly differ in clinical focus, acquired data and epidemiologic research questions.
-%
-Starting 1997 with a cohort consisting of 4.308 subjects, the SHIP, located in northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
+After the pioneering Rotterdam study (started in 1990) several MR imaging studies initiatives have evolved.
+%
+They slightly differ in clinical focus, acquired data and epidemiological research questions.
+%
+Starting in 1997 with a cohort consisting of 4.308 subjects, the SHIP, located in northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
 %
@@
 %
-Subjects were examined in a 5-year time span, continuously adding new parameters including MRI scans in the last iteration of 2012.
+Subjects were examined in a 5-year time span, continuously adding new parameters including MRI scans in the last iteration \cite{Hegenscheid2009}.
 %
@@
 \\\\
-The work of Turkay and colleagues is closest to ours albeit our focus on processing medical image data and variables with categorical manifestations \cite{Turkay2013}.
+Their focus on hypothesis generation based on parallel assessment of multiple data features makes the work of Turkay and colleagues is closest to ours albeit our emphasis on processing medical image data and variables with categorical manifestations \cite{Turkay2013}.
 %
@@
 %
-Statistical measures of continuous variables such as mean, standard deviation, skewness, or inter-quartile range are used to create \emph{dimension plots} that make them comparable with respect to the derived descriptive measures.
+%Statistical measures of continuous variables such as mean, standard deviation, skewness, or inter-quartile range are used to create \emph{dimension plots} that make them comparable with respect to the derived descriptive measures.
+Statistical measures of continuous variables such as mean, standard deviation, skewness, or inter-quartile range are used to create \emph{dimension plots} that make them comparable with respect to the derived descriptive measures such as voxel number of a segmented structure.
 %
@@
 %
-%The technique was applied to variables generated by segmenting the brain into 45 parts and measure the voxel number, volume and properties of the intensity values.
+%The technique was applied to features obtained by segmenting the brain into 45 parts and measure the voxel number, volume and properties of the intensity values.
 %
@@
 \paragraph{Visualizing Image and Non-Image Data. }
-Gresh and colleagues proposed \texttt{WEAVE}, one of the first systems which analyzed medical image and non-image data using linked views \cite{Gresh2000}.
+Gresh and colleagues proposed \texttt{WEAVE}, one of the first systems, which concurrently analyzed medical image and non-image data using linked views \cite{Gresh2000}.
 %
@@
 %
-Their understanding of a cohort differs from the understanding of the term in an epidemiological context.
+Their understanding of a cohort differs from the understanding of the term in an epidemiological context by denoting every subject group divided by parameter as individual cohort.
 
@@
 %
-It is useful to gain an overview over numerous variables and their distributions.
+They are useful to gain an overview over numerous variables and their distributions.
 %This technique is also useful to gain an overview over numerous variables and their distributions.
@@
 %
-With little effort, linked views can be created, but the data processing possibilities such as derivation of new variables or the volume rendering capabilities are very limited.
+With little effort, linked views can be created, but the data processing possibilities such as derivation of new variables or the 3D rendering capabilities are very limited.
 %
@@
 % Image Data
-\paragraph{Visualizing Image Data.}
+% \paragraph{Visualizing Image Data.}
+\paragraph{Visualizing Shape Variance.}
 Comparing tissue between many subjects in an epidemiological context requires methods which allow for shape variance visualizations.
@@
 When applying clustering techniques on the non-image data it was found that \texttt{k-Prototypes} and \texttt{DBSCAN} is appropriate in the epidemiological context but is strongly dependent on the chosen variables and distance measure \cite{Klemm2014BVM}.
+%
+Niemann and colleagues presented a interactive data mining tool for assessment of risk factors on hepatic steatosis, the fatty liver disease \cite{Niemann2014}.
+%
+Association rules created by data mining methods can be analyzed interactively with their presented tool and highlight potentially overlooked features.
 % *********************** / Our own Work ***********************
@@
 %
-Oeltze and colleagues combined a linked view representation of results from a statistical analysis with feature localizations of the tissue perfusion flow with the goal of its evaluation \cite{Oeltze2007}.
+Oeltze and colleagues combined a linked view representation of results from a statistical analysis with feature localizations of the tissue perfusion with the goal of its evaluation \cite{Oeltze2007}.
 %
@@
 %
-Introducing the \emph{IVA} principle to the epidemiological application domain does aim to complement its weaknesses rather than replacing the existing workflow.
+Introducing the \emph{IVA} principle to the epidemiological application domain does aim to compensate its weaknesses rather than replacing the existing workflow.
 %The workflow we propose by introducing the \emph{IVA} principle into the epidemiological application domain does not aim to replace the existing workflow but to complement its weaknesses.
@@
 %
-Statistical tests on a hypothesis-describing feature list yields a value deciding whether the data supports the hypothesis or not.
+Statistical tests on features associated to a hypothesis yields a value deciding whether the data supports the hypothesis or not.
 %A list of features describing the hypothesis is compiled and analyzed using statistical tests. 

### 2014-03-28 draft.tex
completed Steffens correction

@@
 %\documentclass[electronic,journal]{style/vgtc}     % electronic version, journal
-
+% TODO: Reference Nicolas Kruchten
 %% Uncomment one of the lines above depending on where your paper is
@@
 %
-%Analyzing these data requires image segmentation, extraction of key figures, and shape based subject grouping.
+%Analyzing these data requires image segmentation, extraction of key figures, and shape-based subject grouping.
 %
@@
 % %
-% Analyzing these data requires image segmentation, extraction of key figures, and shape based subject grouping.
+% Analyzing these data requires image segmentation, extraction of key figures, and shape-based subject grouping.
 % %
@@
 \paragraph{Domain and Range Variables. }
-In the \emph{IVA} context, data is divided into two major view types.
-%
-The human body exposing shape information for the \emph{physical view} \cite{Oeltze2013}.
-%
-This information space is usually displayed via volume rendering techniques \cite{Oeltze2007}.
-%
-%Spatiotemporal views include data from the \emph{Object Space Domain} renders data centered to the focused domain, in our example the medical image data space.
-%
-These variables spanning the 3D-space are also referred to as \emph{domain variables}.
-%
-\emph{Range variables} in the epidemiological context can be divided twofold:
-%
+In the \emph{IVA} context, data is characterized by a combination of independent variables, such as space and/or time, and dependent variables, like temperature or pressure.
+%
+Two kinds of views are employed to inspect the data:
 \begin{itemize} 
-	\item Variables derived from the image data. 
-	%
-	These measures quantify shape information to allow for comparison and can also be used to brush in the image space.
-	%
-	%These variables describe image data and can also be used to brush in the image space.
-	\item Epidemiological socio-demographic or medical attribute data.
-	%
-	These values are associated with every subject represented in the image space. 
-	%
-	They do not describe shape information.
-	%
-	This is the data epidemiologists usually want to correlate with image data.
+	\item \emph{physical views} \cite{Oeltze2013}, such as direct volume rendering, show information in the context of the spatiotemporal observation space \cite{Oeltze2007}, while
+	\item \emph{attribute views}, such as scatter plots and parallel coordinates, show relationships between multiple data attributes.
 \end{itemize}
-%Other data associated to elements in the object space are considered to be the \emph{Attribute Space Domain} and are displayed using various range based views using statistical graphics and information visualization techniques like scatterplots, bar charts or parallel coordinates \cite{Oeltze2013, Oeltze2007}.
-%
-%Specific for the epidemiological application domain is that both Object- and Feature Space Domain influence the outcome of interest.
-% \textbf{Object Space}\\
-% - Medical Image Data / Spine Segmentations\\
-% \textbf{Attribute Space}\\
-% - SHIP-Variables\\
-% - Derived visualizations
-% 
-% ? Where to incorporate Levels of IVA
-
-\subsection{IVA Patterns}
-%
-\emph{IVA} knows three different exploration patterns, handling associations of domain and range variables.
-%The explorative procedures when analyzing data using \emph{IVA} can be divided into three different patterns, handlinassociationon between domain and range variables.
-
-\paragraph{Local Investigation}
-This pattern projects information from domain space to the range perspective.
-%
-This step is more complicated in epidemiological context compared to other \emph{IVA} application domains.
-%As opposing to other \emph{IVA} application domains, this step is more complicated in the epidemiological context.
-%
-Shape information can not be brushed using ROI-selections but rather has to employ techniques that specify local deformation changes \cite{Hermann2014} or subjects that belong to a shape class.
-%
-\emph{Feature selection} methods strongly depend on the type of segmentation used to extract the tissue of interest.
-%
-Model-based segmentations or masks yield data structures capable of calculating mean shapes and distances between individuals or subject groups.
-%
-Feature selection is also possible by applying clustering algorithms in order to get shape-groups \cite{Klemm2013VMV}.
-%
-These algorithms can be used to investigate associations between shape-groups and other non-image based variables.
+Transferred to epidemiological data, the residential area of cohort subjects could be interpreted as space, the different assessment cycles of a longitudinal study as time, and the image and non-image data as dependent variables.
+%
+However, our current work neglects geographical and temporal aspects.
+%
+Instead we employ a more abstract model and consider the subjects as living in a joint image space where each of them is represented by a particular segmented organ or structure.
+%
+For instance, the spine is segmented over all subjects and all spines are co-registered spanning a joint space.
+%
+Two types of dependent variables then exist: 
+\begin{enumerate}
+	\item the socio-demographic data and medical examination results and 
+	\item variables derived from the segmented structures, e.g., spinal curvature or misalignment of the vertebrae.
+\end{enumerate}
+An alternative of the image space would be the shape space generated by extracting the major modes of variation from all segmentation results \cite{Busking2010a}.
+%
+Based on our abstract model, the three analysis patterns of \emph{IVA} can be employed.
+
+\paragraph{Local Investigation:} refers to the inspection of dependent variables with respect to certain subsets of the image or shape space.
+%
+For instance, the epidemiologist selects several spines with a common characteristic in the image or shape space and wishes to inspect the associated dependent variables in an attribute view \cite{Hermann2014}.
+%
+The selection step requires dedicated interaction techniques for defining a subset.
+%
+Alternatively, derived shape-related variables opposed in an attribute view or automatic techniques for shape clustering may be employed \cite{Klemm2013VMV}.
+%
+Clustering algorithms can be used to investigate associations between shape-groups and other non-image based variables.
 %
 Analysis of outliers can indicate segmentation errors or an group of subjects sharing a pathology.
-%
-%Another application is the outlier analysis. 
-%
-%Outliers can indicate segmentation errors or an outstanding group of individuals who may share a pathology.
-
-% Local Investigation
-% - Projection from Image Space to Attribute data. This is more difficult in our application, because we currently only brush on derived features.\\
-% - Clustering in Image Space yields Groups which can be analyzed using the attribute space\\
-% - Create a Pipeline Overview over different Levels of different IVA Patterns and Stages
-
-% \subsection{Feature Selection}
-
-% \subsection{Brushing the information space}
-% - derive new groups which serve as input for the feature selection
-
-\paragraph{Feature Localization}
-% Subset selection of range variables yields features
-The vast majority of features are considered to be dependent with respect to the image domain in the \emph{IVA} context.
-%As described before, the vast majority of data points are considered to be dependent with respect to the image domain in the \emph{IVA} context.
-%
-Selecting subjects based on image derived data can be seen as additional possibility of shape-related grouping.
-%
-The epidemiologist is interested in the shape of subjects within a range of a set of variables that describe the current hypothesis.
-%
-Epidemiologists may categorize data into groups that fit their hypothesis formulation.
-%
-Continuous variables such as age are for example often divided in categories like young, aged and elderly.
-%
-%Categorization is strongly dependent on the hypothesis and therefore requires suitable brushing techniques as described in Section~\ref{sec:AdaptiveFeatureVisualization}.
-%
-%To gain insight intassociationons between SHIP-
-
-% - Selection of Information using Bar Charts, Scatterplots or Parallel Coordinates which projects the selection into the Object space\\
-% - this selection is for categorical data already given implicitly bei projecting the 3D-View onto the Bar Charts/Mosaik Plots!\\
-% - this aims to locale features of the data!\\
-% - Gain Information about SHIP-Variables by putting them into the context of each other.
+
+\paragraph{Feature Localization:} refers to the search for structures in the image or shape space with a defined characteristic.
+%
+The epidemiologist may be interested in all female subjects with lower back pain and wishes to see the corresponding spines in a physical 3D view.
+
+\paragraph{Multivariate Analysis:} refers to an investigation of multi-variate properties of the dependent data by specifying a feature in one attribute view and at the same time analyzing the values distribution with respect to other variables in other attribute views.
+%
+The epidemiologist may define a feature in a scatter plot of BMI and age and inspect the result in a histogram of body height.
+%
+These associations may also summarized using pivot tables which are popular in epidemiology.
+
+% ---------------------------- old part ----------------------------
+% In the \emph{IVA} context, data is divided into two major view types.
+% %
+% The human body exposing shape information for the \emph{physical view} \cite{Oeltze2013}.
+% %
+% This information space is usually displayed via volume rendering techniques \cite{Oeltze2007}.
+% %
+% %Spatiotemporal views include data from the \emph{Object Space Domain} renders data centered to the focused domain, in our example the medical image data space.
+% %
+% These variables spanning the 3D-space are also referred to as \emph{domain variables}.
+% %
+% \emph{Range variables} in the epidemiological context can be divided twofold:
+% %
+% \begin{itemize}
+% 	\item Variables derived from the image data. 
+% 	%
+% 	These measures quantify shape information to allow for comparison and can also be used to brush in the image space.
+% 	%
+% 	%These variables describe image data and can also be used to brush in the image space.
+% 	\item Epidemiological socio-demographic or medical attribute data.
+% 	%
+% 	These values are associated with every subject represented in the image space. 
+% 	%
+% 	They do not describe shape information.
+% 	%
+% 	This is the data epidemiologists usually want to correlate with image data.
+% \end{itemize}
+% %Other data associated to elements in the object space are considered to be the \emph{Attribute Space Domain} and are displayed using various range based views using statistical graphics and information visualization techniques like scatterplots, bar charts or parallel coordinates \cite{Oeltze2013, Oeltze2007}.
+% %
+% %Specific for the epidemiological application domain is that both Object- and Feature Space Domain influence the outcome of interest.
+% % \textbf{Object Space}\\
+% % - Medical Image Data / Spine Segmentations\\
+% % \textbf{Attribute Space}\\
+% % - SHIP-Variables\\
+% % - Derived visualizations
+% % 
+% % ? Where to incorporate Levels of IVA
+% 
+% \subsection{IVA Patterns}
+% %
+% \emph{IVA} knows three different exploration patterns, handling associations of domain and range variables.
+% %The explorative procedures when analyzing data using \emph{IVA} can be divided into three different patterns, handlinassociationon between domain and range variables.
+% 
+% \paragraph{Local Investigation}
+% This pattern projects information from domain space to the range perspective.
+% %
+% This step is more complicated in epidemiological context compared to other \emph{IVA} application domains.
+% %As opposing to other \emph{IVA} application domains, this step is more complicated in the epidemiological context.
+% %
+% Shape information can not be brushed using ROI-selections but rather has to employ techniques that specify local deformation changes \cite{Hermann2014} or subjects that belong to a shape class.
+% %
+% \emph{Feature selection} methods strongly depend on the type of segmentation used to extract the tissue of interest.
+% %
+% Model-based segmentations or masks yield data structures capable of calculating mean shapes and distances between individuals or subject groups.
+% %
+% Feature selection is also possible by applying clustering algorithms in order to get shape-groups \cite{Klemm2013VMV}.
+% %
+% These algorithms can be used to investigate associations between shape-groups and other non-image based variables.
+% %
+% Analysis of outliers can indicate segmentation errors or an group of subjects sharing a pathology.
+% %
+% %Another application is the outlier analysis. 
+% %
+% %Outliers can indicate segmentation errors or an outstanding group of individuals who may share a pathology.
+% 
+% % Local Investigation
+% % - Projection from Image Space to Attribute data. This is more difficult in our application, because we currently only brush on derived features.\\
+% % - Clustering in Image Space yields Groups which can be analyzed using the attribute space\\
+% % - Create a Pipeline Overview over different Levels of different IVA Patterns and Stages
+% 
+% % \subsection{Feature Selection}
+% 
+% % \subsection{Brushing the information space}
+% % - derive new groups which serve as input for the feature selection
+% 
+% \paragraph{Feature Localization}
+% % Subset selection of range variables yields features
+% The vast majority of features are considered to be dependent with respect to the image domain in the \emph{IVA} context.
+% %As described before, the vast majority of data points are considered to be dependent with respect to the image domain in the \emph{IVA} context.
+% %
+% Selecting subjects based on image derived data can be seen as additional possibility of shape-related grouping.
+% %
+% The epidemiologist is interested in the shape of subjects within a range of a set of variables that describe the current hypothesis.
+% %
+% Epidemiologists may categorize data into groups that fit their hypothesis formulation.
+% %
+% Continuous variables such as age are for example often divided in categories like young, aged and elderly.
+% %
+% %Categorization is strongly dependent on the hypothesis and therefore requires suitable brushing techniques as described in Section~\ref{sec:AdaptiveFeatureVisualization}.
+% %
+% %To gain insight intassociationons between SHIP-
+% 
+% % - Selection of Information using Bar Charts, Scatterplots or Parallel Coordinates which projects the selection into the Object space\\
+% % - this selection is for categorical data already given implicitly bei projecting the 3D-View onto the Bar Charts/Mosaik Plots!\\
+% % - this aims to locale features of the data!\\
+% % - Gain Information about SHIP-Variables by putting them into the context of each other.
 The Pivot Table allows for direct numerical\\ analysis, while the information visualizations allow for better insight of the combination
-
-\paragraph{Multivariate Analysis}
-%Introduced in the information visualization, the multivariate analysis means selecting subjects in like in the feature localization step and get have the result highlighted in another linked view displaying other non-image parameter.
-%Introduced in the information visualization, the multivariate analysis incorporates brushing and linking of of views displaying non-image parameter.
-Multivariate analysis incorporates brushing and linking of of views displaying non-image parameter.
-%
-The type of the parameters which are compared determines the information visualization technique used.
-%
-%This represents for the most part the usual approach which essentially is the analysis of how variables interact with each other, only in a visual analysis context.
-%
-Statistic measures are needed which describe how variables correlate given the selected groups.
-%The need of statistic measures which describe how variables correlate with each other given the selected groups is special for the application domain.
-%
-These associations are also summarized using pivot tables which are popular in epidemiology.
-% - Selection of Elements in Scatterplots, Barcharts or Parallel Coordinates Views - observe how selection changes another view - this allows for multivariate analysis\\
-% - Becker, R.A., Cleveland, W.S.: Brushing scatterplots. Technometrics 29(2) (1987)\\
-% - Wang Baldonado, M.Q., Woodruff, A., Kuchinsky, A.: Guidelines for using multiple views in information visualization.
+% 
+% \paragraph{Multivariate Analysis}
+% %Introduced in the information visualization, the multivariate analysis means selecting subjects in like in the feature localization step and get have the result highlighted in another linked view displaying other non-image parameter.
+% %Introduced in the information visualization, the multivariate analysis incorporates brushing and linking of of views displaying non-image parameter.
+% Multivariate analysis incorporates brushing and linking of of views displaying non-image parameter.
+% %
+% The type of the parameters which are compared determines the information visualization technique used.
+% %
+% %This represents for the most part the usual approach which essentially is the analysis of how variables interact with each other, only in a visual analysis context.
+% %
+% Statistic measures are needed which describe how variables correlate given the selected groups.
+% %The need of statistic measures which describe how variables correlate with each other given the selected groups is special for the application domain.
+% %
+% These associations are also summarized using pivot tables which are popular in epidemiology.
+% % - Selection of Elements in Scatterplots, Barcharts or Parallel Coordinates Views - observe how selection changes another view - this allows for multivariate analysis\\
+% % - Becker, R.A., Cleveland, W.S.: Brushing scatterplots. Technometrics 29(2) (1987)\\
+% % - Wang Baldonado, M.Q., Woodruff, A., Kuchinsky, A.: Guidelines for using multiple views in information visualization.
 In: AVI ’00: Proceedings of the working conference on Ad- vanced visual interfaces, pp.
@@
 DOI\\ http://doi.acm.org/10.1145/345513.345271
-% - By brushing individual parameters or create new binnings of parameter it is possible to see how they change in coordinated views. This is already implemented in the Cargo framework\\
-% - by creating comparative 3d Visualizations it is possible to assess the influence of non-image parameters to the visual space.
+% % - By brushing individual parameters or create new binnings of parameter it is possible to see how they change in coordinated views. This is already implemented in the Cargo framework\\
+% % - by creating comparative 3d Visualizations it is possible to assess the influence of non-image parameters to the visual space.
+% 
+% ---------------------------- old part ----------------------------
 
@@
 %
-Including the data dictionary, which stores information about each manifestation of a feature, helps to get a detailed description of data variables and the meaning and unit of measurement of their values.
+%Including the data dictionary, which stores information about each manifestation of a feature, helps to get a detailed description of data variables and the meaning and unit of measurement of their values.
+A data dictionary stores information about each manifestation of a feature and is stored as lookup table to get a detailed description of data variables and the meaning and unit of measurement of their values.
 %Exporting the data dictionary, which stores information about each manifestation of a feature, is also an important step to get a detailed description of data variables and the meaning and unit of measurement of their values.
@@
 %
-Hypothesis-generation with focus on image-data starts with a shape based clustering.
+Hypothesis-generation with focus on image-data starts with a shape-based clustering.
 %
@@
 %
+Brushing methods are subdivided using the previously described \emph{IVA} patterns.
+%
 Brushes act as feature as they divide the subject space just like categorical variables and lead therefore to a new variable selection.
 %
-Selecting features also triggers a contingency analysis highlighting variables interacting with selected features, which is described in the following section.
-
+Selecting features also triggers a \emph{multivariate analysis} using contingency values (described in the following section) to highlight variables interacting with selected features.
+%
 %The starting point of \emph{IVA Loop} is determined whether the analysis is hypothesis-driven or exploratory.
@@
 %
-A sample workflow using interaction- and visualization techniques described in the following section can be seed in Figure~\ref{fig:visualization}.
+A sample workflow using interaction- and visualization techniques described in the next section can be seen in Figure~\ref{fig:visualization}.
 
@@
  \caption{Adjacency matrix of 131 features with a grand total of 17.161 combinations.
+ %
  Similarity is calculated using the \emph{Cram\'{e}rs' V} contingency value.
+ %
  Color brightness encodes association strength.
+ %Mouse-over focuses a entry and highlights the feature names for better readability.
+ Mouser-over an entry highlight the feature names for better readability.
+ %
  The enlarged excerpt shows associations for shape clusters of subjects with and without diagnosed spine attrition, which show associations between gender, weight, body height and smoking behavior.
@@
 \subsubsection{System Structure} \label{Structure and Workflow}
-We divide the workspace into four major parts as seen in Figure~\ref{fig:visualization}.
+We divide the workspace into four major parts as seen in Figure~\ref{fig:visualization} and Figure~\ref{fig:similarity}.
 \begin{itemize}
@@
 	\item The interactive Pivot Table gives detailed numerical information of the features in the canvas view. This view on the data come natural to epidemiologists.
-	\item The contingency view depicting relations for features in the canvas.
+	\item The contingency view depicting relations for features in the canvas in an adjacency matrix.
 \end{itemize}
@@
 %
-Features can be dragged from the sidebar into the canvas area.
+Variables can be dragged from the sidebar into the canvas area for a \emph{feature localization}.
 %
@@
 \paragraph{Adaptive Feature Visualization.} \label{sec:AdaptiveFeatureVisualization}
-Inspired be previously the discussed \texttt{GPloms} \cite{Francois2013} the visualization type is chosen dynamically based on the variable types and number of visualizations which need to be displayed.
+The visualization type is, inspired by \texttt{GPloms} \cite{Francois2013}, dynamically chosen based on the feature types and number.
+%
+Figure \ref{fig:visualization} describes this dynamic adjustment.
+%Inspired by the previously discussed \texttt{GPloms} \cite{Francois2013} the visualization type is chosen dynamically based on the variable types and number of visualizations which need to be displayed.
 %
@@
 %
-Epidemiologists are used to process groups based on table representations so decided to introduce an interactive pivot table.
+Epidemiologists are used to process groups based on table representations so we decided to introduce an interactive pivot table.
 %When opening a random epidemiological paper the reader will in almost all cases find some sort of pivot table to present the data.
@@
 %
-Pivot table quickly get confusing and cluttered when they are divided into to many subgroups.
+Pivot tables quickly get confusing and cluttered when they are divided into many subgroups.
 %
@@
 %
-This also applies on the designation of Row or Column Variables.
+This also applies on the designation of row or column variables.
 %
@@
 \paragraph{Automated Feature Suggestion using a Contingency Matrix.}
-As discussed previously, highlighting potential interesting values in the data set is one major benefit of the \emph{IVA} powered approach.
+As discussed previously, highlighting potential interesting values in the data set is one major benefit of the \emph{IVA} powered approach and is categorized in the \emph{multivariate analysis} pattern analyzing features outside the shape-space.
 %
@@
 %
-Therefore are manifestations who are only exposed by small subject groups not assessable using this technique.
+Therefore are manifestations, which are only exposed by small subject groups not assessable using this technique.
 %
@@
 	High confidentiality standards of the data can be met by restricting access via a account system
-	\item Recent developments in WebGL applications running in browsers with near native performance push the development into to the web which results in many open source libraries which are well documented, rich in examples and driven by active communities.
+	\item Recent developments in WebGL applications running in browsers with near native performance push the development into to the web, which results in many open source libraries which are well documented, rich in examples and driven by active communities.
+We use WebGL for rendering shape information.
 	\item Web technologies are free.
@@
 %
-Many methods which specialized libraries, like the Visualization Toolkit (VTK) or R for statistics have build in, need to be written from scratch in order to fit in the context.
-%
-
-The back-end is written using \texttt{NodeJS}\footnote{Developed by Joyent Inc, \url{nodejs.org}}, which based on the Google V8 Javascript runtime environment.
+%Many methods which specialized libraries, like the Visualization Toolkit (VTK) or R for statistics have build in, need to be written from scratch in order to fit in the context.
+Due to missing ports of sophisticated specialized libraries/languages, such as the Visualization Toolkit (VTK) or R for statistics, many standard methods need to be written from scratch for the web.
+%Many methods, which specialized libraries, like the Visualization Toolkit (VTK) or R for statistics, have build in, need to be written from scratch in order to fit in the context.
+%
+
+The back-end is written using \texttt{NodeJS}\footnote{Developed by Joyent Inc, \url{nodejs.org}}, which is based on the Google V8 Javascript runtime environment.
 %
@@
 All parameter steered console applications can be incorporated in this context.
-
-The following section describes how our presented \emph{IVA} workflow works for a epidemiological use case.
+\\\\
+The following section describes how our presented \emph{IVA} workflow works for an epidemiological use case.
 
@@
 %
-It is one of the most common reasons for an adult to see a physicians in the western civilization \cite{Backpain}.
+It is one of the most common reasons for an adult to see a physician in the western civilization \cite{Backpain}.
 %
@@
 %
-Determining risk factors in this area can lead to \cite{Fletcher2012} (evtl. weglassen, Dopplung bei Epidemiological Background?):
+Determining risk factors in this area can lead to \cite{Fletcher2012}:
 \begin{itemize}
-	\item a better understanding of effects of preventive measures such as occupational health and safety regulations
-	\item prognostic features for diagnosis and treatment of lumbar back pain
-	\item determination of particularly effected risk groups
+	\item a better understanding of effects of preventive measures such as occupational health and safety regulations,
+	\item prognostic features for diagnosis and treatment of lumbar back pain,
+	\item determination of particularly effected risk groups.
 \end{itemize}
@@
 %
-Our clinical partners compiled a feature list which is a tradeoff between complexity and limitations of the responsible ethics committee. (vllt. etwas hart formuliert).
+Our epidemiological collaborators compiled a feature list which is a tradeoff between complexity and limitations of the responsible ethics committee.
 %
@@
 \subsection{Data Preprocessing} \label{application:Data Preprocessing}
-Following methods described in Section~\ref{Data Preprocessing}, the data is preprocessed as follows for the presented prototype.
+The data are preprocessed as described in Section~\ref{Data Preprocessing}.
+%Following methods described in Section~\ref{Data Preprocessing}, the data is preprocessed as follows for the presented prototype.
 %Transformation operations on the data to prepare it for the presented prototype are denoted as data preprocessing.
@@
 %
-Each feature is stored as object which contains: 
+Each feature is stored as an object which contains: 
 \begin{itemize}
@@
 %
-For clustering purposes we extracted the centerline of the spine canal of the lumbar spine canal which captures information about lordosis and scoliosis which are medical terms for spine curvature \cite{Klemm2013VMV}.
+For clustering purposes we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis which are medical terms for spine curvature \cite{Klemm2013VMV}.
 
@@
 %
-The tetrahedron-based model detection model described in Section~\ref{application:Data Preprocessing} consists of corresponding grid points for each structure instance.
+The tetrahedron-based detection model described in Section~\ref{application:Data Preprocessing} consists of corresponding grid points for each structure instance.
 %
@@
 \\\\
-Shape based clustering is carried out via Agglomerative Hierarchical clustering of the spine canal, centerlines which are described in Section~\ref{application:Data Preprocessing} \cite{Klemm2013VMV}.
-%
-Since it is not possible to determine the number of clusters in a given group, it is automatically computed using a \emph{knee/elbow} point which is described as tradeoff between number of cluster and a cluster evaluation metric \cite{Salvador2004}.
+Shape-based clustering is carried out via Agglomerative Hierarchical clustering of the spine canal, centerlines which are described in Section~\ref{application:Data Preprocessing} \cite{Klemm2013VMV}.
+%
+Since the "correct" number of clusters in a given group is unknown, an estimate is computed by means of the knee/elbow method \cite{Salvador2004}.
+%Since it is not possible to determine the number of clusters in a given group, it is automatically computed using a \emph{knee/elbow} point which is described as tradeoff between number of cluster and a cluster evaluation metric \cite{Salvador2004}.
 %
@@
 %
-\subsubsection{Hypothesis free Analysis} \label{Hypothesis free Analysis}
+The use of \emph{IVA} patterns as presented in Section~\ref{Image Centric Cohort Study Data in Interactive Visual Analysis Context} is denoted accordingly.
+%
+\subsubsection{Hypothesis-Free Analysis} \label{Hypothesis free Analysis}
 \begin{figure}[htb]
@@
 %
-Shape based hypothesis free exploration starts with a shape-grouping step achieved by image-based clustering.
+%Shape-based hypothesis free exploration starts with a shape-grouping step achieved by shape-based clustering.
+Shape-based hypothesis free exploration starts with the \emph{local investigation} pattern by a shape-grouping step using shape-based clustering.
 %
@@
 %
-Looking at the \emph{Cram\'{e}rs' V} contingency values of the group reveal associations of this group with employment status, body size, age, thyroid nodules and blood-fat value.
-%
-TODO Feedback Greifswald, 
-While all observed features seem as plausible associations related to back pain, the values indicate subjects with chronic back pain which radiates into the legs.
+Performing a \emph{multivariate analysis} by looking at the \emph{Cram\'{e}rs' V} contingency values of Cluster~8 reveals associations of this group with employment status, body size, age, thyroid nodules and blood-fat value.
+%
+While all observed features seem as plausible associations related to back pain, the values indicate that it contains subjects with chronic back pain radiating to the legs.
 %
@@
 
-\subsubsection{Hypothesis based Analysis} \label{Hypothesis based Analysis}
+\subsubsection{Hypothesis-Based Analysis} \label{Hypothesis based Analysis}
 \begin{figure}[htb]
@@
 %
-If the user has already a hypothesis about a relation between a non-image feature regarding shape the workflow slightly differs from the hypothesis free analysis.
-%
-The starting point of the analysis is the selection of a feature of interest by dragging it into the canvas area and view the subjects distribution as well as their shape differences.
-%
-In our use case, the epidemiologist was interested in the questionnaire answer "Did you experience back pain in the past three month".
+If the user has already a hypothesis about a relation between a non-image feature regarding shape the workflow slightly differs from the hypothesis-free analysis.
+%
+The starting point of the analysis is a \emph{feature investigation}, where a feature of interest is selected by dragging it into the canvas area and view the subjects distribution as well as their shape differences.
+%
+In our use case, the epidemiologist was interested in the dichotomous questionnaire answer to "Did you experience back pain in the past three month".
 %
@@
 
-\emph{Cram\'{e}rs' V} contingency values highlighted relationships of this cluster with joint degeneration, meat eating habits, preoccupation, back pain, neck or shoulder pain and waist circumference.
+\emph{Multivariate analysis} using \emph{Cram\'{e}rs' V} contingency values highlighted relationships of this cluster with joint degeneration, meat eating habits, preoccupation, back pain, neck or shoulder pain and waist circumference.
 %
@@
 %
-Our clinical partner were interested in the fact that this group presented a correlation with waist circumference.
+Our epidemiological collaborators were interested in the fact that this group presented a correlation with waist circumference.
 %
@@
 \subsubsection{Follow-Up Tasks and Concluding Domain-Expert Feedback}
+%
 Defining a causal relationship solely based on observed correlations of two features is \emph{cum hoc ergo propter hoc}--correlation does not automatically imply causation \cite{Tufte2003}.
@@
 \\\\
+The prototype was demonstrated to the domain experts (with background on epidemiology and radiology) via screen sharing.
+%
+The software was operated by a computer scientist following instructions of the domain experts.
+%
 The presented methods guide attention to features which where not in the focus of attention and expectance of clinical researchers.
@@
 %
-Shape based clustering assesses variability of a anatomical structure in the context of non-image features such as disease indicators or lifestyle factors.
+Shape-based clustering assesses variability of a anatomical structure in the context of non-image features such as disease indicators or lifestyle factors.
 %

### 2014-03-29 draft.tex
Added Figure Working files to repo, improved images

@@
  %\includegraphics[width=1.5in]{figures/InteractionLoop}
- \includegraphics[width=3.5in]{figures/placeholder}
- \caption{\emph{Image is not final}. (a) "Did you experience back pain in the past three month" Yes No; (b) Clustering of Yes}
+ \includegraphics[width=3.5in]{figures/hypothesisbased}
+ \caption{(a) Dichotomous questionnaire answer to "Did you experience back pain in the past three month".
+ %
+ Mean-Shapes between the groups show no difference.
+ %
+ (b) Shape-based clustering for all subjects who suffered from back-pain yields 6 groups. 
+ %
+ Note that the difference in subject count is due to the missing shape information for some subjects.
+ %
+% The total count of subjects in (b) is smaller 
+ }
  \label{fig:hypopthesisbased}

### 2014-03-29 draft.tex
Included changes from Bernhard

@@
 %
-The data is analyzed using a priori defined hypotheses to find statistically significant correlations between variables (\emph{associations}).
+The data are analyzed using a priori defined hypotheses to find statistically significant correlations between variables (\emph{associations}).
 %
@@
 %
-%Shape based grouping of subjects is facilitated using clustering techniques which operate on surface meshes extracted from the image segmentation.
+%Shape-based grouping of subjects is facilitated using clustering techniques which operate on surface meshes extracted from the image segmentation.
 } % end of abstract
@@
 % %
-% Shape based grouping of subjects is facilitated using clustering techniques which operate on surface meshes extracted from the image segmentation.
+% Shape-based grouping of subjects is facilitated using clustering techniques which operate on surface meshes extracted from the image segmentation.
 % } % end of abstract
@@
 %
-These are used to determine environmental and lifestyle factors as well as medical attributes which may influence a condition of interest.
-%
-The data variables necessary are gathered using structured interviews and clinical examinations.
+These are used to determine environmental and lifestyle factors as well as medical examinations results, which may influence a condition of interest.
+%
+The data variables, that are potentially useful, are gathered using structured interviews and clinical examinations.
 %
@@
 %
+The cohort consists of several thousands of persons randomly selected to avoid any bias.
+%
+Due to this random selection, the majority of the cohort does not suffer from a specific health problem.
+%
+A large size of the cohort is therefore essential to investigate difference between healthy and diseased people.
+%
 % The sample is drawn randomized to avoid selection \cite{Fletcher2012}.
@@
 %These data needs to be segmented to analyze its features for correlations with non-image data.
-These data need to be concurrently analyzed with the non-image data.
-This requires their segmentation and the derivation of features such as kidney volume, curvature of the spine or tissue texture of the liver.
+The segmentation of these data would allow to analyze shapes of anatomical structures and to correlate them with other variables.
+%These data needs to be concurrently analyzed with the non-image data.
+%
+%This requires their segmentation and the derivation of features such as kidney volume, curvature of the spine or tissue texture of the liver.
 %
@@
 %
-Semi-automatic techniques are more promising but challenging since the used methods as MRI and ultrasound are subject to inhomogeneity and noise.
+Semi-automatic techniques are more promising but challenging since the used modalities as Magnetic resonance imaging (MRI) and ultrasound are subject to inhomogeneity and noise.
 %
@@
 %
-The standard workflow lacks methods which highlight features of interest the epidemiologists did not consider.
+The standard workflow lacks methods, which identifies correlation the epidemiologists did not consider.
 %The hypothesis driven analysis approach of these data based on expectations of epidemiologists leaves out features which potentially interact with a disease outside.
@@
 %
-Intended as an extension to the well established epidemiological tools it provides a way to rapidly validate hypothesis as well as trigger \emph{hypothesis generation} using Data Mining methods such as clustering.
+Intended as an extension to the well established epidemiological tools it provides a way to rapidly validate hypothesis as well as trigger \emph{hypothesis generation} using data mining methods such as clustering.
+%
+Hypothesis generation gains importance since the number of epidemiological variables increases and the focus shifts towards more complex relations involving more than two variables.
 %
@@
 We applied our approach to a data set compiled to analyze diseases related to the lumbar spine and aim to determine features, which indicate pathological changes.
+%
+This data set comprises 127 variables and 2 sequences of MRI data from 2.333 patients.
 \\\\
@@
 \subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
+\begin{figure}[htb]
+ \centering
+ \includegraphics[width=3.0in]{figures/workflow_comparison}
+ \caption{\emph{IVA} tools are able to complement parts of the epidemiological workflow, not replace it.
+The appropriate combination of statistical- and interactive driven analysis shows promising potential to unveil the information in the data.
+(a) shows the standard epidemiological workflow, (b) the IVA supported one.
+The iterative red highlighted part is called the \emph{IVA Loop}.}
+  \label{fig:WorkflowComparison}
+\end{figure}
 %Epidemiology diversity is reflected in the different experts working at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
@@
 %
-Epidemiologists follow a workflow mainly driven by statistic tools to validate hypotheses about disease specific risk factors.
-%
-Following Thew and colleagues, the workflow, shown in Figure~\ref{fig:WorkflowComparison} (a), can be characterized as follows \cite{Thew2009}:
+Epidemiologists follow a workflow mainly driven by statistic tools to validate hypotheses about disease-specific risk factors.
+%
+Following Thew and colleagues, the workflow can be characterized as follows \cite{Thew2009}:
 %
@@
 \end{enumerate}
+The workflow is shown in Figure~\ref{fig:WorkflowComparison} (a) and serves as orientation for our approach although it doesn't consider any kind of image data.
+%
+The extension, we suggest, are motivated by the potential of image data and the attempt to support \emph{hypothesis generation}.
 %In the following we discuss the problems and challenges regarding the workflow.
@@
 %
+It is difficult to achieve, since many physicians are involved when thousands of test persons are examined and interviewed.
+%
+Thus, both intra- and inter observer variability needs to be low for all aspects of a cohort study examination.
+%
 Longitudinal studies require the acquired attributes to be comparable to evaluate them.
 %
-If the data acquisition process changes, a information bias is introduced to the data, hampering inference in acquisition cycles.
+If the data acquisition process changes, an information bias is introduced to the data, hampering inference in and between acquisition cycles.
 %
@@
 %
-Statistical correlations are prone to \emph{confounding}, meaning that the association of two features are influenced by a third feature which needs to isolated.
+Statistical correlations are prone to \emph{confounding}, meaning that the association of two features are influenced by a third feature which needs to be isolated.
 %
@@
 \subsection{Epidemiological Data} \label{EpidemiologicalData}
-Epidemiological data is highly heterogenous and incomplete.
+Epidemiological data are highly heterogenous and incomplete.
 %
@@
 
-\paragraph{Image acquisition.} \label{ImageAcquisition} Imaging techniques emitting hazardous amounts of radiation for the subject are not suited for ethical reasons.
+\paragraph{Image acquisition.} \label{ImageAcquisition} Imaging techniques involving ionizing radiation for the subject are not suited for ethical reasons.
 %
@@
 %
-Since reliable and efficient segmentation techniques for these data are not available in general, the epidemiologists are forced to measure the data by hand, which is a very tedious work with respect to the number of necessary landmarks and number of subjects.
-%
-Information derived by landmarks such as top and bottom point are also not nearly as expressive and versatile as segmentation masks describing the whole vertebra.
+Since reliable and efficient segmentation techniques for these data are not available in general, epidemiologists are forced to measure the data by hand, which is a very tedious work with respect to the number of necessary landmarks and number of subjects.
+%
+Information derived by landmarks such as top and bottom point are by far not as expressive and versatile as segmentation masks describing the whole vertebra.
 %Information derived by landmarks such as top and bottom point of a vertebra are also not nearly as expressive and versatile as segmentation masks describing the whole vertebra.
@@
 %
-This gains even more momentum when analyzing multiple time steps.
+%This gains even more momentum when analyzing multiple time steps.
+This gains even more importance when analyzing multiple time steps.
 %
@@
 %
-The protocols for analyzing the subjects between \texttt{SHIP} and \texttt{SHIP-Trend} remained the same, making them comparable.
+The protocols for examining the subjects between \texttt{SHIP} and \texttt{SHIP-Trend} remained the same, making them comparable.
 %
@@
 %
-The Principal Component Analysis (PCA) and similar techniques are able to reduce the dimension by extracting most expressive components, but make the influence of each variable hard to determine.
+The Principal Component Analysis (PCA) and similar techniques are able to reduce the dimension by extracting most expressive components, but make the influence of each variable hard to convey.
 \\\\
@@
 \paragraph{Commercial Data Visualization.}
-Commercial systems such as \texttt{Tableau}\footnote{Owned by Tableau Software; \url{www.tableausoftware.com}} or \texttt{Spotfire}\footnote{Owned by TIBCO; \url{spotfire.tibco.com}} provide a rich user interface that allows to apply Visual Analytics techniques without the need of writing any code.
+Commercial systems such as \texttt{Tableau}\footnote{Owned by Tableau Software; \url{www.tableausoftware.com}} or \texttt{Spotfire}\footnote{Owned by TIBCO; \url{spotfire.tibco.com}} provide a rich user interface that enable Visual Analytics approach without the need of writing any code.
 %
@@
 %
-Via mesh morphing interpolated views can be created by the user in a separate view as well as comparisons in a contour view.
+Interpolated views can be created by the user in a separate view as well as comparisons in a contour view.
+%
+Interpolation is carried out by mesh morphing.
 %
@@
 %
-We incorporate the idea of combining 3D-Shape rendering with information visualization techniques.
+Derived from this we incorporate the idea of combining 3D-Shape rendering with information visualization techniques.
 %
@@
 %
-Hermann and colleagues identify local deformation changes by investigating shape related difference \cite{Hermann2014}. 
+Hermann and colleagues identify local deformation changes by investigating shape-related difference \cite{Hermann2014}. 
 %
@@
 \paragraph{SHIP-Data analysis.}
-Klemm and colleagues visualized lumbar spine variabilities based on an semi-automatic shape-detection algorithm of 490 participants of the \texttt{SHIP-2} cohort \cite{Klemm2013VMV}.
+Klemm and colleagues visualized lumbar spine variabilities based on an semi-automatic shape detection algorithm of 490 participants of the \texttt{SHIP-2} cohort \cite{Klemm2013VMV}.
 %
@@
 %
-
-While we take similar steps when analyzing the data such as employing statistical tests, our data is mostly independent from the medical image data and is not describing it.
+This was focused on individual patients and not on cohort study data. 
+%
+
+%While we take similar steps when analyzing the data such as employing statistical tests, our data is mostly independent from the medical image data and is not describing it.
 %--except the variables derived the segmentation model itself.
+While we take similar steps when analyzing the data such as employing statistical tests, our data are not associated in a geographical or temporal context, which affects the techniques described in the following section.
 
@@
 \section{Image Centric Cohort Study Data in Interactive Visual Analysis Context} \label{Image Centric Cohort Study Data in Interactive Visual Analysis Context}
-\begin{figure}[htb]
- \centering
- \includegraphics[width=3.0in]{figures/workflow_comparison}
- \caption{\emph{IVA} tools are able to complement parts of the epidemiological workflow, not replace it.
-The appropriate combination of statistical- and interactive driven analysis shows promising potential to unveil the information in the data.
-(a) shows the standard epidemiological workflow, (b) the IVA supported one.
-The iterative red highlighted part is called the \emph{IVA Loop}.}
-  \label{fig:WorkflowComparison}
-\end{figure}
 %- Acceptance plays a big role - you have to pick up the user by presenting information they are used to in a new way\\
 %
-Subsection~\ref{EpidemiologicalWorkflow} described the epidemiological workflow as a sequence of steps taken by domain experts which need to comprise reproducibility and statistical integrity.
+We described the epidemiological workflow as a sequence of steps taken by domain experts which need to comprise reproducibility and statistical integrity (recall Subsection~\ref{EpidemiologicalWorkflow}).
 %
@@
 %
-This has to be handled with care since \emph{overfitting} of expectations to the data is an imminent danger as described by Turkay and colleagues \cite{Turkay2013}.
+This has to be handled with care since \emph{overfitting} of expectations to the data are an imminent danger \cite{Turkay2013}.
 
 \paragraph{Domain and Range Variables. }
-In the \emph{IVA} context, data is characterized by a combination of independent variables, such as space and/or time, and dependent variables, like temperature or pressure.
+In the \emph{IVA} context, data are characterized by a combination of independent variables, such as space and/or time, and dependent variables, like temperature or pressure.
 %
@@
 %
-The epidemiologist may define a feature in a scatter plot of BMI and age and inspect the result in a histogram of body height.
+Epidemiologists may define a feature in a scatter plot of BMI and age and inspect the result in a histogram of body height.
 %
@@
 % %
-% Selecting subjects based on image derived data can be seen as additional possibility of shape-related grouping.
-% %
-% The epidemiologist is interested in the shape of subjects within a range of a set of variables that describe the current hypothesis.
+% Selecting subjects based on image-derived data can be seen as additional possibility of shape-related grouping.
+% %
+% Epidemiologist are interested in the shape of subjects within a range of a set of variables that describe the current hypothesis.
 % %
@@
 %
-Missing data is denoted using Error codes indicating its cause ranging from ethical to medical and personal issues.
+Missing data are denoted using error codes indicating its cause ranging from ethical to medical and personal issues.
 %The reasons for missing data have a wide range from ethical to medical and personal issues.
@@
 \paragraph{Image data. }
-Processing the image data associated to each subject consists for the most part of information extraction about a anatomical structure.
+Processing the image data associated to each subject consists for the most part of information extraction about an anatomical structure.
 %
@@
 %
-Intensity-based comparisons can be achieved using rigid image registration, but model based results however are preferable \cite{Klemm2012}.
+%Intensity-based comparisons can be achieved using rigid image registration, but model based results however are preferable \cite{Klemm2012}.
 %
@@
 
-Morphometric variables are derived to allow for statistical comparison of the tissue which incorporate mostly positions, diameters, volumes and relative distances and alignment to other structures.
+Morphometric variables are derived to allow for statistical comparison of the tissue, which incorporate mostly positions, diameters, volumes and relative distances and alignment to other structures.
 % 
@@
  \caption{Detailed Version of the \emph{IVA Loop}.
-Usually starting with a selection of Feature of interest (user-driven or via data mining techniques), the data is mapped using a visualization technique appropriate for the selected data types.
-The data is visualized in the range and domain space, which can then be brushed, yielding new groups which can be investigated using further features.
+Usually starting with a selection of Feature of interest (user-driven or via data mining techniques), the data are mapped using a visualization technique appropriate for the selected data types.
+The data are visualized in the range and domain space, which can then be brushed, yielding new groups, which can be investigated using further features.
 Note that adjacent steps are directly connected via feedback loops allowing for iterative refinement and give as much freedom to the user as possible.}
@@
 %
-A feature selection can also be derived from a shape-based clustering which creates shape groups.
-%
-Hypothesis-generation with focus on image-data starts with a shape-based clustering.
-%
-The feature is mapped using a automatically chosen visualization appropriate for its data type.
-%
-The visualization techniques have to combine both image-and non image data in order to set domain and range data in relation to each other.
+A feature selection can also be derived from a shape-based clustering, which creates shape groups.
+%
+\emph{Hypothesis-generation} with focus on image data starts with a shape-based clustering.
+%
+The feature is mapped using a automatically chosen visualization appropriate for its data type (described in detail in the following section).
+%
+The visualization techniques have to combine both image-and non-image data in order to set domain and range data in relation to each other.
 %
@@
 %
-Brushes act as feature as they divide the subject space just like categorical variables and lead therefore to a new variable selection.
+Brushed regions are treated like features as they divide the subject space just like categorical features. %and lead therefore to a new feature selection.
 %
@@
  \includegraphics[width=1\textwidth, resolution=300]{figures/visualization}
- \caption{(Left) Screenshot from the front-end which is divided as follows: (a) Sidebar which contains all features as well as the groups defined in the analysis process; (b) Canvas area where features can be added via drag and drop and the visualization chosen automatically according to the data type; (c) Interactive Pivot Table which shows exact numbers for each displayed variable combination.
+ \caption{(Left) Screenshot from the front-end, which is divided as follows: (a) Sidebar containing all features as well as the groups defined in the analysis process; (b) Canvas area where features can be added via drag and drop and the visualization chosen automatically according to the data type; (c) Interactive Pivot Table showing the exact numbers for each displayed variable combination.
 The data displayed is used to analyze the lumbar spine.
@@
 %
-Suitability of an interaction- and visualization technique for epidemiological data depends on its ability to intuitively compare multiple data features at once while highlighting new interesting associations.
+The suitability of an interaction- and visualization technique for epidemiological data depends on its ability to intuitively compare multiple data features at once while highlighting new associations.
 %
@@
 %
-Visual evaluation of data is therefore as important as methods allowing for numerical data analysis.
+Visual evaluation of data are therefore as important as methods allowing for numerical data analysis.
 %
@@
 	\item The sidebar, which contains all epidemiological features. Cluster results are treated like features and are part of the sidebar as well.
-	\item The canvas which holds all visualizations. Elements can be added, arranged, resized and removed freely.
+	\item The canvas holding all visualizations. Elements can be added, arranged, resized and removed freely.
 	\item The interactive Pivot Table gives detailed numerical information of the features in the canvas view. This view on the data come natural to epidemiologists.
@@
 % %
-% Shape based clustering can either be applied to all subjects or subgroups.
+% Shape-based clustering can either be applied to all subjects or subgroups.
 % 
@@
 \paragraph{Sidebar.}
-An overview of all features is presented in a sidebar where they are categorized to different types such as somatometric, disease- or lifestyle related, pain indicators and laboratory data.
+An overview of all features is presented in a sidebar where they are categorized to different types such as somatometric (measurements of the dimensions of the human body), disease- or lifestyle related, pain indicators and laboratory data.
 %
@@
 %
-Groups are treated like features since they work exactly the same way by dividing the subject space into labeled categories.
+Groups are treated like features since they work also by dividing the subject space into labeled categories.
 %
@@
 %
+Categorical data are mapped either using bar charts (single features) or mosaic plots (multiple features).
+%
+Continuous data are discretized and therefore treated as categorical.
+%
+It is, however, also possible to plot it using scatterplots (single features) or parallel coordinates (multiple features).
+%
 Figure \ref{fig:visualization} describes this dynamic adjustment.
@@
 % TODO Implement MOSAIC PLOTS, Scatterplots, Parallel Coordinates
-Following Tuftes concept of \emph{small multiples} \cite{Tufte1983} the medical image data is directly included into the plot by including color coded mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
+Following Tuftes concept of \emph{small multiples} \cite{Tufte1983} the medical image data are directly included into the plot by including color coded mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
 %If possible the medical image data is directly included into the plot as well by including mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
@@
 %
-Epidemiologists are used to process groups based on table representations so we decided to introduce an interactive pivot table.
+Epidemiologists are used to process groups based on table representations.
+%
+Thus we decided to introduce an interactive pivot table.
 %When opening a random epidemiological paper the reader will in almost all cases find some sort of pivot table to present the data.
 %
-As seen in Figure~\ref{fig:visualization} (c) it is a good way to display how many subjects are in each group.
-%
-Pivot tables quickly get confusing and cluttered when they are divided into many subgroups.
+%As seen in Figure~\ref{fig:visualization} (c) it is a good way to display how many subjects are in each group.
+Pivot tables clearly convey how many subjects are in each group (see Figure~\ref{fig:visualization} (c))
+%
+However, these quickly get confusing and cluttered when they are divided into many subgroups.
 %
@@
 %
-This also applies on the designation of row or column variables.
+This also applies to the assignment of row or column variables.
 %
@@
 %
-Turkay and colleagues used the approach to calculate various key figures based in the distribution functions of each feature derived from the image data \cite{Turkay2013}.
+Turkay and colleagues used the approach to calculate various key figures based on the distribution functions of each feature derived from the image data \cite{Turkay2013}.
 %
@@
 
-It shares the same restrictions as \emph{Pearson's $X^2$} which include next to others expected count larger than 5 for $80\%$ of the contingency table entry and no expected value can be smaller than one.
+%It shares the same restrictions as \emph{Pearson's $X^2$} which include next to others expected count larger than 5 for $80\%$ of the contingency table entry and no expected value can be smaller than one.
+It shares the same restrictions as \emph{Pearson's $X^2$}.
+%
+The expected counts in the contingency table have to be larger than 5 for $80\%$ of the entries and no expected value must be smaller than one!
 %
@@
 %
-This aims to highlight features which may interact with the focused hypothesis as well as trigger new hypotheses.
+This aims to highlight features possibly interacting with the focused hypothesis as well as trigger new hypotheses.
 %
@@
 \\\\
-In the following Section we discuss how we implemented the presented methods using free on open Web standards.
+In the following section we discuss how we implemented the presented methods using free on open Web standards.
 % \subsubsection{Automated Feature Suggestion}
@@
 %
-To provide a fast communication loop between method development and expert input, we decided to rely on modern web technologies which benefits from various advantages:
+To provide a fast communication loop between method development and expert input, we decided to rely on modern web technologies, which benefits from various advantages:
 \begin{itemize}
@@
 	\item The client-server structure allows for employing heavy computation on a server machine and transfer results to the client.
-	\item Since image data for several thousand subjects claims hundreds of gigabytes disk space it can remain safely on the server and elements can be transferred on demand.
+	\item Since image data for several thousand subjects requires hundreds of gigabytes disk space it can remain safely on the server and elements can be transferred on demand.
 	%
 	High confidentiality standards of the data can be met by restricting access via a account system
-	\item Recent developments in WebGL applications running in browsers with near native performance push the development into to the web, which results in many open source libraries which are well documented, rich in examples and driven by active communities.
+	\item Recent developments in WebGL applications running in browsers with near native performance push the development into to the web resulting in many open source libraries, which are well documented, rich in examples and driven by active communities.
 We use WebGL for rendering shape information.
@@
 %
-Image data is available as raw DICOM files as well as segmented meshes, which can be used to compare subjects.
+Image data are available as raw DICOM files as well as meshes representing segmentation masks, which can be used to compare subjects.
 %
@@
 %
-\texttt{WebGL} rendering is done using the \texttt{Three.js}\footnote{Originally developed by Ricardo Cabello, \url{threejs.org}} and allows GPU Accelerated data rendering.
+\texttt{WebGL} rendering is done using the \texttt{Three.js}\footnote{Originally developed by Ricardo Cabello, \url{threejs.org}} and allows GPU-accelerated data rendering.
 %
@@
 \section{Application}
-We applied the presented set of techniques to a data set which is compiled to analyze lower back pain. 
+We applied the presented set of techniques to a data set compiled to analyze lower back pain. 
 %
@@
 %
-If at all, only a few shape related features are included in comparable studies, for example by Lang and colleagues \cite{Lang2011}.
-%
-To our knowledge, this is the first approach on analyzing shape related information of the whole lumbar spine with other epidemiological features.
+If at all, only a few shape-related features are included in comparable studies, for example by Lang and colleagues \cite{Lang2011}.
+%
+To our knowledge, this is the first approach, where shape-related information of the whole lumbar spine is analyzed along with other epidemiological features.
 %
@@
 %
-Characterizing the healthy aging process of the spine is a large stretch goal that allows to determine age-normalized probabilities for spine-related diseases by incorporating individual risk factors.
+Characterizing the healthy aging process of the spine is a long term goal that allows to determine age-normalized probabilities for spine-related diseases by incorporating individual risk factors.
 
@@
 %
-Our epidemiological collaborators compiled a feature list which is a tradeoff between complexity and limitations of the responsible ethics committee.
+Our epidemiological collaborators compiled a feature list, which is a tradeoff between complexity and limitations of the responsible ethics committee.
 %
@@
 %
-There are 136 features describing diagnosed diseases, lifestyle factors, women specific factors, pain indicators, laboratory values, somatometric variables and are ordered accordingly.
+There are 136 features describing diagnosed diseases, lifestyle factors, women-specific factors, pain indicators, laboratory values, somatometric variables and are ordered accordingly.
 %
@@
 %
-The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence (676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
+%The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence (676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
+The spine protocol consisted of sagittal T1- and T2-weighted turbo-spin-echo sequences with rather large slice distance only just sufficient for characterizing the lumbar spine \cite{Hegenscheid2013}.
 
@@
 
-To ensure fast and easy data access outside of statistical processors like \texttt{SPSS} or \texttt{STATA}, the data was exported to the JSON file format which can easily be parsed by modern programming languages.
-%
-Each feature is stored as an object which contains: 
+To ensure fast and easy data access outside of statistical processors like \texttt{SPSS}, the data was exported to the JSON file format, that can easily be parsed by modern programming languages.
+%
+Each feature is stored as an object containging: 
 \begin{itemize}
-	\item the data as array of values--categorical values and error codes are stored using IDs
-	\item the data type (continuous, nominal, ordinal, dichotomous)
-	\item a detailed description of the variable
-	\item the data dictionary which translates value- or error IDs to the actual values
+	\item the data as array of values--categorical values and error codes are stored using IDs,
+	\item the data type (continuous, nominal, ordinal, dichotomous),
+	\item a detailed description of the variable, and
+	\item the data dictionary translating value- or error IDs to the actual values.
 \end{itemize}
@@
 %
-In epidemiology, continuous data is usually categorized into ordinal groups of equal size.
+In epidemiology, continuous data are usually categorized into ordinal groups of equal size.
 %
@@
 %
-To allow for hypothesis generation we set the number of groups to five if not specified otherwise.
-
-\paragraph{Image-Data.} \label{Image-Data}
+To allow for hypothesis generation, we set the number of groups to five if not specified otherwise.
+
+\paragraph{Image Data.} \label{Image-Data}
 The lumbar spine was detected in the image data using a hierarchical finite element method according to Rak and colleagues \cite{Rak2013}.
@@
 %
-Two user defined landmarks on the top and bottom of the L3-vertebra are used to obtain an initial height estimation of the model.
+Two user-defined landmarks on the top and bottom of the L3-vertebra are used to obtain an initial height estimation of the model.
 %Two more user- defined landmarks are the top and bottom of the L3-vertebra, which transforms size of the model accordingly to match the L3-vertebra .
 %
-It uses a weighted sum of T1- and T2-weighted MRI images to detect the lumbar spine shape.
+It uses a weighted sum of T1- and T2-weighted MR images to detect the lumbar spine shape.
 %
@@
 %
-We obtained and work with 983 models.
-%
-For clustering purposes we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis which are medical terms for spine curvature \cite{Klemm2013VMV}.
+We obtained and work with 983 tetrahedron models of the lumbar spine.
+%
+For clustering purposes we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (medical terms for spine curvature \cite{Klemm2013VMV}).
 
@@
 \\\\
-Shape-based clustering is carried out via Agglomerative Hierarchical clustering of the spine canal, centerlines which are described in Section~\ref{application:Data Preprocessing} \cite{Klemm2013VMV}.
+Shape-based clustering is carried out via Agglomerative Hierarchical clustering of the spine canal centerlines, (recall Section~\ref{application:Data Preprocessing} and \cite{Klemm2013VMV}).
 %
@@
 %
-For details, see \cite{Klemm2013VMV}.
+%For details, see \cite{Klemm2013VMV}.
 %
@@
 %
-Analyzing the data set without a prior hypothesis requires a starting point which gives a overview over the data first \cite{Shneiderman1996}.
-%
-%Shape-based hypothesis free exploration starts with a shape-grouping step achieved by shape-based clustering.
-Shape-based hypothesis free exploration starts with the \emph{local investigation} pattern by a shape-grouping step using shape-based clustering.
+Analyzing the data set without a prior hypothesis requires a starting point, giving an overview over the data first \cite{Shneiderman1996}.
+%
+%Shape-based hypothesis-free exploration starts with a shape-grouping step achieved by shape-based clustering.
+Shape-based hypothesis-free exploration starts with the \emph{local investigation} pattern by a shape-grouping step using shape-based clustering.
 %
@@
 %
-Cluster 8 was of special interest because of its large distance to the mean shape while still exposing the second highest subject count.
+Cluster 8 has the 2nd largest number of elements and was therefore of special interest.
+%Cluster 8 was of special interest because of its large distance to the mean shape while still exposing the second highest subject count.
 %
@@
 %
-Metabolic parameters such as blood-fat and blood-sugar are also possibly associated features.
-%
-Employment status may be a confounder describing the lifestyle of a subject.
+Metabolic parameters, such as blood-fat and blood-sugar, are also possibly associated features.
+%
+Employment status is a feature, which relates to many different lifestyle factors or age and might therefore act as confounder.
 %TODO Implement Color coding only in a certain direction!\\
@@
 \\\\
-While this approach does not assume any hypothesis, the data is met with a selection bias when compiling the list of related features by the domain experts.
-%
-It is arguable if this first filtering step, which is purely based on expert experience, is really a disadvantage, since it rules out many parameters which may interact with the presented features but the value of this knowledge would be small \cite{Wiley2008}.
+While this approach does not assume any hypothesis, the 136 epidemiological features were selected by domain experts.
+%the data are met with a selection bias when compiling the list of related features by the domain experts.
+%
+It is arguable whether this first filtering step, which is purely based on expert experience, represents a disadvantage, since it rules out many parameters possibly interacting with the presented features but the value of this knowledge would be small \cite{Wiley2008}.
+% While this approach does not assume any hypothesis, the data are met with a selection bias when compiling the list of related features by the domain experts.
+% %
+% It is arguable if this first filtering step, which is purely based on expert experience, is really a disadvantage, since it rules out many parameters which may interact with the presented features but the value of this knowledge would be small \cite{Wiley2008}.
 
@@
 %
-In our use case, the epidemiologist was interested in the dichotomous questionnaire answer to "Did you experience back pain in the past three month".
+In our use case, epidemiologists were interested in the dichotomous questionnaire answer to "Did you experience back pain in the past three month".
 %
@@
 %
-Since the focus is on subjects which suffer from back pain, the clustering result of these subjects are then drawn into the canvas area, yielding six cluster as seen in Figure~\ref{fig:hypopthesisbased} (b).
-%
-Cluster 5 stood out for having a so-called hyperlordosis, a strong curvature of the lumbar spine which is a indicator for back pain.
+Since the focus is on subjects suffering from back pain, the clustering result of these subjects are then drawn into the canvas area, yielding six cluster as seen in Figure~\ref{fig:hypopthesisbased} (b).
+%
+Cluster 5 stood out for having a so-called hyperlordosis, a strong curvature of the lumbar spine associated with back pain.
 
@@
 %
-It indicates that waist circumference rather than the BMI interacts with unusual shaped spines for subjects with lumbar back pain.
+It indicates that waist circumference rather than the BMI interacts with unusually shaped spines for subjects with lumbar back pain.
 %
@@
 %
-Defining a causal relationship solely based on observed correlations of two features is \emph{cum hoc ergo propter hoc}--correlation does not automatically imply causation \cite{Tufte2003}.
-%
-The observed correlations need to be carefully checked for confounder and medical soundness!
+Correlation does not automatically imply causation \cite{Tufte2003}.
+%Defining a causal relationship solely based on observed correlations of two features is \emph{cum hoc ergo propter hoc}--correlation does not automatically imply causation \cite{Tufte2003}.
+%
+The observed correlations need to be carefully checked for confounders and medical soundness!
 %
@@
 
-Features that potentially interact with a disease related condition need to be validated.
+Features that potentially interact with a disease-related condition need to be validated.
 %
@@
 %
-The presented methods guide attention to features which where not in the focus of attention and expectance of clinical researchers.
-%
-The explorative nature of the methods work well for gathering associations which may act as confounder, as outcome of a disease or as an actual cause or risk factor.
+The presented methods guide attention to features not in the focus of attention and expectance of clinical researchers.
+%
+The explorative nature of the methods work well for gathering associations possibly acting as confounder, as outcome of a disease or as an actual cause or risk factor.
 %"\emph{Correlation does not imply causation}" is a famous saying in statistics, meaning that correlation does not necessarily imply causal relationships \cite{}.
@@
 %
-
 %On the feature level, displaying MRI scans is interesting for the subjects in the outlier clusters because they are highly likely to exhibit pathologies.
@@
 %Seen as an extension to the existing statistics driven workflow, image-centric analysis of epidemiological data from a IVA approach by illuminating the data-black box.
-The presented are classified as extension to the existing statistics driven epidemiological.
+The presented \emph{IVA} approach is classified as extension to the existing statistics driven epidemiological.
 %
@@
 %
-The use of pivot tables is familiar to epidemiologist while embracing the power of interactive adjustment of the features shown.
+The use of pivot tables is familiar to epidemiologists while embracing the power of interactive adjustment of the features shown.
 %
@@
 \\\\\
-Our future work comprises a number of improvements:
+A number of improvements is left open for future work:
 \begin{itemize}
 	\item dynamic discretization of continuous variables to better fit the data distributions
-	\item shape brushing methods to intuitively query subjects using image-data
+	\item shape brushing methods to intuitively query subjects using image data
 	\item incorporate more statistical methods and views that are familiar to the epidemiologists (odds ratios, box plots)
 	\item calculation of contingency differences for selected sub groups--highlight how feature correlation change for a given groups
-	\item more visualization techniques which incorporate both image and non-image data
+	\item integration of further visualization techniques, which incorporate both image and non-image data
 	\item support control-groups to check resilience of observations as suggested by Fletcher and Fletcher \cite{Fletcher2012}.
@@
 % - UI more flexible by hiding UI-panes
-% - More Shape based filtering - reorder by size, curvature, etc. ...
+% - More Shape-based filtering - reorder by size, curvature, etc. ...
 % - Color code different Aspects - Only difference in x or y direction

### 2014-03-29 draft.tex
Added two reworked figures

@@
 \end{figure}
-Our proposed \emph{IVA} workflow knows three major steps as illustrated in Figure~\ref{fig:InteractionLoop}: Feature selection, -visualization and -brushing.
-%
-An hypothesis-driven analysis usually starts with the selection of features.
-%
-A feature selection can also be derived from a shape-based clustering, which creates shape groups.
-%
-\emph{Hypothesis-generation} with focus on image data starts with a shape-based clustering.
-%
-The feature is mapped using a automatically chosen visualization appropriate for its data type (described in detail in the following section).
-%
-The visualization techniques have to combine both image-and non-image data in order to set domain and range data in relation to each other.
-%
-Subsequently the visualization can be either brushed or new variables can be added to the analysis.
-%
-Brushing methods are subdivided using the previously described \emph{IVA} patterns.
-%
-Brushed regions are treated like features as they divide the subject space just like categorical features. %and lead therefore to a new feature selection.
-%
-Selecting features also triggers a \emph{multivariate analysis} using contingency values (described in the following section) to highlight variables interacting with selected features.
-%
-%The starting point of \emph{IVA Loop} is determined whether the analysis is hypothesis-driven or exploratory.
-%
-%An analysis with the aim of hypothesis generation of image-based data starts with a clustering of the structure 
-%
-A sample workflow using interaction- and visualization techniques described in the next section can be seen in Figure~\ref{fig:visualization}.
-
-\subsection{Interaction- and Visualization Techniques} \label{Interaction- and Visualization Techniques}
 %-------------------------------- Figure --------------------------------
-% TODO: VISUALIZATION Figure sollte umgebaut werden, sodass sie die Schritte bei der Analyse hervorhebt
 \begin{figure*}[htb]
@@
 %-------------------------------- /Figure -------------------------------
+Our proposed \emph{IVA} workflow knows three major steps as illustrated in Figure~\ref{fig:InteractionLoop}: Feature selection, -visualization and -brushing.
+%
+An hypothesis-driven analysis usually starts with the selection of features.
+%
+A feature selection can also be derived from a shape-based clustering, which creates shape groups.
+%
+\emph{Hypothesis-generation} with focus on image data starts with a shape-based clustering.
+%
+The feature is mapped using a automatically chosen visualization appropriate for its data type (described in detail in the following section).
+%
+The visualization techniques have to combine both image-and non-image data in order to set domain and range data in relation to each other.
+%
+Subsequently the visualization can be either brushed or new variables can be added to the analysis.
+%
+Brushing methods are subdivided using the previously described \emph{IVA} patterns.
+%
+Brushed regions are treated like features as they divide the subject space just like categorical features. %and lead therefore to a new feature selection.
+%
+Selecting features also triggers a \emph{multivariate analysis} using contingency values (described in the following section) to highlight variables interacting with selected features.
+%
+%The starting point of \emph{IVA Loop} is determined whether the analysis is hypothesis-driven or exploratory.
+%
+%An analysis with the aim of hypothesis generation of image-based data starts with a clustering of the structure 
+%
+A sample workflow using interaction- and visualization techniques described in the next section can be seen in Figure~\ref{fig:visualization}.
+
+\subsection{Interaction- and Visualization Techniques} \label{Interaction- and Visualization Techniques}
 %We employ the fourth and highest level which includes next to brushing and linking, advanced brushing, extraction of attributes the development of visualizations custom tailored to the data sets.
@@
  \includegraphics[width=3.5in]{figures/technologies}
- \caption{\emph{Image is not final}. The front-end solution (left) uses state of the art web technologies such as HTML5/CSS3, WebGL and SVG to display the data.
+ \caption{The front-end solution (left) uses state of the art web technologies such as HTML5/CSS3, WebGL and SVG to display the data.
  % 
@@
  %
- Client-Server communication is accomplished via the WebSocket protocol.
+ Client-Server communication is accomplished via the Websocket protocol.
  }

### 2014-03-30 draft.tex
added changes from Kai

@@
 %\documentclass[electronic,journal]{style/vgtc}     % electronic version, journal
-% TODO: Reference Nicolas Kruchten
 %% Uncomment one of the lines above depending on where your paper is
@@
 %
-Semi-automatic techniques are more promising but challenging since the used modalities as Magnetic resonance imaging (MRI) and ultrasound are subject to inhomogeneity and noise.
+Semi-automatic techniques are more promising but challenging since the employed modalities, such as Magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
 %
@@
 %
-This also applies to the chosen landmarks which are used to quantify medical image data information.
+This also applies to the chosen landmarks used to quantify medical image data information.
 %
@@
 %
-% TODO Define Epidemiological Outcome - 5Ds:
+%Epidemiological Outcome - 5Ds:
 % Death
@@
 (a) shows the standard epidemiological workflow, (b) the IVA supported one.
-The iterative red highlighted part is called the \emph{IVA Loop}.}
+The iterative red highlighted part is called the \emph{IVA Loop} and is described in more detail in Figure~\ref{fig:InteractionLoop}.}
   \label{fig:WorkflowComparison}
@@
 %
-	\item A set of attributes depicting conditions affected by the hypothesis is compiled accordingly.
-% TODO: Bernhard: Effect size of Attribute ?
-	\item Confounding features are adjusted so that they do not influence the effect size of a attribute.
+	\item A set of features depicting conditions affected by the hypothesis is compiled accordingly.
+	%\item Confounding features are adjusted so that they do not influence the effect size of an attribute.
+	\item Confounding features are identified and taken into account, for example using stratification % adjusted so that they do not influence the effect size of a attribute.
 %
@@
 %
-Longitudinal studies require the acquired attributes to be comparable to evaluate them.
+Longitudinal studies require the acquired attributes to be comparable for evaluation.
 %
@@
 %
-For ethical, legal or medical reasons some features can not be gathered for each subject.
+For ethical, legal or medical reasons some features cannot be gathered for each subject.
 %
@@
 
-\paragraph{Image acquisition.} \label{ImageAcquisition} Imaging techniques involving ionizing radiation for the subject are not suited for ethical reasons.
+\paragraph{Image acquisition.} \label{ImageAcquisition} Imaging techniques involving ionizing radiation for the subject are not suitable for ethical reasons.
 %
@@
 %
-Segmentation masks labeling the voxels of an anatomical structure would be ideal since many different key figures, e.g. volume, largest diameter or aspect ratio, can be derived from them.
+Segmentation masks labeling the voxels of an anatomical structure would be ideal since many different key figures, e.g., volume, largest diameter or aspect ratio, can be derived from them.
 %
@@
 %
-The Principal Component Analysis (PCA) and similar techniques are able to reduce the dimension by extracting most expressive components, but make the influence of each variable hard to convey.
+The principal component analysis (PCA) and similar techniques are able to reduce the dimension by extracting most expressive components, but make the influence of each variable hard to convey.
 \\\\
@@
 %
-They incorporated data mining methods such as dividing the data space using a k-nearest-neighbor technique and the PCA.
+They incorporated data mining methods such as dividing the data space by using a k-nearest-neighbor technique and the PCA.
 %
@@
 %
-Dai and colleagues explored risk factors by incorporating choropleth maps of epidemiological features (e.g.
-mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines \cite{Dai2005}.
+Dai and colleagues explored risk factors by incorporating choropleth maps of epidemiological features (e.g., mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines \cite{Dai2005}.
 %A similar approach was taken by Dai and colleagues for risk factor exploration as they also incorporate choropleth maps of epidemiological factors (e.g.
@@
 %
-The distribution of shapes in a space derived from a PCA is plotted by Busking and Colleagues in a 2D-projected plane of the space \cite{Busking2010a}.
+The distribution of shapes in a space derived from a PCA is plotted by Busking and colleagues in a 2D-projected plane of the space \cite{Busking2010a}.
 %
@@
 %
-Derived from this we incorporate the idea of combining 3D-Shape rendering with information visualization techniques.
+Derived from this we incorporate the idea of combining 3D-shape rendering with information visualization techniques.
 %
@@
 %
-As proof of concept a relation between size of the segmented shape and measured size of the subjects was shown.
+As proof of concept, a relation between size of the segmented shape and measured size of the subjects was shown.
 %
@@
 %
-When applying clustering techniques on the non-image data it was found that \texttt{k-Prototypes} and \texttt{DBSCAN} is appropriate in the epidemiological context but is strongly dependent on the chosen variables and distance measure \cite{Klemm2014BVM}.
+When applying clustering techniques on the non-image data it was found that \texttt{k-Prototypes} and \texttt{DBSCAN} is appropriate in the epidemiological context, but is strongly dependent on the chosen variables and distance measure \cite{Klemm2014BVM}.
 %
@@
 %
-If a segmentation yields only binary masks separating the structures, algorithms such as Growing and Adaptive Shapes can be applied creating a surface grid where each point is comparable throughout the population \cite{Ferrarini2007}.
+If a segmentation yields only binary masks separating the structures, algorithms such as \emph{Growing and Adaptive Shapes} can be applied creating a surface grid where each point is comparable throughout the population \cite{Ferrarini2007}.
 %
@@
  % TODO: Recreate Image: Also include Cramér´s V which is part of the visualization
- \caption{Detailed Version of the \emph{IVA Loop}.
+ \caption{Detailed Version of the \emph{IVA Loop} seen in Figure~\ref{fig:WorkflowComparison}.
 Usually starting with a selection of Feature of interest (user-driven or via data mining techniques), the data are mapped using a visualization technique appropriate for the selected data types.
@@
  \includegraphics[width=1\textwidth, resolution=300]{figures/visualization}
- \caption{(Left) Screenshot from the front-end, which is divided as follows: (a) Sidebar containing all features as well as the groups defined in the analysis process; (b) Canvas area where features can be added via drag and drop and the visualization chosen automatically according to the data type; (c) Interactive Pivot Table showing the exact numbers for each displayed variable combination.
+ \caption{(Left) Screenshot from the front-end, which is divided as follows: (a) Sidebar containing all features as well as the groups defined in the analysis process; (b) Canvas area where features can be added via drag and drop and the visualization chosen automatically according to the data type; (c) Interactive pivot table showing the exact numbers for each displayed variable combination.
 The data displayed is used to analyze the lumbar spine.
@@
 % TODO Einbauen
+% Overfitting bei SHIP-Trend
 % Hypothesis generation bears the chance of \emph{overfitting} the data to expectations.
@@
 	\item The canvas holding all visualizations. Elements can be added, arranged, resized and removed freely.
-	\item The interactive Pivot Table gives detailed numerical information of the features in the canvas view. This view on the data come natural to epidemiologists.
+	\item The interactive pivot table gives detailed numerical information of the features in the canvas view. This view on the data come natural to epidemiologists.
 	\item The contingency view depicting relations for features in the canvas in an adjacency matrix.
@@
  \includegraphics[width=3.5in]{figures/technologies}
- \caption{The front-end solution (left) uses state of the art web technologies such as HTML5/CSS3, WebGL and SVG to display the data.
- % 
- The NodeJS based back-end (right) stores all image- and non-image data and transfers it to connected clients.
+ \caption{The front-end solution (left) uses state of the art web technologies such as \emph{HTML5}/\emph{CSS3}, \emph{WebGL} and \emph{SVG} to display the data.
+ % 
+ The \emph{NodeJS} based back-end (right) stores all image- and non-image data and transfers it to connected clients.
  %
@@
  %
- Client-Server communication is accomplished via the Websocket protocol.
+ Client-Server communication is accomplished via the \emph{Websocket} protocol.
  }
@@
 %
+The pivot table implementation was adapted using \texttt{PivotTable.js}\footnote{Developed by Nicolas Kruchten, \url{http://nicolas.kruchten.com/pivottable}}.
+%
 \texttt{WebGL} rendering is done using the \texttt{Three.js}\footnote{Originally developed by Ricardo Cabello, \url{threejs.org}} and allows GPU-accelerated data rendering.
@@
 %
-If at all, only a few shape-related features are included in comparable studies, for example by Lang and colleagues \cite{Lang2011}.
+%If at all, only a few shape-related features are included in comparable studies, for example by Lang and colleagues \cite{Lang2011}.
+In comparable studies, only a few shape-related features are included, for example by Lang and colleagues \cite{Lang2011}.
 %
@@
 %
-There are 136 features describing diagnosed diseases, lifestyle factors, women-specific factors, pain indicators, laboratory values, somatometric variables and are ordered accordingly.
+There are 136 features describing diagnosed diseases, lifestyle factors, women-specific factors, pain indicators, laboratory values, somatometric variables. %and are ordered accordingly.
 %
@@
 
-To ensure fast and easy data access outside of statistical processors like \texttt{SPSS}, the data was exported to the JSON file format, that can easily be parsed by modern programming languages.
+To ensure a fast and easy data access outside of statistical processors like \texttt{SPSS}, the data was exported to the JSON file format, that can easily be parsed by modern programming languages.
 %
@@
 \begin{itemize}
-	\item the data as array of values--categorical values and error codes are stored using IDs,
+	\item the data as array of values--categorical data and error codes are stored using IDs,
 	\item the data type (continuous, nominal, ordinal, dichotomous),
@@
 To allow for hypothesis generation, we set the number of groups to five if not specified otherwise.
+% TODO Kai: Das mussn noch geklärt werden
 
@@
  \includegraphics[width=3.5in]{figures/hypothesisfree}
- \caption{\emph{Image is not final}.
-(a) Clustering result of all subjects.
+ \caption{(a) Clustering result of all subjects.
 The bar chart height indicates the number of subjects in the cluster.
-The difference to the mean-shape is color-coded where blue represents no difference and red a large difference.}
+The difference to the mean-shape is color-coded where blue represents no difference and red a large difference.
+(b) A pivot table relating cluster results to gender and employment status.
+Cluster 8 is striking because of the high proportion of women while Cluster 2, 3, 4 contain almost only men.}
  \label{fig:hypothesisfree}
@@
 %
-Other shapes differ with respect to size, such as Cluster 2, 8, 10 and 3 where the last one also is more straight, which is usual for subjects with larger body height.
-%
-Cluster 4, 7 and 5 contain outliers, characterized by their unusual shape and small number.
+Other shapes differ with respect to size, such as Cluster 2, 8, 10, 3 where the last one also represents a more more straight spine, which is usual for subjects with larger body size.
+%
+Cluster 4, 7, 5 contain outliers, characterized by their unusual shape and small number.
 %
@@
 %
-Performing a \emph{multivariate analysis} by looking at the \emph{Cram\'{e}rs' V} contingency values of Cluster~8 reveals associations of this group with employment status, body size, age, thyroid nodules and blood-fat value.
+Performing a \emph{multivariate analysis} by looking at the \emph{Cram\'{e}rs' V} contingency values of Cluster~8 reveals associations of this group with employment status, gender, body size, age, thyroid nodules and blood-fat value.
+%
+Another \emph{multivariate analysis} using a pivot table set gender and employment status in relation to clustering affiliation, showing that Cluster~8 contains mostly women and also has a larger unemployment (see Figure~\ref{fig:hypothesisfree}), while the overall employment rate of women and men in the data set is almost exactly 50\%.
 %
@@
 %
-If the user has already a hypothesis about a relation between a non-image feature regarding shape the workflow slightly differs from the hypothesis-free analysis.
+If the user proposed a hypothesis about a relation between a non-image feature regarding shape the workflow slightly differs from the hypothesis-free analysis.
 %
@@
 %
-While the Body Mass Index (BMI) is a key figure for assessing height and weight of a subject, it does not tell us anything on how the weight is distributed in the body.
+While the body mass index (BMI) is a key figure for assessing height and weight of a subject, it does not tell anything on how the weight is distributed in the body.
 %
@@
 %
-Our finding follows the recent trends that indicate that BMI is not a good measure for assessing body-shape since healthy weight is dependent on many other measures \cite{Ahima2013}.
+Our finding follows the recent trends indicating that BMI is not a good measure for assessing body-shape since healthy weight is dependent on many other measures \cite{Ahima2013}.
 %
@@
 %
-To increase the probability of the observation not to be random, they are cross check for associations in \texttt{SHIP-TREND} as a second, independent population sample, which was examined with methods identical to SHIP.
+To increase the probability of the observation not to be random and to avoid \emph{overfitting}, they are cross check for associations in \texttt{SHIP-TREND} as a second, independent population sample, which was examined with methods identical to SHIP.
 %If features are identified as interesting for a disease, the epidemiologists look for associations in another cohort, the \texttt{SHIP-TREND}.

### 2014-03-30 draft.tex
Proofread the Paper

@@
 %% hyperreferences for more convenient online viewing.
-
+% TODO: Incomplete data: Cramers V kann nur für bestimmte mindestgrößen funktionieren - entweder Daten unvollständig oderz u wneige Probanden in einer Gruppe
 %% Please use one of the ``review'' options in combination with the
@@
 %
-The data are analyzed using a priori defined hypotheses to find statistically significant correlations between variables (\emph{associations}).
-%
-Modern cohort studies also incorporate medical image data.
+The data are analyzed using a priori defined hypotheses to find statistically significant correlations between features (\emph{associations}).
+%
+Modern cohort studies incorporate medical image data.
 %
@@
 %
-The statistically driven epidemiological workflow only allows to determine \emph{associations} between image-derived metrics such as distances extracted from landmarks of the segmentation model.
+The statistically driven epidemiological workflow only allows to determine \emph{associations} between image-derived metrics, such as distances extracted from landmarks of the segmentation model.
 \\\\
 %(New Solution).
-We propose an Interactive Visual Analysis (\emph{IVA}) approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic variables and attributes derived from the image data.
+We propose an Interactive Visual Analysis (\emph{IVA}) approach, that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic features and attributes derived from the image data.
 %
@@
 %(Results). 
-It shows to be a more flexible comparison between image and non-image data.
+It allows a more flexible comparison between image and non-image data.
 %
 %(Implications).
-The new framework allows for hypothesis validation and hypothesis generation by incorporating human pattern recognition as well as data mining methods.
-%
-Using all reliable information from the image segmentation linked to non-image variables aims to unveil \emph{associations} by applying an iterative analysis approach.
+The new framework enables hypotheses validation and -generation by incorporating human pattern recognition as well as data mining methods.
+%
+Using all reliable information from the image segmentation linked to non-image features aims to unveil \emph{associations} by applying an iterative analysis approach.
 
@@
 %
-On the other hand, observations made by clinicians in the daily routine are translated into hypothesis for epidemiological research.
-%
-These are used to determine environmental and lifestyle factors as well as medical examinations results, which may influence a condition of interest.
-%
-The data variables, that are potentially useful, are gathered using structured interviews and clinical examinations.
+On the other hand, observations made by clinicians in the daily routine are translated into hypotheses for epidemiological research.
+%
+These are used to determine environmental and lifestyle factors as well as medical examinations results, which may influence a disease.
+%
+Potentially useful data variables (\emph{features}) are gathered using structured interviews and clinical examinations.
 %
@@
 %These data needs to be segmented to analyze its features for correlations with non-image data.
-The segmentation of these data would allow to analyze shapes of anatomical structures and to correlate them with other variables.
+Segmentation of these data would allow for anatomical structure analysis and enable correlation with other features.
+%The segmentation of these data would allow to analyze shapes of anatomical structures and to correlate them with other features.
 %These data needs to be concurrently analyzed with the non-image data.
@@
 %
-Semi-automatic techniques are more promising but challenging since the employed modalities, such as Magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
-%
-Analyzing spatial data with respect to other epidemiological factors requires techniques which reach beyond standard statistical methods.
+Semi-automatic techniques are more promising but also challenging since the employed modalities, such as Magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
+%
+Analyzing spatial data with respect to other epidemiological factors requires techniques, which reach beyond standard statistical methods.
 
@@
 %
-Intended as an extension to the well established epidemiological tools it provides a way to rapidly validate hypothesis as well as trigger \emph{hypothesis generation} using data mining methods such as clustering.
-%
-Hypothesis generation gains importance since the number of epidemiological variables increases and the focus shifts towards more complex relations involving more than two variables.
-%
-Easy publishing of developed methods driven by modern web technologies intends to trigger a fast feedback loop between us and the epidemiologists.
+Intended as an extension to the well established epidemiological tools it provides a way to rapidly validate hypotheses as well as trigger \emph{hypothesis generation} using data mining methods, such as clustering.
+%
+\emph{Hypothesis generation} gains importance since the number of epidemiological features increases and the focus shifts towards more complex relations involving more than two features.
+%
+Easy exchange of developed methods driven by modern web technologies intends to trigger a fast feedback loop between us and the epidemiologists.
 % http://wiki.infowiss.net/Gulf_of_execution ?
@@
 %
-This data set comprises 127 variables and 2 sequences of MRI data from 2.333 patients.
+This data set comprises 127 features and 2 sequences of MRI data from 2.333 patients.
 \\\\
@@
 %	\item providing an overview over the workflow for analyzing cohort study data to gain insight into the large diverse subject spaces.
-	\item an Interactive Visual Analysis workflow for analyzing image-based epidemiological data including both hypothesis-driven and for hypothesis-generation based on a characterization of the standard epidemiological workflow
-	\item visualization techniques, which combine both information visualization and 3D rendering of organ shapes as well as combining them with epidemiological graphics and key figures.
-	\item highlighting interesting subject groups and feature associations using shape-based clustering and statistical contingency measures
+	\item an Interactive Visual Analysis workflow for analyzing image-based epidemiological data both hypothesis-driven and for hypothesis-generation based on a characterization of the standard epidemiological workflow,
+	%\item an Interactive Visual Analysis workflow for analyzing image-based epidemiological data including both hypothesis-driven and for hypothesis-generation based on a characterization of the standard epidemiological workflow
+	\item visualization techniques, which incorporate both information visualization and 3D rendering of organ shapes as well as combining them with epidemiological graphics and key figures,
+	\item highlighting interesting subject groups and feature associations using shape-based clustering and statistical contingency measures,
 	%Applying Interactive Visual Analysis toolset to image-based analysis in epidemiology with focus hypothesis generation
@@
 The appropriate combination of statistical- and interactive driven analysis shows promising potential to unveil the information in the data.
-(a) shows the standard epidemiological workflow, (b) the IVA supported one.
+(a) shows the standard epidemiological workflow, (b) the \emph{IVA} supported one.
 The iterative red highlighted part is called the \emph{IVA Loop} and is described in more detail in Figure~\ref{fig:InteractionLoop}.}
@@
 %Epidemiology diversity is reflected in the different experts working at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
-The diversity of epidemiology is reflected in the different experts who work at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
+The diversity of epidemiology is reflected in the different experts who work at cohort studies, ranging from specialized doctors, to medical computer scientists with focus on biometrics, and statisticians.
 %Many different experts work at epidemiological studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
@@
 %	\item Hypotheses are based most commonly on observations made by clinicians in their daily routine.
-	\item A Hypothesis is derived from observations made by clinicians in their daily routine.
+	\item A hypothesis is derived from observations made by clinicians in their daily routine.
 %
@@
 	%\item Confounding features are adjusted so that they do not influence the effect size of an attribute.
-	\item Confounding features are identified and taken into account, for example using stratification % adjusted so that they do not influence the effect size of a attribute.
-%
-	\item Statistical methods such as regression analysis assess the association of selected features with the investigated condition.%are applied to measure the effect size of attributes to the outcome of interest.
+	\item Confounding features are identified and taken into account (for example using stratification). % adjusted so that they do not influence the effect size of a attribute.
+%
+	\item Statistical methods, such as regression analysis, assess the association of selected features with the investigated disease.%are applied to measure the effect size of attributes to the outcome of interest.
 	%\item Statistical methods such as regression analysis are applied to measure the effect size of attributes to the outcome of interest.
@@
 %
-The extension, we suggest, are motivated by the potential of image data and the attempt to support \emph{hypothesis generation}.
+The extension we suggest, is motivated by the potential of image data and the attempt to support \emph{hypothesis generation}.
 %In the following we discuss the problems and challenges regarding the workflow.
@@
 %
-Age for example is divided into groups (e.g. in 20 year-steps) when investigating its influence on a condition.
-%
-These groups depend strongly on the condition of interest and therefore there is no standard on how to categorize these.
-
-To determine, whether a subject is prone to be affected by a certain disease, \emph{relative risks} are expressed through the evaluation of p-values which indicate statistical significance.
-%
-Statistical correlations are prone to \emph{confounding}, meaning that the association of two features are influenced by a third feature which needs to be isolated.
+Age for example is divided into groups (e.g. in 20-year-steps) when investigating its influence on a condition.
+%
+These groups strongly depend on the condition of interest and therefore there is no standard on how to categorize them.
+
+To determine, whether a subject is prone to be affected by a certain disease, \emph{relative risks} are expressed through the evaluation of p-values, which indicate statistical significance.
+%
+Statistical correlations are prone to \emph{confounding}, meaning that the association of two features are influenced by a third feature, which needs to be isolated.
 %
@@
 %MRI data is more expensive to obtain than CT data but does not affect the subjects health and is therefore the main method for collecting cohort study imaging data.
-MRI the main method for collecting cohort study imaging data.
+Therefore is MRI the main method for collecting cohort study imaging data.
 %
@@
 %
-Starting in 1997 with a cohort consisting of 4.308 subjects, the SHIP, located in northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
+Starting in 1997 with a cohort consisting of 4.308 subjects, the \texttt{SHIP}, located in northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
 %
@@
 %Einfuehren von Helwigs Terminologie?
-Designing a visualization which conveys all data aspects equally is challenging.
-%
-Given the number of features of epidemiological data sets and their different manifestations, the strength of different visualization techniques need to be combined \cite{Buja91, Konyha2009}.
-%
-The principal component analysis (PCA) and similar techniques are able to reduce the dimension by extracting most expressive components, but make the influence of each variable hard to convey.
+Designing a visualization, which conveys all data aspects equally, is challenging.
+%
+Given the number of features of epidemiological data sets and their different manifestations, the strength of different visualization techniques needs to be combined \cite{Buja91, Konyha2009}.
+%
+The principal component analysis (PCA) and similar techniques are able to reduce the dimensions by extracting most expressive components, but make the influence of each variable hard to convey.
 \\\\
-Their focus on hypothesis generation based on parallel assessment of multiple data features makes the work of Turkay and colleagues is closest to ours albeit our emphasis on processing medical image data and variables with categorical manifestations \cite{Turkay2013}.
+Their focus on hypothesis generation based on parallel assessment of multiple data features makes the work of Turkay and colleagues closest to ours albeit our emphasis on processing medical image data and variables with categorical manifestations \cite{Turkay2013}.
 %
@@
 %Statistical measures of continuous variables such as mean, standard deviation, skewness, or inter-quartile range are used to create \emph{dimension plots} that make them comparable with respect to the derived descriptive measures.
-Statistical measures of continuous variables such as mean, standard deviation, skewness, or inter-quartile range are used to create \emph{dimension plots} that make them comparable with respect to the derived descriptive measures such as voxel number of a segmented structure.
+Statistical measures of continuous variables, such as mean, standard deviation, skewness, or inter-quartile range, are used to create \emph{dimension plots} that make them comparable with respect to the derived descriptive measures, such as voxel number of a segmented structure.
 %
@@
 %
-Steenwijk and colleagues employ a relational database to organize the data to visualize subject data using linked views such as parallel coordinates, scatterplots and time plots \cite{Steenwijk2010}.
+Steenwijk and colleagues employ a relational database to organize the data for visualizing subject data using linked views such as parallel coordinates, scatterplots and time plots \cite{Steenwijk2010}.
 %
@@
 %
-Their understanding of a cohort differs from the understanding of the term in an epidemiological context by denoting every subject group divided by parameter as individual cohort.
+Their understanding of a cohort differs from the understanding of the term in an epidemiological context by denoting every parameter divided subject group as individual cohort.
 
@@
 %
-Their findings yielded a \emph{Concept Map} which linked cancer-related associations via graph edges.
+Their findings yielded a \emph{Concept Map}, which linked cancer-related associations via graph edges.
 %From their findings regarding theassociationn of cancer-related socio-demographic factors are drawn in a \emph{Concept Map} where related factors are connected via graph-edges.
@@
 %
-Derived from this we incorporate the idea of combining 3D-shape rendering with information visualization techniques.
+Derived from this, we incorporate the idea of combining 3D-shape rendering with information visualization techniques.
 %
@@
 %
-This work focuses on incorporating these derived data as new features of the overall data set, making it possible to include it into the hypothesis validation and generation process.
+This work focuses on incorporating these derived data as new features of the overall data set, making it possible to include it into the hypothesis validation and -generation process.
 %
@@
 %- Acceptance plays a big role - you have to pick up the user by presenting information they are used to in a new way\\
-%
+% TODO: Epidemiologist compiled list ist sehr weit gefasst - bei Begründung ob die Einschränkung der Liste ein Problem darstellt
 We described the epidemiological workflow as a sequence of steps taken by domain experts which need to comprise reproducibility and statistical integrity (recall Subsection~\ref{EpidemiologicalWorkflow}).
@@
 Introducing the \emph{IVA} principle to the epidemiological application domain does aim to compensate its weaknesses rather than replacing the existing workflow.
-%The workflow we propose by introducing the \emph{IVA} principle into the epidemiological application domain does not aim to replace the existing workflow but to complement its weaknesses.
-%
-In the current state the workflow treats the data like a black box.
+%The workflow we propose by introducing the \emph{IVA} principle into the epidemiological application domain does not aim at replacing the existing workflow but to complement its weaknesses.
+%
+In the current state, the workflow treats the data like a black box.
 %
@@
 This becomes even more important when the workflow is adapted to the analysis of the medical image data, where
-domain experts annotate landmarks tediously to derive metrics such as diameters. %which are then handled like other features and analyzed using the same set of statistical tools.
+domain experts annotate landmarks tediously to derive metrics, such as diameters. %which are then handled like other features and analyzed using the same set of statistical tools.
 %
@@
 \\\\
-\emph{IVA} tries to illuminate the black box by making the domain experts part of an iterative feature list selection process, which is shown in Figure~\ref{fig:WorkflowComparison} (b) as part of the epidemiological workflow.
+\emph{IVA} tries to illuminate the black box by making the domain experts part of an iterative feature selection process, which is shown in Figure~\ref{fig:WorkflowComparison} (b) as part of the epidemiological workflow.
 %
@@
 %
-This has to be handled with care since \emph{overfitting} of expectations to the data are an imminent danger \cite{Turkay2013}.
+This has to be handled with care since \emph{overfitting} of expectations to the data is an imminent danger \cite{Turkay2013}.
 
@@
 %
-Clustering algorithms can be used to investigate associations between shape-groups and other non-image based variables.
+Clustering algorithms can be used to investigate associations between shape-groups and other non-image-based variables.
 %
@@
 %
-Epidemiologists may define a feature in a scatter plot of BMI and age and inspect the result in a histogram of body height.
+Epidemiologists may define a feature in a scatter plot of the body mass index (BMI) and age to inspect the result in a histogram of body height.
 %
@@
 %
-Data obtained using questionnaires or medical tests are often stored using statistical packages such as \texttt{SPSS} or \texttt{Stata}, which have a proprietary data format with limited export capabilities.
+Data obtained using questionnaires or medical tests are often stored using statistical packages such as \texttt{SPSS}, which have a proprietary data format with limited export capabilities.
 %
@@
 %
-This can be achieved using data wrangling tools such as \texttt{OpenRefine}\footnote{Developed by Google, Open Source; \url{http://openrefine.org/}}, which also validates the data (find missing data, clean up bad formatting, transform scales).
+This can be achieved using data wrangling tools such as \texttt{OpenRefine}\footnote{Developed by Google, Open Source; \url{http://openrefine.org/}}, which also validate the data (find missing data, clean up bad formatting, transform scales).
 %In order to verify that the conversion worked as expected and the data is valid, it is good practice to use data wrangling tools such as \texttt{OpenRefine} to validate the data (find missing data, clean up bad formatting).
@@
 %
-%Therefore, these are also included as error codes which have to marked as such in the data dictionary.
+%Therefore, these are also included as error codes which have to be marked as such in the data dictionary.
 \paragraph{Image data. }
@@
 %
-Model-based approaches have shown in principle to be effective for segmentation \cite{Gloger2010, Gloger2012} and detection \cite{Rak2013}.
+%Model-based approaches have shown in principle to be effective for segmentation \cite{Gloger2010, Gloger2012} and detection \cite{Rak2013}.
+%Model-based segmentation approaches are based on assumptions related to object shape, size, location and texture.
+Model-based approaches have shown in principle to be effective for detection \cite{Rak2013} and segmentation \cite{Gloger2010, Gloger2012}.
 %
@@
 
-Morphometric variables are derived to allow for statistical comparison of the tissue, which incorporate mostly positions, diameters, volumes and relative distances and alignment to other structures.
+Morphometric features are derived to allow for statistical comparison of the tissue, which incorporate mostly positions, diameters, volumes and relative distances and alignment to other structures.
 % 
@@
  \caption{Detailed Version of the \emph{IVA Loop} seen in Figure~\ref{fig:WorkflowComparison}.
-Usually starting with a selection of Feature of interest (user-driven or via data mining techniques), the data are mapped using a visualization technique appropriate for the selected data types.
+Usually starting with a selection of a feature of interest (user-driven or via data mining techniques), the data are mapped using a visualization technique appropriate for the selected data types.
 The data are visualized in the range and domain space, which can then be brushed, yielding new groups, which can be investigated using further features.
@@
  \includegraphics[width=1\textwidth, resolution=300]{figures/visualization}
- \caption{(Left) Screenshot from the front-end, which is divided as follows: (a) Sidebar containing all features as well as the groups defined in the analysis process; (b) Canvas area where features can be added via drag and drop and the visualization chosen automatically according to the data type; (c) Interactive pivot table showing the exact numbers for each displayed variable combination.
+ \caption{(Left) Screenshot from the front-end, which is divided as follows: (a) The Sidebar containing all features as well as the groups defined in the analysis process; (b) the canvas area where features can be added via drag and drop and the visualization chosen automatically according to the data type; (c) the interactive pivot table showing the exact numbers for each displayed feature combination.
 The data displayed is used to analyze the lumbar spine.
@@
 %
-Subsequently the visualization can be either brushed or new variables can be added to the analysis.
+Subsequently the visualization can be either brushed or new features can be added to the analysis.
 %
@@
 %
-Selecting features also triggers a \emph{multivariate analysis} using contingency values (described in the following section) to highlight variables interacting with selected features.
+Selecting features also triggers a \emph{multivariate analysis} using contingency values (described in the following section) to highlight features associated with selected features.
 %
@@
 %
-The chances of a method to be adapted by epidemiologists increases highly if their domain language is used and the methods reflect routines they are used to.
+%The chances of a method to be adapted by epidemiologists increases highly if their domain language is used and the methods reflect routines they are used to.
+The methods have to use domain language and reflect routines of epidemiologists in order to be incorporated into their research.
+%The chances of a method to be adapted by epidemiologists increases highly if their domain language is used and routines they are used to are reflected in the methods.
 %
@@
  %Mouse-over focuses a entry and highlights the feature names for better readability.
- Mouser-over an entry highlight the feature names for better readability.
+ Mouse-over an entry highlights the feature names for better readability.
  %
@@
 \subsubsection{System Structure} \label{Structure and Workflow}
-We divide the workspace into four major parts as seen in Figure~\ref{fig:visualization} and Figure~\ref{fig:similarity}.
+We divide the workspace into four major parts as seen in Figure~\ref{fig:visualization} and \ref{fig:similarity}.
 \begin{itemize}
@@
 	\item The canvas holding all visualizations. Elements can be added, arranged, resized and removed freely.
-	\item The interactive pivot table gives detailed numerical information of the features in the canvas view. This view on the data come natural to epidemiologists.
+	\item The interactive pivot table gives detailed numerical information of the features in the canvas view. This view on the data comes natural to epidemiologists.
 	\item The contingency view depicting relations for features in the canvas in an adjacency matrix.
@@
 \paragraph{Sidebar.}
-An overview of all features is presented in a sidebar where they are categorized to different types such as somatometric (measurements of the dimensions of the human body), disease- or lifestyle related, pain indicators and laboratory data.
+An overview of all features is presented in a sidebar where they are categorized to different types, such as somatometric (measurements of the dimensions of the human body), disease- or lifestyle-related, pain indicators and laboratory data.
 %
@@
 %
-Groups are treated like features since they work also by dividing the subject space into labeled categories.
-%
-Variables can be dragged from the sidebar into the canvas area for a \emph{feature localization}.
-%
-This triggers an adaptive feature visualization suitable for the current data type.
+Groups are treated like features since they act the same by dividing the subject space into labeled categories.
+%
+Variables can be dragged from the sidebar into the canvas area for a \emph{feature localization} which works as follows.
+%
+%This triggers an adaptive feature visualization suitable for the current data type.
 %Bar charts show the distribution of manifestations of each variable in the sidebar.
@@
 \paragraph{Adaptive Feature Visualization.} \label{sec:AdaptiveFeatureVisualization}
-The visualization type is, inspired by \texttt{GPloms} \cite{Francois2013}, dynamically chosen based on the feature types and number.
+The visualization type is, inspired by \texttt{GPLOMS} \cite{Francois2013}, dynamically chosen based on the feature types and number.
 %
@@
 %
-Continuous data are discretized and therefore treated as categorical.
+Continuous data are discretized and therefore treated as categorical data.
 %
@@
 % TODO Implement MOSAIC PLOTS, Scatterplots, Parallel Coordinates
-Following Tuftes concept of \emph{small multiples} \cite{Tufte1983} the medical image data are directly included into the plot by including color coded mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
+Following Tuftes concept of \emph{small multiples} \cite{Tufte1983}, the medical image data are directly incorporated into the plot by including color coded mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
 %If possible the medical image data is directly included into the plot as well by including mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
@@
 %As seen in Figure~\ref{fig:visualization} (c) it is a good way to display how many subjects are in each group.
-Pivot tables clearly convey how many subjects are in each group (see Figure~\ref{fig:visualization} (c))
-%
-However, these quickly get confusing and cluttered when they are divided into many subgroups.
+These tables clearly convey the subject count in each group (see Figure~\ref{fig:visualization} (c)).
+%
+However, they quickly get confusing and cluttered when they are divided into many subgroups.
 %
@@
 %
-To allow better comparison with respect to features the values of each cell can also be displayed as percentage of the feature represented of either the row or the column.
+To allow better comparison with respect to features, the values of each cell can also be displayed as percentage of the feature represented of either the row or the column.
 % TODO IMPLEMENT
@@
 \paragraph{Automated Feature Suggestion using a Contingency Matrix.}
-As discussed previously, highlighting potential interesting values in the data set is one major benefit of the \emph{IVA} powered approach and is categorized in the \emph{multivariate analysis} pattern analyzing features outside the shape-space.
+% TODO: Has to be triggered since visualization is very large
+As discussed previously, highlighting potentially interesting values in the data set is one major benefit of the \emph{IVA} powered approach and belongs to the \emph{multivariate analysis} pattern,  analyzing features outside the shape-space.
 %
@@
 %
-$V$ assumes values between $0$, meaning two variables are completely independent and $1$ indicating they are the same.
+$V$ assumes values between $0$, meaning two variables are completely independent, and $1$, indicating they are the same.
 %
@@
 %
-Therefore are manifestations, which are only exposed by small subject groups not assessable using this technique.
-%
-This makes them not assessable for epidemiological analysis as well since statistical validation needs a minimum count to be seen as valid.
+Some manifestations and feature combinations, which are only exposed by small subject groups, cannot be assessed with this technique.
+%
+They cannot be included into epidemiological analysis as well since statistical validation needs a minimum count to be seen as valid.
+%This makes them not assessable for epidemiological analysis as well since statistical validation needs a minimum count to be seen as valid.
 % Limitations and usual space
-The contingency matrix highlights contingencies of visualized features with features in the data set.
-%
-This aims to highlight features possibly interacting with the focused hypothesis as well as trigger new hypotheses.
+The contingency matrix highlights correlations between all data set features. %of visualized features with features in the data set.
+%
+This aims to highlight features possibly associated with the focused hypothesis as well as trigger generation of new hypotheses.
 %
@@
 %
-The distinction whether an association is a confounder or and effect depends on the context defined by the hypothesis and is a decision to be made by the domain expert.
+The distinction, whether an association is a confounder or an effect, depends on the context defined by the hypothesis and is a decision to be made by the domain expert.
 \\\\
-In the following section we discuss how we implemented the presented methods using free on open Web standards.
+In the following section we discuss how we implemented the presented methods using free on open web standards.
 % \subsubsection{Automated Feature Suggestion}
@@
 % %
-% These difference are then highlighted in a separate tab of the side bar "Interactions".
+% These differences are then highlighted in a separate tab of the side bar "Interactions".
 % %
@@
 % \\\\
-% In the following sections we will discuss details on the implementation which relies on modern Web-Technologies.
+% In the following sections we will discuss details of the implementation which relies on modern Web-Technologies.
 % \\\\
@@
  %
- All computation heavy operations like calculation of mean-shapes or -distances as well as statistical processing is done by the server to keep hardware requirements of client systems low. 
- %
- Client-Server communication is accomplished via the \emph{Websocket} protocol.
+ All computation-heavy operations, such as calculation of mean-shapes or -distances, as well as statistical processing is done by the server to keep hardware requirements of client systems low. 
+ %
+ Client-server communication is accomplished via the \emph{Websocket} protocol.
  }
@@
 \begin{itemize}
-	\item No additional software needs to be installed, most people use decent state-of-the-art web browser, even on mobile devices.
+	\item No additional software needs to be installed, most people use a decent state-of-the-art web browser, even on mobile devices.
 	\item The client-server structure allows for employing heavy computation on a server machine and transfer results to the client.
@@
 	High confidentiality standards of the data can be met by restricting access via a account system
-	\item Recent developments in WebGL applications running in browsers with near native performance push the development into to the web resulting in many open source libraries, which are well documented, rich in examples and driven by active communities.
-We use WebGL for rendering shape information.
+	\item Recent developments in \emph{WebGL} applications running in browsers with near native performance push the development into to the web resulting in many open source libraries, which are well documented, rich in examples and driven by active communities.
+We use \emph{WebGL} for rendering shape information.
 	\item Web technologies are free.
@@
 %Many methods which specialized libraries, like the Visualization Toolkit (VTK) or R for statistics have build in, need to be written from scratch in order to fit in the context.
-Due to missing ports of sophisticated specialized libraries/languages, such as the Visualization Toolkit (VTK) or R for statistics, many standard methods need to be written from scratch for the web.
+Due to missing ports of sophisticated specialized libraries/languages, such as the \texttt{Visualization Toolkit} (\texttt{VTK})\footnote{Developed by Kitware Inc; \url{www.vtk.org}} or \texttt{R}\footnote{Open Source; \url{www.r-project.org}} for statistics, many standard methods need to be written from scratch for the web.
 %Many methods, which specialized libraries, like the Visualization Toolkit (VTK) or R for statistics, have build in, need to be written from scratch in order to fit in the context.
@@
 %
-Image data are available as raw DICOM files as well as meshes representing segmentation masks, which can be used to compare subjects.
-%
-On client connection the requested files are transferred.
-%
-The server processes calculation heavy statistical tasks such as calculation of \emph{Cram\'{e}rs' V} values for all variable combinations in order to keep the computation time on the client as low as possible.
+Image data are available as raw DICOM files as well as meshes representing segmentation masks of anatomical structures, which can be used to compare subjects.
+%
+The requested data is transmitted when a client connects.
+%
+The server processes calculation heavy statistical tasks such as calculation of \emph{Cram\'{e}rs' V} values for all feature combinations in order to keep the computation time on the client as low as possible.
 %
@@
 %
-Information Visualizations such as scatterplots and bar charts are created using the popular \texttt{Data-Driven Documents (D3.js)} \cite{D3} library, which works well for attaching data to visible elements like vector graphics and provides powerful transformation and mapping tools.
+Information visualizations such as scatterplots and bar charts are created using the popular \texttt{Data-Driven Documents (D3.js)} library \cite{D3}, which works well for attaching data to visible elements like vector graphics and provides powerful transformation and mapping tools.
 %
@@
 %
-\texttt{WebGL} rendering is done using the \texttt{Three.js}\footnote{Originally developed by Ricardo Cabello, \url{threejs.org}} and allows GPU-accelerated data rendering.
-%
-Communication between Client and Server runs through the WebSockets protocol.
-%
-Since our clustering algorithms are written in \texttt{MatLab}\footnote{Owned by The MathWorks, \url{www.mathworks.com}} we hat to access them using the \texttt{NodeJS} Server.
+\texttt{WebGL} rendering is carried out using the \texttt{Three.js}\footnote{Originally developed by Ricardo Cabello, \url{threejs.org}} and allows GPU-accelerated data rendering.
+%
+Communication between client and server is enabled via the WebSockets protocol.
+%
+Since our clustering algorithms are written in \texttt{MatLab}\footnote{Owned by The MathWorks, \url{www.mathworks.com}}, we hat to access them using the \texttt{NodeJS} server.
 %
@@
 %
-Epidemiological analysis of lumbar back pain such the work of Harreby and colleagues \cite{Harreby1996} is largely focused on non-image information.
+Epidemiological analysis of lumbar back pain, such as the work of Harreby and colleagues \cite{Harreby1996}, is largely focused on non-image information.
 %
 %If at all, only a few shape-related features are included in comparable studies, for example by Lang and colleagues \cite{Lang2011}.
-In comparable studies, only a few shape-related features are included, for example by Lang and colleagues \cite{Lang2011}.
+%In comparable studies, only a few shape-related features are included, for example by Lang and colleagues \cite{Lang2011}.
+In comparable studies, only a few shape-related features are included \cite{Lang2011}.
 %
@@
 	\item a better understanding of effects of preventive measures such as occupational health and safety regulations,
-	\item prognostic features for diagnosis and treatment of lumbar back pain,
-	\item determination of particularly effected risk groups.
+	\item prognostic features for diagnosis and treatment of lumbar back pain, and
+	\item determination of particularly affected risk groups.
 \end{itemize}
@@
 %
-Characterizing the healthy aging process of the spine is a long term goal that allows to determine age-normalized probabilities for spine-related diseases by incorporating individual risk factors.
-
-Data confidentiality and ethical reasons prohibit us from accessing the complete data space of the SHIP feature space.
+Characterizing the healthy aging process of the spine is a long term goal for determining age-normalized probabilities for spine-related diseases by incorporating individual risk factors.
+
+Data confidentiality and ethical reasons prohibit us from accessing the complete \texttt{SHIP} feature space.
 %
@@
 %
-\subsection{The Lumbar Spine Dataset}
+\subsection{The Lumbar Spine Data Set}
 We divide the data set in image and non-image data.
 %
-There are 136 features describing diagnosed diseases, lifestyle factors, women-specific factors, pain indicators, laboratory values, somatometric variables. %and are ordered accordingly.
+There are 136 features describing diagnosed diseases, lifestyle factors, women-specific factors, pain indicators, laboratory values, somatometric features. %and are ordered accordingly.
 %
@@
 
-To ensure a fast and easy data access outside of statistical processors like \texttt{SPSS}, the data was exported to the JSON file format, that can easily be parsed by modern programming languages.
-%
-Each feature is stored as an object containging: 
+To ensure a fast and easy data access outside of statistical processors like \texttt{SPSS}, the data was exported to the JSON file format.
+%
+Each feature is stored as an object containing: 
 \begin{itemize}
@@
 	\item the data type (continuous, nominal, ordinal, dichotomous),
-	\item a detailed description of the variable, and
+	\item a detailed description of the feature, and
 	\item the data dictionary translating value- or error IDs to the actual values.
@@
 \paragraph{Image Data.} \label{Image-Data}
-The lumbar spine was detected in the image data using a hierarchical finite element method according to Rak and colleagues \cite{Rak2013}.
+The lumbar spine was detected in the image data using a hierarchical finite element method by Rak and colleagues \cite{Rak2013}.
 %
@@
 %
-For clustering purposes we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (medical terms for spine curvature \cite{Klemm2013VMV}).
+For clustering purposes, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature \cite{Klemm2013VMV}).
 
@@
 %
-For dichotomous variables, the color codes distances between mean shapes of the two groups, for variables with more than two manifestation it encodes the distance to the global mean shape of all subjects (Figure~\ref{fig:hypothesisfree}).
+For dichotomous variables, the color codes distances between mean shapes of the two groups, for variables with more than two manifestation it encodes the distance to the global mean shape of all subjects (see Figure~\ref{fig:hypothesisfree}).
 \\\\
@@
 
-\subsection{Exploratory Analysis of the Lumbar Spine Dataset}
-%
-%We analyzed the given dataset expert guided to assess its suitability for supporting both hypothesis-free analysis as well as generation.
-Expert guided analysis assessed the suitability of our approach for supporting both hypothesis-free analysis as well as hypothesis generation. 
+\subsection{Exploratory Analysis of the Lumbar Spine Data Set}
+%
+%We analyzed the given dataset expert-guided to assess its suitability for supporting both hypothesis-free analysis as well as generation.
+Expert-guided analysis assessed the suitability of our approach for supporting both hypothesis-free analysis as well as hypothesis generation. 
 %
@@
 The difference to the mean-shape is color-coded where blue represents no difference and red a large difference.
-(b) A pivot table relating cluster results to gender and employment status.
-Cluster 8 is striking because of the high proportion of women while Cluster 2, 3, 4 contain almost only men.}
+(b) A pivot table relates cluster results to gender and employment status.
+Cluster 8 is striking because of the high proportion of women, while Cluster 2, 3, 4 contain almost only men.}
  \label{fig:hypothesisfree}
@@
 
-Cluster 9 represents the subjects with average shape.
-%
-Other shapes differ with respect to size, such as Cluster 2, 8, 10, 3 where the last one also represents a more more straight spine, which is usual for subjects with larger body size.
-%
-Cluster 4, 7, 5 contain outliers, characterized by their unusual shape and small number.
-%
-Cluster 8 has the 2nd largest number of elements and was therefore of special interest.
+Cluster 9 represents subjects with average shape.
+%
+Other shapes differ with respect to size, such as Cluster 2, 3, 8, 10 where the last one also represents a more more straight spine, which is usual for subjects with larger body size.
+%
+Cluster 4, 5, 7 contain outliers, characterized by their unusual shape and small number.
+%
+Cluster 8 has the second largest number of elements and was therefore of special interest.
 %Cluster 8 was of special interest because of its large distance to the mean shape while still exposing the second highest subject count.
@@
 %
-Another \emph{multivariate analysis} using a pivot table set gender and employment status in relation to clustering affiliation, showing that Cluster~8 contains mostly women and also has a larger unemployment (see Figure~\ref{fig:hypothesisfree}), while the overall employment rate of women and men in the data set is almost exactly 50\%.
+Another \emph{multivariate analysis} using a pivot table set gender and employment status in relation to clustering affiliation, showing that Cluster~8 contains mostly women and also has a larger unemployment rate (see Figure~\ref{fig:hypothesisfree}), while the overall employment rate of women and men in the data set is almost exactly 50\%.
 %
@@
 %the data are met with a selection bias when compiling the list of related features by the domain experts.
-%
-It is arguable whether this first filtering step, which is purely based on expert experience, represents a disadvantage, since it rules out many parameters possibly interacting with the presented features but the value of this knowledge would be small \cite{Wiley2008}.
+% TODO: In der Einleitung erwähnen, dass man bisher nur maximal 3 Variablen gleichzeitig betrachten kann, damit man unsere Technik etwas in den Kontext rückt
+%
+This filtering step based on expert experience rules out possible confounder and useless associations and is therefore arguable, whether this represents a disadvantage \cite{Wiley2008}.
+%It is arguable whether this first filtering step, which is purely based on expert experience, represents a disadvantage, since it rules out many parameters possibly associated with the presented features, but the value of this knowledge would be small \cite{Wiley2008}.
+%
 % While this approach does not assume any hypothesis, the data are met with a selection bias when compiling the list of related features by the domain experts.
@@
  \includegraphics[width=3.5in]{figures/hypothesisbased}
- \caption{(a) Dichotomous questionnaire answer to "Did you experience back pain in the past three month".
- %
- Mean-Shapes between the groups show no difference.
+ \caption{(a) Dichotomous questionnaire answer to "\emph{Did you experience back pain in the past three month}".
+ %
+ Mean-shapes between the groups show no difference.
  %
@@
 %
-If the user proposed a hypothesis about a relation between a non-image feature regarding shape the workflow slightly differs from the hypothesis-free analysis.
-%
-The starting point of the analysis is a \emph{feature investigation}, where a feature of interest is selected by dragging it into the canvas area and view the subjects distribution as well as their shape differences.
-%
-In our use case, epidemiologists were interested in the dichotomous questionnaire answer to "Did you experience back pain in the past three month".
-%
-The mean shapes of the resulting visualization as seen in Figure~\ref{fig:hypopthesisbased} (a) show no difference between the two groups.
+If the user proposed a hypothesis about a relation between a non-image feature regarding shape, the workflow slightly differs from the hypothesis-free analysis.
+%
+The starting point follows the \emph{feature investigation} pattern, where a feature of interest is selected by dragging it into the canvas area and view the subjects distribution as well as their shape differences.
+%
+In our use case, epidemiologists were interested in the dichotomous questionnaire answer to "\emph{Did you experience back pain in the past three month}".
+%
+The mean shapes of the resulting visualization show no difference between the two groups (see Figure~\ref{fig:hypopthesisbased} (a)).
 %
@@
 %
-Since the focus is on subjects suffering from back pain, the clustering result of these subjects are then drawn into the canvas area, yielding six cluster as seen in Figure~\ref{fig:hypopthesisbased} (b).
-%
-Cluster 5 stood out for having a so-called hyperlordosis, a strong curvature of the lumbar spine associated with back pain.
+Since the focus is on subjects suffering from back pain, the clustering result of these subjects are then drawn into the canvas area, yielding six clusters as seen in Figure~\ref{fig:hypopthesisbased} (b).
+%
+Cluster 5 stood out for having a so-called \emph{hyperlordosis}, a strong curvature of the lumbar spine associated with back pain.
 
@@
 
-It is well known that overweight is a indicator for back pain.
-%
-While the body mass index (BMI) is a key figure for assessing height and weight of a subject, it does not tell anything on how the weight is distributed in the body.
+% TODO Bernhard fragen, was ihn an der Reihenfolge stört
+It is well known that overweight is a risk factor for back pain.
+%
+While the BMI is a key figure for assessing height and weight of a subject, it does not tell anything on how the weight is distributed in the body.
 %
@@
 %
-To increase the probability of the observation not to be random and to avoid \emph{overfitting}, they are cross check for associations in \texttt{SHIP-TREND} as a second, independent population sample, which was examined with methods identical to SHIP.
+To increase the probability of the observation not to be random and to avoid \emph{overfitting}, they are cross checked for associations in \texttt{SHIP-TREND} as a second, independent population sample, which was examined with methods identical to \texttt{SHIP}.
 %If features are identified as interesting for a disease, the epidemiologists look for associations in another cohort, the \texttt{SHIP-TREND}.
@@
 \section{Summary and Conclusion}
+% TODO: Include IVA Pattern here?
 %Seen as an extension to the existing statistics driven workflow, image-centric analysis of epidemiological data from a IVA approach by illuminating the data-black box.
@@
 A number of improvements is left open for future work:
+%TODO: Fragen ob die Liste so klar geht mit der Formulierung und ohne Kommas
 \begin{itemize}
@@
 %
-This requires a large number of subjects, especially if the investigated features are rare and are only presented by a few subjects.
+This requires a large number of subjects, especially if the investigated features are rare and only presented by a few subjects.
 %
@@
 \\\\
-As the number of image-centric cohort studies, participating subjects, gathered features and imaging modalities rises and advances towards comparability between cohort studies are made, the gap between data complexity and assessability increases.
+As the number of image-centric cohort studies, participating subjects, gathered features and imaging modalities rises and advances towards comparability between cohort studies are made, the gap between data complexity and analysability increases.
 %
@@
 %
-We believe web technologies pave the way to allow as many experts as possible to analyze this data and provide a fast exchange between users and developers employing many different devices.
+We believe web technologies pave the way to allow as many experts as possible to analyze this data.
+%
+They allow a fast exchange between users and developers and employ many different devices.
 %

### 2014-03-30 draft.tex
Included Bernhards feedback

@@
 %
-Also, the introduction of preventive measures in medicine and beyond, are based on epidemiological research, where, for example, subgroups with increased risk are identified \cite{Fletcher2012}.
+Also, the introduction of preventive measures in medicine and beyond are based on epidemiological research, where, for example, subgroups with increased risk are identified \cite{Fletcher2012}.
 %
@@
 %
-Methods like regression analysis check the attribute list for statistical soundness.
+Methods like regression analysis are used to check the attribute list for statistical soundness.
 %
@@
 %
-A large size of the cohort is therefore essential to investigate difference between healthy and diseased people.
+A large size of the cohort is therefore essential to investigate differences between healthy and diseased people.
 %
@@
 %
-Semi-automatic techniques are more promising but also challenging since the employed modalities, such as Magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
+Semi-automatic techniques more promising but also challenging since the employed modalities, such as Magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
 %
@@
 %
-The standard workflow lacks methods, which identifies correlation the epidemiologists did not consider.
+The standard workflow lacks methods, which identify correlations the epidemiologists did not consider.
 %The hypothesis driven analysis approach of these data based on expectations of epidemiologists leaves out features which potentially interact with a disease outside.
@@
  \includegraphics[width=3.0in]{figures/workflow_comparison}
- \caption{\emph{IVA} tools are able to complement parts of the epidemiological workflow, not replace it.
+ \caption{\emph{IVA} tools complement parts of the epidemiological workflow, not replace it.
 The appropriate combination of statistical- and interactive driven analysis shows promising potential to unveil the information in the data.
@@
 \end{enumerate}
-The workflow is shown in Figure~\ref{fig:WorkflowComparison} (a) and serves as orientation for our approach although it doesn't consider any kind of image data.
-%
-The extension we suggest, is motivated by the potential of image data and the attempt to support \emph{hypothesis generation}.
+The workflow is shown in Figure~\ref{fig:WorkflowComparison} (a) and serves as orientation for our approach although it does not consider any kind of image data.
+%
+We focus, on the potential of image data and attempt to support \emph{hypothesis generation}.
+%The extension we suggest, is motivated by the potential of image data and the attempt to support \emph{hypothesis generation}.
 %In the following we discuss the problems and challenges regarding the workflow.
@@
 
-In longitudinal cohort design, grouping subjects using epidemiological features is essential in order to allow per-group risk determination.
+In longitudinal cohort studies, grouping subjects using epidemiological features is essential in order to allow per-group risk determination.
 %
@@
 %
-Information derived by landmarks such as top and bottom point are by far not as expressive and versatile as segmentation masks describing the whole vertebra.
+Information derived by landmarks, such as top and bottom point are by far not as expressive and versatile as segmentation masks describing the whole vertebra.
 %Information derived by landmarks such as top and bottom point of a vertebra are also not nearly as expressive and versatile as segmentation masks describing the whole vertebra.
 %
-They are also prone to a high inter-observer variability and hard to reproduce.
+They are also prone to a high inter-observer variability and are hard to reproduce.
 %
@@
 \\\\
-Their focus on hypothesis generation based on parallel assessment of multiple data features makes the work of Turkay and colleagues closest to ours albeit our emphasis on processing medical image data and variables with categorical manifestations \cite{Turkay2013}.
+Their focus on \emph{hypothesis generation} using parallel assessment of multiple data features makes the work of Turkay and colleagues closest to ours albeit our emphasis on processing medical image data and variables with categorical manifestations \cite{Turkay2013}.
 %
@@
 %\emph{Over-fitting} of hypotheses to the data may be imposed  is a danger pointed out by the authors when hypotheses to the data based on observations of changes in these plots the measure highlights only subsets of statistical changes.
-Hypotheses based on observations of changes in these plots may impose \emph{overfitting} to the data because the measure highlights only subsets of statistical changes.
-%
-Our approach sticks more to the information extracted from the segmented image data and derive variable associations with non-image epidemiological factors.
+Hypotheses based on observations of changes in these plots may impose \emph{overfitting} to the data because the measures highlights only subsets of statistical changes.
+%
+Our approach sticks more to the information extracted from the segmented image data and variable associations with non-image epidemiological factors.
 % Disadvantage - Nur 2 abgeleitete Variablen sichtbar, schlecht mit anderen Variablen in Verbindung zu bekommen (müsste jedoch auch gehen) - viele Dinge gehen durch die Abstraktion aber auch flöten!
@@
 \paragraph{Commercial Data Visualization.}
-Commercial systems such as \texttt{Tableau}\footnote{Owned by Tableau Software; \url{www.tableausoftware.com}} or \texttt{Spotfire}\footnote{Owned by TIBCO; \url{spotfire.tibco.com}} provide a rich user interface that enable Visual Analytics approach without the need of writing any code.
+Commercial systems such as \texttt{Tableau}\footnote{Owned by Tableau Software; \url{www.tableausoftware.com}} or \texttt{Spotfire}\footnote{Owned by TIBCO; \url{spotfire.tibco.com}} provide a rich user interface that enable a Visual Analytics approach without the need of writing any code.
 %
@@
 %
-These systems share limitations in extensibility to a specific problem domain.
+These systems with their origin in business intelligence applications do not support epidemiological workflows, in particular workflows that also consider medical image data.
+%These systems share limitations in extensibility to a specific problem domain.
 %Spotfire offers the possibilities to interact with the \texttt{R} statistical programming language 
@@
 \paragraph{Visualizing Shape Variance.}
-Comparing tissue between many subjects in an epidemiological context requires methods which allow for shape variance visualizations.
-%
-Caban and colleagues investigated the suitability of variance visualizations of shape distribution models and concluded in their user study that users favor spherical glyph representations over deformation grids and likelihood volumes \cite{Caban2011}.
+Comparing tissue between many subjects requires methods which allow for shape variance visualizations.
+%
+Caban and colleagues investigated the suitability of variance visualizations of shape distribution models and concluded that users favor spherical glyph representations over deformation grids and likelihood volumes \cite{Caban2011}.
 %
@@
 %
-Derived from this, we incorporate the idea of combining 3D-shape rendering with information visualization techniques.
+We incorporate the idea of combining 3D shape rendering with information visualization techniques.
 %
@@
 %
-Hermann and colleagues identify local deformation changes by investigating shape-related difference \cite{Hermann2014}. 
-%
-The user specifies a deformation of interest and showing corresponding changes in the shape using covariance tensors.
+Hermann and colleagues identify local deformation changes by investigating shape-related difference on rodent mandibles \cite{Hermann2014}.
+% TODO: Blöde formulierung
+Specifying a change in shape of a certain area by the user shows how other parts of the model would react using covariance tensors.
+%Covariance tensors indicate how a model is deformed when the user given a user specified deformation.
+%The user specifies a deformation of interest and showing corresponding changes in the shape using covariance tensors.
 %
@@
 %Including the data dictionary, which stores information about each manifestation of a feature, helps to get a detailed description of data variables and the meaning and unit of measurement of their values.
-A data dictionary stores information about each manifestation of a feature and is stored as lookup table to get a detailed description of data variables and the meaning and unit of measurement of their values.
+A data dictionary stores information about each manifestation of a feature and is stored as lookup table to get a detailed description of data variables, the meaning and unit of measurement of their values.
 %Exporting the data dictionary, which stores information about each manifestation of a feature, is also an important step to get a detailed description of data variables and the meaning and unit of measurement of their values.
 %
-Missing data are denoted using error codes indicating its cause ranging from ethical to medical and personal issues.
+Missing data are denoted using error codes indicating its cause ranging from ethical to medical and personal issues (recall Subsection~\ref{EpidemiologicalData}).
 %The reasons for missing data have a wide range from ethical to medical and personal issues.
@@
 \paragraph{Image data. }
-Processing the image data associated to each subject consists for the most part of information extraction about an anatomical structure.
+%Gerade weil der Satz davor so kompliziert war, ist unklar, was "this" bedeutet.
+Ich denke zunächst muss gesagt werden, was typische Ziele sind (anhand von Beispielen, wie Abstände zwischen anatomischen Landmarken, Ausdehnungen (also längste Durchmesser).
+Wichtig wären aber konkrete relevante Beispiele.
+Dann wird gesagt, was man dafür machen muss und wie man das macht.
+Information about anatomical structure, such as diameter or volumes, is extracted from the image data.
+%Processing the image data associated to each subject consists for the most part of information extraction about an anatomical structure.
 %
@@
 % - Extraction of Image Parameter
-\subsection{Analysis workflow}
+\subsection{Analysis Workflow}
 \begin{figure}[htb]
@@
  % TODO: Recreate Image: Also include Cramér´s V which is part of the visualization
- \caption{Detailed Version of the \emph{IVA Loop} seen in Figure~\ref{fig:WorkflowComparison}.
+ \caption{Detailed \emph{IVA Loop} as seen in Figure~\ref{fig:WorkflowComparison}.
 Usually starting with a selection of a feature of interest (user-driven or via data mining techniques), the data are mapped using a visualization technique appropriate for the selected data types.
@@
 % TODO Implement MOSAIC PLOTS, Scatterplots, Parallel Coordinates
-Following Tuftes concept of \emph{small multiples} \cite{Tufte1983}, the medical image data are directly incorporated into the plot by including color coded mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
+Following Tuftes concept of \emph{small multiples} \cite{Tufte1983}, the medical image data are directly incorporated into the plot by including color-coded mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
 %If possible the medical image data is directly included into the plot as well by including mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
 %
-The 3D-plots can be navigated using standard mouse inputs, the camera is synchronized between all views to enable direct comparison.
+The 3D plots can be navigated using standard mouse inputs, the camera is synchronized between all views to enable direct comparison.
 %
@@
 \end{equation}
-where $X^2$ equals \emph{Pearson's chi squared}, $N$ is the total of observations and $k$ either the row or column count, depending on which one is lower.
+where $X^2$ equals \emph{Pearson's chi squared}, $N$ is the total number of observations and $k$ either the row or column count, depending on which one is lower.
 %
@@
 %
-The expected counts in the contingency table have to be larger than 5 for $80\%$ of the entries and no expected value must be smaller than one!
+The expected counts in the contingency table have to be larger than 5 for $80\%$ of the entries and no expected value must be smaller than one \cite{Cochran1952}!
 %
@@
 %
-To provide a fast communication loop between method development and expert input, we decided to rely on modern web technologies, which benefits from various advantages:
+To provide a fast communication loop between method development and expert input, we decided to rely on modern web technologies.
+%
+In addition to the obvious advantages of web technologies, the following aspects are crucial for our work:
 \begin{itemize}
-	\item No additional software needs to be installed, most people use a decent state-of-the-art web browser, even on mobile devices.
+%	\item No additional software needs to be installed, most people use a decent state-of-the-art web browser, even on mobile devices.
 	\item The client-server structure allows for employing heavy computation on a server machine and transfer results to the client.
@@
 We use \emph{WebGL} for rendering shape information.
-	\item Web technologies are free.
+%	\item Web technologies are free.
 \end{itemize}
@@
 %
-All parameter steered console applications can be incorporated in this context.
+All parameter-steered console applications can be incorporated in this context.
 \\\\
@@
 %
-The method has proven to produce comprehensible results on a preliminary data set and was able to reproduce textbook knowledge \cite{Klemm2013VMV}.
+The method has proven to produce sound results on a preliminary data set and was able to reproduce textbook knowledge \cite{Klemm2013VMV}.
 
@@
 Metabolic parameters, such as blood-fat and blood-sugar, are also possibly associated features.
-%
-Employment status is a feature, which relates to many different lifestyle factors or age and might therefore act as confounder.
+% TODO: Mit Bernhard klären
+Employment status is a feature relating to many different lifestyle factors such as income or nutrition as well as age and might act as confounder.
 %TODO Implement Color coding only in a certain direction!\\
@@
  \includegraphics[width=3.5in]{figures/hypothesisbased}
- \caption{(a) Dichotomous questionnaire answer to "\emph{Did you experience back pain in the past three month}".
+ \caption{(a) Dichotomous questionnaire answer to "\emph{Did you experience back pain in the past three month?}".
  %
@@
 %
-In our use case, epidemiologists were interested in the dichotomous questionnaire answer to "\emph{Did you experience back pain in the past three month}".
+In our use case, epidemiologists were interested in the dichotomous questionnaire answer to "\emph{Did you experience back pain in the past three month?}".
 %
@@
 % TODO Bernhard fragen, was ihn an der Reihenfolge stört
-It is well known that overweight is a risk factor for back pain.
+It is well known that overweight is an risk factor for back pain.
 %
@@
 %
-Our epidemiological collaborators were interested in the fact that this group presented a correlation with waist circumference.
+Our epidemiological collaborators were interested in the correlation with waist circumference presented by this group.
+%Our epidemiological collaborators were interested that this group presented a correlation with waist circumference.
 %
@@
 Correlation does not automatically imply causation \cite{Tufte2003}.
+%
+In particular, when large amounts of attributes are automatically checked for strong correlations, just by chance hundreds of them reach at least a moderately significant correlation, such as $p>0.05$
 %Defining a causal relationship solely based on observed correlations of two features is \emph{cum hoc ergo propter hoc}--correlation does not automatically imply causation \cite{Tufte2003}.
@@
 %
-The iterative nature of the \emph{IVA} approach reflects versatility as both \emph{hypothesis driven analysis} and \emph{hypothesis generation} is allowed.
+The iterative nature of the \emph{IVA} approach reflects versatility as both \emph{hypothesis-driven analysis} and \emph{hypothesis generation} is allowed.
 \\\\\

### 2014-03-31 draft.tex
Added changes from Manuel

@@
 
-Statistics tools such as \texttt{SPSS}\footnote{Product of IBM; \url{www.ibm.com/software/de/analytics/spss/}} play a major role for analyzing epidemiological data.
+Statistics tools such as \texttt{SPSS}\footnote{Product of IBM; \url{http://ibm.com/software/analytics/spss/}} play a major role for analyzing epidemiological data.
 %
@@
 %
-Segmentation masks labeling the voxels of an anatomical structure would be ideal since many different key figures, e.g., volume, largest diameter or aspect ratio, can be derived from them.
+Segmentation masks representing the voxels of an anatomical structure would be ideal since many different key figures, e.g., volume, largest diameter or aspect ratio, can be derived from them.
 %
@@
 %
-Morphometric information from landmarks comprises thickness, diameter or length of a structure as well as grey-value distribution in an area (used for determining the type of tisse).
+Morphometric information from landmarks comprises thickness, diameter or length of a structure as well as grey-value distribution in an area (used for determining the type of tissue).
 
@@
 \paragraph{Commercial Data Visualization.}
-Commercial systems such as \texttt{Tableau}\footnote{Owned by Tableau Software; \url{www.tableausoftware.com}} or \texttt{Spotfire}\footnote{Owned by TIBCO; \url{spotfire.tibco.com}} provide a rich user interface that enable a Visual Analytics approach without the need of writing any code.
+Commercial systems such as \texttt{Tableau}\footnote{Owned by Tableau Software; \url{http://tableausoftware.com}} or \texttt{Spotfire}\footnote{Owned by TIBCO; \url{http://spotfire.tibco.com}} provide a rich user interface that enable a Visual Analytics approach without the need of writing any code.
 %
@@
 %
-This can be achieved using data wrangling tools such as \texttt{OpenRefine}\footnote{Developed by Google, Open Source; \url{http://openrefine.org/}}, which also validate the data (find missing data, clean up bad formatting, transform scales).
+This can be achieved using data wrangling tools such as \texttt{OpenRefine}\footnote{Developed by Google, Open Source; \url{http://openrefine.org}}, which also validate the data (find missing data, clean up bad formatting, transform scales).
 %In order to verify that the conversion worked as expected and the data is valid, it is good practice to use data wrangling tools such as \texttt{OpenRefine} to validate the data (find missing data, clean up bad formatting).
@@
 %Including the data dictionary, which stores information about each manifestation of a feature, helps to get a detailed description of data variables and the meaning and unit of measurement of their values.
-A data dictionary stores information about each manifestation of a feature and is stored as lookup table to get a detailed description of data variables, the meaning and unit of measurement of their values.
+%A data dictionary stores information about each manifestation of a feature and is stored as lookup table to get a detailed description of data variables and the meaning and unit of measurement of their values
+A data dictionary stores information about each manifestation of a feature.
+%
+Detailed description of data variables, its meaning as well as unit of measurement are stored as a lookup table.
 %Exporting the data dictionary, which stores information about each manifestation of a feature, is also an important step to get a detailed description of data variables and the meaning and unit of measurement of their values.
@@
  \caption{Detailed \emph{IVA Loop} as seen in Figure~\ref{fig:WorkflowComparison}.
+ %
  Usually starting with a selection of a feature of interest (user-driven or via data mining techniques), the data are mapped using a visualization technique appropriate for the selected data types.
-The data are visualized in the range and domain space, which can then be brushed, yielding new groups, which can be investigated using further features.
+ %
+ The data are visualized in the range and domain space, which can then be brushed, yielding new groups, to be investigated using further features.
+ % The data are visualized in the range and domain space, which can then be brushed, yielding new groups, to be investigated using further features. 
+ %
  Note that adjacent steps are directly connected via feedback loops allowing for iterative refinement and give as much freedom to the user as possible.}
@@
 %
-A feature selection can also be derived from a shape-based clustering, which creates shape groups.
+A feature selection from a shape-based clustering, which creates shape groups.
 %
@@
 %
-To allow better comparison with respect to features, the values of each cell can also be displayed as percentage of the feature represented of either the row or the column.
+To allow better comparison with respect to features, the values of each cell can also be displayed as percentage of the feature represented of either the row or column.
 % TODO IMPLEMENT
@@
  %
- All computation-heavy operations, such as calculation of mean-shapes or -distances, as well as statistical processing is done by the server to keep hardware requirements of client systems low. 
+ All computation-heavy operations, such as calculation of mean shapes or -distances, as well as statistical processing is done by the server to keep hardware requirements of client systems low. 
  %
@@
 %	\item No additional software needs to be installed, most people use a decent state-of-the-art web browser, even on mobile devices.
-	\item The client-server structure allows for employing heavy computation on a server machine and transfer results to the client.
+	\item The client-server structure allows for employing heavy computation on a capable server machine and transfer results to the client.
 	\item Since image data for several thousand subjects requires hundreds of gigabytes disk space it can remain safely on the server and elements can be transferred on demand.
 	%
-	High confidentiality standards of the data can be met by restricting access via a account system
-	\item Recent developments in \emph{WebGL} applications running in browsers with near native performance push the development into to the web resulting in many open source libraries, which are well documented, rich in examples and driven by active communities.
+	High confidentiality standards of the data can be met by restricting access via an account system
+	\item Recent developments in \emph{WebGL} applications running in browsers with near native performance push the development further into to the web resulting in many open source libraries, which are well documented, rich in examples and driven by active communities.
 We use \emph{WebGL} for rendering shape information.
@@
 %Many methods which specialized libraries, like the Visualization Toolkit (VTK) or R for statistics have build in, need to be written from scratch in order to fit in the context.
-Due to missing ports of sophisticated specialized libraries/languages, such as the \texttt{Visualization Toolkit} (\texttt{VTK})\footnote{Developed by Kitware Inc; \url{www.vtk.org}} or \texttt{R}\footnote{Open Source; \url{www.r-project.org}} for statistics, many standard methods need to be written from scratch for the web.
+Due to missing ports of sophisticated and specialized libraries/languages, such as the \texttt{Visualization Toolkit} (\texttt{VTK})\footnote{Developed by Kitware Inc; \url{http://vtk.org}} or \texttt{R}\footnote{Open Source; \url{http://r-project.org}} for statistics, many standard methods need to be written from scratch for the web.
 %Many methods, which specialized libraries, like the Visualization Toolkit (VTK) or R for statistics, have build in, need to be written from scratch in order to fit in the context.
@@
 
-The back-end is written using \texttt{NodeJS}\footnote{Developed by Joyent Inc, \url{nodejs.org}}, which is based on the Google V8 Javascript runtime environment.
+The back-end is written using \texttt{NodeJS}\footnote{Developed by Joyent Inc, \url{http://nodejs.org}}, which is based on the Google V8 Javascript runtime environment.
 %
@@
 \\\\
-Non-image data for all subjects including the data dictionary is stored in a JSON file on the Server.
+Non-image data for all subjects including the data dictionary is stored in a JSON file on the server.
 %
@@
 %
-The server processes calculation heavy statistical tasks such as calculation of \emph{Cram\'{e}rs' V} values for all feature combinations in order to keep the computation time on the client as low as possible.
-%
-
-The front-end is created using \texttt{Twitter Bootstrap}\footnote{Developed by Twitter, \url{getbootstrap.com}} as foundation for the layout and basic UI elements using \texttt{HTML5}, \texttt{CSS3} and \texttt{Javascript}.
+The server performs heavy statistical tasks such as calculation of \emph{Cram\'{e}rs' V} values for all feature combinations in order to keep the computation time on the client as low as possible.
+%
+
+The front-end is created using \texttt{Bootstrap}\footnote{Developed by Twitter, \url{http://getbootstrap.com}} as foundation for the layout and basic UI elements using \texttt{HTML5}, \texttt{CSS3} and \texttt{Javascript}.
 %
@@
 %
-\texttt{WebGL} rendering is carried out using the \texttt{Three.js}\footnote{Originally developed by Ricardo Cabello, \url{threejs.org}} and allows GPU-accelerated data rendering.
+\texttt{WebGL} rendering is carried out by \texttt{Three.js}\footnote{Originally developed by Ricardo Cabello, \url{http://threejs.org}} and allows GPU-accelerated data rendering.
 %
@@
 %
-Since our clustering algorithms are written in \texttt{MatLab}\footnote{Owned by The MathWorks, \url{www.mathworks.com}}, we hat to access them using the \texttt{NodeJS} server.
+Since our clustering algorithms are written in \texttt{MatLab}\footnote{Owned by The MathWorks, \url{http://mathworks.com}}, we hat to access them using the \texttt{NodeJS} server.
 %
@@
 \\\\
-The following section describes how our presented \emph{IVA} workflow works for an epidemiological use case.
-
+%The following section describes how our presented \emph{IVA} workflow works for an epidemiological use case.
+The following section describes how our presented \emph{IVA} workflow is used for an epidemiological use case.
 \section{Application}
@@
 %
-To our knowledge, this is the first approach, where shape-related information of the whole lumbar spine is analyzed along with other epidemiological features.
+To our knowledge, this is the first approach where shape-related information of the whole lumbar spine is analyzed along with other epidemiological features.
 %
@@
 %
-We obtained and work with 983 tetrahedron models of the lumbar spine.
+We obtained and worked with 983 tetrahedron models of the lumbar spine.
 %
@@
 %
-This information is used to calculate mean-shapes as described in Section~\ref{Interaction- and Visualization Techniques}.
+This information is used to calculate mean shapes as described in Section~\ref{Interaction- and Visualization Techniques}.
 
@@
 %
-For dichotomous variables, the color codes distances between mean shapes of the two groups, for variables with more than two manifestation it encodes the distance to the global mean shape of all subjects (see Figure~\ref{fig:hypothesisfree}).
+For dichotomous variables, the color codes distances between mean shapes of the two groups, for variables with more than two manifestations it encodes the distance to the global mean shape of all subjects (see Figure~\ref{fig:hypothesisfree}).
 \\\\
-Shape-based clustering is carried out via Agglomerative Hierarchical clustering of the spine canal centerlines, (recall Section~\ref{application:Data Preprocessing} and \cite{Klemm2013VMV}).
+Shape-based clustering is carried out via agglomerative hierarchical clustering of the spine canal centerlines, (recall Section~\ref{application:Data Preprocessing} and \cite{Klemm2013VMV}).
 %
@@
 The bar chart height indicates the number of subjects in the cluster.
-The difference to the mean-shape is color-coded where blue represents no difference and red a large difference.
+The difference to the mean shape is color-coded where blue represents no difference and red a large difference.
 (b) A pivot table relates cluster results to gender and employment status.
@@
  %
- Mean-shapes between the groups show no difference.
+ Mean shapes between the groups show no difference.
  %
@@
 %
-Either there are no differences or the variance information was lost in the mean-shape calculation.
+Either there are no differences or the variance information was lost in the mean shape calculation.
 %
@@
 %
-The explorative nature of the methods work well for gathering associations possibly acting as confounder, as outcome of a disease or as an actual cause or risk factor.
+The explorative nature of the methods works well for gathering associations possibly acting as confounder, as outcome of a disease or as an actual cause or risk factor.
 %"\emph{Correlation does not imply causation}" is a famous saying in statistics, meaning that correlation does not necessarily imply causal relationships \cite{}.
@@
 % - Calculation of contingency based on the contingency differences - how does the current selection differ compared to the whole data set?
-To reduce the number of false positive findings, the data space can also be randomly be cut in half the hypothesis then can be cross-validated for statistical soundness.
+To reduce the number of false positive findings, the data space can also be randomly cut in half the hypothesis then can be cross-validated for statistical soundness.
 %

### 2014-03-31 draft.tex
Added changes from Steffi

@@
 %(New Solution).
-We propose an Interactive Visual Analysis (\emph{IVA}) approach, that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic features and attributes derived from the image data.
+We propose an Interactive Visual Analysis (\emph{IVA}) approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic features and attributes derived from the image data.
 %
@@
 %(Implications).
-The new framework enables hypotheses validation and -generation by incorporating human pattern recognition as well as data mining methods.
+The new framework enables hypotheses validation and generation by incorporating human pattern recognition as well as data mining methods.
 %
@@
 %
-Also, the introduction of preventive measures in medicine and beyond are based on epidemiological research, where, for example, subgroups with increased risk are identified \cite{Fletcher2012}.
+Also, the introduction of preventive measures in medicine and beyond is based on epidemiological research, where, for example, subgroups with increased risk are identified \cite{Fletcher2012}.
 %
@@
 %
-These are used to determine environmental and lifestyle factors as well as medical examinations results, which may influence a disease.
+These are used to determine environmental and lifestyle factors as well as medical examination results that may influence a disease.
 %
@@
 %
-Semi-automatic techniques more promising but also challenging since the employed modalities, such as Magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
-%
-Analyzing spatial data with respect to other epidemiological factors requires techniques, which reach beyond standard statistical methods.
+Semi-automatic techniques are more promising but also challenging, since the employed modalities, such as magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
+%
+Analyzing spatial data with respect to other epidemiological factors requires techniques that reach beyond standard statistical methods.
 
@@
 %
-The standard workflow lacks methods, which identify correlations the epidemiologists did not consider.
+The standard workflow lacks methods, which identify correlations that the epidemiologists did not consider.
 %The hypothesis driven analysis approach of these data based on expectations of epidemiologists leaves out features which potentially interact with a disease outside.
@@
 \\\\
-We propose an Interactive Visual Analysis (\emph{IVA}) approach \cite{Thomas2005} to provide a way to analyze image- and non-image data.
+We propose an Interactive Visual Analysis (\emph{IVA}) approach \cite{Thomas2005} to provide a way to analyze image and non-image data.
 %
@@
 %
-Intended as an extension to the well established epidemiological tools it provides a way to rapidly validate hypotheses as well as trigger \emph{hypothesis generation} using data mining methods, such as clustering.
+Intended as an extension to the well-established epidemiological tools it provides a way to rapidly validate hypotheses and to trigger \emph{hypothesis generation} using data mining methods, such as clustering.
 %
@@
 %
-Easy exchange of developed methods driven by modern web technologies intends to trigger a fast feedback loop between us and the epidemiologists.
+The easy exchange of developed methods driven by modern web technologies intends to trigger a fast feedback loop between us and the epidemiologists.
 % http://wiki.infowiss.net/Gulf_of_execution ?
@@
 %
-This data set comprises 127 features and 2 sequences of MRI data from 2.333 patients.
+This data set comprises 127 features and 2 sequences of MRI data from 2,333 patients.
 \\\\
@@
  \includegraphics[width=3.0in]{figures/workflow_comparison}
- \caption{\emph{IVA} tools complement parts of the epidemiological workflow, not replace it.
-The appropriate combination of statistical- and interactive driven analysis shows promising potential to unveil the information in the data.
+ \caption{\emph{IVA} tools complement parts of the epidemiological workflow instead of replacing them.
+The appropriate combination of statistical and interactive-driven analysis shows promising potential to unveil the information in the data.
 (a) shows the standard epidemiological workflow, (b) the \emph{IVA} supported one.
@@
 %Epidemiology diversity is reflected in the different experts working at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
-The diversity of epidemiology is reflected in the different experts who work at cohort studies, ranging from specialized doctors, to medical computer scientists with focus on biometrics, and statisticians.
+The diversity of epidemiology is reflected in the different experts who work at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics, and statisticians.
 %Many different experts work at epidemiological studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
@@
 %
-Thus, both intra- and inter observer variability needs to be low for all aspects of a cohort study examination.
+Thus, both intra- and inter-observer variability needs to be low for all aspects of a cohort study examination.
 %
@@
 %
-Age for example is divided into groups (e.g. in 20-year-steps) when investigating its influence on a condition.
+Age for example is divided into groups (e.g. in 20 year steps) when investigating its influence on a condition.
 %
@@
 
-To determine, whether a subject is prone to be affected by a certain disease, \emph{relative risks} are expressed through the evaluation of p-values, which indicate statistical significance.
-%
-Statistical correlations are prone to \emph{confounding}, meaning that the association of two features are influenced by a third feature, which needs to be isolated.
-%
-A famous example for a confounder is the correlation observable between shoe size and mortality where it can be observed that people with larger shoes have a smaller life expectation than people with small shoe size.
-%
-Shoe size actually is associated with gender where women have a smaller feet than men but have a longer life expectation.
+To determine whether a subject is prone to be affected by a certain disease, \emph{relative risks} are expressed through the evaluation of p-values, which indicate statistical significance.
+%
+Statistical correlations are prone to \emph{confounding}, meaning that the association of two features is influenced by a third feature, which needs to be isolated.
+%
+A famous example for a confounder is the correlation observable between shoe size and mortality, where it can be observed that people with larger shoe size have a smaller life expectation than people with small shoe size.
+%
+The shoe size is actually associated with gender, where women have smaller feet than men but also a longer life expectation.
 % Statistical correlations are prone to \emph{confounding}, meaning that two features are dependent and therefore should be normalized with respect to each other.
@@
 %
-The most obvious example is women-specific questions about menstrual status or number of born children.
-%
-Follow-up examinations or -questions for conditions like medications taken after a diagnosed disease also yield features only available for a small amount of subjects.
-%
-
-Indicators for medical conditions as well as questions about a subjects' lifestyle are also often \emph{dichotomous}--they have two manifestations (\emph{Yes} or \emph{No}).
+The most obvious example are women-specific questions about menstrual status or number of born children.
+%
+Follow-up examinations or questions about conditions like medications taken after a diagnosed disease also yield features only available for a small amount of subjects.
+%
+
+Indicators for medical conditions as well as questions about a subject's lifestyle are also often \emph{dichotomous}--they have two manifestations (\emph{Yes} or \emph{No}).
 %
@@
 %MRI data is more expensive to obtain than CT data but does not affect the subjects health and is therefore the main method for collecting cohort study imaging data.
-Therefore is MRI the main method for collecting cohort study imaging data.
+Therefore, MRI is the main method for collecting cohort study imaging data.
 %
@@
 %
-Segmentation masks representing the voxels of an anatomical structure would be ideal since many different key figures, e.g., volume, largest diameter or aspect ratio, can be derived from them.
-%
-Since reliable and efficient segmentation techniques for these data are not available in general, epidemiologists are forced to measure the data by hand, which is a very tedious work with respect to the number of necessary landmarks and number of subjects.
-%
-Information derived by landmarks, such as top and bottom point are by far not as expressive and versatile as segmentation masks describing the whole vertebra.
+Segmentation masks representing the voxels of an anatomical structure would be ideal, since many different key figures, e.g., volume, largest diameter or aspect ratio, can be derived from them.
+%
+Since reliable and efficient segmentation techniques for these data are not available in general, epidemiologists are forced to measure the data by hand, which is a very tedious work with respect to the number of necessary landmarks and the number of subjects.
+%
+Information derived by landmarks, such as top and bottom point, are by far not as expressive and versatile as segmentation masks describing the whole vertebra.
 %Information derived by landmarks such as top and bottom point of a vertebra are also not nearly as expressive and versatile as segmentation masks describing the whole vertebra.
@@
 %
-Morphometric information from landmarks comprises thickness, diameter or length of a structure as well as grey-value distribution in an area (used for determining the type of tissue).
+Morphometric information from landmarks comprises thickness, diameter or length of a structure as well as grey value distribution in an area (used for determining the type of tissue).
 
@@
 \subsection{The Study of Health in Pomerania (SHIP)}
-After the pioneering Rotterdam study (started in 1990) several MR imaging studies initiatives have evolved.
+After the pioneering Rotterdam study (started in 1990), several MR imaging study initiatives have evolved.
 %
@@
 %
-Starting in 1997 with a cohort consisting of 4.308 subjects, the \texttt{SHIP}, located in northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
+Starting in 1997 with a cohort consisting of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
 %
@@
 \\\\
-Their focus on \emph{hypothesis generation} using parallel assessment of multiple data features makes the work of Turkay and colleagues closest to ours albeit our emphasis on processing medical image data and variables with categorical manifestations \cite{Turkay2013}.
+Their focus on \emph{hypothesis generation} using parallel assessment of multiple data features makes the work of Turkay and colleagues closest to ours albeit our emphasis on processing models derived from medical image data segmentation and variables with categorical manifestations \cite{Turkay2013}.
 %
@@
 %\emph{Over-fitting} of hypotheses to the data may be imposed  is a danger pointed out by the authors when hypotheses to the data based on observations of changes in these plots the measure highlights only subsets of statistical changes.
-Hypotheses based on observations of changes in these plots may impose \emph{overfitting} to the data because the measures highlights only subsets of statistical changes.
+Hypotheses based on observations of changes in these plots may impose \emph{overfitting} to the data because the measures highlight only subsets of statistical changes.
 %
@@
 \paragraph{Visualizing Image and Non-Image Data. }
-Gresh and colleagues proposed \texttt{WEAVE}, one of the first systems, which concurrently analyzed medical image and non-image data using linked views \cite{Gresh2000}.
-%
-Blaas and colleagues presented a similar system which analyzed medical image data and variables derived from them using views from the feature- and physical space \cite{Blaas2007}.
-%
-They incorporated data mining methods such as dividing the data space by using a k-nearest-neighbor technique and the PCA.
+Gresh and colleagues proposed \texttt{WEAVE}, one of the first systems which concurrently analyzed medical image and non-image data using linked views \cite{Gresh2000}.
+%
+Blaas and colleagues presented a similar system which analyzed medical image data and variables derived from them using views from the feature and physical space \cite{Blaas2007}.
+%
+They incorporated data mining methods such as dividing the data space by using a k-nearest neighbor technique and the PCA.
 %
@@
 %
-Their understanding of a cohort differs from the understanding of the term in an epidemiological context by denoting every parameter divided subject group as individual cohort.
+Their understanding of a cohort differs from the understanding of the term in an epidemiological context by denoting every parameter-divided subject group as individual cohort.
 
@@
 %From their findings regarding theassociationn of cancer-related socio-demographic factors are drawn in a \emph{Concept Map} where related factors are connected via graph-edges.
-Chui and colleagues visualized associations in time-dependent epidemiological data using time-series plots highlighting risk factors differences in age and gender \cite{Chui2011}.
+Chui and colleagues visualized associations in time-dependent epidemiological data using time-series plots highlighting risk factor differences in age and gender \cite{Chui2011}.
 
 \paragraph{Commercial Data Visualization.}
-Commercial systems such as \texttt{Tableau}\footnote{Owned by Tableau Software; \url{http://tableausoftware.com}} or \texttt{Spotfire}\footnote{Owned by TIBCO; \url{http://spotfire.tibco.com}} provide a rich user interface that enable a Visual Analytics approach without the need of writing any code.
+Commercial systems such as \texttt{Tableau}\footnote{Owned by Tableau Software; \url{http://tableausoftware.com}} or \texttt{Spotfire}\footnote{Owned by TIBCO; \url{http://spotfire.tibco.com}} provide a rich user interface that enables a Visual Analytics approach without the need of writing any code.
 %
@@
 %
-Distance to the mean shape is color-coded.
+The distance to the mean shape is color-coded.
 %Differences between structures are highlighted using color mapping of the difference to the mean shape, but is rather hard to recognize due to small renderings of each subject in the shape-space.
@@
 %
-Hermann and colleagues identify local deformation changes by investigating shape-related difference on rodent mandibles \cite{Hermann2014}.
+Hermann and colleagues identify local deformation changes by investigating shape-related differences on rodent mandibles \cite{Hermann2014}.
 % TODO: Blöde formulierung
@@
 % Prior Work
-\paragraph{SHIP-Data analysis.}
+\paragraph{SHIP Data Analysis.}
 Klemm and colleagues visualized lumbar spine variabilities based on an semi-automatic shape detection algorithm of 490 participants of the \texttt{SHIP-2} cohort \cite{Klemm2013VMV}.
@@
 %
-As proof of concept, a relation between size of the segmented shape and measured size of the subjects was shown.
-%
-This work focuses on incorporating these derived data as new features of the overall data set, making it possible to include it into the hypothesis validation and -generation process.
+As proof of concept, a relation between the size of the segmented shape and measured size of the subjects was shown.
+%
+This work focuses on incorporating these derived data as new features of the overall data set, making it possible to include it into the hypothesis validation and generation process.
 %
@@
 %
-Niemann and colleagues presented a interactive data mining tool for assessment of risk factors on hepatic steatosis, the fatty liver disease \cite{Niemann2014}.
+Niemann and colleagues presented an interactive data mining tool for the assessment of risk factors on hepatic steatosis, the fatty liver disease \cite{Niemann2014}.
 %
@@
 %--except the variables derived the segmentation model itself.
-While we take similar steps when analyzing the data such as employing statistical tests, our data are not associated in a geographical or temporal context, which affects the techniques described in the following section.
+While we take similar steps when analyzing the data, such as employing statistical tests, our data are not associated in a geographical or temporal context, which affects the techniques described in the following section.
 
@@
 %
-Figure~\ref{fig:WorkflowComparison} (a) shows this workflow as consecutive series of steps.
-%
-Introducing the \emph{IVA} principle to the epidemiological application domain does aim to compensate its weaknesses rather than replacing the existing workflow.
+Figure~\ref{fig:WorkflowComparison} (a) shows this workflow as a consecutive series of steps.
+%
+Introducing the \emph{IVA} principle to the epidemiological application domain aims to compensate its weaknesses rather than replacing the existing workflow.
 %The workflow we propose by introducing the \emph{IVA} principle into the epidemiological application domain does not aim at replacing the existing workflow but to complement its weaknesses.
@@
 %
-Statistical tests on features associated to a hypothesis yields a value deciding whether the data supports the hypothesis or not.
+Statistical tests on features associated to a hypothesis yield a value for deciding whether the data supports the hypothesis or not.
 %A list of features describing the hypothesis is compiled and analyzed using statistical tests. 
@@
 %
-It is easily possible that information left out would heavily influence the result.
+It is easily possible that left out information would heavily influence the result.
 %Not only does this leave out the majority of the information in the medical image data by abstracting it to single values, it is easily possible that information left out would heavily influence the result.
@@
 %
-This has to be handled with care since \emph{overfitting} of expectations to the data is an imminent danger \cite{Turkay2013}.
+This has to be handled with care, since \emph{overfitting} of expectations to the data is an imminent danger \cite{Turkay2013}.
 
@@
 %
-Two types of dependent variables then exist: 
+Then, two types of dependent variables exist: 
 \begin{enumerate}
@@
 %
-Clustering algorithms can be used to investigate associations between shape-groups and other non-image-based variables.
-%
-Analysis of outliers can indicate segmentation errors or an group of subjects sharing a pathology.
+Clustering algorithms can be used to investigate associations between shape groups and other non-image-based variables.
+%
+Analysis of outliers can indicate segmentation errors or a group of subjects sharing a pathology.
 
@@
 
-\paragraph{Multivariate Analysis:} refers to an investigation of multi-variate properties of the dependent data by specifying a feature in one attribute view and at the same time analyzing the values distribution with respect to other variables in other attribute views.
+\paragraph{Multivariate Analysis:} refers to an investigation of multi-variate properties of the dependent data by specifying a feature in one attribute view and at the same time analyzing the value distribution with respect to other variables in other attribute views.
 %
@@
 %
-These associations may also summarized using pivot tables which are popular in epidemiology.
+These associations may also be summarized using pivot tables which are popular in epidemiology.
 
@@
 \subsection{Data Preprocessing} \label{Data Preprocessing}
-Transformation operations on the data to prepare it for a \emph{IVA} system are denoted as data preprocessing.
-\paragraph{Non-Image data. }
+Transformation operations on the data to prepare it for an \emph{IVA} system are denoted as data preprocessing.
+\paragraph{Non-Image Data. }
 %To include heterogenous epidemiological data in an \emph{IVA}-framework it is necessary to process it to obtain standardized views to the available features.
@@
 %
-Exporting the data in the respective tool to a CSV file and then convert it to file types that are easily manageable such as JSON or XML makes it readable for all modern programming languages.
+Exporting the data in the respective tool to a CSV file and then converting it to file types that are easily manageable, such as JSON or XML, makes it readable for all modern programming languages.
 %The best solution for us was to simply export the data in the respective tool to a CSV file and then convert it to data types which are easier manageable such as JSON or XML using our own classes.
 %
-This can be achieved using data wrangling tools such as \texttt{OpenRefine}\footnote{Developed by Google, Open Source; \url{http://openrefine.org}}, which also validate the data (find missing data, clean up bad formatting, transform scales).
+This can be achieved by using data wrangling tools such as \texttt{OpenRefine}\footnote{Developed by Google, Open Source; \url{http://openrefine.org}}, which also validate the data (find missing data, clean up bad formatting, transform scales).
 %In order to verify that the conversion worked as expected and the data is valid, it is good practice to use data wrangling tools such as \texttt{OpenRefine} to validate the data (find missing data, clean up bad formatting).
@@
 %
-Missing data are denoted using error codes indicating its cause ranging from ethical to medical and personal issues (recall Subsection~\ref{EpidemiologicalData}).
+Missing data are denoted using error codes indicating their cause ranging from ethical to medical and personal issues (recall Subsection~\ref{EpidemiologicalData}).
 %The reasons for missing data have a wide range from ethical to medical and personal issues.
@@
 %Therefore, these are also included as error codes which have to be marked as such in the data dictionary.
-\paragraph{Image data. }
+\paragraph{Image Data. }
 %Gerade weil der Satz davor so kompliziert war, ist unklar, was "this" bedeutet.
@@
 %
-This is either done manually by experts setting landmarks (sometimes supported by algorithms connecting the land marks such as graph cuts \cite{GraphCut}) or by a (semi)-automatic detection, registration and segmentation.
-%
-These algorithms have to deal with a large inter-subject variability of the anatomical structure and need to produce reproducible results \cite{Preim2014}.
+This is either done manually by experts setting landmarks (sometimes supported by algorithms connecting the landmarks such as graph cuts \cite{GraphCut}) or by a (semi-)automatic detection, registration and segmentation.
+%
+These algorithms have to deal with a large inter-subject variability of the anatomical structure and need to create reproducible results \cite{Preim2014}.
 %
@@
 %Model-based segmentation approaches are based on assumptions related to object shape, size, location and texture.
-Model-based approaches have shown in principle to be effective for detection \cite{Rak2013} and segmentation \cite{Gloger2010, Gloger2012}.
+In principle, model-based approaches have shown to be effective for detection \cite{Rak2013} and segmentation \cite{Gloger2010, Gloger2012}.
 %
@@
 %
-Grey value comparison is usually used to measure the quantity of fat, water, and--application specific--iron content (liver) or the distribution of grey and white brain tissue.
-
-Morphometric features are derived to allow for statistical comparison of the tissue, which incorporate mostly positions, diameters, volumes and relative distances and alignment to other structures.
+Grey value comparison is usually used to measure the quantity of fat, water, and--application-specific--the iron content (liver) or the distribution of grey and white brain tissue.
+
+Morphometric features are derived to allow for statistical comparison of the tissue, which incorporates mostly positions, diameters, volumes and relative distances and alignment to other structures.
 % 
@@
  %
- Note that adjacent steps are directly connected via feedback loops allowing for iterative refinement and give as much freedom to the user as possible.}
+ Note that adjacent steps are directly connected via feedback loops allowing for an iterative refinement and giving as much freedom to the user as possible.}
  \label{fig:InteractionLoop}
@@
  \includegraphics[width=1\textwidth, resolution=300]{figures/visualization}
- \caption{(Left) Screenshot from the front-end, which is divided as follows: (a) The Sidebar containing all features as well as the groups defined in the analysis process; (b) the canvas area where features can be added via drag and drop and the visualization chosen automatically according to the data type; (c) the interactive pivot table showing the exact numbers for each displayed feature combination.
+ \caption{(Left) Screenshot from the front-end, which is divided as follows: (a) The sidebar containing all features as well as the groups defined in the analysis process; (b) the canvas area where features can be added via drag and drop and the visualization can be chosen automatically according to the data type; (c) the interactive pivot table showing the exact numbers for each displayed feature combination.
 The data displayed is used to analyze the lumbar spine.
@@
 %-------------------------------- /Figure -------------------------------
-Our proposed \emph{IVA} workflow knows three major steps as illustrated in Figure~\ref{fig:InteractionLoop}: Feature selection, -visualization and -brushing.
-%
-An hypothesis-driven analysis usually starts with the selection of features.
+Our proposed \emph{IVA} workflow knows three major steps as illustrated in Figure~\ref{fig:InteractionLoop}: Feature selection, visualization and brushing.
+%
+A hypothesis-driven analysis usually starts with the selection of features.
 %
@@
 %
-\emph{Hypothesis-generation} with focus on image data starts with a shape-based clustering.
-%
-The feature is mapped using a automatically chosen visualization appropriate for its data type (described in detail in the following section).
+\emph{Hypothesis generation} with focus on image data starts with a shape-based clustering.
+%
+The feature is mapped using an automatically chosen visualization appropriate for its data type (described in detail in the following section).
 %
@@
 %
-Subsequently the visualization can be either brushed or new features can be added to the analysis.
+In our system, the visualization can either be brushed or new features can be added to the analysis. 
+%Subsequently the visualization can be either brushed or new features can be added to the analysis.
 %
@@
 %
-A sample workflow using interaction- and visualization techniques described in the next section can be seen in Figure~\ref{fig:visualization}.
-
-\subsection{Interaction- and Visualization Techniques} \label{Interaction- and Visualization Techniques}
+A sample workflow using interaction and visualization techniques described in the next section can be seen in Figure~\ref{fig:visualization}.
+
+\subsection{Interaction and Visualization Techniques} \label{Interaction- and Visualization Techniques}
 %We employ the fourth and highest level which includes next to brushing and linking, advanced brushing, extraction of attributes the development of visualizations custom tailored to the data sets.
@@
 %
-The suitability of an interaction- and visualization technique for epidemiological data depends on its ability to intuitively compare multiple data features at once while highlighting new associations.
+The suitability of an interaction and visualization technique for epidemiological data depends on its ability to intuitively compare multiple data features at once while highlighting new associations.
 %
@@
 %
-Visual evaluation of data are therefore as important as methods allowing for numerical data analysis.
+Visual evaluations of data are therefore as important as methods allowing for numerical data analysis.
 %
@@
  \includegraphics[width=3.5in]{figures/similarity_matrix}
- \caption{Adjacency matrix of 131 features with a grand total of 17.161 combinations.
- %
- Similarity is calculated using the \emph{Cram\'{e}rs' V} contingency value.
+ \caption{Adjacency matrix of 131 features with a grand total of 17,161 combinations.
+ %
+ Similarity is calculated using the \emph{Cram\'{e}r's V} contingency value.
  %
@@
 \subsubsection{System Structure} \label{Structure and Workflow}
-We divide the workspace into four major parts as seen in Figure~\ref{fig:visualization} and \ref{fig:similarity}.
+We divide the workspace into four major parts, as illustrated in Figure~\ref{fig:visualization} and \ref{fig:similarity}.
 \begin{itemize}
@@
 \paragraph{Sidebar.}
-An overview of all features is presented in a sidebar where they are categorized to different types, such as somatometric (measurements of the dimensions of the human body), disease- or lifestyle-related, pain indicators and laboratory data.
-%
-It also contains subject groups either defined by brushing or automated shape-clustering.
+An overview of all features is presented in a sidebar where they are categorized into different types, such as somatometric (measurements of the dimensions of the human body), disease- or lifestyle-related, pain indicators and laboratory data.
+%
+It also contains subject groups either defined by brushing or automated shape clustering.
 %
@@
 %
-Categorical data are mapped either using bar charts (single features) or mosaic plots (multiple features).
+Categorical data are either mapped using bar charts (single features) or mosaic plots (multiple features).
 %
 Continuous data are discretized and therefore treated as categorical data.
+%
+Too many groups potentially generate sparse bins not suited for statistical evaluation. 
+%
+Not enough groups overgeneralize information.
+%
+Adaptive discretization is an option, but imposes overfitting to the data.
+%
+Drawn conclusions based on statistical relationships derived from groups already biased by feature distribution are heavily influenced by the used discretization.
 %
@@
 % TODO Implement MOSAIC PLOTS, Scatterplots, Parallel Coordinates
-Following Tuftes concept of \emph{small multiples} \cite{Tufte1983}, the medical image data are directly incorporated into the plot by including color-coded mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
+Following Tufte's concept of \emph{small multiples} \cite{Tufte1983}, the medical image data are directly incorporated into the plot by including color-coded mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
 %If possible the medical image data is directly included into the plot as well by including mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
@@
 %
-Thus we decided to introduce an interactive pivot table.
+Thus, we decided to introduce an interactive pivot table.
 %When opening a random epidemiological paper the reader will in almost all cases find some sort of pivot table to present the data.
@@
 % TODO: Has to be triggered since visualization is very large
-As discussed previously, highlighting potentially interesting values in the data set is one major benefit of the \emph{IVA} powered approach and belongs to the \emph{multivariate analysis} pattern,  analyzing features outside the shape-space.
+As previously discussed, highlighting potentially interesting values in the data set is one major benefit of the \emph{IVA}-powered approach and belongs to the \emph{multivariate analysis} pattern,  analyzing features outside the shape space.
 %
@@
 %
-The \emph{Cram\'{e}rs' V} contingency coefficient can be used to calculate coherences between categorical variables \cite{CramerV}.
+The \emph{Cram\'{e}r's V} contingency coefficient can be used to calculate coherences between categorical variables \cite{CramerV}.
 %
@@
 %
-\emph{Cram\'{e}rs' V} is defined as:
+\emph{Cram\'{e}r's V} is defined as:
 \begin{equation}
@@
 \end{equation}
-where $X^2$ equals \emph{Pearson's chi squared}, $N$ is the total number of observations and $k$ either the row or column count, depending on which one is lower.
-%
-$V$ assumes values between $0$, meaning two variables are completely independent, and $1$, indicating they are the same.
-%
-\emph{Cram\'{e}rs' V} is always positive and does therefore not allow statements about dependency direction.
+where $X^2$ equals \emph{Pearson's chi squared}, $N$ is the total number of observations and $k$ is either the row or column count, depending on which one is lower.
+%
+$V$ assumes values between $0$, meaning that two variables are completely independent, and $1$ indicating they are the same.
+%
+\emph{Cram\'{e}r's V} is always positive and does therefore not allow statements about dependency direction.
 
@@
 %
-The expected counts in the contingency table have to be larger than 5 for $80\%$ of the entries and no expected value must be smaller than one \cite{Cochran1952}!
+The expected counts in the contingency table have to be larger than 5 for $80\%$ of the entries and no expected value must be smaller than one \cite{Cochran1952}.
 %
@@
 %
-They cannot be included into epidemiological analysis as well since statistical validation needs a minimum count to be seen as valid.
+They cannot be included into the epidemiological analysis, since statistical validation needs a minimum count to be seen as valid.
 %This makes them not assessable for epidemiological analysis as well since statistical validation needs a minimum count to be seen as valid.
@@
 %
-The distinction, whether an association is a confounder or an effect, depends on the context defined by the hypothesis and is a decision to be made by the domain expert.
+The distinction whether an association is a confounder or an effect, depends on the context defined by the hypothesis and is a decision to be made by the domain expert.
 \\\\
-In the following section we discuss how we implemented the presented methods using free on open web standards.
+In the following section we discuss how we implemented the presented methods using free, open web standards.
 % \subsubsection{Automated Feature Suggestion}
@@
  \includegraphics[width=3.5in]{figures/technologies}
- \caption{The front-end solution (left) uses state of the art web technologies such as \emph{HTML5}/\emph{CSS3}, \emph{WebGL} and \emph{SVG} to display the data.
- % 
- The \emph{NodeJS} based back-end (right) stores all image- and non-image data and transfers it to connected clients.
- %
- All computation-heavy operations, such as calculation of mean shapes or -distances, as well as statistical processing is done by the server to keep hardware requirements of client systems low. 
+ \caption{The front-end solution (left) uses state-of-the-art web technologies such as \emph{HTML5}/\emph{CSS3}, \emph{WebGL} and \emph{SVG} to display the data.
+ % 
+ The \emph{NodeJS} based back-end (right) stores all image and non-image data and transfers it to connected clients.
+ %
+ All computation-heavy operations, such as calculation of mean shapes or distances, as well as statistical processing are done by the server to keep hardware requirements of client systems low. 
  %
@@
 %	\item No additional software needs to be installed, most people use a decent state-of-the-art web browser, even on mobile devices.
-	\item The client-server structure allows for employing heavy computation on a capable server machine and transfer results to the client.
-	\item Since image data for several thousand subjects requires hundreds of gigabytes disk space it can remain safely on the server and elements can be transferred on demand.
-	%
-	High confidentiality standards of the data can be met by restricting access via an account system
-	\item Recent developments in \emph{WebGL} applications running in browsers with near native performance push the development further into to the web resulting in many open source libraries, which are well documented, rich in examples and driven by active communities.
+	\item The client-server structure allows for employing heavy computation on a capable server machine and transfering results to the client.
+	\item Since image data for several thousand subjects requires hundreds of gigabytes disk space, it can remain safely on the server and elements can be transferred on demand.
+	%
+	High confidentiality standards of the data can be met by restricting access via an account system.
+	\item Recent developments in \emph{WebGL} applications running in browsers with near-native performance push the development further into to the web resulting in many open source libraries, which are well documented, rich in examples and driven by active communities.
 We use \emph{WebGL} for rendering shape information.
@@
 %
-The server performs heavy statistical tasks such as calculation of \emph{Cram\'{e}rs' V} values for all feature combinations in order to keep the computation time on the client as low as possible.
+The server performs heavy statistical tasks, such as calculation of \emph{Cram\'{e}r's V} values for all feature combinations in order to keep the computation time on the client as low as possible.
 %
@@
 %
-Since our clustering algorithms are written in \texttt{MatLab}\footnote{Owned by The MathWorks, \url{http://mathworks.com}}, we hat to access them using the \texttt{NodeJS} server.
-%
-We accomplished this by converting it to a parameterized standalone console application, that is spawned by \texttt{NodeJS} on client request and then reads the result from the console standard-out and returns it in a proper format to the client.
+Since our clustering algorithms are written in \texttt{MatLab}\footnote{Owned by The MathWorks, \url{http://mathworks.com}}, we had to access them using the \texttt{NodeJS} server.
+%
+We accomplished this by converting it to a parameterized standalone console application that is spawned by \texttt{NodeJS} on client request and then reads the result from the console standard out and returns it in a proper format to the client.
 %
@@
 %
-It is one of the most common reasons for an adult to see a physician in the western civilization \cite{Backpain}.
+This is one of the most common reasons for an adult to see a physician in the Western civilization \cite{Backpain}.
 %
@@
 %
-To our knowledge, this is the first approach where shape-related information of the whole lumbar spine is analyzed along with other epidemiological features.
+To our knowledge, this is the first approach, where shape-related information of the whole lumbar spine is analyzed along with other epidemiological features.
 %
@@
 %
-Characterizing the healthy aging process of the spine is a long term goal for determining age-normalized probabilities for spine-related diseases by incorporating individual risk factors.
+Characterizing the healthy aging process of the spine is a long-term goal for determining age-normalized probabilities for spine-related diseases by incorporating individual risk factors.
 
@@
 %
-There are 136 features describing diagnosed diseases, lifestyle factors, women-specific factors, pain indicators, laboratory values, somatometric features. %and are ordered accordingly.
+There are 136 features describing diagnosed diseases, lifestyle factors, women-specific factors, pain indicators, laboratory values and somatometric features. %and are ordered accordingly.
 %
@@
 %The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence (676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
-The spine protocol consisted of sagittal T1- and T2-weighted turbo-spin-echo sequences with rather large slice distance only just sufficient for characterizing the lumbar spine \cite{Hegenscheid2013}.
+The spine protocol consisted of sagittal T1- and T2-weighted turbo-spin-echo sequences with rather large slice distance only sufficient for characterizing the lumbar spine \cite{Hegenscheid2013}.
 
@@
 	\item a detailed description of the feature, and
-	\item the data dictionary translating value- or error IDs to the actual values.
+	\item the data dictionary translating value or error IDs to the actual values.
 \end{itemize}
@@
 % Binning: https://www.researchgate.net/post/What_metric_best_captures_the_information_loss_when_binning_a_large_data-set_into_a_histogram_representation
-Continuous variables are discretized to allow for \emph{Cram\'{e}rs' V} contingency coefficient assessment.
+Continuous variables are discretized to allow for \emph{Cram\'{e}r's V} contingency coefficient assessment.
 %
@@
 Since the number of categories often strongly depends on the hypothesis, the discretization steps can be adapted dynamically.
-%
-To allow for hypothesis generation, we set the number of groups to five if not specified otherwise.
-% TODO Kai: Das mussn noch geklärt werden
+% TODO: Bernhard, reicht es hier die Zahl 5 zu nennen oder muss das auch schon bei der adaptiven Visualisierung passieren?
+In consultation with the collaborating epidemiologists, we set the number of groups to five, to allow for contingency assessment.
 
@@
 %
-This semi-automatic method requires the user to initialize the Tetrahedron-based finite element models (FEM) with a click on the L3-vertebra.
-%
-Two user-defined landmarks on the top and bottom of the L3-vertebra are used to obtain an initial height estimation of the model.
+This semi-automatic method requires the user to initialize the tetrahedron-based finite element models (FEM) with a click on the L3 vertebra.
+%
+Two user-defined landmarks on the top and bottom of the L3 vertebra are used to obtain an initial height estimation of the model.
 %Two more user- defined landmarks are the top and bottom of the L3-vertebra, which transforms size of the model accordingly to match the L3-vertebra .
@@
 %
-The registered models capture resilient information about shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae \cite{Klemm2013VMV}.
-%
-Due to incorrect initialization, strongly deformed spines, contrast-differences and artifacts, the model was not able to detect lumbar spines for all subjects.
+The registered models capture resilient information about the shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae \cite{Klemm2013VMV}.
+%
+Due to incorrect initialization, strongly deformed spines, contrast differences and artifacts, the model was not able to detect lumbar spines for all subjects.
 %
@@
 %
-This allows for calculation of shape-distance and similarity.
+This allows for calculation of shape distance and similarity.
 %
@@
 \\\\
-Shape-based clustering is carried out via agglomerative hierarchical clustering of the spine canal centerlines, (recall Section~\ref{application:Data Preprocessing} and \cite{Klemm2013VMV}).
-%
-Since the "correct" number of clusters in a given group is unknown, an estimate is computed by means of the knee/elbow method \cite{Salvador2004}.
+Shape-based clustering is carried out via agglomerative hierarchical clustering of the spine canal centerlines (recall Section~\ref{application:Data Preprocessing} and \cite{Klemm2013VMV}).
+%
+Since the ``correct'' number of clusters in a given group is unknown, an estimate is computed by means of the knee/elbow method \cite{Salvador2004}.
 %Since it is not possible to determine the number of clusters in a given group, it is automatically computed using a \emph{knee/elbow} point which is described as tradeoff between number of cluster and a cluster evaluation metric \cite{Salvador2004}.
@@
 The bar chart height indicates the number of subjects in the cluster.
-The difference to the mean shape is color-coded where blue represents no difference and red a large difference.
+The difference to the mean shape is color-coded, whereas blue represents no difference and red a large difference.
 (b) A pivot table relates cluster results to gender and employment status.
-Cluster 8 is striking because of the high proportion of women, while Cluster 2, 3, 4 contain almost only men.}
+Cluster 8 is striking because of the high proportion of women, while cluster 2, 3, 4 contain almost only men.}
  \label{fig:hypothesisfree}
@@
 %
-Analyzing the data set without a prior hypothesis requires a starting point, giving an overview over the data first \cite{Shneiderman1996}.
+Analyzing the data set without prior hypothesis requires a starting point giving an overview over the data first \cite{Shneiderman1996}.
 %
 %Shape-based hypothesis-free exploration starts with a shape-grouping step achieved by shape-based clustering.
-Shape-based hypothesis-free exploration starts with the \emph{local investigation} pattern by a shape-grouping step using shape-based clustering.
+Shape-based hypothesis-free exploration starts with the \emph{local investigation} pattern by a shape grouping step using shape-based clustering.
 %
@@
 %
-Other shapes differ with respect to size, such as Cluster 2, 3, 8, 10 where the last one also represents a more more straight spine, which is usual for subjects with larger body size.
+Other shapes differ with respect to size, such as cluster 2, 3, 8, 10 where the last one also represents a more straight spine, which is usual for subjects with larger body size.
 %
@@
 %
-Performing a \emph{multivariate analysis} by looking at the \emph{Cram\'{e}rs' V} contingency values of Cluster~8 reveals associations of this group with employment status, gender, body size, age, thyroid nodules and blood-fat value.
-%
-Another \emph{multivariate analysis} using a pivot table set gender and employment status in relation to clustering affiliation, showing that Cluster~8 contains mostly women and also has a larger unemployment rate (see Figure~\ref{fig:hypothesisfree}), while the overall employment rate of women and men in the data set is almost exactly 50\%.
-%
-While all observed features seem as plausible associations related to back pain, the values indicate that it contains subjects with chronic back pain radiating to the legs.
-%
-Metabolic parameters, such as blood-fat and blood-sugar, are also possibly associated features.
+Performing a \emph{multivariate analysis} by looking at \emph{Cram\'{e}r's V} contingency values of cluster~8 reveals associations of this group with employment status, gender, body size, age, thyroid nodules and blood fat value.
+%
+Another \emph{multivariate analysis} using a pivot table set gender and employment status in relation to clustering affiliation, showing that cluster~8 contains mostly women and also has a larger unemployment rate (see Figure~\ref{fig:hypothesisfree}), while the overall employment rate of women and men in the data set is almost exactly 50\%.
+%
+While all observed features seem to be plausible associations related to back pain, the values indicate that it contains subjects with chronic back pain radiating to the legs.
+%
+Metabolic parameters, such as blood fat and blood sugar, are also possibly associated features.
 % TODO: Mit Bernhard klären
-Employment status is a feature relating to many different lifestyle factors such as income or nutrition as well as age and might act as confounder.
+The employment status is a feature relating to many different lifestyle factors such as income or nutrition as well as age and might act as confounder.
 %TODO Implement Color coding only in a certain direction!\\
@@
 %
-This filtering step based on expert experience rules out possible confounder and useless associations and is therefore arguable, whether this represents a disadvantage \cite{Wiley2008}.
+This filtering step based on expert experience rules out possible confounder and useless associations and it is therefore arguable whether this represents a disadvantage \cite{Wiley2008}.
 %It is arguable whether this first filtering step, which is purely based on expert experience, represents a disadvantage, since it rules out many parameters possibly associated with the presented features, but the value of this knowledge would be small \cite{Wiley2008}.
@@
  \includegraphics[width=3.5in]{figures/hypothesisbased}
- \caption{(a) Dichotomous questionnaire answer to "\emph{Did you experience back pain in the past three month?}".
+ \caption{(a) Dichotomous questionnaire answer to ``emph{Did you experience back pain in the past three months?}''.
  %
@@
  %
- (b) Shape-based clustering for all subjects who suffered from back-pain yields 6 groups. 
+ (b) Shape-based clustering for all subjects who suffered from back pain yields 6 groups. 
  %
@@
 %
-If the user proposed a hypothesis about a relation between a non-image feature regarding shape, the workflow slightly differs from the hypothesis-free analysis.
-%
-The starting point follows the \emph{feature investigation} pattern, where a feature of interest is selected by dragging it into the canvas area and view the subjects distribution as well as their shape differences.
-%
-In our use case, epidemiologists were interested in the dichotomous questionnaire answer to "\emph{Did you experience back pain in the past three month?}".
+If the user proposes a hypothesis about a relation between a non-image feature regarding shape, the workflow slightly differs from the hypothesis-free analysis.
+%
+The starting point follows the \emph{feature investigation} pattern, where a feature of interest is selected by dragging it into the canvas area and viewing the subject's distribution as well as their shape differences.
+%
+In our use case, epidemiologists were interested in the dichotomous questionnaire answer to ``\emph{Did you experience back pain in the past three months?}''.
 %
@@
 %
-Since the focus is on subjects suffering from back pain, the clustering result of these subjects are then drawn into the canvas area, yielding six clusters as seen in Figure~\ref{fig:hypopthesisbased} (b).
+Since the focus is on subjects suffering from back pain, the clustering results of these subjects are then drawn into the canvas area, yielding six clusters as seen in Figure~\ref{fig:hypopthesisbased} (b).
 %
@@
 
-\emph{Multivariate analysis} using \emph{Cram\'{e}rs' V} contingency values highlighted relationships of this cluster with joint degeneration, meat eating habits, preoccupation, back pain, neck or shoulder pain and waist circumference.
+\emph{Multivariate analysis} using \emph{Cram\'{e}r's V} contingency values highlighted relationships of this cluster with joint degeneration, meat eating habits, preoccupation, back pain, neck or shoulder pain and waist circumference.
 %
@@
 % TODO Bernhard fragen, was ihn an der Reihenfolge stört
-It is well known that overweight is an risk factor for back pain.
-%
-While the BMI is a key figure for assessing height and weight of a subject, it does not tell anything on how the weight is distributed in the body.
+It is well known that overweight is a risk factor for back pain.
+%
+While the BMI is a key figure for assessing height and weight of a subject, it does not tell anything about how the weight is distributed in the body.
 %
@@
 %
-Our finding follows the recent trends indicating that BMI is not a good measure for assessing body-shape since healthy weight is dependent on many other measures \cite{Ahima2013}.
-%
-It indicates that waist circumference rather than the BMI interacts with unusually shaped spines for subjects with lumbar back pain.
+Our finding follows the recent trends indicating that BMI is not a good measure for assessing the body shape, since healthy weight is dependent on many other measures \cite{Ahima2013}.
+%
+It indicates that the waist circumference rather than the BMI interacts with unusually shaped spines for subjects with lumbar back pain.
 %
@@
 %
-In particular, when large amounts of attributes are automatically checked for strong correlations, just by chance hundreds of them reach at least a moderately significant correlation, such as $p>0.05$
+In particular, when large amounts of attributes are automatically checked for strong correlations, just by chance hundreds of them reach at least a moderately significant correlation, such as $p>0.05$.
 %Defining a causal relationship solely based on observed correlations of two features is \emph{cum hoc ergo propter hoc}--correlation does not automatically imply causation \cite{Tufte2003}.
 %
-The observed correlations need to be carefully checked for confounders and medical soundness!
+The observed correlations need to be carefully checked for confounders and medical soundness.
 %
@@
 %
-To increase the probability of the observation not to be random and to avoid \emph{overfitting}, they are cross checked for associations in \texttt{SHIP-TREND} as a second, independent population sample, which was examined with methods identical to \texttt{SHIP}.
+To increase the probability of the observation not to be random and to avoid \emph{overfitting}, they are cross-checked for associations in \texttt{SHIP-TREND} as a second, independent population sample, which was examined with methods identical to \texttt{SHIP}.
 %If features are identified as interesting for a disease, the epidemiologists look for associations in another cohort, the \texttt{SHIP-TREND}.
@@
 %
-The presented methods guide attention to features not in the focus of attention and expectance of clinical researchers.
-%
-The explorative nature of the methods works well for gathering associations possibly acting as confounder, as outcome of a disease or as an actual cause or risk factor.
+The presented methods guide the attention to features that are not in the focus of attention and expectance of clinical researchers.
+%
+The explorative nature of the methods work well for gathering associations possibly acting as confounder, as outcome of a disease or as an actual cause or risk factor.
 %"\emph{Correlation does not imply causation}" is a famous saying in statistics, meaning that correlation does not necessarily imply causal relationships \cite{}.
@@
 %
-The combination of multiple views with shape information help to connect many different information sources to make the large information spaces cognitive feasible.
+The combination of multiple views with shape information helps to connect many different information sources to make the large information spaces cognitively feasible.
 %
@@
 %Seen as an extension to the existing statistics driven workflow, image-centric analysis of epidemiological data from a IVA approach by illuminating the data-black box.
-The presented \emph{IVA} approach is classified as extension to the existing statistics driven epidemiological.
-%
-Image-centric epidemiological analysis benefits from a IVA approach by illuminating the data-black box.
-%
-Visualization of multivariate data using connected views and different views allows to get fast visual feedback about subject groups.
+The presented \emph{IVA} approach is classified as extension to the existing statistics-driven epidemiological approach.
+%
+Image-centric epidemiological analysis benefits from an IVA approach by illuminating the data-black box.
+%
+The Visualization of multivariate data using connected views and different views allows to get fast visual feedback about subject groups.
 %
@@
 %
-The use of pivot tables is familiar to epidemiologists while embracing the power of interactive adjustment of the features shown.
-%
-Automatic suggestion of correlations using contingency methods like \emph{Cram\'{e}rs' V} trigger \emph{hypothesis generation} by highlighting features potentially overlooked by the experts.
-%
-Shape-based clustering assesses variability of a anatomical structure in the context of non-image features such as disease indicators or lifestyle factors.
-%
-The iterative nature of the \emph{IVA} approach reflects versatility as both \emph{hypothesis-driven analysis} and \emph{hypothesis generation} is allowed.
+The use of pivot tables is familiar to epidemiologists while embracing the power of interactive adjustment of the shown features.
+%
+The automatic suggestion of correlations using contingency methods like \emph{Cram\'{e}r's V} triggers \emph{hypothesis generation} by highlighting features potentially overlooked by the experts.
+%
+Shape-based clustering assesses the variability of a anatomical structure in the context of non-image features such as disease indicators or lifestyle factors.
+%
+The iterative nature of the \emph{IVA} approach reflects versatility, as both \emph{hypothesis-driven analysis} and \emph{hypothesis generation} are allowed.
 \\\\\
@@
 \begin{itemize}
-	\item dynamic discretization of continuous variables to better fit the data distributions
+%	\item dynamic discretization of continuous variables to better fit the data distributions
 	\item shape brushing methods to intuitively query subjects using image data
-	\item incorporate more statistical methods and views that are familiar to the epidemiologists (odds ratios, box plots)
-	\item calculation of contingency differences for selected sub groups--highlight how feature correlation change for a given groups
+	\item incorporating more statistical methods and views that are familiar to the epidemiologists (odds ratios, box plots)
+	\item calculation of contingency differences for selected sub groups--highlighting how the feature correlation changes for a given group
 	\item integration of further visualization techniques, which incorporate both image and non-image data
-	\item support control-groups to check resilience of observations as suggested by Fletcher and Fletcher \cite{Fletcher2012}.
+	\item support of control groups to check the resilience of observations \cite{Fletcher2012}.
+%	\item understand how the existing methods are applied to the data, iteratively adapt to the epidemiologiical workflow
+%	\item investigate which parts of the methods are applied by domain experts and what aspecunderstand how the existing methods are applied to the data, iteratively adapt to the epidemiologiical workflow
+	\item adapt the shape visualization to other capture organ data with different variance type (such as texture of liver or white/gray matter distribution in the brain)
 \end{itemize}
@@
 % - Calculation of contingency based on the contingency differences - how does the current selection differ compared to the whole data set?
-To reduce the number of false positive findings, the data space can also be randomly cut in half the hypothesis then can be cross-validated for statistical soundness.
+To reduce the number of false positive findings, the data space can also be randomly cut in half.
+%
+Then the hypothesis then can be cross-validated for statistical soundness.
 %
@@
 \\\\
-As the number of image-centric cohort studies, participating subjects, gathered features and imaging modalities rises and advances towards comparability between cohort studies are made, the gap between data complexity and analysability increases.
+As the number of image-centric cohort studies, participating subjects, gathered features and imaging modalities rises, and advances towards comparability between cohort studies are made, the gap between data complexity and analyzability increases.
 %
@@
 %
-We believe web technologies pave the way to allow as many experts as possible to analyze this data.
+We believe that web technologies pave the way to allow as many experts as possible to analyze this data.
 %

### 2014-03-31 draft.tex
Final Blind Version

@@
 %% hyperreferences for more convenient online viewing.
-% TODO: Incomplete data: Cramers V kann nur für bestimmte mindestgrößen funktionieren - entweder Daten unvollständig oderz u wneige Probanden in einer Gruppe
 %% Please use one of the ``review'' options in combination with the
@@
 %% OnlineID. Otherwise, you may safely leave it at ``0''.
-\onlineid{0}
+\onlineid{194}
 
@@
 %% indicate IEEE Member or Student Member in form indicated below
-\author{--blind--}
+\author{Submission 194}
 \authorfooter{
@@
 %
-On the other hand, observations made by clinicians in the daily routine are translated into hypotheses for epidemiological research.
+Observations made by clinicians in the daily routine are translated into hypotheses for epidemiological research.
 %
@@
 %
-Methods like regression analysis are used to check the attribute list for statistical soundness.
+Methods like regression analysis are employed to check the attribute list for statistical soundness.
 %
@@
 The standard workflow lacks methods, which identify correlations that the epidemiologists did not consider.
+%
+%Also, statistics tools are limited 
+Also, only a small subset of variables can be concurrently analyzed, depending on the statistical measure.
 %The hypothesis driven analysis approach of these data based on expectations of epidemiologists leaves out features which potentially interact with a disease outside.
@@
 \\\\
-We propose an Interactive Visual Analysis (\emph{IVA}) approach \cite{Thomas2005} to provide a way to analyze image and non-image data.
-%
-Visual queries and direct feedback of Visual Analytics systems allow for a fast exploration of the data space.
+%We propose an Interactive Visual Analysis (\emph{IVA}) approach \cite{Thomas2005} to provide a way to analyze image and non-image data.
+We propose an Interactive Visual Analysis (\emph{IVA}) approach \cite{Thomas2005} for the combined analysis of image and non-image data.
+%
+Visual queries and direct feedback of Visual Analytics systems allow for a fast exploration of the data space incorporating many different variables.
 %
@@
 %	\item providing an overview over the workflow for analyzing cohort study data to gain insight into the large diverse subject spaces.
-	\item an Interactive Visual Analysis workflow for analyzing image-based epidemiological data both hypothesis-driven and for hypothesis-generation based on a characterization of the standard epidemiological workflow,
+	%\item an Interactive Visual Analysis workflow for analyzing image-based epidemiological data both hypothesis-driven and for hypothesis generation--based on a characterization of the standard epidemiological workflow,
+	\item an Interactive Visual Analysis workflow for cohort study data to allow both, hypothesis-driven analysis and hypothesis generation--based on a characterization of the standard epidemiological workflow,
 	%\item an Interactive Visual Analysis workflow for analyzing image-based epidemiological data including both hypothesis-driven and for hypothesis-generation based on a characterization of the standard epidemiological workflow
@@
 % 	\item Provide visualization techniques which combine both information visualization and 3D rendering of organ shapes as well as combining them with well known epidemiological graphics and key figures.
-	\item implementing the presented methods as a web framework based on \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}.
+	\item an implementation of the presented methods in a web framework based on \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}.
 \end{itemize}
@@
 %
-A famous example for a confounder is the correlation observable between shoe size and mortality, where it can be observed that people with larger shoe size have a smaller life expectation than people with small shoe size.
+A famous example for a confounder is the association between shoe size and mortality, where it can be observed that people with larger shoe size have a smaller life expectation than people with small shoe size.
 %
@@
 %
-Information derived by landmarks, such as top and bottom point, are by far not as expressive and versatile as segmentation masks describing the whole vertebra.
+Information derived by landmarks, such as top and bottom point of a vertebra, are by far not as expressive and versatile as segmentation masks describing its whole shape.
+%Information derived by landmarks, such as top and bottom point, are by far not as expressive and versatile as segmentation masks describing the whole vertebra.
 %Information derived by landmarks such as top and bottom point of a vertebra are also not nearly as expressive and versatile as segmentation masks describing the whole vertebra.
@@
 Hermann and colleagues identify local deformation changes by investigating shape-related differences on rodent mandibles \cite{Hermann2014}.
-% TODO: Blöde formulierung
-Specifying a change in shape of a certain area by the user shows how other parts of the model would react using covariance tensors.
+%
+User specified regions of interests are mapped to associated anatomic covariation using tensor visualization.
+%Anatomic covariation induced by user input is displayed using tensors 
+%Specifying a change in shape of a certain area by the user shows how other parts of the model would react using covariance tensors.
 %Covariance tensors indicate how a model is deformed when the user given a user specified deformation.
@@
 %- Acceptance plays a big role - you have to pick up the user by presenting information they are used to in a new way\\
-% TODO: Epidemiologist compiled list ist sehr weit gefasst - bei Begründung ob die Einschränkung der Liste ein Problem darstellt
 We described the epidemiological workflow as a sequence of steps taken by domain experts which need to comprise reproducibility and statistical integrity (recall Subsection~\ref{EpidemiologicalWorkflow}).
@@
 %
-For instance, the spine is segmented over all subjects and all spines are co-registered spanning a joint space.
+For instance, the lumbar spine is segmented over all subjects and all lumbar spines are co-registered spanning a joint space.
 %
@@
 %
-For instance, the epidemiologist selects several spines with a common characteristic in the image or shape space and wishes to inspect the associated dependent variables in an attribute view \cite{Hermann2014}.
+For instance, the epidemiologist selects several lumbar spines with a common characteristic in the image or shape space and wishes to inspect the associated dependent variables in an attribute view \cite{Hermann2014}.
 %
@@
  \includegraphics[width=3.0in]{figures/InteractionLoop}
- % TODO: Recreate Image: Also include Cramér´s V which is part of the visualization
- \caption{Detailed \emph{IVA Loop} as seen in Figure~\ref{fig:WorkflowComparison}.
+ \caption{Detailed \emph{IVA Loop} as extension from Figure~\ref{fig:WorkflowComparison}.
  %
@@
 
-% TODO Einbauen
-% Overfitting bei SHIP-Trend
-% Hypothesis generation bears the chance of \emph{overfitting} the data to expectations.
-% %
-% To avoid this, a timeline needs to be introduced which keeps track how many variable variations the user evaluated before coming to an conclusion.
-% %
-% Since epidemiologists are used to process groups based on table representations we decided to introduce an interactive solution in form of a Pivot Table.
-
 \begin{figure}[htb]
@@
  \includegraphics[width=3.5in]{figures/similarity_matrix}
- \caption{Adjacency matrix of 131 features with a grand total of 17,161 combinations.
+ \caption{Adjacency matrix of 129 features (127 data set variables, 2 cluster results) with a grand total of 16,641 combinations.
  %
@@
 \begin{itemize}
-	\item The sidebar, which contains all epidemiological features. Cluster results are treated like features and are part of the sidebar as well.
-	\item The canvas holding all visualizations. Elements can be added, arranged, resized and removed freely.
-	\item The interactive pivot table gives detailed numerical information of the features in the canvas view. This view on the data comes natural to epidemiologists.
-	\item The contingency view depicting relations for features in the canvas in an adjacency matrix.
+	\item The \emph{sidebar}, which contains all epidemiological features. Cluster results are treated like features and are part of the sidebar as well.
+	\item The \emph{canvas} holding all visualizations. Elements can be added, arranged, resized and removed freely.
+	\item The interactive \emph{pivot table} gives detailed numerical information of the features in the canvas view. This view on the data comes natural to epidemiologists.
+	\item The \emph{contingency view} depicts relations for features in the canvas in an adjacency matrix.
 \end{itemize}
 %
-% TODO Include:
 % The sidebar holds all available variables as well as groups either derived by user input or automatic clustering.
@@
 %
-Continuous data are discretized and therefore treated as categorical data.
+Figure \ref{fig:visualization} describes this dynamic adjustment.
+%
+%Continuous data are discretized and therefore treated as categorical data.
+Continuous data can be visualized using scatterplots (single features) or parallel coordinates (multiple features), but in epidemiology, this data type is usually categorized into ordinal groups of \emph{equal size}.
+%
+Since the number of categories often depends on the hypothesis, the discretization steps can be adapted dynamically.
 %
@@
 %
-It is, however, also possible to plot it using scatterplots (single features) or parallel coordinates (multiple features).
-%
-Figure \ref{fig:visualization} describes this dynamic adjustment.
+Therefore we follow the convention to use bins of equal size.
+
 %Inspired by the previously discussed \texttt{GPloms} \cite{Francois2013} the visualization type is chosen dynamically based on the variable types and number of visualizations which need to be displayed.
@@
 %
-Mapping the distance from a group mean shape to the global mean to color allows to assess local shape changes.
+%Mapping the distance from a group mean shape to the global mean to color allows to assess local shape
+Mapping the distance from a group mean shape to the global mean using color allows to assess local shape changes.
 %%TODO Adaptive visualization!
-If a feature is dropped on an existing plot, the visualization changes dynamically to properly make them comparable. 
+If a feature is dropped on an existing plot, the visualization changes dynamically to properly make them comparable (Figure~\ref{fig:visualization} (right)).
 %
@@
 \paragraph{Automated Feature Suggestion using a Contingency Matrix.}
-% TODO: Has to be triggered since visualization is very large
 As previously discussed, highlighting potentially interesting values in the data set is one major benefit of the \emph{IVA}-powered approach and belongs to the \emph{multivariate analysis} pattern,  analyzing features outside the shape space.
@@
 The distinction whether an association is a confounder or an effect, depends on the context defined by the hypothesis and is a decision to be made by the domain expert.
-\\\\
-In the following section we discuss how we implemented the presented methods using free, open web standards.
 % \subsubsection{Automated Feature Suggestion}
@@
 \end{figure}
+In this section we discuss how we implemented the presented methods using free, open web standards.
 %
@@
 %
-\texttt{WebGL} rendering is carried out by \texttt{Three.js}\footnote{Originally developed by Ricardo Cabello, \url{http://threejs.org}} and allows GPU-accelerated data rendering.
+%\texttt{WebGL} rendering is carried out by \texttt{Three.js}\footnote{Originally developed by Ricardo Cabello, \url{http://threejs.org}} and allows GPU-accelerated data rendering.
+\texttt{Three.js}\footnote{Originally developed by Ricardo Cabello, \url{http://threejs.org}} allows GPU-accelerated data rendering using \texttt{WebGL}.
 %
@@
 %The following section describes how our presented \emph{IVA} workflow works for an epidemiological use case.
-The following section describes how our presented \emph{IVA} workflow is used for an epidemiological use case.
 \section{Application}
+This section describes how our presented \emph{IVA} workflow is used for an epidemiological use case.
+%
 We applied the presented set of techniques to a data set compiled to analyze lower back pain. 
@@
 %
-There are 136 features describing diagnosed diseases, lifestyle factors, women-specific factors, pain indicators, laboratory values and somatometric features. %and are ordered accordingly.
+There are 127 features describing diagnosed diseases, lifestyle factors, women-specific factors, pain indicators, laboratory values and somatometric features. %and are ordered accordingly.
 %
@@
 %
-In epidemiology, continuous data are usually categorized into ordinal groups of equal size.
-%
-Since the number of categories often strongly depends on the hypothesis, the discretization steps can be adapted dynamically.
-% TODO: Bernhard, reicht es hier die Zahl 5 zu nennen oder muss das auch schon bei der adaptiven Visualisierung passieren?
 In consultation with the collaborating epidemiologists, we set the number of groups to five, to allow for contingency assessment.
@@
 %
-While all observed features seem to be plausible associations related to back pain, the values indicate that it contains subjects with chronic back pain radiating to the legs.
+While all observed features seem to be plausible associations related to back pain, the values indicate that cluster~8 contains subjects with chronic back pain radiating to the legs.
 %
@@
 %TODO Implement Color coding only in a certain direction!\\
-%TODO Selection of the variables and display them using the pivot table!
 \\\\
-While this approach does not assume any hypothesis, the 136 epidemiological features were selected by domain experts.
+While this approach does not assume any hypothesis, the 127 epidemiological features were selected by domain experts with a wide range on potential influential factors.
+%While this approach does not assume any hypothesis, the 127 epidemiological features were selected by domain experts and also includes features, which are only remotly unter verdacht to be associating with back pain.
 %the data are met with a selection bias when compiling the list of related features by the domain experts.
-% TODO: In der Einleitung erwähnen, dass man bisher nur maximal 3 Variablen gleichzeitig betrachten kann, damit man unsere Technik etwas in den Kontext rückt
 %
@@
  \includegraphics[width=3.5in]{figures/hypothesisbased}
- \caption{(a) Dichotomous questionnaire answer to ``emph{Did you experience back pain in the past three months?}''.
+ \caption{(a) Dichotomous questionnaire answer to ``\emph{Did you experience back pain in the past three months?}''.
  %
@@
 \section{Summary and Conclusion}
-% TODO: Include IVA Pattern here?
+In this paper, we presented an \emph{IVA} framework for the analysis of complex image-centric epidemiological data.
+%
+Hence, the framework allows for both, hypothesis-driven analysis and hypothesis generation.
 %Seen as an extension to the existing statistics driven workflow, image-centric analysis of epidemiological data from a IVA approach by illuminating the data-black box.
-The presented \emph{IVA} approach is classified as extension to the existing statistics-driven epidemiological approach.
-%
-Image-centric epidemiological analysis benefits from an IVA approach by illuminating the data-black box.
-%
-The Visualization of multivariate data using connected views and different views allows to get fast visual feedback about subject groups.
+%The presented \emph{IVA} approach is classified as extension to the existing statistics-driven epidemiological approach.
+%
+%Image-centric epidemiological analysis benefits from an IVA approach by illuminating the data-black box.
+%
+The visualization of multivariate data using connected views and different views allows to get fast visual feedback about subject groups.
 %
@@
 %
-The iterative nature of the \emph{IVA} approach reflects versatility, as both \emph{hypothesis-driven analysis} and \emph{hypothesis generation} are allowed.
+%The iterative nature of the \emph{IVA} approach reflects versatility, as both \emph{hypothesis-driven analysis} and \emph{hypothesis generation} are allowed.
+
+Our clinical partners from the \texttt{SHIP} are for the first time able to assess shape information of the lumbar spine and its influence to diseases.
+%
+Findings from analyzing lumbar back pain using the \emph{IVA} approach range from deriving shape-based groups of subjects to detailed description of features potentially associated with the disease, such as waist circumference, thyroid nodules and blood fat values.
+%
+Other somatometric features, such as BMI, are not as influential as expected.
+%
 \\\\\
-A number of improvements is left open for future work:
+A number of improvements is left open for future work, e.g.,
 %TODO: Fragen ob die Liste so klar geht mit der Formulierung und ohne Kommas
@@
 %	\item dynamic discretization of continuous variables to better fit the data distributions
-	\item shape brushing methods to intuitively query subjects using image data
-	\item incorporating more statistical methods and views that are familiar to the epidemiologists (odds ratios, box plots)
-	\item calculation of contingency differences for selected sub groups--highlighting how the feature correlation changes for a given group
-	\item integration of further visualization techniques, which incorporate both image and non-image data
-	\item support of control groups to check the resilience of observations \cite{Fletcher2012}.
+	\item shape brushing methods to intuitively query subjects using image data,
+	\item the inclusion of more statistical methods and views that are familiar to the epidemiologists (odds ratios, box plots), or
+	%\item calculation of contingency differences for selected sub groups--highlighting how the feature correlation changes for a given group
+	%\item integration of further visualization techniques, which incorporate both image and non-image data
+	%\item support of control groups to check the resilience of observations \cite{Fletcher2012}.
 %	\item understand how the existing methods are applied to the data, iteratively adapt to the epidemiologiical workflow
 %	\item investigate which parts of the methods are applied by domain experts and what aspecunderstand how the existing methods are applied to the data, iteratively adapt to the epidemiologiical workflow
-	\item adapt the shape visualization to other capture organ data with different variance type (such as texture of liver or white/gray matter distribution in the brain)
+	\item adapt the shape visualization to explore other organ data with different variance type (such as texture of liver or white/gray matter distribution in the brain).
 \end{itemize}
@@
 % - Calculation of contingency based on the contingency differences - how does the current selection differ compared to the whole data set?
-To reduce the number of false positive findings, the data space can also be randomly cut in half.
-%
-Then the hypothesis then can be cross-validated for statistical soundness.
-%
-This requires a large number of subjects, especially if the investigated features are rare and only presented by a few subjects.
+%
+% To reduce the number of false positive findings, the data space can also be randomly cut in half.
+% %
+% Then the hypothesis then can be cross-validated for statistical soundness.
+% %
+% This requires a large number of subjects, especially if the investigated features are rare and only presented by a few subjects.
 %
 %As suggested by Fletcher and Fletcher, we plan to validate the hypothesis using the \texttt{SHIP-TREND} cohort \cite{Fletcher2012}.
-\\\\
+% \\\\
 As the number of image-centric cohort studies, participating subjects, gathered features and imaging modalities rises, and advances towards comparability between cohort studies are made, the gap between data complexity and analyzability increases.
@@
 %
-We believe that web technologies pave the way to allow as many experts as possible to analyze this data.
+%We believe that web technologies pave the way to allow as many experts as possible to analyze this data.
+We believe that web technologies pave the way to analyze this data in a convenient way.
 %
@@
 This work was supported by the DFG Priority Program 1335: Scalable Visual Analytics.}
-
+\clearpage
+\newpage
 \bibliographystyle{abbrv}

### 2014-06-10 draft.tex
Cleaned up comments in the blind submission. Added authors and acknowledgements.

@@
 %% indicate IEEE Member or Student Member in form indicated below
-\author{Submission 194}
+\author{P. Klemm$^{1}$, K. Lawonn$^{1}$, M. Rak$^{1}$, B. Preim$^{1}$, K. Toennies$^{1}$, K. Hegenscheid$^{2}$, H. Völzke$^{2}$, S. Oeltze$^{1}$}
 \authorfooter{
 %% insert punctuation at end of each item
-% \item
-%  Otto-v.-Guericke-University Magdeburg
+\item
+ $^1$Otto-von-Guericke University Magdeburg, Germany.
+\item
+ $^2$Ernst-Moritz-Arndt-University Greifswald, Germany.
 }
@@
 %
-Cohort studies comprise heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors.
-%
-The data are analyzed using a priori defined hypotheses to find statistically significant correlations between features (\emph{associations}).
-%
-Modern cohort studies incorporate medical image data.
-%
-%Analyzing these data requires image segmentation, extraction of key figures, and shape-based subject grouping.
-%
-The statistically driven epidemiological workflow only allows to determine \emph{associations} between image-derived metrics, such as distances extracted from landmarks of the segmentation model.
-\\\\
+Cohort studies comprise heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors and more recently, medical image data.
+%
+%The data are analyzed using a priori defined hypotheses to find statistically significant correlations between features (\emph{associations}).
+%
+%Modern cohort studies incorporate medical image data.
+%
+%
+%The statistically driven epidemiological workflow only allows to determine \emph{associations} between image-derived metrics, such as distances extracted from landmarks of the segmentation model.
+%\\\\
 %(New Solution).
@@
 This is achieved by combining brushing and linking enabled coordinated information visualization views and interactive 3D shape renderings with epidemiological data representations such as pivot tables and key figures as association measures.
-%incorporating coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
-%
-%Multiple coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
-%
 %(Validation).
@@
 Using all reliable information from the image segmentation linked to non-image features aims to unveil \emph{associations} by applying an iterative analysis approach.
-
-%
-%Similarity measures between data variables are used to compute interesting changes in variable associations for the current variable selection.
-%
-%Shape-based grouping of subjects is facilitated using clustering techniques which operate on surface meshes extracted from the image segmentation.
 } % end of abstract
-
-% \abstract{Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease specific risk factors.
-% %
-% Cohort studies comprise heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors.
-% %
-% %Using well established statistical methods the data is hypothesis driven analyzed to find statistically significant variable correlations (\emph{associations}).
-% Driven by a priori defined hypotheses, the data is analyzed using well established statistical methods in order to find statistically significant correlations between variables (\emph{associations}).
-% %
-% Modern cohort studies also incorporate medical image data.
-% %
-% Analyzing these data requires image segmentation, extraction of key figures, and shape-based subject grouping.
-% %
-% The statistically driven epidemiological workflow only allows to determine \emph{associations} between image-derived metrics such as distances extracted from landmarks of the segmentation model.
-% \\\\
-% We propose an Interactive Visual Analysis approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic variables and attributes derived from the image data.
-% %
-% Using all reliable information from the image segmentation linked to non-image variables aims to unveil \emph{associations} by applying an iterative analysis approach.
-% %As extension to the epidemiological workflow it aims to unveil associations based on all reliable shape information rather than abstracting it to a few key figures.
-% %Iterative analysis between ...
-allows for extracting associations which without the visiual analysis driven approach maybe would be overlooked.
-%Einsicht in Interaktionen zwischen nicht-bilddaten und bilddaten auf basis aller verfügbarer Informationen
-% %Iterative Analyse von Organstrukturen im Bezug auf Bilddaten 
-% %Iterative analysis of image-based epidemiological datasets aims to provide insight into associations between organ shape and % in order to provide insights into \emph{associations} by incorporating additional information
-% %It extends the epidemiological workflow with an iterative Analysis process and 
-% It allows for hypothesis validation and hypothesis generation by incorporating human pattern recognition as well as data mining methods.
-% %
-% Multiple coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
-% %
-% Similarity measures between data variables are used to compute interesting changes in variable associations for the current variable selection.
-% %
-% Shape-based grouping of subjects is facilitated using clustering techniques which operate on surface meshes extracted from the image segmentation.
-% } % end of abstract
 
@@
 %
-%These data needs to be segmented to analyze its features for correlations with non-image data.
 Segmentation of these data would allow for anatomical structure analysis and enable correlation with other features.
-%The segmentation of these data would allow to analyze shapes of anatomical structures and to correlate them with other features.
-%These data needs to be concurrently analyzed with the non-image data.
-%
-%This requires their segmentation and the derivation of features such as kidney volume, curvature of the spine or tissue texture of the liver.
-%
-%Manual segmentation is not feasible since it leads to a high variability which is at odds with the high reproducibility required in epidemiology research.
 %
@@
 %
-%Also, statistics tools are limited 
 Also, only a small subset of variables can be concurrently analyzed, depending on the statistical measure.
-%The hypothesis driven analysis approach of these data based on expectations of epidemiologists leaves out features which potentially interact with a disease outside.
-%The hypothesis driven analysis approach of these data leaves out features which potentially interact with a disease because they they are it is outside of the erwartungshaltung of the epidemiologists.
-%leaves out features which potentially interact with a disease because they are usually outside of the practical experience of the epidemiologists.
-%
-%Analyzing these data for correlating features requires preprocessing steps as segmentation of the structures.
-%
-%Manual segmentation is not feasible since it leads to a high variability which is att odds with the high reproducibility required in epidemiology research. 
-%
-%Semi-automatic techniques are more promising but challenging since MRI and ultrasound are subject to inhomogeneity and noise
-%Due to the lack of sophisticated segmentation algorithms this is often achieved by manually specifying landmarks by radiologists which is costly and prone to inter- and intra observer variability.
-%
-% Since it is unethical to expose people to radiation, non-harming imaging such as Magnetic Resonance Imaging (MRI) or Ultrasound Imaging is used.
-% %
-% To quantify these data it is necessary to label each voxel regarding structure affiliation (\emph{segmentation}).
-% %
-% Manual segmentation carried out by radiological experts is possible but very costly and prone to inter- and intra observer variability.
-% %
-% Segmentation algorithms allow for (semi)-automated analysis of the data but require sophisticated methods due to high inter-subject variability caused by the subject diversity.
-% %
-% Analyzing spatial data with respect to other epidemiological factors requires techniques which reach beyond standard statistical methods.
 \\\\
-%We propose an Interactive Visual Analysis (\emph{IVA}) approach \cite{Thomas2005} to provide a way to analyze image and non-image data.
 We propose an Interactive Visual Analysis (\emph{IVA}) approach \cite{Thomas2005} for the combined analysis of image and non-image data.
@@
 \begin{itemize}
-%	\item providing an overview over the workflow for analyzing cohort study data to gain insight into the large diverse subject spaces.
-	%\item an Interactive Visual Analysis workflow for analyzing image-based epidemiological data both hypothesis-driven and for hypothesis generation--based on a characterization of the standard epidemiological workflow,
 	\item an Interactive Visual Analysis workflow for cohort study data to allow both, hypothesis-driven analysis and hypothesis generation--based on a characterization of the standard epidemiological workflow,
-	%\item an Interactive Visual Analysis workflow for analyzing image-based epidemiological data including both hypothesis-driven and for hypothesis-generation based on a characterization of the standard epidemiological workflow
 	\item visualization techniques, which incorporate both information visualization and 3D rendering of organ shapes as well as combining them with epidemiological graphics and key figures,
 	\item highlighting interesting subject groups and feature associations using shape-based clustering and statistical contingency measures,
-	%Applying Interactive Visual Analysis toolset to image-based analysis in epidemiology with focus hypothesis generation
-	%\item Applying Interactive Visual Analysis toolset to image-based analysis in epidemiology with focus hypothesis generation
-	% \item Describing a Interactive Visual Analysis system aiming to support hypothesis generation .. of image-based cohort study analysis
-% 		\item Characterize the problems of epidemiological context which  %Applying the Interactive Visual Analysis technique set to the epidemiological problem domain by characterizing requirements of this context.
-% 	\item Applying the Interactive Visual Analysis technique set to the epidemiological problem domain by characterizing requirements of this context.
-% 	\item Provide an overview over the workflow for analyzing cohort study data to gain insight into the large diverse subject spaces.
-% 	\item Provide visualization techniques which combine both information visualization and 3D rendering of organ shapes as well as combining them with well known epidemiological graphics and key figures.
 	\item an implementation of the presented methods in a web framework based on \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}.
 \end{itemize}
-
-%Statistical correlations derived from analyzing the data itself may be misleading since
-
-%Include Implementation details - web based; how are image data included; what technologies are used?
 
@@
 
-%In this section we give insight into the epidemiological workflow when analyzing cohort study data and derive several challenges. %study data to identify the problems we address in this paper.
 In this section we describe the epidemiological workflow and associated requirements.
-% give insight into the epidemiological workflow when analyzing cohort study data and derive several challenges
-%
-%Epidemiological Outcome - 5Ds:
-% Death
-% Disease
-% Discomfort
-% Disability
-% Dissatisfaction
-%
+
 \subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
@@
 \end{figure}
-%Epidemiology diversity is reflected in the different experts working at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
 The diversity of epidemiology is reflected in the different experts who work at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics, and statisticians.
-%Many different experts work at epidemiological studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
 %
@@
 \begin{enumerate}
-%	\item Hypotheses are based most commonly on observations made by clinicians in their daily routine.
 	\item A hypothesis is derived from observations made by clinicians in their daily routine.
@@
 	\item A set of features depicting conditions affected by the hypothesis is compiled accordingly.
-	%\item Confounding features are adjusted so that they do not influence the effect size of an attribute.
-	\item Confounding features are identified and taken into account (for example using stratification). % adjusted so that they do not influence the effect size of a attribute.
-%
-	\item Statistical methods, such as regression analysis, assess the association of selected features with the investigated disease.%are applied to measure the effect size of attributes to the outcome of interest.
-	%\item Statistical methods such as regression analysis are applied to measure the effect size of attributes to the outcome of interest.
+%
+	\item Confounding features are identified and taken into account (for example using stratification).
+%
+	\item Statistical methods, such as regression analysis, assess the association of selected features with the investigated disease.
+%
 \end{enumerate}
@@
 We focus, on the potential of image data and attempt to support \emph{hypothesis generation}.
-%The extension we suggest, is motivated by the potential of image data and the attempt to support \emph{hypothesis generation}.
-%In the following we discuss the problems and challenges regarding the workflow.
 
@@
 %
-%This underlines the high quality standards to methods processing the data, whether to extract additional parameters or gain insight.
 
@@
 %
-%Grouping is carried out hypothesis driven.
 Grouping depends on the underlying hypothesis.
@@
 The shoe size is actually associated with gender, where women have smaller feet than men but also a longer life expectation.
-% Statistical correlations are prone to \emph{confounding}, meaning that two features are dependent and therefore should be normalized with respect to each other.
-%
-%Statistics tools such as \texttt{SPSS}\footnote{Product of IBM; \url{www.ibm.com/software/de/analytics/spss/}} and \texttt{Stata}\footnote{Product of Stata Corp.; \url{http://www.stata.com/}} play a major role for analyzing epidemiological data.
 
@@
 %
-%Often data are derived from acquired variables to either group or threshold values or to get information derived from reviewed data such as breast density data for women.
-%
 For ethical, legal or medical reasons some features cannot be gathered for each subject.
@@
 %
-%This allows for the calculation of \emph{odds ratios} which describe the relation of two \emph{dichotomous} variables, allowing for direct comparison of their influences.
-%
 Dichotomous data can also be derived by aggregating features to yield only two manifestations (e.g. subjects younger or older than 50 years).
@@
 %
-%MRI data is more expensive to obtain than CT data but does not affect the subjects health and is therefore the main method for collecting cohort study imaging data.
 Therefore, MRI is the main method for collecting cohort study imaging data.
@@
 This often yields image resolutions inferior to those of clinical day-to-day practice, which makes their analysis more challenging.
-%The equipment used to gather medical image data is kept, if possible, on the initial software and hardware version to ensure comparability in and between acquisition cycles.
+%
 The equipment used to gather medical image data is kept, if possible, on the initial software and hardware version during a longitudinal study to ensure comparability in and between acquisition cycles.
@@
 Information derived by landmarks, such as top and bottom point of a vertebra, are by far not as expressive and versatile as segmentation masks describing its whole shape.
-%Information derived by landmarks, such as top and bottom point, are by far not as expressive and versatile as segmentation masks describing the whole vertebra.
-%Information derived by landmarks such as top and bottom point of a vertebra are also not nearly as expressive and versatile as segmentation masks describing the whole vertebra.
 %
@@
 %
-%This gains even more momentum when analyzing multiple time steps.
 This gains even more importance when analyzing multiple time steps.
@@
 %
-%In addition to other MRI data, breast MRI scans were acquired for women.
-%
-%A second cohort \texttt{SHIP-Trend} was established in 2008 to acquire data about a younger population.
 A second cohort \texttt{SHIP-Trend} was established in 2008.
@@
 \section{Prior and Related Work}
-%Einfuehren von Helwigs Terminologie?
 Designing a visualization, which conveys all data aspects equally, is challenging.
@@
 %
-%Statistical measures of continuous variables such as mean, standard deviation, skewness, or inter-quartile range are used to create \emph{dimension plots} that make them comparable with respect to the derived descriptive measures.
 Statistical measures of continuous variables, such as mean, standard deviation, skewness, or inter-quartile range, are used to create \emph{dimension plots} that make them comparable with respect to the derived descriptive measures, such as voxel number of a segmented structure.
@@
 They are useful to gain an overview over numerous variables and their distributions.
-%This technique is also useful to gain an overview over numerous variables and their distributions.
 %
@@
 Dai and colleagues explored risk factors by incorporating choropleth maps of epidemiological features (e.g., mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines \cite{Dai2005}.
-%A similar approach was taken by Dai and colleagues for risk factor exploration as they also incorporate choropleth maps of epidemiological factors (e.g.
-mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines \cite{Dai2005}.
 %
 Their findings yielded a \emph{Concept Map}, which linked cancer-related associations via graph edges.
-%From their findings regarding theassociationn of cancer-related socio-demographic factors are drawn in a \emph{Concept Map} where related factors are connected via graph-edges.
+%
 Chui and colleagues visualized associations in time-dependent epidemiological data using time-series plots highlighting risk factor differences in age and gender \cite{Chui2011}.
@@
 These systems with their origin in business intelligence applications do not support epidemiological workflows, in particular workflows that also consider medical image data.
-%These systems share limitations in extensibility to a specific problem domain.
-%Spotfire offers the possibilities to interact with the \texttt{R} statistical programming language 
 % Image Data
@@
 \section{Image Centric Cohort Study Data in Interactive Visual Analysis Context} \label{Image Centric Cohort Study Data in Interactive Visual Analysis Context}
-%- Acceptance plays a big role - you have to pick up the user by presenting information they are used to in a new way\\
 We described the epidemiological workflow as a sequence of steps taken by domain experts which need to comprise reproducibility and statistical integrity (recall Subsection~\ref{EpidemiologicalWorkflow}).
@@
 Introducing the \emph{IVA} principle to the epidemiological application domain aims to compensate its weaknesses rather than replacing the existing workflow.
-%The workflow we propose by introducing the \emph{IVA} principle into the epidemiological application domain does not aim at replacing the existing workflow but to complement its weaknesses.
 %
@@
 Statistical tests on features associated to a hypothesis yield a value for deciding whether the data supports the hypothesis or not.
-%A list of features describing the hypothesis is compiled and analyzed using statistical tests. 
-%
-%The resulting value decides whether the data supports the hypothesis or not.
 %
 Features not included in the analysis may potentially support the chosen hypothesis by discriminating the population in the expected way, but are not highlighted in any way.
-%It would be possible that there are actually features of the data set which support the hypothesis by discriminating the population in the expected way, but with this approach they are not highlighted in any way.
-%
-%Adapting this workflow to medical image data analysis also highlights this problem.
+%
 This becomes even more important when the workflow is adapted to the analysis of the medical image data, where
-domain experts annotate landmarks tediously to derive metrics, such as diameters. %which are then handled like other features and analyzed using the same set of statistical tools.
+domain experts annotate landmarks tediously to derive metrics, such as diameters.
 %
@@
 It is easily possible that left out information would heavily influence the result.
-%Not only does this leave out the majority of the information in the medical image data by abstracting it to single values, it is easily possible that information left out would heavily influence the result.
 %
 Considering more complex parts of the data would make those results more trustworthy and also could identify possible anatomical confounders--an epidemiolgical research result in itself.
-%Statistical tests check for validity of the number but not for their completeness or plausibility!
 \\\\
@@
 %
-%Figure~\ref{fig:WorkflowComparison} (b) highlights the iterative process as part of the epidemiolgical workflow.
-%
 It also aims to project back into the hypothesis formulation step to amplify hypothesis generation.
-%Note that it also aims to project back into the hypothesis formulation step to amplify hypothesis generation 
 %
@@
 
-% ---------------------------- old part ----------------------------
-% In the \emph{IVA} context, data is divided into two major view types.
-% %
-% The human body exposing shape information for the \emph{physical view} \cite{Oeltze2013}.
-% %
-% This information space is usually displayed via volume rendering techniques \cite{Oeltze2007}.
-% %
-% %Spatiotemporal views include data from the \emph{Object Space Domain} renders data centered to the focused domain, in our example the medical image data space.
-% %
-% These variables spanning the 3D-space are also referred to as \emph{domain variables}.
-% %
-% \emph{Range variables} in the epidemiological context can be divided twofold:
-% %
-% \begin{itemize}
-% 	\item Variables derived from the image data. 
-% 	%
-% 	These measures quantify shape information to allow for comparison and can also be used to brush in the image space.
-% 	%
-% 	%These variables describe image data and can also be used to brush in the image space.
-% 	\item Epidemiological socio-demographic or medical attribute data.
-% 	%
-% 	These values are associated with every subject represented in the image space. 
-% 	%
-% 	They do not describe shape information.
-% 	%
-% 	This is the data epidemiologists usually want to correlate with image data.
-% \end{itemize}
-% %Other data associated to elements in the object space are considered to be the \emph{Attribute Space Domain} and are displayed using various range based views using statistical graphics and information visualization techniques like scatterplots, bar charts or parallel coordinates \cite{Oeltze2013, Oeltze2007}.
-% %
-% %Specific for the epidemiological application domain is that both Object- and Feature Space Domain influence the outcome of interest.
-% % \textbf{Object Space}\\
-% % - Medical Image Data / Spine Segmentations\\
-% % \textbf{Attribute Space}\\
-% % - SHIP-Variables\\
-% % - Derived visualizations
-% % 
-% % ? Where to incorporate Levels of IVA
-% 
-% \subsection{IVA Patterns}
-% %
-% \emph{IVA} knows three different exploration patterns, handling associations of domain and range variables.
-% %The explorative procedures when analyzing data using \emph{IVA} can be divided into three different patterns, handlinassociationon between domain and range variables.
-% 
-% \paragraph{Local Investigation}
-% This pattern projects information from domain space to the range perspective.
-% %
-% This step is more complicated in epidemiological context compared to other \emph{IVA} application domains.
-% %As opposing to other \emph{IVA} application domains, this step is more complicated in the epidemiological context.
-% %
-% Shape information can not be brushed using ROI-selections but rather has to employ techniques that specify local deformation changes \cite{Hermann2014} or subjects that belong to a shape class.
-% %
-% \emph{Feature selection} methods strongly depend on the type of segmentation used to extract the tissue of interest.
-% %
-% Model-based segmentations or masks yield data structures capable of calculating mean shapes and distances between individuals or subject groups.
-% %
-% Feature selection is also possible by applying clustering algorithms in order to get shape-groups \cite{Klemm2013VMV}.
-% %
-% These algorithms can be used to investigate associations between shape-groups and other non-image based variables.
-% %
-% Analysis of outliers can indicate segmentation errors or an group of subjects sharing a pathology.
-% %
-% %Another application is the outlier analysis. 
-% %
-% %Outliers can indicate segmentation errors or an outstanding group of individuals who may share a pathology.
-% 
-% % Local Investigation
-% % - Projection from Image Space to Attribute data. This is more difficult in our application, because we currently only brush on derived features.\\
-% % - Clustering in Image Space yields Groups which can be analyzed using the attribute space\\
-% % - Create a Pipeline Overview over different Levels of different IVA Patterns and Stages
-% 
-% % \subsection{Feature Selection}
-% 
-% % \subsection{Brushing the information space}
-% % - derive new groups which serve as input for the feature selection
-% 
-% \paragraph{Feature Localization}
-% % Subset selection of range variables yields features
-% The vast majority of features are considered to be dependent with respect to the image domain in the \emph{IVA} context.
-% %As described before, the vast majority of data points are considered to be dependent with respect to the image domain in the \emph{IVA} context.
-% %
-% Selecting subjects based on image-derived data can be seen as additional possibility of shape-related grouping.
-% %
-% Epidemiologist are interested in the shape of subjects within a range of a set of variables that describe the current hypothesis.
-% %
-% Epidemiologists may categorize data into groups that fit their hypothesis formulation.
-% %
-% Continuous variables such as age are for example often divided in categories like young, aged and elderly.
-% %
-% %Categorization is strongly dependent on the hypothesis and therefore requires suitable brushing techniques as described in Section~\ref{sec:AdaptiveFeatureVisualization}.
-% %
-% %To gain insight intassociationons between SHIP-
-% 
-% % - Selection of Information using Bar Charts, Scatterplots or Parallel Coordinates which projects the selection into the Object space\\
-% % - this selection is for categorical data already given implicitly bei projecting the 3D-View onto the Bar Charts/Mosaik Plots!\\
-% % - this aims to locale features of the data!\\
-% % - Gain Information about SHIP-Variables by putting them into the context of each other.
-The Pivot Table allows for direct numerical\\ analysis, while the information visualizations allow for better insight of the combination
-% 
-% \paragraph{Multivariate Analysis}
-% %Introduced in the information visualization, the multivariate analysis means selecting subjects in like in the feature localization step and get have the result highlighted in another linked view displaying other non-image parameter.
-% %Introduced in the information visualization, the multivariate analysis incorporates brushing and linking of of views displaying non-image parameter.
-% Multivariate analysis incorporates brushing and linking of of views displaying non-image parameter.
-% %
-% The type of the parameters which are compared determines the information visualization technique used.
-% %
-% %This represents for the most part the usual approach which essentially is the analysis of how variables interact with each other, only in a visual analysis context.
-% %
-% Statistic measures are needed which describe how variables correlate given the selected groups.
-% %The need of statistic measures which describe how variables correlate with each other given the selected groups is special for the application domain.
-% %
-% These associations are also summarized using pivot tables which are popular in epidemiology.
-% % - Selection of Elements in Scatterplots, Barcharts or Parallel Coordinates Views - observe how selection changes another view - this allows for multivariate analysis\\
-% % - Becker, R.A., Cleveland, W.S.: Brushing scatterplots. Technometrics 29(2) (1987)\\
-% % - Wang Baldonado, M.Q., Woodruff, A., Kuchinsky, A.: Guidelines for using multiple views in information visualization.
-In: AVI ’00: Proceedings of the working conference on Ad- vanced visual interfaces, pp.
-110–119.
-ACM Press, New York, NY, USA (2000).
-DOI\\ http://doi.acm.org/10.1145/345513.345271
-% % - By brushing individual parameters or create new binnings of parameter it is possible to see how they change in coordinated views. This is already implemented in the Cargo framework\\
-% % - by creating comparative 3d Visualizations it is possible to assess the influence of non-image parameters to the visual space.
-% 
-% ---------------------------- old part ----------------------------
-
 \subsection{Data Preprocessing} \label{Data Preprocessing}
@@
 \paragraph{Non-Image Data. }
-%To include heterogenous epidemiological data in an \emph{IVA}-framework it is necessary to process it to obtain standardized views to the available features.
 %
 Multimodal features require different techniques.
-%Due to the different acquisition modalities there have to be different techniques incorporated.
 %
@@
 Exporting the data in the respective tool to a CSV file and then converting it to file types that are easily manageable, such as JSON or XML, makes it readable for all modern programming languages.
-%The best solution for us was to simply export the data in the respective tool to a CSV file and then convert it to data types which are easier manageable such as JSON or XML using our own classes.
 %
 This can be achieved by using data wrangling tools such as \texttt{OpenRefine}\footnote{Developed by Google, Open Source; \url{http://openrefine.org}}, which also validate the data (find missing data, clean up bad formatting, transform scales).
-%In order to verify that the conversion worked as expected and the data is valid, it is good practice to use data wrangling tools such as \texttt{OpenRefine} to validate the data (find missing data, clean up bad formatting).
 %
@@
 \paragraph{Image Data. }
-%Gerade weil der Satz davor so kompliziert war, ist unklar, was "this" bedeutet.
-Ich denke zunächst muss gesagt werden, was typische Ziele sind (anhand von Beispielen, wie Abstände zwischen anatomischen Landmarken, Ausdehnungen (also längste Durchmesser).
-Wichtig wären aber konkrete relevante Beispiele.
-Dann wird gesagt, was man dafür machen muss und wie man das macht.
 Information about anatomical structure, such as diameter or volumes, is extracted from the image data.
-%Processing the image data associated to each subject consists for the most part of information extraction about an anatomical structure.
 %
@@
 %
-%Model-based approaches have shown in principle to be effective for segmentation \cite{Gloger2010, Gloger2012} and detection \cite{Rak2013}.
-%Model-based segmentation approaches are based on assumptions related to object shape, size, location and texture.
 In principle, model-based approaches have shown to be effective for detection \cite{Rak2013} and segmentation \cite{Gloger2010, Gloger2012}.
@@
 If a segmentation yields only binary masks separating the structures, algorithms such as \emph{Growing and Adaptive Shapes} can be applied creating a surface grid where each point is comparable throughout the population \cite{Ferrarini2007}.
-%
-%Intensity-based comparisons can be achieved using rigid image registration, but model based results however are preferable \cite{Klemm2012}.
 %
@@
  The data are visualized in the range and domain space, which can then be brushed, yielding new groups, to be investigated using further features.
- % The data are visualized in the range and domain space, which can then be brushed, yielding new groups, to be investigated using further features. 
  %
@@
 In our system, the visualization can either be brushed or new features can be added to the analysis. 
-%Subsequently the visualization can be either brushed or new features can be added to the analysis.
 %
@@
 %
-Brushed regions are treated like features as they divide the subject space just like categorical features. %and lead therefore to a new feature selection.
+Brushed regions are treated like features as they divide the subject space just like categorical features.
 %
@@
 %
-%The starting point of \emph{IVA Loop} is determined whether the analysis is hypothesis-driven or exploratory.
-%
-%An analysis with the aim of hypothesis generation of image-based data starts with a clustering of the structure 
-%
 A sample workflow using interaction and visualization techniques described in the next section can be seen in Figure~\ref{fig:visualization}.
@@
 \subsection{Interaction and Visualization Techniques} \label{Interaction- and Visualization Techniques}
-%We employ the fourth and highest level which includes next to brushing and linking, advanced brushing, extraction of attributes the development of visualizations custom tailored to the data sets.
-%
-%\emph{IVA} levels define different levels interaction. 
 %
@@
 %
-%Since epidemiologists come from different expertise, their different background has to be taken into account.
-%
-%The chances of a method to be adapted by epidemiologists increases highly if their domain language is used and the methods reflect routines they are used to.
 The methods have to use domain language and reflect routines of epidemiologists in order to be incorporated into their research.
-%The chances of a method to be adapted by epidemiologists increases highly if their domain language is used and routines they are used to are reflected in the methods.
 %
@@
 In the following sections we present the different parts of our proposed \emph{IVA} system for image-based cohort study analysis.
-%Chosing or creating well suited visualization- anassociationon techniques for epidemiological data
-%The employed techniques need to respect the epidemiological requirements.
 
@@
 %
-%Continuous data are discretized and therefore treated as categorical data.
 Continuous data can be visualized using scatterplots (single features) or parallel coordinates (multiple features), but in epidemiology, this data type is usually categorized into ordinal groups of \emph{equal size}.
@@
 
-%Inspired by the previously discussed \texttt{GPloms} \cite{Francois2013} the visualization type is chosen dynamically based on the variable types and number of visualizations which need to be displayed.
 %
@@
 Following Tufte's concept of \emph{small multiples} \cite{Tufte1983}, the medical image data are directly incorporated into the plot by including color-coded mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
-%If possible the medical image data is directly included into the plot as well by including mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
 %
@@
 %
-%Mapping the distance from a group mean shape to the global mean to color allows to assess local shape
 Mapping the distance from a group mean shape to the global mean using color allows to assess local shape changes.
@@
 Thus, we decided to introduce an interactive pivot table.
-%When opening a random epidemiological paper the reader will in almost all cases find some sort of pivot table to present the data.
-%
-%As seen in Figure~\ref{fig:visualization} (c) it is a good way to display how many subjects are in each group.
+%
 These tables clearly convey the subject count in each group (see Figure~\ref{fig:visualization} (c)).
@@
 % TODO IMPLEMENT
-%The mean shape for each resulting sub-group is also displayed for each subject.
 
@@
 
-%It shares the same restrictions as \emph{Pearson's $X^2$} which include next to others expected count larger than 5 for $80\%$ of the contingency table entry and no expected value can be smaller than one.
 It shares the same restrictions as \emph{Pearson's $X^2$}.
@@
 They cannot be included into the epidemiological analysis, since statistical validation needs a minimum count to be seen as valid.
-%This makes them not assessable for epidemiological analysis as well since statistical validation needs a minimum count to be seen as valid.
-% Limitations and usual space
+%
 The contingency matrix highlights correlations between all data set features. %of visualized features with features in the data set.
@@
 %
-%Many methods which specialized libraries, like the Visualization Toolkit (VTK) or R for statistics have build in, need to be written from scratch in order to fit in the context.
 Due to missing ports of sophisticated and specialized libraries/languages, such as the \texttt{Visualization Toolkit} (\texttt{VTK})\footnote{Developed by Kitware Inc; \url{http://vtk.org}} or \texttt{R}\footnote{Open Source; \url{http://r-project.org}} for statistics, many standard methods need to be written from scratch for the web.
-%Many methods, which specialized libraries, like the Visualization Toolkit (VTK) or R for statistics, have build in, need to be written from scratch in order to fit in the context.
 %
@@
 %
-%\texttt{WebGL} rendering is carried out by \texttt{Three.js}\footnote{Originally developed by Ricardo Cabello, \url{http://threejs.org}} and allows GPU-accelerated data rendering.
 \texttt{Three.js}\footnote{Originally developed by Ricardo Cabello, \url{http://threejs.org}} allows GPU-accelerated data rendering using \texttt{WebGL}.
@@
 \\\\
-%The following section describes how our presented \emph{IVA} workflow works for an epidemiological use case.
 \section{Application}
@@
 %
-%If at all, only a few shape-related features are included in comparable studies, for example by Lang and colleagues \cite{Lang2011}.
-%In comparable studies, only a few shape-related features are included, for example by Lang and colleagues \cite{Lang2011}.
 In comparable studies, only a few shape-related features are included \cite{Lang2011}.
@@
 %
-%The goal is to find features which \emph{interact} with lower back pain based on image and non-image data.
-%
 \subsection{The Lumbar Spine Data Set}
@@
 %
-There are 127 features describing diagnosed diseases, lifestyle factors, women-specific factors, pain indicators, laboratory values and somatometric features. %and are ordered accordingly.
+There are 127 features describing diagnosed diseases, lifestyle factors, women-specific factors, pain indicators, laboratory values and somatometric features.
 %
@@
 The data are preprocessed as described in Section~\ref{Data Preprocessing}.
-%Following methods described in Section~\ref{Data Preprocessing}, the data is preprocessed as follows for the presented prototype.
-%Transformation operations on the data to prepare it for the presented prototype are denoted as data preprocessing.
 %
@@
 Two user-defined landmarks on the top and bottom of the L3 vertebra are used to obtain an initial height estimation of the model.
-%Two more user- defined landmarks are the top and bottom of the L3-vertebra, which transforms size of the model accordingly to match the L3-vertebra .
 %
@@
 \subsection{Shape Visualization and Clustering}
-%The detection algorithm described in Section~\ref{Data Preprocessing} yields a model that determines voxel belonging to the lumbar spine.
 %
@@
 Since the ``correct'' number of clusters in a given group is unknown, an estimate is computed by means of the knee/elbow method \cite{Salvador2004}.
-%Since it is not possible to determine the number of clusters in a given group, it is automatically computed using a \emph{knee/elbow} point which is described as tradeoff between number of cluster and a cluster evaluation metric \cite{Salvador2004}.
-%
-%For details, see \cite{Klemm2013VMV}.
 %
@@
 %
-%We analyzed the given dataset expert-guided to assess its suitability for supporting both hypothesis-free analysis as well as generation.
 Expert-guided analysis assessed the suitability of our approach for supporting both hypothesis-free analysis as well as hypothesis generation. 
@@
 %
-%Shape-based hypothesis-free exploration starts with a shape-grouping step achieved by shape-based clustering.
 Shape-based hypothesis-free exploration starts with the \emph{local investigation} pattern by a shape grouping step using shape-based clustering.
@@
 Cluster 8 has the second largest number of elements and was therefore of special interest.
-%Cluster 8 was of special interest because of its large distance to the mean shape while still exposing the second highest subject count.
 %
@@
 Metabolic parameters, such as blood fat and blood sugar, are also possibly associated features.
-% TODO: Mit Bernhard klären
+%
 The employment status is a feature relating to many different lifestyle factors such as income or nutrition as well as age and might act as confounder.
@@
 While this approach does not assume any hypothesis, the 127 epidemiological features were selected by domain experts with a wide range on potential influential factors.
-%While this approach does not assume any hypothesis, the 127 epidemiological features were selected by domain experts and also includes features, which are only remotly unter verdacht to be associating with back pain.
-%the data are met with a selection bias when compiling the list of related features by the domain experts.
 %
@@
 
-% TODO Bernhard fragen, was ihn an der Reihenfolge stört
 It is well known that overweight is a risk factor for back pain.
@@
 Our epidemiological collaborators were interested in the correlation with waist circumference presented by this group.
-%Our epidemiological collaborators were interested that this group presented a correlation with waist circumference.
 %
@@
 To increase the probability of the observation not to be random and to avoid \emph{overfitting}, they are cross-checked for associations in \texttt{SHIP-TREND} as a second, independent population sample, which was examined with methods identical to \texttt{SHIP}.
-%If features are identified as interesting for a disease, the epidemiologists look for associations in another cohort, the \texttt{SHIP-TREND}.
-%
-%If the correlation is observed there as well, it is likely not random.
 \\\\
@@
 %
-%On the feature level, displaying MRI scans is interesting for the subjects in the outlier clusters because they are highly likely to exhibit pathologies.
 Displaying MRI scans for the subjects in the outlier cluster is promising because they are highly likely to exhibit pathologies.
@@
 Hence, the framework allows for both, hypothesis-driven analysis and hypothesis generation.
-%Seen as an extension to the existing statistics driven workflow, image-centric analysis of epidemiological data from a IVA approach by illuminating the data-black box.
-%The presented \emph{IVA} approach is classified as extension to the existing statistics-driven epidemiological approach.
-%
-%Image-centric epidemiological analysis benefits from an IVA approach by illuminating the data-black box.
 %
@@
 Shape-based clustering assesses the variability of a anatomical structure in the context of non-image features such as disease indicators or lifestyle factors.
-%
-%The iterative nature of the \emph{IVA} approach reflects versatility, as both \emph{hypothesis-driven analysis} and \emph{hypothesis generation} are allowed.
 
@@
 A number of improvements is left open for future work, e.g.,
-%TODO: Fragen ob die Liste so klar geht mit der Formulierung und ohne Kommas
+
 \begin{itemize}
@@
 %
-%We believe that web technologies pave the way to allow as many experts as possible to analyze this data.
 We believe that web technologies pave the way to analyze this data in a convenient way.
@@
 
-%With increasing number of image-centric cohort studies, >evalueted< subjects, features collected and major advantages in methods that allow for comparabality of findings in different studies the gap between
-
 %% if specified like this the section will be committed in review mode
-%\acknowledgments{SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no.
+\acknowledgments{SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no.
 03ZIK012), the Ministry of Cultural Affairs as well as the Social Ministry of the Federal State of Mecklenburg-West Pomerania.

### 2014-06-10 draft.tex
added some minor changes of the reviewer. Added changelog to keep track for cover letter

@@
 \usepackage{times}
+% TODO: Self added Packages - need to be removed afterwards
+%\usepackage{babel}
+\usepackage[normalem]{ulem}
+\usepackage{color}
+\newcommand{\rem}[1]{\textcolor{red}{\sout{#1}}}
+\newcommand{\add}[1]{\textcolor{blue}{\uline{#1}}}
 
@@
 %
-Cohort studies comprise heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors and more recently, medical image data.
+Cohort studies comprise of heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors and more recently, medical image data.
 %
@@
 %
-The cohort consists of several thousands of persons randomly selected to avoid any bias.
-%
-Due to this random selection, the majority of the cohort does not suffer from a specific health problem.
+The cohort consists of several thousands of people who are randomly selected to avoid any bias.
+%
+\rem{Due to this random selection, the majority of the cohort does not suffer from a specific health problem.}
+\add{The people are selected without focus on a certain disease.}
 %
@@
 \emph{Hypothesis generation} gains importance since the number of epidemiological features increases and the focus shifts towards more complex relations involving more than two features.
-%
-The easy exchange of developed methods driven by modern web technologies intends to trigger a fast feedback loop between us and the epidemiologists.
-% http://wiki.infowiss.net/Gulf_of_execution ?
-We applied our approach to a data set compiled to analyze diseases related to the lumbar spine and aim to determine features, which indicate pathological changes.
-%
-This data set comprises 127 features and 2 sequences of MRI data from 2,333 patients.
 \\\\
@@
 	\item highlighting interesting subject groups and feature associations using shape-based clustering and statistical contingency measures,
-	\item an implementation of the presented methods in a web framework based on \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}.
+%	\item an implementation of the presented methods in a web framework based on \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}.
 \end{itemize}
+%
+\rem{The easy exchange of developed methods driven by modern web technologies intends to trigger a fast feedback loop between us and the epidemiologists}.
+% http://wiki.infowiss.net/Gulf_of_execution ?
+We applied our approach to a data set compiled to analyze diseases related to the lumbar spine and aim to determine features, which indicate pathological changes.
+%
+This data set comprises 127 features and 2 sequences of MRI data from 2,333 patients.
+%
+\add{We implement the presented methods using modern web technologies, such as \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}, making them easily accessible for the domain experts.
+%
+This aims to trigger a fast feedback loop between us and the epidemiologists.}
 
@@
 %
-Following Thew and colleagues, the workflow can be characterized as follows \cite{Thew2009}:
+Following Thew et al \cite{Thew2009}, the workflow can be characterized as follows:
 %
@@
 The shoe size is actually associated with gender, where women have smaller feet than men but also a longer life expectation.
-
-Statistics tools such as \texttt{SPSS}\footnote{Product of IBM; \url{http://ibm.com/software/analytics/spss/}} play a major role for analyzing epidemiological data.
+%
+\add{An extreme form of confounding is called \emph{Simpson's Paradox} \cite{SimpsonsParadox}, where the confounder reverses the observed effect between two features.
+%
+%An example is the higher mortality rate of non-insulin dependent diabetes opposed to insulin dependent diabetes.
+A famous example is the Berkeley University gender bias case, where there were significantly less women admitted then man.
+%
+Analysis later showed that women applied mostly for strongly restricted subjects and when only looking acceptance divided by departments there was actually a small bias toward women.
+%
+%For example, insulin dependent and non-insulin dependent diabetes shows 
+%In a famous example, researchers concluded that a newer treatment for kidney stones was more effective than traditional surgery, but it was later revealed that the newer treatment was more often being used on small kidney stones.
+%For example, comparing mortality rate of non-insulin dependent 
+%When for example surgeons are ranked regarding diseased patients, without including the risk of the surgeries, the results may be distorted.
+}
+
+Statistical tools such as \texttt{SPSS}\footnote{Product of IBM; \url{http://ibm.com/software/analytics/spss/}} play a major role for analyzing epidemiological data.
 %
@@
 
-Indicators for medical conditions as well as questions about a subject's lifestyle are also often \emph{dichotomous}--they have two manifestations (\emph{Yes} or \emph{No}).
+Indicators for medical conditions as well as questions about a subject's lifestyle are \rem{also} often \emph{dichotomous}--they have two manifestations (\emph{Yes} or \emph{No}).
 %
@@
 %
-The equipment used to gather medical image data is kept, if possible, on the initial software and hardware version during a longitudinal study to ensure comparability in and between acquisition cycles.
+%The equipment used to gather medical image data is not updated or changed during a longitudinal study to ensure comparability in and between acquisition cycles.
+The equipment used to gather medical image data is \rem{kept, if possible, on the initial software and hardware version} \add{not updated or changed} during a longitudinal study to ensure comparability in and between acquisition cycles.
 
@@
 \\\\
-Their focus on \emph{hypothesis generation} using parallel assessment of multiple data features makes the work of Turkay and colleagues closest to ours albeit our emphasis on processing models derived from medical image data segmentation and variables with categorical manifestations \cite{Turkay2013}.
+%Their focus on \emph{hypothesis generation} using parallel assessment of multiple data features makes the work of Turkay and colleagues closest to ours albeit our emphasis on processing models derived from medical image data segmentation and variables with categorical manifestations \cite{Turkay2013}.
+\rem{Their} \add{Turkay and colleagues \cite{Turkay2013}} focus on \emph{hypothesis generation} using parallel assessment of multiple data features makes \rem{the work of Turkay and colleagues} \add{their work} closest to ours albeit our emphasis on processing models derived from medical image data segmentation and variables with categorical manifestations \rem{\cite{Turkay2013}}.
 %
@@
 \paragraph{Visualizing Image and Non-Image Data. }
-Gresh and colleagues proposed \texttt{WEAVE}, one of the first systems which concurrently analyzed medical image and non-image data using linked views \cite{Gresh2000}.
-%
-Blaas and colleagues presented a similar system which analyzed medical image data and variables derived from them using views from the feature and physical space \cite{Blaas2007}.
+Gresh \rem{and colleagues} \add{et al. \cite{Gresh2000}} proposed \texttt{WEAVE}, one of the first systems which concurrently analyzed medical image and non-image data using linked views.
+%
+Blaas \rem{and colleagues} \add{et al. \cite{Blaas2007}} presented a similar system which analyzed medical image data and variables derived from them using views from the feature and physical space.
 %
@@
 %
-Steenwijk and colleagues employ a relational database to organize the data for visualizing subject data using linked views such as parallel coordinates, scatterplots and time plots \cite{Steenwijk2010}.
-%
-Zhang and colleagues provide a web-based system for analyzing subject groups with linked views and batch-processing capabilities for categorizing new subject entries into the data set \cite{Zhang2012}.
+Steenwijk \rem{and colleagues} \add{et al.
+\cite{Steenwijk2010}} employ a relational database to organize the data for visualizing subject data using linked views such as parallel coordinates, scatterplots and time plots.
+%
+Zhang \rem{and colleagues} \add{et al.
+\cite{Zhang2012}} provide a web-based system for analyzing subject groups with linked views and batch-processing capabilities for categorizing new subject entries into the data set.
 %
@@
 %
-Dai and colleagues explored risk factors by incorporating choropleth maps of epidemiological features (e.g., mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines \cite{Dai2005}.
+Dai \rem{and colleagues} \add{et al.
+\cite{Dai2005}} explored risk factors by incorporating choropleth maps of epidemiological features (e.g., mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines.
 %
@@
 %
-Chui and colleagues visualized associations in time-dependent epidemiological data using time-series plots highlighting risk factor differences in age and gender \cite{Chui2011}.
+Chui \rem{and colleagues} \add{et al. \cite{Chui2011}} visualized associations in time-dependent epidemiological data using time-series plots highlighting risk factor differences in age and gender.
 
@@
 %
-Caban and colleagues investigated the suitability of variance visualizations of shape distribution models and concluded that users favor spherical glyph representations over deformation grids and likelihood volumes \cite{Caban2011}.
-%
-The distribution of shapes in a space derived from a PCA is plotted by Busking and colleagues in a 2D-projected plane of the space \cite{Busking2010a}.
+Caban \rem{and colleagues} \add{et al.
+\cite{Caban2011}} investigated the suitability of variance visualizations of shape distribution models and concluded that users favor spherical glyph representations over deformation grids and likelihood volumes.
+%
+The distribution of shapes in a space derived from a PCA is plotted by Busking \rem{and colleagues} \add{et al. \cite{Busking2010a}} in a 2D-projected plane of the space.
 %
@@
 %
-Hermann and colleagues identify local deformation changes by investigating shape-related differences on rodent mandibles \cite{Hermann2014}.
+Hermann \rem{and colleagues} \add{et al. \cite{Hermann2014}} identify local deformation changes by investigating shape-related differences on rodent mandibles.
 %
@@
 \paragraph{SHIP Data Analysis.}
-Klemm and colleagues visualized lumbar spine variabilities based on an semi-automatic shape detection algorithm of 490 participants of the \texttt{SHIP-2} cohort \cite{Klemm2013VMV}.
+Klemm et al. \cite{Klemm2013VMV} visualized lumbar spine variabilities based on an semi-automatic shape detection algorithm of 490 participants of the \texttt{SHIP-2} cohort.
 %
@@
 %
-Niemann and colleagues presented an interactive data mining tool for the assessment of risk factors on hepatic steatosis, the fatty liver disease \cite{Niemann2014}.
+Niemann et al. \cite{Niemann2014} presented an interactive data mining tool for the assessment of risk factors on hepatic steatosis, the fatty liver disease.
 %
@@
 %
-Oeltze and colleagues combined a linked view representation of results from a statistical analysis with feature localizations of the tissue perfusion with the goal of its evaluation \cite{Oeltze2007}.
+Oeltze et al. \cite{Oeltze2007} combined a linked view representation of results from a statistical analysis with feature localizations of the tissue perfusion with the goal of its evaluation.
 %
@@
 Features can be added freely on the canvas via drag and drop.
-Dropping the \emph{age} parameter on the already plotted \emph{body size} container creates a mosaic plot combining both features (right).
+Dropping the \emph{gender} parameter on the already plotted \emph{body size} container creates a mosaic plot combining both features (right).
  }
@@
 %
-The methods have to use domain language and reflect routines of epidemiologists in order to be incorporated into their research.
+The methods have to use domain language and reflect \add{the} routines \rem{of}\add{that} epidemiologists \add{take} in order to be incorporated into their research.
 %
@@
 % TODO Implement MOSAIC PLOTS, Scatterplots, Parallel Coordinates
-Following Tufte's concept of \emph{small multiples} \cite{Tufte1983}, the medical image data are directly incorporated into the plot by including color-coded mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
+Following Tufte's concept of \emph{small multiples} \cite{Tufte1983}, \add{information derived from the} medical image data are directly incorporated into the plot by including color-coded mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
 %
@@
 %
-Turkay and colleagues used the approach to calculate various key figures based on the distribution functions of each feature derived from the image data \cite{Turkay2013}.
+Turkay et al. \cite{Turkay2013} used the approach to calculate various key figures based on the distribution functions of each feature derived from the image data.
 %
@@
 %
-Contingency is visualized using an interactive adjacency matrix with association power mapped to color hue.
+Contingency is visualized using an interactive adjacency matrix with association power mapped to color brightness.
 %
@@
 %
-Epidemiological analysis of lumbar back pain, such as the work of Harreby and colleagues \cite{Harreby1996}, is largely focused on non-image information.
+Epidemiological analysis of lumbar back pain, such as the work of Harreby et al. \cite{Harreby1996}, is largely focused on non-image information.
 %
@@
 \subsection{The Lumbar Spine Data Set}
-We divide the data set in image and non-image data.
+We divide the data set into image and non-image data.
 %
@@
 \paragraph{Image Data.} \label{Image-Data}
-The lumbar spine was detected in the image data using a hierarchical finite element method by Rak and colleagues \cite{Rak2013}.
+The lumbar spine was detected in the image data using a hierarchical finite element method by Rak et al. \cite{Rak2013}.
 %
@@
 %
-Another \emph{multivariate analysis} using a pivot table set gender and employment status in relation to clustering affiliation, showing that cluster~8 contains mostly women and also has a larger unemployment rate (see Figure~\ref{fig:hypothesisfree}), while the overall employment rate of women and men in the data set is almost exactly 50\%.
+Another \emph{multivariate analysis} using a pivot table set gender and employment status in relation to clustering affiliation\add{.} 
+%
+\add{It S}hows that cluster~8 contains mostly women and also has a larger unemployment rate (see Figure~\ref{fig:hypothesisfree}), while the overall employment rate of women and men in the data set is almost exactly 50\%.
 %
@@
 %
-Correlation does not automatically imply causation \cite{Tufte2003}.
+Correlation does not automatically imply causation. %\cite{Tufte2003}.
 %

### 2014-06-16 draft.tex
First Version of rewrite of application section

@@
 \usepackage[normalem]{ulem}
-\usepackage{color}
+%\usepackage{color}
+\usepackage{xcolor}
 \newcommand{\rem}[1]{\textcolor{red}{\sout{#1}}}
 \newcommand{\add}[1]{\textcolor{blue}{\uline{#1}}}
+\newcommand{\com}[1]{\textcolor{orange}{\uline{#1}}}
 
@@
 %(The problem).
+\com{ToDo: Reviewer-Feedback: The abstract should be formulated more concisely and focus on What? 
+Why? How? What follows?}
 Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease-specific risk factors.
@@
 \rem{Due to this random selection, the majority of the cohort does not suffer from a specific health problem.}
-\add{The people are selected without focus on a certain disease.}
+\add{The subjects are selected without focus on a certain disease.}
 %
@@
 
-\subsection{Epidemiological Data} \label{EpidemiologicalData}
+\section{Epidemiological Data} \label{EpidemiologicalData}
+\com{Primary Reviewer Feedback: Another weakness is on the data analysis; while there is information about the data within the text (including 2.2, 4.1, and 5.2) I would like to see a section that focuses on the data and the challenges thereof.
+This chapter needs to be enhanced! Maybe own section!}
 Epidemiological data are highly heterogenous and incomplete.
@@
 
-\paragraph{Image acquisition.} \label{ImageAcquisition} Imaging techniques involving ionizing radiation for the subject are not suitable for ethical reasons.
+\subsection{Image acquisition} \label{ImageAcquisition} Imaging techniques involving ionizing radiation for the subject are not suitable for ethical reasons.
 %
@@
 
-\paragraph{Image analysis.} \label{ImageAnalysis} Decisions have to be made on how image data are \emph{compared} and \emph{quantified}.
+\subsection{Image analysis} \label{ImageAnalysis} Decisions have to be made on how image data are \emph{compared} and \emph{quantified}.
 %
@@
 
-\subsubsection{System Structure} \label{Structure and Workflow}
+\section{System Structure and Design} \label{Structure and Workflow}
+\com{Primary Reviewer Feedback: There is little detail about the design, and how it was decided to create this particular design; what interaction was made with the users, how did they influence this design?}
+% \subsubsection{System Structure}
 We divide the workspace into four major parts, as illustrated in Figure~\ref{fig:visualization} and \ref{fig:similarity}.
@@
 \section{Application}
+\add{This Chapter needs to be reworked completely, since it was subject of the reviewer criticism.}
+This section describes how our presented \emph{IVA} workflow is used for an epidemiological use case.
+%
+We conducted a case study with two domain experts on a data set compiled to analyze lower back pain. 
+%
+This is one of the most common reasons for an adult to see a physician in the Western civilization \cite{Backpain}.
+%
+Epidemiological analysis of lumbar back pain, such as the work of Harreby et al. \cite{Harreby1996}, is largely focused on non-image information.
+%
+In comparable studies, only a few shape-related features are included \cite{Lang2011}.
+%
+To our knowledge, this is the first approach, where shape-related information of the whole lumbar spine is analyzed along with other epidemiological features.
+%
+Determining risk factors in this area can lead to \cite{Fletcher2012}:
+\begin{itemize}
+	\item a better understanding of effects of preventive measures such as occupational health and safety regulations,
+	\item prognostic features for diagnosis and treatment of lumbar back pain, and
+	\item determination of particularly affected risk groups.
+\end{itemize}
+%
+Characterizing the healthy aging process of the spine is a long-term goal for determining age-normalized probabilities for spine-related diseases by incorporating individual risk factors.
+%
+\subsection{The Lumbar Spine Data Set}
+We divide the data set into image and non-image data.
+%
+There are 127 features describing diagnosed diseases, lifestyle factors, women-specific factors, pain indicators, laboratory values and somatometric features.
+%
+The image data was acquired on a 1.5~Tesla scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany) by four trained technicians in a standardized way.
+%
+%The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence (676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
+The spine protocol consisted of sagittal T1- and T2-weighted turbo-spin-echo sequences with rather large slice distance only sufficient for characterizing the lumbar spine \cite{Hegenscheid2013}.
+
+\subsubsection{Data Preprocessing} \label{application:Data Preprocessing}
+The data are preprocessed as described in Section~\ref{Data Preprocessing}.
+%
+\paragraph{Non-Image Data.}
+
+To ensure a fast and easy data access outside of statistical processors like \texttt{SPSS}, the data was exported to the JSON file format.
+%
+Each feature is stored as an object containing: 
+\begin{itemize}
+	\item the data as array of values--categorical data and error codes are stored using IDs,
+	\item the data type (continuous, nominal, ordinal, dichotomous),
+	\item a detailed description of the feature, and
+	\item the data dictionary translating value or error IDs to the actual values.
+\end{itemize}
+%
+% Binning: https://www.researchgate.net/post/What_metric_best_captures_the_information_loss_when_binning_a_large_data-set_into_a_histogram_representation
+Continuous variables are discretized to allow for \emph{Cram\'{e}r's V} contingency coefficient assessment.
+%
+In consultation with the collaborating epidemiologists, we set the number of groups to five, to allow for contingency assessment.
+
+\paragraph{Image Data.} \label{Image-Data}
+The lumbar spine was detected in the image data using a hierarchical finite element method by Rak et al. \cite{Rak2013}.
+%
+This semi-automatic method requires the user to initialize the tetrahedron-based finite element models (FEM) with a click on the L3 vertebra.
+%
+Two user-defined landmarks on the top and bottom of the L3 vertebra are used to obtain an initial height estimation of the model.
+%
+It uses a weighted sum of T1- and T2-weighted MR images to detect the lumbar spine shape.
+%
+The registered models capture resilient information about the shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae \cite{Klemm2013VMV}.
+%
+Due to incorrect initialization, strongly deformed spines, contrast differences and artifacts, the model was not able to detect lumbar spines for all subjects.
+%
+We obtained and worked with 983 tetrahedron models of the lumbar spine.
+%
+For clustering purposes, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature \cite{Klemm2013VMV}).
+
+\subsubsection{Shape Visualization and Clustering}
+%
+The tetrahedron-based detection model described in Section~\ref{application:Data Preprocessing} consists of corresponding grid points for each structure instance.
+%
+This allows for calculation of shape distance and similarity.
+%
+This information is used to calculate mean shapes as described in Section~\ref{Interaction- and Visualization Techniques}.
+
+Shape distance is mapped onto color.
+%
+For dichotomous variables, the color codes distances between mean shapes of the two groups, for variables with more than two manifestations it encodes the distance to the global mean shape of all subjects (see Figure~\ref{fig:hypothesisfree}).
+\\\\
+Shape-based clustering is carried out via agglomerative hierarchical clustering of the spine canal centerlines (recall Section~\ref{application:Data Preprocessing} and \cite{Klemm2013VMV}).
+%
+Since the ``correct'' number of clusters in a given group is unknown, an estimate is computed by means of the knee/elbow method \cite{Salvador2004}.
+%
+The method has proven to produce sound results on a preliminary data set and was able to reproduce textbook knowledge \cite{Klemm2013VMV}.
+\subsection{Domain Experts, Setup and Procedure}
+%
+%Lam et al. divide evaluations of Visual Analysis systems in two categories, 
+%
+\com{Anmerkung von Steffen einbauen: Evaluierung geht um fachliches Feedback - Wissen über epidemiologie UND rücken und wissen um die Daten - deswegen schwer ganz viele Experten ranzuziehen}
+Guided by the work of Lam et al \cite{Lam2012}, we conducted an investigation of \emph{Visual Data Analysis and Reasoning} (VDAR).
+%
+This approach aims to characterizes the ability of a system to explore data, discover knowledge, generate hypotheses and help formulating decisions.
+%
+Since it is hard to quantify when these goals are achieved, Lam et al. suggest case studies for the VDAR analysis.
+
+We conducted a case study with two epidemiological domain experts.
+%
+Both of them are practicing clinicians with focus on epidemiological research.
+%
+The first clinician is an specialist in internal medicine and also the leader of the SHIP.
+%
+One clinician is an radiologist and responsible for the image acquisition of the SHIP and has a deep understanding on the available image data.
+%
+\paragraph{Setup.} To bypass the large geographical distance between us and the domain experts and to take their busy schedule into account, the case study was done completely web based.
+%
+Since one major reason for the focus on web technologies, no software needed to be exchanged.
+%
+A website link was send to the users, which they could open with their browser of choice and start using the tool.
+%
+To make the user input comprehensible, we observed their monitor via screen sharing.
+%
+Communication was enabled via webcam supported voice over ip.
+%
+The total setup time including installing the screen sharing application was about five minutes.
+%
+Video-recordings of the sessions made it possible to evaluate the sessions later on.
+\com{TODO: Name der Screensharing Software mit angeben (Teamviewer)?}
+\paragraph{Procedure.}
+At first, controlled mouse and keyboard of the clinicians pc and demonstrated the basic functionalities of the system.
+%
+This included the Heatmap, the correlation view, how to introduce parameter visualization, how the color coding of differences works and the pivot tables.
+%
+As they understood the concepts, we handed over the mouse and keyboard control and only observed from this point on.
+%
+The epidemiologists were given two tasks: one hypothesis-free analysis of the data and one starting with an assumption.
+
+% \subsection{Exploratory Analysis of the Lumbar Spine Data Set}
+\subsection{Hypothesis-free analysis}
+Analyzing the data set without prior hypothesis requires a starting point giving an overview over the data first \cite{Shneiderman1996}.
+%
+With our tool, there are two ways to achieve this.
+%
+Performing a \emph{local investigation} by clustering all subjects or use the contingency matrix for a \emph{multivariate analysis}.
+%
+The latter was chosen by both domain experts.
+%
+They were excited by the possibility to look at all variables in the context of each other.
+%
+They stated that it "illumunates the black box of the data", making it possible to look at the data in an unbiased way.
+%
+Due to the dense mapping of information in the contingency matrix it was suggested to make this visualization full screen.
+%
+The radiologist was looking for correlations with shape related features in the data, finding that they correlate strong with each other as well as with leg pain, age, body height and hormone replacement therapy.
+%
+
+After this initial overview, both experts performed a \emph{multivariate analysis} by introducing features, such as age, waist circumference, weight or lumbar spine centerline curvature as bar charts views into the canvas area and selected subgroups to see how they are distributed and if they could observe unusual behavior in the mean-shapes.
+%
+
+% TODO: Contingency Matrix or Adjecency Matrix
+\com{00.44 Clustering-Ergebnis Weiblich und Arbeitspensum Alter}
+
+\subsection{Hypothesis-based analysis}
+The radiologist performed a \emph{local investigation} by dragging the age variable into the canvas area, observing that age-bins 36-52 years and 52-68 years are nearly identical to the global mean, while 20-36 years and 68-84 years differ in size.
+%
+Performing a \emph{multivariate analysis} by selecting only female subjects via the additionally added gender feature it was observed, that distance was closest to subjects of 68-84 years old and largest to 20-36 years old.
+%
+The expert argued that this is most likely due to the shrinking of the spine with increasing age, making it more similar to the female average spine which is smaller than the global average \com{(00.29)}.
+%
+
+A second analysis was carried out on body weight.
+%\com{Video 00.29. Analysis of Gender and Age - older subjects are closer to females, younger to males. This is due to shrinking of the spine in the aging process.}
+\com{00.46 Second: Analysis of Female subjects via clustering, selction of cluster 2, analysis using Body weight and Age.
+Clustering Result if 1 gives large distance between other subjects.
+00.49 How are can the differences be explained?}
+\com{00.46 Feedback Body-Size/Body weight.
+Some features are sparse, suggestion was made to use quantile/quintile, resulting in equally occupied results.
+This also depends on what the user is intersted in - for clinical use the outliers may be interesting (57.30).
+(58.30) From perspective of a radiologists, quartile and quintiles make sense when establishing reference values for diseases; clinical relevant parameter, einflussfaktoren oder risikofaktoren zu identifizieren, je nach dem welche hypothese man verwendet.}
+\com{Backpain three month > age.
+There are more subjects reporting backpain than those who dont.
+Since influences heavily the average shape - since most people have back pain, the average shape already is associated with back pain, which makes it difficult to extract parameters which reliably describe back pain.
+(1.05) The radiologist pointed out that for the mosaic view it would be beneficial to abstract the shape and curvature and categorize it into shapes s, i form.
+It was also pointed out that in the plot back pain>age, back pain can not be differenciated because the difference is mainly encoded via the age, more variables need to be added (gender, use of pain medications.
+this would be good for future work, providing a better way to subdivide the data set and construct groups.
+The presented method is very good to gain an overview).}
+\subsection{Lessons Learned}
+\com{ToDo: Noch einmal auf die VDAR Results eingehen}
+%
+The radiologist pointed out the unusual strong similarity of the L3 vertebrae throughout the population.
+%
+The medical explanation is that it represents a angular point of curvature of the lumbar spine.
+%
+A seconds explanation is the use of the L3 vertebra is initialization point of the lumbar spine model \com{(00.54)}.
+%
+The domain experts see another great potential in communicating insights efficiently using the presented visualizations \com{(1.08)}.
+%
+The radiologist pointed out that the image data representations would be better comparable if they would be abstracted into representative land marks, such as centers of the vertebrae and cardinal points of the lumbar spine canal.
+%
+Both experts emphasized the potential of the \emph{multivariate analysis} capabilities of the adjacency matrix for getting insight into a large amount of features simultaneously.
+%
+It is also useful to verify established but still controversial risk factors, such as the metabolic syndrome for coronary heart disease and whether the data set provides even more suitable risk factors.
+%
+Creating adjacency matrices for sub groups, such as different age bins can help to characterize the aging process by deriving age-specific risk factors.
+%
+%She also pointed out the benefits of an automatic categorization of the subjects into shape groups.%$$, which allows for even faster comparison.
+\com{Employ a more abstract visualization of the spine to capture reliable information and enable better comparison between subjects}
+%
+%Since many large population studies incorporate medical image data this gains importance.
+%\com{1.08 It was pointed out that this technique is also good for epidemiological publication because it is often more insigthfull to show these images instead of long tables. }
+\com{ToDo Fazit über Web-Austausch, das Experten begeistert davon waren und dass es hervorragend funktioniert.
+Kleine Änderungen können sogar live programmiert werden ohne das etwas neu kompiliert werden muss (hier wäre ein Beispiel nett, z.B.
+Größe der Schrift irgendwo.
+Außerdem bringt der Browser viele Funktionen mit, wie z.B.
+die Suchfunktionen.)}
+
+\com{Small User Interface Suggestions: Search-Bar for Subjects, making the elements of the contingency matrix selectable; See which Subject is currently selected}
+\com{Enhancement contingency matrix: Display exact value of correlations, also information about distribution. More focus on this}
+\com{Außerdem ein Anwendungsgebiet für die Matrix: Einschätzen metabolisches syndrom, eine Menge von Risikofaktoren die für Koronare Herzkrankheit Marker darstellen und die bereits länger in der Kritik sind.
+Hier kann getestet werden, ob es hierfür nicht bessere Indikatoren noch geben kann!
+(39.32)}
+%\com{Also mosaic plots can be incorporated for several sub groups to see where the main differences are (for different age groups to characterize age)}
+\com{1.20 More exact segmentations are useful since curvature changes are often the last change happening in a pathological process, after local deformations of the vertebrae}
+
+% ---------------------------------------------------------------------------------------------------------------------------------------
+% ---------------------------------------------------------------------------------------------------------------------------------------
+% ---------------------------------------------------------------------------------------------------------------------------------------
+\section{Old Application Section}
+%
 This section describes how our presented \emph{IVA} workflow is used for an epidemiological use case.
@@
 \section{Summary and Conclusion}
+% TODO: Future work Bulletliste zusammenkürzen
 In this paper, we presented an \emph{IVA} framework for the analysis of complex image-centric epidemiological data.

### 2014-06-17 draft.tex
continued rewrite of application section

@@
 %% indicate IEEE Member or Student Member in form indicated below
-\author{P. Klemm$^{1}$, K. Lawonn$^{1}$, M. Rak$^{1}$, B. Preim$^{1}$, K. Toennies$^{1}$, K. Hegenscheid$^{2}$, H. Völzke$^{2}$, S. Oeltze$^{1}$}
+\author{P. Klemm$^{1}$, S. Oeltze$^{1}$, K. Lawonn$^{1}$, K. Hegenscheid$^{2}$, H. V{\"o}lzke$^{2}$, B. Preim$^{1}$}
 \authorfooter{
@@
 %
+\com{Multivariate analysis using the adjacency matrix may have also be explained here, depending on the Application chapter.}
+%
 The feature is mapped using an automatically chosen visualization appropriate for its data type (described in detail in the following section).
@@
 \section{Application}
-\add{This Chapter needs to be reworked completely, since it was subject of the reviewer criticism.}
 This section describes how our presented \emph{IVA} workflow is used for an epidemiological use case.
@@
 
-We conducted a case study with two epidemiological domain experts.
+We conducted a qualitative case study with two epidemiological domain experts.
 %
@@
 % \subsection{Exploratory Analysis of the Lumbar Spine Data Set}
-\subsection{Hypothesis-free analysis}
+\subsection{Case 1: Hypothesis-free analysis}
 Analyzing the data set without prior hypothesis requires a starting point giving an overview over the data first \cite{Shneiderman1996}.
@@
 Performing a \emph{local investigation} by clustering all subjects or use the contingency matrix for a \emph{multivariate analysis}.
-%
-The latter was chosen by both domain experts.
-%
-They were excited by the possibility to look at all variables in the context of each other.
-%
-They stated that it "illumunates the black box of the data", making it possible to look at the data in an unbiased way.
+
+\paragraph{Analysis 1.}
+%
+The latter was chosen first by both domain experts.
+%
+%They pointed out the high potential they see in this visualization, because it allows to look at the data in an unbiased way.
+They pointed out the benefits of the possibility to look at all variables in the context of each other.
+%
+To cite the radiologist, it "illumunates the black box of the data", making it possible to look at the data in an unbiased way.
 %
@@
 %
-
-After this initial overview, both experts performed a \emph{multivariate analysis} by introducing features, such as age, waist circumference, weight or lumbar spine centerline curvature as bar charts views into the canvas area and selected subgroups to see how they are distributed and if they could observe unusual behavior in the mean-shapes.
-%
-
-% TODO: Contingency Matrix or Adjecency Matrix
-\com{00.44 Clustering-Ergebnis Weiblich und Arbeitspensum Alter}
-
-\subsection{Hypothesis-based analysis}
-The radiologist performed a \emph{local investigation} by dragging the age variable into the canvas area, observing that age-bins 36-52 years and 52-68 years are nearly identical to the global mean, while 20-36 years and 68-84 years differ in size.
-%
-Performing a \emph{multivariate analysis} by selecting only female subjects via the additionally added gender feature it was observed, that distance was closest to subjects of 68-84 years old and largest to 20-36 years old.
-%
-The expert argued that this is most likely due to the shrinking of the spine with increasing age, making it more similar to the female average spine which is smaller than the global average \com{(00.29)}.
-%
-
-A second analysis was carried out on body weight.
-%\com{Video 00.29. Analysis of Gender and Age - older subjects are closer to females, younger to males. This is due to shrinking of the spine in the aging process.}
-\com{00.46 Second: Analysis of Female subjects via clustering, selction of cluster 2, analysis using Body weight and Age.
-Clustering Result if 1 gives large distance between other subjects.
-00.49 How are can the differences be explained?}
-\com{00.46 Feedback Body-Size/Body weight.
-Some features are sparse, suggestion was made to use quantile/quintile, resulting in equally occupied results.
-This also depends on what the user is intersted in - for clinical use the outliers may be interesting (57.30).
-(58.30) From perspective of a radiologists, quartile and quintiles make sense when establishing reference values for diseases; clinical relevant parameter, einflussfaktoren oder risikofaktoren zu identifizieren, je nach dem welche hypothese man verwendet.}
-\com{Backpain three month > age.
-There are more subjects reporting backpain than those who dont.
-Since influences heavily the average shape - since most people have back pain, the average shape already is associated with back pain, which makes it difficult to extract parameters which reliably describe back pain.
-(1.05) The radiologist pointed out that for the mosaic view it would be beneficial to abstract the shape and curvature and categorize it into shapes s, i form.
-It was also pointed out that in the plot back pain>age, back pain can not be differenciated because the difference is mainly encoded via the age, more variables need to be added (gender, use of pain medications.
-this would be good for future work, providing a better way to subdivide the data set and construct groups.
-The presented method is very good to gain an overview).}
-\subsection{Lessons Learned}
-\com{ToDo: Noch einmal auf die VDAR Results eingehen}
-%
-The radiologist pointed out the unusual strong similarity of the L3 vertebrae throughout the population.
-%
-The medical explanation is that it represents a angular point of curvature of the lumbar spine.
-%
-A seconds explanation is the use of the L3 vertebra is initialization point of the lumbar spine model \com{(00.54)}.
-%
-The domain experts see another great potential in communicating insights efficiently using the presented visualizations \com{(1.08)}.
-%
-The radiologist pointed out that the image data representations would be better comparable if they would be abstracted into representative land marks, such as centers of the vertebrae and cardinal points of the lumbar spine canal.
-%
-Both experts emphasized the potential of the \emph{multivariate analysis} capabilities of the adjacency matrix for getting insight into a large amount of features simultaneously.
-%
-It is also useful to verify established but still controversial risk factors, such as the metabolic syndrome for coronary heart disease and whether the data set provides even more suitable risk factors.
-%
-Creating adjacency matrices for sub groups, such as different age bins can help to characterize the aging process by deriving age-specific risk factors.
-%
-%She also pointed out the benefits of an automatic categorization of the subjects into shape groups.%$$, which allows for even faster comparison.
-\com{Employ a more abstract visualization of the spine to capture reliable information and enable better comparison between subjects}
-%
-%Since many large population studies incorporate medical image data this gains importance.
-%\com{1.08 It was pointed out that this technique is also good for epidemiological publication because it is often more insigthfull to show these images instead of long tables. }
-\com{ToDo Fazit über Web-Austausch, das Experten begeistert davon waren und dass es hervorragend funktioniert.
-Kleine Änderungen können sogar live programmiert werden ohne das etwas neu kompiliert werden muss (hier wäre ein Beispiel nett, z.B.
-Größe der Schrift irgendwo.
-Außerdem bringt der Browser viele Funktionen mit, wie z.B.
-die Suchfunktionen.)}
-
-\com{Small User Interface Suggestions: Search-Bar for Subjects, making the elements of the contingency matrix selectable; See which Subject is currently selected}
-\com{Enhancement contingency matrix: Display exact value of correlations, also information about distribution. More focus on this}
-\com{Außerdem ein Anwendungsgebiet für die Matrix: Einschätzen metabolisches syndrom, eine Menge von Risikofaktoren die für Koronare Herzkrankheit Marker darstellen und die bereits länger in der Kritik sind.
-Hier kann getestet werden, ob es hierfür nicht bessere Indikatoren noch geben kann!
-(39.32)}
-%\com{Also mosaic plots can be incorporated for several sub groups to see where the main differences are (for different age groups to characterize age)}
-\com{1.20 More exact segmentations are useful since curvature changes are often the last change happening in a pathological process, after local deformations of the vertebrae}
-
-% ---------------------------------------------------------------------------------------------------------------------------------------
-% ---------------------------------------------------------------------------------------------------------------------------------------
-% ---------------------------------------------------------------------------------------------------------------------------------------
-\section{Old Application Section}
-%
-This section describes how our presented \emph{IVA} workflow is used for an epidemiological use case.
-%
-We applied the presented set of techniques to a data set compiled to analyze lower back pain. 
-%
-This is one of the most common reasons for an adult to see a physician in the Western civilization \cite{Backpain}.
-%
-Epidemiological analysis of lumbar back pain, such as the work of Harreby et al. \cite{Harreby1996}, is largely focused on non-image information.
-%
-In comparable studies, only a few shape-related features are included \cite{Lang2011}.
-%
-To our knowledge, this is the first approach, where shape-related information of the whole lumbar spine is analyzed along with other epidemiological features.
-%
-Determining risk factors in this area can lead to \cite{Fletcher2012}:
-\begin{itemize}
-	\item a better understanding of effects of preventive measures such as occupational health and safety regulations,
-	\item prognostic features for diagnosis and treatment of lumbar back pain, and
-	\item determination of particularly affected risk groups.
-\end{itemize}
-% Determining risk factors in this area can lead to \cite{Fletcher2012} (evtl. weglassen, Dopplung bei Epidemiological Background?):
-% \begin{itemize}
-% 	\item a better understanding of effects of preventive measures such as occupational health and safety regulations
-% 	\item prognostic features for diagnosis and treatment of lumbar back pain
-% 	\item determination of particularly effected risk groups
-% \end{itemize}
-%
-Characterizing the healthy aging process of the spine is a long-term goal for determining age-normalized probabilities for spine-related diseases by incorporating individual risk factors.
-
-Data confidentiality and ethical reasons prohibit us from accessing the complete \texttt{SHIP} feature space.
-%
-Our epidemiological collaborators compiled a feature list, which is a tradeoff between complexity and limitations of the responsible ethics committee.
-%
-\subsection{The Lumbar Spine Data Set}
-We divide the data set into image and non-image data.
-%
-There are 127 features describing diagnosed diseases, lifestyle factors, women-specific factors, pain indicators, laboratory values and somatometric features.
-%
-The image data was acquired on a 1.5~Tesla scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany) by four trained technicians in a standardized way.
-%
-%The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence (676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
-The spine protocol consisted of sagittal T1- and T2-weighted turbo-spin-echo sequences with rather large slice distance only sufficient for characterizing the lumbar spine \cite{Hegenscheid2013}.
-
-\subsection{Data Preprocessing} \label{application:Data Preprocessing}
-The data are preprocessed as described in Section~\ref{Data Preprocessing}.
-%
-\paragraph{Non-Image Data.}
-
-To ensure a fast and easy data access outside of statistical processors like \texttt{SPSS}, the data was exported to the JSON file format.
-%
-Each feature is stored as an object containing: 
-\begin{itemize}
-	\item the data as array of values--categorical data and error codes are stored using IDs,
-	\item the data type (continuous, nominal, ordinal, dichotomous),
-	\item a detailed description of the feature, and
-	\item the data dictionary translating value or error IDs to the actual values.
-\end{itemize}
-%
-% Binning: https://www.researchgate.net/post/What_metric_best_captures_the_information_loss_when_binning_a_large_data-set_into_a_histogram_representation
-Continuous variables are discretized to allow for \emph{Cram\'{e}r's V} contingency coefficient assessment.
-%
-In consultation with the collaborating epidemiologists, we set the number of groups to five, to allow for contingency assessment.
-
-\paragraph{Image Data.} \label{Image-Data}
-The lumbar spine was detected in the image data using a hierarchical finite element method by Rak et al. \cite{Rak2013}.
-%
-This semi-automatic method requires the user to initialize the tetrahedron-based finite element models (FEM) with a click on the L3 vertebra.
-%
-Two user-defined landmarks on the top and bottom of the L3 vertebra are used to obtain an initial height estimation of the model.
-%
-It uses a weighted sum of T1- and T2-weighted MR images to detect the lumbar spine shape.
-%
-The registered models capture resilient information about the shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae \cite{Klemm2013VMV}.
-%
-Due to incorrect initialization, strongly deformed spines, contrast differences and artifacts, the model was not able to detect lumbar spines for all subjects.
-%
-We obtained and worked with 983 tetrahedron models of the lumbar spine.
-%
-For clustering purposes, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature \cite{Klemm2013VMV}).
-
-\subsection{Shape Visualization and Clustering}
-%
-The tetrahedron-based detection model described in Section~\ref{application:Data Preprocessing} consists of corresponding grid points for each structure instance.
-%
-This allows for calculation of shape distance and similarity.
-%
-This information is used to calculate mean shapes as described in Section~\ref{Interaction- and Visualization Techniques}.
-
-Shape distance is mapped onto color.
-%
-For dichotomous variables, the color codes distances between mean shapes of the two groups, for variables with more than two manifestations it encodes the distance to the global mean shape of all subjects (see Figure~\ref{fig:hypothesisfree}).
-\\\\
-Shape-based clustering is carried out via agglomerative hierarchical clustering of the spine canal centerlines (recall Section~\ref{application:Data Preprocessing} and \cite{Klemm2013VMV}).
-%
-Since the ``correct'' number of clusters in a given group is unknown, an estimate is computed by means of the knee/elbow method \cite{Salvador2004}.
-%
-The method has proven to produce sound results on a preliminary data set and was able to reproduce textbook knowledge \cite{Klemm2013VMV}.
-
-\subsection{Exploratory Analysis of the Lumbar Spine Data Set}
-%
-Expert-guided analysis assessed the suitability of our approach for supporting both hypothesis-free analysis as well as hypothesis generation. 
-%
-The use of \emph{IVA} patterns as presented in Section~\ref{Image Centric Cohort Study Data in Interactive Visual Analysis Context} is denoted accordingly.
-%
-\subsubsection{Hypothesis-Free Analysis} \label{Hypothesis free Analysis}
+The internist inspected features which are not derived from the images, such as parameters strongly related with age, such as income, blood fat values or number of born children.
+%
+This also shows, that even though the visualization may treat every feature equally, the domain experts is more likely interested in features they are familiar with and have personal clinical experience with.
+%
+%Many correlations observed in this view
+
+After this initial overview, the expert performed a \emph{multivariate analysis} by introducing features, such as age, waist circumference, weight or lumbar spine centerline curvature as bar charts views into the canvas area and selected subgroups to see how they are distributed and if they could observe unusual behavior in the mean-shapes.
+%
++ Cramer test 
+\com{Allgemein: Workflow noch einmal grob beschreiben - Overview first, details on demand}
+%
+\paragraph{Analysis 2.}
 \begin{figure}[htb]
@@
 \end{figure}
-%
-Analyzing the data set without prior hypothesis requires a starting point giving an overview over the data first \cite{Shneiderman1996}.
-%
-Shape-based hypothesis-free exploration starts with the \emph{local investigation} pattern by a shape grouping step using shape-based clustering.
+\com{ToDo: Rewrite}
+%
+The internist started also with a \emph{multivariate analysis} using the contingency table, but found no associations he was interested in.
+%
+Therefore, he applied the \emph{local investigation} pattern by a shape grouping step using shape-based clustering via dragging the "All subjects" feature into the canvas area, triggering the shape clustering.
+%Shape-based hypothesis-free exploration starts with the \emph{local investigation} pattern by a shape grouping step using shape-based clustering.
 %
@@
 %
-Performing a \emph{multivariate analysis} by looking at \emph{Cram\'{e}r's V} contingency values of cluster~8 reveals associations of this group with employment status, gender, body size, age, thyroid nodules and blood fat value.
+The user opened the contingency pane to performing a \emph{multivariate analysis} by looking at \emph{Cram\'{e}r's V} contingency values of cluster~8.
+%
+This revealed associations of this group with employment status, gender, body size, age, thyroid nodules and blood fat value.
 %
@@
 %
-\add{It S}hows that cluster~8 contains mostly women and also has a larger unemployment rate (see Figure~\ref{fig:hypothesisfree}), while the overall employment rate of women and men in the data set is almost exactly 50\%.
+\add{It s}hows that cluster~8 contains mostly women and also has a larger unemployment rate (see Figure~\ref{fig:hypothesisfree}), while the overall employment rate of women and men in the data set is almost exactly 50\%.
 %
@@
 The employment status is a feature relating to many different lifestyle factors such as income or nutrition as well as age and might act as confounder.
-%TODO Implement Color coding only in a certain direction!\\
-\\\\
-While this approach does not assume any hypothesis, the 127 epidemiological features were selected by domain experts with a wide range on potential influential factors.
-%
-This filtering step based on expert experience rules out possible confounder and useless associations and it is therefore arguable whether this represents a disadvantage \cite{Wiley2008}.
-%It is arguable whether this first filtering step, which is purely based on expert experience, represents a disadvantage, since it rules out many parameters possibly associated with the presented features, but the value of this knowledge would be small \cite{Wiley2008}.
-%
-% While this approach does not assume any hypothesis, the data are met with a selection bias when compiling the list of related features by the domain experts.
-% %
-% It is arguable if this first filtering step, which is purely based on expert experience, is really a disadvantage, since it rules out many parameters which may interact with the presented features but the value of this knowledge would be small \cite{Wiley2008}.
-
-\subsubsection{Hypothesis-Based Analysis} \label{Hypothesis based Analysis}
+
+% TODO: Contingency Matrix or Adjecency Matrix
+\com{00.44 Clustering-Ergebnis Weiblich und Arbeitspensum Alter}
+
+\subsection{Case 2: Hypothesis-based analysis}
+If the user proposes a hypothesis about a relation between a non-image feature regarding shape, the workflow slightly differs from the hypothesis-free analysis.
+%
+The starting point follows the \emph{feature investigation} pattern, where a feature of interest is selected by dragging it into the canvas area and viewing the subject's distribution as well as their shape differences.
+\paragraph{Analysis 1}
+The radiologist performed a \emph{feature investigation} by dragging the age variable into the canvas area, observing that age-bins 36-52 years and 52-68 years are nearly identical to the global mean, while 20-36 years and 68-84 years differ in size.
+%
+Performing a \emph{multivariate analysis} by selecting only female subjects via the additionally added gender feature it was observed, that distance was closest to subjects of 68-84 years old and largest to 20-36 years old.
+%
+The expert argued that this is most likely due to the shrinking of the spine with increasing age, making it more similar to the female average spine which is smaller than the global average \com{(00.29)}.
+%
+
+\paragraph{Analysis 2.}
+Hypothesis: "Back pain is associated with age and therefore there differences in the lumbar spine shape are observable." \com{TODO Image Mosaic Plot Age, Backpain}.
+To validate this hypothesis, a \emph{multivariate analysis} was performed by introducing the dichotomous feature "\emph{Did you experience backpain in the last three moth?}" together with the age as mosaic plot by dropping both features on the canvas area.
+%
+The experts were not able to observe the expected effect in the visualization.
+%
+Reasons for this are twofold.
+%
+For one, age influences the lumbar spine shape very heavily, as seen in Fig~\ref{?}, while the differences between subjects with and without back pain are small.
+%
+The major difference seen in the visualization are therefore presented by the age feature, masking differences via the back pain parameter.
+
+The second explanation is the commonality of back pain in our society.
+%
+As seen in Fig~\ref{?}, subjects reporting back pain are even the majority, which makes it difficult to extract parameters which reliably describe back pain.
+%
+A \emph{local investigation} using the contingency table showed a strong association between back pain with gender and body height.
+%
+Body height was explained to be a confounder for gender, since female subjects smaller on average than male subjects.
+%
+The analysis solely based on shape only accentuated gender differences, not in back pain.
+%This makes it hard to analyze shape related features for this disease
+%
+A next step would be to use a variable describing the strength of back pain and only analyzing strong back pain!
+%
+
+\com{ToDo: Ability to better create sub groups would be useful as well as better segmentation}
+%gender, use of pain medications. this would be good for future work, providing a better way to subdivide the data set and construct groups. The presented method is very good to gain an overview
+
+%
+%\com{Video 00.29. Analysis of Gender and Age - older subjects are closer to females, younger to males. This is due to shrinking of the spine in the aging process.}
+%\com{Backpain three month > age.
+There are more subjects reporting backpain than those who dont.
+Since influences heavily the average shape - since most people have back pain, the average shape already is associated with back pain, which makes it difficult to extract parameters which reliably describe back pain.
+(1.05) The radiologist pointed out that for the mosaic view it would be beneficial to abstract the shape and curvature and categorize it into shapes s, i form.
+It was also pointed out that in the plot back pain>age, back pain can not be differenciated because the difference is mainly encoded via the age, more variables need to be added (gender, use of pain medications.
+this would be good for future work, providing a better way to subdivide the data set and construct groups.
+The presented method is very good to gain an overview).}
+
+\paragraph{Analysis 3}
+\com{ToDo: Needs to be rewritten}
 \begin{figure}[htb]
@@
 %
-If the user proposes a hypothesis about a relation between a non-image feature regarding shape, the workflow slightly differs from the hypothesis-free analysis.
-%
-The starting point follows the \emph{feature investigation} pattern, where a feature of interest is selected by dragging it into the canvas area and viewing the subject's distribution as well as their shape differences.
-%
 In our use case, epidemiologists were interested in the dichotomous questionnaire answer to ``\emph{Did you experience back pain in the past three months?}''.
@@
 
-\subsubsection{Follow-Up Tasks and Concluding Domain-Expert Feedback}
-%
-Correlation does not automatically imply causation. %\cite{Tufte2003}.
-%
-In particular, when large amounts of attributes are automatically checked for strong correlations, just by chance hundreds of them reach at least a moderately significant correlation, such as $p>0.05$.
-%Defining a causal relationship solely based on observed correlations of two features is \emph{cum hoc ergo propter hoc}--correlation does not automatically imply causation \cite{Tufte2003}.
-%
-The observed correlations need to be carefully checked for confounders and medical soundness.
-%
-Statisticians validate causal inferences of the drawn conclusions.
-
-Features that potentially interact with a disease-related condition need to be validated.
-%
-To increase the probability of the observation not to be random and to avoid \emph{overfitting}, they are cross-checked for associations in \texttt{SHIP-TREND} as a second, independent population sample, which was examined with methods identical to \texttt{SHIP}.
-\\\\
-The prototype was demonstrated to the domain experts (with background on epidemiology and radiology) via screen sharing.
-%
-The software was operated by a computer scientist following instructions of the domain experts.
-%
-The presented methods guide the attention to features that are not in the focus of attention and expectance of clinical researchers.
-%
-The explorative nature of the methods work well for gathering associations possibly acting as confounder, as outcome of a disease or as an actual cause or risk factor.
-%"\emph{Correlation does not imply causation}" is a famous saying in statistics, meaning that correlation does not necessarily imply causal relationships \cite{}.
-%
-This distinction is hard to make and requires a lot of clinical experience.
-%
-The combination of multiple views with shape information helps to connect many different information sources to make the large information spaces cognitively feasible.
-%
-Displaying MRI scans for the subjects in the outlier cluster is promising because they are highly likely to exhibit pathologies.
-%
-%We plan to include a \texttt{DICOM}-viewer to meet this wish.
-
-% - Describe steps from gathering Information from the raw image files (segmentation, abstraction, visualization)\\
-% - Problem of sparse differences - Visualization has to be more abstract to emphasize the differences.\\
-% - Input of Epidemiologists goes here!\\
+\com{00.46 Second: Analysis of Female subjects via clustering, selction of cluster 2, analysis using Body weight and Age.
+Clustering Result if 1 gives large distance between other subjects.
+00.49 How are can the differences be explained?}
+\com{00.46 Feedback Body-Size/Body weight.
+Some features are sparse, suggestion was made to use quantile/quintile, resulting in equally occupied results.
+This also depends on what the user is intersted in - for clinical use the outliers may be interesting (57.30).
+(58.30) From perspective of a radiologists, quartile and quintiles make sense when establishing reference values for diseases; clinical relevant parameter, einflussfaktoren oder risikofaktoren zu identifizieren, je nach dem welche hypothese man verwendet.}
+
+% ---------------------------------------------------------------------------------------------------------
+\subsection{Further Feedback and Lessons Learned}
+\com{ToDo: Noch einmal auf die VDAR Results eingehen}
+%\com{ToDo: Lessons Learned der einzelnen Cases - Fokus bei Hypothesenfreier Analyse muss etwa eine gute Übersichtsvisualisierung sein, da es genau das ist was den Epidemiologen hierzu im Moment fehlt}
+%
+Both domain experts concluded positively over the benefits of the \emph{IVA} approach when analyzing image based cohort study data.
+%
+One expert emphasized the way the image data is included into information visualizations which comes much more natural to her due to her background in radiology.
+%
+%Both experts quickly understood the basic concepts of the visualizations and were able to 
+%
+The domain experts see another great potential in communicating insights efficiently using the presented visualizations \com{(1.08)}.
+%
+The experts proposed to improve the visual comparison between image data representations by abstracting them into representative land marks, such as centers of the vertebrae and cardinal points of the lumbar spine canal.
+%
+\paragraph{Multivariate analysis is most important for hypothesis generation.}
+Both experts emphasized the potential of the \emph{multivariate analysis} capabilities of the adjacency matrix for getting insight into a large amount of features simultaneously.
+%
+It is also useful to verify established but still controversial risk factors, such as the metabolic syndrome for coronary heart disease and whether the data set provides even more suitable risk factors \com{(39.32)}.
+%
+Creating adjacency matrices for sub groups, such as different age bins can help to characterize the aging process by deriving age-specific risk factors.
+%
+%\paragraph{Hypothesis driven analysis would benefit from more advanced grouping.}
+%While the existing brushing and linking techniques already allow the comparison of image data 
+\com{(TODO weitere Unterteilung der Variablen und anschließende Anbindung an Statistik Tools)}
+
+\paragraph{Segmenation quality is crucial.}
+The radiologist pointed out the unusual strong similarity of the L3 vertebrae throughout the population.
+%
+The medical explanation is that it represents a angular point of curvature of the lumbar spine.
+%
+A seconds explanation is the use of the L3 vertebra is initialization point of the lumbar spine model \com{(00.54)}.
+
+The experts also emphasized that associations related to shape strongly depend on the segmentation quality.
+%
+The lumbar spine model used in this case study captures deformation of the spine canal well, but lacks precise definition in vertebrae height and shape.
+%
+Since deformation of the spine canal is commonly the last pathological stage of pathological lumbar spine deformation and is preceded by vertebrae deformation, the system would strongly benefit from more precise segmentation results capturing these prior changes.
+
+\paragraph{Usage of different Categorizations depending on expected outcome.}
+When categorizing numerical variables into equal groups it is possible to create very sparse categories due to outliers, for example when analyzing body weight.
+%
+These outliers are of high interest for finding pathological subjects, but not wanted when new parameters describing risk factors are established.
+%
+The experts therefore suggested two modes of the tool.
+%
+The outlier mode still creates categories of equal size, producing sparse categories for outliers.
+%
+Balanced categories are created on the second mode, which uses quartiles or quintiles to set borders between categories.
+\paragraph{Web technologies are well suited for rapid feedback.}
+The web-based approach for both implementing the prototype as well as getting feedback via voice over ip conference calls worked very well.
+%
+Since the software does not need to be compiled, small changes can even be made on the fly during a testing session.
+%
+The large data base associated with image based epidemiological data remains on the server machine and has not to be moved tediously using external hard disks.
+%\com{ToDo Fazit über Web-Austausch, das Experten begeistert davon waren und dass es hervorragend funktioniert.
+Kleine Änderungen können sogar live programmiert werden ohne das etwas neu kompiliert werden muss (hier wäre ein Beispiel nett, z.B.
+Größe der Schrift irgendwo.
+Außerdem bringt der Browser viele Funktionen mit, wie z.B.
+die Suchfunktionen.)}
+%She also pointed out the benefits of an automatic categorization of the subjects into shape groups.%$$, which allows for even faster comparison.
+%\com{Employ a more abstract visualization of the spine to capture reliable information and enable better comparison between subjects}
+%
+%Since many large population studies incorporate medical image data this gains importance.
+%\com{1.08 It was pointed out that this technique is also good for epidemiological publication because it is often more insigthfull to show these images instead of long tables.}
+
+%\com{Small User Interface Suggestions: Search-Bar for Subjects, making the elements of the contingency matrix selectable; See which Subject is currently selected}
+%\com{Enhancement contingency matrix: Display exact value of correlations, also information about distribution. More focus on this}
+%\com{Außerdem ein Anwendungsgebiet für die Matrix: Einschätzen metabolisches syndrom, eine Menge von Risikofaktoren die für Koronare Herzkrankheit Marker darstellen und die bereits länger in der Kritik sind.
+Hier kann getestet werden, ob es hierfür nicht bessere Indikatoren noch geben kann!
+(39.32)}
+%\com{Also mosaic plots can be incorporated for several sub groups to see where the main differences are (for different age groups to characterize age)}
+%\com{1.20 More exact segmentations are useful since curvature changes are often the last change happening in a pathological process, after local deformations of the vertebrae}
+
+% ---------------------------------------------------------------------------------------------------------------------------------------
+% ---------------------------------------------------------------------------------------------------------------------------------------
+% ---------------------------------------------------------------------------------------------------------------------------------------
+% \section{Old Application Section}
+% %
+% This section describes how our presented \emph{IVA} workflow is used for an epidemiological use case.
+% %
+% We applied the presented set of techniques to a data set compiled to analyze lower back pain. 
+% %
+% This is one of the most common reasons for an adult to see a physician in the Western civilization \cite{Backpain}.
+% %
+% Epidemiological analysis of lumbar back pain, such as the work of Harreby et al. \cite{Harreby1996}, is largely focused on non-image information.
+% %
+% In comparable studies, only a few shape-related features are included \cite{Lang2011}.
+% %
+% To our knowledge, this is the first approach, where shape-related information of the whole lumbar spine is analyzed along with other epidemiological features.
+% %
+% Determining risk factors in this area can lead to \cite{Fletcher2012}:
+% \begin{itemize}
+% 	\item a better understanding of effects of preventive measures such as occupational health and safety regulations,
+% 	\item prognostic features for diagnosis and treatment of lumbar back pain, and
+% 	\item determination of particularly affected risk groups.
+% \end{itemize}
+% % Determining risk factors in this area can lead to \cite{Fletcher2012} (evtl. weglassen, Dopplung bei Epidemiological Background?):
+% % \begin{itemize}
+% % 	\item a better understanding of effects of preventive measures such as occupational health and safety regulations
+% % 	\item prognostic features for diagnosis and treatment of lumbar back pain
+% % 	\item determination of particularly effected risk groups
+% % \end{itemize}
+% %
+% Characterizing the healthy aging process of the spine is a long-term goal for determining age-normalized probabilities for spine-related diseases by incorporating individual risk factors.
+% 
+% Data confidentiality and ethical reasons prohibit us from accessing the complete \texttt{SHIP} feature space.
+% %
+% Our epidemiological collaborators compiled a feature list, which is a tradeoff between complexity and limitations of the responsible ethics committee.
+% %
+% \subsection{The Lumbar Spine Data Set}
+% We divide the data set into image and non-image data.
+% %
+% There are 127 features describing diagnosed diseases, lifestyle factors, women-specific factors, pain indicators, laboratory values and somatometric features.
+% %
+% The image data was acquired on a 1.5~Tesla scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany) by four trained technicians in a standardized way.
+% %
+% %The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence (676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
+% The spine protocol consisted of sagittal T1- and T2-weighted turbo-spin-echo sequences with rather large slice distance only sufficient for characterizing the lumbar spine \cite{Hegenscheid2013}.
+% 
+% \subsection{Data Preprocessing} \label{application:Data Preprocessing}
+% The data are preprocessed as described in Section~\ref{Data Preprocessing}.
+% %
+% \paragraph{Non-Image Data.} 
+% 
+% To ensure a fast and easy data access outside of statistical processors like \texttt{SPSS}, the data was exported to the JSON file format.
+% %
+% Each feature is stored as an object containing: 
+% \begin{itemize}
+% 	\item the data as array of values--categorical data and error codes are stored using IDs,
+% 	\item the data type (continuous, nominal, ordinal, dichotomous),
+% 	\item a detailed description of the feature, and
+% 	\item the data dictionary translating value or error IDs to the actual values.
+% \end{itemize}
+% %
+% % Binning: https://www.researchgate.net/post/What_metric_best_captures_the_information_loss_when_binning_a_large_data-set_into_a_histogram_representation
+% Continuous variables are discretized to allow for \emph{Cram\'{e}r's V} contingency coefficient assessment.
+% %
+% In consultation with the collaborating epidemiologists, we set the number of groups to five, to allow for contingency assessment.
+% 
+% \paragraph{Image Data.} \label{Image-Data}
+% The lumbar spine was detected in the image data using a hierarchical finite element method by Rak et al. \cite{Rak2013}.
+% %
+% This semi-automatic method requires the user to initialize the tetrahedron-based finite element models (FEM) with a click on the L3 vertebra.
+% %
+% Two user-defined landmarks on the top and bottom of the L3 vertebra are used to obtain an initial height estimation of the model.
+% %
+% It uses a weighted sum of T1- and T2-weighted MR images to detect the lumbar spine shape.
+% %
+% The registered models capture resilient information about the shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae \cite{Klemm2013VMV}.
+% %
+% Due to incorrect initialization, strongly deformed spines, contrast differences and artifacts, the model was not able to detect lumbar spines for all subjects.
+% %
+% We obtained and worked with 983 tetrahedron models of the lumbar spine.
+% %
+% For clustering purposes, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature \cite{Klemm2013VMV}).
+% 
+% \subsection{Shape Visualization and Clustering}
+% %
+% The tetrahedron-based detection model described in Section~\ref{application:Data Preprocessing} consists of corresponding grid points for each structure instance.
+% %
+% This allows for calculation of shape distance and similarity.
+% %
+% This information is used to calculate mean shapes as described in Section~\ref{Interaction- and Visualization Techniques}.
+% 
+% Shape distance is mapped onto color.
+% %
+% For dichotomous variables, the color codes distances between mean shapes of the two groups, for variables with more than two manifestations it encodes the distance to the global mean shape of all subjects (see Figure~\ref{fig:hypothesisfree}).
+% \\\\
+% Shape-based clustering is carried out via agglomerative hierarchical clustering of the spine canal centerlines (recall Section~\ref{application:Data Preprocessing} and \cite{Klemm2013VMV}).
+% %
+% Since the ``correct'' number of clusters in a given group is unknown, an estimate is computed by means of the knee/elbow method \cite{Salvador2004}.
+% %
+% The method has proven to produce sound results on a preliminary data set and was able to reproduce textbook knowledge \cite{Klemm2013VMV}.
+% 
+% \subsection{Exploratory Analysis of the Lumbar Spine Data Set}
+% %
+% Expert-guided analysis assessed the suitability of our approach for supporting both hypothesis-free analysis as well as hypothesis generation. 
+% %
+% The use of \emph{IVA} patterns as presented in Section~\ref{Image Centric Cohort Study Data in Interactive Visual Analysis Context} is denoted accordingly.
+% %
+% \subsubsection{Hypothesis-Free Analysis} \label{Hypothesis free Analysis}
+% \begin{figure}[htb]
+%  \centering
+%  %\includegraphics[width=1.5in]{figures/InteractionLoop}
+%  \includegraphics[width=3.5in]{figures/hypothesisfree}
+%  \caption{(a) Clustering result of all subjects.
+The bar chart height indicates the number of subjects in the cluster.
+The difference to the mean shape is color-coded, whereas blue represents no difference and red a large difference.
+(b) A pivot table relates cluster results to gender and employment status.
+Cluster 8 is striking because of the high proportion of women, while cluster 2, 3, 4 contain almost only men.}
+%  \label{fig:hypothesisfree}
+% \end{figure}
+% %
+% Analyzing the data set without prior hypothesis requires a starting point giving an overview over the data first \cite{Shneiderman1996}.
+% %
+% Shape-based hypothesis-free exploration starts with the \emph{local investigation} pattern by a shape grouping step using shape-based clustering.
+% %
+% The results for this step can be seen in Figure~\ref{fig:hypothesisfree} (a).
+% 
+% Cluster 9 represents subjects with average shape.
+% %
+% Other shapes differ with respect to size, such as cluster 2, 3, 8, 10 where the last one also represents a more straight spine, which is usual for subjects with larger body size.
+% %
+% Cluster 4, 5, 7 contain outliers, characterized by their unusual shape and small number.
+% %
+% Cluster 8 has the second largest number of elements and was therefore of special interest.
+% %
+% Performing a \emph{multivariate analysis} by looking at \emph{Cram\'{e}r's V} contingency values of cluster~8 reveals associations of this group with employment status, gender, body size, age, thyroid nodules and blood fat value.
+% %
+% Another \emph{multivariate analysis} using a pivot table set gender and employment status in relation to clustering affiliation\add{.} 
+% %
+% \add{It s}hows that cluster~8 contains mostly women and also has a larger unemployment rate (see Figure~\ref{fig:hypothesisfree}), while the overall employment rate of women and men in the data set is almost exactly 50\%.
+% %
+% While all observed features seem to be plausible associations related to back pain, the values indicate that cluster~8 contains subjects with chronic back pain radiating to the legs.
+% %
+% Metabolic parameters, such as blood fat and blood sugar, are also possibly associated features.
+% %
+% The employment status is a feature relating to many different lifestyle factors such as income or nutrition as well as age and might act as confounder.
+% \\\\
+% While this approach does not assume any hypothesis, the 127 epidemiological features were selected by domain experts with a wide range on potential influential factors.
+% %
+% This filtering step based on expert experience rules out possible confounder and useless associations and it is therefore arguable whether this represents a disadvantage \cite{Wiley2008}.
+% %It is arguable whether this first filtering step, which is purely based on expert experience, represents a disadvantage, since it rules out many parameters possibly associated with the presented features, but the value of this knowledge would be small \cite{Wiley2008}.
+% %
+% % While this approach does not assume any hypothesis, the data are met with a selection bias when compiling the list of related features by the domain experts.
+% % %
+% % It is arguable if this first filtering step, which is purely based on expert experience, is really a disadvantage, since it rules out many parameters which may interact with the presented features but the value of this knowledge would be small \cite{Wiley2008}.
+% 
+% \subsubsection{Hypothesis-Based Analysis} \label{Hypothesis based Analysis}
+% \begin{figure}[htb]
+%  \centering
+%  %\includegraphics[width=1.5in]{figures/InteractionLoop}
+%  \includegraphics[width=3.5in]{figures/hypothesisbased}
+%  \caption{(a) Dichotomous questionnaire answer to ``\emph{Did you experience back pain in the past three months?}''.
+%  %
+%  Mean shapes between the groups show no difference.
+%  %
+%  (b) Shape-based clustering for all subjects who suffered from back pain yields 6 groups. 
+%  %
+%  Note that the difference in subject count is due to the missing shape information for some subjects.
+%  %
+% % The total count of subjects in (b) is smaller 
+%  }
+%  \label{fig:hypopthesisbased}
+% \end{figure}
+% %
+% If the user proposes a hypothesis about a relation between a non-image feature regarding shape, the workflow slightly differs from the hypothesis-free analysis.
+% %
+% The starting point follows the \emph{feature investigation} pattern, where a feature of interest is selected by dragging it into the canvas area and viewing the subject's distribution as well as their shape differences.
+% %
+% In our use case, epidemiologists were interested in the dichotomous questionnaire answer to ``\emph{Did you experience back pain in the past three months?}''.
+% %
+% The mean shapes of the resulting visualization show no difference between the two groups (see Figure~\ref{fig:hypopthesisbased} (a)).
+% %
+% Either there are no differences or the variance information was lost in the mean shape calculation.
+% %
+% Since the focus is on subjects suffering from back pain, the clustering results of these subjects are then drawn into the canvas area, yielding six clusters as seen in Figure~\ref{fig:hypopthesisbased} (b).
+% %
+% Cluster 5 stood out for having a so-called \emph{hyperlordosis}, a strong curvature of the lumbar spine associated with back pain.
+% 
+% \emph{Multivariate analysis} using \emph{Cram\'{e}r's V} contingency values highlighted relationships of this cluster with joint degeneration, meat eating habits, preoccupation, back pain, neck or shoulder pain and waist circumference.
+% %
+% Since the prior selection only yields subjects that report back pain, the pain indicators specify the pain localization for the subjects.
+% 
+% It is well known that overweight is a risk factor for back pain.
+% %
+% While the BMI is a key figure for assessing height and weight of a subject, it does not tell anything about how the weight is distributed in the body.
+% %
+% Our epidemiological collaborators were interested in the correlation with waist circumference presented by this group.
+% %
+% Our finding follows the recent trends indicating that BMI is not a good measure for assessing the body shape, since healthy weight is dependent on many other measures \cite{Ahima2013}.
+% %
+% It indicates that the waist circumference rather than the BMI interacts with unusually shaped spines for subjects with lumbar back pain.
+% %
+% The influence of the parameter is now in the focus of further analysis.
+% 
+% \subsubsection{Follow-Up Tasks and Concluding Domain-Expert Feedback}
+% %
+% Correlation does not automatically imply causation. %\cite{Tufte2003}.
+% %
+% In particular, when large amounts of attributes are automatically checked for strong correlations, just by chance hundreds of them reach at least a moderately significant correlation, such as $p>0.05$.
+% %Defining a causal relationship solely based on observed correlations of two features is \emph{cum hoc ergo propter hoc}--correlation does not automatically imply causation \cite{Tufte2003}.
+% %
+% The observed correlations need to be carefully checked for confounders and medical soundness.
+% %
+% Statisticians validate causal inferences of the drawn conclusions.
+% 
+% Features that potentially interact with a disease-related condition need to be validated.
+% %
+% To increase the probability of the observation not to be random and to avoid \emph{overfitting}, they are cross-checked for associations in \texttt{SHIP-TREND} as a second, independent population sample, which was examined with methods identical to \texttt{SHIP}.
+% \\\\
+% The prototype was demonstrated to the domain experts (with background on epidemiology and radiology) via screen sharing.
+% %
+% The software was operated by a computer scientist following instructions of the domain experts.
+% %
+% The presented methods guide the attention to features that are not in the focus of attention and expectance of clinical researchers.
+% %
+% The explorative nature of the methods work well for gathering associations possibly acting as confounder, as outcome of a disease or as an actual cause or risk factor.
+% %"\emph{Correlation does not imply causation}" is a famous saying in statistics, meaning that correlation does not necessarily imply causal relationships \cite{}.
+% %
+% This distinction is hard to make and requires a lot of clinical experience.
+% %
+% The combination of multiple views with shape information helps to connect many different information sources to make the large information spaces cognitively feasible.
+% %
+% Displaying MRI scans for the subjects in the outlier cluster is promising because they are highly likely to exhibit pathologies.
+% %
+% %We plan to include a \texttt{DICOM}-viewer to meet this wish.
+% 
+% % - Describe steps from gathering Information from the raw image files (segmentation, abstraction, visualization)\\
+% % - Problem of sparse differences - Visualization has to be more abstract to emphasize the differences.\\
+% % - Input of Epidemiologists goes here!\\
+% ---------------------------------------------------------------------------------------------------------------------------------------
+% ---------------------------------------------------------------------------------------------------------------------------------------
+% ---------------------------------------------------------------------------------------------------------------------------------------
 

### 2014-06-17 draft.tex
Worked on the Application section

@@
 %
-\com{Multivariate analysis using the adjacency matrix may have also be explained here, depending on the Application chapter.}
+\com{ToDo: Multivariate analysis using the adjacency matrix may have also be explained here, depending on the Application chapter.}
 %
@@
 
+% TODO: This can be an Section of it's own? Or part of the design?
 \subsection{Implementation}
@@
 %
-\com{Anmerkung von Steffen einbauen: Evaluierung geht um fachliches Feedback - Wissen über epidemiologie UND rücken und wissen um die Daten - deswegen schwer ganz viele Experten ranzuziehen}
-Guided by the work of Lam et al \cite{Lam2012}, we conducted an investigation of \emph{Visual Data Analysis and Reasoning} (VDAR).
+%\com{Anmerkung von Steffen einbauen: Evaluierung geht um fachliches Feedback - Wissen über epidemiologie UND rücken und wissen um die Daten - deswegen schwer ganz viele Experten ranzuziehen}
+Guided by the work of Lam et al \cite{Lam2012}, we conducted an investigation of \emph{Visual Data Analysis and Reasoning} (\emph{VDAR}).
 %
@@
 %
-Since it is hard to quantify when these goals are achieved, Lam et al. suggest case studies for the VDAR analysis.
-
-We conducted a qualitative case study with two epidemiological domain experts.
-%
-Both of them are practicing clinicians with focus on epidemiological research.
-%
-The first clinician is an specialist in internal medicine and also the leader of the SHIP.
-%
-One clinician is an radiologist and responsible for the image acquisition of the SHIP and has a deep understanding on the available image data.
+Since it is hard to quantify when these goals are achieved, Lam et al.
+suggest case studies for the \emph{VDAR} analysis by applying the think aloud protocol to understand the domain experts observations, inferences and conclusions when using the system.
+
+We conducted a qualitative case study with two epidemiological domain experts who also co-authored this publication.
+%
+Both of them are practicing physicians with focus on epidemiological research.
+%
+One is an specialist in internal medicine and also the head of the SHIP, the other an radiologist and responsible for the MRI-data acquisition of the entire SHIP.
+%and has a deep understanding on the available image data.
+%
+(Their expertise and experience with the data in their day to day research make them aware of the problems associated with the analysis of medical cohort study data.)
 %
@@
 Video-recordings of the sessions made it possible to evaluate the sessions later on.
-\com{TODO: Name der Screensharing Software mit angeben (Teamviewer)?}
+%
 \paragraph{Procedure.}
-At first, controlled mouse and keyboard of the clinicians pc and demonstrated the basic functionalities of the system.
+At first, we controlled mouse and keyboard of the participants pc and demonstrated the basic functionalities of the system.
 %
@@
 %They pointed out the high potential they see in this visualization, because it allows to look at the data in an unbiased way.
-They pointed out the benefits of the possibility to look at all variables in the context of each other.
-%
-To cite the radiologist, it "illumunates the black box of the data", making it possible to look at the data in an unbiased way.
+%They pointed out the benefits of the possibility to look at all variables in the context of each other.
+Before, they were not able to to look at all variables in the context of each other.
+%
+%
+To cite one expert, it "illumunates the data black box", making it possible to look at the data unbiased from assumptions.
+%
+The radiologist was looking for correlations with shape related features in the data, finding that they correlate with \emph{leg pain}, \emph{age}, \emph{body height} and \emph{hormone replacement therapy status}.
+%
+The internist inspected non-image-derived features, such as age-associated parameters like \emph{income}, \emph{blood fat values} or \emph{number of born children}.
+%
+This also shows, that even though the visualization may treat every feature equally, the domain experts is more likely interested in features they are familiar with and have personal clinical experience with.
 %
 Due to the dense mapping of information in the contingency matrix it was suggested to make this visualization full screen.
-%
-The radiologist was looking for correlations with shape related features in the data, finding that they correlate strong with each other as well as with leg pain, age, body height and hormone replacement therapy.
-%
-The internist inspected features which are not derived from the images, such as parameters strongly related with age, such as income, blood fat values or number of born children.
-%
-This also shows, that even though the visualization may treat every feature equally, the domain experts is more likely interested in features they are familiar with and have personal clinical experience with.
-%
 %Many correlations observed in this view
 
-After this initial overview, the expert performed a \emph{multivariate analysis} by introducing features, such as age, waist circumference, weight or lumbar spine centerline curvature as bar charts views into the canvas area and selected subgroups to see how they are distributed and if they could observe unusual behavior in the mean-shapes.
-%
-+ Cramer test 
-\com{Allgemein: Workflow noch einmal grob beschreiben - Overview first, details on demand}
+After this initial overview, the expert performed a \emph{multivariate analysis} by introducing features, such as \emph{age}, \emph{waist circumference}, \emph{weight} or \emph{lumbar spine canal curvature} as bar charts views into the canvas area and selected subgroups to see how they are distributed and if they could observe unusual behavior in the mean-shapes.
+%
+This pointed out problems with the used categorization method splitting numerical variables into equally sized bins.
+%
+If an feature contains outliers, such as waist circumference (e.g. subjects with morbid obesity) this approach leads to sparse categories, making it hard to calculate associations.
+%
+Solutions for this were discussed with the domain experts and are described in more detail in Section~\ref{Lessons Learned}.
+%
+%\com{Allgemein: Workflow noch einmal grob beschreiben - Overview first, details on demand}
 %
@@
 
-Cluster 9 represents subjects with average shape.
-%
-Other shapes differ with respect to size, such as cluster 2, 3, 8, 10 where the last one also represents a more straight spine, which is usual for subjects with larger body size.
-%
-Cluster 4, 5, 7 contain outliers, characterized by their unusual shape and small number.
-%
-Cluster 8 has the second largest number of elements and was therefore of special interest.
-%
-The user opened the contingency pane to performing a \emph{multivariate analysis} by looking at \emph{Cram\'{e}r's V} contingency values of cluster~8.
-%
-This revealed associations of this group with employment status, gender, body size, age, thyroid nodules and blood fat value.
-%
-Another \emph{multivariate analysis} using a pivot table set gender and employment status in relation to clustering affiliation\add{.} 
-%
-\add{It s}hows that cluster~8 contains mostly women and also has a larger unemployment rate (see Figure~\ref{fig:hypothesisfree}), while the overall employment rate of women and men in the data set is almost exactly 50\%.
-%
-While all observed features seem to be plausible associations related to back pain, the values indicate that cluster~8 contains subjects with chronic back pain radiating to the legs.
-%
-Metabolic parameters, such as blood fat and blood sugar, are also possibly associated features.
-%
-The employment status is a feature relating to many different lifestyle factors such as income or nutrition as well as age and might act as confounder.
-
+Cluster 4 represents subjects with average shape.
+%
+Other shapes differ with respect to size, such as cluster 2, 3, 7, where the last one and cluster 5 also represent a more straight spine, which is usual for subjects with larger body size.
+%
+Cluster 1 and 6 contain outliers, characterized by their unusual shape and small number.
+%
+%Cluster 8 has the second largest number of elements and was therefore of special interest.
+%
+%To perform an automatic feature suggestion using a \emph{multivariate analysis} by looking at \emph{Cram\'{e}r's V} contingency values of in the contingency pane.
+To get an overview over the suggested features, the user opened the contingency pane to perform a \emph{multivariate analysis} by looking at \emph{Cram\'{e}r's V} contingency values of all clusters, revealing a strong correlation with \emph{gender} and \emph{body size}.
+%
+Therefore another \emph{multivariate analysis} was carried out by the feature \emph{gender} to the canvas from the contingency pane and selecting all female subjects (Fig.~\ref{ToDo}).
+%
+Cluster 1 contained almost only female subjects.
+%
+Contingency values for this cluster revealed correlations with \emph{leg fatigue}, \emph{physical heavy work}, \emph{body weight}, \emph{dyspnoea} and \emph{headache intensity}.
+%
+%This revealed associations of this group with employment status, gender, body size, age, thyroid nodules and blood fat value.
+%
+Since it is an pain indicator, headache was of special interest and was further investigated by incorporating a pivot table setting \emph{headache intensity} in relation to cluster affiliation.
+%
+Cluster 1 subjects report a much higher chance for heavy headaches than other subjects.
+%Another \emph{multivariate analysis} using a pivot table set gender and employment status in relation to clustering affiliation\add{.} 
+%
+%\add{It s}hows that cluster~8 contains mostly women and also has a larger unemployment rate (see Figure~\ref{fig:hypothesisfree}), while the overall employment rate of women and men in the data set is almost exactly 50\%.
+%
+%While all observed features seem to be plausible associations related to back pain, the values indicate that cluster~8 contains subjects with chronic back pain radiating to the legs.
+%
+%Metabolic parameters, such as blood fat and blood sugar, are also possibly associated features.
+%
+%The employment status is a feature relating to many different lifestyle factors such as income or nutrition as well as age and might act as confounder.
+\\\\
+The experts emphasized the importance of methods providing overview over the data for hypothesis generation.
+%
+With the presented \emph{IVA} approaches they were quickly able to confirm medical knowledge as well as elaborate new hypothesis.
 % TODO: Contingency Matrix or Adjecency Matrix
-\com{00.44 Clustering-Ergebnis Weiblich und Arbeitspensum Alter}
+%\com{00.44 Clustering-Ergebnis Weiblich und Arbeitspensum Alter}
 
@@
 The starting point follows the \emph{feature investigation} pattern, where a feature of interest is selected by dragging it into the canvas area and viewing the subject's distribution as well as their shape differences.
-\paragraph{Analysis 1}
-The radiologist performed a \emph{feature investigation} by dragging the age variable into the canvas area, observing that age-bins 36-52 years and 52-68 years are nearly identical to the global mean, while 20-36 years and 68-84 years differ in size.
-%
-Performing a \emph{multivariate analysis} by selecting only female subjects via the additionally added gender feature it was observed, that distance was closest to subjects of 68-84 years old and largest to 20-36 years old.
-%
-The expert argued that this is most likely due to the shrinking of the spine with increasing age, making it more similar to the female average spine which is smaller than the global average \com{(00.29)}.
-%
-
-\paragraph{Analysis 2.}
-Hypothesis: "Back pain is associated with age and therefore there differences in the lumbar spine shape are observable." \com{TODO Image Mosaic Plot Age, Backpain}.
-To validate this hypothesis, a \emph{multivariate analysis} was performed by introducing the dichotomous feature "\emph{Did you experience backpain in the last three moth?}" together with the age as mosaic plot by dropping both features on the canvas area.
+% \paragraph{Analysis 1.}
+% The radiologist performed a \emph{feature investigation} by dragging the age variable into the canvas area, observing that age-bins 36-52 years and 52-68 years are nearly identical to the global mean, while 20-36 years and 68-84 years differ in size.
+% %
+% Performing a \emph{multivariate analysis} by selecting only female subjects via the additionally added gender feature it was observed, that distance was closest to subjects of 68-84 years old and largest to 20-36 years old.
+% %
+% The expert argued that this is most likely due to the shrinking of the spine with increasing age, making it more similar to the female average spine which is smaller than the global average \com{(00.29)}.
+% %
+% 
+\paragraph{Analysis 1.}
+Hypothesis: ``\emph{Back pain is associated with age and lumbar spine shape}''.
+%Hypothesis: "Back pain is associated with age and therefore there differences in the lumbar spine shape are observable." 
+\com{TODO Image Mosaic Plot Age, Backpain}.
+%
+To validate this hypothesis, a \emph{multivariate analysis} was performed by introducing the dichotomous feature ``\emph{Did you experience back pain in the last three month?}'' together with the age as mosaic plot by dropping both features on the canvas area.
 %
@@
 %
-For one, age influences the lumbar spine shape very heavily, as seen in Fig~\ref{?}, while the differences between subjects with and without back pain are small.
+For one, age influences the lumbar spine shape very heavily, as seen in Fig~\ref{ToDoA}, while the differences between subjects with and without back pain are small.
 %
@@
 %
-As seen in Fig~\ref{?}, subjects reporting back pain are even the majority, which makes it difficult to extract parameters which reliably describe back pain.
+As seen in Fig~\ref{ToDoA}, subjects reporting back pain are even the majority, which makes it difficult to extract parameters which reliably describe back pain.
 %
@@
 %
-A next step would be to use a variable describing the strength of back pain and only analyzing strong back pain!
-%
-
-\com{ToDo: Ability to better create sub groups would be useful as well as better segmentation}
+The next step is to use a variable describing the strength of back pain and only analyzing strong back pain!
+
+The epidemiologists pointed out that they would like to see a more intuitive and fast way to select subgroups based on different features to make full use of the analysis capabilities.
+%
+We discuss this in detail in Section~\ref{Lessons Learned}.
 %gender, use of pain medications. this would be good for future work, providing a better way to subdivide the data set and construct groups. The presented method is very good to gain an overview
@@
 
-\paragraph{Analysis 3}
-\com{ToDo: Needs to be rewritten}
-\begin{figure}[htb]
- \centering
- %\includegraphics[width=1.5in]{figures/InteractionLoop}
- \includegraphics[width=3.5in]{figures/hypothesisbased}
- \caption{(a) Dichotomous questionnaire answer to ``\emph{Did you experience back pain in the past three months?}''.
- %
- Mean shapes between the groups show no difference.
- %
- (b) Shape-based clustering for all subjects who suffered from back pain yields 6 groups. 
- %
- Note that the difference in subject count is due to the missing shape information for some subjects.
- %
-% The total count of subjects in (b) is smaller 
- }
- \label{fig:hypopthesisbased}
-\end{figure}
-%
-In our use case, epidemiologists were interested in the dichotomous questionnaire answer to ``\emph{Did you experience back pain in the past three months?}''.
-%
-The mean shapes of the resulting visualization show no difference between the two groups (see Figure~\ref{fig:hypopthesisbased} (a)).
-%
-Either there are no differences or the variance information was lost in the mean shape calculation.
-%
-Since the focus is on subjects suffering from back pain, the clustering results of these subjects are then drawn into the canvas area, yielding six clusters as seen in Figure~\ref{fig:hypopthesisbased} (b).
-%
-Cluster 5 stood out for having a so-called \emph{hyperlordosis}, a strong curvature of the lumbar spine associated with back pain.
-
-\emph{Multivariate analysis} using \emph{Cram\'{e}r's V} contingency values highlighted relationships of this cluster with joint degeneration, meat eating habits, preoccupation, back pain, neck or shoulder pain and waist circumference.
-%
-Since the prior selection only yields subjects that report back pain, the pain indicators specify the pain localization for the subjects.
-
-It is well known that overweight is a risk factor for back pain.
-%
-While the BMI is a key figure for assessing height and weight of a subject, it does not tell anything about how the weight is distributed in the body.
-%
-Our epidemiological collaborators were interested in the correlation with waist circumference presented by this group.
-%
-Our finding follows the recent trends indicating that BMI is not a good measure for assessing the body shape, since healthy weight is dependent on many other measures \cite{Ahima2013}.
-%
-It indicates that the waist circumference rather than the BMI interacts with unusually shaped spines for subjects with lumbar back pain.
-%
-The influence of the parameter is now in the focus of further analysis.
-
-\com{00.46 Second: Analysis of Female subjects via clustering, selction of cluster 2, analysis using Body weight and Age.
+\paragraph{Analysis 2.}
+Hypothesis: ``\emph{Back pain is related to lumbar spine deformation.}''.
+%
+The previously discussed analysis puts the suitability of the lumbar spine segmentation for analyzing back pain into question.
+%
+The results of the analysis lead to this analysis.
+%
+This hypothesis is analyzed by dropping the feature ``\emph{Did you experience back pain in the past three months?}'' into the canvas are, as seen in previous Fig.~\ref{ToDoA}.
+%
+Then, the features derived from the clustering results for both subject groups are dropped into the canvas area as well (Fig.~\ref{ToDoB} (a)).
+%
+While the clustering algorithm finds only three homogenous cluster close to the mean shape for subjects reporting no back pain.
+%
+The cluster analysis for back pain yields diverse clusters with various pathological shape classes.
+%
+Cluster~5 represents most of the subjects and is very similar to the global mean shape.
+%
+Cluster~1 and 2 present a \emph{hyperlordosis}, a strong curvature of the lumbar spine, while Cluster 3 and 4 present a more straight shape.
+%
+A \emph{multivariate analysis} using the pivot tables put gender and strong back pain in context to cluster affiliation (Fig.~\ref{ToDoB} (b)). 
+%
+It shows, that subjects in Cluster~1, 2 and 6 reporting strong back pain while at the same time they also have a considerably higher share of females.
+%
+To check for unusual correlations, the expert opened the \emph{Cram\'{e}r's V} contingency table.
+%
+It depicted strong association with \emph{body fat}, \emph{body weight} and \emph{blood pressure} (Cluster~1) \emph{alcohol consumption} and \emph{attentiveness disorder} (Cluster~2), and \emph{amount of sleep} (Cluster~6).
+%
+For the experts, these observations are a starting point for a number of new hypothesis about possible relationships, for example association between overweight and Cluster~1.
+%
+\\\\
+In summary it can be stated that hypothesis-driven analysis leads to hypothesis generation by design of the tool.
+%
+It is not designed to statistically validate hypothesis like statistical processors.
+%
+It rather triggers the analysis of potentially associated features with a pathology of interest.
+%Zusammenfassend kann gesagt werden, dass hypothesenanalyse und hypothesengenerierung durch das design des Tools hand in hand gehen, es istn icht dafür geeignet statistisch bis ins detail hypothesen zu validieren, sondern triggert eher die analyse potentiell assoziierter variablen, die später verfeinert werden.
+%
+%By introducing the \emph{gender} feature and selecting all male subjects it was shown, that Cluster 1, 2 and 6 pr
+%\emph{Multivariate analysis} using \emph{Cram\'{e}r's V} contingency values highlighted relationships of this cluster with joint degeneration, meat eating habits, preoccupation, back pain, neck or shoulder pain and waist circumference.
+
+% \paragraph{Analysis 3}
+% \com{ToDo: Needs to be rewritten}
+% \begin{figure}[htb]
+%  \centering
+%  %\includegraphics[width=1.5in]{figures/InteractionLoop}
+%  \includegraphics[width=3.5in]{figures/hypothesisbased}
+%  \caption{(a) Dichotomous questionnaire answer to ``\emph{Did you experience back pain in the past three months?}''.
+%  %
+%  Mean shapes between the groups show no difference.
+%  %
+%  (b) Shape-based clustering for all subjects who suffered from back pain yields 6 groups. 
+%  %
+%  Note that the difference in subject count is due to the missing shape information for some subjects.
+%  %
+% % The total count of subjects in (b) is smaller 
+%  }
+%  \label{fig:hypopthesisbased}
+% \end{figure}
+% %
+% In our use case, epidemiologists were interested in the dichotomous questionnaire answer to ``\emph{Did you experience back pain in the past three months?}''.
+% %
+% The mean shapes of the resulting visualization show no difference between the two groups (see Figure~\ref{fig:hypopthesisbased} (a)).
+% %
+% Either there are no differences or the variance information was lost in the mean shape calculation.
+% %
+% Since the focus is on subjects suffering from back pain, the clustering results of these subjects are then drawn into the canvas area, yielding six clusters as seen in Figure~\ref{fig:hypopthesisbased} (b).
+% %
+% Cluster 5 stood out for having a so-called \emph{hyperlordosis}, a strong curvature of the lumbar spine associated with back pain.
+% 
+% \emph{Multivariate analysis} using \emph{Cram\'{e}r's V} contingency values highlighted relationships of this cluster with joint degeneration, meat eating habits, preoccupation, back pain, neck or shoulder pain and waist circumference.
+% %
+% Since the prior selection only yields subjects that report back pain, the pain indicators specify the pain localization for the subjects.
+% 
+% It is well known that overweight is a risk factor for back pain.
+% %
+% While the BMI is a key figure for assessing height and weight of a subject, it does not tell anything about how the weight is distributed in the body.
+% %
+% Our epidemiological collaborators were interested in the correlation with waist circumference presented by this group.
+% %
+% Our finding follows the recent trends indicating that BMI is not a good measure for assessing the body shape, since healthy weight is dependent on many other measures \cite{Ahima2013}.
+% %
+% It indicates that the waist circumference rather than the BMI interacts with unusually shaped spines for subjects with lumbar back pain.
+% %
+% The influence of the parameter is now in the focus of further analysis.
+
+%\com{00.46 Second: Analysis of Female subjects via clustering, selection of cluster 2, analysis using Body weight and Age.
 Clustering Result if 1 gives large distance between other subjects.
 00.49 How are can the differences be explained?}
-\com{00.46 Feedback Body-Size/Body weight.
+%\com{00.46 Feedback Body-Size/Body weight.
 Some features are sparse, suggestion was made to use quantile/quintile, resulting in equally occupied results.
@@
 % ---------------------------------------------------------------------------------------------------------
-\subsection{Further Feedback and Lessons Learned}
-\com{ToDo: Noch einmal auf die VDAR Results eingehen}
+\subsection{Further Feedback and Lessons Learned} \label{Lessons Learned}
 %\com{ToDo: Lessons Learned der einzelnen Cases - Fokus bei Hypothesenfreier Analyse muss etwa eine gute Übersichtsvisualisierung sein, da es genau das ist was den Epidemiologen hierzu im Moment fehlt}
@@
 The domain experts see another great potential in communicating insights efficiently using the presented visualizations \com{(1.08)}.
-%
-The experts proposed to improve the visual comparison between image data representations by abstracting them into representative land marks, such as centers of the vertebrae and cardinal points of the lumbar spine canal.
 %
@@
 Creating adjacency matrices for sub groups, such as different age bins can help to characterize the aging process by deriving age-specific risk factors.
+
+\emph{Multivariate analysis} can also be improved by more ways of brushing the data as well as creating sub groups for comparison as a result of the hypothesis-driven analysis case.
+%
+To small feature ranges resulting in very sparse groups could potentially hinder the ability to calculate statistical resilient measures, since they require a minimum amount of subjects exhibiting the selected feature ranges.
+%This also has to respect the number of subjects available for epidemiological studies
 %
@@
 %While the existing brushing and linking techniques already allow the comparison of image data 
-\com{(TODO weitere Unterteilung der Variablen und anschließende Anbindung an Statistik Tools)}
-
-\paragraph{Segmenation quality is crucial.}
+%\com{(TODO weitere Unterteilung der Variablen und anschließende Anbindung an Statistik Tools)}
+
+\paragraph{Segmentation quality is crucial.}
 The radiologist pointed out the unusual strong similarity of the L3 vertebrae throughout the population.
@@
 Since deformation of the spine canal is commonly the last pathological stage of pathological lumbar spine deformation and is preceded by vertebrae deformation, the system would strongly benefit from more precise segmentation results capturing these prior changes.
-
-\paragraph{Usage of different Categorizations depending on expected outcome.}
+%
+For the visual comparison itself, the radiologist proposed an abstraction the representation them into representative land marks, such as centers of the vertebrae and cardinal points of the lumbar spine canal.
+
+\paragraph{Usage of different categorizations depending on expected outcome.}
 When categorizing numerical variables into equal groups it is possible to create very sparse categories due to outliers, for example when analyzing body weight.
@@
 The large data base associated with image based epidemiological data remains on the server machine and has not to be moved tediously using external hard disks.
+%
+This approach is well suited for the \emph{VDAR} approach to assess user thought processes using the think aloud technique.
+%The \emph{VDAR} approach to assess user thought processes using the think aloud technique worked well with the web-based approach for both implementing the prototype as well as getting feedback via voice over ip conference calls worked very well.
+\\\\
+Since it is an entirely different approach to their usual workflows which are heavily influenced by statistical tools, it is to highlight that they had fun browsing through the data set in the presented manner and to dig out new information about the data.
+%
+This underlines the accessibility of \emph{IVA} tools, which amplifies the natural curiosity of the domain experts in complex medical feature associations.
 %\com{ToDo Fazit über Web-Austausch, das Experten begeistert davon waren und dass es hervorragend funktioniert.
@@
 %
+\com{This needs to be adjusted to fit the new Application section}
 Findings from analyzing lumbar back pain using the \emph{IVA} approach range from deriving shape-based groups of subjects to detailed description of features potentially associated with the disease, such as waist circumference, thyroid nodules and blood fat values.

### 2014-06-18 draft.tex
Added new figure for application part and finished it

@@
 \section{Application}
-This section describes how our presented \emph{IVA} workflow is used for an epidemiological use case.
-%
-We conducted a case study with two domain experts on a data set compiled to analyze lower back pain. 
+This section describes how the presented \emph{IVA} workflow is used in epidemiological application.
+%
+We conducted a qualitative case study with two domain experts on a data set compiled to analyze lower back pain. 
 %
@@
 
+%-------------------------------- Figure --------------------------------
+\begin{figure*}[htb]
+ \centering
+ \includegraphics[width=1\textwidth, resolution=300]{figures/evaluation}
+ \caption{Various case study results.
+ %
+ (a) Mean curvature of lumbar spine canal plotted against the mean shape of 58-74 years old female subjects (light-blue bars).
+ %
+ Note the high amount of this subject group in the third group. The last group only contains four outlier-subjects.
+ %
+ (b) Clustering of all subjects yields seven groups, where Cluster~4 assembles the mean. The light blue bars indicate the share of females in the group.
+ %
+ (c) A mosaic plot mapping age against dichotomous questionnaire answer to ``\emph{Did you experience back pain in the past three months?}''.
+ (d) Clustering result of ``\emph{Did you experience back pain in the past three months?}'' Yes/no with female share in each group. 
+ %
+ Cluster 1 and 6 for answer ``\emph{Yes}'' contain mostly women.
+ %
+ The pivot table shows how many subjects with strong back pain are in each cluster for answer ``\emph{Yes}''.
+ %
+ Subjects in Cluster 1, 2 and 6 report strong back pain more often than subjects in other clusters.
+%Female subjects are selected as reference for all plots, except for (c) The light blue selection in the bar charts represent the female share. }
+ }
+ \label{fig:application}
+\end{figure*}
+%-------------------------------- /Figure -------------------------------
 \subsubsection{Shape Visualization and Clustering}
@@
 The method has proven to produce sound results on a preliminary data set and was able to reproduce textbook knowledge \cite{Klemm2013VMV}.
-\subsection{Domain Experts, Setup and Procedure}
+\subsection{Participants, Setup and Procedure}
 %
@@
 %
-Since it is hard to quantify when these goals are achieved, Lam et al.
+Since it is hard to quantify when these outcomes, Lam et al.
 suggest case studies for the \emph{VDAR} analysis by applying the think aloud protocol to understand the domain experts observations, inferences and conclusions when using the system.
@@
 %
-One is an specialist in internal medicine and also the head of the SHIP, the other an radiologist and responsible for the MRI-data acquisition of the entire SHIP.
-%and has a deep understanding on the available image data.
-%
-(Their expertise and experience with the data in their day to day research make them aware of the problems associated with the analysis of medical cohort study data.)
-%
-\paragraph{Setup.} To bypass the large geographical distance between us and the domain experts and to take their busy schedule into account, the case study was done completely web based.
-%
-Since one major reason for the focus on web technologies, no software needed to be exchanged.
-%
-A website link was send to the users, which they could open with their browser of choice and start using the tool.
+One is an specialist in internal medicine and also the head of the SHIP, the second an radiologist and responsible for the MRI-data acquisition of the study.
+\com{Years of experience} %and has a deep understanding on the available image data.
+%
+%(Their expertise and experience with the data in their day to day research make them aware of the problems associated with the analysis of medical cohort study data.)
+%
+\paragraph{Setup.} To bypass the large geographical distance between us and the domain experts and to take their busy schedule as physicians into account, the case study was done completely web-based.
+%
+%Our focus on web technologies right from the beginning made it easy to 
+%Since one major reason for the focus on web technologies, no software needed to be exchanged.
+The experts accessed the prototype by simply entering the website link into their browser of choice, no exchange of software was required.
+%Since our prototype ran on a local server, simply needed the homepage link to access the prototype.
+%
+%A website link was send to the users, which they could open with their browser of choice and start using the tool.
 %
@@
 %
-Video-recordings of the sessions made it possible to evaluate the sessions later on.
+Video-recordings of the sessions allowed a detailed evaluation afterwards.%made a possible to evaluate the sessions later on.
 %
@@
 %
-This included the Heatmap, the correlation view, how to introduce parameter visualization, how the color coding of differences works and the pivot tables.
+%This included the contingency matrix, the correlation view, how to introduce feature visualizations, how the color coding of differences works and the pivot tables.
 %
@@
 The epidemiologists were given two tasks: one hypothesis-free analysis of the data and one starting with an assumption.
+%
+For each case we conducted one analysis with each expert.
 
 % \subsection{Exploratory Analysis of the Lumbar Spine Data Set}
-\subsection{Case 1: Hypothesis-free analysis}
+\subsection{Case 1: Hypothesis-free Analysis} \label{Hypothesis-free analysis}
 Analyzing the data set without prior hypothesis requires a starting point giving an overview over the data first \cite{Shneiderman1996}.
@@
 Performing a \emph{local investigation} by clustering all subjects or use the contingency matrix for a \emph{multivariate analysis}.
-
-\paragraph{Analysis 1.}
 %
 The latter was chosen first by both domain experts.
+% The latter was chosen first by the domain expert with radiological background.
 %
@@
 %
-To cite one expert, it "illumunates the data black box", making it possible to look at the data unbiased from assumptions.
+To cite one expert, the contingency matrix ``illumunates the data black box'', making it possible to look at the data unbiased from assumptions.
+
+\paragraph{Analysis 1.}
 %
 The radiologist was looking for correlations with shape related features in the data, finding that they correlate with \emph{leg pain}, \emph{age}, \emph{body height} and \emph{hormone replacement therapy status}.
-%
-The internist inspected non-image-derived features, such as age-associated parameters like \emph{income}, \emph{blood fat values} or \emph{number of born children}.
-%
-This also shows, that even though the visualization may treat every feature equally, the domain experts is more likely interested in features they are familiar with and have personal clinical experience with.
 %
@@
 %
-This pointed out problems with the used categorization method splitting numerical variables into equally sized bins.
-%
-If an feature contains outliers, such as waist circumference (e.g. subjects with morbid obesity) this approach leads to sparse categories, making it hard to calculate associations.
-%
-Solutions for this were discussed with the domain experts and are described in more detail in Section~\ref{Lessons Learned}.
+This pointed out problems with the used categorization method splitting numerical variables into equally sized ordinal bins.
+%
+If an feature contains outliers, such as \emph{waist circumference} (e.g. by subjects with morbid obesity) this approach leads to sparse categories, making it hard to calculate associations.
+%
+The proposed expert solution for this is categorization using quantiles/quartiles and is described in detail in Section~\ref{Lessons Learned}.
+
+A \emph{multivariate analysis} using the \emph{Cram\'{e}r's V} contingency values for subjects with strong lumbar spine curvature showed, that these subjects are prominetly females between 58-74 years who also report pain radiating from their back into other body regions Fig.~\ref{fig:application}(a).
 %
@@
 \paragraph{Analysis 2.}
-\begin{figure}[htb]
- \centering
- %\includegraphics[width=1.5in]{figures/InteractionLoop}
- \includegraphics[width=3.5in]{figures/hypothesisfree}
- \caption{(a) Clustering result of all subjects.
+% \begin{figure}[htb]
+%  \centering
+%  %\includegraphics[width=1.5in]{figures/InteractionLoop}
+%  \includegraphics[width=3.5in]{figures/hypothesisfree}
+%  \caption{(a) Clustering result of all subjects.
 The bar chart height indicates the number of subjects in the cluster.
@@
 Cluster 8 is striking because of the high proportion of women, while cluster 2, 3, 4 contain almost only men.}
- \label{fig:hypothesisfree}
-\end{figure}
-\com{ToDo: Rewrite}
-%
-The internist started also with a \emph{multivariate analysis} using the contingency table, but found no associations he was interested in.
-%
-Therefore, he applied the \emph{local investigation} pattern by a shape grouping step using shape-based clustering via dragging the "All subjects" feature into the canvas area, triggering the shape clustering.
+%  \label{fig:hypothesisfree}
+% \end{figure}
+%
+The second expert started also with a \emph{multivariate analysis} using the contingency matrix to analyze non-image-derived features, such as age-associated parameters like \emph{income}, \emph{blood fat values} or \emph{number of born children}, but found no associations of interest.
+%
+Therefore, he applied the \emph{local investigation} pattern by a shape grouping step using shape-based clustering via dragging the \emph{All subjects} from the sidebar into the canvas area, triggering the shape clustering.
 %Shape-based hypothesis-free exploration starts with the \emph{local investigation} pattern by a shape grouping step using shape-based clustering.
 %
-The results for this step can be seen in Figure~\ref{fig:hypothesisfree} (a).
+The results for this step can be seen in Figure~\ref{fig:application} (b).
 
@@
 %
-Therefore another \emph{multivariate analysis} was carried out by the feature \emph{gender} to the canvas from the contingency pane and selecting all female subjects (Fig.~\ref{ToDo}).
+Therefore another \emph{multivariate analysis} was carried out by the feature \emph{gender} to the canvas from the contingency pane and selecting all female subjects (Figure~\ref{fig:application} (b)).
 %
@@
 %
-Cluster 1 subjects report a much higher chance for heavy headaches than other subjects.
+It was found, that cluster~1 subjects report heavy headaches more frequently than other subjects.
 %Another \emph{multivariate analysis} using a pivot table set gender and employment status in relation to clustering affiliation\add{.} 
@@
 With the presented \emph{IVA} approaches they were quickly able to confirm medical knowledge as well as elaborate new hypothesis.
+%
+We observed, that even though the visualization may treat every feature equally, the domain experts is more likely interested in features they are familiar with and have personal clinical experience with.
 % TODO: Contingency Matrix or Adjecency Matrix
@@
 
-\subsection{Case 2: Hypothesis-based analysis}
+\subsection{Case 2: Hypothesis-based Analysis}
 If the user proposes a hypothesis about a relation between a non-image feature regarding shape, the workflow slightly differs from the hypothesis-free analysis.
@@
 %Hypothesis: "Back pain is associated with age and therefore there differences in the lumbar spine shape are observable." 
-\com{TODO Image Mosaic Plot Age, Backpain}.
-%
-To validate this hypothesis, a \emph{multivariate analysis} was performed by introducing the dichotomous feature ``\emph{Did you experience back pain in the last three month?}'' together with the age as mosaic plot by dropping both features on the canvas area.
-%
-The experts were not able to observe the expected effect in the visualization.
+%
+To validate this hypothesis, a \emph{feature investigation} was performed by introducing the dichotomous feature ``\emph{Did you experience back pain in the last three month?}'' together with the age as mosaic plot by dropping both features on the canvas area.
+%
+The expert was not able to observe the expected effect in the visualization.
 %
@@
 %
-For one, age influences the lumbar spine shape very heavily, as seen in Fig~\ref{ToDoA}, while the differences between subjects with and without back pain are small.
+For one, age influences the lumbar spine shape, as seen in Fig~\ref{fig:application}~(c), while the differences between subjects with and without back pain are small.
 %
 The major difference seen in the visualization are therefore presented by the age feature, masking differences via the back pain parameter.
-
+%
 The second explanation is the commonality of back pain in our society.
 %
-As seen in Fig~\ref{ToDoA}, subjects reporting back pain are even the majority, which makes it difficult to extract parameters which reliably describe back pain.
-%
-A \emph{local investigation} using the contingency table showed a strong association between back pain with gender and body height.
-%
-Body height was explained to be a confounder for gender, since female subjects smaller on average than male subjects.
-%
-The analysis solely based on shape only accentuated gender differences, not in back pain.
+As seen in Fig~\ref{fig:application}~(c), subjects reporting back pain are even the majority, which makes it difficult to extract parameters which reliably describe back pain.
+%
+A \emph{multivariate analysis} using the contingency table showed a strong association between \emph{back pain} with \emph{gender} and \emph{body height}.
+%
+\emph{Body height} was explained to be a confounder for \emph{gender}, since female subjects smaller on average than male subjects.
+%
+The analysis solely based on shape accentuated differences in \emph{gender}, not \emph{back pain}.
 %This makes it hard to analyze shape related features for this disease
@@
 
-The epidemiologists pointed out that they would like to see a more intuitive and fast way to select subgroups based on different features to make full use of the analysis capabilities.
-%
-We discuss this in detail in Section~\ref{Lessons Learned}.
+The epidemiologists pointed out that they would like to see a more intuitive and fast way to select subgroups based on different features to make full use of the analysis capabilities, as discussed in Section~\ref{Lessons Learned}.
 %gender, use of pain medications. this would be good for future work, providing a better way to subdivide the data set and construct groups. The presented method is very good to gain an overview
@@
 %
-The previously discussed analysis puts the suitability of the lumbar spine segmentation for analyzing back pain into question.
-%
-The results of the analysis lead to this analysis.
-%
-This hypothesis is analyzed by dropping the feature ``\emph{Did you experience back pain in the past three months?}'' into the canvas are, as seen in previous Fig.~\ref{ToDoA}.
-%
-Then, the features derived from the clustering results for both subject groups are dropped into the canvas area as well (Fig.~\ref{ToDoB} (a)).
+The previously discussed analysis puts the suitability of the lumbar spine segmentation for analyzing back pain into question, leading to this analysis.
+%
+Therefore the feature ``\emph{Did you experience back pain in the past three months?}'' is dropped into the canvas area.%, as seen in previous Fig.~\ref{ToDoA}.
+%
+Then, the features derived from the clustering results for both subject groups are dropped into the canvas area as well (Fig.~\ref{fig:application} (d)).
 %
@@
 %
-A \emph{multivariate analysis} using the pivot tables put gender and strong back pain in context to cluster affiliation (Fig.~\ref{ToDoB} (b)). 
-%
-It shows, that subjects in Cluster~1, 2 and 6 reporting strong back pain while at the same time they also have a considerably higher share of females.
-%
-To check for unusual correlations, the expert opened the \emph{Cram\'{e}r's V} contingency table.
+A \emph{multivariate analysis} using the pivot tables put gender and strong back pain in context to cluster affiliation (Fig.~\ref{fig:application} (d)). 
+%
+It shows, that subjects in Cluster~1, 2 and 6 reported strong back pain, while at the same time they also have a considerably higher share of females.
+%
+To check for unusual correlations, the expert used the \emph{Cram\'{e}r's V} contingency table.
 %
@@
 %
-It is not designed to statistically validate hypothesis like statistical processors.
+It is not suited and intended to statistically validate hypothesis like statistical processors.
 %
@@
 %
-The domain experts see another great potential in communicating insights efficiently using the presented visualizations \com{(1.08)}.
+Great potential is also seen in communicating insights efficiently using the presented visualizations \com{(1.08)}.
 %
@@
 %
-Since deformation of the spine canal is commonly the last pathological stage of pathological lumbar spine deformation and is preceded by vertebrae deformation, the system would strongly benefit from more precise segmentation results capturing these prior changes.
+Since deformation of the spine canal is the last stage of pathological lumbar spine deformation and is preceded by vertebrae deformation, the system would strongly benefit from more precise segmentation results capturing these prior changes.
 %
@@
 %
-These outliers are of high interest for finding pathological subjects, but not wanted when new parameters describing risk factors are established.
+These outliers only of high interest for finding pathological subjects. %but not wanted when new parameters describing risk factors are established.
 %

### 2014-06-19 draft.tex
Reworked Design chapter

@@
 Morphometric features are derived to allow for statistical comparison of the tissue, which incorporates mostly positions, diameters, volumes and relative distances and alignment to other structures.
-% 
-% - Data Format\\
-% - Missing Data\\
-% - Registration of Image Data (VMV'12)\\
-% - Extraction of Image Parameter
-\subsection{Analysis Workflow}
+%-------------------------------- Figure --------------------------------
 \begin{figure}[htb]
@@
 \end{figure}
+%------------------------------ / Figure --------------------------------
+\subsection{Analysis Workflow}
 %-------------------------------- Figure --------------------------------
@@
  \includegraphics[width=1\textwidth, resolution=300]{figures/visualization}
- \caption{(Left) Screenshot from the front-end, which is divided as follows: (a) The sidebar containing all features as well as the groups defined in the analysis process; (b) the canvas area where features can be added via drag and drop and the visualization can be chosen automatically according to the data type; (c) the interactive pivot table showing the exact numbers for each displayed feature combination.
+ \caption{\com{Add: Explanation of the color coding}(Left) Screenshot from the front-end, which is divided as follows: (a) The sidebar containing all features as well as the groups defined in the analysis process; (b) the canvas area where features can be added via drag and drop and the visualization can be chosen automatically according to the data type; (c) the interactive pivot table showing the exact numbers for each displayed feature combination.
 The data displayed is used to analyze the lumbar spine.
@@
 
-\subsection{Interaction and Visualization Techniques} \label{Interaction- and Visualization Techniques}
+% \subsection{Interaction and Visualization Techniques} \label{Interaction- and Visualization Techniques}
+\section{System Design and Visualization Techniques} \label{Interaction- and Visualization Techniques}
 %
@@
 
-\section{System Structure and Design} \label{Structure and Workflow}
-\com{Primary Reviewer Feedback: There is little detail about the design, and how it was decided to create this particular design; what interaction was made with the users, how did they influence this design?}
+% \subsection{System Structure and Design} \label{Structure and Workflow}
+\subsection{ToDo Rename.System Structure and Design} \label{Structure and Workflow}
+% \begin{itemize}
+% 	\item [?] Was sind die requirements des tools?
+% 	\item [?] how it was decided to create this particular design
+% 	\item which kind of User interaction was made?
+% 	\item Spezielle Anforderungen durch Webseitenformat, weggehen von allgemeinen UI Elementen wie Menubar, keine Rechtsklickmenüs.
+% 	\item Entschieden eine saubere UI zu bauen, Unterscheide durch Whitespace akzentuieren. Interface in Fig. 3 zu sehen.
+% 	\item Wie wird interagiert, wie wird gebrusht, was wird gelinkt?
+% 	\item wie wird das Clustering initialisiert?
+% 	\item auf Kontingenzmatrix mit verweisen als Teil des Interfaces (dieses auch noch anpassen)
+% \end{itemize}
+Early it became that we have to rely a lot on online communication with our clinical partners due to their busy schedule and our large spatial distance towards each other.
+%
+Hence we build our system using modern web technologies, described in detail in Section~\ref{implementation}.
+%
+By running the prototypes on server machines, software exchange became as easy as sharing a website link, giving us the opportunity to include the clinical experts easily in the development process.
+%
+Incorporating the \emph{IVA} workflow for image-centric cohort study data requires \emph{overview visualizations} as well as \emph{multivariate visualizations}, which bring image-derived information in context to non-image features.
+
+The focus on web technologies is not without tradeoffs.
+%
+Classical UI elements, such as the menu bar or custom right click menus are technically possible, but not common in this domain.
+%
+In favor of a clean layout, we designed the system without such components.
+%
+Since the previously described \emph{IVA} workflow allows for many different ways to analyze the data, we tried to make the interface as minimalistic and discrete as possible, treating the resulting space as \emph{canvas} for the data.
+%
 % \subsubsection{System Structure}
@@
 \begin{itemize}
+	\item The \emph{canvas} holding all visualizations. Elements can be added, arranged, resized and removed freely.
+	%
 	\item The \emph{sidebar}, which contains all epidemiological features. Cluster results are treated like features and are part of the sidebar as well.
-	\item The \emph{canvas} holding all visualizations. Elements can be added, arranged, resized and removed freely.
+	%
 	\item The interactive \emph{pivot table} gives detailed numerical information of the features in the canvas view. This view on the data comes natural to epidemiologists.
+	%
 	\item The \emph{contingency view} depicts relations for features in the canvas in an adjacency matrix.
 \end{itemize}
-%
-% The sidebar holds all available variables as well as groups either derived by user input or automatic clustering.
-% 
-% Features can be dragged from the sidebar into the canvas area.
-% %
-% Doing so will automatically create an information visualization suitable for the current data type.
-% %
-% 3D shape information about the investigated structure is displayed for each variable manifestation.
-% %
-% By dropping variables on existing visualizations the system creates a visualization that allows for comparison (e.g. mosaic plots for ordinal variables).
-% %
-% Elements can be brushed in the information visualizations and are linked to the other representations in the canvas.
-% %
-% Shape-based clustering can either be applied to all subjects or subgroups.
-% 
-% All elements in the canvas view are also represented in the interactive Pivot Table which gives detailed information how the subjects are distributed given the displayed variables.
-% %
-% Details to the different views are presented in the following sub-sections.
+
+\paragraph{System Layout}
+We experimented with several layouts.
+%
+The initial idea was to make all components freely arrange- and resizable on a large \emph{canvas} area.
+%
+This idea was soon dropped since domain experts reported a cluttered workspace, which required a lot of scrolling.
+%
+A well working configuration was the introduction of separate panes for the adjacency matrix, pivot table and sidebar, displayed with a mouse click on the corresponding button and sliding on top of the \emph{canvas} (Figure~\ref{fig:visualization} shows the system with reeled-out pivot table pane).
+%
+All user-generated visualizations follow are part of the \emph{canvas} and can be arranged freely.
+% are the  was to keep the concept of free arrangement for user-added 
+%The resulting system can be seen in Figure~\ref{fig:visualization} (left).
 
 \paragraph{Sidebar.}
+The \emph{sidebar} is the only visible element when the system is opened.
+%
 An overview of all features is presented in a sidebar where they are categorized into different types, such as somatometric (measurements of the dimensions of the human body), disease- or lifestyle-related, pain indicators and laboratory data.
 %
-It also contains subject groups either defined by brushing or automated shape clustering.
+It also contains subject groups defined by automated shape clustering.
 %
@@
 %
-% TODO Implement MOSAIC PLOTS, Scatterplots, Parallel Coordinates
+%
 Following Tufte's concept of \emph{small multiples} \cite{Tufte1983}, \add{information derived from the} medical image data are directly incorporated into the plot by including color-coded mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
@@
 %
-Mapping the distance from a group mean shape to the global mean using color allows to assess local shape changes.
-%%TODO Adaptive visualization!
+Mapping the distance from a group mean shape to the global mean using color allows to assess local shape changes (Figure~\ref{fig:visualization} (e)).
+%
+This is an important information to the epidemiologist, because until now they were not able rapidly validate shape differences based on non-image features.
+%
 If a feature is dropped on an existing plot, the visualization changes dynamically to properly make them comparable (Figure~\ref{fig:visualization} (right)).
 %
-Each plot can be brushed using widgets.
-%
-Brush selections are propagated to all visualizations allowing for fast feature querying.
-% TODO Timeline
-%
-%It is possible to duplicate brushes in order to create new groups which are evenly spread out.
-%
-%A use case for this is when a continuous feature has to be divided into even groups.
+%Elements can be brushed using 
+%Each plot can be brushed using widgets.
+%
+%Brush selections are propagated to all visualizations allowing for fast feature querying.
 
@@
 The distinction whether an association is a confounder or an effect, depends on the context defined by the hypothesis and is a decision to be made by the domain expert.
-% \subsubsection{Automated Feature Suggestion}
-% % https://de.wikipedia.org/wiki/Kontingenzkoeffizient#Cram.C3.A9rs_V
-% 
-% As discussed previously, highlighting potential interesting values in the data set is one major benefit of the \emph{IVA} powered approach.
-% %
-% Turkay and colleagues used the approach to calculate various key figures based in the distribution functions of each feature derived from the image data \cite{Turkay2013}.
-% %
-% Since the majority of our data are categorical features, we have to employ different solutions.
-% %
-% A solution for this problem are odds ratios, which are a standard statistical tool for stating relations between features.
-% %
-% Odds ratios can only be calculated for $2\times2$ contingency tables which usually represent the presence of a condition in a population divided by a characteristic (e.g.
-male/female subjects with presence or absence of back pain).
-% %
-% To calculate odds ratios for variables with more than two manifestations, we calculate local odds ratios for each possible combination, yielding a matrix for each feature combination \cite{Rudas1998}.
-% %
-% When the subjects are divided into groups and the calculation is carried out again for all feature combination, the difference in sum of the odds weighted with the number of manifestations per feature can be used to indicate if the feature combination yields a difference.
-% %
-% These differences are then highlighted in a separate tab of the side bar "Interactions".
-% %
-% Interestinassociationons then can be assessed creating a linked view using the standard drag and drop workflow.
-% \\\\
-% In the following sections we will discuss details of the implementation which relies on modern Web-Technologies.
-% \\\\
-% ToDo\\
-% - This can be improved--summing up the values possibly not the cleverest solution--calculation of variance etc. possible\\
-% - Matrix Visualization?
+%
+The contingency matrix visualization is an appropriate \emph{overview} visualization, something the epidemiological community lacks and is in great need of.
+%\com{Contingency Matrix excellent for Overview}
+\paragraph{Initialization and Interaction}
+Using feature suggestion allows to initialize the system with a set of potentially interesting. visualizations.
+%
+After testing and domain expert feedback we dropped this idea.
+%
+Reasons for this are twofold.
+%
+Very often are high correlations obvious, such as gender with menstrual status.
+%
+Also, we observed that the variables of interest is dependent on the specialization of the domain expert (explained in detail in Section~\ref{application}).
+
+Subject clustering is triggered automatically for each variable after it was added to the canvas by the user.
+%
+A status indicator at the bottom of the screen keeps the user informed about the pending clustering result, since the process can take up to ten seconds.
+%Since a clustering process can take up to ten seconds, a status indicator at the bottom of the screen keeps the user informed.
+%
+%\com{Clustering also automatically triggered}
 
 % TODO: This can be an Section of it's own? Or part of the design?
-\subsection{Implementation}
+\subsection{Implementation} \label{implementation}
 \begin{figure}[htb]
@@
 \\\\
-\section{Application}
-This section describes how the presented \emph{IVA} workflow is used in epidemiological application.
+\section{Application} \label{application}
+This section describes how the presented \emph{IVA} workflow is used in the epidemiological application.
 %
@@
 %
-To our knowledge, this is the first approach, where shape-related information of the whole lumbar spine is analyzed along with other epidemiological features.
+%To our knowledge, this is the first approach, where shape-related information of the whole lumbar spine is analyzed along with other epidemiological features.
 %
@@
 %
-Since it is hard to quantify when these outcomes, Lam et al.
+Since it is hard to quantify these outcomes, Lam et al.
 suggest case studies for the \emph{VDAR} analysis by applying the think aloud protocol to understand the domain experts observations, inferences and conclusions when using the system.
 
-We conducted a qualitative case study with two epidemiological domain experts who also co-authored this publication.
+Our participants are two epidemiological domain experts who also co-authored this publication.
 %
@@
 %
-One is an specialist in internal medicine and also the head of the SHIP, the second an radiologist and responsible for the MRI-data acquisition of the study.
+One is a specialist in internal medicine and also the head of the SHIP, the second a radiologist and responsible for the MRI data acquisition of the study.
 \com{Years of experience} %and has a deep understanding on the available image data.
@@
 %
-\paragraph{Setup.} To bypass the large geographical distance between us and the domain experts and to take their busy schedule as physicians into account, the case study was done completely web-based.
+\paragraph{Setup.} To bypass the large geographical distance between us and the domain experts and to take their busy schedule into account, the case study was done completely web-based.
 %
@@
 \paragraph{Procedure.}
-At first, we controlled mouse and keyboard of the participants pc and demonstrated the basic functionalities of the system.
+At first, we controlled mouse and keyboard of the participants PC and demonstrated the basic functionalities of the system.
 %
@@
 %
-The radiologist was looking for correlations with shape related features in the data, finding that they correlate with \emph{leg pain}, \emph{age}, \emph{body height} and \emph{hormone replacement therapy status}.
+The radiologist was looking for correlations with shape-related features in the data, finding that they correlate with \emph{leg pain}, \emph{age}, \emph{body height} and \emph{hormone replacement therapy status}.
 %
@@
 %
-This pointed out problems with the used categorization method splitting numerical variables into equally sized ordinal bins.
-%
-If an feature contains outliers, such as \emph{waist circumference} (e.g. by subjects with morbid obesity) this approach leads to sparse categories, making it hard to calculate associations.
+This pointed out problems with the used categorization method splitting numerical variables into equally-sized ordinal bins.
+%
+If a feature contains outliers, such as \emph{waist circumference} (e.g. by subjects with morbid obesity) this approach leads to sparse categories, making it hard to calculate associations.
 %
@@
 
-A \emph{multivariate analysis} using the \emph{Cram\'{e}r's V} contingency values for subjects with strong lumbar spine curvature showed, that these subjects are prominetly females between 58-74 years who also report pain radiating from their back into other body regions Fig.~\ref{fig:application}(a).
+A \emph{multivariate analysis} using the \emph{Cram\'{e}r's V} contingency values for subjects with strong lumbar spine curvature showed, that these subjects are primarily females between 58-74 years who also report pain radiating from their back into other body regions Figure~\ref{fig:application}(a).
 %
@@
 %
-The second expert started also with a \emph{multivariate analysis} using the contingency matrix to analyze non-image-derived features, such as age-associated parameters like \emph{income}, \emph{blood fat values} or \emph{number of born children}, but found no associations of interest.
+The second expert started also with a \emph{multivariate analysis} using the contingency matrix to analyze non-image features, such as age-associated parameters like \emph{income}, \emph{blood fat values} or \emph{number of born children}, but found no associations of interest.
 %
@@
 %
-Contingency values for this cluster revealed correlations with \emph{leg fatigue}, \emph{physical heavy work}, \emph{body weight}, \emph{dyspnoea} and \emph{headache intensity}.
+Contingency values for this cluster revealed correlations with \emph{leg fatigue}, \emph{physically heavy work}, \emph{body weight}, \emph{dyspnoea} and \emph{headache intensity}.
 %
@@
 \\\\
-The experts emphasized the importance of methods providing overview over the data for hypothesis generation.
+The experts emphasized the importance of methods providing an overview over the data for hypothesis generation.
 %
@@
 %
-We observed, that even though the visualization may treat every feature equally, the domain experts is more likely interested in features they are familiar with and have personal clinical experience with.
+We observed, that even though the visualization may treat every feature equally, the domain experts are more likely interested in features they are familiar with and have personal clinical experience with.
 % TODO: Contingency Matrix or Adjecency Matrix
@@
 %
-For one, age influences the lumbar spine shape, as seen in Fig~\ref{fig:application}~(c), while the differences between subjects with and without back pain are small.
-%
-The major difference seen in the visualization are therefore presented by the age feature, masking differences via the back pain parameter.
+For one, age influences the lumbar spine shape, as seen in Figure~\ref{fig:application}~(c), while the differences between subjects with and without back pain are small.
+%
+The major difference seen in the visualization are therefore related to the age feature, masking differences via the back pain parameter.
 %
@@
 %
-As seen in Fig~\ref{fig:application}~(c), subjects reporting back pain are even the majority, which makes it difficult to extract parameters which reliably describe back pain.
+As seen in Figure~\ref{fig:application}~(c), subjects reporting back pain are even the majority, which makes it difficult to extract parameters which reliably describe back pain.
 %
@@
 %
-\emph{Body height} was explained to be a confounder for \emph{gender}, since female subjects smaller on average than male subjects.
+\emph{Body height} was explained to be a confounder for \emph{gender}, since female subjects are smaller on average than male subjects.
 %
@@
 %
-The next step is to use a variable describing the strength of back pain and only analyzing strong back pain!
+The next step is to use a variable describing the strength of back pain and only analyzing strong back pain.
 
@@
 \paragraph{Analysis 2.}
-Hypothesis: ``\emph{Back pain is related to lumbar spine deformation.}''.
+Hypothesis: ``\emph{Back pain is related to lumbar spine deformation}''.
 %
@@
 %
-Therefore the feature ``\emph{Did you experience back pain in the past three months?}'' is dropped into the canvas area.%, as seen in previous Fig.~\ref{ToDoA}.
-%
-Then, the features derived from the clustering results for both subject groups are dropped into the canvas area as well (Fig.~\ref{fig:application} (d)).
-%
-While the clustering algorithm finds only three homogenous cluster close to the mean shape for subjects reporting no back pain.
+Therefore the feature ``\emph{Did you experience back pain in the past three months?}'' is dropped into the canvas area. %, as seen in previous Fig.~\ref{ToDoA}.
+%
+Then, the features derived from the clustering results for both subject groups are dropped into the canvas area as well (Figure~\ref{fig:application} (d)).
+%
+While the clustering algorithm finds only three homogenous clusters close to the mean shape for subjects reporting no back pain.
 %
@@
 %
-A \emph{multivariate analysis} using the pivot tables put gender and strong back pain in context to cluster affiliation (Fig.~\ref{fig:application} (d)). 
+A \emph{multivariate analysis} using the pivot tables put gender and strong back pain in context to cluster affiliation (Figure~\ref{fig:application} (d)). 
 %
@@
 %
-Both domain experts concluded positively over the benefits of the \emph{IVA} approach when analyzing image based cohort study data.
+Both domain experts concluded positively over the benefits of the \emph{IVA} approach when analyzing image-based cohort study data.
 %
@@
 %
-The medical explanation is that it represents a angular point of curvature of the lumbar spine.
+The medical explanation is that it represents an angular point of curvature of the lumbar spine.
 %
@@
 %
-For the visual comparison itself, the radiologist proposed an abstraction the representation them into representative land marks, such as centers of the vertebrae and cardinal points of the lumbar spine canal.
+For the visual comparison itself, the radiologist proposed an abstraction the representation them into representative landmarks, such as centers of the vertebrae and cardinal points of the lumbar spine canal.
 
@@
 %
-These outliers only of high interest for finding pathological subjects. %but not wanted when new parameters describing risk factors are established.
+These outliers are only of high interest for finding pathological subjects. %but not wanted when new parameters describing risk factors are established.
 %
@@
 %
-The large data base associated with image based epidemiological data remains on the server machine and has not to be moved tediously using external hard disks.
+The large data base associated with image-based epidemiological data remains on the server machine and has not to be moved tediously using external hard disks.
 %

### 2014-06-20 draft.tex
Added new Matrix Visualization Figure; Worked on Data Chapter but its not quite done yet

@@
 %
-\add{We implement the presented methods using modern web technologies, such as \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}, making them easily accessible for the domain experts.
+\add{We implemented the presented methods using modern web technologies, such as \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}, making them easily accessible for the domain experts.
 %
@@
 
-\section{Epidemiological Data} \label{EpidemiologicalData}
+\subsection{Epidemiological Data} \label{EpidemiologicalData}
 \com{Primary Reviewer Feedback: Another weakness is on the data analysis; while there is information about the data within the text (including 2.2, 4.1, and 5.2) I would like to see a section that focuses on the data and the challenges thereof.
 This chapter needs to be enhanced! Maybe own section!}
+\begin{itemize}
+	\item \com{OpenRefine erwaehnen? Wurde von uns ja nicht mehr verwendet}
+	\item \com{"I was unsure which type of Image data you had - a better description and the analysis of the data would help here"}
+	\item \com{Export from SPSS to CSV/JSON - "How do you ensure the semantics of the data is carried over (e.g. for ordinal data)"}
+\end{itemize}
+\add{Access to epidemiological data is limited and usually controlled by ethics committees.
+%
+Large studies, such as the SHIP require written proposals reasoning about each requested features and the expected outcome.
+%
+Since the number of features and modalities is steadily increasing, the epidemiological community is aware of the lack of specialized analysis methods, making an acceptance more likely.
+%In our experience, the applying for a large amount of features required for a \underline{IVA} system requires 
+%
+Due to the high sensibility of the data, all identity revealing information of a person, such as name or place of residence is anonymized.
+}
+% --- PARAGRAPH ---
+
 Epidemiological data are highly heterogenous and incomplete.
@@
 
-\subsection{Image acquisition} \label{ImageAcquisition} Imaging techniques involving ionizing radiation for the subject are not suitable for ethical reasons.
+\subsubsection{Image acquisition} \label{ImageAcquisition} Imaging techniques involving ionizing radiation for the subject are not suitable for ethical reasons.
 %
@@
 
-\subsection{Image analysis} \label{ImageAnalysis} Decisions have to be made on how image data are \emph{compared} and \emph{quantified}.
+\subsubsection{Image analysis} \label{ImageAnalysis}
+\com{Take up huge amount of space!}
+%
+Decisions have to be made on how image data are \emph{compared} and \emph{quantified}.
 %
@@
 % \\
+
+\subsection{The Study of Health in Pomerania (SHIP)}
+After the pioneering Rotterdam study (started in 1990), several MR imaging study initiatives have evolved.
+%
+They slightly differ in clinical focus, acquired data and epidemiological research questions.
+%
+Starting in 1997 with a cohort consisting of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
+%
+Data is collected without focus on a group of diseases.
+%
+This allows the data set to be queried regarding many different diseases and conditions.
+%
+Subjects were examined in a 5-year time span, continuously adding new parameters including MRI scans in the last iteration \cite{Hegenscheid2009}.
+%
+The MRI protocol features a rich number of sequences.
+%
+A second cohort \texttt{SHIP-Trend} was established in 2008.
+%
+The protocols for examining the subjects between \texttt{SHIP} and \texttt{SHIP-Trend} remained the same, making them comparable.
+%
+The overall examination time for each person attending the study is two days.
+
+% \subsection{Data Analysis Challenges}
+% The challenges of analyzing epidemiological data are diverse.
+% %
+% Risk factors of a disease are rarely limited to one epidemiological feature and most often feature combinations.
+% %
+% Confounding effects between these features need to be identified and eliminated, which requires large experience of the experts.
+% %
+% Analyzing large heterogenous feature spaces is difficult due to missing techniques providing \emph{overview visualizations}, forcing epidemiologists to stick to their hypothesis-driven workflow described in Section~\ref{EpidemiologicalWorkflow}.
+%
+% Another great problem is how image data is analyzed in context with non-image data.
+% %
+% Epidemiological use of image data is for the most part restricted
+
 % - Missing Data\\
@@
 % 	- show image data of multiple subjects
-
-\subsection{The Study of Health in Pomerania (SHIP)}
-After the pioneering Rotterdam study (started in 1990), several MR imaging study initiatives have evolved.
-%
-They slightly differ in clinical focus, acquired data and epidemiological research questions.
-%
-Starting in 1997 with a cohort consisting of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
-%
-Data is collected without focus on a group of diseases.
-%
-This allows the data set to be queried regarding many different diseases and conditions.
-%
-Subjects were examined in a 5-year time span, continuously adding new parameters including MRI scans in the last iteration \cite{Hegenscheid2009}.
-%
-The MRI protocol features a rich number of sequences.
-%
-A second cohort \texttt{SHIP-Trend} was established in 2008.
-%
-The protocols for examining the subjects between \texttt{SHIP} and \texttt{SHIP-Trend} remained the same, making them comparable.
-%
-The overall examination time for each person attending the study is two days.
-
+% ---------------------------------------------------------------------------------
+% ----------------------------------- Section -------------------------------------
+% ---------------------------------------------------------------------------------
 \section{Prior and Related Work}
@@
 %
-This can be achieved by using data wrangling tools such as \texttt{OpenRefine}\footnote{Developed by Google, Open Source; \url{http://openrefine.org}}, which also validate the data (find missing data, clean up bad formatting, transform scales).
-%
-%Including the data dictionary, which stores information about each manifestation of a feature, helps to get a detailed description of data variables and the meaning and unit of measurement of their values.
-%A data dictionary stores information about each manifestation of a feature and is stored as lookup table to get a detailed description of data variables and the meaning and unit of measurement of their values
+%This can be achieved by using data wrangling tools such as \texttt{OpenRefine}\footnote{Developed by Google, Open Source; \url{http://openrefine.org}}, which also validate the data (find missing data, clean up bad formatting, transform scales).
+%
 A data dictionary stores information about each manifestation of a feature.
@@
 Detailed description of data variables, its meaning as well as unit of measurement are stored as a lookup table.
+%
+\add{
+Since it lacks export methods for data dictionaries, we used \texttt{SPSS} to export our data to the 'SAS v9+' format, which writes out the data labels and saved the data values as non-labeled CSV.
+%
+A short script combined both data sources to a JSON file.
+%
+The data types had to be transferred manually.
+}
 %Exporting the data dictionary, which stores information about each manifestation of a feature, is also an important step to get a detailed description of data variables and the meaning and unit of measurement of their values.
@@
  \includegraphics[width=1\textwidth, resolution=300]{figures/visualization}
- \caption{\com{Add: Explanation of the color coding}(Left) Screenshot from the front-end, which is divided as follows: (a) The sidebar containing all features as well as the groups defined in the analysis process; (b) the canvas area where features can be added via drag and drop and the visualization can be chosen automatically according to the data type; (c) the interactive pivot table showing the exact numbers for each displayed feature combination.
-The data displayed is used to analyze the lumbar spine.
-Features can be added freely on the canvas via drag and drop.
+ \caption{\com{Add: Explanation of the color coding, ToDo: Selektion in Abbildung 3 einbauen und daran Selektion erklären}
+ %
+ (Left) Screenshot from the front-end, which is divided as follows: (a) The sidebar containing all features as well as the groups defined in the analysis process; (b) the canvas area where features can be added via drag and drop and the visualization can be chosen automatically according to the data type; (c) the interactive pivot table showing the exact numbers for each displayed feature combination.
+ %
+ The data displayed is used to analyze the lumbar spine. Features can be added freely on the canvas via drag and drop.
+ %
  Dropping the \emph{gender} parameter on the already plotted \emph{body size} container creates a mosaic plot combining both features (right).
@@
 % \begin{itemize}
-% 	\item [?] Was sind die requirements des tools?
-% 	\item [?] how it was decided to create this particular design
-% 	\item which kind of User interaction was made?
-% 	\item Spezielle Anforderungen durch Webseitenformat, weggehen von allgemeinen UI Elementen wie Menubar, keine Rechtsklickmenüs.
-% 	\item Entschieden eine saubere UI zu bauen, Unterscheide durch Whitespace akzentuieren. Interface in Fig. 3 zu sehen.
-% 	\item Wie wird interagiert, wie wird gebrusht, was wird gelinkt?
-% 	\item wie wird das Clustering initialisiert?
+% 	\item [x] Was sind die requirements des tools?
+% 	\item [x] how it was decided to create this particular design
+% 	\item [x] which kind of User interaction was made?
+% 	\item [x] Spezielle Anforderungen durch Webseitenformat, weggehen von allgemeinen UI Elementen wie Menubar, keine Rechtsklickmenüs.
+% 	\item [x] Entschieden eine saubere UI zu bauen, Unterscheide durch Whitespace akzentuieren. Interface in Fig. 3 zu sehen.
+% 	\item [x] Wie wird interagiert, wie wird gebrusht, was wird gelinkt?
+% 	\item [x] wie wird das Clustering initialisiert?
 % 	\item auf Kontingenzmatrix mit verweisen als Teil des Interfaces (dieses auch noch anpassen)
 % \end{itemize}
-Early it became that we have to rely a lot on online communication with our clinical partners due to their busy schedule and our large spatial distance towards each other.
+Early it became clear that we have to rely a lot on online communication with our clinical partners due to their busy schedule and our large spatial distance towards each other.
 %
@@
 	%
-	\item The \emph{contingency view} depicts relations for features in the canvas in an adjacency matrix.
+	\item The \emph{contingency view} depicts relations for features in the canvas in an adjacency matrix and a \emph{contingency list}.
 \end{itemize}
@@
 If a feature is dropped on an existing plot, the visualization changes dynamically to properly make them comparable (Figure~\ref{fig:visualization} (right)).
-%
+% --- PARAGRAPH ---
+
+Subject groups can be brushed via a double-click on it's representative in the visualizations.
+%
+Holding down the shift key allows to select multiple manifestations at once.
+%
+Brushed groups act as reference for the shape-visualization, calculating distances based on the mean-shape of the brushed selection.
+%
+%This allows to highlight distances between subjects
+The share of subjects of this sub group is linked to all other views (Figure~\ref{fig:visualization} (left)).
+%
+If the user for example selects all female subjects in a visualization of gender distribution, all other displayed meshes are color coded with their distance to the female mean and the share of female subjects is highlighted in the information visualization.
 %Elements can be brushed using 
@@
 The contingency matrix visualization is an appropriate \emph{overview} visualization, something the epidemiological community lacks and is in great need of.
-%\com{Contingency Matrix excellent for Overview}
-\paragraph{Initialization and Interaction}
+
+\paragraph{Contingency Pane}
+Dropping a feature into the canvas area adds an entry for each manifestation of it to the \emph{contingency matrix}.
+%
+Testing sessions revealed that it was very tedious to open the matrix every time a new feature is added. 
+%A simple list of features, which correlate with the last added feature.
+%
+The result was the \emph{contingency pane}, a table containing correlating features for the last added visualizations in descending order of the \emph{Cram\'{e}r's V} value.
+%
+\emph{Contingency pane} entries can be dragged and dropped into the \emph{canvas area} just like features in the \emph{sidebar}.
+% Build after user request
+%
+\paragraph{Initialization and Clustering}
 Using feature suggestion allows to initialize the system with a set of potentially interesting. visualizations.
@@
 Also, we observed that the variables of interest is dependent on the specialization of the domain expert (explained in detail in Section~\ref{application}).
+% --- PARAGRAPH ---
 
@@
 A status indicator at the bottom of the screen keeps the user informed about the pending clustering result, since the process can take up to ten seconds.
+%
+Clustering result are listed in their own category in the \emph{sidbar}.
 %Since a clustering process can take up to ten seconds, a status indicator at the bottom of the screen keeps the user informed.

### 2014-06-21 draft.tex
Edited Data Chapter. Added Bernhards Review.

@@
 Cohort studies often include medical image data.
-%
-Segmentation of these data would allow for anatomical structure analysis and enable correlation with other features.
+% .
+Segmentation of these data enables an anatomical structure analysis and an analysis of possible correlations with other features.
 %
@@
 %
-\rem{The easy exchange of developed methods driven by modern web technologies intends to trigger a fast feedback loop between us and the epidemiologists}.
+\rem{The easy exchange of developed methods driven by modern web technologies intends to trigger a fast feedback loop between computer scientists and epidemiologists}.
 % http://wiki.infowiss.net/Gulf_of_execution ?
@@
  \caption{\emph{IVA} tools complement parts of the epidemiological workflow instead of replacing them.
+ %
  The appropriate combination of statistical and interactive-driven analysis shows promising potential to unveil the information in the data.
-(a) shows the standard epidemiological workflow, (b) the \emph{IVA} supported one.
-The iterative red highlighted part is called the \emph{IVA Loop} and is described in more detail in Figure~\ref{fig:InteractionLoop}.}
+ %
+ The standard epidemiology workflow consists of four steps (a).
+ %
+ IVA complements this workflow and is used to elaborate the feature selection (b).
+ %
+ The iterative red highlighted part is called the \emph{IVA Loop} and is described in more detail in Figure~\ref{fig:InteractionLoop}.
+ %(a) shows the standard epidemiological workflow, (b) the \emph{IVA} supported one.
+The iterative red highlighted part is called the \emph{IVA Loop} and is described in more detail in Figure~\ref{fig:InteractionLoop}.
+ %
+ }
   \label{fig:WorkflowComparison}
@@
 \end{enumerate}
-The workflow is shown in Figure~\ref{fig:WorkflowComparison} (a) and serves as orientation for our approach although it does not consider any kind of image data.
+The workflow is shown in Figure~\ref{fig:WorkflowComparison} (a) and serves as orientation for our approach although it does not consider image data.
 %
@@
 %
-Longitudinal studies require the acquired attributes to be comparable for evaluation.
-%
-If the data acquisition process changes, an information bias is introduced to the data, hampering inference in and between acquisition cycles.
+Longitudinal studies require the acquired features to be comparable for evaluation.
+%
+%If the data acquisition process changes, an information bias is introduced to the data, hampering inference in and between acquisition cycles.
 %
@@
 %
-These groups strongly depend on the condition of interest and therefore there is no standard on how to categorize them.
-
-To determine whether a subject is prone to be affected by a certain disease, \emph{relative risks} are expressed through the evaluation of p-values, which indicate statistical significance.
+These groups strongly depend on the condition of interest and therefore there is no standard for their categorization.
+% --- NEW PARAGRAPH
+
+%
+To determine whether a subject is prone to be affected by a certain disease, \emph{relative risks} are determined including confidence intervals that indicate the certainty of that feature being a risk factor.
+%To determine whether a subject is prone to be affected by a certain disease, \emph{relative risks} are expressed through the evaluation of p-values, which indicate statistical significance.
 %
@@
 %
-Analysis later showed that women applied mostly for strongly restricted subjects and when only looking acceptance divided by departments there was actually a small bias toward women.
+Analysis later showed that women applied mostly for strongly restricted subjects and when only looking at acceptance divided by departments there was actually a small bias toward women.
 %
@@
 \subsection{Epidemiological Data} \label{EpidemiologicalData}
-\com{Primary Reviewer Feedback: Another weakness is on the data analysis; while there is information about the data within the text (including 2.2, 4.1, and 5.2) I would like to see a section that focuses on the data and the challenges thereof.
-This chapter needs to be enhanced! Maybe own section!}
-\begin{itemize}
-	\item \com{OpenRefine erwaehnen? Wurde von uns ja nicht mehr verwendet}
-	\item \com{"I was unsure which type of Image data you had - a better description and the analysis of the data would help here"}
-	\item \com{Export from SPSS to CSV/JSON - "How do you ensure the semantics of the data is carried over (e.g. for ordinal data)"}
-\end{itemize}
+% \com{Primary Reviewer Feedback: Another weakness is on the data analysis; while there is information about the data within the text (including 2.2, 4.1, and 5.2) I would like to see a section that focuses on the data and the challenges thereof.
+% This chapter needs to be enhanced! Maybe own section!}
+%
+% \com{Bernhard: In Section 2 reichen 1-2 allgemeine Sätze, die typische Probleme der Daten beschreiben.}
+% \begin{itemize}
+% 	\item \com{OpenRefine erwaehnen? Wurde von uns ja nicht mehr verwendet}
+% 	\item \com{"I was unsure which type of Image data you had - a better description and the analysis of the data would help here"}
+% 	\item \com{Export from SPSS to CSV/JSON - "How do you ensure the semantics of the data is carried over (e.g. for ordinal data)"}
+% \end{itemize}
 \add{Access to epidemiological data is limited and usually controlled by ethics committees.
@@
 %
-For ethical, legal or medical reasons some features cannot be gathered for each subject.
+For ethical or medical reasons some features cannot be gathered for each subject.
 %
@@
 Medical examinations mostly comprise categorical (e.g. levels of back pain) and continuous values (e.g. age or body size).
+%
+\add{Analysis of the data is usually carried out by calculating correlations,
+which is challenging due to the data type heterogeneity.
+%
+Parameter correlation can also be associated with confounding, which can not be automatically predicted but rather has to be judged by a domain expert.
+%
+Sparse populated features are also hard to assess statistically as to few data samples have a higher chance to distort the real underlying distributions.
+%Also, correlation can both be of epidemiological interest but 
+}
 
@@
 %
-Decisions have to be made on how image data are \emph{compared} and \emph{quantified}.
+Decisions have to be made about \emph{comparison} and \emph{quantification} of image data.
 %
@@
 %
-Since reliable and efficient segmentation techniques for these data are not available in general, epidemiologists are forced to measure the data by hand, which is a very tedious work with respect to the number of necessary landmarks and the number of subjects.
+Since reliable and efficient segmentation techniques are not available in general, epidemiologists are forced to measure the data by hand, which is a very tedious work with respect to the number of necessary landmarks and the number of subjects.
 %
@@
 \subsection{The Study of Health in Pomerania (SHIP)}
-After the pioneering Rotterdam study (started in 1990), several MR imaging study initiatives have evolved.
+After the pioneering Rotterdam study (started in 1990), several MR imaging study initiatives were initiated.
 %
@@
 %
-Starting in 1997 with a cohort consisting of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
+Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
 %
@@
 %
-This allows the data set to be queried regarding many different diseases and conditions.
+This allows the data set to be queried regarding many diseases and conditions.
 %
@@
 \section{Prior and Related Work}
-Designing a visualization, which conveys all data aspects equally, is challenging.
+%Designing a visualization, which conveys all data aspects equally, is challenging.
 %
@@
 %Their focus on \emph{hypothesis generation} using parallel assessment of multiple data features makes the work of Turkay and colleagues closest to ours albeit our emphasis on processing models derived from medical image data segmentation and variables with categorical manifestations \cite{Turkay2013}.
-\rem{Their} \add{Turkay and colleagues \cite{Turkay2013}} focus on \emph{hypothesis generation} using parallel assessment of multiple data features makes \rem{the work of Turkay and colleagues} \add{their work} closest to ours albeit our emphasis on processing models derived from medical image data segmentation and variables with categorical manifestations \rem{\cite{Turkay2013}}.
+\rem{Their} \add{Turkay and colleagues \cite{Turkay2013}} focus on \emph{hypothesis generation} using parallel assessment of multiple data features makes \rem{the work of Turkay and colleagues} \add{their work} closest to ours albeit our emphasis is on processing models derived from medical image data segmentation and variables with categorical manifestations \rem{\cite{Turkay2013}}.
 %
@@
 %
-Statistical measures of continuous variables, such as mean, standard deviation, skewness, or inter-quartile range, are used to create \emph{dimension plots} that make them comparable with respect to the derived descriptive measures, such as voxel number of a segmented structure.
-%
-%These transform dimensions into data points and make them comparable with respect to the derived descriptive measures, making them comparable.
+Statistical measures of continuous features, such as mean, standard deviation, skewness, or inter-quartile range, are used to create \emph{dimension plots}.
+%that make them comparable with respect to the derived descriptive measures, such as voxel number of a segmented structure.
+%
+These transform dimensions into data points and make them comparable with respect to the derived descriptive measures, making them comparable.
 %
@@
 %\emph{Over-fitting} of hypotheses to the data may be imposed  is a danger pointed out by the authors when hypotheses to the data based on observations of changes in these plots the measure highlights only subsets of statistical changes.
-Hypotheses based on observations of changes in these plots may impose \emph{overfitting} to the data because the measures highlight only subsets of statistical changes.
+Hypotheses based on observations of changes in these plots may impose \emph{overfitting} to the data because the measures highlight only parts of the statistical changes.
 %
@@
 \paragraph{Visualizing Heterogenous Non-Image Data.}
-Generalized Pairs Plots (\texttt{GPLOM´S}) are an information visualization technique comparing heterogenous variables pairwise using a plot-matrix grouped by type \cite{Francois2013}.
+Generalized Pairs Plots (\texttt{GPLOM´S}) are an information visualization technique comparing heterogenous features pairwise using a plot-matrix grouped by type \cite{Francois2013}.
 %
@@
 \paragraph{Commercial Data Visualization.}
-Commercial systems such as \texttt{Tableau}\footnote{Owned by Tableau Software; \url{http://tableausoftware.com}} or \texttt{Spotfire}\footnote{Owned by TIBCO; \url{http://spotfire.tibco.com}} provide a rich user interface that enables a Visual Analytics approach without the need of writing any code.
-%
-With little effort, linked views can be created, but the data processing possibilities such as derivation of new variables or the 3D rendering capabilities are very limited.
-%
-These systems with their origin in business intelligence applications do not support epidemiological workflows, in particular workflows that also consider medical image data.
+Commercial systems such as \texttt{Tableau}\footnote{Owned by Tableau Software; \url{http://tableausoftware.com}} or \texttt{Spotfire}\footnote{Owned by TIBCO; \url{http://spotfire.tibco.com}} provide a rich user interface that enables a Visual Analytics approach.
+%
+With little effort, linked views can be created, but the data processing possibilities, such as derivation of new variables or the 3D rendering capabilities, are very limited.
+%
+These systems with their origin in business intelligence applications do not support epidemiological workflows well, in particular workflows that also consider medical image data.
 % Image Data
@@
 \paragraph{Visualizing Shape Variance.}
-Comparing tissue between many subjects requires methods which allow for shape variance visualizations.
+Comparing tissue between many subjects requires shape variance visualizations.
 %
@@
 %
-User specified regions of interests are mapped to associated anatomic covariation using tensor visualization.
+User-specified regions of interests are mapped to associated anatomic covariation using tensor visualization.
 %Anatomic covariation induced by user input is displayed using tensors 
@@
 %
-This method allowed for rapid hypotheses validation and was able to reproduce textbook knowledge.
+This method enables rapid hypotheses validation and was able to reproduce textbook knowledge.
 %
@@
 \paragraph{SHIP Data Analysis.}
-Klemm et al. \cite{Klemm2013VMV} visualized lumbar spine variabilities based on an semi-automatic shape detection algorithm of 490 participants of the \texttt{SHIP-2} cohort.
+Klemm et al. \cite{Klemm2013VMV} visualized lumbar spine variabilities based on a semi-automatic shape detection algorithm of 490 participants of the \texttt{SHIP-2} cohort.
 %
@@
 %
-When applying clustering techniques on the non-image data it was found that \texttt{k-Prototypes} and \texttt{DBSCAN} is appropriate in the epidemiological context, but is strongly dependent on the chosen variables and distance measure \cite{Klemm2014BVM}.
-%
-Niemann et al. \cite{Niemann2014} presented an interactive data mining tool for the assessment of risk factors on hepatic steatosis, the fatty liver disease.
-%
-Association rules created by data mining methods can be analyzed interactively with their presented tool and highlight potentially overlooked features.
+When applying clustering techniques to the non-image data it was found that \texttt{k-Prototypes} and \texttt{DBSCAN} are appropriate in the epidemiological context, but is strongly dependent on the chosen variables and distance measure \cite{Klemm2014BVM}.
+%
+Niemann et al. \cite{Niemann2014} presented an interactive data mining tool for the assessment of risk factors of hepatic steatosis, the fatty liver disease.
+%
+Association rules created by data mining methods can be analyzed interactively with their tool and highlight potentially overlooked features.
 % *********************** / Our own Work ***********************
@@
 %
-Oeltze et al. \cite{Oeltze2007} combined a linked view representation of results from a statistical analysis with feature localizations of the tissue perfusion with the goal of its evaluation.
+Oeltze et al. \cite{Oeltze2007} combined a linked view representation of results from a statistical analysis with feature localizations of the tissue perfusion. % with the goal of its evaluation.
 %
@@
 %--except the variables derived the segmentation model itself.
-While we take similar steps when analyzing the data, such as employing statistical tests, our data are not associated in a geographical or temporal context, which affects the techniques described in the following section.
+While we take similar steps when analyzing the data, such as employing statistical tests, our data are not associated to a geographical or temporal context, which affects the techniques described in the following section.
 
@@
 \section{Image Centric Cohort Study Data in Interactive Visual Analysis Context} \label{Image Centric Cohort Study Data in Interactive Visual Analysis Context}
-We described the epidemiological workflow as a sequence of steps taken by domain experts which need to comprise reproducibility and statistical integrity (recall Subsection~\ref{EpidemiologicalWorkflow}).
+We described the epidemiological workflow followed by domain experts where reproducibility and statistical integrity are important aspects (recall Subsection~\ref{EpidemiologicalWorkflow}).
 %
@@
 %
-Introducing the \emph{IVA} principle to the epidemiological application domain aims to compensate its weaknesses rather than replacing the existing workflow.
+%Introducing the \emph{IVA} principle to the epidemiological domain aims to compensate its weaknesses rather than replacing the existing workflow.
+Introducing the \emph{IVA} principle to the epidemiological domain aims to compensate the weaknesses of the existing workflow, rather than replacing it.
 %
@@
 This becomes even more important when the workflow is adapted to the analysis of the medical image data, where
-domain experts annotate landmarks tediously to derive metrics, such as diameters.
+domain experts annotate landmarks tediously to derive measures, such as diameters.
 %
@@
 %
-It is easily possible that left out information would heavily influence the result.
+It is easily possible that omitted information would heavily influence the result.
 %
@@
 \\\\
-\emph{IVA} tries to illuminate the black box by making the domain experts part of an iterative feature selection process, which is shown in Figure~\ref{fig:WorkflowComparison} (b) as part of the epidemiological workflow.
+\emph{IVA} tries to illuminate the black box by making the domain experts part of an iterative feature selection process (see Fig.~\ref{fig:WorkflowComparison} (b)).
+%\emph{IVA} tries to illuminate the black box by making the domain experts part of an iterative feature selection process, which is shown in Figure~\ref{fig:WorkflowComparison} (b) as part of the epidemiological workflow.
 %
@@
 \begin{itemize} 
-	\item \emph{physical views} \cite{Oeltze2013}, such as direct volume rendering, show information in the context of the spatiotemporal observation space \cite{Oeltze2007}, while
+	\item \emph{physical views} \cite{Oeltze2013}, such as direct volume rendering, show information in the context of the spatio-temporal observation space \cite{Oeltze2007}, while
 	\item \emph{attribute views}, such as scatter plots and parallel coordinates, show relationships between multiple data attributes.
 \end{itemize}
-Transferred to epidemiological data, the residential area of cohort subjects could be interpreted as space, the different assessment cycles of a longitudinal study as time, and the image and non-image data as dependent variables.
+Transferred to epidemiological data, the residential area of cohort subjects could be interpreted as \emph{space}, the different assessment cycles of a longitudinal study as \emph{time}, and the image and non-image data as \emph{dependent variables}.
 %
@@
 %
-Clustering algorithms can be used to investigate associations between shape groups and other non-image-based variables.
+Clustering algorithms can be used to investigate associations between shape groups and other non-image variables.
 %
@@
 %
-These associations may also be summarized using pivot tables which are popular in epidemiology.
+These associations may also be summarized using pivot tables which are widely used in epidemiology.
 
 \subsection{Data Preprocessing} \label{Data Preprocessing}
-Transformation operations on the data to prepare it for an \emph{IVA} system are denoted as data preprocessing.
+A number of transformations are carried out in the data preprocessing stage.
+%Transformation operations on the data to prepare it for an \emph{IVA} system are denoted as data preprocessing.
 \paragraph{Non-Image Data. }
 %
-Multimodal features require different techniques.
+%Multimodal features require different techniques.
 %
@@
 %
-\add{
-Since it lacks export methods for data dictionaries, we used \texttt{SPSS} to export our data to the 'SAS v9+' format, which writes out the data labels and saved the data values as non-labeled CSV.
-%
-A short script combined both data sources to a JSON file.
-%
-The data types had to be transferred manually.
-}
 %Exporting the data dictionary, which stores information about each manifestation of a feature, is also an important step to get a detailed description of data variables and the meaning and unit of measurement of their values.
@@
 \paragraph{Image Data. }
-Information about anatomical structure, such as diameter or volumes, is extracted from the image data.
+Information about anatomical structures, such as diameter or volumes, is extracted from the image data.
 %
@@
 %
-In principle, model-based approaches have shown to be effective for detection \cite{Rak2013} and segmentation \cite{Gloger2010, Gloger2012}.
+In principle, model-based approaches are effective for detection \cite{Rak2013} and segmentation \cite{Gloger2010, Gloger2012}.
 %
@@
  %
- (Left) Screenshot from the front-end, which is divided as follows: (a) The sidebar containing all features as well as the groups defined in the analysis process; (b) the canvas area where features can be added via drag and drop and the visualization can be chosen automatically according to the data type; (c) the interactive pivot table showing the exact numbers for each displayed feature combination.
+ (Left) Screenshot from the front-end, which is divided as follows: (a) The sidebar containing all features as well as the groups defined in the analysis process; (b) the canvas area where features can be added via drag and drop and the visualization is chosen automatically according to the data type; (c) the interactive pivot table showing the exact numbers for each displayed feature combination.
  %
@@
 %-------------------------------- /Figure -------------------------------
-Our proposed \emph{IVA} workflow knows three major steps as illustrated in Figure~\ref{fig:InteractionLoop}: Feature selection, visualization and brushing.
+Our proposed \emph{IVA} workflow consists of three major steps as illustrated in Figure~\ref{fig:InteractionLoop}: Feature selection, visualization and brushing.
 %
@@
 %
-A feature selection from a shape-based clustering, which creates shape groups.
+A feature selection derived from a shape-based clustering, which creates shape groups.
 %
@@
 %
-Brushed regions are treated like features as they divide the subject space just like categorical features.
+Brushed regions are treated like features, as they divide the subject space just like categorical features.
 %
@@
 	%
-	\item The \emph{sidebar}, which contains all epidemiological features. Cluster results are treated like features and are part of the sidebar as well.
-	%
-	\item The interactive \emph{pivot table} gives detailed numerical information of the features in the canvas view. This view on the data comes natural to epidemiologists.
+	\item The \emph{sidebar}, which contains all epidemiological features. As cluster results group features like categorical features and are part of the sidebar as well.
+	%
+	\item The interactive \emph{pivot table} gives detailed numerical information of the features in the canvas view. This view on the data is familiar to epidemiologists.
 	%
@@
 %
-Categorical data are either mapped using bar charts (single features) or mosaic plots (multiple features).
+Categorical data are either mapped to bar charts (single features) or mosaic plots (multiple features).
 %
@@
 %
-Continuous data can be visualized using scatterplots (single features) or parallel coordinates (multiple features), but in epidemiology, this data type is usually categorized into ordinal groups of \emph{equal size}.
+Continuous data can be visualized using scatterplots (two features) or parallel coordinates (multiple features), but in epidemiology, this data type is usually categorized into ordinal groups of \emph{equal size}.
 %
@@
 %
-Adaptive discretization is an option, but imposes overfitting to the data.
-%
-Drawn conclusions based on statistical relationships derived from groups already biased by feature distribution are heavily influenced by the used discretization.
+Adaptive discretization is an option, but imposes possible overfitting to the data.
+%
+Conclusions based on statistical relationships derived from groups already biased by feature distribution are heavily influenced by the used discretization.
 %
@@
 %
-The 3D plots can be navigated using standard mouse inputs, the camera is synchronized between all views to enable direct comparison.
+The 3D plots can be navigated using standard mouse input, the camera is synchronized between all views to enable direct comparison.
 %
@@
 %
-If a feature is dropped on an existing plot, the visualization changes dynamically to properly make them comparable (Figure~\ref{fig:visualization} (right)).
+If a feature is dropped on an existing plot, the visualization changes dynamically to make them comparable (Figure~\ref{fig:visualization} (right)).
 % --- PARAGRAPH ---
@@
 %
-The contingency matrix highlights correlations between all data set features. %of visualized features with features in the data set.
+The contingency matrix highlights correlations between all features. %of visualized features with features in the data set.
 %
@@
 %
-Due to missing ports of sophisticated and specialized libraries/languages, such as the \texttt{Visualization Toolkit} (\texttt{VTK})\footnote{Developed by Kitware Inc; \url{http://vtk.org}} or \texttt{R}\footnote{Open Source; \url{http://r-project.org}} for statistics, many standard methods need to be written from scratch for the web.
+Due to missing ports of sophisticated and specialized libraries/languages, such as the \texttt{Visualization Toolkit} (\texttt{VTK})\footnote{Developed by Kitware Inc; \url{http://vtk.org}} or \texttt{R}\footnote{Open Source; \url{http://r-project.org}} for statistics, many standard methods need to be written from scratch.
 %
@@
 %
+\add{The data contains 30 metric, 7 nominal, 29 ordinal and 62 dichotomous epidemiological features.
+%
+Somatometric features include measures of the human body, such as body height, weight and body fat percentage as well as gender.
+%
+These measures are reliable and complete.
+%
+Other features, such as pain indicators or lifestyle indicators such as physical activity are more subjective and less reliable.
+%
+There are also features missing for each subject, such as features building upon each other (e.g. Do you have high blood pressure? Which medication is prescribed against it?).
+%
+Therefore some manifestations are sparsely populated, which makes statistical evaluation challenging.
+%
+%Another challenge the 
+}
+%
+
 The image data was acquired on a 1.5~Tesla scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany) by four trained technicians in a standardized way.
@@
 To ensure a fast and easy data access outside of statistical processors like \texttt{SPSS}, the data was exported to the JSON file format.
+%
+\add{
+Since it lacks export methods for data dictionaries, we used \texttt{SPSS} to export our data to the 'SAS v9+' format, which writes out the data labels and saved the data values as non-labeled CSV.
+%
+A short script combined both data sources to a JSON file.
+%
+The data types had to be transferred manually.
+}
 %
@@
 %
-Shape-based clustering assesses the variability of a anatomical structure in the context of non-image features such as disease indicators or lifestyle factors.
+Shape-based clustering assesses the variability of an anatomical structure in the context of non-image features such as disease indicators or lifestyle factors.
 

### 2014-06-21 draft.tex
Added further Reviewer comments

@@
 %(Validation).
-The presented concepts are applied expert-guided to gather and evaluate hypotheses about the aging process of the lumbar spine.
+The presented concepts are applied by experts to gather and evaluate hypotheses about the aging process of the lumbar spine.
 %
@@
 %
-A hypothesis-driven analysis usually starts with the selection of features.
-%
-A feature selection derived from a shape-based clustering, which creates shape groups.
+A hypothesis-driven analysis usually starts with the selection of features, or shape groups derived from a shape-based clustering.
 %
@@
 \begin{itemize}
-	\item The \emph{canvas} holding all visualizations. Elements can be added, arranged, resized and removed freely.
-	%
-	\item The \emph{sidebar}, which contains all epidemiological features. As cluster results group features like categorical features and are part of the sidebar as well.
-	%
-	\item The interactive \emph{pivot table} gives detailed numerical information of the features in the canvas view. This view on the data is familiar to epidemiologists.
-	%
-	\item The \emph{contingency view} depicts relations for features in the canvas in an adjacency matrix and a \emph{contingency list}.
+		\item The \emph{sidebar}, which contains all epidemiological features.
+As cluster results group features like categorical features and are part of the sidebar as well (\add{Fig.~\ref{fig:visualization} (a)}).
+	%
+	\item The \emph{canvas} holding all visualizations. Elements can be added, arranged, resized and removed freely (\add{Fig.~\ref{fig:visualization} (b))}.
+	%
+	\item The interactive \emph{pivot table} gives detailed numerical information of the features in the canvas view.
+This view on the data is familiar to epidemiologists (\add{Fig.~\ref{fig:visualization} (c)}).
+	%
+	\item The \emph{contingency view} depicts relations for features in the canvas in an adjacency matrix (\add{Fig.~\ref{fig:similarity}}) and a \emph{contingency list}.
 \end{itemize}
@@
 %
-Epidemiologists are used to process groups based on table representations.
+Epidemiologists are used to analyze groups based on table representations.
 %
@@
 %
-Due to missing ports of sophisticated and specialized libraries/languages, such as the \texttt{Visualization Toolkit} (\texttt{VTK})\footnote{Developed by Kitware Inc; \url{http://vtk.org}} or \texttt{R}\footnote{Open Source; \url{http://r-project.org}} for statistics, many standard methods need to be written from scratch.
+Sophisticated libraries/languages, such as the \texttt{Visualization Toolkit} (\texttt{VTK})\footnote{Developed by Kitware Inc; \url{http://vtk.org}} or \texttt{R}\footnote{Open Source; \url{http://r-project.org}} for statistics, are either not available at all or only accessible through complex client-server systems.
+%
+Therefore, many standard methods had to be written from scratch.
+%
+\add{The user needs a fully updated modern web browser, which supports these new technologies, such as WebGL.}
+%Due to missing ports of sophisticated and specialized libraries/languages, such as the \texttt{Visualization Toolkit} (\texttt{VTK})\footnote{Developed by Kitware Inc; \url{http://vtk.org}} or \texttt{R}\footnote{Open Source; \url{http://r-project.org}} for statistics, many standard methods need to be written from scratch.
+%Due to missing ports of sophisticated and specialized libraries/languages, such as the \texttt{Visualization Toolkit} (\texttt{VTK})\footnote{Developed by Kitware Inc; \url{http://vtk.org}} or \texttt{R}\footnote{Open Source; \url{http://r-project.org}} for statistics, many standard methods need to be written from scratch.
 %
@@
 %
-There are 127 features describing diagnosed diseases, lifestyle factors, women-specific factors, pain indicators, laboratory values and somatometric features.
+\add{
+There are 127 features describing diagnosed diseases, lifestyle factors, women-specific factors, pain indicators, laboratory values and somatometric features for 6.753 subjects (4.420 from \texttt{SHIP-Trend-0} and 2.333 from \texttt{SHIP-2}).
+%
+Since data acquisition protocols between these two cohorts are identical, the features between the two cohorts are comparable.
+}
 %
@@
 
-The image data was acquired on a 1.5~Tesla scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany) by four trained technicians in a standardized way.
-%
-%The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence (676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
-The spine protocol consisted of sagittal T1- and T2-weighted turbo-spin-echo sequences with rather large slice distance only sufficient for characterizing the lumbar spine \cite{Hegenscheid2013}.
+The MRI image data was acquired for each subject on a 1.5~Tesla scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany) by four trained technicians in a standardized way.
+%
+\add{The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence (676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
+}
+%The spine protocol consisted of sagittal T1- and T2-weighted turbo-spin-echo sequences with rather large slice distance only sufficient for characterizing the lumbar spine \cite{Hegenscheid2013}.
 
@@
 %
-We obtained and worked with 983 tetrahedron models of the lumbar spine.
+We obtained and worked with 2.540 tetrahedron models of the lumbar spine.
 %
@@
 This underlines the accessibility of \emph{IVA} tools, which amplifies the natural curiosity of the domain experts in complex medical feature associations.
-%\com{ToDo Fazit über Web-Austausch, das Experten begeistert davon waren und dass es hervorragend funktioniert.
-Kleine Änderungen können sogar live programmiert werden ohne das etwas neu kompiliert werden muss (hier wäre ein Beispiel nett, z.B.
-Größe der Schrift irgendwo.
-Außerdem bringt der Browser viele Funktionen mit, wie z.B.
-die Suchfunktionen.)}
-%She also pointed out the benefits of an automatic categorization of the subjects into shape groups.%$$, which allows for even faster comparison.
-%\com{Employ a more abstract visualization of the spine to capture reliable information and enable better comparison between subjects}
-%
-%Since many large population studies incorporate medical image data this gains importance.
-%\com{1.08 It was pointed out that this technique is also good for epidemiological publication because it is often more insigthfull to show these images instead of long tables.}
-
-%\com{Small User Interface Suggestions: Search-Bar for Subjects, making the elements of the contingency matrix selectable; See which Subject is currently selected}
-%\com{Enhancement contingency matrix: Display exact value of correlations, also information about distribution. More focus on this}
-%\com{Außerdem ein Anwendungsgebiet für die Matrix: Einschätzen metabolisches syndrom, eine Menge von Risikofaktoren die für Koronare Herzkrankheit Marker darstellen und die bereits länger in der Kritik sind.
-Hier kann getestet werden, ob es hierfür nicht bessere Indikatoren noch geben kann!
-(39.32)}
-%\com{Also mosaic plots can be incorporated for several sub groups to see where the main differences are (for different age groups to characterize age)}
-%\com{1.20 More exact segmentations are useful since curvature changes are often the last change happening in a pathological process, after local deformations of the vertebrae}
-
-% ---------------------------------------------------------------------------------------------------------------------------------------
-% ---------------------------------------------------------------------------------------------------------------------------------------
-% ---------------------------------------------------------------------------------------------------------------------------------------
-% \section{Old Application Section}
-% %
-% This section describes how our presented \emph{IVA} workflow is used for an epidemiological use case.
-% %
-% We applied the presented set of techniques to a data set compiled to analyze lower back pain. 
-% %
-% This is one of the most common reasons for an adult to see a physician in the Western civilization \cite{Backpain}.
-% %
-% Epidemiological analysis of lumbar back pain, such as the work of Harreby et al. \cite{Harreby1996}, is largely focused on non-image information.
-% %
-% In comparable studies, only a few shape-related features are included \cite{Lang2011}.
-% %
-% To our knowledge, this is the first approach, where shape-related information of the whole lumbar spine is analyzed along with other epidemiological features.
-% %
-% Determining risk factors in this area can lead to \cite{Fletcher2012}:
-% \begin{itemize}
-% 	\item a better understanding of effects of preventive measures such as occupational health and safety regulations,
-% 	\item prognostic features for diagnosis and treatment of lumbar back pain, and
-% 	\item determination of particularly affected risk groups.
-% \end{itemize}
-% % Determining risk factors in this area can lead to \cite{Fletcher2012} (evtl. weglassen, Dopplung bei Epidemiological Background?):
-% % \begin{itemize}
-% % 	\item a better understanding of effects of preventive measures such as occupational health and safety regulations
-% % 	\item prognostic features for diagnosis and treatment of lumbar back pain
-% % 	\item determination of particularly effected risk groups
-% % \end{itemize}
-% %
-% Characterizing the healthy aging process of the spine is a long-term goal for determining age-normalized probabilities for spine-related diseases by incorporating individual risk factors.
-% 
-% Data confidentiality and ethical reasons prohibit us from accessing the complete \texttt{SHIP} feature space.
-% %
-% Our epidemiological collaborators compiled a feature list, which is a tradeoff between complexity and limitations of the responsible ethics committee.
-% %
-% \subsection{The Lumbar Spine Data Set}
-% We divide the data set into image and non-image data.
-% %
-% There are 127 features describing diagnosed diseases, lifestyle factors, women-specific factors, pain indicators, laboratory values and somatometric features.
-% %
-% The image data was acquired on a 1.5~Tesla scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany) by four trained technicians in a standardized way.
-% %
-% %The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence (676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
-% The spine protocol consisted of sagittal T1- and T2-weighted turbo-spin-echo sequences with rather large slice distance only sufficient for characterizing the lumbar spine \cite{Hegenscheid2013}.
-% 
-% \subsection{Data Preprocessing} \label{application:Data Preprocessing}
-% The data are preprocessed as described in Section~\ref{Data Preprocessing}.
-% %
-% \paragraph{Non-Image Data.} 
-% 
-% To ensure a fast and easy data access outside of statistical processors like \texttt{SPSS}, the data was exported to the JSON file format.
-% %
-% Each feature is stored as an object containing: 
-% \begin{itemize}
-% 	\item the data as array of values--categorical data and error codes are stored using IDs,
-% 	\item the data type (continuous, nominal, ordinal, dichotomous),
-% 	\item a detailed description of the feature, and
-% 	\item the data dictionary translating value or error IDs to the actual values.
-% \end{itemize}
-% %
-% % Binning: https://www.researchgate.net/post/What_metric_best_captures_the_information_loss_when_binning_a_large_data-set_into_a_histogram_representation
-% Continuous variables are discretized to allow for \emph{Cram\'{e}r's V} contingency coefficient assessment.
-% %
-% In consultation with the collaborating epidemiologists, we set the number of groups to five, to allow for contingency assessment.
-% 
-% \paragraph{Image Data.} \label{Image-Data}
-% The lumbar spine was detected in the image data using a hierarchical finite element method by Rak et al. \cite{Rak2013}.
-% %
-% This semi-automatic method requires the user to initialize the tetrahedron-based finite element models (FEM) with a click on the L3 vertebra.
-% %
-% Two user-defined landmarks on the top and bottom of the L3 vertebra are used to obtain an initial height estimation of the model.
-% %
-% It uses a weighted sum of T1- and T2-weighted MR images to detect the lumbar spine shape.
-% %
-% The registered models capture resilient information about the shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae \cite{Klemm2013VMV}.
-% %
-% Due to incorrect initialization, strongly deformed spines, contrast differences and artifacts, the model was not able to detect lumbar spines for all subjects.
-% %
-% We obtained and worked with 983 tetrahedron models of the lumbar spine.
-% %
-% For clustering purposes, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature \cite{Klemm2013VMV}).
-% 
-% \subsection{Shape Visualization and Clustering}
-% %
-% The tetrahedron-based detection model described in Section~\ref{application:Data Preprocessing} consists of corresponding grid points for each structure instance.
-% %
-% This allows for calculation of shape distance and similarity.
-% %
-% This information is used to calculate mean shapes as described in Section~\ref{Interaction- and Visualization Techniques}.
-% 
-% Shape distance is mapped onto color.
-% %
-% For dichotomous variables, the color codes distances between mean shapes of the two groups, for variables with more than two manifestations it encodes the distance to the global mean shape of all subjects (see Figure~\ref{fig:hypothesisfree}).
-% \\\\
-% Shape-based clustering is carried out via agglomerative hierarchical clustering of the spine canal centerlines (recall Section~\ref{application:Data Preprocessing} and \cite{Klemm2013VMV}).
-% %
-% Since the ``correct'' number of clusters in a given group is unknown, an estimate is computed by means of the knee/elbow method \cite{Salvador2004}.
-% %
-% The method has proven to produce sound results on a preliminary data set and was able to reproduce textbook knowledge \cite{Klemm2013VMV}.
-% 
-% \subsection{Exploratory Analysis of the Lumbar Spine Data Set}
-% %
-% Expert-guided analysis assessed the suitability of our approach for supporting both hypothesis-free analysis as well as hypothesis generation. 
-% %
-% The use of \emph{IVA} patterns as presented in Section~\ref{Image Centric Cohort Study Data in Interactive Visual Analysis Context} is denoted accordingly.
-% %
-% \subsubsection{Hypothesis-Free Analysis} \label{Hypothesis free Analysis}
-% \begin{figure}[htb]
-%  \centering
-%  %\includegraphics[width=1.5in]{figures/InteractionLoop}
-%  \includegraphics[width=3.5in]{figures/hypothesisfree}
-%  \caption{(a) Clustering result of all subjects.
-The bar chart height indicates the number of subjects in the cluster.
-The difference to the mean shape is color-coded, whereas blue represents no difference and red a large difference.
-(b) A pivot table relates cluster results to gender and employment status.
-Cluster 8 is striking because of the high proportion of women, while cluster 2, 3, 4 contain almost only men.}
-%  \label{fig:hypothesisfree}
-% \end{figure}
-% %
-% Analyzing the data set without prior hypothesis requires a starting point giving an overview over the data first \cite{Shneiderman1996}.
-% %
-% Shape-based hypothesis-free exploration starts with the \emph{local investigation} pattern by a shape grouping step using shape-based clustering.
-% %
-% The results for this step can be seen in Figure~\ref{fig:hypothesisfree} (a).
-% 
-% Cluster 9 represents subjects with average shape.
-% %
-% Other shapes differ with respect to size, such as cluster 2, 3, 8, 10 where the last one also represents a more straight spine, which is usual for subjects with larger body size.
-% %
-% Cluster 4, 5, 7 contain outliers, characterized by their unusual shape and small number.
-% %
-% Cluster 8 has the second largest number of elements and was therefore of special interest.
-% %
-% Performing a \emph{multivariate analysis} by looking at \emph{Cram\'{e}r's V} contingency values of cluster~8 reveals associations of this group with employment status, gender, body size, age, thyroid nodules and blood fat value.
-% %
-% Another \emph{multivariate analysis} using a pivot table set gender and employment status in relation to clustering affiliation\add{.} 
-% %
-% \add{It s}hows that cluster~8 contains mostly women and also has a larger unemployment rate (see Figure~\ref{fig:hypothesisfree}), while the overall employment rate of women and men in the data set is almost exactly 50\%.
-% %
-% While all observed features seem to be plausible associations related to back pain, the values indicate that cluster~8 contains subjects with chronic back pain radiating to the legs.
-% %
-% Metabolic parameters, such as blood fat and blood sugar, are also possibly associated features.
-% %
-% The employment status is a feature relating to many different lifestyle factors such as income or nutrition as well as age and might act as confounder.
-% \\\\
-% While this approach does not assume any hypothesis, the 127 epidemiological features were selected by domain experts with a wide range on potential influential factors.
-% %
-% This filtering step based on expert experience rules out possible confounder and useless associations and it is therefore arguable whether this represents a disadvantage \cite{Wiley2008}.
-% %It is arguable whether this first filtering step, which is purely based on expert experience, represents a disadvantage, since it rules out many parameters possibly associated with the presented features, but the value of this knowledge would be small \cite{Wiley2008}.
-% %
-% % While this approach does not assume any hypothesis, the data are met with a selection bias when compiling the list of related features by the domain experts.
-% % %
-% % It is arguable if this first filtering step, which is purely based on expert experience, is really a disadvantage, since it rules out many parameters which may interact with the presented features but the value of this knowledge would be small \cite{Wiley2008}.
-% 
-% \subsubsection{Hypothesis-Based Analysis} \label{Hypothesis based Analysis}
-% \begin{figure}[htb]
-%  \centering
-%  %\includegraphics[width=1.5in]{figures/InteractionLoop}
-%  \includegraphics[width=3.5in]{figures/hypothesisbased}
-%  \caption{(a) Dichotomous questionnaire answer to ``\emph{Did you experience back pain in the past three months?}''.
-%  %
-%  Mean shapes between the groups show no difference.
-%  %
-%  (b) Shape-based clustering for all subjects who suffered from back pain yields 6 groups. 
-%  %
-%  Note that the difference in subject count is due to the missing shape information for some subjects.
-%  %
-% % The total count of subjects in (b) is smaller 
-%  }
-%  \label{fig:hypopthesisbased}
-% \end{figure}
-% %
-% If the user proposes a hypothesis about a relation between a non-image feature regarding shape, the workflow slightly differs from the hypothesis-free analysis.
-% %
-% The starting point follows the \emph{feature investigation} pattern, where a feature of interest is selected by dragging it into the canvas area and viewing the subject's distribution as well as their shape differences.
-% %
-% In our use case, epidemiologists were interested in the dichotomous questionnaire answer to ``\emph{Did you experience back pain in the past three months?}''.
-% %
-% The mean shapes of the resulting visualization show no difference between the two groups (see Figure~\ref{fig:hypopthesisbased} (a)).
-% %
-% Either there are no differences or the variance information was lost in the mean shape calculation.
-% %
-% Since the focus is on subjects suffering from back pain, the clustering results of these subjects are then drawn into the canvas area, yielding six clusters as seen in Figure~\ref{fig:hypopthesisbased} (b).
-% %
-% Cluster 5 stood out for having a so-called \emph{hyperlordosis}, a strong curvature of the lumbar spine associated with back pain.
-% 
-% \emph{Multivariate analysis} using \emph{Cram\'{e}r's V} contingency values highlighted relationships of this cluster with joint degeneration, meat eating habits, preoccupation, back pain, neck or shoulder pain and waist circumference.
-% %
-% Since the prior selection only yields subjects that report back pain, the pain indicators specify the pain localization for the subjects.
-% 
-% It is well known that overweight is a risk factor for back pain.
-% %
-% While the BMI is a key figure for assessing height and weight of a subject, it does not tell anything about how the weight is distributed in the body.
-% %
-% Our epidemiological collaborators were interested in the correlation with waist circumference presented by this group.
-% %
-% Our finding follows the recent trends indicating that BMI is not a good measure for assessing the body shape, since healthy weight is dependent on many other measures \cite{Ahima2013}.
-% %
-% It indicates that the waist circumference rather than the BMI interacts with unusually shaped spines for subjects with lumbar back pain.
-% %
-% The influence of the parameter is now in the focus of further analysis.
-% 
-% \subsubsection{Follow-Up Tasks and Concluding Domain-Expert Feedback}
-% %
-% Correlation does not automatically imply causation. %\cite{Tufte2003}.
-% %
-% In particular, when large amounts of attributes are automatically checked for strong correlations, just by chance hundreds of them reach at least a moderately significant correlation, such as $p>0.05$.
-% %Defining a causal relationship solely based on observed correlations of two features is \emph{cum hoc ergo propter hoc}--correlation does not automatically imply causation \cite{Tufte2003}.
-% %
-% The observed correlations need to be carefully checked for confounders and medical soundness.
-% %
-% Statisticians validate causal inferences of the drawn conclusions.
-% 
-% Features that potentially interact with a disease-related condition need to be validated.
-% %
-% To increase the probability of the observation not to be random and to avoid \emph{overfitting}, they are cross-checked for associations in \texttt{SHIP-TREND} as a second, independent population sample, which was examined with methods identical to \texttt{SHIP}.
-% \\\\
-% The prototype was demonstrated to the domain experts (with background on epidemiology and radiology) via screen sharing.
-% %
-% The software was operated by a computer scientist following instructions of the domain experts.
-% %
-% The presented methods guide the attention to features that are not in the focus of attention and expectance of clinical researchers.
-% %
-% The explorative nature of the methods work well for gathering associations possibly acting as confounder, as outcome of a disease or as an actual cause or risk factor.
-% %"\emph{Correlation does not imply causation}" is a famous saying in statistics, meaning that correlation does not necessarily imply causal relationships \cite{}.
-% %
-% This distinction is hard to make and requires a lot of clinical experience.
-% %
-% The combination of multiple views with shape information helps to connect many different information sources to make the large information spaces cognitively feasible.
-% %
-% Displaying MRI scans for the subjects in the outlier cluster is promising because they are highly likely to exhibit pathologies.
-% %
-% %We plan to include a \texttt{DICOM}-viewer to meet this wish.
-% 
-% % - Describe steps from gathering Information from the raw image files (segmentation, abstraction, visualization)\\
-% % - Problem of sparse differences - Visualization has to be more abstract to emphasize the differences.\\
-% % - Input of Epidemiologists goes here!\\
-% ---------------------------------------------------------------------------------------------------------------------------------------
-% ---------------------------------------------------------------------------------------------------------------------------------------
-% ---------------------------------------------------------------------------------------------------------------------------------------
 

### 2014-06-21 draft.tex
Started Work on Related Work

@@
 \section{Prior and Related Work}
+\com{\textbf{Reviewer 1}: The related work section includes merely a set of related tools.
+I would have liked to see a more comprehensive section that discusses the previous related work in more detail and describes why they are important and relevant to this work and how your work differs or improves upon theirs.\\
+%
+Quality: The work contains many references.
+But there are potentially some missing; e.g, the framework talks about utilizing multiple coordinated views (although the details of what is actually linked is vague) and therefore cited papers such as Baldonado's work in AVI 2000, or Roberts 2007 on Coordinated multiple views, or Weaver's Cross-filtered Views for multidimensional visual analysis 2010 should be cited.
+The paper really proposes a framework; however it was unclear how this was implemented into the presented system.
+Therefore this should be improved.\\\\
+%
+\textbf{Reviewer 2}
+- I would suggest to also cite the original Generalized Pairs Plot and 
+not only the adaptation as GPLOMS [21]: J. W. Emerson and W. A. Green. 
+gpairs: The Generalized Pairs Plot, 2012. R package version 1.1.\\
+%
+- A related process model for Visual Analytics that focuses on hypotheses 
+and models: T. Lammarsch, W. Aigner, A. Bertone, S. Miksch, and A. Rind, 
+“Towards a Concept how the Structure of Time can Support the Visual 
+Analytics Process,” in Proceedings of International Workshop on Visual 
+Analytics (EuroVA 2011) in conjunction with EuroVis 2011, Goslar, 
+Germany, 2011, pp. 9–12.\\
+%
+- VISITORS is related work in visualization of medical variables of 
+cohorts: D. Klimov, Y. Shahar, and M. Taieb-Maimon, “Intelligent 
+visualization and exploration of time-oriented data of multiple 
+patients,” Artificial Intelligence in Medicine, vol. 49, no. 1, pp. 
+11–31, May 2010.\\
+%
+- another, forthcoming related paper: P. Angelelli, S. Oeltze, C. Turkay, 
+J. Haasz, E. Hodneland, A. Lundervold, A. Lundervold, B. Preim, and H. 
+Hauser, “Interactive Visual Analysis of Heterogeneous Cohort Study 
+Data,” IEEE Computer Graphics and Applications, vol. Early Access 
+Online, 2014.\\\\
+%
+\textbf{Reviewer 3}
+What the prior work section needs at the end is a paragraph stating how their work differs from the work they summarise i.e.
+it uses more non-image data than Turkay and uses different visualisations, and more importantly differentiation from Steenwijk et al’s very similar work (saying they use a different meaning for the word “cohort” does not seem justification enough, given the large overlap), only then can we be sure there is a novel contribution here.}
 %Designing a visualization, which conveys all data aspects equally, is challenging.

### 2014-06-22 draft.tex
Added revised figures and removed old 'add' entries

@@
 %
-\rem{Due to this random selection, the majority of the cohort does not suffer from a specific health problem.}
-\add{The subjects are selected without focus on a certain disease.}
+The subjects are selected without focus on a certain disease.
 %
@@
 \end{itemize}
-%
-\rem{The easy exchange of developed methods driven by modern web technologies intends to trigger a fast feedback loop between computer scientists and epidemiologists}.
 % http://wiki.infowiss.net/Gulf_of_execution ?
@@
 %
-\add{We implemented the presented methods using modern web technologies, such as \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}, making them easily accessible for the domain experts.
-%
-This aims to trigger a fast feedback loop between us and the epidemiologists.}
+We implemented the presented methods using modern web technologies, such as \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}, making them easily accessible for the domain experts.
+%
+This aims to trigger a fast feedback loop between us and the epidemiologists.
 
@@
 
-Indicators for medical conditions as well as questions about a subject's lifestyle are \rem{also} often \emph{dichotomous}--they have two manifestations (\emph{Yes} or \emph{No}).
+Indicators for medical conditions as well as questions about a subject's lifestyle are often \emph{dichotomous}--they have two manifestations (\emph{Yes} or \emph{No}).
 %
@@
 %The equipment used to gather medical image data is not updated or changed during a longitudinal study to ensure comparability in and between acquisition cycles.
-The equipment used to gather medical image data is \rem{kept, if possible, on the initial software and hardware version} \add{not updated or changed} during a longitudinal study to ensure comparability in and between acquisition cycles.
+The equipment used to gather medical image data is not updated or changed during a longitudinal study to ensure comparability in and between acquisition cycles.
 
@@
 %Their focus on \emph{hypothesis generation} using parallel assessment of multiple data features makes the work of Turkay and colleagues closest to ours albeit our emphasis on processing models derived from medical image data segmentation and variables with categorical manifestations \cite{Turkay2013}.
-\rem{Their} \add{Turkay and colleagues \cite{Turkay2013}} focus on \emph{hypothesis generation} using parallel assessment of multiple data features makes \rem{the work of Turkay and colleagues} \add{their work} closest to ours albeit our emphasis is on processing models derived from medical image data segmentation and variables with categorical manifestations \rem{\cite{Turkay2013}}.
+Turkay and colleagues \cite{Turkay2013} focus on \emph{hypothesis generation} using parallel assessment of multiple data features makes their work closest to ours albeit our emphasis is on processing models derived from medical image data segmentation and variables with categorical manifestations.
 %
@@
 \paragraph{Visualizing Image and Non-Image Data. }
-Gresh \rem{and colleagues} \add{et al. \cite{Gresh2000}} proposed \texttt{WEAVE}, one of the first systems which concurrently analyzed medical image and non-image data using linked views.
-%
-Blaas \rem{and colleagues} \add{et al. \cite{Blaas2007}} presented a similar system which analyzed medical image data and variables derived from them using views from the feature and physical space.
+Gresh et al. \cite{Gresh2000} proposed \texttt{WEAVE}, one of the first systems which concurrently analyzed medical image and non-image data using linked views.
+%
+Blaas et al. \cite{Blaas2007} presented a similar system which analyzed medical image data and variables derived from them using views from the feature and physical space.
 %
@@
 %
-Steenwijk \rem{and colleagues} \add{et al.
-\cite{Steenwijk2010}} employ a relational database to organize the data for visualizing subject data using linked views such as parallel coordinates, scatterplots and time plots.
-%
-Zhang \rem{and colleagues} \add{et al.
-\cite{Zhang2012}} provide a web-based system for analyzing subject groups with linked views and batch-processing capabilities for categorizing new subject entries into the data set.
+Steenwijk et al. \cite{Steenwijk2010} employ a relational database to organize the data for visualizing subject data using linked views such as parallel coordinates, scatterplots and time plots.
+%
+Zhang et al. \cite{Zhang2012} provide a web-based system for analyzing subject groups with linked views and batch-processing capabilities for categorizing new subject entries into the data set.
 %
@@
 %
-Dai \rem{and colleagues} \add{et al.
-\cite{Dai2005}} explored risk factors by incorporating choropleth maps of epidemiological features (e.g., mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines.
+Dai et al.
+\cite{Dai2005} explored risk factors by incorporating choropleth maps of epidemiological features (e.g., mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines.
 %
@@
 %
-Chui \rem{and colleagues} \add{et al. \cite{Chui2011}} visualized associations in time-dependent epidemiological data using time-series plots highlighting risk factor differences in age and gender.
+Chui et al. \cite{Chui2011} visualized associations in time-dependent epidemiological data using time-series plots highlighting risk factor differences in age and gender.
 
@@
 %
-Caban \rem{and colleagues} \add{et al.
-\cite{Caban2011}} investigated the suitability of variance visualizations of shape distribution models and concluded that users favor spherical glyph representations over deformation grids and likelihood volumes.
-%
-The distribution of shapes in a space derived from a PCA is plotted by Busking \rem{and colleagues} \add{et al. \cite{Busking2010a}} in a 2D-projected plane of the space.
+Caban et al.
+\cite{Caban2011} investigated the suitability of variance visualizations of shape distribution models and concluded that users favor spherical glyph representations over deformation grids and likelihood volumes.
+%
+The distribution of shapes in a space derived from a PCA is plotted by Busking et al. \cite{Busking2010a} in a 2D-projected plane of the space.
 %
@@
 %
-Hermann \rem{and colleagues} \add{et al. \cite{Hermann2014}} identify local deformation changes by investigating shape-related differences on rodent mandibles.
+Hermann et al. \cite{Hermann2014} identify local deformation changes by investigating shape-related differences on rodent mandibles.
 %
@@
  Dropping the \emph{gender} parameter on the already plotted \emph{body size} container creates a mosaic plot combining both features (right).
+ %
+ \com{3d-Koordinatensystem Kodierung durch Farbe (grün, rot, nuancen)}
+ %The lumbar spine models encode distance to the mean model using  \emph{color} 
  }
@@
 %
-The methods have to use domain language and reflect \add{the} routines \rem{of}\add{that} epidemiologists \add{take} in order to be incorporated into their research.
+The methods have to use domain language and reflect the routines that epidemiologists take in order to be incorporated into their research.
 %
@@
 %
-Following Tufte's concept of \emph{small multiples} \cite{Tufte1983}, \add{information derived from the} medical image data are directly incorporated into the plot by including color-coded mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
+Following Tufte's concept of \emph{small multiples} \cite{Tufte1983}, information derived from the medical image data are directly incorporated into the plot by including color-coded mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
 %
@@
 %
-Mapping the distance from a group mean shape to the global mean using color allows to assess local shape changes (Figure~\ref{fig:visualization} (e)).
+The distance from a group mean shape is mapped to the global mean using color allows to assess local shape changes (Figure~\ref{fig:visualization}).
 %
@@
  %
- Note the high amount of this subject group in the third group. The last group only contains four outlier-subjects.
+ Note the high amount of this subject group relative to the total count in the third group compared. The last group only contains four outlier-subjects.
  %

### 2014-06-22 draft.tex
Worked on Related Work section

@@
 it uses more non-image data than Turkay and uses different visualisations, and more importantly differentiation from Steenwijk et al’s very similar work (saying they use a different meaning for the word “cohort” does not seem justification enough, given the large overlap), only then can we be sure there is a novel contribution here.}
+
+\add{%Multiple coordinated views are appropriate for analyzing epidemiological data, because it suits Baldonado et al's \emph{Rule of Diversity} \cite{Baldonado2000}.
+%
+This section describes prior and related work and covers a broad range of visual analysis methods covering both image and non-image data.
+%
+We divide these section accordingly and also present lessons learned and from each publication.\\
+% --- PARAGRAPH --- % TODO Add Paragraph when removing ADD
+Baldonado et al. \cite{Baldonado2000} presented rules for designing multiple coordinated views.
+%
+They point out the cost-benefit-tradeoff introduced by cognitive overhead when connecting multiple views over more complex single views introduced by tasks, such as view comparison or context switching.
+%
+LESSON LEARNED HIERAUS??
+%
+Weaver et al extracted guidelines for cross-filtering multiple views by incorporating \emph{views} mapping data to visual elements, \emph{brushes} for selecting these elements and \emph{switches} linking brushing results between \emph{views}.
+%
+Our system follows the same rules for selecting subject groups, but our goal is to judge feature relations and potential outcomes.
+}
 %Designing a visualization, which conveys all data aspects equally, is challenging.
-%
+% --- PARAGRAPH ---
+
 Given the number of features of epidemiological data sets and their different manifestations, the strength of different visualization techniques needs to be combined \cite{Buja91, Konyha2009}.
@@
 %
-Our approach sticks more to the information extracted from the segmented image data and variable associations with non-image epidemiological factors.
+%Our approach sticks more to the information extracted from the segmented image data and variable associations with non-image epidemiological factors.
+\add{Our approach is to use information extracted from the segmented image data (such as 3D meshes) and variable associations with non-image epidemiological factors.}
 % Disadvantage - Nur 2 abgeleitete Variablen sichtbar, schlecht mit anderen Variablen in Verbindung zu bekommen (müsste jedoch auch gehen) - viele Dinge gehen durch die Abstraktion aber auch flöten!
 
-\paragraph{Visualizing Image and Non-Image Data. }
+\paragraph{Visualizing Image and Non-Image Data.}
+\add{One major focus is the connection of image and non-image data.}
+%TODO aus Steenwijk: Visual data analysis combining multi-dimensional single patient medical data with real-time linked scatter plots and parallel coordi- nates representations was first investigated in the WEAVE system of Gresh et al.
+[14].
+A number of aspects with regards to the specifi- cation of this kind of focus+context visualization were formalized by Doleisch et al.
+[8].
 Gresh et al. \cite{Gresh2000} proposed \texttt{WEAVE}, one of the first systems which concurrently analyzed medical image and non-image data using linked views.
 %
+%\add{The functionality is limited to project spatial related data to surface renderings of the image data and use By projecting surface models presented in 3D views to information visualizations, such as scatterplots, the authors were to support medical resea}
+%
 Blaas et al. \cite{Blaas2007} presented a similar system which analyzed medical image data and variables derived from them using views from the feature and physical space.
-%
-They incorporated data mining methods such as dividing the data space by using a k-nearest neighbor technique and the PCA.
+% From Blaas2007 - Tasks:
+%- Investigating combinations of features for finding spatial separations of point sets in physical space (segmentation)
+%- Classification of feature point sets (pattern classification)
+%- Finding clusters, patterns, and relations in the data
+\add{In both paper the data points plotted in information visualizations are part of the spatio-temporal domain and have a colocation in the image data, making it possible to plot them onto the image data representation to investigate interesting combinations.
+%
+Epidemiological features often don't have an exact spatial location in the image data.}
+%
+\rem{They incorporated data mining methods such as dividing the data space by using a k-nearest neighbor technique and the PCA.}
 %
@@
 \paragraph{Visualizing Heterogenous Non-Image Data.}
-Generalized Pairs Plots (\texttt{GPLOM´S}) are an information visualization technique comparing heterogenous features pairwise using a plot-matrix grouped by type \cite{Francois2013}.
+Generalized Pairs Plots (\texttt{GPLOM´S}) are an information visualization technique comparing heterogenous features pairwise using a plot-matrix grouped by type \cite{GPLOMS, Francois2013}.
 %
@@
 While we take similar steps when analyzing the data, such as employing statistical tests, our data are not associated to a geographical or temporal context, which affects the techniques described in the following section.
+\\\\
+\add{The uniqueness of the \emph{IVA} workflow applied to image-based cohort study data compared to the discussed work are twofold.
+%
+(1) We incorporate 3D models abstracting shape information (e.g.
+for the spine) fused with non-image data visualizations, allowing to analyze local physiological changes related to non-image parameters.
+%
+(2) Epidemiologists already have an well established workflow for statistically validating hypothesis. 
+%
+Therefore, we focus on hypothesis generation by discovering new relationships associated with shape information.
+%
+( (3) Following Weaver et al's data abstraction, we also want to create overview visualizations, that allow an unbiased feature relationship assessment.)
+}
 

### 2014-06-23 draft.tex
Nearly everything in the Related Work section

@@
 \section{Prior and Related Work}
-\com{\textbf{Reviewer 1}: The related work section includes merely a set of related tools.
+% \com{\textbf{Reviewer 1}: The related work section includes merely a set of related tools.
 I would have liked to see a more comprehensive section that discusses the previous related work in more detail and describes why they are important and relevant to this work and how your work differs or improves upon theirs.\\
-%
-Quality: The work contains many references.
+% %
+% Quality: The work contains many references.
 But there are potentially some missing; e.g, the framework talks about utilizing multiple coordinated views (although the details of what is actually linked is vague) and therefore cited papers such as Baldonado's work in AVI 2000, or Roberts 2007 on Coordinated multiple views, or Weaver's Cross-filtered Views for multidimensional visual analysis 2010 should be cited.
@@
 Therefore this should be improved.\\\\
-%
-\textbf{Reviewer 2}
-- I would suggest to also cite the original Generalized Pairs Plot and 
-not only the adaptation as GPLOMS [21]: J. W. Emerson and W. A. Green. 
-gpairs: The Generalized Pairs Plot, 2012. R package version 1.1.\\
-%
-- A related process model for Visual Analytics that focuses on hypotheses 
-and models: T. Lammarsch, W. Aigner, A. Bertone, S. Miksch, and A. Rind, 
-“Towards a Concept how the Structure of Time can Support the Visual 
-Analytics Process,” in Proceedings of International Workshop on Visual 
-Analytics (EuroVA 2011) in conjunction with EuroVis 2011, Goslar, 
-Germany, 2011, pp. 9–12.\\
-%
-- VISITORS is related work in visualization of medical variables of 
-cohorts: D. Klimov, Y. Shahar, and M. Taieb-Maimon, “Intelligent 
-visualization and exploration of time-oriented data of multiple 
-patients,” Artificial Intelligence in Medicine, vol. 49, no. 1, pp. 
-11–31, May 2010.\\
-%
-- another, forthcoming related paper: P. Angelelli, S. Oeltze, C. Turkay, 
-J. Haasz, E. Hodneland, A. Lundervold, A. Lundervold, B. Preim, and H. 
-Hauser, “Interactive Visual Analysis of Heterogeneous Cohort Study 
-Data,” IEEE Computer Graphics and Applications, vol. Early Access 
-Online, 2014.\\\\
-%
-\textbf{Reviewer 3}
-What the prior work section needs at the end is a paragraph stating how their work differs from the work they summarise i.e.
+% %
+% \textbf{Reviewer 2}
+% - I would suggest to also cite the original Generalized Pairs Plot and
+% not only the adaptation as GPLOMS [21]: J. W. Emerson and W. A. Green.
+% gpairs: The Generalized Pairs Plot, 2012. R package version 1.1.\\
+% %
+% - A related process model for Visual Analytics that focuses on hypotheses
+% and models: T. Lammarsch, W. Aigner, A. Bertone, S. Miksch, and A. Rind,
+% “Towards a Concept how the Structure of Time can Support the Visual
+% Analytics Process,” in Proceedings of International Workshop on Visual
+% Analytics (EuroVA 2011) in conjunction with EuroVis 2011, Goslar,
+% Germany, 2011, pp. 9–12.\\
+% %
+% - VISITORS is related work in visualization of medical variables of
+% cohorts: D. Klimov, Y. Shahar, and M. Taieb-Maimon, “Intelligent
+% visualization and exploration of time-oriented data of multiple
+% patients,” Artificial Intelligence in Medicine, vol. 49, no. 1, pp.
+% 11–31, May 2010.\\
+% %
+% - another, forthcoming related paper: P. Angelelli, S. Oeltze, C. Turkay,
+% J. Haasz, E. Hodneland, A. Lundervold, A. Lundervold, B. Preim, and H.
+% Hauser, “Interactive Visual Analysis of Heterogeneous Cohort Study
+% Data,” IEEE Computer Graphics and Applications, vol. Early Access
+% Online, 2014.\\\\
+% %
+% \textbf{Reviewer 3}
+% What the prior work section needs at the end is a paragraph stating how their work differs from the work they summarise i.e.
 it uses more non-image data than Turkay and uses different visualisations, and more importantly differentiation from Steenwijk et al’s very similar work (saying they use a different meaning for the word “cohort” does not seem justification enough, given the large overlap), only then can we be sure there is a novel contribution here.}
@@
 We divide these section accordingly and also present lessons learned and from each publication.\\
+% --- PARAGRAPH --- % TODO Add Paragraph when removing ADD
+Lammarsch et al. \cite{Lammarsch2011} create a terminological overview of a VA process and how it is connected.
+%
+They define a model as a representation of system entities, phenomena and processes.
+%
+Further, hypothesis are models, where the results are not compared with real-world data (\emph{validation}).
+%
+The VA process is basis is also reflected in our \emph{IVA}-Loop.\\
 % --- PARAGRAPH --- % TODO Add Paragraph when removing ADD
@@
 %
+\add{Information about medical image data is incorporated via mapper, which extract comparable metrics about the data.
+%
+Display of medical data is possible for single subjects, for example to look at outliers.
+%
+While we use a similar approach when analyzing non-image data, our analysis process also includes \emph{overview visualizations} and statistical suggestions of potentially interesting features.}
+
+\paragraph{Visualizing Heterogenous Non-Image Data.}
 Zhang et al. \cite{Zhang2012} provide a web-based system for analyzing subject groups with linked views and batch-processing capabilities for categorizing new subject entries into the data set.
@@
 Their understanding of a cohort differs from the understanding of the term in an epidemiological context by denoting every parameter-divided subject group as individual cohort.
-
-\paragraph{Visualizing Heterogenous Non-Image Data.}
+%
+\add{The paper also states little about kind of data used and how the system reacts to it, probably due to the short length.
+%
+We employ the idea of adding features via drag and drop into a working area.
+%
+Unfortunately, the author don't go into detail about how their algorithm of identifying similar subjects works or what and if they employ statistical measures.}
+
 Generalized Pairs Plots (\texttt{GPLOM´S}) are an information visualization technique comparing heterogenous features pairwise using a plot-matrix grouped by type \cite{GPLOMS, Francois2013}.
@@
 %
+\add{The resulting matrix provides a good \emph{overview visualization} but uses up a lot of screen space when applied to our data, which consists of 127 features.}
+%
 %Brushing, linking and filtering are feasible with \texttt{GPLOMS}, but have limitations such as making only one category brushable at a time.
@@
 %
-%The inspiration on the chosen visualization techniques stems from this publication.
+\add{The inspiration on the chosen visualization techniques stems from this publication.}
 %
@@
 %
+\add{While their goal to identify possible risk factors using socio-economic and health data is similar to ours, they focus on iteratively refining defined hypothesis and also focus on geographical data.
+%
+We employ the use of small multiples in this publication for incorporating heterogenous data types for comparability.}
+%
 Chui et al. \cite{Chui2011} visualized associations in time-dependent epidemiological data using time-series plots highlighting risk factor differences in age and gender.
+%
+\add{The work shows well, how different visualization modalities allow to get insight into these data sets, while focusing on the time-aspect, which is not present for our data.}
 
@@
 %
-%Applying our data sets to this technique yielded a cluttered shape space due to the many subjects.
-%
-%The data needs to abstracted or summarized in order to work in this context.
+\add{Applying our data sets to this technique yielded a cluttered shape space due to the many subjects.
+%
+The data needs to abstracted or summarized in order to work in this context.}
 %
@@
 %
-%This requires a geographic colocation of associated features.
+\add{This requires a geographic colocation of associated features.}
 %
@@
 % Prior Work
-\paragraph{SHIP Data Analysis.}
+\paragraph{Prior Work.}
 Klemm et al. \cite{Klemm2013VMV} visualized lumbar spine variabilities based on a semi-automatic shape detection algorithm of 490 participants of the \texttt{SHIP-2} cohort.
@@
 %
+\com{Hier noch einer Einschaetzung, oder reicht das schon?}
+%
 Niemann et al. \cite{Niemann2014} presented an interactive data mining tool for the assessment of risk factors of hepatic steatosis, the fatty liver disease.
@@
 Association rules created by data mining methods can be analyzed interactively with their tool and highlight potentially overlooked features.
+%
+\add{We combine the idea of suggesting potentially interesting features by adding them to the \emph{IVA}-workflow in order to amplify hypothesis generation.}
 % *********************** / Our own Work ***********************
@@
  \centering
- \includegraphics[width=3.0in]{figures/InteractionLoop}
+ \includegraphics[width=2.3in]{figures/InteractionLoop}
  \caption{Detailed \emph{IVA Loop} as extension from Figure~\ref{fig:WorkflowComparison}.

### 2014-06-23 draft.tex
reviewed related work section

@@
 %
-This section describes prior and related work and covers a broad range of visual analysis methods covering both image and non-image data.
-%
-We divide these section accordingly and also present lessons learned and from each publication.\\
-% --- PARAGRAPH --- % TODO Add Paragraph when removing ADD
-Lammarsch et al. \cite{Lammarsch2011} create a terminological overview of a VA process and how it is connected.
-%
-They define a model as a representation of system entities, phenomena and processes.
-%
-Further, hypothesis are models, where the results are not compared with real-world data (\emph{validation}).
-%
-The VA process is basis is also reflected in our \emph{IVA}-Loop.\\
-% --- PARAGRAPH --- % TODO Add Paragraph when removing ADD
-Baldonado et al. \cite{Baldonado2000} presented rules for designing multiple coordinated views.
-%
-They point out the cost-benefit-tradeoff introduced by cognitive overhead when connecting multiple views over more complex single views introduced by tasks, such as view comparison or context switching.
-%
-LESSON LEARNED HIERAUS??
-%
-Weaver et al extracted guidelines for cross-filtering multiple views by incorporating \emph{views} mapping data to visual elements, \emph{brushes} for selecting these elements and \emph{switches} linking brushing results between \emph{views}.
-%
-Our system follows the same rules for selecting subject groups, but our goal is to judge feature relations and potential outcomes.
+This section describes prior and related work and covers a broad range of visual analysis methods incorporating both image and non-image data.
+%
+%We divide these section accordingly and also present lessons learned and from each publication.\\
 }
@@
 The method is strongly dependent on the descriptive measures of the epidemiological factors.
-
+%
 %\emph{Over-fitting} of hypotheses to the data may be imposed  is a danger pointed out by the authors when hypotheses to the data based on observations of changes in these plots the measure highlights only subsets of statistical changes.
@@
 %Our approach sticks more to the information extracted from the segmented image data and variable associations with non-image epidemiological factors.
-\add{Our approach is to use information extracted from the segmented image data (such as 3D meshes) and variable associations with non-image epidemiological factors.}
+\add{Our approach uses information extracted from the segmented image data (such as 3D meshes) and variable associations with non-image epidemiological factors.}
 % Disadvantage - Nur 2 abgeleitete Variablen sichtbar, schlecht mit anderen Variablen in Verbindung zu bekommen (müsste jedoch auch gehen) - viele Dinge gehen durch die Abstraktion aber auch flöten!
@@
 \paragraph{Visualizing Image and Non-Image Data.}
-\add{One major focus is the connection of image and non-image data.}
+%\add{One major focus is the connection of image and non-image data.}
 %TODO aus Steenwijk: Visual data analysis combining multi-dimensional single patient medical data with real-time linked scatter plots and parallel coordi- nates representations was first investigated in the WEAVE system of Gresh et al.
@@
 %
-Blaas et al. \cite{Blaas2007} presented a similar system which analyzed medical image data and variables derived from them using views from the feature and physical space.
+Blaas et al. \cite{Blaas2007} presented a similar system, which also analyzed medical image data and variables derived from them using views from the feature and physical space.
 % From Blaas2007 - Tasks:
@@
 %- Finding clusters, patterns, and relations in the data
-\add{In both paper the data points plotted in information visualizations are part of the spatio-temporal domain and have a colocation in the image data, making it possible to plot them onto the image data representation to investigate interesting combinations.
+\add{The parameter in both paper are part of the spatio-temporal domain and have a colocation in the image data, making it possible to plot them onto the image data representation to investigate interesting combinations.
 %
@@
 %
-\add{Information about medical image data is incorporated via mapper, which extract comparable metrics about the data.
-%
-Display of medical data is possible for single subjects, for example to look at outliers.
-%
-While we use a similar approach when analyzing non-image data, our analysis process also includes \emph{overview visualizations} and statistical suggestions of potentially interesting features.}
+\add{Information about medical image data is incorporated via mappers, which extract comparable metrics about the data.
+%
+Medical image data can be displayed individually for subjects, e.g. for analyzing outliers.
+%Display of medical data is possible for single subjects, for example to look at outliers.
+%
+While we use a similar approach when analyzing non-image data, our process also includes \emph{overview visualizations} and statistical suggestions of potentially interesting features.}
 
@@
 %
-\add{The paper also states little about kind of data used and how the system reacts to it, probably due to the short length.
-%
-We employ the idea of adding features via drag and drop into a working area.
-%
-Unfortunately, the author don't go into detail about how their algorithm of identifying similar subjects works or what and if they employ statistical measures.}
+\add{Due to the short paper length, detail is missing on the data type and their algorithm of identifying similar subjects works or what and if they employ statistical measures.
+%
+We employ the idea of adding features via drag and drop into a canvas area.}
 
@@
 %
-\add{The resulting matrix provides a good \emph{overview visualization} but uses up a lot of screen space when applied to our data, which consists of 127 features.}
+\add{The resulting matrix provides a \emph{overview visualization}, but requires a lot of screen space for many features (127 in our application scenario).}
 %
@@
 %
-\add{The inspiration on the chosen visualization techniques stems from this publication.}
+\add{We incorporate the idea of adaptive type-dependent visualizations.}
+%\add{The inspiration on the chosen visualization techniques stems from this publication.}
 %
@@
 %
-\add{The work shows well, how different visualization modalities allow to get insight into these data sets, while focusing on the time-aspect, which is not present for our data.}
+\add{%The connection of time series plots with histograms and heat maps allows for fast insight into complex relationships and 
+While the works shows, how different visualization modalities provide insight into these data sets, it focuses on the time aspect, which is not present for our data.}
 
@@
 %--except the variables derived the segmentation model itself.
-While we take similar steps when analyzing the data, such as employing statistical tests, our data are not associated to a geographical or temporal context, which affects the techniques described in the following section.
+\rem{While we take similar steps when analyzing the data, such as employing statistical tests, our data are not associated to a geographical or temporal context, which affects the techniques described in the following section.}
+% --- PARAGRAPH --- % TODO Add Paragraph when removing ADD
+A workflow and definition of terminology of Visual Analysis (VA) techniques is provided by Lammarsch et al. \cite{Lammarsch2011}.
+%
+They define a model as a representation of system entities, phenomena and processes and hypothesis as models whose outcomes are not compared with real-world data (\emph{validation}).
+%
+The VA-process is also reflected in our \emph{IVA}-Loop.\\
+% --- PARAGRAPH --- % TODO Add Paragraph when removing ADD
+Baldonado et al. \cite{Baldonado2000} presented rules for designing multiple coordinated views.
+%
+They point out the cost-benefit-tradeoff introduced by the cognitive overhead by mentally connecting multiple views over more complex single views.
+%introduced by tasks, such as view comparison or context switching.
+%
+LESSON LEARNED HIERAUS??
+%
+Weaver et al.
+\cite{Weaver2010} extracted guidelines for cross-filtering multiple views by incorporating \emph{views} mapping data to visual elements, \emph{brushes} for selecting these elements and \emph{switches} for linking brushing results between \emph{views}.
+%
+Our system follows the same rules for selecting subject groups, but our goal is to judge feature relations and potential outcomes.
 \\\\
-\add{The uniqueness of the \emph{IVA} workflow applied to image-based cohort study data compared to the discussed work are twofold.
+\com{Dieser Abschnitt wurde hinzugefuegt, weil mehrere Reviewer meinten, dass der Related Work Teil zu wenig die Einzigartigkeit der Arbeit hervorbringt.}
+%Related work focused on epidemiological data often deals with time-dependent data, hypothesis generation by guidance through parameter suggestions is seldom.
+\add{The uniqueness of the \emph{IVA} workflow applied to image-based cohort study data compared to the discussed work are threefold.
 %
@@
 %
-( (3) Following Weaver et al's data abstraction, we also want to create overview visualizations, that allow an unbiased feature relationship assessment.)
+(3) Overview visualizations using statistical abstractions aim to provide an unbiased feature relationship assessment.
 }
@@
 %	\item No additional software needs to be installed, most people use a decent state-of-the-art web browser, even on mobile devices.
-	\item The client-server structure allows for employing heavy computation on a capable server machine and transfering results to the client.
-	\item Since image data for several thousand subjects requires hundreds of gigabytes disk space, it can remain safely on the server and elements can be transferred on demand.
-	%
-	High confidentiality standards of the data can be met by restricting access via an account system.
+	\item The client-server structure allows for employing heavy computation on a server machine and transferring results to the client.
+	\item \add{Disk-space demanding image data remains on the server and elements can be transferred on demand.}
+	%\item Since image data for thousands of subjects requires hundreds of gigabytes disk space, it can remain safely on the server and elements can be transferred on demand.
+	%
+	High confidentiality standards of the data are met by \add{password protecting the access.}%restricting access via an account system.
 	\item Recent developments in \emph{WebGL} applications running in browsers with near-native performance push the development further into to the web resulting in many open source libraries, which are well documented, rich in examples and driven by active communities.
@@
 	\item a detailed description of the feature, and
-	\item the data dictionary translating value or error IDs to the actual values.
+	\item the data dictionary translating value or error IDs to values.
 \end{itemize}
@@
 \\\\
-Since it is an entirely different approach to their usual workflows which are heavily influenced by statistical tools, it is to highlight that they had fun browsing through the data set in the presented manner and to dig out new information about the data.
+Since it is an entirely different approach to their usual workflows, which are heavily influenced by statistical tools, it is to highlight that they had fun browsing through the data set in the presented manner and to dig out new information about the data.
 %
@@
 %
-\\\\\
-A number of improvements is left open for future work, e.g.,
-
-\begin{itemize}
-%	\item dynamic discretization of continuous variables to better fit the data distributions
-	\item shape brushing methods to intuitively query subjects using image data,
-	\item the inclusion of more statistical methods and views that are familiar to the epidemiologists (odds ratios, box plots), or
-	%\item calculation of contingency differences for selected sub groups--highlighting how the feature correlation changes for a given group
-	%\item integration of further visualization techniques, which incorporate both image and non-image data
-	%\item support of control groups to check the resilience of observations \cite{Fletcher2012}.
-%	\item understand how the existing methods are applied to the data, iteratively adapt to the epidemiologiical workflow
-%	\item investigate which parts of the methods are applied by domain experts and what aspecunderstand how the existing methods are applied to the data, iteratively adapt to the epidemiologiical workflow
-	\item adapt the shape visualization to explore other organ data with different variance type (such as texture of liver or white/gray matter distribution in the brain).
-\end{itemize}
+\add{A number of improvements is left open for future work, such as shape brushing methods to intuitively query subjects using image data or the inclusion of more statistical methods and views that are familiar to the epidemiologists (odds ratios, box plots).}
+
+% \begin{itemize}
+% %	\item dynamic discretization of continuous variables to better fit the data distributions
+% 	\item shape brushing methods to intuitively query subjects using image data,
+% 	\item the inclusion of more statistical methods and views that are familiar to the epidemiologists (odds ratios, box plots), or
+% 	%\item calculation of contingency differences for selected sub groups--highlighting how the feature correlation changes for a given group
+% 	%\item integration of further visualization techniques, which incorporate both image and non-image data
+% 	%\item support of control groups to check the resilience of observations \cite{Fletcher2012}.
+% %	\item understand how the existing methods are applied to the data, iteratively adapt to the epidemiologiical workflow
+% %	\item investigate which parts of the methods are applied by domain experts and what aspecunderstand how the existing methods are applied to the data, iteratively adapt to the epidemiologiical workflow
+% 	\item adapt the shape visualization to explore other organ data with different variance type (such as texture of liver or white/gray matter distribution in the brain).
+% \end{itemize}
 % These benefits 
@@
 %% if specified like this the section will be committed in review mode
+\begin{small}
 \acknowledgments{SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no.
@@
 This work was supported by the DFG Priority Program 1335: Scalable Visual Analytics.}
+\end{small}
 \clearpage

### 2014-06-23 draft.tex
Reworked Figures

@@
 %(The problem).
-\com{ToDo: Reviewer-Feedback: The abstract should be formulated more concisely and focus on What? 
-Why? How? What follows?}
-Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease-specific risk factors.
-%
-Cohort studies comprise of heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors and more recently, medical image data.
 %
@@
 %\\\\
-%(New Solution).
-We propose an Interactive Visual Analysis (\emph{IVA}) approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic features and attributes derived from the image data.
-%
-This is achieved by combining brushing and linking enabled coordinated information visualization views and interactive 3D shape renderings with epidemiological data representations such as pivot tables and key figures as association measures.
-%(Validation).
-The presented concepts are applied by experts to gather and evaluate hypotheses about the aging process of the lumbar spine.
-%
-%(Results). 
-It allows a more flexible comparison between image and non-image data.
-%
-%(Implications).
-The new framework enables hypotheses validation and generation by incorporating human pattern recognition as well as data mining methods.
-%
-Using all reliable information from the image segmentation linked to non-image features aims to unveil \emph{associations} by applying an iterative analysis approach.
+% ----------- OLD ABSTRACT ---------
+% Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease-specific risk factors.
+% %
+% Cohort studies comprise of heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors and more recently, medical image data.
+% %(New Solution).
+% We propose an Interactive Visual Analysis (\emph{IVA}) approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic features and attributes derived from the image data.
+% %
+% This is achieved by combining brushing and linking enabled coordinated information visualization views and interactive 3D shape renderings with epidemiological data representations such as pivot tables and key figures as association measures.
+% %(Validation).
+% The presented concepts are applied by experts to gather and evaluate hypotheses about the aging process of the lumbar spine.
+% %
+% %(Results).
+% It allows a more flexible comparison between image and non-image data.
+% %
+% %(Implications).
+% The new framework enables hypotheses validation and generation by incorporating human pattern recognition as well as data mining methods.
+% %
+% Using all reliable information from the image segmentation linked to non-image features aims to unveil \emph{associations} by applying an iterative analysis approach.
+% ----------- /OLD ABSTRACT ---------
+\add{
+Large-scale longitudinal epidemiological studies investigate groups of people with common characteristics or experiences (a cohort) to characterize health and identify disease-specific risk factors.
+%
+Cohort study data comprise thousands of individuals and hundreds of heterogenuous variables, describing the medical condition as well as demographic and lifestyle factors, and more recently, medical image data.
+%
+In the standard epidemiological workflow, a hypothesis is given and validated by analyzing a subset of possibly relevant variables over manually defined subgroups of the cohort, e.g., based on age or body weight.
+%
+A potential impact of the neglected variables remains unnoticed, data-driven group definition is not supported, and capabilities for integrating the image data into the analysis are missing.\\
+%
+We propose an Interactive Visual Analysis (IVA) approach that enables epidemiologists to rapidly investigate the entire data pool for both, hypothesis validation and generation.
+%
+Furthermore, we suggest a strategy for incorporating the image data which involves shape-based object detection and the derivation of attributes describing the object shape.
+%
+The concurrent investigation of image-based and non-image data is realized in a web-based multiple coordinated view system, comprising standard views from information visualization and epidemiological data representations such as pivot tables.
+%
+To account for the numerous categorical variables, a mosaic plot is integrated.
+%
+The views are equipped with brushing facilities and augmented by 3D shape renderings of the segmented objects, e.g., each bar in a histogram is overlayed with a mean shape of the associated subgroup of the cohort.
+%
+We integrate clustering of variables and object shape for data-driven subgroup definition and statistical key figures for measuring the association between variables.
+%
+We demonstrate the IVA approach by validating and generating hypotheses related to lower back pain and the aging process of the lumbar spine.}
 } % end of abstract

### 2014-06-24 draft.tex
fixed error in visualization figure and other stuff

@@
  \includegraphics[width=1\textwidth, resolution=300]{figures/visualization}
- \caption{\com{Add: Explanation of the color coding, ToDo: Selektion in Abbildung 3 einbauen und daran Selektion erklären}
- %
- (Left) Screenshot from the front-end, which is divided as follows: (a) The sidebar containing all features as well as the groups defined in the analysis process; (b) the canvas area where features can be added via drag and drop and the visualization is chosen automatically according to the data type; (c) the interactive pivot table showing the exact numbers for each displayed feature combination.
+ \caption{
+ %
+ (Left) Screenshot from the front-end, which is divided as follows: (a) The sidebar containing all features as well as the groups defined in the analysis process; (b) the canvas area where features can be added via drag and drop and the visualization is chosen automatically according to the data type; (c) the interactive pivot table showing the exact numbers for each displayed feature combination; \add{(d) Buttons to open panes containing the contingency matrix, contingency pane and pivot table}.
  %
@@
  %
- \com{3d-Koordinatensystem Kodierung durch Farbe (grün, rot, nuancen)}
+ \add{In a prior step, the user selected all subjects with diagnosed thyroid disorder.
+ %
+ These subjects are shown as shade in the visualization, denoting their share.
+ %Their shares are denoted in the bar chart as blue and the mosaic plot as dark shade.
+ %
+ Subjects between 153.5--170~cm body size are more affected by thyroid disorder (box plot) and are mostly female (mosaic plot).}
+ %
+ \add{Distance to the mean mesh of subjects with thyroid disorder is encoded as red for x axis, blue to y axis and green to z axis.}
  %The lumbar spine models encode distance to the mean model using  \emph{color} 

### 2014-06-24 draft.tex
added Bernhards review

@@
 %% indicate IEEE Member or Student Member in form indicated below
-\author{P. Klemm$^{1}$, S. Oeltze$^{1}$, K. Lawonn$^{1}$, K. Hegenscheid$^{2}$, H. V{\"o}lzke$^{2}$, B. Preim$^{1}$}
+\author{Paul Klemm$^{1}$, Steffen Oeltze-Jafra$^{1}$, Kai Lawonn$^{1}$, Katrin Hegenscheid$^{2}$, Henry V{\"o}lzke$^{2}$, Bernhard Preim$^{1}$}
 \authorfooter{
@@
 % ----------- /OLD ABSTRACT ---------
+\com{Epidemiological Features Begriff einfuehren}
 \add{
@@
 %
-Also, only a small subset of variables can be concurrently analyzed, depending on the statistical measure.
+Also, only a small subset of variables can be concurrently analyzed. %depending on the statistical measure.
 \\\\
@@
 %
-We implemented the presented methods using modern web technologies, such as \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}, making them easily accessible for the domain experts.
+We implemented the presented methods using modern web technologies, such as \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}, making them easily accessible for the domain experts to enable a fast feedback loop.
 %
@@
  \includegraphics[width=3.0in]{figures/workflow_comparison}
- \caption{\emph{IVA} tools complement parts of the epidemiological workflow instead of replacing them.
- %
- The appropriate combination of statistical and interactive-driven analysis shows promising potential to unveil the information in the data.
- %
- The standard epidemiology workflow consists of four steps (a).
- %
- IVA complements this workflow and is used to elaborate the feature selection (b).
- %
- The iterative red highlighted part is called the \emph{IVA Loop} and is described in more detail in Figure~\ref{fig:InteractionLoop}.
+ \caption{
+ (a) The standard epidemiology workflow consists of four steps.
+ %
+ (b) \emph{IVA} tools complement parts of this workflow instead of replacing them.
+ %
+ The combination of statistical and interactive-driven analysis shows promising potential to unveil information in the data.
+ %
+ %IVA complements this workflow and is used to elaborate the feature selection.
+ %
+ We call the iterative red highlighted part \emph{IVA Loop}, described in detail in Figure~\ref{fig:InteractionLoop}.
  %(a) shows the standard epidemiological workflow, (b) the \emph{IVA} supported one.
@@
 
-Reproducibility of results is an epidemiological key requirement.
+\emph{Reproducibility} of results is an epidemiological key requirement.
 %
@@
 
-In longitudinal cohort studies, grouping subjects using epidemiological features is essential in order to allow per-group risk determination.
+Grouping subjects using epidemiological features is essential in cohort studies to allow per-group risk determination.
 %
@@
 %
-Age for example is divided into groups (e.g. in 20 year steps) when investigating its influence on a condition.
+Age for example is divided into groups (e.g. in 20 year steps) when investigating its influence.
 %
@@
 %
-A famous example for a confounder is the association between shoe size and mortality, where it can be observed that people with larger shoe size have a smaller life expectation than people with small shoe size.
-%
-The shoe size is actually associated with gender, where women have smaller feet than men but also a longer life expectation.
+A famous example for a confounder is the association between shoe size and mortality, where it can be observed that people with larger shoe size have a smaller life expectation.
+%than people with small shoe size.
+%
+The shoe size is actually associated with gender, where women have smaller feet and a longer life expectation.
 %
@@
 %An example is the higher mortality rate of non-insulin dependent diabetes opposed to insulin dependent diabetes.
-A famous example is the Berkeley University gender bias case, where there were significantly less women admitted then man.
+A famous example is the Berkeley University gender bias case, where significantly less women were admitted.
 %
@@
 %
-Large studies, such as the SHIP require written proposals reasoning about each requested features and the expected outcome.
+Large studies, such as the SHIP, require written proposals reasoning about each requested features and the expected outcome.
 %
@@
 %
-Due to the high sensibility of the data, all identity revealing information of a person, such as name or place of residence is anonymized.
+Due to the high sensibility of the data, all identity-revealing information of a person, such as name or place of residence is anonymized.
 }
@@
 
-Epidemiological data are highly heterogenous and incomplete.
+Epidemiological data are strongly heterogenous and incomplete.
 %
@@
 %
-For ethical or medical reasons some features cannot be gathered for each subject.
-%
-The most obvious example are women-specific questions about menstrual status or number of born children.
-%
-Follow-up examinations or questions about conditions like medications taken after a diagnosed disease also yield features only available for a small amount of subjects.
+For ethical or medical reasons some features cannot be gathered for each subject, e.g. women-specific questions about menstrual status or number of born children.
+%
+Follow-up examinations or questions about conditions such as medications taken after a diagnosed disease also yield features only available for a small amount of subjects.
 %
@@
 %
-Medical examinations mostly comprise categorical (e.g. levels of back pain) and continuous values (e.g. age or body size).
+Medical examinations comprise categorical (e.g. levels of back pain) and continuous values (e.g. age or body size).
 %
@@
 %
-Sparse populated features are also hard to assess statistically as to few data samples have a higher chance to distort the real underlying distributions.
+Sparse populated features are also hard to assess statistically as too few data samples have a higher chance to distort the real underlying distributions.
 %Also, correlation can both be of epidemiological interest but 
@@
 
-\subsubsection{Image acquisition} \label{ImageAcquisition} Imaging techniques involving ionizing radiation for the subject are not suitable for ethical reasons.
+\paragraph{Image Acquisition.} Imaging techniques involving ionizing radiation for the subject are not suitable for ethical reasons.
 %
@@
 
-\subsubsection{Image analysis} \label{ImageAnalysis}
+\paragraph{Image Analysis.}
 \com{Take up huge amount of space!}
@@
 %
-Since reliable and efficient segmentation techniques are not available in general, epidemiologists are forced to measure the data by hand, which is a very tedious work with respect to the number of necessary landmarks and the number of subjects.
+Since reliable and efficient segmentation techniques are not available in general, epidemiologists are forced to measure the data by hand.
+%, which is a very tedious work with respect to the number of necessary landmarks and the number of subjects.
 %
@@
 %
-They are also prone to a high inter-observer variability and are hard to reproduce.
-%
-This gains even more importance when analyzing multiple time steps.
+They are also prone to a high inter-observer variability.
+%
+%This gains even more importance when analyzing multiple time steps.
 %
@@
 %
-The MRI protocol features a rich number of sequences.
+%The MRI protocol features many of sequences.
 %
@@
 %
-This section describes prior and related work and covers a broad range of visual analysis methods incorporating both image and non-image data.
+This section describes prior and related work and covers visual analysis methods incorporating both image and non-image data.
 %
@@
 
-\paragraph{Interactive Visual Analysis}
+\paragraph{Interactive Visual Analysis.}
 % Interactive Visual Analysis - Steffens Paper
@@
 \section{Image Centric Cohort Study Data in Interactive Visual Analysis Context} \label{Image Centric Cohort Study Data in Interactive Visual Analysis Context}
-We described the epidemiological workflow followed by domain experts where reproducibility and statistical integrity are important aspects (recall Subsection~\ref{EpidemiologicalWorkflow}).
-%
-Figure~\ref{fig:WorkflowComparison} (a) shows this workflow as a consecutive series of steps.
+We described the epidemiological workflow and emphasized the reproducibility and statistical integrity (recall Subsection~\ref{EpidemiologicalWorkflow}).
+%
+%Figure~\ref{fig:WorkflowComparison} (a) shows this workflow as a consecutive series of steps.
 %
 %Introducing the \emph{IVA} principle to the epidemiological domain aims to compensate its weaknesses rather than replacing the existing workflow.
-Introducing the \emph{IVA} principle to the epidemiological domain aims to compensate the weaknesses of the existing workflow, rather than replacing it.
+Introducing the \emph{IVA} principle to the epidemiological domain aims to compensate the weaknesses of the existing workflow, rather than replacing it (recall Fig.~\ref{fig:WorkflowComparison}).
 %
@@
 %
-Statistical tests on features associated to a hypothesis yield a value for deciding whether the data supports the hypothesis or not.
-%
-Features not included in the analysis may potentially support the chosen hypothesis by discriminating the population in the expected way, but are not highlighted in any way.
+Statistical tests on features associated to a hypothesis yield a value for deciding whether the data supports the hypothesis.
+%
+Features not included in the analysis may potentially support the chosen hypothesis by discriminating the population in the expected way, but are not highlighted.
 %
@@
 %
-It is easily possible that omitted information would heavily influence the result.
+%It is easily possible that omitted information would heavily influence the result.
 %
@@
 
-\paragraph{Local Investigation:} refers to the inspection of dependent variables with respect to certain subsets of the image or shape space.
-%
-For instance, the epidemiologist selects several lumbar spines with a common characteristic in the image or shape space and wishes to inspect the associated dependent variables in an attribute view \cite{Hermann2014}.
+\paragraph{Local Investigation.} Refers to the inspection of dependent variables with respect to certain subsets of the image or shape space.
+%
+For instance, the epidemiologist selects several lumbar spines with a common characteristic in the image or shape space and inspects the associated dependent variables in an attribute view \cite{Hermann2014}.
 %
@@
 
-\paragraph{Feature Localization:} refers to the search for structures in the image or shape space with a defined characteristic.
+\paragraph{Feature Localization.} Refers to the search for structures in the image or shape space with a defined characteristic.
 %
@@
 
-\paragraph{Multivariate Analysis:} refers to an investigation of multi-variate properties of the dependent data by specifying a feature in one attribute view and at the same time analyzing the value distribution with respect to other variables in other attribute views.
+\paragraph{Multivariate Analysis.} Refers to an investigation of multi-variate properties of the dependent data by specifying a feature in one attribute view and at the same time analyzing the value distribution with respect to other variables in other attribute views.
 %
@@
 %
-Data obtained using questionnaires or medical tests are often stored using statistical packages such as \texttt{SPSS}, which have a proprietary data format with limited export capabilities.
+Data obtained using questionnaires or medical tests are often stored using statistical packages such as \texttt{SPSS}, which have a proprietary data format. %with limited export capabilities.
 %
@@
 %
-This is either done manually by experts setting landmarks (sometimes supported by algorithms connecting the landmarks such as graph cuts \cite{GraphCut}) or by a (semi-)automatic detection, registration and segmentation.
-%
-These algorithms have to deal with a large inter-subject variability of the anatomical structure and need to create reproducible results \cite{Preim2014}.
-%
-In principle, model-based approaches are effective for detection \cite{Rak2013} and segmentation \cite{Gloger2010, Gloger2012}.
+This is either done manually by experts setting landmarks or by a (semi-)automatic detection, registration and segmentation.
+%
+These algorithms have to deal with a large inter-subject variability of the anatomical structure \cite{Preim2014}.
+%
+In principle, model-based approaches are effective for detection \cite{Rak2013} and segmentation \cite{Gloger2010}.%\cite{Gloger2010, Gloger2012}.
 %
@@
 %
-Grey value comparison is usually used to measure the quantity of fat, water, and--application-specific--the iron content (liver) or the distribution of grey and white brain tissue.
+Grey value comparison is used to measure the quantity of fat, water, and--application-specific--the iron content (liver) or the distribution of grey and white brain tissue.
 
@@
 %
-Brushing methods are subdivided using the previously described \emph{IVA} patterns.
+%Brushing methods are subdivided using the previously described \emph{IVA} patterns.
 %
@@
 % \subsection{Interaction and Visualization Techniques} \label{Interaction- and Visualization Techniques}
-\section{System Design and Visualization Techniques} \label{Interaction- and Visualization Techniques}
-%
-The suitability of an interaction and visualization technique for epidemiological data depends on its ability to intuitively compare multiple data features at once while highlighting new associations.
-%
-The methods have to use domain language and reflect the routines that epidemiologists take in order to be incorporated into their research.
+\section{System Design and Implementation} \label{Interaction- and Visualization Techniques}
+%
+The suitability of an interaction and visualization technique for epidemiological data depends on its ability to compare multiple data features at once while highlighting new associations.
+%
+The methods have to reflect the routines that epidemiologists take in order to be incorporated into their research.
 %
@@
  %Mouse-over focuses a entry and highlights the feature names for better readability.
- Mouse-over an entry highlights the feature names for better readability.
+ Mouse-over an entry enlarges the feature names for better readability.
  %
@@
 % \subsection{System Structure and Design} \label{Structure and Workflow}
-\subsection{ToDo Rename.System Structure and Design} \label{Structure and Workflow}
+\subsection{Design and Visualization Techniques} \label{Structure and Workflow}
 % \begin{itemize}
@@
 % \end{itemize}
-Early it became clear that we have to rely a lot on online communication with our clinical partners due to their busy schedule and our large spatial distance towards each other.
+%Early it became clear that we have to rely a lot on online communication with our clinical partners due to their busy schedule and our large spatial distance towards each other.
+\add{Early it became clear that we have to rely a lot on online communication due to the busy schedule and our large spatial distance towards each other.}
 %
@@
 %
-Classical UI elements, such as the menu bar or custom right click menus are technically possible, but not common in this domain.
+Classical UI elements, such as the menu bar or custom right click menus, are technically possible, but not common in this domain.
 %
@@
 %
-Since the previously described \emph{IVA} workflow allows for many different ways to analyze the data, we tried to make the interface as minimalistic and discrete as possible, treating the resulting space as \emph{canvas} for the data.
+Since the previously described \emph{IVA} workflow allows for many different ways to analyze the data, we tried to make the interface as minimalistic as possible, treating the resulting space as \emph{canvas} for the data.
 %
@@
 
-\paragraph{System Layout}
+\paragraph{System Layout.}
 We experimented with several layouts.
@@
 %
-A well working configuration was the introduction of separate panes for the adjacency matrix, pivot table and sidebar, displayed with a mouse click on the corresponding button and sliding on top of the \emph{canvas} (Figure~\ref{fig:visualization} shows the system with reeled-out pivot table pane).
+The introduction of separate panes for the adjacency matrix, pivot table and sidebar, displayed with a mouse click on the corresponding button and sliding on top of the \emph{canvas} was considered more feasible (Figure~\ref{fig:visualization} shows the system with reeled-out pivot table pane).
 %
@@
 \paragraph{Sidebar.}
-The \emph{sidebar} is the only visible element when the system is opened.
+The \emph{sidebar} is the only visible element when the system is opened (Fig.~\ref{fig:application} (a)).
 %
@@
 \paragraph{Adaptive Feature Visualization.} \label{sec:AdaptiveFeatureVisualization}
-The visualization type is, inspired by \texttt{GPLOMS} \cite{Francois2013}, dynamically chosen based on the feature types and number.
+The visualization type, inspired by \texttt{GPLOMS} \cite{GPLOMS, Francois2013}, is dynamically chosen based on the feature types and number to allow for \emph{multivariate analysis}.
 %
@@
 %
-The distance from a group mean shape is mapped to the global mean using color allows to assess local shape changes (Figure~\ref{fig:visualization}).
+The distance from a group mean shape is mapped to the global mean using color.
+%
+This allows to assess local shape changes (Fig.~\ref{fig:visualization}).
 %
@@
 %
-If a feature is dropped on an existing plot, the visualization changes dynamically to make them comparable (Figure~\ref{fig:visualization} (right)).
+If a feature is dropped on an existing plot, the visualization changes dynamically to make them comparable (Fig.~\ref{fig:visualization} (right)).
 % --- PARAGRAPH ---
 
-Subject groups can be brushed via a double-click on it's representative in the visualizations.
-%
-Holding down the shift key allows to select multiple manifestations at once.
+To support \emph{feature localization}, subject groups can be brushed via a double-click on its representative in the visualizations.
+%
+Holding down the shift key allows to select multiple manifestations.
 %
@@
 %This allows to highlight distances between subjects
-The share of subjects of this sub group is linked to all other views (Figure~\ref{fig:visualization} (left)).
-%
-If the user for example selects all female subjects in a visualization of gender distribution, all other displayed meshes are color coded with their distance to the female mean and the share of female subjects is highlighted in the information visualization.
+The share of subjects of this subgroup is linked to all other views (Fig.~\ref{fig:visualization} (left)).
+%
+If the user selects all female subjects in a visualization of gender distribution, all other displayed meshes are color-coded with their distance to the female mean and the share of female subjects is highlighted in the information visualization.
 %Elements can be brushed using 
@@
 \paragraph{Pivot Tables.}
-Pivot tables are frequently used to present the data in epidemiological publications.
-%
-Epidemiologists are used to analyze groups based on table representations.
+%Pivot tables are frequently used to present the data in epidemiological publications.
+%
+Epidemiologists are used to perform \emph{multivariate analysis} of groups based on table representations.
 %
@@
 %
-However, they quickly get confusing and cluttered when they are divided into many subgroups.
+However, they quickly get confusing when they are divided into many subgroups.
 %
@@
 \paragraph{Automated Feature Suggestion using a Contingency Matrix.}
-As previously discussed, highlighting potentially interesting values in the data set is one major benefit of the \emph{IVA}-powered approach and belongs to the \emph{multivariate analysis} pattern,  analyzing features outside the shape space.
-%
-Turkay et al. \cite{Turkay2013} used the approach to calculate various key figures based on the distribution functions of each feature derived from the image data.
+As previously discussed, highlighting potentially interesting values in the data set is one major benefits of the \emph{IVA}-powered approach and belongs to the \emph{multivariate analysis} pattern,  analyzing features outside the shape space.
+%
+Turkay et al. \cite{Turkay2013} used the approach to calculate key figures based on the distribution functions of each feature derived from the image data.
 %
@@
 %
-$V$ assumes values between $0$, meaning that two variables are completely independent, and $1$ indicating they are the same.
-%
-\emph{Cram\'{e}r's V} is always positive and does therefore not allow statements about dependency direction.
+$V$ yields values between $0$, meaning that two variables are completely independent, and $1$ indicating they are the same.
+%
+Since \emph{Cram\'{e}r's V} is always positive, it does not allow to infer about dependency direction.
 
@@
 %
-They cannot be included into the epidemiological analysis, since statistical validation needs a minimum count to be seen as valid.
+They cannot be included into the epidemiological analysis, since statistical validation needs a minimum count to be valid.
 %
@@
 
-\paragraph{Contingency Pane}
+\paragraph{Contingency Pane.}
 Dropping a feature into the canvas area adds an entry for each manifestation of it to the \emph{contingency matrix}.
 %
-Testing sessions revealed that it was very tedious to open the matrix every time a new feature is added. 
+Testing sessions revealed that it was tedious to open the matrix every time a new feature is added. 
 %A simple list of features, which correlate with the last added feature.
 %
-The result was the \emph{contingency pane}, a table containing correlating features for the last added visualizations in descending order of the \emph{Cram\'{e}r's V} value.
+As a consequence, the \emph{contingency pane}, a table containing correlating features for the last added visualizations in descending order of the \emph{Cram\'{e}r's V} value.
 %
@@
 %
-\paragraph{Initialization and Clustering}
-Using feature suggestion allows to initialize the system with a set of potentially interesting. visualizations.
+\paragraph{Initialization and Clustering.}
+Using feature suggestion allows to initialize the system with a set of potentially interesting visualizations.
 %
@@
 %
-Also, we observed that the variables of interest is dependent on the specialization of the domain expert (explained in detail in Section~\ref{application}).
+Also, we observed that the variables of interest are dependent on the specialization of the domain expert (explained in detail in Section~\ref{application}).
 % --- PARAGRAPH ---
 
-Subject clustering is triggered automatically for each variable after it was added to the canvas by the user.
+Subject clustering is triggered automatically as \emph{local investigation} for each feature after it was added to the canvas by the user.
 %
@@
 \end{figure}
-In this section we discuss how we implemented the presented methods using free, open web standards.
+In this section we discuss how we implemented the presented methods using open web standards.
 %
@@
 	High confidentiality standards of the data are met by \add{password protecting the access.}%restricting access via an account system.
-	\item Recent developments in \emph{WebGL} applications running in browsers with near-native performance push the development further into to the web resulting in many open source libraries, which are well documented, rich in examples and driven by active communities.
+	\item Recent developments in \emph{WebGL} applications running in browsers with near-native performance result in many open source libraries, which are well documented and driven by active communities.
 We use \emph{WebGL} for rendering shape information.
@@
 %
-Due to its event-driven non-blocking I/O model it is fast and does not freeze in case of heavy workload like mesh calculation.
+Due to its event-driven non-blocking I/O model it is fast and does not freeze in case of heavy workload such as mesh processing.
 \\\\
@@
 %
-Image data are available as raw DICOM files as well as meshes representing segmentation masks of anatomical structures, which can be used to compare subjects.
+Image data are available as raw DICOM files.
+%
+Segmentation masks of anatomical structures are represented as meshes, suited for comparing subjects.
 %
@@
 %
-We accomplished this by converting it to a parameterized standalone console application that is spawned by \texttt{NodeJS} on client request and then reads the result from the console standard out and returns it in a proper format to the client.
+We accomplished this by converting it to a parameterized standalone console application that is spawned by \texttt{NodeJS} on client request and then reads the result from the console standard out and returns it to the client.
 %
@@
 %
-We conducted a qualitative case study with two domain experts on a data set compiled to analyze lower back pain. 
+We conducted a qualitative evaluation with two domain experts on a data set compiled to analyze lower back pain. 
 %
@@
 \begin{itemize}
-	\item a better understanding of effects of preventive measures such as occupational health and safety regulations,
+	\item a better understanding of effects of preventive measures, such as occupational health and safety regulations,
 	\item prognostic features for diagnosis and treatment of lumbar back pain, and
@@
 \subsection{The Lumbar Spine Data Set}
-We divide the data set into image and non-image data.
+%We divide the data set into image and non-image data.
 %
@@
 %
-Other features, such as pain indicators or lifestyle indicators such as physical activity are more subjective and less reliable.
+Other features, such as pain indicators or lifestyle indicators (e.g. physical activity) are more subjective and less reliable.
 %
@@
 
-The MRI image data was acquired for each subject on a 1.5~Tesla scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany) by four trained technicians in a standardized way.
-%
-\add{The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence (676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
+The MRI data was acquired for each subject on a 1.5~Tesla scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany) by four trained technicians in a standardized way.
+%
+\add{The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013.} %(676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
 }
@@
 \add{
-Since it lacks export methods for data dictionaries, we used \texttt{SPSS} to export our data to the 'SAS v9+' format, which writes out the data labels and saved the data values as non-labeled CSV.
+Since it lacks export methods for data dictionaries, we used \texttt{SPSS} to export our data to the 'SAS v9+' format, which saves the data labels, and exported the data values as non-labeled CSV.
 %
@@
 %
-In consultation with the collaborating epidemiologists, we set the number of groups to five, to allow for contingency assessment.
+We set the number of groups to five, to allow for contingency assessment.
 
@@
 %
-For clustering purposes, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature \cite{Klemm2013VMV}).
+For clustering, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature \cite{Klemm2013VMV}).
 
@@
 %
-The tetrahedron-based detection model described in Section~\ref{application:Data Preprocessing} consists of corresponding grid points for each structure instance.
-%
-This allows for calculation of shape distance and similarity.
+The tetrahedron-based detection model consists of corresponding grid points for each structure instance.
+%
+This allows to calculate shape distance and similarity.
 %
 This information is used to calculate mean shapes as described in Section~\ref{Interaction- and Visualization Techniques}.
-
-Shape distance is mapped onto color.
-%
-For dichotomous variables, the color codes distances between mean shapes of the two groups, for variables with more than two manifestations it encodes the distance to the global mean shape of all subjects (see Figure~\ref{fig:hypothesisfree}).
-\\\\
+%
+Shape distance between to mean meshes is mapped to color (recall Fig.~\ref{fig:visualization}).
+%
+%For dichotomous variables, the color codes distances between mean shapes of the two groups, for variables with more than two manifestations it encodes the distance to the global mean shape of all subjects).
+%
+
 Shape-based clustering is carried out via agglomerative hierarchical clustering of the spine canal centerlines (recall Section~\ref{application:Data Preprocessing} and \cite{Klemm2013VMV}).
@@
 Since it is hard to quantify these outcomes, Lam et al.
-suggest case studies for the \emph{VDAR} analysis by applying the think aloud protocol to understand the domain experts observations, inferences and conclusions when using the system.
+suggest case studies for the \emph{VDAR} by applying the think aloud protocol to understand the domain experts observations, inferences and conclusions when using the system.
 
@@
 %
-Both of them are practicing physicians with focus on epidemiological research.
-%
-One is a specialist in internal medicine and also the head of the SHIP, the second a radiologist and responsible for the MRI data acquisition of the study.
-\com{Years of experience} %and has a deep understanding on the available image data.
+HV and KH are practicing physicians with focus on epidemiological research.
+%
+HV is a specialist in internal medicine (23 years of experience) and also the head of the SHIP, KH a radiologist (9 years of experience) and responsible for the MRI data acquisition of the study.
+%and has a deep understanding on the available image data.
 %
@@
 %
-\paragraph{Setup.} To bypass the large geographical distance between us and the domain experts and to take their busy schedule into account, the case study was done completely web-based.
+\paragraph{Setup.} Due to the large geographical distance, the evaluation was done completely web-based.
 %
@@
 %
-The radiologist was looking for correlations with shape-related features in the data, finding that they correlate with \emph{leg pain}, \emph{age}, \emph{body height} and \emph{hormone replacement therapy status}.
+The radiologist (KH) was looking for correlations with shape-related features in the data, finding that they correlate with \emph{leg pain}, \emph{age}, \emph{body height} and \emph{hormone replacement therapy status}.
 %
@@
 
-After this initial overview, the expert performed a \emph{multivariate analysis} by introducing features, such as \emph{age}, \emph{waist circumference}, \emph{weight} or \emph{lumbar spine canal curvature} as bar charts views into the canvas area and selected subgroups to see how they are distributed and if they could observe unusual behavior in the mean-shapes.
+After this initial overview, KH performed a \emph{multivariate analysis} by introducing features, such as \emph{age}, \emph{waist circumference}, \emph{weight} or \emph{lumbar spine canal curvature} as bar charts views into the canvas area and selected subgroups to see how they are distributed and if they could observe unusual behavior in the mean-shapes.
 %
@@
 %
-The second expert started also with a \emph{multivariate analysis} using the contingency matrix to analyze non-image features, such as age-associated parameters like \emph{income}, \emph{blood fat values} or \emph{number of born children}, but found no associations of interest.
+HV started also with a \emph{multivariate analysis} using the contingency matrix to analyze non-image features, such as age-associated parameters like \emph{income}, \emph{blood fat values} or \emph{number of born children}, but found no associations of interest.
 %
@@
 %
-Therefore another \emph{multivariate analysis} was carried out by the feature \emph{gender} to the canvas from the contingency pane and selecting all female subjects (Figure~\ref{fig:application} (b)).
-%
-Cluster 1 contained almost only female subjects.
+Therefore another \emph{multivariate analysis} was carried out by the feature \emph{gender} to the canvas from the contingency pane and selecting all female subjects (Fig.~\ref{fig:application} (b)).
+%
+Cluster 1 contained primarily female subjects.
 %
@@
 %
-Since it is an pain indicator, headache was of special interest and was further investigated by incorporating a pivot table setting \emph{headache intensity} in relation to cluster affiliation.
+Since it is a pain indicator, headache was of special interest and was further investigated by incorporating a pivot table setting \emph{headache intensity} in relation to cluster affiliation.
 %
@@
 %
-We observed, that even though the visualization may treat every feature equally, the domain experts are more likely interested in features they are familiar with and have personal clinical experience with.
+We observed, that the domain experts are more likely interested in features they are familiar with and have personal clinical experience with.
 % TODO: Contingency Matrix or Adjecency Matrix
@@
 %
-To validate this hypothesis, a \emph{feature investigation} was performed by introducing the dichotomous feature ``\emph{Did you experience back pain in the last three month?}'' together with the age as mosaic plot by dropping both features on the canvas area.
-%
-The expert was not able to observe the expected effect in the visualization.
+To validate this hypothesis, a \emph{feature investigation} was performed by introducing the dichotomous feature ``\emph{Did you experience back pain in the last three months?}'' together with the age as mosaic plot by dropping both features on the canvas area.
+%
+HV was not able to observe the expected effect in the visualization.
 %
@@
 %
-The major difference seen in the visualization are therefore related to the age feature, masking differences via the back pain parameter.
+The major differences seen in the visualization are therefore related to the age feature, masking differences via the back pain parameter.
 %
@@
 %
-The analysis solely based on shape accentuated differences in \emph{gender}, not \emph{back pain}.
+The analysis solely based on shape accentuated body height differences in \emph{gender}, which clouded the differences of \emph{back pain}.
 %This makes it hard to analyze shape related features for this disease
 %
-The next step is to use a variable describing the strength of back pain and only analyzing strong back pain.
+%The next step is to use a variable describing the strength of back pain and only analyzing strong back pain.
 
@@
 %
-Then, the features derived from the clustering results for both subject groups are dropped into the canvas area as well (Figure~\ref{fig:application} (d)).
+Then, the features derived from the clustering results for both subject groups are dropped into the canvas area as well (Fig.~\ref{fig:application} (d)).
 %
@@
 %
-A \emph{multivariate analysis} using the pivot tables put gender and strong back pain in context to cluster affiliation (Figure~\ref{fig:application} (d)). 
+A \emph{multivariate analysis} using the pivot tables put gender and strong back pain in context to cluster affiliation (Fig.~\ref{fig:application} (d)). 
 %
@@
 %
-It depicted strong association with \emph{body fat}, \emph{body weight} and \emph{blood pressure} (Cluster~1) \emph{alcohol consumption} and \emph{attentiveness disorder} (Cluster~2), and \emph{amount of sleep} (Cluster~6).
+It depicted strong associations with \emph{body fat}, \emph{body weight} and \emph{blood pressure} (Cluster~1) \emph{alcohol consumption} and \emph{attentiveness disorder} (Cluster~2), and \emph{amount of sleep} (Cluster~6).
 %
@@
 \\\\
-In summary it can be stated that hypothesis-driven analysis leads to hypothesis generation by design of the tool.
+In summary it can be stated that hypothesis-driven analysis leads to hypothesis generation by design of the framework.
 %
@@
 %
-One expert emphasized the way the image data is included into information visualizations which comes much more natural to her due to her background in radiology.
+KH emphasized the way the image data is included into information visualizations which comes much more natural to her due to her background in radiology.
 %
@@
 %
-Creating adjacency matrices for sub groups, such as different age bins can help to characterize the aging process by deriving age-specific risk factors.
-
-\emph{Multivariate analysis} can also be improved by more ways of brushing the data as well as creating sub groups for comparison as a result of the hypothesis-driven analysis case.
-%
-To small feature ranges resulting in very sparse groups could potentially hinder the ability to calculate statistical resilient measures, since they require a minimum amount of subjects exhibiting the selected feature ranges.
+Creating adjacency matrices for subgroups, such as different age bins can help to characterize the aging process by deriving age-specific risk factors.
+
+\emph{Multivariate analysis} can also be improved by more ways of brushing the data as well as creating subgroups for comparison as a result of the hypothesis-driven analysis case.
+%
+Too small feature ranges yielding sparse groups could potentially hinder the ability to calculate statistical resilient measures, since they require a minimum amount of subjects exhibiting the selected feature ranges.
 %This also has to respect the number of subjects available for epidemiological studies
@@
 %
-For the visual comparison itself, the radiologist proposed an abstraction the representation them into representative landmarks, such as centers of the vertebrae and cardinal points of the lumbar spine canal.
+For the visual comparison itself, the radiologist proposed an abstraction of the representation into landmarks, such as centers of the vertebrae and cardinal points of the lumbar spine canal.
 
 \paragraph{Usage of different categorizations depending on expected outcome.}
-When categorizing numerical variables into equal groups it is possible to create very sparse categories due to outliers, for example when analyzing body weight.
+When categorizing numerical variables into equal groups it is possible to create sparse categories due to outliers, for example when analyzing body weight.
 %
@@
 %
-Balanced categories are created on the second mode, which uses quartiles or quintiles to set borders between categories.
+Balanced categories are created in the second mode, which uses quartiles or quintiles to set borders between categories.
 \paragraph{Web technologies are well suited for rapid feedback.}
@@
 \\\\
-Since it is an entirely different approach to their usual workflows, which are heavily influenced by statistical tools, it is to highlight that they had fun browsing through the data set in the presented manner and to dig out new information about the data.
+Although it is an entirely different approach to their usual workflows it is to highlight that the epidemiologists liked browsing through the data set in the presented manner and to dig out new information about the data.
 %
@@
 
-Our clinical partners from the \texttt{SHIP} are for the first time able to assess shape information of the lumbar spine and its influence to diseases.
+Epidemiologists are for the first time able to assess shape information of the lumbar spine and its influence to diseases.
 %

### 2014-06-24 draft.tex
Enhanced Related Work with Steffens part

@@
 \section{Prior and Related Work}
+
+This section describes prior and related work and covers visual analysis 
+methods incorporating both image and non-image data.
+
+\paragraph{Visual Analysis of Image and Non-Image Data.}
+
+Our work is closest to that of Steenwijk \cite{Steenwijk2010}, Turkey \cite{Turkay2013}, and Angelelli \cite{Angelelli2014} and colleagues\add{, who employ multiple coordinated view systems for the analysis of cohort study.}
+%
+%They employ multiple coordinated view systems for the analysis of cohort study data and demonstrate their approaches based on an aging study \cite{Turkay2013, Angelelli2014} and a cohort of patients suspected of having a special type of rheumatic disease \cite{Steenwijk2010}.
+
+Steenwijk et al. \cite{Steenwijk2010} propose a relational database to organize cohort study data for a visual analysis based on linked views such as parallel coordinates, scatterplots and time plots.
+%
+Information about medical image data is incorporated via mappers, which extract comparable metrics about the data.
+%
+Medical image data can be displayed individually for subjects, e.g., for analyzing outliers.
+%
+\add{While we use a similar approach when analyzing non-image data, our process also includes \emph{overview visualizations} and statistical suggestions of potentially interesting features.}
+%
+
+Turkey et al. \cite{Turkay2013} present \emph{hypothesis generation} based on descriptive statistics of the data dimensions.
+%
+Key figures describing the distribution of data values, e.g., standard deviation and interquartile range, are computed per dimension and analyzed by pairs in a \emph{deviation plot}.
+%
+%Dimensions can represent non-image data as well as variables derived from image data, e.g., volume of a brain structure.
+%
+The \emph{dual analysis} of data items and dimensions in multiple linked views led to several hypotheses in analysis sessions with domain experts.
+%
+%Care must be taken when observing changes in the deviation plot since the resulting hypotheses may impose \emph{overfitting} to the data because the measures highlight only parts of the statistical changes.
+Hypotheses based on observations in the deviation plot may impose \emph{overfitting} to the data because the measures highlight only parts of the statistical changes.
+%
+\add{Our approach uses information extracted from the segmented image data (such as 3D meshes) and variable associations with non-image epidemiological factors.}
+
+Angelelli et al. \cite{Angelelli2014} focus on a suitable data organization for the interactive visual analysis of heterogeneous cohort study data.
+%
+\rem{They propose a data-cube based model able to handle partially overlapping data subsets during the interactive visualization.}
+%
+Brain image data was integrated into the analysis by first, segmenting brain regions and tracking neural pathways and then, deriving attributes from both, e.g., volume and fractional anisotropy.
+%
+The data-cube model facilitates the seamless integration of heterogeneous image-based and non-image data.
+%
+A multiple coordinated view framework linked spatial and non-spatial data views.
+%
+%The approach is a demonstrated by a case-study analysis of selected aspects of brain connectivity.
+% --- Steffens Fazit ---
+Our integration of image data into the analysis is similar to the work of Steenwijk \cite{Steenwijk2010} and Angelelli \cite{Angelelli2014} and colleagues.
+%
+While they offer a single spatial view for visualizing image-based information of one subgroup of the cohort, we provide multiple, small views showing the information of subgroups and its respective deviation from the entire cohort.
+%
+We also integrate the data-driven definition of subgroups using clustering, statistically-based suggestions of potentially interesting features, and adaptive feature visualization.
+%tailored to epidemiological data, e.g., a mosaic plot for the frequent categorical variables.
+% %
+% %Another unique characteristic is our web-based implementation and the integration of D3.js.
+% %
+% %The latter offers an extensive set of information visualization plots, which are at the disposal of our system.
+% %
+% --- / Steffens Fazit ---
+
+Gresh et al. \cite{Gresh2000} proposed \texttt{WEAVE}, one of the first systems which concurrently analyzed medical image and non-image data using linked views.
+%
+Blaas et al. \cite{Blaas2007} presented a similar system, which also analyzed medical image data and variables derived from them using views from the feature and physical space.
+%
+Both works are restricted to the analysis of one case at a time and to non-image data with a unique spatial reference, e.g., voltage simulated across the heart muscle.
+%
+In epidemiology, multiple cases must be concurrently investigated and non-image data often lacks a spatial reference, such as gender and age.
+%
+
 % \com{\textbf{Reviewer 1}: The related work section includes merely a set of related tools.
@@
 
-\add{%Multiple coordinated views are appropriate for analyzing epidemiological data, because it suits Baldonado et al's \emph{Rule of Diversity} \cite{Baldonado2000}.
-%
-This section describes prior and related work and covers visual analysis methods incorporating both image and non-image data.
-%
-%We divide these section accordingly and also present lessons learned and from each publication.\\
-}
-%Designing a visualization, which conveys all data aspects equally, is challenging.
-% --- PARAGRAPH ---
-
-Given the number of features of epidemiological data sets and their different manifestations, the strength of different visualization techniques needs to be combined \cite{Buja91, Konyha2009}.
-%
-The principal component analysis (PCA) and similar techniques are able to reduce the dimensions by extracting most expressive components, but make the influence of each variable hard to convey.
-\\\\
-%Their focus on \emph{hypothesis generation} using parallel assessment of multiple data features makes the work of Turkay and colleagues closest to ours albeit our emphasis on processing models derived from medical image data segmentation and variables with categorical manifestations \cite{Turkay2013}.
-Turkay and colleagues \cite{Turkay2013} focus on \emph{hypothesis generation} using parallel assessment of multiple data features makes their work closest to ours albeit our emphasis is on processing models derived from medical image data segmentation and variables with categorical manifestations.
-%
-Their methods aim to amplify a hypothesis generation process for analyzing data of a Norwegian aging study.
-%
-Statistical measures of continuous features, such as mean, standard deviation, skewness, or inter-quartile range, are used to create \emph{dimension plots}.
+% ------------------ OLD SECTION ---------------------
+% \add{%Multiple coordinated views are appropriate for analyzing epidemiological data, because it suits Baldonado et al's \emph{Rule of Diversity} \cite{Baldonado2000}.
+% %
+% This section describes prior and related work and covers visual analysis methods incorporating both image and non-image data.
+% %
+% %We divide these section accordingly and also present lessons learned and from each publication.\\
+% }
+% %Designing a visualization, which conveys all data aspects equally, is challenging.
+% % --- PARAGRAPH ---
+%
+% Given the number of features of epidemiological data sets and their different manifestations, the strength of different visualization techniques needs to be combined \cite{Buja91, Konyha2009}.
+% %
+% The principal component analysis (PCA) and similar techniques are able to reduce the dimensions by extracting most expressive components, but make the influence of each variable hard to convey.
+% \\\\
+% %Their focus on \emph{hypothesis generation} using parallel assessment of multiple data features makes the work of Turkay et al.
+closest to ours albeit our emphasis on processing models derived from medical image data segmentation and variables with categorical manifestations \cite{Turkay2013}.
+% Turkay et al.
+\cite{Turkay2013} focus on \emph{hypothesis generation} using parallel assessment of multiple data features makes their work closest to ours albeit our emphasis is on processing models derived from medical image data segmentation and variables with categorical manifestations.
+% %
+% Their methods aim to amplify a hypothesis generation process for analyzing data of a Norwegian aging study.
+% %
+% Statistical measures of continuous features, such as mean, standard deviation, skewness, or inter-quartile range, are used to create \emph{dimension plots}.
 %that make them comparable with respect to the derived descriptive measures, such as voxel number of a segmented structure.
-%
-These transform dimensions into data points and make them comparable with respect to the derived descriptive measures, making them comparable.
-%
-%This not only allows for comparing all continuous variables in a single plot but make their distribution change comprehensible.
-%
-%This requires a good descriptive measure which captures the kind of change the user is interested in or which reflects unexpected data behavior.
-%
-%The technique was applied to features obtained by segmenting the brain into 45 parts and measure the voxel number, volume and properties of the intensity values.
-%
-The method is strongly dependent on the descriptive measures of the epidemiological factors.
-%
-%\emph{Over-fitting} of hypotheses to the data may be imposed  is a danger pointed out by the authors when hypotheses to the data based on observations of changes in these plots the measure highlights only subsets of statistical changes.
-Hypotheses based on observations of changes in these plots may impose \emph{overfitting} to the data because the measures highlight only parts of the statistical changes.
-%
-%Our approach sticks more to the information extracted from the segmented image data and variable associations with non-image epidemiological factors.
-\add{Our approach uses information extracted from the segmented image data (such as 3D meshes) and variable associations with non-image epidemiological factors.}
-% Disadvantage - Nur 2 abgeleitete Variablen sichtbar, schlecht mit anderen Variablen in Verbindung zu bekommen (müsste jedoch auch gehen) - viele Dinge gehen durch die Abstraktion aber auch flöten!
-
-\paragraph{Visualizing Image and Non-Image Data.}
-%\add{One major focus is the connection of image and non-image data.}
-%TODO aus Steenwijk: Visual data analysis combining multi-dimensional single patient medical data with real-time linked scatter plots and parallel coordi- nates representations was first investigated in the WEAVE system of Gresh et al.
+% %
+% These transform dimensions into data points and make them comparable with respect to the derived descriptive measures, making them comparable.
+% %
+% %This not only allows for comparing all continuous variables in a single plot but make their distribution change comprehensible.
+% %
+% %This requires a good descriptive measure which captures the kind of change the user is interested in or which reflects unexpected data behavior.
+% %
+% %The technique was applied to features obtained by segmenting the brain into 45 parts and measure the voxel number, volume and properties of the intensity values.
+% %
+% The method is strongly dependent on the descriptive measures of the epidemiological factors.
+% %
+% %\emph{Over-fitting} of hypotheses to the data may be imposed  is a danger pointed out by the authors when hypotheses to the data based on observations of changes in these plots the measure highlights only subsets of statistical changes.
+% Hypotheses based on observations of changes in these plots may impose \emph{overfitting} to the data because the measures highlight only parts of the statistical changes.
+% %
+% %Our approach sticks more to the information extracted from the segmented image data and variable associations with non-image epidemiological factors.
+% \add{Our approach uses information extracted from the segmented image data (such as 3D meshes) and variable associations with non-image epidemiological factors.}
+% % Disadvantage - Nur 2 abgeleitete Variablen sichtbar, schlecht mit anderen Variablen in Verbindung zu bekommen (müsste jedoch auch gehen) - viele Dinge gehen durch die Abstraktion aber auch flöten!
+%
+% \paragraph{Visual Analysis of Image and Non-Image Data.}
+% %\add{One major focus is the connection of image and non-image data.}
+% %TODO aus Steenwijk: Visual data analysis combining multi-dimensional single patient medical data with real-time linked scatter plots and parallel coordi- nates representations was first investigated in the WEAVE system of Gresh et al.
 [14].
@@
 [8].
-Gresh et al. \cite{Gresh2000} proposed \texttt{WEAVE}, one of the first systems which concurrently analyzed medical image and non-image data using linked views.
-%
-%\add{The functionality is limited to project spatial related data to surface renderings of the image data and use By projecting surface models presented in 3D views to information visualizations, such as scatterplots, the authors were to support medical resea}
-%
-Blaas et al. \cite{Blaas2007} presented a similar system, which also analyzed medical image data and variables derived from them using views from the feature and physical space.
-% From Blaas2007 - Tasks:
-%- Investigating combinations of features for finding spatial separations of point sets in physical space (segmentation)
-%- Classification of feature point sets (pattern classification)
-%- Finding clusters, patterns, and relations in the data
-\add{The parameter in both paper are part of the spatio-temporal domain and have a colocation in the image data, making it possible to plot them onto the image data representation to investigate interesting combinations.
-%
-Epidemiological features often don't have an exact spatial location in the image data.}
-%
-\rem{They incorporated data mining methods such as dividing the data space by using a k-nearest neighbor technique and the PCA.}
-%
-Steenwijk et al. \cite{Steenwijk2010} employ a relational database to organize the data for visualizing subject data using linked views such as parallel coordinates, scatterplots and time plots.
-%
-\add{Information about medical image data is incorporated via mappers, which extract comparable metrics about the data.
-%
-Medical image data can be displayed individually for subjects, e.g. for analyzing outliers.
-%Display of medical data is possible for single subjects, for example to look at outliers.
-%
-While we use a similar approach when analyzing non-image data, our process also includes \emph{overview visualizations} and statistical suggestions of potentially interesting features.}
-
-\paragraph{Visualizing Heterogenous Non-Image Data.}
+% Gresh et al. \cite{Gresh2000} proposed \texttt{WEAVE}, one of the first systems which concurrently analyzed medical image and non-image data using linked views.
+% %
+% %\add{The functionality is limited to project spatial related data to surface renderings of the image data and use By projecting surface models presented in 3D views to information visualizations, such as scatterplots, the authors were to support medical resea}
+% %
+% Blaas et al. \cite{Blaas2007} presented a similar system, which also analyzed medical image data and variables derived from them using views from the feature and physical space.
+% % From Blaas2007 - Tasks:
+% %- Investigating combinations of features for finding spatial separations of point sets in physical space (segmentation)
+% %- Classification of feature point sets (pattern classification)
+% %- Finding clusters, patterns, and relations in the data
+% \add{The parameter in both paper are part of the spatio-temporal domain and have a colocation in the image data, making it possible to plot them onto the image data representation to investigate interesting combinations.
+% %
+% Epidemiological features often don't have an exact spatial location in the image data.}
+% %
+% Steenwijk et al. \cite{Steenwijk2010} employ a relational database to organize the data for visualizing subject data using linked views such as parallel coordinates, scatterplots and time plots.
+% %
+% \add{Information about medical image data is incorporated via mappers, which extract comparable metrics about the data.
+% %
+% Medical image data can be displayed individually for subjects, e.g. for analyzing outliers.
+% %Display of medical data is possible for single subjects, for example to look at outliers.
+% %
+% While we use a similar approach when analyzing non-image data, our process also includes \emph{overview visualizations} and statistical suggestions of potentially interesting features.}
+% ------------------ / OLD SECTION ---------------------
+
+\paragraph{Visual Analysis of Heterogenous Non-Image Data.}
 Zhang et al. \cite{Zhang2012} provide a web-based system for analyzing subject groups with linked views and batch-processing capabilities for categorizing new subject entries into the data set.
@@
 %
-%By plotting p-values in ventricle surfaces, Chou and colleagues were able to map disease-associated values directly on a 3D tissue representations \cite{Chou2009}.
+%By plotting p-values in ventricle surfaces, Chou et al. were able to map disease-associated values directly on a 3D tissue representations \cite{Chou2009}.
 %
@@
 %
-Oeltze et al. \cite{Oeltze2007} combined a linked view representation of results from a statistical analysis with feature localizations of the tissue perfusion. % with the goal of its evaluation.
-%
-This was focused on individual patients and not on cohort study data. 
-%
-
+Oeltze et al. proposed a multiple coordinated view approach for the analysis of medical perfusion data \cite{Oeltze2007} and biological multi-channel fluorescence microscopy data \cite{Oeltze2011}.
+%
+The approach is restricted to the investigation of a single subject at a time.
+%
 %While we take similar steps when analyzing the data such as employing statistical tests, our data is mostly independent from the medical image data and is not describing it.
 %--except the variables derived the segmentation model itself.
-\rem{While we take similar steps when analyzing the data, such as employing statistical tests, our data are not associated to a geographical or temporal context, which affects the techniques described in the following section.}
-% --- PARAGRAPH --- % TODO Add Paragraph when removing ADD
+% --- PARAGRAPH ---
+
 A workflow and definition of terminology of Visual Analysis (VA) techniques is provided by Lammarsch et al. \cite{Lammarsch2011}.
@@
 %
-LESSON LEARNED HIERAUS??
-%
 Weaver et al.
@@
 \\\\
-\com{Dieser Abschnitt wurde hinzugefuegt, weil mehrere Reviewer meinten, dass der Related Work Teil zu wenig die Einzigartigkeit der Arbeit hervorbringt.}
 %Related work focused on epidemiological data often deals with time-dependent data, hypothesis generation by guidance through parameter suggestions is seldom.
@@
 %
-\add{The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013.} %(676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
+\add{The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
+%(676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
 }

### 2014-06-25 draft.tex
reduction marathon, changing image sizes and much of the text

@@
 % ----------- /OLD ABSTRACT ---------
-\com{Epidemiological Features Begriff einfuehren}
-\add{
-Large-scale longitudinal epidemiological studies investigate groups of people with common characteristics or experiences (a cohort) to characterize health and identify disease-specific risk factors.
-%
-Cohort study data comprise thousands of individuals and hundreds of heterogenuous variables, describing the medical condition as well as demographic and lifestyle factors, and more recently, medical image data.
-%
-In the standard epidemiological workflow, a hypothesis is given and validated by analyzing a subset of possibly relevant variables over manually defined subgroups of the cohort, e.g., based on age or body weight.
-%
-A potential impact of the neglected variables remains unnoticed, data-driven group definition is not supported, and capabilities for integrating the image data into the analysis are missing.\\
-%
-We propose an Interactive Visual Analysis (IVA) approach that enables epidemiologists to rapidly investigate the entire data pool for both, hypothesis validation and generation.
-%
-Furthermore, we suggest a strategy for incorporating the image data which involves shape-based object detection and the derivation of attributes describing the object shape.
+\add{%Large-scale longitudinal epidemiological studies investigate groups of people with common characteristics or experiences (a cohort) to characterize health and identify disease-specific risk factors.
+%
+%Cohort study data comprise thousands of individuals and hundreds of heterogenuous variables (\emph{features}), describing the medical condition as well as demographic and lifestyle factors, and more recently, medical image data.
+Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease-specific risk factors.
+%
+Cohort studies comprise of heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors and more recently, medical image data.
+%
+%In the standard epidemiological workflow, a hypothesis is given and validated by analyzing a subset of possibly relevant variables over manually defined subgroups of the cohort, e.g., based on age or body weight.
+%
+%A potential impact of the neglected variables remains unnoticed, data-driven group definition is not supported, and capabilities for integrating the image data into the analysis are missing.\\
+%
+We propose an Interactive Visual Analysis (\emph{IVA}) approach that enables epidemiologists to rapidly investigate the entire data pool for \emph{hypothesis validation} and \emph{generation}.
+%
+We incorporate image data, which involves shape-based object detection and the derivation of attributes describing the object shape.
 %
@@
 %
-To account for the numerous categorical variables, a mosaic plot is integrated.
-%
-The views are equipped with brushing facilities and augmented by 3D shape renderings of the segmented objects, e.g., each bar in a histogram is overlayed with a mean shape of the associated subgroup of the cohort.
-%
-We integrate clustering of variables and object shape for data-driven subgroup definition and statistical key figures for measuring the association between variables.
-%
-We demonstrate the IVA approach by validating and generating hypotheses related to lower back pain and the aging process of the lumbar spine.}
+%To account for the numerous categorical variables, a mosaic plot is integrated.
+%
+The views are equipped with brushing facilities and augmented by 3D shape renderings of the segmented objects, e.g., each bar in a histogram is overlaid with a mean shape of the associated subgroup of the cohort.
+%
+We integrate an \emph{overview visualization}, clustering of features and object shape for data-driven subgroup definition and statistical key figures for measuring the association between variables.
+%
+We demonstrate the \emph{IVA} approach by validating and generating hypotheses related to lower back pain as part of a qualitative evaluation.}%and the aging process of the lumbar spine.}
 } % end of abstract
@@
 \maketitle
-
 %% \section{Introduction} %for journal use above \firstsection{..} instead
@@
 \emph{Hypothesis generation} gains importance since the number of epidemiological features increases and the focus shifts towards more complex relations involving more than two features.
-\\\\
 Our contributions are:
 \begin{itemize}
-	\item an Interactive Visual Analysis workflow for cohort study data to allow both, hypothesis-driven analysis and hypothesis generation--based on a characterization of the standard epidemiological workflow,
+	\item an \add{IVA} workflow for cohort study data to allow both, hypothesis-driven analysis and hypothesis generation--based on a characterization of the standard epidemiological workflow,
 	\item visualization techniques, which incorporate both information visualization and 3D rendering of organ shapes as well as combining them with epidemiological graphics and key figures,
-	\item highlighting interesting subject groups and feature associations using shape-based clustering and statistical contingency measures,
+	\item highlighting subject groups and feature associations using shape-based clustering and statistical contingency measures.
+	%\item highlighting interesting subject groups and feature associations using shape-based clustering and statistical contingency measures.
 %	\item an implementation of the presented methods in a web framework based on \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}.
@@
 %
-Following Thew et al \cite{Thew2009}, the workflow can be characterized as follows:
+Following Thew et al. \cite{Thew2009}, the workflow can be characterized as follows:
 %
@@
 %
-
+%
 Grouping subjects using epidemiological features is essential in cohort studies to allow per-group risk determination.
@@
 %
-To determine whether a subject is prone to be affected by a certain disease, \emph{relative risks} are determined including confidence intervals that indicate the certainty of that feature being a risk factor.
+\emph{Relative risks} are determined to detect if a subject is prone to be affected by a certain disease.
+%
+This includes confidence intervals indicating the certainty of that feature being a risk factor.
 %To determine whether a subject is prone to be affected by a certain disease, \emph{relative risks} are expressed through the evaluation of p-values, which indicate statistical significance.
@@
 %
-\add{An extreme form of confounding is called \emph{Simpson's Paradox} \cite{SimpsonsParadox}, where the confounder reverses the observed effect between two features.
-%
-%An example is the higher mortality rate of non-insulin dependent diabetes opposed to insulin dependent diabetes.
-A famous example is the Berkeley University gender bias case, where significantly less women were admitted.
-%
-Analysis later showed that women applied mostly for strongly restricted subjects and when only looking at acceptance divided by departments there was actually a small bias toward women.
+%\add{An extreme form of confounding is called \emph{Simpson's Paradox} \cite{SimpsonsParadox}, where the confounder reverses the observed effect between two features.
+%
+%A famous example is the Berkeley University gender bias case, where significantly less women were admitted.
+%
+%Analysis later showed that women applied mostly for strongly restricted subjects and when only looking at acceptance divided by departments there was actually a small bias toward women.
 %
@@
 %When for example surgeons are ranked regarding diseased patients, without including the risk of the surgeries, the results may be distorted.
-}
-
-Statistical tools such as \texttt{SPSS}\footnote{Product of IBM; \url{http://ibm.com/software/analytics/spss/}} play a major role for analyzing epidemiological data.
+%
+
+Statistical tools such as \texttt{SPSS}\footnote{Product of IBM; \href{http://ibm.com/software/analytics/spss/}{\texttt{ibm.com/software/analytics/spss/}}} play a major role for analyzing epidemiological data.
 %
@@
 % \end{itemize}
-\add{Access to epidemiological data is limited and usually controlled by ethics committees.
-%
-Large studies, such as the SHIP, require written proposals reasoning about each requested features and the expected outcome.
-%
-Since the number of features and modalities is steadily increasing, the epidemiological community is aware of the lack of specialized analysis methods, making an acceptance more likely.
-%In our experience, the applying for a large amount of features required for a \underline{IVA} system requires 
-%
-Due to the high sensibility of the data, all identity-revealing information of a person, such as name or place of residence is anonymized.
-}
+% \rem{Access to epidemiological data is limited and usually controlled by ethics committees.
+% %
+% Large studies, such as the SHIP, require written proposals reasoning about each requested features and the expected outcome.
+% %
+% Since the number of features and modalities is steadily increasing, the epidemiological community is aware of the lack of specialized analysis methods, making an acceptance more likely.
+% %
+% Due to the high sensibility of the data, all identity-revealing information of a person, such as name or place of residence is anonymized.
+% }
 % --- PARAGRAPH ---
@@
 %
-\add{Analysis of the data is usually carried out by calculating correlations,
+\add{Data analysis is usually carried out by calculating correlations,
 which is challenging due to the data type heterogeneity.
 %
-Parameter correlation can also be associated with confounding, which can not be automatically predicted but rather has to be judged by a domain expert.
-%
-Sparse populated features are also hard to assess statistically as too few data samples have a higher chance to distort the real underlying distributions.
+Parameter correlation can also be associated with confounding, which can not be automatically predicted, it has to be judged by a domain expert.
+%
+Sparse populated features are hard to assess statistically.
+%
+Too few data samples may distort the real underlying distributions.
 %Also, correlation can both be of epidemiological interest but 
@@
 %
-%The equipment used to gather medical image data is not updated or changed during a longitudinal study to ensure comparability in and between acquisition cycles.
-The equipment used to gather medical image data is not updated or changed during a longitudinal study to ensure comparability in and between acquisition cycles.
+%\rem{The equipment used to gather medical image data is not updated or changed during a longitudinal study to ensure comparability in and between acquisition cycles.}
 
 \paragraph{Image Analysis.}
-\com{Take up huge amount of space!}
 %
@@
 %
-Morphometric information from landmarks comprises thickness, diameter or length of a structure as well as grey value distribution in an area (used for determining the type of tissue).
+Morphometric information from landmarks comprises thickness, diameter or length of a structure as well as grey value distribution in an area. %\rem{(for determining the tissue type).}
 
@@
 %
-Data is collected without focus on a group of diseases.
-%
-This allows the data set to be queried regarding many diseases and conditions.
+Data are collected without focus on a group of diseases.
+%
+This allows to query the data regarding many diseases and conditions.
 %
@@
 %
-A second cohort \texttt{SHIP-Trend} was established in 2008.
-%
-The protocols for examining the subjects between \texttt{SHIP} and \texttt{SHIP-Trend} remained the same, making them comparable.
-%
-The overall examination time for each person attending the study is two days.
+% \rem{A second cohort \texttt{SHIP-Trend} was established in 2008.
+% %
+% The protocols for examining the subjects between \texttt{SHIP} and \texttt{SHIP-Trend} remained the same, making them comparable.
+% %
+% The overall examination time for each person attending the study is two days.}
 
@@
 %
-\rem{They propose a data-cube based model able to handle partially overlapping data subsets during the interactive visualization.}
+%\rem{They propose a data-cube based model able to handle partially overlapping data subsets during the interactive visualization.}
 %
@@
 % --- Steffens Fazit ---
-Our integration of image data into the analysis is similar to the work of Steenwijk \cite{Steenwijk2010} and Angelelli \cite{Angelelli2014} and colleagues.
-%
-While they offer a single spatial view for visualizing image-based information of one subgroup of the cohort, we provide multiple, small views showing the information of subgroups and its respective deviation from the entire cohort.
-%
-We also integrate the data-driven definition of subgroups using clustering, statistically-based suggestions of potentially interesting features, and adaptive feature visualization.
-%tailored to epidemiological data, e.g., a mosaic plot for the frequent categorical variables.
+%
+\add{Our integration of image data into the analysis is similar to the work of Steenwijk \cite{Steenwijk2010} and Angelelli \cite{Angelelli2014} and colleagues.
+%
+While they offer a single spatial view for visualizing image-based information of one subgroup of the cohort, we provide multiple, small views showing the information of subgroups and its respective deviation from the entire cohort.}
+%
+%\rem{We also integrate the data-driven definition of subgroups using clustering, statistically-based suggestions of potentially interesting features, and adaptive feature visualization.} %tailored to epidemiological data, e.g., a mosaic plot for the frequent categorical variables.
 % %
@@
 %
-
 % \com{\textbf{Reviewer 1}: The related work section includes merely a set of related tools.
@@
 % ------------------ / OLD SECTION ---------------------
-
 \paragraph{Visual Analysis of Heterogenous Non-Image Data.}
@@
 %
-\add{While their goal to identify possible risk factors using socio-economic and health data is similar to ours, they focus on iteratively refining defined hypothesis and also focus on geographical data.
+\add{While their goal to identify possible risk factors using socio-economic and health data are similar to ours, they focus on iteratively refining defined hypothesis and also focus on geographical data.
 %
@@
 While the works shows, how different visualization modalities provide insight into these data sets, it focuses on the time aspect, which is not present for our data.}
-
-\paragraph{Commercial Data Visualization.}
-Commercial systems such as \texttt{Tableau}\footnote{Owned by Tableau Software; \url{http://tableausoftware.com}} or \texttt{Spotfire}\footnote{Owned by TIBCO; \url{http://spotfire.tibco.com}} provide a rich user interface that enables a Visual Analytics approach.
-%
-With little effort, linked views can be created, but the data processing possibilities, such as derivation of new variables or the 3D rendering capabilities, are very limited.
-%
-These systems with their origin in business intelligence applications do not support epidemiological workflows well, in particular workflows that also consider medical image data.
+%
+% \paragraph{Commercial Data Visualization.}
+% Commercial systems such as \texttt{Tableau}\footnote{Owned by Tableau Software; \href{http://tableausoftware.com}{\texttt{tableausoftware.com}}} or \texttt{Spotfire}\footnote{Owned by TIBCO; \href{http://spotfire.tibco.com}{\texttt{spotfire.tibco.com}}} provide a rich user interface that enables a Visual Analytics approach.
+% %
+% With little effort, linked views can be created, but the data processing possibilities, such as derivation of new variables or the 3D rendering capabilities, are very limited.
+% %
+% These systems with their origin in business intelligence applications do not support epidemiological workflows well, in particular workflows that also consider medical image data.
 % Image Data
@@
 %
-\com{Hier noch einer Einschaetzung, oder reicht das schon?}
-%
 Niemann et al. \cite{Niemann2014} presented an interactive data mining tool for the assessment of risk factors of hepatic steatosis, the fatty liver disease.
@@
 
-A workflow and definition of terminology of Visual Analysis (VA) techniques is provided by Lammarsch et al. \cite{Lammarsch2011}.
+Lammarsch et al. \cite{Lammarsch2011} provided a workflow and terminology definition of Visual Analysis techniques.
 %
@@
 %
-The VA-process is also reflected in our \emph{IVA}-Loop.\\
-% --- PARAGRAPH --- % TODO Add Paragraph when removing ADD
+The VA-process is also reflected in our \emph{IVA}-Loop.
+% --- PARAGRAPH ---
+
 Baldonado et al. \cite{Baldonado2000} presented rules for designing multiple coordinated views.
@@
 %
-However, our current work neglects geographical and temporal aspects.
-%
-Instead we employ a more abstract model and consider the subjects as living in a joint image space where each of them is represented by a particular segmented organ or structure.
+Our current work neglects geographical and temporal aspects.
+%
+Instead we employ a abstract model and consider the subjects as living in a joint image space where each of them is represented by a segmented organ or structure.
 %
@@
 %
-Then, two types of dependent variables exist: 
-\begin{enumerate}
-	\item the socio-demographic data and medical examination results and 
-	\item variables derived from the segmented structures, e.g., spinal curvature or misalignment of the vertebrae.
-\end{enumerate}
+Then, two types of dependent variables exist: the socio-demographic data and medical examination results, and variables derived from the segmented structures, e.g., spinal curvature or misalignment of the vertebrae.
+% \begin{enumerate}
+% 	\item the socio-demographic data and medical examination results and
+% 	\item variables derived from the segmented structures, e.g., spinal curvature or misalignment of the vertebrae.
+% \end{enumerate}
 An alternative of the image space would be the shape space generated by extracting the major modes of variation from all segmentation results \cite{Busking2010a}.
@@
 
-\paragraph{Local Investigation.} Refers to the inspection of dependent variables with respect to certain subsets of the image or shape space.
+\paragraph{Local Investigation.} Refers to the inspection of dependent variables with respect to subsets of the image or shape space.
 %
@@
 
-\paragraph{Multivariate Analysis.} Refers to an investigation of multi-variate properties of the dependent data by specifying a feature in one attribute view and at the same time analyzing the value distribution with respect to other variables in other attribute views.
+\paragraph{Multivariate Analysis.} Refers to an investigation of multi-variate properties of the dependent data by specifying a feature in one attribute view while analyzing the value distribution with respect to other variables in other attribute views.
 %
@@
 \subsection{Data Preprocessing} \label{Data Preprocessing}
-A number of transformations are carried out in the data preprocessing stage.
+%A number of transformations are carried out in the data preprocessing stage.
 %Transformation operations on the data to prepare it for an \emph{IVA} system are denoted as data preprocessing.
@@
 %
-Exporting the data in the respective tool to a CSV file and then converting it to file types that are easily manageable, such as JSON or XML, makes it readable for all modern programming languages.
-%
-%This can be achieved by using data wrangling tools such as \texttt{OpenRefine}\footnote{Developed by Google, Open Source; \url{http://openrefine.org}}, which also validate the data (find missing data, clean up bad formatting, transform scales).
+Exporting the data in the respective tool to a CSV file and then converting it to file types that are easily manageable, such as JSON, makes it readable for modern programming languages.
+%
+%This can be achieved by using data wrangling tools such as \texttt{OpenRefine}\footnote{Developed by Google, Open Source; \href{http://openrefine.org}{\texttt{openrefine.org}}}, which also validate the data (find missing data, clean up bad formatting, transform scales).
 %
@@
 %
-In principle, model-based approaches are effective for detection \cite{Rak2013} and segmentation \cite{Gloger2010}.%\cite{Gloger2010, Gloger2012}.
+In principle, model-based approaches are effective for detection \cite{Rak2013} and segmentation \cite{Gloger2010}. %\cite{Gloger2010, Gloger2012}.
 %
@@
 Grey value comparison is used to measure the quantity of fat, water, and--application-specific--the iron content (liver) or the distribution of grey and white brain tissue.
-
+%
 Morphometric features are derived to allow for statistical comparison of the tissue, which incorporates mostly positions, diameters, volumes and relative distances and alignment to other structures.
@@
  \centering
- \includegraphics[width=2.3in]{figures/InteractionLoop}
+ \includegraphics[width=2.0in]{figures/InteractionLoop}
  \caption{Detailed \emph{IVA Loop} as extension from Figure~\ref{fig:WorkflowComparison}.
@@
  %
- The data are visualized in the range and domain space, which can then be brushed, yielding new groups, to be investigated using further features.
- %
- Note that adjacent steps are directly connected via feedback loops allowing for an iterative refinement and giving as much freedom to the user as possible.}
+ The data are visualized in the range and domain space, which can be brushed, yielding new groups, to be investigated using further features.
+ %
+ Note that adjacent steps are directly connected via feedback loops, allowing for an iterative refinement and giving as much freedom to the user as possible.}
  \label{fig:InteractionLoop}
@@
 %
-\emph{Hypothesis generation} with focus on image data starts with a shape-based clustering.
-%
-\com{ToDo: Multivariate analysis using the adjacency matrix may have also be explained here, depending on the Application chapter.}
+\add{Hypothesis generation with focus on image data starts with a shape-based clustering or an \emph{overview visualization} of all features.}
 %
@@
 %
-The visualization techniques have to combine both image-and non-image data in order to set domain and range data in relation to each other.
+The visualization techniques have to combine both image- and non-image data to set domain and range data in relation to each other.
 %
@@
 %
-Brushed regions are treated like features, as they divide the subject space just like categorical features.
-%
-Selecting features also triggers a \emph{multivariate analysis} using contingency values (described in the following section) to highlight features associated with selected features.
+Brushed regions are treated like categorical features, as they divide the subject space in the same way.
+%
+Selecting features also triggers a \emph{multivariate analysis} using contingency values (described in the following section) to highlight \add{associated} features.  %\rem{with selected features.}
 %
@@
 In the following sections we present the different parts of our proposed \emph{IVA} system for image-based cohort study analysis.
-
-\begin{figure}[htb]
- \centering
- \includegraphics[width=3.5in]{figures/similarity_matrix}
- \caption{Adjacency matrix of 129 features (127 data set variables, 2 cluster results) with a grand total of 16,641 combinations.
- %
- Similarity is calculated using the \emph{Cram\'{e}r's V} contingency value.
- %
- Color brightness encodes association strength.
- %Mouse-over focuses a entry and highlights the feature names for better readability.
- Mouse-over an entry enlarges the feature names for better readability.
- %
- The enlarged excerpt shows associations for shape clusters of subjects with and without diagnosed spine attrition, which show associations between gender, weight, body height and smoking behavior.
- }
- \label{fig:similarity}
-\end{figure}
-
 % \subsection{System Structure and Design} \label{Structure and Workflow}
@@
 %
-An overview of all features is presented in a sidebar where they are categorized into different types, such as somatometric (measurements of the dimensions of the human body), disease- or lifestyle-related, pain indicators and laboratory data.
+It categorizes all features into different types, such as somatometric (measurements of the human body dimensions), disease- or lifestyle-related, pain indicators and laboratory data.
 %
@@
 %
-Variables can be dragged from the sidebar into the canvas area for a \emph{feature localization} which works as follows.
+Variables can be dragged from the sidebar into the canvas area for a \emph{feature localization}, which works as follows.
 %
@@
 \paragraph{Adaptive Feature Visualization.} \label{sec:AdaptiveFeatureVisualization}
+\begin{figure}[htb]
+ \centering
+ \includegraphics[width=3.5in]{figures/similarity_matrix}
+ \caption{Adjacency matrix of 129 features (127 data set variables, 2 cluster results) with a grand total of 16,641 combinations.
+ %
+ Similarity is calculated using the \emph{Cram\'{e}r's V} contingency value.
+ %
+ Color brightness encodes association strength.
+ %Mouse-over focuses a entry and highlights the feature names for better readability.
+ Mouse-over an entry enlarges the feature names for better readability.
+ %
+ The enlarged excerpt shows associations for shape clusters of subjects with and without diagnosed spine attrition, which show associations between gender, weight, body height and smoking behavior.
+ }
+ \label{fig:similarity}
+\end{figure}
 The visualization type, inspired by \texttt{GPLOMS} \cite{GPLOMS, Francois2013}, is dynamically chosen based on the feature types and number to allow for \emph{multivariate analysis}.
@@
 %
-This allows to assess local shape changes (Fig.~\ref{fig:visualization}).
-%
-This is an important information to the epidemiologist, because until now they were not able rapidly validate shape differences based on non-image features.
-%
-If a feature is dropped on an existing plot, the visualization changes dynamically to make them comparable (Fig.~\ref{fig:visualization} (right)).
+This allows to assess local shape changes (Fig.~\ref{fig:visualization}) and is an important information to the epidemiologist.
+%
+Until now they were not able rapidly validate shape differences based on non-image features.
+%
+Dropping a feature on an existing plot adapts the visualization dynamically to allow for comparison (Fig.~\ref{fig:visualization} (right)).
 % --- PARAGRAPH ---
@@
 \paragraph{Automated Feature Suggestion using a Contingency Matrix.}
-As previously discussed, highlighting potentially interesting values in the data set is one major benefits of the \emph{IVA}-powered approach and belongs to the \emph{multivariate analysis} pattern,  analyzing features outside the shape space.
+%\rem{As previously discussed}
+Highlighting potentially interesting values in the data set is one major benefits of the \emph{IVA}-powered approach and belongs to the \emph{multivariate analysis} pattern,  analyzing features outside the shape space.
 %
@@
 %
-Since \emph{Cram\'{e}r's V} is always positive, it does not allow to infer about dependency direction.
+%Since \emph{Cram\'{e}r's V} is always positive, 
+\emph{Cram\'{e}r's V} does not allow to infer about dependency direction.
+%
 
@@
 %
-This aims to highlight features possibly associated with the focused hypothesis as well as trigger generation of new hypotheses.
+This aims to highlight features possibly associated with the focused hypothesis as well as trigger new hypotheses.
 %
@@
 %
-The contingency matrix visualization is an appropriate \emph{overview} visualization, something the epidemiological community lacks and is in great need of.
+The contingency matrix visualization is an \emph{overview visualization}, something the epidemiological community lacks and is in great need of.
 
@@
 %
-As a consequence, the \emph{contingency pane}, a table containing correlating features for the last added visualizations in descending order of the \emph{Cram\'{e}r's V} value.
+As a consequence, the \emph{contingency pane}, a table containing correlating features for the last added visualizations in descending order of the \emph{Cram\'{e}r's V} value was added.
 %
@@
  \includegraphics[width=3.5in]{figures/technologies}
- \caption{The front-end solution (left) uses state-of-the-art web technologies such as \emph{HTML5}/\emph{CSS3}, \emph{WebGL} and \emph{SVG} to display the data.
+ \caption{The front-end solution (left) uses \emph{HTML5}/\emph{CSS3}, \emph{WebGL} and \emph{SVG} to display the data.
  % 
@@
  %
- All computation-heavy operations, such as calculation of mean shapes or distances, as well as statistical processing are done by the server to keep hardware requirements of client systems low. 
+ All computation-heavy operations, such as calculation of mean shapes or distances, are performed on the server-side. 
+ %All computation-heavy operations, such as calculation of mean shapes or distances, as well as statistical processing are done by the server to keep hardware requirements of client systems low. 
  %
@@
 %
-Sophisticated libraries/languages, such as the \texttt{Visualization Toolkit} (\texttt{VTK})\footnote{Developed by Kitware Inc; \url{http://vtk.org}} or \texttt{R}\footnote{Open Source; \url{http://r-project.org}} for statistics, are either not available at all or only accessible through complex client-server systems.
+Sophisticated libraries/languages, such as the \texttt{Visualization Toolkit} (\texttt{VTK})\footnote{Developed by Kitware Inc; \href{http://vtk.org}{\texttt{vtk.org}}} or \texttt{R}\footnote{Open Source; \href{http://r-project.org}{\texttt{r-project.org}}} for statistics, are either not available at all or only accessible through complex client-server systems.
 %
@@
 %
-\add{The user needs a fully updated modern web browser, which supports these new technologies, such as WebGL.}
-%Due to missing ports of sophisticated and specialized libraries/languages, such as the \texttt{Visualization Toolkit} (\texttt{VTK})\footnote{Developed by Kitware Inc; \url{http://vtk.org}} or \texttt{R}\footnote{Open Source; \url{http://r-project.org}} for statistics, many standard methods need to be written from scratch.
-%Due to missing ports of sophisticated and specialized libraries/languages, such as the \texttt{Visualization Toolkit} (\texttt{VTK})\footnote{Developed by Kitware Inc; \url{http://vtk.org}} or \texttt{R}\footnote{Open Source; \url{http://r-project.org}} for statistics, many standard methods need to be written from scratch.
-%
-
-The back-end is written using \texttt{NodeJS}\footnote{Developed by Joyent Inc, \url{http://nodejs.org}}, which is based on the Google V8 Javascript runtime environment.
+%\rem{The user needs a fully updated modern web browser, which supports these new technologies, such as WebGL.}
+%Due to missing ports of sophisticated and specialized libraries/languages, such as the \texttt{Visualization Toolkit} (\texttt{VTK})\footnote{Developed by Kitware Inc; \href{http://vtk.org}{\texttt{vtk.org}}} or \texttt{R}\footnote{Open Source; \href{http://r-project.{\texttt{r-project}}org}} for statistics, many standard methods need to be written from scratch.
+%Due to missing ports of sophisticated and specialized libraries/languages, such as the \texttt{Visualization Toolkit} (\texttt{VTK})\footnote{Developed by Kitware Inc; \href{http://vtk.org}{\texttt{vtk.org}}} or \texttt{R}\footnote{Open Source; \href{http://r-project.{\texttt{r-project}}org}} for statistics, many standard methods need to be written from scratch.
+%
+
+The back-end is written using \texttt{NodeJS}\footnote{Developed by Joyent Inc, \href{http://nodejs.org}{\texttt{nodejs.org}}}, which is based on the Google V8 Javascript runtime environment.
 %
@@
 %
-The requested data is transmitted when a client connects.
+The requested data are transmitted when a client connects.
 %
@@
 
-The front-end is created using \texttt{Bootstrap}\footnote{Developed by Twitter, \url{http://getbootstrap.com}} as foundation for the layout and basic UI elements using \texttt{HTML5}, \texttt{CSS3} and \texttt{Javascript}.
+The front-end is created using \texttt{Bootstrap}\footnote{Developed by Twitter, \href{http://getbootstrap.com}{\texttt{getbootstrap.com}}} as foundation for the layout and basic UI elements using \texttt{HTML5}, \texttt{CSS3} and \texttt{Javascript}.
 %
@@
 %
-The pivot table implementation was adapted using \texttt{PivotTable.js}\footnote{Developed by Nicolas Kruchten, \url{http://nicolas.kruchten.com/pivottable}}.
-%
-\texttt{Three.js}\footnote{Originally developed by Ricardo Cabello, \url{http://threejs.org}} allows GPU-accelerated data rendering using \texttt{WebGL}.
+The pivot table implementation was adapted using \texttt{PivotTable.js}\footnote{Developed by N. Kruchten, \href{http://nicolas.kruchten.com/pivottable}{\texttt{nicolas.kruchten.com/pivottable}}}.
+%
+\texttt{Three.js}\footnote{Originally developed by R. Cabello, \href{http://threejs.org}{\texttt{threejs.org}}} allows GPU-accelerated data rendering using \texttt{WebGL}.
 %
@@
 %
-Since our clustering algorithms are written in \texttt{MatLab}\footnote{Owned by The MathWorks, \url{http://mathworks.com}}, we had to access them using the \texttt{NodeJS} server.
+Since our clustering algorithms are written in \texttt{MatLab}\footnote{Owned by The MathWorks, \href{http://mathworks.com}{\texttt{mathworks.com}}}, we had to access them using the \texttt{NodeJS} server.
 %
@@
 %
-Determining risk factors in this area can lead to \cite{Fletcher2012}:
-\begin{itemize}
-	\item a better understanding of effects of preventive measures, such as occupational health and safety regulations,
-	\item prognostic features for diagnosis and treatment of lumbar back pain, and
-	\item determination of particularly affected risk groups.
-\end{itemize}
+Determining risk factors in this area can lead to particularly affected risk groups, prognostic features for diagnosis and treatment of lumbar back pain and a better understanding of effects of preventive measures, such as occupational health and safety regulations \cite{Fletcher2012}.
+% Determining risk factors in this area can lead to \cite{Fletcher2012}
+% \begin{itemize}
+% 	\item a better understanding of effects of preventive measures, such as occupational health and safety regulations,
+% 	\item prognostic features for diagnosis and treatment of lumbar back pain, and
+% 	\item determination of particularly affected risk groups.
+% \end{itemize}
 %
@@
 %
-Each feature is stored as an object containing: 
-\begin{itemize}
-	\item the data as array of values--categorical data and error codes are stored using IDs,
-	\item the data type (continuous, nominal, ordinal, dichotomous),
-	\item a detailed description of the feature, and
-	\item the data dictionary translating value or error IDs to values.
-\end{itemize}
+\add{Each feature is stored as an object containing the data as array it's data type, a detailed description and data dictionary translating value or error IDs to values.}
+% \begin{itemize}
+% 	\item the data as array of values--categorical data and error codes are stored using IDs,
+% 	\item the data type (continuous, nominal, ordinal, dichotomous),
+% 	\item a detailed description of the feature, and
+% 	\item the data dictionary translating value or error IDs to values.
+% \end{itemize}
+% Each feature is stored as an object containing:
+% \begin{itemize}
+% 	\item the data as array of values--categorical data and error codes are stored using IDs,
+% 	\item the data type (continuous, nominal, ordinal, dichotomous),
+% 	\item a detailed description of the feature, and
+% 	\item the data dictionary translating value or error IDs to values.
+% \end{itemize}
 %
@@
 %
-This approach aims to characterizes the ability of a system to explore data, discover knowledge, generate hypotheses and help formulating decisions.
+This approach aims to characterize the systems ability to explore data, discover knowledge, generate hypotheses and help formulating decisions.
 %
@@
 %Since one major reason for the focus on web technologies, no software needed to be exchanged.
-The experts accessed the prototype by simply entering the website link into their browser of choice, no exchange of software was required.
+%The experts accessed the prototype by simply entering the website link into their browser of choice, no exchange of software was required.
+The experts accessed the prototype by entering the website link into their browser of choice, no exchange of software was required.
 %Since our prototype ran on a local server, simply needed the homepage link to access the prototype.
@@
 %
-To make the user input comprehensible, we observed their monitor via screen sharing.
+User input was observed using screen sharing.
 %
@@
 \subsection{Case 1: Hypothesis-free Analysis} \label{Hypothesis-free analysis}
-Analyzing the data set without prior hypothesis requires a starting point giving an overview over the data first \cite{Shneiderman1996}.
+Analyzing the data set without prior hypothesis requires a starting point giving an overview over the data \cite{Shneiderman1996}.
 %
@@
 %
-The latter was chosen first by both domain experts.
+The latter was chosen first by both experts.
 % The latter was chosen first by the domain expert with radiological background.
@@
 %
-For one, age influences the lumbar spine shape, as seen in Figure~\ref{fig:application}~(c), while the differences between subjects with and without back pain are small.
+Age influences the lumbar spine shape, while the differences between subjects with and without back pain are small (Fig.~\ref{fig:application}~(c)).
 %
@@
 %
-As seen in Figure~\ref{fig:application}~(c), subjects reporting back pain are even the majority, which makes it difficult to extract parameters which reliably describe back pain.
+As seen in Figure~\ref{fig:application}~(c), subjects reporting back pain are the majority, which makes it difficult to extract parameters which reliably describe back pain.
 %
@@
 %
-\emph{Body height} was explained to be a confounder for \emph{gender}, since female subjects are smaller on average than male subjects.
+\emph{Body height} was explained as a confounder for \emph{gender}, since female subjects are smaller on average than male subjects.
 %
@@
 %
-KH emphasized the way the image data is included into information visualizations which comes much more natural to her due to her background in radiology.
+KH emphasized the way the image data are included into information visualizations which comes much more natural to her due to her background in radiology.
 %
@@
 %
-For the visual comparison itself, the radiologist proposed an abstraction of the representation into landmarks, such as centers of the vertebrae and cardinal points of the lumbar spine canal.
+For the visual comparison, KH proposed an abstraction of the representation into landmarks, such as centers of the vertebrae and cardinal points of the lumbar spine canal.
 
@@
 %The \emph{VDAR} approach to assess user thought processes using the think aloud technique worked well with the web-based approach for both implementing the prototype as well as getting feedback via voice over ip conference calls worked very well.
-\\\\
-Although it is an entirely different approach to their usual workflows it is to highlight that the epidemiologists liked browsing through the data set in the presented manner and to dig out new information about the data.
-%
-This underlines the accessibility of \emph{IVA} tools, which amplifies the natural curiosity of the domain experts in complex medical feature associations.
+%\\\\
+% \rem{Although it is an entirely different approach to their usual workflows it is to highlight that the epidemiologists liked browsing through the data set in the presented manner and to dig out new information about the data.
+% %
+% This underlines the accessibility of \emph{IVA} tools, which amplifies the natural curiosity of the domain experts in complex medical feature associations.}
 
 \section{Summary and Conclusion}
-% TODO: Future work Bulletliste zusammenkürzen
-In this paper, we presented an \emph{IVA} framework for the analysis of complex image-centric epidemiological data.
+We presented an \emph{IVA} framework for the analysis of complex image-centric epidemiological data.
 %
@@
 %
-\com{This needs to be adjusted to fit the new Application section}
-Findings from analyzing lumbar back pain using the \emph{IVA} approach range from deriving shape-based groups of subjects to detailed description of features potentially associated with the disease, such as waist circumference, thyroid nodules and blood fat values.
-%
-Other somatometric features, such as BMI, are not as influential as expected.
+%\com{This needs to be adjusted to fit the new Application section}
+Findings from analyzing lumbar back pain using the \emph{IVA} approach range from deriving shape-based groups of subjects to detailed description of features potentially associated with the disease, such as waist circumference, alcohol consumption and attentiveness disorder.
+%
+%\rem{Other somatometric features, such as BMI, are not as influential as expected.}
 %
@@
 %
-They allow a fast exchange between users and developers and employ many different devices.
+They allow a fast exchange between users and developers and employ different devices.
 %

### 2014-06-25 draft.tex
further text cuts

@@
 Cohort studies often include medical image data.
-% .
-Segmentation of these data enables an anatomical structure analysis and an analysis of possible correlations with other features.
-%
-Semi-automatic techniques are more promising but also challenging, since the employed modalities, such as magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
-%
-Analyzing spatial data with respect to other epidemiological factors requires techniques that reach beyond standard statistical methods.
+% 
+\add{The concurrent analysis of image data and non-spatial epidemiological factors requires techniques that reach beyond standard statistical methods.
+%
+For instance, segmentation of the image data is required for an analysis of anatomical structure and of possible correlations between this structure and epidemiological factors.
+%
+Semi-automatic segementation techniques are more promising but also challenging, since the employed modalities, such as magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.}
+% Segmentation of these data enables an anatomical structure analysis and an analysis of possible correlations with other features.
+% %
+% Semi-automatic techniques are more promising but also challenging, since the employed modalities, such as magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
+%
 
@@
 %
-The standard workflow lacks methods, which identify correlations that the epidemiologists did not consider.
-%
-Also, only a small subset of variables can be concurrently analyzed. %depending on the statistical measure.
+%The standard workflow lacks methods, which identify correlations that the epidemiologists did not consider.
+%
+%Also, only a small subset of variables can be concurrently analyzed. %depending on the statistical measure.
+The standard workflow lacks methods, which automatically identify correlations possibly buried deep in the data or overseen by the epidemiologists.
+%
+Also, only a small subset of factors can be concurrently analyzed.
 \\\\
@@
 % http://wiki.infowiss.net/Gulf_of_execution ?
-We applied our approach to a data set compiled to analyze diseases related to the lumbar spine and aim to determine features, which indicate pathological changes.
-%
-This data set comprises 127 features and 2 sequences of MRI data from 2,333 patients.
+We applied our approach to a data set compiled to analyze lower back pain and aim to determine features, which indicate pathological changes.
+%
+This data set comprises 127 features and 2 sequences of MRI data from 6.753 patients.
 %
@@
 %
-This aims to trigger a fast feedback loop between us and the epidemiologists.
+This aims to trigger a fast feedback loop. %between us and the epidemiologists.
 
@@
  \centering
- \includegraphics[width=3.0in]{figures/workflow_comparison}
+ \includegraphics[width=2.8in]{figures/workflow_comparison}
  \caption{
@@
 %
-A famous example for a confounder is the association between shoe size and mortality, where it can be observed that people with larger shoe size have a smaller life expectation.
-%than people with small shoe size.
+A famous example is the association between shoe size and mortality, where it can be observed that people with larger shoe size have a smaller life expectation. %than people with small shoe size.
 %
@@
 %
-Graphic data representation is largely used to present results rather than gaining insight.
+%Graphic data representation is largely used to present results rather than gaining insight.
+Epidemiologists employ static graphical data representations primarily at the very end of an analysis session for presenting results or observing trends in the data.
 
@@
 %
-While they offer a single spatial view for visualizing image-based information of one subgroup of the cohort, we provide multiple, small views showing the information of subgroups and its respective deviation from the entire cohort.}
+While they offer a single spatial view for visualizing image-based information of one subgroup of the cohort, we provide multiple views showing the information of subgroups and their respective deviation from the entire cohort.}
 %
@@
 %
-\add{While their goal to identify possible risk factors using socio-economic and health data are similar to ours, they focus on iteratively refining defined hypothesis and also focus on geographical data.
-%
-We employ the use of small multiples in this publication for incorporating heterogenous data types for comparability.}
+\add{While their goal to identify possible risk factors using socio-economic and health data are similar to ours, they focus on iteratively refining defined hypothesis and on geographical data.
+%
+We employ the use of small multiples for incorporating heterogenous data types for comparability.}
 %
@@
 %
-\add{Applying our data sets to this technique yielded a cluttered shape space due to the many subjects.
+\add{Applying this technique to our data sets yielded a cluttered shape space due to the many subjects.
 %
@@
 %
-When applying clustering techniques to the non-image data it was found that \texttt{k-Prototypes} and \texttt{DBSCAN} are appropriate in the epidemiological context, but is strongly dependent on the chosen variables and distance measure \cite{Klemm2014BVM}.
+When applying clustering techniques to the non-image data it was found that \texttt{k-Prototypes} and \texttt{DBSCAN} are appropriate, but is strongly dependent on the chosen variables and distance measure \cite{Klemm2014BVM}.
 %
@@
 %
-\add{We combine the idea of suggesting potentially interesting features by adding them to the \emph{IVA}-workflow in order to amplify hypothesis generation.}
+%\add{We combine the idea of suggesting potentially interesting features by adding them to the \emph{IVA}-workflow in order to amplify hypothesis generation.}
 % *********************** / Our own Work ***********************
@@
 % Interactive Visual Analysis - Steffens Paper
-The strength of the \emph{IVA} approach described in the next section is its versatility with respect to the application field \cite{Konyha2009}.
+The strength of the \emph{IVA} approach is its versatility with respect to the application field \cite{Konyha2009}.
 %
@@
 %Related work focused on epidemiological data often deals with time-dependent data, hypothesis generation by guidance through parameter suggestions is seldom.
-\add{The uniqueness of the \emph{IVA} workflow applied to image-based cohort study data compared to the discussed work are threefold.
-%
-(1) We incorporate 3D models abstracting shape information (e.g.
-for the spine) fused with non-image data visualizations, allowing to analyze local physiological changes related to non-image parameters.
-%
-(2) Epidemiologists already have an well established workflow for statistically validating hypothesis. 
-%
-Therefore, we focus on hypothesis generation by discovering new relationships associated with shape information.
+\add{%The uniqueness of the \emph{IVA} workflow applied to image-based cohort study data compared to the discussed work is threefold.
+The uniqueness of our workflow compared to the discussed work is threefold.
+%
+(1) We incorporate 3D models abstracting shape information fused with non-image data visualizations, allowing to analyze local physiological changes related to non-image parameters.
+%
+(2) We focus on hypothesis generation by discovering new relationships associated with shape information, as epidemiologists already have an well established workflow for statistically validating hypothesis.
 %
@@
 \begin{itemize} 
-	\item \emph{physical views} \cite{Oeltze2013}, such as direct volume rendering, show information in the context of the spatio-temporal observation space \cite{Oeltze2007}, while
+	\item \emph{physical views} \cite{Oeltze2013}, such as volume rendering, show information in the context of the spatio-temporal observation space \cite{Oeltze2007}, while
 	\item \emph{attribute views}, such as scatter plots and parallel coordinates, show relationships between multiple data attributes.
@@
  %
- (Left) Screenshot from the front-end, which is divided as follows: (a) The sidebar containing all features as well as the groups defined in the analysis process; (b) the canvas area where features can be added via drag and drop and the visualization is chosen automatically according to the data type; (c) the interactive pivot table showing the exact numbers for each displayed feature combination; \add{(d) Buttons to open panes containing the contingency matrix, contingency pane and pivot table}.
+ (Left) Screenshot from the front-end, which is divided as follows: (a) The sidebar containing all features as well as the groups defined in the analysis process; (b) the canvas area where features can be added via drag and drop and the visualization is chosen automatically according to the data type; (c) the interactive pivot table showing the exact numbers for each displayed feature combination; \add{(d) buttons to open panes containing the contingency matrix, contingency pane and pivot table}.
  %
@@
  %
- Subjects between 153.5--170~cm body size are more affected by thyroid disorder (box plot) and are mostly female (mosaic plot).}
+ Subjects between 153.5-170~cm body size are more affected by thyroid disorder (box plot) and are mostly female (mosaic plot).}
  %
@@
  \includegraphics[width=3.5in]{figures/similarity_matrix}
- \caption{Adjacency matrix of 129 features (127 data set variables, 2 cluster results) with a grand total of 16,641 combinations.
+ \caption{Adjacency matrix of 129 features (127 data set variables, 2 cluster results) showing 16,641 combinations.
  %
@@
 %
-We set the number of groups to five, to allow for contingency assessment.
+In accordance with epidemiological publications, we set the number of groups to five, to allow for contingency assessment.
 
@@
  %
- Note the high amount of this subject group relative to the total count in the third group compared. The last group only contains four outlier-subjects.
+ Note the high amount of this subject group relative to the total count in the third group. The last group contains four outliers.
  %
@@
 Great potential is also seen in communicating insights efficiently using the presented visualizations \com{(1.08)}.
-%
-\paragraph{Multivariate analysis is most important for hypothesis generation.}
+
+\textbf{Multivariate analysis is most important for hypothesis generation.}
 Both experts emphasized the potential of the \emph{multivariate analysis} capabilities of the adjacency matrix for getting insight into a large amount of features simultaneously.
@@
 Creating adjacency matrices for subgroups, such as different age bins can help to characterize the aging process by deriving age-specific risk factors.
-
+%
 \emph{Multivariate analysis} can also be improved by more ways of brushing the data as well as creating subgroups for comparison as a result of the hypothesis-driven analysis case.
@@
 %
-%\paragraph{Hypothesis driven analysis would benefit from more advanced grouping.}
+%\textbf{Hypothesis driven analysis would benefit from more advanced grouping.}
 %While the existing brushing and linking techniques already allow the comparison of image data 
@@
 
-\paragraph{Segmentation quality is crucial.}
+\textbf{Segmentation quality is crucial.}
 The radiologist pointed out the unusual strong similarity of the L3 vertebrae throughout the population.
@@
 A seconds explanation is the use of the L3 vertebra is initialization point of the lumbar spine model \com{(00.54)}.
-
+%
 The experts also emphasized that associations related to shape strongly depend on the segmentation quality.
@@
 
-\paragraph{Usage of different categorizations depending on expected outcome.}
+\textbf{Usage of different categorizations depending on expected outcome.}
 When categorizing numerical variables into equal groups it is possible to create sparse categories due to outliers, for example when analyzing body weight.
@@
 Balanced categories are created in the second mode, which uses quartiles or quintiles to set borders between categories.
-\paragraph{Web technologies are well suited for rapid feedback.}
+%
+
+\textbf{Web technologies are well suited for rapid feedback.}
 The web-based approach for both implementing the prototype as well as getting feedback via voice over ip conference calls worked very well.

### 2014-06-26 draft.tex
Completed Coverletter

@@
 %Also, only a small subset of variables can be concurrently analyzed. %depending on the statistical measure.
-The standard workflow lacks methods, which automatically identify correlations possibly buried deep in the data or overseen by the epidemiologists.
+The standard workflow lacks methods for automatically identifying correlations possibly buried deep in the data or overseen by the expert.
 %
 Also, only a small subset of factors can be concurrently analyzed.
-\\\\
+%
+
 We propose an Interactive Visual Analysis (\emph{IVA}) approach \cite{Thomas2005} for the combined analysis of image and non-image data.

### 2014-06-26 draft.tex
Removed all 'add' commands

@@
 % ----------- /OLD ABSTRACT ---------
-\add{%Large-scale longitudinal epidemiological studies investigate groups of people with common characteristics or experiences (a cohort) to characterize health and identify disease-specific risk factors.
+%Large-scale longitudinal epidemiological studies investigate groups of people with common characteristics or experiences (a cohort) to characterize health and identify disease-specific risk factors.
 %
@@
 %
-We demonstrate the \emph{IVA} approach by validating and generating hypotheses related to lower back pain as part of a qualitative evaluation.}%and the aging process of the lumbar spine.}
+We demonstrate the \emph{IVA} approach by validating and generating hypotheses related to lower back pain as part of a qualitative evaluation. %and the aging process of the lumbar spine.}
 } % end of abstract
@@
 % 
-\add{The concurrent analysis of image data and non-spatial epidemiological factors requires techniques that reach beyond standard statistical methods.
+The concurrent analysis of image data and non-spatial epidemiological factors requires techniques that reach beyond standard statistical methods.
 %
@@
 %
-Semi-automatic segementation techniques are more promising but also challenging, since the employed modalities, such as magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.}
+Semi-automatic segementation techniques are more promising but also challenging, since the employed modalities, such as magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
 % Segmentation of these data enables an anatomical structure analysis and an analysis of possible correlations with other features.
@@
 Our contributions are:
-\begin{itemize}
-	\item an \add{IVA} workflow for cohort study data to allow both, hypothesis-driven analysis and hypothesis generation--based on a characterization of the standard epidemiological workflow,
+\begin{itemize} \itemsep0.1em
+	\item an IVA workflow for cohort study data to allow both, hypothesis-driven analysis and hypothesis generation--based on a characterization of the standard epidemiological workflow,
 	\item visualization techniques, which incorporate both information visualization and 3D rendering of organ shapes as well as combining them with epidemiological graphics and key figures,
@@
 %
-This data set comprises 127 features and 2 sequences of MRI data from 6.753 patients.
+This data set comprises 127 features and 2 sequences of MRI data from 6,753 patients.
 %
@@
 %
-\add{Data analysis is usually carried out by calculating correlations,
+Data analysis is usually carried out by calculating correlations,
 which is challenging due to the data type heterogeneity.
@@
 %Also, correlation can both be of epidemiological interest but
-}
 
@@
 
-Our work is closest to that of Steenwijk \cite{Steenwijk2010}, Turkey \cite{Turkay2013}, and Angelelli \cite{Angelelli2014} and colleagues\add{, who employ multiple coordinated view systems for the analysis of cohort study.}
+Our work is closest to that of Steenwijk \cite{Steenwijk2010}, Turkey \cite{Turkay2013}, and Angelelli \cite{Angelelli2014} and colleagues, who employ multiple coordinated view systems for the analysis of cohort study.
 %
@@
 %
-\add{While we use a similar approach when analyzing non-image data, our process also includes \emph{overview visualizations} and statistical suggestions of potentially interesting features.}
+While we use a similar approach when analyzing non-image data, our process also includes \emph{overview visualizations} and statistical suggestions of potentially interesting features.
 %
@@
 %
-\add{Our approach uses information extracted from the segmented image data (such as 3D meshes) and variable associations with non-image epidemiological factors.}
+Our approach uses information extracted from the segmented image data (such as 3D meshes) and variable associations with non-image epidemiological factors.
 
@@
 %
-\add{Our integration of image data into the analysis is similar to the work of Steenwijk \cite{Steenwijk2010} and Angelelli \cite{Angelelli2014} and colleagues.
-%
-While they offer a single spatial view for visualizing image-based information of one subgroup of the cohort, we provide multiple views showing the information of subgroups and their respective deviation from the entire cohort.}
+Our integration of image data into the analysis is similar to the work of Steenwijk \cite{Steenwijk2010} and Angelelli \cite{Angelelli2014} and colleagues.
+%
+While they offer a single spatial view for visualizing image-based information of one subgroup of the cohort, we provide multiple views showing the information of subgroups and their respective deviation from the entire cohort.
 %
@@
 %
-\add{Due to the short paper length, detail is missing on the data type and their algorithm of identifying similar subjects works or what and if they employ statistical measures.
-%
-We employ the idea of adding features via drag and drop into a canvas area.}
+Due to the short paper length, detail is missing on the data type and their algorithm of identifying similar subjects works or what and if they employ statistical measures.
+%
+We employ the idea of adding features via drag and drop into a canvas area.
 
@@
 %
-\add{The resulting matrix provides a \emph{overview visualization}, but requires a lot of screen space for many features (127 in our application scenario).}
+The resulting matrix provides a \emph{overview visualization}, but requires a lot of screen space for many features (127 in our application scenario).
 %
@@
 %
-\add{We incorporate the idea of adaptive type-dependent visualizations.}
+We incorporate the idea of adaptive type-dependent visualizations.
 %\add{The inspiration on the chosen visualization techniques stems from this publication.}
@@
 %
-\add{While their goal to identify possible risk factors using socio-economic and health data are similar to ours, they focus on iteratively refining defined hypothesis and on geographical data.
-%
-We employ the use of small multiples for incorporating heterogenous data types for comparability.}
+While their goal to identify possible risk factors using socio-economic and health data are similar to ours, they focus on iteratively refining defined hypothesis and on geographical data.
+%
+We employ the use of small multiples for incorporating heterogenous data types for comparability.
 %
@@
 %
-\add{%The connection of time series plots with histograms and heat maps allows for fast insight into complex relationships and 
-While the works shows, how different visualization modalities provide insight into these data sets, it focuses on the time aspect, which is not present for our data.}
+%The connection of time series plots with histograms and heat maps allows for fast insight into complex relationships and 
+While the works shows, how different visualization modalities provide insight into these data sets, it focuses on the time aspect, which is not present for our data.
 %
@@
 %
-\add{Applying this technique to our data sets yielded a cluttered shape space due to the many subjects.
-%
-The data needs to abstracted or summarized in order to work in this context.}
+Applying this technique to our data sets yielded a cluttered shape space due to the many subjects.
+%
+The data needs to abstracted or summarized in order to work in this context.
 %
@@
 %
-\add{This requires a geographic colocation of associated features.}
+This requires a geographic colocation of associated features.
 %
@@
 %Related work focused on epidemiological data often deals with time-dependent data, hypothesis generation by guidance through parameter suggestions is seldom.
-\add{%The uniqueness of the \emph{IVA} workflow applied to image-based cohort study data compared to the discussed work is threefold.
+%The uniqueness of the \emph{IVA} workflow applied to image-based cohort study data compared to the discussed work is threefold.
 The uniqueness of our workflow compared to the discussed work is threefold.
@@
 (3) Overview visualizations using statistical abstractions aim to provide an unbiased feature relationship assessment.
-}
 
@@
 Two kinds of views are employed to inspect the data:
-\begin{itemize} 
-	\item \emph{physical views} \cite{Oeltze2013}, such as volume rendering, show information in the context of the spatio-temporal observation space \cite{Oeltze2007}, while
+\begin{itemize} \itemsep0.2em
+	\item \emph{physical views} \cite{Oeltze2013}, e.g. volume rendering, show information in the context of the spatio-temporal observation space \cite{Oeltze2007}, while
 	\item \emph{attribute views}, such as scatter plots and parallel coordinates, show relationships between multiple data attributes.
@@
  %
- (Left) Screenshot from the front-end, which is divided as follows: (a) The sidebar containing all features as well as the groups defined in the analysis process; (b) the canvas area where features can be added via drag and drop and the visualization is chosen automatically according to the data type; (c) the interactive pivot table showing the exact numbers for each displayed feature combination; \add{(d) buttons to open panes containing the contingency matrix, contingency pane and pivot table}.
+ (Left) Screenshot from the front-end, which is divided as follows: (a) The sidebar containing all features as well as the groups defined in the analysis process; (b) the canvas area where features can be added via drag and drop and the visualization is chosen automatically according to the data type; (c) the interactive pivot table showing the exact numbers for each displayed feature combination; (d) buttons to open panes containing the contingency matrix, contingency pane and pivot table.
  %
@@
  %
- \add{In a prior step, the user selected all subjects with diagnosed thyroid disorder.
+ In a prior step, the user selected all subjects with diagnosed thyroid disorder.
  %
@@
  %
- Subjects between 153.5-170~cm body size are more affected by thyroid disorder (box plot) and are mostly female (mosaic plot).}
- %
- \add{Distance to the mean mesh of subjects with thyroid disorder is encoded as red for x axis, blue to y axis and green to z axis.}
+ Subjects between 153.5-170~cm body size are more affected by thyroid disorder (box plot) and are mostly female (mosaic plot).
+ %
+ Distance to the mean mesh of subjects with thyroid disorder is encoded as red for x axis, blue to y axis and green to z axis.
  %The lumbar spine models encode distance to the mean model using  \emph{color} 
@@
 %
-\add{Hypothesis generation with focus on image data starts with a shape-based clustering or an \emph{overview visualization} of all features.}
+Hypothesis generation with focus on image data starts with a shape-based clustering or an \emph{overview visualization} of all features.
 %
@@
 %
-Selecting features also triggers a \emph{multivariate analysis} using contingency values (described in the following section) to highlight \add{associated} features.  %\rem{with selected features.}
+Selecting features also triggers a \emph{multivariate analysis} using contingency values (described in the following section) to highlight associated features.  %\rem{with selected features.}
 %
@@
 %Early it became clear that we have to rely a lot on online communication with our clinical partners due to their busy schedule and our large spatial distance towards each other.
-\add{Early it became clear that we have to rely a lot on online communication due to the busy schedule and our large spatial distance towards each other.}
+Early it became clear that we have to rely a lot on online communication due to the busy schedule and our large spatial distance towards each other.
 %
@@
 We divide the workspace into four major parts, as illustrated in Figure~\ref{fig:visualization} and \ref{fig:similarity}.
-\begin{itemize}
+\begin{itemize} \itemsep0.2em
 		\item The \emph{sidebar}, which contains all epidemiological features.
-As cluster results group features like categorical features and are part of the sidebar as well (\add{Fig.~\ref{fig:visualization} (a)}).
-	%
-	\item The \emph{canvas} holding all visualizations. Elements can be added, arranged, resized and removed freely (\add{Fig.~\ref{fig:visualization} (b))}.
+As cluster results group features like categorical features and are part of the sidebar as well (Fig.~\ref{fig:visualization} (a)).
+	%
+	\item The \emph{canvas} holding all visualizations. Elements can be added, arranged, resized and removed freely (Fig.~\ref{fig:visualization} (b)).
 	%
 	\item The interactive \emph{pivot table} gives detailed numerical information of the features in the canvas view.
-This view on the data is familiar to epidemiologists (\add{Fig.~\ref{fig:visualization} (c)}).
-	%
-	\item The \emph{contingency view} depicts relations for features in the canvas in an adjacency matrix (\add{Fig.~\ref{fig:similarity}}) and a \emph{contingency list}.
+This view on the data is familiar to epidemiologists (Fig.~\ref{fig:visualization} (c)).
+	%
+	\item The \emph{contingency view} depicts relations for features in the canvas in an adjacency matrix (Fig.~\ref{fig:similarity}) and a \emph{contingency list}.
 \end{itemize}
@@
 \paragraph{Sidebar.}
-The \emph{sidebar} is the only visible element when the system is opened (Fig.~\ref{fig:application} (a)).
-%
-It categorizes all features into different types, such as somatometric (measurements of the human body dimensions), disease- or lifestyle-related, pain indicators and laboratory data.
+Only the \emph{sidebar} is visible at system start.
+%The \emph{sidebar} is the only visible element when the system is opened (Fig.~\ref{fig:visualization} (a)).
+%
+It categorizes all features into different types, such as somatometric (measurements of the human body dimensions), disease- or lifestyle-related, pain indicators and laboratory data (Fig.~\ref{fig:visualization} (a)).
 %
@@
 In addition to the obvious advantages of web technologies, the following aspects are crucial for our work:
-\begin{itemize}
+\begin{itemize} %\itemsep0.2em
 %	\item No additional software needs to be installed, most people use a decent state-of-the-art web browser, even on mobile devices.
 	\item The client-server structure allows for employing heavy computation on a server machine and transferring results to the client.
-	\item \add{Disk-space demanding image data remains on the server and elements can be transferred on demand.}
+	\item Disk-space demanding image data remains on the server and elements can be transferred on demand.
 	%\item Since image data for thousands of subjects requires hundreds of gigabytes disk space, it can remain safely on the server and elements can be transferred on demand.
 	%
-	High confidentiality standards of the data are met by \add{password protecting the access.}%restricting access via an account system.
+	High confidentiality standards of the data are met by password protecting the access.%restricting access via an account system.
 	\item Recent developments in \emph{WebGL} applications running in browsers with near-native performance result in many open source libraries, which are well documented and driven by active communities.
@@
 %
-\add{
-There are 127 features describing diagnosed diseases, lifestyle factors, women-specific factors, pain indicators, laboratory values and somatometric features for 6.753 subjects (4.420 from \texttt{SHIP-Trend-0} and 2.333 from \texttt{SHIP-2}).
+There are 127 features describing diagnosed diseases, lifestyle factors, women-specific factors, pain indicators, laboratory values and somatometric features for 6,753 subjects (4,420 from \texttt{SHIP-Trend-0} and 2,333 from \texttt{SHIP-2}).
 %
 Since data acquisition protocols between these two cohorts are identical, the features between the two cohorts are comparable.
-}
-%
-\add{The data contains 30 metric, 7 nominal, 29 ordinal and 62 dichotomous epidemiological features.
+%
+The data contains 30 metric, 7 nominal, 29 ordinal and 62 dichotomous epidemiological features.
 %
@@
 %Another challenge the 
-}
 %
@@
 %
-\add{The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
+The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
 %(676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
-}
 %The spine protocol consisted of sagittal T1- and T2-weighted turbo-spin-echo sequences with rather large slice distance only sufficient for characterizing the lumbar spine \cite{Hegenscheid2013}.
@@
 %
-\add{
 Since it lacks export methods for data dictionaries, we used \texttt{SPSS} to export our data to the 'SAS v9+' format, which saves the data labels, and exported the data values as non-labeled CSV.
@@
 The data types had to be transferred manually.
-}
-%
-\add{Each feature is stored as an object containing the data as array it's data type, a detailed description and data dictionary translating value or error IDs to values.}
+%
+Each feature is stored as an object containing the data as array it's data type, a detailed description and data dictionary translating value or error IDs to values.
 % \begin{itemize}
@@
 %
-\add{A number of improvements is left open for future work, such as shape brushing methods to intuitively query subjects using image data or the inclusion of more statistical methods and views that are familiar to the epidemiologists (odds ratios, box plots).}
+A number of improvements is left open for future work, such as shape brushing methods to intuitively query subjects using image data or the inclusion of more statistical methods and views that are familiar to the epidemiologists (odds ratios, box plots).
 

### 2014-06-26 draft.tex
First version with proper length

@@
 %
-The cohort consists of several thousands of people who are randomly selected to avoid any bias.
+The cohort consists of several thousands of people, randomly selected to avoid any bias.
 %
@@
 %
-A large size of the cohort is therefore essential to investigate differences between healthy and diseased people.
+A large cohort size is essential to investigate differences between healthy and diseased people.
 %
@@
 %
-Semi-automatic segementation techniques are more promising but also challenging, since the employed modalities, such as magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
+Semi-automatic segmentation techniques are promising but also challenging, since the employed modalities, such as magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
 % Segmentation of these data enables an anatomical structure analysis and an analysis of possible correlations with other features.
@@
 \begin{itemize} \itemsep0.1em
-	\item an IVA workflow for cohort study data to allow both, hypothesis-driven analysis and hypothesis generation--based on a characterization of the standard epidemiological workflow,
+	\item an IVA workflow for cohort study data to allow both, hypothesis-driven analysis and hypothesis generation,%--based on a characterization of the standard epidemiological workflow,
 	\item visualization techniques, which incorporate both information visualization and 3D rendering of organ shapes as well as combining them with epidemiological graphics and key figures,
@@
 %
-This data set comprises 127 features and 2 sequences of MRI data from 6,753 patients.
-%
-We implemented the presented methods using modern web technologies, such as \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}, making them easily accessible for the domain experts to enable a fast feedback loop.
-%
-This aims to trigger a fast feedback loop. %between us and the epidemiologists.
+This data set comprises 127 features and 2 sequences of MRI data from 6,753 subjects.
+%
+We implemented the presented methods using modern web technologies to make them easily accessible for the domain experts to enable a fast feedback loop.
+%We implemented the presented methods using modern web technologies, such as \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}, making them easily accessible for the domain experts to enable a fast feedback loop.
+%
+%This aims to trigger a fast feedback loop. %between us and the epidemiologists.
 
@@
 %
-\begin{enumerate}
+\begin{enumerate}\itemsep0.1em
 	\item A hypothesis is derived from observations made by clinicians in their daily routine.
@@
 \end{enumerate}
-The workflow is shown in Figure~\ref{fig:WorkflowComparison} (a) and serves as orientation for our approach although it does not consider image data.
+The workflow is shown in Figure~\ref{fig:WorkflowComparison} (a) and serves as orientation for our approach. %although it does not consider image data.
 %
@@
 %
-Statistical correlations are prone to \emph{confounding}, meaning that the association of two features is influenced by a third feature, which needs to be isolated.
-%
-A famous example is the association between shoe size and mortality, where it can be observed that people with larger shoe size have a smaller life expectation. %than people with small shoe size.
-%
-The shoe size is actually associated with gender, where women have smaller feet and a longer life expectation.
 %
@@
 %
-Parameter correlation can also be associated with confounding, which can not be automatically predicted, it has to be judged by a domain expert.
+Parameter correlation can also be associated with confounding, which can not be automatically predicted. 
+%
+It has to be judged by a domain expert.
 %
@@
 Too few data samples may distort the real underlying distributions.
+%
+Statistical correlations are prone to \emph{confounding}, meaning that the association of two features is influenced by a third feature, which needs to be isolated.
+%
+A famous example is the association between shoe size and mortality, where it can be observed that people with larger shoe size have a smaller life expectation. %than people with small shoe size.
+%
+The shoe size is actually associated with gender, where women have smaller feet and a longer life expectation.
 %Also, correlation can both be of epidemiological interest but
@@
 %
-This often yields image resolutions inferior to those of clinical day-to-day practice, which makes their analysis more challenging.
+This often yields image resolutions inferior to those of clinical practice.%, which makes their analysis more challenging.
 %
@@
 %
-Segmentation masks representing the voxels of an anatomical structure would be ideal, since many different key figures, e.g., volume, largest diameter or aspect ratio, can be derived from them.
+Segmentation masks representing the voxels of an anatomical structure would be ideal, since key figures, e.g., volume, largest diameter or aspect ratio, can be derived from them.
 %
@@
 
-Angelelli et al. \cite{Angelelli2014} focus on a suitable data organization for the interactive visual analysis of heterogeneous cohort study data.
+Angelelli et al. \cite{Angelelli2014} focus on a data organization for the interactive visual analysis of heterogeneous cohort study data.
 %
@@
 %
-The resulting matrix provides a \emph{overview visualization}, but requires a lot of screen space for many features (127 in our application scenario).
+The resulting matrix provides an \emph{overview visualization}, but requires a lot of screen space for many features (127 in our application scenario).
 %
@@
 %
-While their goal to identify possible risk factors using socio-economic and health data are similar to ours, they focus on iteratively refining defined hypothesis and on geographical data.
+While their goal to identify possible risk factors using socio-economic and health data similar to ours, they focus on iteratively refining defined hypothesis and on geographical data.
 %
@@
 %The connection of time series plots with histograms and heat maps allows for fast insight into complex relationships and 
-While the works shows, how different visualization modalities provide insight into these data sets, it focuses on the time aspect, which is not present for our data.
+While the works shows, how different visualization techniques provide insight into these data sets, it focuses on the time aspect, which is not present for our data.
 %
@@
 %
-Applying this technique to our data sets yielded a cluttered shape space due to the many subjects.
-%
-The data needs to abstracted or summarized in order to work in this context.
+Applying this technique to our data yielded a cluttered shape space due to the high subject count.
+%
+The data needs to be abstracted to work in this context.
 %
@@
 %
-This work focuses on incorporating these derived data as new features of the overall data set, making it possible to include it into the hypothesis validation and generation process.
+This work focuses on incorporating these derived data as new features, making it possible to include it into the hypothesis validation and generation process.
 %
@@
 %
-The VA-process is also reflected in our \emph{IVA}-Loop.
+The VA-process is also reflected in our \emph{IVA} loop.
 % --- PARAGRAPH ---
@@
 %
-(2) We focus on hypothesis generation by discovering new relationships associated with shape information, as epidemiologists already have an well established workflow for statistically validating hypothesis.
+(2) We focus on hypothesis generation by discovering new relationships associated with shape information.
+%, as epidemiologists already have an well established workflow for statistically validating hypothesis.
 %
@@
 \\\\
-\emph{IVA} tries to illuminate the black box by making the domain experts part of an iterative feature selection process (see Fig.~\ref{fig:WorkflowComparison} (b)).
+\emph{IVA} tries to illuminate the black box by making the domain experts part of an iterative feature selection process (see Fig.~\ref{fig:WorkflowComparison}~b).
 %\emph{IVA} tries to illuminate the black box by making the domain experts part of an iterative feature selection process, which is shown in Figure~\ref{fig:WorkflowComparison} (b) as part of the epidemiological workflow.
@@
 %
-Instead we employ a abstract model and consider the subjects as living in a joint image space where each of them is represented by a segmented organ or structure.
+Instead we employ an abstract model and consider the subjects as living in a joint image space where each of them is represented by a segmented organ or structure.
 %
@@
 
-\paragraph{Local Investigation.} Refers to the inspection of dependent variables with respect to subsets of the image or shape space.
+\paragraph{Local Investigation} refers to the inspection of dependent variables with respect to subsets of the image or shape space.
 %
@@
 
-\paragraph{Feature Localization.} Refers to the search for structures in the image or shape space with a defined characteristic.
+\paragraph{Feature Localization} refers to the search for structures in the image or shape space with a defined characteristic.
 %
@@
 
-\paragraph{Multivariate Analysis.} Refers to an investigation of multi-variate properties of the dependent data by specifying a feature in one attribute view while analyzing the value distribution with respect to other variables in other attribute views.
+\paragraph{Multivariate Analysis} refers to an investigation of multi-variate properties of the dependent data by specifying a feature in one attribute view while analyzing the value distribution with respect to other variables in other attribute views.
 %
@@
 %
-Exporting the data in the respective tool to a CSV file and then converting it to file types that are easily manageable, such as JSON, makes it readable for modern programming languages.
+Exporting the data in the respective tool to a \texttt{CSV} file and then converting it to file types that are easily manageable, such as \texttt{JSON}, makes it readable for modern programming languages.
 %
@@
 %
-If a segmentation yields only binary masks separating the structures, algorithms such as \emph{Growing and Adaptive Shapes} can be applied creating a surface grid where each point is comparable throughout the population \cite{Ferrarini2007}.
+If a segmentation yields only binary masks, algorithms such as \emph{Growing and Adaptive Shapes} can be applied for creating a surface grid where each point is comparable throughout the population \cite{Ferrarini2007}.
 %
@@
 %
-The suitability of an interaction and visualization technique for epidemiological data depends on its ability to compare multiple data features at once while highlighting new associations.
-%
-The methods have to reflect the routines that epidemiologists take in order to be incorporated into their research.
+The suitability of visualization techniques for epidemiological data depends on its ability to compare multiple data features while highlighting associations.
+%
+%The methods have to reflect the routines that epidemiologists take in order to be incorporated into their research.
 %
@@
 %
-In the following sections we present the different parts of our proposed \emph{IVA} system for image-based cohort study analysis.
+In the following sections we present the different parts of system.
 % \subsection{System Structure and Design} \label{Structure and Workflow}
@@
 %Early it became clear that we have to rely a lot on online communication with our clinical partners due to their busy schedule and our large spatial distance towards each other.
-Early it became clear that we have to rely a lot on online communication due to the busy schedule and our large spatial distance towards each other.
-%
-Hence we build our system using modern web technologies, described in detail in Section~\ref{implementation}.
-%
-By running the prototypes on server machines, software exchange became as easy as sharing a website link, giving us the opportunity to include the clinical experts easily in the development process.
+Early it became clear that we have to rely a lot on online communication due to the large spatial distance towards each other.
+%
+Hence we build our system using web technologies.%, described in detail in Section~\ref{implementation}.
+%
+By running the prototypes on server machines, software exchange became as easy as sharing a website link, giving us the opportunity to include the clinical experts in the development process with little effort.
 %
@@
 		\item The \emph{sidebar}, which contains all epidemiological features.
-As cluster results group features like categorical features and are part of the sidebar as well (Fig.~\ref{fig:visualization} (a)).
-	%
-	\item The \emph{canvas} holding all visualizations. Elements can be added, arranged, resized and removed freely (Fig.~\ref{fig:visualization} (b)).
+As cluster results group features like categorical features and are part of the sidebar as well (Fig.~\ref{fig:visualization}~a).
+	%
+	\item The \emph{canvas} holding all visualizations. Elements can be added, arranged, resized and removed freely (Fig.~\ref{fig:visualization}~b).
 	%
 	\item The interactive \emph{pivot table} gives detailed numerical information of the features in the canvas view.
-This view on the data is familiar to epidemiologists (Fig.~\ref{fig:visualization} (c)).
+This view on the data is familiar to epidemiologists (Fig.~\ref{fig:visualization}~c).
 	%
@@
 %
-It categorizes all features into different types, such as somatometric (measurements of the human body dimensions), disease- or lifestyle-related, pain indicators and laboratory data (Fig.~\ref{fig:visualization} (a)).
+It categorizes all features into different types, such as somatometric (measurements of the human body dimensions), disease- or lifestyle-related, pain indicators and laboratory data (Fig.~\ref{fig:visualization}~a).
 %
@@
 %
-Groups are treated like features since they act the same by dividing the subject space into labeled categories.
+\add{Groups are treated like dichotomous features.} %since they act the same by dividing the subject space into labeled categories.
 %
@@
 %
-Following Tufte's concept of \emph{small multiples} \cite{Tufte1983}, information derived from the medical image data are directly incorporated into the plot by including color-coded mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
+Following Tufte's concept of \emph{small multiples} \cite{Tufte1983}, information derived from the medical image data are directly incorporated into the plot by including color-coded mean shapes for each manifestation (Figure~\ref{fig:visualization}~b).
 %
@@
 %
-Dropping a feature on an existing plot adapts the visualization dynamically to allow for comparison (Fig.~\ref{fig:visualization} (right)).
+Dropping a feature on an existing plot adapts the visualization dynamically to allow for comparison (Fig.~\ref{fig:visualization}~right).
 % --- PARAGRAPH ---
@@
 %This allows to highlight distances between subjects
-The share of subjects of this subgroup is linked to all other views (Fig.~\ref{fig:visualization} (left)).
+The share of subjects of this subgroup is linked to all other views (Fig.~\ref{fig:visualization}~left).
 %
@@
 %
-These tables clearly convey the subject count in each group (see Figure~\ref{fig:visualization} (c)).
+These tables clearly convey the subject count in each group (see Figure~\ref{fig:visualization}~c).
 %
@@
 \end{itemize}
-These advantages do not come without drawbacks.
-%
-Sophisticated libraries/languages, such as the \texttt{Visualization Toolkit} (\texttt{VTK})\footnote{Developed by Kitware Inc; \href{http://vtk.org}{\texttt{vtk.org}}} or \texttt{R}\footnote{Open Source; \href{http://r-project.org}{\texttt{r-project.org}}} for statistics, are either not available at all or only accessible through complex client-server systems.
+%These advantages do not come without drawbacks.
+%
+However, sophisticated libraries/languages, such as the \texttt{Visualization Toolkit}\footnote{Developed by Kitware Inc; \href{http://vtk.org}{\texttt{vtk.org}}} or \texttt{R}\footnote{Open Source; \href{http://r-project.org}{\texttt{r-project.org}}} for statistics, are either not available at all or only accessible through complex client-server systems.
 %
@@
 %
-
-The back-end is written using \texttt{NodeJS}\footnote{Developed by Joyent Inc, \href{http://nodejs.org}{\texttt{nodejs.org}}}, which is based on the Google V8 Javascript runtime environment.
-%
-Due to its event-driven non-blocking I/O model it is fast and does not freeze in case of heavy workload such as mesh processing.
-\\\\
-Non-image data for all subjects including the data dictionary is stored in a JSON file on the server.
-%
-Image data are available as raw DICOM files.
+%
+The back-end is realized using \texttt{NodeJS}\footnote{Developed by Joyent Inc, \href{http://nodejs.org}{\texttt{nodejs.org}}}, which is based on the Google V8 Javascript runtime environment.
+%
+Due to its event-driven non-blocking I/O model it is fast and responding even with heavy workload, such as mesh processing.
+%
+Non-image data for all subjects including the data dictionary is stored in a \texttt{JSON} file on the server.
+%
+Image data are available as raw \texttt{DICOM} files.
 %
@@
 %
-The pivot table implementation was adapted using \texttt{PivotTable.js}\footnote{Developed by N. Kruchten, \href{http://nicolas.kruchten.com/pivottable}{\texttt{nicolas.kruchten.com/pivottable}}}.
+The pivot table implementation uses \texttt{PivotTable.js}.\footnote{Developed by N. Kruchten, \href{http://nicolas.kruchten.com/pivottable}{\texttt{nicolas.kruchten.com/pivottable}}}
 %
@@
 %
-Communication between client and server is enabled via the WebSockets protocol.
+The WebSockets protocol handles the client-server communication. 
+%Communication between client and server is enabled via the WebSockets protocol.
+%Communication between client and server is enabled via the WebSockets protocol.
 %
@@
 %
-We accomplished this by converting it to a parameterized standalone console application that is spawned by \texttt{NodeJS} on client request and then reads the result from the console standard out and returns it to the client.
+We accomplished this by converting it to a parameterized standalone console application, spawned by \texttt{NodeJS} on client request.
+%
+The result is read from the console standard out and returns it to the client.
 %
 All parameter-steered console applications can be incorporated in this context.
-\\\\
+%
 \section{Application} \label{application}
@@
 %
-This is one of the most common reasons for an adult to see a physician in the Western civilization \cite{Backpain}.
+This is one of the most common diseases in the Western civilization \cite{Backpain}.
+%This is one of the most common reasons for an adult to see a physician in the Western civilization \cite{Backpain}.
 %
@@
 
-To ensure a fast and easy data access outside of statistical processors like \texttt{SPSS}, the data was exported to the JSON file format.
-%
-Since it lacks export methods for data dictionaries, we used \texttt{SPSS} to export our data to the 'SAS v9+' format, which saves the data labels, and exported the data values as non-labeled CSV.
-%
-A short script combined both data sources to a JSON file.
+To ensure a fast and easy data access outside of statistical processors like \texttt{SPSS}, the data was exported to the \texttt{JSON} file format.
+%
+Since it lacks export methods for data dictionaries, we used \texttt{SPSS} to export our data to the \texttt{SAS v9+} format, which saves the data labels, and exported the data values as non-labeled \texttt{CSV}.
+%
+A short script combined both data sources to a \texttt{JSON} file.
 %
@@
 %
-In accordance with epidemiological publications, we set the number of groups to five, to allow for contingency assessment.
+According to epidemiological publications, we set the number of groups to five, to allow for contingency assessment.
 
@@
 %
-Two user-defined landmarks on the top and bottom of the L3 vertebra are used to obtain an initial height estimation of the model.
+Two user-defined landmarks on the top and bottom of the L3 vertebra describe an initial model height estimation.
 %
@@
 %
-Shape distance between to mean meshes is mapped to color (recall Fig.~\ref{fig:visualization}).
+Shape distance between meshes is mapped to color (recall Fig.~\ref{fig:visualization}).
 %
@@
 %\com{Anmerkung von Steffen einbauen: Evaluierung geht um fachliches Feedback - Wissen über epidemiologie UND rücken und wissen um die Daten - deswegen schwer ganz viele Experten ranzuziehen}
-Guided by the work of Lam et al \cite{Lam2012}, we conducted an investigation of \emph{Visual Data Analysis and Reasoning} (\emph{VDAR}).
+Inspired by Lam et al. \cite{Lam2012}, we conducted an investigation of \emph{Visual Data Analysis and Reasoning} (\emph{VDAR}).
 %
@@
 %
-HV and KH are practicing physicians with focus on epidemiological research.
-%
-HV is a specialist in internal medicine (23 years of experience) and also the head of the SHIP, KH a radiologist (9 years of experience) and responsible for the MRI data acquisition of the study.
+HV and KH are physicians with focus on epidemiological research.
+%
+HV is a specialist in internal medicine (23 years of experience) and head of the SHIP, KH a radiologist (9 years of experience) and responsible for the SHIP MRI data acquisition.
 %and has a deep understanding on the available image data.
@@
 %The experts accessed the prototype by simply entering the website link into their browser of choice, no exchange of software was required.
-The experts accessed the prototype by entering the website link into their browser of choice, no exchange of software was required.
+The experts accessed the prototype by entering the website link into their browser. %of choice, no exchange of software was required.
 %Since our prototype ran on a local server, simply needed the homepage link to access the prototype.
@@
 %
-If a feature contains outliers, such as \emph{waist circumference} (e.g. by subjects with morbid obesity) this approach leads to sparse categories, making it hard to calculate associations.
+If a feature contains outliers, such as \emph{waist circumference} (e.g. by subjects with morbid obesity), this approach leads to sparse categories, making it hard to calculate associations.
 %
@@
 %
-Therefore, he applied the \emph{local investigation} pattern by a shape grouping step using shape-based clustering via dragging the \emph{All subjects} from the sidebar into the canvas area, triggering the shape clustering.
+Therefore, he applied the \emph{local investigation} pattern by a shape grouping step using shape-based clustering via dragging the \emph{All subjects} from the sidebar into the canvas area, triggering the shape clustering (Fig.~\ref{fig:application} b).
 %Shape-based hypothesis-free exploration starts with the \emph{local investigation} pattern by a shape grouping step using shape-based clustering.
 %
-The results for this step can be seen in Figure~\ref{fig:application} (b).
+%The results for this step can be seen in Figure~\ref{fig:application} (b).
 
@@
 %
-Therefore another \emph{multivariate analysis} was carried out by the feature \emph{gender} to the canvas from the contingency pane and selecting all female subjects (Fig.~\ref{fig:application} (b)).
+Therefore another \emph{multivariate analysis} was carried out by the feature \emph{gender} to the canvas from the contingency pane and selecting all female subjects (Fig.~\ref{fig:application}~b).
 %
@@
 %
-Age influences the lumbar spine shape, while the differences between subjects with and without back pain are small (Fig.~\ref{fig:application}~(c)).
-%
-The major differences seen in the visualization are therefore related to the age feature, masking differences via the back pain parameter.
+Age influences the lumbar spine shape, while the differences between subjects with and without back pain are small (Fig.~\ref{fig:application}~c).
+%
+The major differences seen in the visualization are therefore related to the age feature, masking differences related to the back pain parameter.
 %
@@
 %
-Then, the features derived from the clustering results for both subject groups are dropped into the canvas area as well (Fig.~\ref{fig:application} (d)).
+Then, the features derived from the clustering results for both subject groups are dropped into the canvas area as well (Fig.~\ref{fig:application}~d).
 %
@@
 %
-A \emph{multivariate analysis} using the pivot tables put gender and strong back pain in context to cluster affiliation (Fig.~\ref{fig:application} (d)). 
+A \emph{multivariate analysis} using the pivot tables put gender and strong back pain in context to cluster affiliation (Fig.~\ref{fig:application}~d). 
 %
@@
 %
-Both domain experts concluded positively over the benefits of the \emph{IVA} approach when analyzing image-based cohort study data.
+Both domain experts concluded positively over the \emph{IVA} approach. %when analyzing image-based cohort study data.
 %
@@
 %
-Great potential is also seen in communicating insights efficiently using the presented visualizations \com{(1.08)}.
+Great potential is also seen in communicating insights efficiently using the presented visualizations.
 
@@
 %
-It is also useful to verify established but still controversial risk factors, such as the metabolic syndrome for coronary heart disease and whether the data set provides even more suitable risk factors \com{(39.32)}.
+It is also useful to verify established but still controversial risk factors, such as the metabolic syndrome for coronary heart disease and whether the data set provides even more suitable risk factors.
 %
@@
 %
-\emph{Multivariate analysis} can also be improved by more ways of brushing the data as well as creating subgroups for comparison as a result of the hypothesis-driven analysis case.
-%
-Too small feature ranges yielding sparse groups could potentially hinder the ability to calculate statistical resilient measures, since they require a minimum amount of subjects exhibiting the selected feature ranges.
+\emph{Multivariate analysis} can be improved by more ways of brushing the data as well as creating subgroups for comparison as a result of the hypothesis-driven analysis case.
+%
+Too small feature ranges yielding sparse groups could hinder the calculation of statistical resilient measures, since they require a minimum amount of subjects exhibiting the selected feature ranges.
 %This also has to respect the number of subjects available for epidemiological studies
@@
 %
-A seconds explanation is the use of the L3 vertebra is initialization point of the lumbar spine model \com{(00.54)}.
+A second explanation is the use of the L3 vertebra as initialization point of the lumbar spine model.
 %
@@
 \textbf{Usage of different categorizations depending on expected outcome.}
-When categorizing numerical variables into equal groups it is possible to create sparse categories due to outliers, for example when analyzing body weight.
+Categorizing numerical variables into equal groups possibly creates sparse categories due to outliers.%, for example when analyzing body weight.
 %
@@
 \section{Summary and Conclusion}
-We presented an \emph{IVA} framework for the analysis of complex image-centric epidemiological data.
+We presented an \emph{IVA} framework for the analysis of image-centric epidemiological data.
 %
@@
 %
-The automatic suggestion of correlations using contingency methods like \emph{Cram\'{e}r's V} triggers \emph{hypothesis generation} by highlighting features potentially overlooked by the experts.
+The automatic suggestion of correlations using contingency methods, such as \emph{Cram\'{e}r's V} triggers \emph{hypothesis generation} by highlighting features potentially overlooked by the experts.
 %
@@
 %
-They allow a fast exchange between users and developers and employ different devices.
-%
-Visual analysis shows to be a promising way to clear the view on complex epidemiological data to uncover its secrets.
+%They allow a fast exchange between users and developers and employ different devices.
+%
+%Visual analysis shows to be a promising way to clear the view on complex epidemiological data to uncover its secrets.
 
@@
 The University of Greifswald is a member of the ‘Centre of Knowledge Interchange’ program of the Siemens AG.
-This work was supported by the DFG Priority Program 1335: Scalable Visual Analytics.}
+This work was supported by the DFG Priority Program 1335: Scalable Visual Analytics.
+We thank Marko Rak and Klaus Toennies for providing the image detection data.}
 \end{small}

### 2014-06-26 draft.tex
minor changes and todos completed

@@
 \usepackage{times}
-% TODO: Self added Packages - need to be removed afterwards
-%\usepackage{babel}
+
+% -- My Own Packages and Commands
 \usepackage[normalem]{ulem}
-%\usepackage{color}
 \usepackage{xcolor}
@@
 %
-We integrate an \emph{overview visualization}, clustering of features and object shape for data-driven subgroup definition and statistical key figures for measuring the association between variables.
+We integrate an \emph{overview visualization}, clustering of features and object shape for data-driven subgroup definition and statistical key figures for measuring the association between features.
 %
@@
 %
-Visual queries and direct feedback of Visual Analytics systems allow for a fast exploration of the data space incorporating many different variables.
+Visual queries and direct feedback of Visual Analytics systems allow for a fast exploration of the data space incorporating many different features.
 %
@@
 %
-Our approach uses information extracted from the segmented image data (such as 3D meshes) and variable associations with non-image epidemiological factors.
+Our approach uses information extracted from the segmented image data (such as 3D meshes) and feature associations with non-image epidemiological factors.
 
@@
 %
-Blaas et al. \cite{Blaas2007} presented a similar system, which also analyzed medical image data and variables derived from them using views from the feature and physical space.
+Blaas et al. \cite{Blaas2007} presented a similar system, which also analyzed medical image data and features derived from them using views from the feature and physical space.
 %
@@
 % %\add{One major focus is the connection of image and non-image data.}
-% %TODO aus Steenwijk: Visual data analysis combining multi-dimensional single patient medical data with real-time linked scatter plots and parallel coordi- nates representations was first investigated in the WEAVE system of Gresh et al.
+% Visual data analysis combining multi-dimensional single patient medical data with real-time linked scatter plots and parallel coordi- nates representations was first investigated in the WEAVE system of Gresh et al.
 [14].
@@
 %
-When applying clustering techniques to the non-image data it was found that \texttt{k-Prototypes} and \texttt{DBSCAN} are appropriate, but is strongly dependent on the chosen variables and distance measure \cite{Klemm2014BVM}.
+When applying clustering techniques to the non-image data it was found that \texttt{k-Prototypes} and \texttt{DBSCAN} are appropriate, but is strongly dependent on the chosen features and distance measure \cite{Klemm2014BVM}.
 %
@@
 %
-Detailed description of data variables, its meaning as well as unit of measurement are stored as a lookup table.
+Detailed description of data features, its meaning as well as unit of measurement are stored as a lookup table.
 %
@@
 	%
-	\item The \emph{contingency view} depicts relations for features in the canvas in an adjacency matrix (Fig.~\ref{fig:similarity}) and a \emph{contingency list}.
+	\item The \emph{contingency view} depicts relations for features in the canvas in an contingency matrix (Fig.~\ref{fig:similarity}) and a \emph{contingency list}.
 \end{itemize}
@@
 %
-The introduction of separate panes for the adjacency matrix, pivot table and sidebar, displayed with a mouse click on the corresponding button and sliding on top of the \emph{canvas} was considered more feasible (Figure~\ref{fig:visualization} shows the system with reeled-out pivot table pane).
+The introduction of separate panes for the contingency matrix, pivot table and sidebar, displayed with a mouse click on the corresponding button and sliding on top of the \emph{canvas} was considered more feasible (Figure~\ref{fig:visualization} shows the system with reeled-out pivot table pane).
 %
@@
 %
-\add{Groups are treated like dichotomous features.} %since they act the same by dividing the subject space into labeled categories.
-%
-Variables can be dragged from the sidebar into the canvas area for a \emph{feature localization}, which works as follows.
+Groups are treated like dichotomous features. %since they act the same by dividing the subject space into labeled categories.
+%
+Features can be dragged from the sidebar into the canvas area for a \emph{feature localization}, which works as follows.
 %
@@
  \includegraphics[width=3.5in]{figures/similarity_matrix}
- \caption{Adjacency matrix of 129 features (127 data set variables, 2 cluster results) showing 16,641 combinations.
+ \caption{Contingency matrix of 129 features (127 data set features, 2 cluster results) showing 16,641 combinations.
  %
@@
 %
-This also applies to the assignment of row or column variables.
-%
-Another way to avoid clutter is the user-driven selection of displayed variables.
+This also applies to the assignment of row or column features.
+%
+Another way to avoid clutter is the user-driven selection of displayed features.
 %
 To allow better comparison with respect to features, the values of each cell can also be displayed as percentage of the feature represented of either the row or column.
-% TODO IMPLEMENT
 
@@
 %
-The \emph{Cram\'{e}r's V} contingency coefficient can be used to calculate coherences between categorical variables \cite{CramerV}.
-%
-It is based on \emph{Pearson's $X^2$} distribution test \cite{ChiSquare}, which uses contingency tables holding the counts of subjects for all possible manifestations of two variables.
+The \emph{Cram\'{e}r's V} contingency coefficient can be used to calculate coherences between categorical features \cite{CramerV}.
+%
+It is based on \emph{Pearson's $X^2$} distribution test \cite{ChiSquare}, which uses contingency tables holding the counts of subjects for all possible manifestations of two features.
 %
@@
 %
-$V$ yields values between $0$, meaning that two variables are completely independent, and $1$ indicating they are the same.
+$V$ yields values between $0$, meaning that two features are completely independent, and $1$ indicating they are the same.
 %
@@
 %
-Contingency is visualized using an interactive adjacency matrix with association power mapped to color brightness.
+Contingency is visualized using an interactive contingency matrix with association power mapped to color brightness.
 %
@@
 %\com{Clustering also automatically triggered}
-
-% TODO: This can be an Section of it's own? Or part of the design?
 \subsection{Implementation} \label{implementation}
@@
 % Binning: https://www.researchgate.net/post/What_metric_best_captures_the_information_loss_when_binning_a_large_data-set_into_a_histogram_representation
-Continuous variables are discretized to allow for \emph{Cram\'{e}r's V} contingency coefficient assessment.
+Continuous features are discretized to allow for \emph{Cram\'{e}r's V} contingency coefficient assessment.
 %
@@
 %They pointed out the benefits of the possibility to look at all variables in the context of each other.
-Before, they were not able to to look at all variables in the context of each other.
+Before, they were not able to to look at all features in the context of each other.
 %
@@
 %
-This pointed out problems with the used categorization method splitting numerical variables into equally-sized ordinal bins.
+This pointed out problems with the used categorization method splitting numerical features into equally-sized ordinal bins.
 %
@@
 We observed, that the domain experts are more likely interested in features they are familiar with and have personal clinical experience with.
-% TODO: Contingency Matrix or Adjecency Matrix
 %\com{00.44 Clustering-Ergebnis Weiblich und Arbeitspensum Alter}
+% TODO: Correlation Matrix or Contingency Matrix??
 
@@
 % \paragraph{Analysis 3}
-% \com{ToDo: Needs to be rewritten}
 % \begin{figure}[htb]
@@
 \subsection{Further Feedback and Lessons Learned} \label{Lessons Learned}
-%\com{ToDo: Lessons Learned der einzelnen Cases - Fokus bei Hypothesenfreier Analyse muss etwa eine gute Übersichtsvisualisierung sein, da es genau das ist was den Epidemiologen hierzu im Moment fehlt}
 %
@@
 \textbf{Multivariate analysis is most important for hypothesis generation.}
-Both experts emphasized the potential of the \emph{multivariate analysis} capabilities of the adjacency matrix for getting insight into a large amount of features simultaneously.
+Both experts emphasized the potential of the \emph{multivariate analysis} capabilities of the contingency matrix for getting insight into a large amount of features simultaneously.
 %
@@
 %
-Creating adjacency matrices for subgroups, such as different age bins can help to characterize the aging process by deriving age-specific risk factors.
+Creating contingency matrices for subgroups, such as different age bins can help to characterize the aging process by deriving age-specific risk factors.
 %
@@
 %While the existing brushing and linking techniques already allow the comparison of image data
-%\com{(TODO weitere Unterteilung der Variablen und anschließende Anbindung an Statistik Tools)}
 
@@
 \textbf{Usage of different categorizations depending on expected outcome.}
-Categorizing numerical variables into equal groups possibly creates sparse categories due to outliers.%, for example when analyzing body weight.
+Categorizing numerical features into equal groups possibly creates sparse categories due to outliers.%, for example when analyzing body weight.
 %

### 2014-06-26 draft.tex
proofread first half

@@
 %
-We integrate an \emph{overview visualization}, clustering of features and object shape for data-driven subgroup definition and statistical key figures for measuring the association between features.
+We integrate an overview visualization, clustering of features and object shape for data-driven subgroup definition and statistical key figures for measuring the association between features.
 %
@@
 % Confounding variables, are often not obvious at all and characterizing them is already an epidemiological result.
-%
+% TODO - Khonya Paper Venue
+% TODO - Introduction Highlight Overview visualization?
+% TODO - Klemm et al. - Klingt ein bisschen shizophren
 Cohort studies often include medical image data.
@@
 %
-For instance, segmentation of the image data is required for an analysis of anatomical structure and of possible correlations between this structure and epidemiological factors.
+For instance, segmentation of the image data is required for an analysis of anatomical structure \com{and of possible} correlations between this structure and epidemiological factors.
 %
@@
  %
- The combination of statistical and interactive-driven analysis shows promising potential to unveil information in the data.
+ The combination of statistical and interactive analysis shows promising potential to unveil information in the data.
  %
@@
 %
-It is difficult to achieve, since many physicians are involved when thousands of test persons are examined and interviewed.
+It is difficult to achieve, since many physicians are involved when thousands of subjects are examined and interviewed.
 %
@@
 
-Our work is closest to that of Steenwijk \cite{Steenwijk2010}, Turkey \cite{Turkay2013}, and Angelelli \cite{Angelelli2014} and colleagues, who employ multiple coordinated view systems for the analysis of cohort study.
+Our work is closest to that of Steenwijk \cite{Steenwijk2010}, Turkey \cite{Turkay2013}, and Angelelli \cite{Angelelli2014} and colleagues, who employ multiple coordinated view systems for the analysis of cohort study data.
 %
@@
 %
-In epidemiology, multiple cases must be concurrently investigated and non-image data often lacks a spatial reference, such as gender and age.
+In epidemiology, multiple cases must be concurrently investigated and non-image data often lacks a spatial reference, such as gender or age.
 %
@@
 %
-Their understanding of a cohort differs from the understanding of the term in an epidemiological context by denoting every parameter-divided subject group as individual cohort.
+Their definition of a cohort differs from the understanding of the term in an epidemiological context by denoting every parameter-divided subject group as individual cohort.
 %
@@
 %
-While their goal to identify possible risk factors using socio-economic and health data similar to ours, they focus on iteratively refining defined hypothesis and on geographical data.
+While their goal to identify possible risk factors using socio-economic and health data is similar to ours, they focus on iteratively refining defined hypothesis and on geographical data.
 %
@@
 %The connection of time series plots with histograms and heat maps allows for fast insight into complex relationships and 
-While the works shows, how different visualization techniques provide insight into these data sets, it focuses on the time aspect, which is not present for our data.
+While the work shows, how different visualization techniques provide insight into these data sets, it focuses on the time aspect, which is not present for our data.
 %

### 2014-06-27 draft.tex
Replaced Features with Variables in the text

@@
 %
-Cohort studies comprise of heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors and more recently, medical image data.
+Cohort studies comprise of heterogenous variables describing the medical condition as well as demographic and lifestyle factors and more recently, medical image data.
 %
@@
 %
-We integrate an overview visualization, clustering of features and object shape for data-driven subgroup definition and statistical key figures for measuring the association between features.
+We integrate an overview visualization, clustering of variables and object shape for data-driven subgroup definition and statistical key figures for measuring the association between variables.
 %
@@
 %
-Potentially useful data variables (\emph{features}) are gathered using structured interviews and clinical examinations.
+Potentially useful data variables are gathered using structured interviews and clinical examinations.
 %
@@
 
-Compiling a list of features for tests of statistical resilience based on experience-driven hypotheses leaves out other features in the data which potentially interact with a disease.
+Compiling a list of variables for tests of statistical resilience based on experience-driven hypotheses leaves out other variables in the data which potentially interact with a disease.
 %
@@
 %
-Visual queries and direct feedback of Visual Analytics systems allow for a fast exploration of the data space incorporating many different features.
+Visual queries and direct feedback of Visual Analytics systems allow for a fast exploration of the data space incorporating many different variables.
 %
@@
 %
-\emph{Hypothesis generation} gains importance since the number of epidemiological features increases and the focus shifts towards more complex relations involving more than two features.
+\emph{Hypothesis generation} gains importance since the number of epidemiological variables increases and the focus shifts towards more complex relations involving more than two variables.
 Our contributions are:
@@
 	\item visualization techniques, which incorporate both information visualization and 3D rendering of organ shapes as well as combining them with epidemiological graphics and key figures,
-	\item highlighting subject groups and feature associations using shape-based clustering and statistical contingency measures.
+	\item highlighting subject groups and variable associations using shape-based clustering and statistical contingency measures.
 	%\item highlighting interesting subject groups and feature associations using shape-based clustering and statistical contingency measures.
@@
 % http://wiki.infowiss.net/Gulf_of_execution ?
-We applied our approach to a data set compiled to analyze lower back pain and aim to determine features, which indicate pathological changes.
-%
-This data set comprises 127 features and 2 sequences of MRI data from 6,753 subjects.
+We applied our approach to a data set compiled to analyze lower back pain and aim to determine variables, which indicate pathological changes.
+%
+This data set comprises 127 variables and 2 sequences of MRI data from 6,753 subjects.
 %
@@
 %
-	\item A set of features depicting conditions affected by the hypothesis is compiled accordingly.
-%
-	\item Confounding features are identified and taken into account (for example using stratification).
-%
-	\item Statistical methods, such as regression analysis, assess the association of selected features with the investigated disease.
+	\item A set of variables depicting conditions affected by the hypothesis is compiled accordingly.
+%
+	\item Confounding variables are identified and taken into account (for example using stratification).
+%
+	\item Statistical methods, such as regression analysis, assess the association of selected variables with the investigated disease.
 %
@@
 %
-Longitudinal studies require the acquired features to be comparable for evaluation.
+Longitudinal studies require the acquired variables to be comparable for evaluation.
 %
@@
 %
-Grouping subjects using epidemiological features is essential in cohort studies to allow per-group risk determination.
+Grouping subjects using epidemiological variables is essential in cohort studies to allow per-group risk determination.
 %
@@
 %
-This includes confidence intervals indicating the certainty of that feature being a risk factor.
+This includes confidence intervals indicating the certainty of that variable being a risk factor.
 %To determine whether a subject is prone to be affected by a certain disease, \emph{relative risks} are expressed through the evaluation of p-values, which indicate statistical significance.
@@
 %
-For ethical or medical reasons some features cannot be gathered for each subject, e.g. women-specific questions about menstrual status or number of born children.
-%
-Follow-up examinations or questions about conditions such as medications taken after a diagnosed disease also yield features only available for a small amount of subjects.
+For ethical or medical reasons some variables cannot be gathered for each subject, e.g. women-specific questions about menstrual status or number of born children.
+%
+Follow-up examinations or questions about conditions such as medications taken after a diagnosed disease also yield variables only available for a small amount of subjects.
 %
@@
 %
-Dichotomous data can also be derived by aggregating features to yield only two manifestations (e.g. subjects younger or older than 50 years).
+Dichotomous data can also be derived by aggregating variables to yield only two manifestations (e.g. subjects younger or older than 50 years).
 %
@@
 %
-Sparse populated features are hard to assess statistically.
+Sparse populated variables are hard to assess statistically.
 %
@@
 %
-Statistical correlations are prone to \emph{confounding}, meaning that the association of two features is influenced by a third feature, which needs to be isolated.
+Statistical correlations are prone to \emph{confounding}, meaning that the association of two variables is influenced by a third variable, which needs to be isolated.
 %
@@
 %
-While we use a similar approach when analyzing non-image data, our process also includes \emph{overview visualizations} and statistical suggestions of potentially interesting features.
+While we use a similar approach when analyzing non-image data, our process also includes \emph{overview visualizations} and statistical suggestions of potentially interesting variables.
 %
@@
 %
-Our approach uses information extracted from the segmented image data (such as 3D meshes) and feature associations with non-image epidemiological factors.
+Our approach uses information extracted from the segmented image data (such as 3D meshes) and variable associations with non-image epidemiological factors.
 
@@
 %
-Blaas et al. \cite{Blaas2007} presented a similar system, which also analyzed medical image data and features derived from them using views from the feature and physical space.
+Blaas et al. \cite{Blaas2007} presented a similar system, which also analyzed medical image data and variables derived from them using views from the variable and physical space.
 %
@@
 %
-We employ the idea of adding features via drag and drop into a canvas area.
-
-Generalized Pairs Plots (\texttt{GPLOM´S}) are an information visualization technique comparing heterogenous features pairwise using a plot-matrix grouped by type \cite{GPLOMS, Francois2013}.
+We employ the idea of adding variables via drag and drop into a canvas area.
+
+Generalized Pairs Plots (\texttt{GPLOM´S}) are an information visualization technique comparing heterogenous variables pairwise using a plot-matrix grouped by type \cite{GPLOMS, Francois2013}.
 %
@@
 %
-The resulting matrix provides an \emph{overview visualization}, but requires a lot of screen space for many features (127 in our application scenario).
+The resulting matrix provides an \emph{overview visualization}, but requires a lot of screen space for many variables (127 in our application scenario).
 %
@@
 Dai et al.
-\cite{Dai2005} explored risk factors by incorporating choropleth maps of epidemiological features (e.g., mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines.
+\cite{Dai2005} explored risk factors by incorporating choropleth maps of epidemiological variables (e.g., mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines.
 %
@@
 %
-This requires a geographic colocation of associated features.
+This requires a geographic colocation of associated variables.
 %
@@
 %
-This work focuses on incorporating these derived data as new features, making it possible to include it into the hypothesis validation and generation process.
-%
-When applying clustering techniques to the non-image data it was found that \texttt{k-Prototypes} and \texttt{DBSCAN} are appropriate, but is strongly dependent on the chosen features and distance measure \cite{Klemm2014BVM}.
+This work focuses on incorporating these derived data as new variables, making it possible to include it into the hypothesis validation and generation process.
+%
+When applying clustering techniques to the non-image data it was found that \texttt{k-Prototypes} and \texttt{DBSCAN} are appropriate, but is strongly dependent on the chosen variables and distance measure \cite{Klemm2014BVM}.
 %
@@
 %
-Association rules created by data mining methods can be analyzed interactively with their tool and highlight potentially overlooked features.
+Association rules created by data mining methods can be analyzed interactively with their tool and highlight potentially overlooked variables.
 %
@@
 %
-Our system follows the same rules for selecting subject groups, but our goal is to judge feature relations and potential outcomes.
+Our system follows the same rules for selecting subject groups, but our goal is to judge variable relations and potential outcomes.
 \\\\
@@
 %
-(3) Overview visualizations using statistical abstractions aim to provide an unbiased feature relationship assessment.
+(3) Overview visualizations using statistical abstractions aim to provide an unbiased variable relationship assessment.
 
@@
 %
-Statistical tests on features associated to a hypothesis yield a value for deciding whether the data supports the hypothesis.
-%
-Features not included in the analysis may potentially support the chosen hypothesis by discriminating the population in the expected way, but are not highlighted.
+Statistical tests on variables associated to a hypothesis yield a value for deciding whether the data supports the hypothesis.
+%
+Variables not included in the analysis may potentially support the chosen hypothesis by discriminating the population in the expected way, but are not highlighted.
 %
@@
 \\\\
-\emph{IVA} tries to illuminate the black box by making the domain experts part of an iterative feature selection process (see Fig.~\ref{fig:WorkflowComparison}~b).
+\emph{IVA} tries to illuminate the black box by making the domain experts part of an iterative variable selection process (see Fig.~\ref{fig:WorkflowComparison}~b).
 %\emph{IVA} tries to illuminate the black box by making the domain experts part of an iterative feature selection process, which is shown in Figure~\ref{fig:WorkflowComparison} (b) as part of the epidemiological workflow.
@@
 
-\paragraph{Multivariate Analysis} refers to an investigation of multi-variate properties of the dependent data by specifying a feature in one attribute view while analyzing the value distribution with respect to other variables in other attribute views.
-%
-Epidemiologists may define a feature in a scatter plot of the body mass index (BMI) and age to inspect the result in a histogram of body height.
+\paragraph{Multivariate Analysis} refers to an investigation of multi-variate properties of the dependent data by specifying a variable in one attribute view while analyzing the value distribution with respect to other variables in other attribute views.
+%
+Epidemiologists may define a variable in a scatter plot of the body mass index (BMI) and age to inspect the result in a histogram of body height.
 %
@@
 %
-A data dictionary stores information about each manifestation of a feature.
-%
-Detailed description of data features, its meaning as well as unit of measurement are stored as a lookup table.
+A data dictionary stores information about each manifestation of a variable.
+%
+Detailed description of data variables, its meaning as well as unit of measurement are stored as a lookup table.
 %
@@
 %
-Morphometric features are derived to allow for statistical comparison of the tissue, which incorporates mostly positions, diameters, volumes and relative distances and alignment to other structures.
+Morphometric variables are derived to allow for statistical comparison of the tissue, which incorporates mostly positions, diameters, volumes and relative distances and alignment to other structures.
 %-------------------------------- Figure --------------------------------
@@
  %
- Usually starting with a selection of a feature of interest (user-driven or via data mining techniques), the data are mapped using a visualization technique appropriate for the selected data types.
- %
- The data are visualized in the range and domain space, which can be brushed, yielding new groups, to be investigated using further features.
+ Usually starting with a selection of a variable of interest (user-driven or via data mining techniques), the data are mapped using a visualization technique appropriate for the selected data types.
+ %
+ The data are visualized in the range and domain space, which can be brushed, yielding new groups, to be investigated using further variables.
  %
@@
  %
- (Left) Screenshot from the front-end, which is divided as follows: (a) The sidebar containing all features as well as the groups defined in the analysis process; (b) the canvas area where features can be added via drag and drop and the visualization is chosen automatically according to the data type; (c) the interactive pivot table showing the exact numbers for each displayed feature combination; (d) buttons to open panes containing the contingency matrix, contingency pane and pivot table.
- %
- The data displayed is used to analyze the lumbar spine. Features can be added freely on the canvas via drag and drop.
- %
- Dropping the \emph{gender} parameter on the already plotted \emph{body size} container creates a mosaic plot combining both features (right).
+ (Left) Screenshot from the front-end, which is divided as follows: (a) The sidebar containing all variables as well as the groups defined in the analysis process; (b) the canvas area where variables can be added via drag and drop and the visualization is chosen automatically according to the data type; (c) the interactive pivot table showing the exact numbers for each displayed variable combination; (d) buttons to open panes containing the contingency matrix, contingency pane and pivot table.
+ %
+ The data displayed is used to analyze the lumbar spine. Variables can be added freely on the canvas via drag and drop.
+ %
+ Dropping the \emph{gender} parameter on the already plotted \emph{body size} container creates a mosaic plot combining both variables (right).
  %
@@
 %-------------------------------- /Figure -------------------------------
-Our proposed \emph{IVA} workflow consists of three major steps as illustrated in Figure~\ref{fig:InteractionLoop}: Feature selection, visualization and brushing.
-%
-A hypothesis-driven analysis usually starts with the selection of features, or shape groups derived from a shape-based clustering.
-%
-Hypothesis generation with focus on image data starts with a shape-based clustering or an \emph{overview visualization} of all features.
-%
-The feature is mapped using an automatically chosen visualization appropriate for its data type (described in detail in the following section).
+Our proposed \emph{IVA} workflow consists of three major steps as illustrated in Figure~\ref{fig:InteractionLoop}: Variable selection, visualization and brushing.
+%
+A hypothesis-driven analysis usually starts with the selection of variables, or shape groups derived from a shape-based clustering.
+%
+Hypothesis generation with focus on image data starts with a shape-based clustering or an \emph{overview visualization} of all variables.
+%
+The variable is mapped using an automatically chosen visualization appropriate for its data type (described in detail in the following section).
 %
@@
 %
-In our system, the visualization can either be brushed or new features can be added to the analysis. 
+In our system, the visualization can either be brushed or new variables can be added to the analysis. 
 %
@@
 %
-Brushed regions are treated like categorical features, as they divide the subject space in the same way.
-%
-Selecting features also triggers a \emph{multivariate analysis} using contingency values (described in the following section) to highlight associated features.  %\rem{with selected features.}
+Brushed regions are treated like categorical variables, as they divide the subject space in the same way.
+%
+Selecting variables also triggers a \emph{multivariate analysis} using contingency values (described in the following section) to highlight associated variables.  %\rem{with selected features.}
 %
@@
 %
-The suitability of visualization techniques for epidemiological data depends on its ability to compare multiple data features while highlighting associations.
+The suitability of visualization techniques for epidemiological data depends on its ability to compare multiple data variables while highlighting associations.
 %
@@
 %
-Incorporating the \emph{IVA} workflow for image-centric cohort study data requires \emph{overview visualizations} as well as \emph{multivariate visualizations}, which bring image-derived information in context to non-image features.
+Incorporating the \emph{IVA} workflow for image-centric cohort study data requires \emph{overview visualizations} as well as \emph{multivariate visualizations}, which bring image-derived information in context to non-image variables.
 
@@
 \begin{itemize} \itemsep0.2em
-		\item The \emph{sidebar}, which contains all epidemiological features.
-As cluster results group features like categorical features and are part of the sidebar as well (Fig.~\ref{fig:visualization}~a).
+		\item The \emph{sidebar}, which contains all epidemiological variables.
+As cluster results group variables like categorical variables and are part of the sidebar as well (Fig.~\ref{fig:visualization}~a).
 	%
@@
 	%
-	\item The interactive \emph{pivot table} gives detailed numerical information of the features in the canvas view.
+	\item The interactive \emph{pivot table} gives detailed numerical information of the variables in the canvas view.
 This view on the data is familiar to epidemiologists (Fig.~\ref{fig:visualization}~c).
 	%
-	\item The \emph{contingency view} depicts relations for features in the canvas in an contingency matrix (Fig.~\ref{fig:similarity}) and a \emph{contingency list}.
+	\item The \emph{contingency view} depicts relations for variables in the canvas in an contingency matrix (Fig.~\ref{fig:similarity}) and a \emph{contingency list}.
 \end{itemize}
@@
 %
-It categorizes all features into different types, such as somatometric (measurements of the human body dimensions), disease- or lifestyle-related, pain indicators and laboratory data (Fig.~\ref{fig:visualization}~a).
+It categorizes all variables into different types, such as somatometric (measurements of the human body dimensions), disease- or lifestyle-related, pain indicators and laboratory data (Fig.~\ref{fig:visualization}~a).
 %
@@
 %
-Groups are treated like dichotomous features. %since they act the same by dividing the subject space into labeled categories.
-%
-Features can be dragged from the sidebar into the canvas area for a \emph{feature localization}, which works as follows.
+Groups are treated like dichotomous variables. %since they act the same by dividing the subject space into labeled categories.
+%
+Variables can be dragged from the sidebar into the canvas area for a \emph{feature localization}, which works as follows.
 %
@@
 
-\paragraph{Adaptive Feature Visualization.} \label{sec:AdaptiveFeatureVisualization}
+\paragraph{Adaptive Variable Visualization.} \label{sec:AdaptiveFeatureVisualization}
 \begin{figure}[htb]
@@
  \includegraphics[width=3.5in]{figures/similarity_matrix}
- \caption{Contingency matrix of 129 features (127 data set features, 2 cluster results) showing 16,641 combinations.
+ \caption{Contingency matrix of 129 variables (127 data set variables, 2 cluster results) showing 16,641 combinations.
  %
@@
  %Mouse-over focuses a entry and highlights the feature names for better readability.
- Mouse-over an entry enlarges the feature names for better readability.
+ Mouse-over an entry enlarges the variable names for better readability.
  %
@@
 \end{figure}
-The visualization type, inspired by \texttt{GPLOMS} \cite{GPLOMS, Francois2013}, is dynamically chosen based on the feature types and number to allow for \emph{multivariate analysis}.
-%
-Categorical data are either mapped to bar charts (single features) or mosaic plots (multiple features).
+The visualization type, inspired by \texttt{GPLOMS} \cite{GPLOMS, Francois2013}, is dynamically chosen based on the variable types and number to allow for \emph{multivariate analysis}.
+%
+Categorical data are either mapped to bar charts (single variables) or mosaic plots (multiple variables).
 %
@@
 %
-Continuous data can be visualized using scatterplots (two features) or parallel coordinates (multiple features), but in epidemiology, this data type is usually categorized into ordinal groups of \emph{equal size}.
+Continuous data can be visualized using scatterplots (two variables) or parallel coordinates (multiple variables), but in epidemiology, this data type is usually categorized into ordinal groups of \emph{equal size}.
 %
@@
 %
-Conclusions based on statistical relationships derived from groups already biased by feature distribution are heavily influenced by the used discretization.
+Conclusions based on statistical relationships derived from groups already biased by variable distribution are heavily influenced by the used discretization.
 %
@@
 %
-Until now they were not able rapidly validate shape differences based on non-image features.
-%
-Dropping a feature on an existing plot adapts the visualization dynamically to allow for comparison (Fig.~\ref{fig:visualization}~right).
+Until now they were not able rapidly validate shape differences based on non-image variables.
+%
+Dropping a variable on an existing plot adapts the visualization dynamically to allow for comparison (Fig.~\ref{fig:visualization}~right).
 % --- PARAGRAPH ---
@@
 %
-This also applies to the assignment of row or column features.
-%
-Another way to avoid clutter is the user-driven selection of displayed features.
-%
-To allow better comparison with respect to features, the values of each cell can also be displayed as percentage of the feature represented of either the row or column.
-
-\paragraph{Automated Feature Suggestion using a Contingency Matrix.}
+This also applies to the assignment of row or column variables.
+%
+Another way to avoid clutter is the user-driven selection of displayed variables.
+%
+To allow better comparison with respect to variables, the values of each cell can also be displayed as percentage of the variable represented of either the row or column.
+
+\paragraph{Automated Variable Suggestion using a Contingency Matrix.}
 %\rem{As previously discussed}
-Highlighting potentially interesting values in the data set is one major benefits of the \emph{IVA}-powered approach and belongs to the \emph{multivariate analysis} pattern,  analyzing features outside the shape space.
-%
-Turkay et al. \cite{Turkay2013} used the approach to calculate key figures based on the distribution functions of each feature derived from the image data.
-%
-Since the majority of our data are categorical features, we have to employ different solutions.
-%
-The \emph{Cram\'{e}r's V} contingency coefficient can be used to calculate coherences between categorical features \cite{CramerV}.
-%
-It is based on \emph{Pearson's $X^2$} distribution test \cite{ChiSquare}, which uses contingency tables holding the counts of subjects for all possible manifestations of two features.
+Highlighting potentially interesting values in the data set is one major benefits of the \emph{IVA}-powered approach and belongs to the \emph{multivariate analysis} pattern,  analyzing variables outside the shape space.
+%
+Turkay et al. \cite{Turkay2013} used the approach to calculate key figures based on the distribution functions of each variable derived from the image data.
+%
+Since the majority of our data are categorical variables, we have to employ different solutions.
+%
+The \emph{Cram\'{e}r's V} contingency coefficient can be used to calculate coherences between categorical variables \cite{CramerV}.
+%
+It is based on \emph{Pearson's $X^2$} distribution test \cite{ChiSquare}, which uses contingency tables holding the counts of subjects for all possible manifestations of two variables.
 %
@@
 %
-$V$ yields values between $0$, meaning that two features are completely independent, and $1$ indicating they are the same.
+$V$ yields values between $0$, meaning that two variables are completely independent, and $1$ indicating they are the same.
 %
@@
 %
-Some manifestations and feature combinations, which are only exposed by small subject groups, cannot be assessed with this technique.
+Some manifestations and variable combinations, which are only exposed by small subject groups, cannot be assessed with this technique.
 %
@@
 %
-The contingency matrix highlights correlations between all features. %of visualized features with features in the data set.
-%
-This aims to highlight features possibly associated with the focused hypothesis as well as trigger new hypotheses.
+The contingency matrix highlights correlations between all variables. %of visualized features with features in the data set.
+%
+This aims to highlight variables possibly associated with the focused hypothesis as well as trigger new hypotheses.
 %
@@
 \paragraph{Contingency Pane.}
-Dropping a feature into the canvas area adds an entry for each manifestation of it to the \emph{contingency matrix}.
-%
-Testing sessions revealed that it was tedious to open the matrix every time a new feature is added. 
+Dropping a variable into the canvas area adds an entry for each manifestation of it to the \emph{contingency matrix}.
+%
+Testing sessions revealed that it was tedious to open the matrix every time a new variable is added. 
 %A simple list of features, which correlate with the last added feature.
 %
-As a consequence, the \emph{contingency pane}, a table containing correlating features for the last added visualizations in descending order of the \emph{Cram\'{e}r's V} value was added.
-%
-\emph{Contingency pane} entries can be dragged and dropped into the \emph{canvas area} just like features in the \emph{sidebar}.
+As a consequence, the \emph{contingency pane}, a table containing correlating variables for the last added visualizations in descending order of the \emph{Cram\'{e}r's V} value was added.
+%
+\emph{Contingency pane} entries can be dragged and dropped into the \emph{canvas area} just like variables in the \emph{sidebar}.
 % Build after user request
@@
 \paragraph{Initialization and Clustering.}
-Using feature suggestion allows to initialize the system with a set of potentially interesting visualizations.
+Using variable suggestion allows to initialize the system with a set of potentially interesting visualizations.
 %
@@
 
-Subject clustering is triggered automatically as \emph{local investigation} for each feature after it was added to the canvas by the user.
+Subject clustering is triggered automatically as \emph{local investigation} for each variable after it was added to the canvas by the user.
 %
@@
 %
-The server performs heavy statistical tasks, such as calculation of \emph{Cram\'{e}r's V} values for all feature combinations in order to keep the computation time on the client as low as possible.
+The server performs heavy statistical tasks, such as calculation of \emph{Cram\'{e}r's V} values for all variable combinations in order to keep the computation time on the client as low as possible.
 %
@@
 %
-In comparable studies, only a few shape-related features are included \cite{Lang2011}.
+In comparable studies, only a few shape-related variables are included \cite{Lang2011}.
 %
@@
 %
-Determining risk factors in this area can lead to particularly affected risk groups, prognostic features for diagnosis and treatment of lumbar back pain and a better understanding of effects of preventive measures, such as occupational health and safety regulations \cite{Fletcher2012}.
+Determining risk factors in this area can lead to particularly affected risk groups, prognostic variables for diagnosis and treatment of lumbar back pain and a better understanding of effects of preventive measures, such as occupational health and safety regulations \cite{Fletcher2012}.
 % Determining risk factors in this area can lead to \cite{Fletcher2012}
@@
 %
-There are 127 features describing diagnosed diseases, lifestyle factors, women-specific factors, pain indicators, laboratory values and somatometric features for 6,753 subjects (4,420 from \texttt{SHIP-Trend-0} and 2,333 from \texttt{SHIP-2}).
-%
-Since data acquisition protocols between these two cohorts are identical, the features between the two cohorts are comparable.
-%
-The data contains 30 metric, 7 nominal, 29 ordinal and 62 dichotomous epidemiological features.
-%
-Somatometric features include measures of the human body, such as body height, weight and body fat percentage as well as gender.
+There are 127 variables describing diagnosed diseases, lifestyle factors, women-specific factors, pain indicators, laboratory values and somatometric variables for 6,753 subjects (4,420 from \texttt{SHIP-Trend-0} and 2,333 from \texttt{SHIP-2}).
+%
+Since data acquisition protocols between these two cohorts are identical, the variables between the two cohorts are comparable.
+%
+The data contains 30 metric, 7 nominal, 29 ordinal and 62 dichotomous variables.
+%
+Somatometric variables include measures of the human body, such as body height, weight and body fat percentage as well as gender.
 %
@@
 %
-Other features, such as pain indicators or lifestyle indicators (e.g. physical activity) are more subjective and less reliable.
-%
-There are also features missing for each subject, such as features building upon each other (e.g. Do you have high blood pressure? Which medication is prescribed against it?).
+Other variables, such as pain indicators or lifestyle indicators (e.g. physical activity) are more subjective and less reliable.
+%
+There are also variables missing for each subject, such as variables building upon each other (e.g. Do you have high blood pressure? Which medication is prescribed against it?).
 %
@@
 %
-Each feature is stored as an object containing the data as array it's data type, a detailed description and data dictionary translating value or error IDs to values.
+Each variable is stored as an object containing the data as array it's data type, a detailed description and data dictionary translating value or error IDs to values.
 % \begin{itemize}
@@
 % Binning: https://www.researchgate.net/post/What_metric_best_captures_the_information_loss_when_binning_a_large_data-set_into_a_histogram_representation
-Continuous features are discretized to allow for \emph{Cram\'{e}r's V} contingency coefficient assessment.
+Continuous variables are discretized to allow for \emph{Cram\'{e}r's V} contingency coefficient assessment.
 %
@@
 %They pointed out the benefits of the possibility to look at all variables in the context of each other.
-Before, they were not able to to look at all features in the context of each other.
+Before, they were not able to to look at all variables in the context of each other.
 %
@@
 %
-The radiologist (KH) was looking for correlations with shape-related features in the data, finding that they correlate with \emph{leg pain}, \emph{age}, \emph{body height} and \emph{hormone replacement therapy status}.
+The radiologist (KH) was looking for correlations with shape-related variables in the data, finding that they correlate with \emph{leg pain}, \emph{age}, \emph{body height} and \emph{hormone replacement therapy status}.
 %
@@
 
-After this initial overview, KH performed a \emph{multivariate analysis} by introducing features, such as \emph{age}, \emph{waist circumference}, \emph{weight} or \emph{lumbar spine canal curvature} as bar charts views into the canvas area and selected subgroups to see how they are distributed and if they could observe unusual behavior in the mean-shapes.
-%
-This pointed out problems with the used categorization method splitting numerical features into equally-sized ordinal bins.
-%
-If a feature contains outliers, such as \emph{waist circumference} (e.g. by subjects with morbid obesity), this approach leads to sparse categories, making it hard to calculate associations.
+After this initial overview, KH performed a \emph{multivariate analysis} by introducing variables, such as \emph{age}, \emph{waist circumference}, \emph{weight} or \emph{lumbar spine canal curvature} as bar charts views into the canvas area and selected subgroups to see how they are distributed and if they could observe unusual behavior in the mean-shapes.
+%
+This pointed out problems with the used categorization method splitting numerical variables into equally-sized ordinal bins.
+%
+If a variable contains outliers, such as \emph{waist circumference} (e.g. by subjects with morbid obesity), this approach leads to sparse categories, making it hard to calculate associations.
 %
@@
 %
-HV started also with a \emph{multivariate analysis} using the contingency matrix to analyze non-image features, such as age-associated parameters like \emph{income}, \emph{blood fat values} or \emph{number of born children}, but found no associations of interest.
+HV started also with a \emph{multivariate analysis} using the contingency matrix to analyze non-image variables, such as age-associated parameters like \emph{income}, \emph{blood fat values} or \emph{number of born children}, but found no associations of interest.
 %
@@
 %To perform an automatic feature suggestion using a \emph{multivariate analysis} by looking at \emph{Cram\'{e}r's V} contingency values of in the contingency pane.
-To get an overview over the suggested features, the user opened the contingency pane to perform a \emph{multivariate analysis} by looking at \emph{Cram\'{e}r's V} contingency values of all clusters, revealing a strong correlation with \emph{gender} and \emph{body size}.
-%
-Therefore another \emph{multivariate analysis} was carried out by the feature \emph{gender} to the canvas from the contingency pane and selecting all female subjects (Fig.~\ref{fig:application}~b).
+To get an overview over the suggested variables, the user opened the contingency pane to perform a \emph{multivariate analysis} by looking at \emph{Cram\'{e}r's V} contingency values of all clusters, revealing a strong correlation with \emph{gender} and \emph{body size}.
+%
+Therefore another \emph{multivariate analysis} was carried out by the variable \emph{gender} to the canvas from the contingency pane and selecting all female subjects (Fig.~\ref{fig:application}~b).
 %
@@
 %
-We observed, that the domain experts are more likely interested in features they are familiar with and have personal clinical experience with.
+We observed, that the domain experts are more likely interested in variables they are familiar with and have personal clinical experience with.
 %\com{00.44 Clustering-Ergebnis Weiblich und Arbeitspensum Alter}
@@
 \subsection{Case 2: Hypothesis-based Analysis}
-If the user proposes a hypothesis about a relation between a non-image feature regarding shape, the workflow slightly differs from the hypothesis-free analysis.
-%
-The starting point follows the \emph{feature investigation} pattern, where a feature of interest is selected by dragging it into the canvas area and viewing the subject's distribution as well as their shape differences.
+If the user proposes a hypothesis about a relation between a non-image variable regarding shape, the workflow slightly differs from the hypothesis-free analysis.
+%
+The starting point follows the \emph{feature localization} pattern, where a variable of interest is selected by dragging it into the canvas area and viewing the subject's distribution as well as their shape differences.
 % \paragraph{Analysis 1.}
@@
 %
-To validate this hypothesis, a \emph{feature investigation} was performed by introducing the dichotomous feature ``\emph{Did you experience back pain in the last three months?}'' together with the age as mosaic plot by dropping both features on the canvas area.
+To validate this hypothesis, a \emph{feature localization} was performed by introducing the dichotomous variable ``\emph{Did you experience back pain in the last three months?}'' together with the age as mosaic plot by dropping both variables on the canvas area.
 %
@@
 %
-The major differences seen in the visualization are therefore related to the age feature, masking differences related to the back pain parameter.
+The major differences seen in the visualization are therefore related to the age variable, masking differences related to the back pain parameter.
 %
@@
 
-The epidemiologists pointed out that they would like to see a more intuitive and fast way to select subgroups based on different features to make full use of the analysis capabilities, as discussed in Section~\ref{Lessons Learned}.
+The epidemiologists pointed out that they would like to see a more intuitive and fast way to select subgroups based on different variables to make full use of the analysis capabilities, as discussed in Section~\ref{Lessons Learned}.
 %gender, use of pain medications. this would be good for future work, providing a better way to subdivide the data set and construct groups. The presented method is very good to gain an overview
@@
 %
-Therefore the feature ``\emph{Did you experience back pain in the past three months?}'' is dropped into the canvas area. %, as seen in previous Fig.~\ref{ToDoA}.
-%
-Then, the features derived from the clustering results for both subject groups are dropped into the canvas area as well (Fig.~\ref{fig:application}~d).
+Therefore the variable ``\emph{Did you experience back pain in the past three months?}'' is dropped into the canvas area. %, as seen in previous Fig.~\ref{ToDoA}.
+%
+Then, the variables derived from the clustering results for both subject groups are dropped into the canvas area as well (Fig.~\ref{fig:application}~d).
 %
@@
 %
-It rather triggers the analysis of potentially associated features with a pathology of interest.
+It rather triggers the analysis of potentially associated variables with a pathology of interest.
 %Zusammenfassend kann gesagt werden, dass hypothesenanalyse und hypothesengenerierung durch das design des Tools hand in hand gehen, es istn icht dafür geeignet statistisch bis ins detail hypothesen zu validieren, sondern triggert eher die analyse potentiell assoziierter variablen, die später verfeinert werden.
@@
 \textbf{Multivariate analysis is most important for hypothesis generation.}
-Both experts emphasized the potential of the \emph{multivariate analysis} capabilities of the contingency matrix for getting insight into a large amount of features simultaneously.
+Both experts emphasized the potential of the \emph{multivariate analysis} capabilities of the contingency matrix for getting insight into a large amount of variables simultaneously.
 %
@@
 %
-Too small feature ranges yielding sparse groups could hinder the calculation of statistical resilient measures, since they require a minimum amount of subjects exhibiting the selected feature ranges.
+Too small variable ranges yielding sparse groups could hinder the calculation of statistical resilient measures, since they require a minimum amount of subjects exhibiting the selected variable ranges.
 %This also has to respect the number of subjects available for epidemiological studies
@@
 \textbf{Usage of different categorizations depending on expected outcome.}
-Categorizing numerical features into equal groups possibly creates sparse categories due to outliers.%, for example when analyzing body weight.
+Categorizing numerical variables into equal groups possibly creates sparse categories due to outliers.%, for example when analyzing body weight.
 %
@@
 %
-The use of pivot tables is familiar to epidemiologists while embracing the power of interactive adjustment of the shown features.
-%
-The automatic suggestion of correlations using contingency methods, such as \emph{Cram\'{e}r's V} triggers \emph{hypothesis generation} by highlighting features potentially overlooked by the experts.
-%
-Shape-based clustering assesses the variability of an anatomical structure in the context of non-image features such as disease indicators or lifestyle factors.
+The use of pivot tables is familiar to epidemiologists while embracing the power of interactive adjustment of the shown variables.
+%
+The automatic suggestion of correlations using contingency methods, such as \emph{Cram\'{e}r's V} triggers \emph{hypothesis generation} by highlighting variables potentially overlooked by the experts.
+%
+Shape-based clustering assesses the variability of an anatomical structure in the context of non-image variables such as disease indicators or lifestyle factors.
 
@@
 %\com{This needs to be adjusted to fit the new Application section}
-Findings from analyzing lumbar back pain using the \emph{IVA} approach range from deriving shape-based groups of subjects to detailed description of features potentially associated with the disease, such as waist circumference, alcohol consumption and attentiveness disorder.
+Findings from analyzing lumbar back pain using the \emph{IVA} approach range from deriving shape-based groups of subjects to detailed description of variables potentially associated with the disease, such as waist circumference, alcohol consumption and attentiveness disorder.
 %
@@
 % \\\\
-As the number of image-centric cohort studies, participating subjects, gathered features and imaging modalities rises, and advances towards comparability between cohort studies are made, the gap between data complexity and analyzability increases.
+As the number of image-centric cohort studies, participating subjects, gathered variables and imaging modalities rises, and advances towards comparability between cohort studies are made, the gap between data complexity and analyzability increases.
 %

### 2014-06-27 draft.tex
Replaced 'features' with 'variables' in the figures

@@
 %
-It is also useful to verify established but still controversial risk factors, such as the metabolic syndrome for coronary heart disease and whether the data set provides even more suitable risk factors.
+It is also useful to verify established but still controversial risk factors, such as the metabolic syndrome for coronary heart disease and whether the data set provides more suitable risk factors.
 %

### 2014-06-27 draft.tex
Added missing email references, cut the introduction to fit the format again

@@
 %% indicate IEEE Member or Student Member in form indicated below
-\author{Paul Klemm$^{1}$, Steffen Oeltze-Jafra$^{1}$, Kai Lawonn$^{1}$, Katrin Hegenscheid$^{2}$, Henry V{\"o}lzke$^{2}$, Bernhard Preim$^{1}$}
+\author{Paul Klemm, Steffen Oeltze-Jafra, Kai Lawonn, Katrin Hegenscheid, Henry V{\"o}lzke, Bernhard Preim}
 \authorfooter{
@@
 \item
- $^1$Otto-von-Guericke University Magdeburg, Germany.
+ Paul Klemm, Steffen Oeltze-Jafra, Kai Lawonn, Bernhard Preim are with Otto-von-Guericke University Magdeburg, Germany. E-mail: \{klemm,oeltze,lawonn,preim\}@ovgu.de
 \item
- $^2$Ernst-Moritz-Arndt-University Greifswald, Germany.
+ Katrin Hegenscheid, Henry V{\"o}lzke are with Ernst-Moritz-Arndt University Greifswald, Germany. E-mail: \{katrin.hegenscheid,voelzke\}@uni-greifswald.de
 }
+
+% \author{Paul Klemm$^{1}$, Steffen Oeltze-Jafra$^{1}$, Kai Lawonn$^{1}$, Katrin Hegenscheid$^{2}$, Henry V{\"o}lzke$^{2}$, Bernhard Preim$^{1}$}
+% \authorfooter{
+% %% insert punctuation at end of each item
+% \item
+%  $^1$Otto-von-Guericke University Magdeburg, Germany.
+% \item
+%  $^2$Ernst-Moritz-Arndt-University Greifswald, Germany.
+% }
 
@@
 %
-Also, the introduction of preventive measures in medicine and beyond is based on epidemiological research, where, for example, subgroups with increased risk are identified \cite{Fletcher2012}.
+The introduction of preventive measures in medicine and beyond is also based on epidemiological research, where, for example, subgroups with increased risk are identified \cite{Fletcher2012}.
 %
@@
 %
-Methods like regression analysis are employed to check the attribute list for statistical soundness.
-%
-\\\\
+Methods like regression analysis check the attribute list for statistical soundness.
+%
+%\\\\
+
 Longitudinal population-based studies, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}, gather as much information as possible about a defined sample of people (a \emph{cohort}).
@@
 %
-We implemented the presented methods using modern web technologies to make them easily accessible for the domain experts to enable a fast feedback loop.
+The method implementation is web based to allow a fast feedback loop with domain experts.
+%We implemented the presented methods using modern web technologies to make them easily accessible for the domain experts to enable a fast feedback loop.
+%We implemented the presented methods using modern web technologies to make them easily accessible for the domain experts to enable a fast feedback loop.
 %We implemented the presented methods using modern web technologies, such as \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}, making them easily accessible for the domain experts to enable a fast feedback loop.
@@
 
-In this section we describe the epidemiological workflow and associated requirements.
+%In this section we describe the epidemiological workflow and requirements.
+This is about the epidemiological workflow and requirements.
 

### 2014-06-27 draft.tex
Included revised version from Steffi

@@
 %
-Cohort studies comprise of heterogenous variables describing the medical condition as well as demographic and lifestyle factors and more recently, medical image data.
+Cohort studies comprise of heterogenous variables describing the medical condition as well as demographic and lifestyle factors and, more recently, medical image data.
 %
@@
 %
-We focus, on the potential of image data and attempt to support \emph{hypothesis generation}.
+We focus on the potential of image data and attempt to support \emph{hypothesis generation}.
 
@@
 %
-Age for example is divided into groups (e.g. in 20 year steps) when investigating its influence.
+Age, for example, is divided into groups (e.g. in 20 year steps) when investigating its influence.
 %
@@
 %
-Parameter correlation can also be associated with confounding, which can not be automatically predicted. 
+Parameter correlation can also be associated with confounding, which cannot be automatically predicted. 
 %
@@
 
-Steenwijk et al. \cite{Steenwijk2010} propose a relational database to organize cohort study data for a visual analysis based on linked views such as parallel coordinates, scatterplots and time plots.
+Steenwijk et al.
+\cite{Steenwijk2010} propose a relational database to organize cohort study data for a visual analysis based on linked views such as parallel coordinates, scatter plots and time plots.
 %
@@
 %
-Brain image data was integrated into the analysis by first, segmenting brain regions and tracking neural pathways and then, deriving attributes from both, e.g., volume and fractional anisotropy.
+Brain image data was integrated into the analysis by first segmenting brain regions and tracking neural pathways and then deriving attributes from both, e.g., volume and fractional anisotropy.
 %
@@
 %
-Due to the short paper length, detail is missing on the data type and their algorithm of identifying similar subjects works or what and if they employ statistical measures.
+Due to the short paper length, detail is missing on the data type and their algorithm of identifying similar subjects works or whether they employ statistical measures.
 %
@@
 %
-Histograms, bar charts, scatterplots and heat maps are used to visualize variable combinations with regard to their type.
+Histograms, bar charts, scatter plots and heat maps are used to visualize variable combinations with regard to their type.
 %
@@
 Dai et al.
-\cite{Dai2005} explored risk factors by incorporating choropleth maps of epidemiological variables (e.g., mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines.
+\cite{Dai2005} explored risk factors by incorporating choropleth maps of epidemiological variables (e.g., mortality rates in a region) with parallel coordinates, bar charts and scatter plots with integrated regression lines.
 %
@@
 %
-While their goal to identify possible risk factors using socio-economic and health data is similar to ours, they focus on iteratively refining defined hypothesis and on geographical data.
+While their goal to identify possible risk factors using socio-economic and health data is similar to ours, they focus on iteratively refining defined hypotheses and on geographical data.
 %
@@
 %The connection of time series plots with histograms and heat maps allows for fast insight into complex relationships and 
-While the work shows, how different visualization techniques provide insight into these data sets, it focuses on the time aspect, which is not present for our data.
+While the work shows how different visualization techniques provide insight into these data sets, it focuses on the time aspect, which is not present for our data.
 %
@@
 \paragraph{Prior Work.}
-Klemm et al. \cite{Klemm2013VMV} visualized lumbar spine variabilities based on a semi-automatic shape detection algorithm of 490 participants of the \texttt{SHIP-2} cohort.
+We \cite{Klemm2013VMV} visualized lumbar spine variabilities based on a semi-automatic shape detection algorithm of 490 participants of the \texttt{SHIP-2} cohort.
 %
@@
 %
-As proof of concept, a relation between the size of the segmented shape and measured size of the subjects was shown.
-%
-This work focuses on incorporating these derived data as new variables, making it possible to include it into the hypothesis validation and generation process.
-%
-When applying clustering techniques to the non-image data it was found that \texttt{k-Prototypes} and \texttt{DBSCAN} are appropriate, but is strongly dependent on the chosen variables and distance measure \cite{Klemm2014BVM}.
+As proof of concept, a relation between the size of the segmented shape and the measured size of the subjects was shown.
+%
+This work focuses on incorporating these derived data as new variables, enabling to include it into the hypothesis validation and generation process.
+%
+When applying clustering techniques to the non-image data it was found that \texttt{k-Prototypes} and \texttt{DBSCAN} are appropriate, but are strongly dependent on the chosen variables and distance measures \cite{Klemm2014BVM}.
 %
@@
 %
-They define a model as a representation of system entities, phenomena and processes and hypothesis as models whose outcomes are not compared with real-world data (\emph{validation}).
-%
-The VA-process is also reflected in our \emph{IVA} loop.
+They define a model as a representation of system entities, phenomena, processes and hypotheses as models whose outcomes are not compared with real-world data (\emph{validation}).
+%
+The VA process is also reflected in our \emph{IVA} loop.
 % --- PARAGRAPH ---
@@
 %
-They point out the cost-benefit-tradeoff introduced by the cognitive overhead by mentally connecting multiple views over more complex single views.
+They point out the cost-benefit tradeoff introduced by the cognitive overhead by mentally connecting multiple views over more complex single views.
 %introduced by tasks, such as view comparison or context switching.
@@
 %\section{Interactive Visual Analyis in Cohort Study Data}
-\section{Image Centric Cohort Study Data in Interactive Visual Analysis Context} \label{Image Centric Cohort Study Data in Interactive Visual Analysis Context}
+\section{Image-Centric Cohort Study Data in Interactive Visual Analysis Context} \label{Image Centric Cohort Study Data in Interactive Visual Analysis Context}
 We described the epidemiological workflow and emphasized the reproducibility and statistical integrity (recall Subsection~\ref{EpidemiologicalWorkflow}).
@@
 %Introducing the \emph{IVA} principle to the epidemiological domain aims to compensate its weaknesses rather than replacing the existing workflow.
-Introducing the \emph{IVA} principle to the epidemiological domain aims to compensate the weaknesses of the existing workflow, rather than replacing it (recall Fig.~\ref{fig:WorkflowComparison}).
+Introducing the \emph{IVA} principle to the epidemiological domain aims to compensate the weaknesses of the existing workflow rather than replacing it (recall Fig.~\ref{fig:WorkflowComparison}).
 %
@@
 %
-Considering more complex parts of the data would make those results more trustworthy and also could identify possible anatomical confounders--an epidemiolgical research result in itself.
+Considering more complex parts of the data would make those results more trustworthy and could also identify possible anatomical confounders--an epidemiolgical research result in itself.
 \\\\
@@
 %
-These associations may also be summarized using pivot tables which are widely used in epidemiology.
+These associations may also be summarized using pivot tables, which are widely used in epidemiology.
 
@@
 %
-This is either done manually by experts setting landmarks or by a (semi-)automatic detection, registration and segmentation.
+This is either done manually by experts setting, landmarks or by a (semi-)automatic detection, registration and segmentation.
 %
@@
 %
-Morphometric variables are derived to allow for statistical comparison of the tissue, which incorporates mostly positions, diameters, volumes and relative distances and alignment to other structures.
+Morphometric variables are derived to allow for statistical comparison of the tissue, which incorporates mostly positions, diameters, volumes, relative distances and alignment to other structures.
 %-------------------------------- Figure --------------------------------
@@
  %
- The data are visualized in the range and domain space, which can be brushed, yielding new groups, to be investigated using further variables.
+ The data are visualized in the range and domain space, which can be brushed, yielding new groups to be investigated using further variables.
  %
@@
  %
- Distance to the mean mesh of subjects with thyroid disorder is encoded as red for x axis, blue to y axis and green to z axis.
+ Distance to the mean mesh of subjects with thyroid disorder is encoded as red for x axis, blue y axis and green z axis.
  %The lumbar spine models encode distance to the mean model using  \emph{color} 
@@
 %-------------------------------- /Figure -------------------------------
-Our proposed \emph{IVA} workflow consists of three major steps as illustrated in Figure~\ref{fig:InteractionLoop}: Variable selection, visualization and brushing.
-%
-A hypothesis-driven analysis usually starts with the selection of variables, or shape groups derived from a shape-based clustering.
+Our proposed \emph{IVA} workflow consists of three major steps, as illustrated in Figure~\ref{fig:InteractionLoop}: Variable selection, visualization and brushing.
+%
+A hypothesis-driven analysis usually starts with the selection of variables or shape groups derived from a shape-based clustering.
 %
@@
 %
-In the following sections we present the different parts of system.
+In the following sections we present the different parts of the system.
 % \subsection{System Structure and Design} \label{Structure and Workflow}
@@
 %
-Hence we build our system using web technologies.%, described in detail in Section~\ref{implementation}.
+Hence, we build our system using web technologies. %, described in detail in Section~\ref{implementation}.
 %
@@
 %
-Classical UI elements, such as the menu bar or custom right click menus, are technically possible, but not common in this domain.
+Classical UI elements, such as the menu bar or custom right-click menus, are technically possible, but not common in this domain.
 %
@@
 		\item The \emph{sidebar}, which contains all epidemiological variables.
-As cluster results group variables like categorical variables and are part of the sidebar as well (Fig.~\ref{fig:visualization}~a).
+The cluster results group variables like categorical variables and are part of the sidebar as well (Fig.~\ref{fig:visualization}~a).
 	%
@@
 %
-This idea was soon dropped since domain experts reported a cluttered workspace, which required a lot of scrolling.
+This idea was soon dropped, since domain experts reported a cluttered workspace, which required a lot of scrolling.
 %
@@
 %
-All user-generated visualizations follow are part of the \emph{canvas} and can be arranged freely.
+All user-generated visualizations are part of the \emph{canvas} and can be arranged freely.
 % are the  was to keep the concept of free arrangement for user-added 
@@
  %Mouse-over focuses a entry and highlights the feature names for better readability.
- Mouse-over an entry enlarges the variable names for better readability.
- %
+ Moving the mouse over an entry enlarges the variable names for better readability.
+ % Potential Cut here
  The enlarged excerpt shows associations for shape clusters of subjects with and without diagnosed spine attrition, which show associations between gender, weight, body height and smoking behavior.
@@
 %
-Continuous data can be visualized using scatterplots (two variables) or parallel coordinates (multiple variables), but in epidemiology, this data type is usually categorized into ordinal groups of \emph{equal size}.
+Continuous data can be visualized using scatter plots (two variables) or parallel coordinates (multiple variables), but in epidemiology, this data type is usually categorized into ordinal groups of \emph{equal size}.
 %
@@
 %
-Therefore we follow the convention to use bins of equal size.
-
-%
-%
-Following Tufte's concept of \emph{small multiples} \cite{Tufte1983}, information derived from the medical image data are directly incorporated into the plot by including color-coded mean shapes for each manifestation (Figure~\ref{fig:visualization}~b).
+Therefore, we follow the convention to use bins of equal size.
+%
+
+Following Tufte's concept of \emph{small multiples} \cite{Tufte1983}, information derived from the medical image data are incorporated into the plot by including color-coded mean shapes for each manifestation (Figure~\ref{fig:visualization}~b).
 %
@@
 %
-Until now they were not able rapidly validate shape differences based on non-image variables.
+Until now, epidemiologists were not able to rapidly validate shape differences based on non-image variables.
 %
@@
 %
-Brushed groups act as reference for the shape-visualization, calculating distances based on the mean-shape of the brushed selection.
+Brushed groups act as reference for the shape visualization, calculating distances based on the mean-shape of the brushed selection.
 %
@@
 %Brush selections are propagated to all visualizations allowing for fast feature querying.
-
 \paragraph{Pivot Tables.}
@@
 %
-To allow better comparison with respect to variables, the values of each cell can also be displayed as percentage of the variable represented of either the row or column.
+To allow for better comparison with respect to variables, the values of each cell can also be displayed as percentage of the variable represented of either the row or column.
 
@@
 %\rem{As previously discussed}
-Highlighting potentially interesting values in the data set is one major benefits of the \emph{IVA}-powered approach and belongs to the \emph{multivariate analysis} pattern,  analyzing variables outside the shape space.
+Highlighting potentially interesting values in the data set is one major benefits of the \emph{IVA}-powered approach and belongs to the \emph{multivariate analysis} pattern for analyzing variables outside the shape space.
 %
@@
 %
-$V$ yields values between $0$, meaning that two variables are completely independent, and $1$ indicating they are the same.
+$V$ yields values between $0$, meaning that two variables are completely independent, and $1$ indicates that they are the same.
 %
 %Since \emph{Cram\'{e}r's V} is always positive, 
-\emph{Cram\'{e}r's V} does not allow to infer about dependency direction.
+\emph{Cram\'{e}r's V} does not allow to infer the dependency direction.
 %
@@
 %
-This aims to highlight variables possibly associated with the focused hypothesis as well as trigger new hypotheses.
+This aims to highlight variables possibly associated with the focused hypothesis and to trigger new hypotheses.
 %
@@
 %
-The distinction whether an association is a confounder or an effect, depends on the context defined by the hypothesis and is a decision to be made by the domain expert.
-%
-The contingency matrix visualization is an \emph{overview visualization}, something the epidemiological community lacks and is in great need of.
+The distinction whether an association is a confounder or an effect depends on the context defined by the hypothesis and is a decision to be made by the domain expert.
+%
+The contingency matrix visualization is an \emph{overview visualization}--something the epidemiological community lacks and is in great need of.
 
@@
 %
-Very often are high correlations obvious, such as gender with menstrual status.
+Very often, high correlations are obvious, such as gender with menstrual status.
 %
@@
 %
-Clustering result are listed in their own category in the \emph{sidbar}.
+Clustering results are listed in their own category in the \emph{sidbar}.
 %Since a clustering process can take up to ten seconds, a status indicator at the bottom of the screen keeps the user informed.
@@
  %
- All computation-heavy operations, such as calculation of mean shapes or distances, are performed on the server-side. 
+ All computation-heavy operations, such as calculation of mean shapes or distances, are performed on the server side. 
  %All computation-heavy operations, such as calculation of mean shapes or distances, as well as statistical processing are done by the server to keep hardware requirements of client systems low. 
@@
 \end{figure}
-In this section we discuss how we implemented the presented methods using open web standards.
+In this section, we discuss how we implemented the presented methods using open web standards.
 %
@@
 	%
-	High confidentiality standards of the data are met by password protecting the access.%restricting access via an account system.
+	High confidentiality standards of the data are met by a password protecting the access.%restricting access via an account system.
 	\item Recent developments in \emph{WebGL} applications running in browsers with near-native performance result in many open source libraries, which are well documented and driven by active communities.
@@
 %
-Information visualizations such as scatterplots and bar charts are created using the popular \texttt{Data-Driven Documents (D3.js)} library \cite{D3}, which works well for attaching data to visible elements like vector graphics and provides powerful transformation and mapping tools.
+Information visualizations such as scatter plots and bar charts are created using the popular \texttt{Data-Driven Documents (D3.js)} library \cite{D3}, which works well for attaching data to visible elements like vector graphics and provides powerful transformation and mapping tools.
 %
@@
 %
-The result is read from the console standard out and returns it to the client.
+The result is read from the console output and is returned to the client.
 %
@@
 %
-Therefore some manifestations are sparsely populated, which makes statistical evaluation challenging.
+Therefore, some manifestations are sparsely populated, which makes statistical evaluation challenging.
 %
@@
 %
-Each variable is stored as an object containing the data as array it's data type, a detailed description and data dictionary translating value or error IDs to values.
+Each variable is stored as an object containing the data as array, its data type, a detailed description and data dictionary translating value or error IDs to values.
 % \begin{itemize}
@@
 %
-We obtained and worked with 2.540 tetrahedron models of the lumbar spine.
+We obtained and worked with 2,540 tetrahedron models of the lumbar spine.
 %
@@
  %
- (b) Clustering of all subjects yields seven groups, where Cluster~4 assembles the mean. The light blue bars indicate the share of females in the group.
- %
- (c) A mosaic plot mapping age against dichotomous questionnaire answer to ``\emph{Did you experience back pain in the past three months?}''.
+ (b) Clustering of all subjects yields seven groups, whereas Cluster~4 assembles the mean. The light blue bars indicate the share of females in the group.
+ %
+ (c) A mosaic plot mapping age against the dichotomous questionnaire answer to ``\emph{Did you experience back pain in the past three months?}''.
  (d) Clustering result of ``\emph{Did you experience back pain in the past three months?}'' Yes/no with female share in each group. 
@@
 %
-Shape distance between meshes is mapped to color (recall Fig.~\ref{fig:visualization}).
+The shape distance between meshes is mapped to color (recall Fig.~\ref{fig:visualization}).
 %
@@
 %
-This approach aims to characterize the systems ability to explore data, discover knowledge, generate hypotheses and help formulating decisions.
+This approach aims to characterize the system's ability to explore data, discover knowledge, generate hypotheses and help formulating decisions.
 %
 Since it is hard to quantify these outcomes, Lam et al.
-suggest case studies for the \emph{VDAR} by applying the think aloud protocol to understand the domain experts observations, inferences and conclusions when using the system.
+suggest case studies for the \emph{VDAR} by applying the think-aloud protocol to understand the domain expert's observations, inferences and conclusions when using the system.
 
@@
 %
-Communication was enabled via webcam supported voice over ip.
+Communication was enabled via webcam-supported voice over ip.
 %
@@
 %
-Video-recordings of the sessions allowed a detailed evaluation afterwards.%made a possible to evaluate the sessions later on.
+Video recordings of the sessions allowed a detailed evaluation afterwards.%made a possible to evaluate the sessions later on.
 %
 \paragraph{Procedure.}
-At first, we controlled mouse and keyboard of the participants PC and demonstrated the basic functionalities of the system.
+At first, we controlled mouse and keyboard of the participants' PC and demonstrated the basic functionalities of the system.
 %
@@
 %
-Performing a \emph{local investigation} by clustering all subjects or use the contingency matrix for a \emph{multivariate analysis}.
+Performing a \emph{local investigation} by clustering all subjects or using the contingency matrix for a \emph{multivariate analysis}.
 %
@@
 %They pointed out the benefits of the possibility to look at all variables in the context of each other.
-Before, they were not able to to look at all variables in the context of each other.
+Before, they were not able to look at all variables in the context of each other.
 %
@@
 
-After this initial overview, KH performed a \emph{multivariate analysis} by introducing variables, such as \emph{age}, \emph{waist circumference}, \emph{weight} or \emph{lumbar spine canal curvature} as bar charts views into the canvas area and selected subgroups to see how they are distributed and if they could observe unusual behavior in the mean-shapes.
+After this initial overview, KH performed a \emph{multivariate analysis} by introducing variables, such as \emph{age}, \emph{waist circumference}, \emph{weight} or \emph{lumbar spine canal curvature} as bar charts views into the canvas area and selected subgroups to see how they are distributed and if they could observe unusual behavior in the mean shapes.
 %
@@
 %
-HV started also with a \emph{multivariate analysis} using the contingency matrix to analyze non-image variables, such as age-associated parameters like \emph{income}, \emph{blood fat values} or \emph{number of born children}, but found no associations of interest.
-%
-Therefore, he applied the \emph{local investigation} pattern by a shape grouping step using shape-based clustering via dragging the \emph{All subjects} from the sidebar into the canvas area, triggering the shape clustering (Fig.~\ref{fig:application} b).
+HV also started with a \emph{multivariate analysis} using the contingency matrix to analyze non-image variables, such as age-associated parameters like \emph{income}, \emph{blood fat values} or \emph{number of born children}, but found no associations of interest.
+%
+Therefore, he applied the \emph{local investigation} pattern by a shape grouping step using shape-based clustering via dragging \emph{All subjects} from the sidebar into the canvas area, triggering the shape clustering (Fig.~\ref{fig:application} b).
 %Shape-based hypothesis-free exploration starts with the \emph{local investigation} pattern by a shape grouping step using shape-based clustering.
@@
 %
-Other shapes differ with respect to size, such as cluster 2, 3, 7, where the last one and cluster 5 also represent a more straight spine, which is usual for subjects with larger body size.
+Other shapes differ with respect to size, such as cluster 2, 3, 7, whereas the last one and cluster 5 also represent a more straight spine, which is usual for subjects with larger body size.
 %
@@
 %To perform an automatic feature suggestion using a \emph{multivariate analysis} by looking at \emph{Cram\'{e}r's V} contingency values of in the contingency pane.
-To get an overview over the suggested variables, the user opened the contingency pane to perform a \emph{multivariate analysis} by looking at \emph{Cram\'{e}r's V} contingency values of all clusters, revealing a strong correlation with \emph{gender} and \emph{body size}.
-%
-Therefore another \emph{multivariate analysis} was carried out by the variable \emph{gender} to the canvas from the contingency pane and selecting all female subjects (Fig.~\ref{fig:application}~b).
+To get an overview of the suggested variables, the user opened the contingency pane to perform a \emph{multivariate analysis} by looking at \emph{Cram\'{e}r's V} contingency values of all clusters, revealing a strong correlation with \emph{gender} and \emph{body size}.
+%
+Therefore, another \emph{multivariate analysis} was carried out by the variable \emph{gender} to the canvas from the contingency pane and by selecting all female subjects (Fig.~\ref{fig:application}~b).
 %
@@
 %
-It was found, that cluster~1 subjects report heavy headaches more frequently than other subjects.
+It was found that cluster~1 subjects report heavy headaches more frequently than other subjects.
 %Another \emph{multivariate analysis} using a pivot table set gender and employment status in relation to clustering affiliation\add{.} 
@@
 %
-With the presented \emph{IVA} approaches they were quickly able to confirm medical knowledge as well as elaborate new hypothesis.
-%
-We observed, that the domain experts are more likely interested in variables they are familiar with and have personal clinical experience with.
+With the presented \emph{IVA} approaches they were quickly able to confirm medical knowledge and to elaborate new hypotheses.
+%
+We observed that the domain experts are more likely interested in variables they are familiar with and have personal clinical experience with.
 %\com{00.44 Clustering-Ergebnis Weiblich und Arbeitspensum Alter}
@@
 %
-The starting point follows the \emph{feature localization} pattern, where a variable of interest is selected by dragging it into the canvas area and viewing the subject's distribution as well as their shape differences.
+The starting point follows the \emph{feature localization} pattern, where a variable of interest is selected by dragging it into the canvas area and viewing the subjects' distribution as well as their shape differences.
 % \paragraph{Analysis 1.}
@@
 %
-As seen in Figure~\ref{fig:application}~(c), subjects reporting back pain are the majority, which makes it difficult to extract parameters which reliably describe back pain.
+As seen in Figure~\ref{fig:application}~(c), subjects reporting back pain are the majority, which makes it difficult to extract parameters that reliably describe back pain.
 %
@@
 %
-\emph{Body height} was explained as a confounder for \emph{gender}, since female subjects are smaller on average than male subjects.
-%
-The analysis solely based on shape accentuated body height differences in \emph{gender}, which clouded the differences of \emph{back pain}.
+\emph{Body height} was explained as a confounder for \emph{gender}, since female subjects on average are smaller than male subjects.
+%
+The analysis solely based on shape-accentuated body height differences in \emph{gender}, which clouded the differences of \emph{back pain}.
 %This makes it hard to analyze shape related features for this disease
@@
 %
-The previously discussed analysis puts the suitability of the lumbar spine segmentation for analyzing back pain into question, leading to this analysis.
-%
-Therefore the variable ``\emph{Did you experience back pain in the past three months?}'' is dropped into the canvas area. %, as seen in previous Fig.~\ref{ToDoA}.
+The previously discussed analysis questions the suitability of the lumbar spine segmentation for analyzing back pain, leading to this analysis.
+%
+Therefore, the variable ``\emph{Did you experience back pain in the past three months?}'' is dropped into the canvas area. %, as seen in previous Fig.~\ref{ToDoA}.
 %
@@
 %
-While the clustering algorithm finds only three homogenous clusters close to the mean shape for subjects reporting no back pain.
+The clustering algorithm finds only three homogenous clusters close to the mean shape for subjects reporting no back pain.
 %
@@
 %
-It shows, that subjects in Cluster~1, 2 and 6 reported strong back pain, while at the same time they also have a considerably higher share of females.
+It shows that subjects in Cluster~1, 2 and 6 reported strong back pain, while at the same time they also have a considerably higher share of females.
 %
@@
 %
-For the experts, these observations are a starting point for a number of new hypothesis about possible relationships, for example association between overweight and Cluster~1.
+For the experts, these observations are a starting point for a number of new hypotheses about possible relationships, for example the association between overweight and Cluster~1.
 %
 \\\\
-In summary it can be stated that hypothesis-driven analysis leads to hypothesis generation by design of the framework.
-%
-It is not suited and intended to statistically validate hypothesis like statistical processors.
+In summary, it can be stated that the hypothesis-driven analysis leads to hypothesis generation by design of the framework.
+%
+It is not suited and intended to statistically validate hypotheses like statistical processors.
 %
@@
 %
-Both domain experts concluded positively over the \emph{IVA} approach. %when analyzing image-based cohort study data.
-%
-KH emphasized the way the image data are included into information visualizations which comes much more natural to her due to her background in radiology.
+Both domain experts rated the \emph{IVA} approach positively. %when analyzing image-based cohort study data.
+%
+KH emphasized the way the image data are included into information visualizations, which comes much more natural to her due to her background in radiology.
 %
@@
 \textbf{Multivariate analysis is most important for hypothesis generation.}
-Both experts emphasized the potential of the \emph{multivariate analysis} capabilities of the contingency matrix for getting insight into a large amount of variables simultaneously.
+Both experts emphasized the potential of the \emph{multivariate analysis} capabilities of the contingency matrix for gaining insight into a large amount of variables simultaneously.
 %
@@
 \textbf{Web technologies are well suited for rapid feedback.}
-The web-based approach for both implementing the prototype as well as getting feedback via voice over ip conference calls worked very well.
+The web-based approach for both implementing the prototype and getting feedback via voice over ip conference calls worked very well.
 %
@@
 %
-This approach is well suited for the \emph{VDAR} approach to assess user thought processes using the think aloud technique.
+This approach is well suited for the \emph{VDAR} approach to assess user thought processes using the think-aloud technique.
 %The \emph{VDAR} approach to assess user thought processes using the think aloud technique worked well with the web-based approach for both implementing the prototype as well as getting feedback via voice over ip conference calls worked very well.
@@
 %
-Hence, the framework allows for both, hypothesis-driven analysis and hypothesis generation.
+Hence, the framework allows both for hypothesis-driven analysis and hypothesis generation.
 %
@@
 %
-Shape-based clustering assesses the variability of an anatomical structure in the context of non-image variables such as disease indicators or lifestyle factors.
+Shape-based clustering assesses the variability of an anatomical structure in the context of non-image variables, such as disease indicators or lifestyle factors.
 
@@
 %\com{This needs to be adjusted to fit the new Application section}
-Findings from analyzing lumbar back pain using the \emph{IVA} approach range from deriving shape-based groups of subjects to detailed description of variables potentially associated with the disease, such as waist circumference, alcohol consumption and attentiveness disorder.
+Findings from analyzing lumbar back pain using the \emph{IVA} approach range from deriving shape-based groups of subjects to detailed descriptions of variables potentially associated with the disease, such as waist circumference, alcohol consumption and attentiveness disorder.
 %

### 2014-06-27 draft.tex
Revised Version from Steffen, new Style

@@
 %
-Cohort studies comprise of heterogenous variables describing the medical condition as well as demographic and lifestyle factors and, more recently, medical image data.
+Cohort studies comprise heterogenous variables describing the medical condition as well as demographic and lifestyle factors and, more recently, medical image data.
 %
@@
 %
-For instance, segmentation of the image data is required for an analysis of anatomical structure \com{and of possible} correlations between this structure and epidemiological factors.
+For instance, segmentation of the image data is required for an analysis of anatomical structure and of possible correlations between this structure and epidemiological factors.
 %
@@
 
-Angelelli et al. \cite{Angelelli2014} focus on a data organization for the interactive visual analysis of heterogeneous cohort study data.
-%
+Angelelli et al. \cite{Angelelli2014} focus on the data organization for an interactive visual analysis of heterogeneous cohort study data.
+%
+The proposed data-cube model facilitates the seamless integration of image-based and non-image data.
 %\rem{They propose a data-cube based model able to handle partially overlapping data subsets during the interactive visualization.}
 %
-Brain image data was integrated into the analysis by first segmenting brain regions and tracking neural pathways and then deriving attributes from both, e.g., volume and fractional anisotropy.
-%
-The data-cube model facilitates the seamless integration of heterogeneous image-based and non-image data.
-%
-A multiple coordinated view framework linked spatial and non-spatial data views.
+In a demonstration of the model, brain image data was integrated into the analysis by first segmenting brain regions and tracking neural pathways and then deriving attributes from both, e.g., volume and fractional anisotropy.
+%
+A multiple coordinated view framework then linked spatial and non-spatial data views.
 %
@@
 %
-Due to the short paper length, detail is missing on the data type and their algorithm of identifying similar subjects works or whether they employ statistical measures.
+Due to the short paper length, detail is missing on the data types and their algorithms of identifying similar subjects' works or whether they employ statistical measures.
 %
@@
 %\section{Interactive Visual Analyis in Cohort Study Data}
-\section{Image-Centric Cohort Study Data in Interactive Visual Analysis Context} \label{Image Centric Cohort Study Data in Interactive Visual Analysis Context}
+\section{Image-Centric Cohort Study Data in an Interactive Visual Analysis Context} \label{Image Centric Cohort Study Data in Interactive Visual Analysis Context}
 We described the epidemiological workflow and emphasized the reproducibility and statistical integrity (recall Subsection~\ref{EpidemiologicalWorkflow}).
@@
 %
-The suitability of visualization techniques for epidemiological data depends on its ability to compare multiple data variables while highlighting associations.
+The suitability of visualization techniques for epidemiological data depends on their ability to compare multiple data variables while highlighting associations.
 %
@@
 %
-In the following sections we present the different parts of the system.
+In the following sections we present the different parts of our system.
 % \subsection{System Structure and Design} \label{Structure and Workflow}
@@
 %
-Hence, we build our system using web technologies. %, described in detail in Section~\ref{implementation}.
+Hence, we built our system using web technologies. %, described in detail in Section~\ref{implementation}.
 %
@@
 %
-Since the previously described \emph{IVA} workflow allows for many different ways to analyze the data, we tried to make the interface as minimalistic as possible, treating the resulting space as \emph{canvas} for the data.
+Since the previously described \emph{IVA} workflow allows for many different ways to analyze the data, we designed the interface as minimalistic as possible, treating the resulting space as \emph{canvas} for the data.
 %
 % \subsubsection{System Structure}
-We divide the workspace into four major parts, as illustrated in Figure~\ref{fig:visualization} and \ref{fig:similarity}.
+We divided the workspace into four major parts, as illustrated in Figure~\ref{fig:visualization} and \ref{fig:similarity}.
 \begin{itemize} \itemsep0.2em
@@
  The enlarged excerpt shows associations for shape clusters of subjects with and without diagnosed spine attrition, which show associations between gender, weight, body height and smoking behavior.
+ %
+ The contingency pane is not shown here.
  }
@@
 %
-Until now, epidemiologists were not able to rapidly validate shape differences based on non-image variables.
+Until now, epidemiologists were not able to inspect shape differences based on non-image variables.
 %
@@
 %\rem{As previously discussed}
-Highlighting potentially interesting values in the data set is one major benefits of the \emph{IVA}-powered approach and belongs to the \emph{multivariate analysis} pattern for analyzing variables outside the shape space.
-%
-Turkay et al. \cite{Turkay2013} used the approach to calculate key figures based on the distribution functions of each variable derived from the image data.
+Highlighting potentially interesting associations in the data set is one major benefits of the \emph{IVA}-powered approach and is part of the \emph{multivariate analysis} pattern for analyzing variables outside the shape space.
+%
+Turkay et al. \cite{Turkay2013} used the approach to calculate statistical key figures based on the distribution functions of each variable derived from the image data.
 %
@@
 %
-The expected counts in the contingency table have to be larger than 5 for $80\%$ of the entries and no expected value must be smaller than one \cite{Cochran1952}.
+The expected counts in the contingency table have to be larger than five for $80\%$ of the entries and no expected value must be smaller than one \cite{Cochran1952}.
 %
@@
 
-Subject clustering is triggered automatically as \emph{local investigation} for each variable after it was added to the canvas by the user.
+Subject clustering is triggered automatically as \emph{local investigation} for a variable after it was added to the canvas by the user.
 %
@@
 	High confidentiality standards of the data are met by a password protecting the access.%restricting access via an account system.
-	\item Recent developments in \emph{WebGL} applications running in browsers with near-native performance result in many open source libraries, which are well documented and driven by active communities.
+	\item Recent developments in \emph{WebGL} applications running in browsers with near-native performance results in many open source libraries, which are well documented and driven by active communities.
 We use \emph{WebGL} for rendering shape information.
@@
 \end{itemize}
-%These advantages do not come without drawbacks.
-%
-However, sophisticated libraries/languages, such as the \texttt{Visualization Toolkit}\footnote{Developed by Kitware Inc; \href{http://vtk.org}{\texttt{vtk.org}}} or \texttt{R}\footnote{Open Source; \href{http://r-project.org}{\texttt{r-project.org}}} for statistics, are either not available at all or only accessible through complex client-server systems.
+These advantages do not come without drawbacks.
+%
+Sophisticated libraries/languages, such as the \texttt{Visualization Toolkit}\footnote{Developed by Kitware Inc; \href{http://vtk.org}{\texttt{vtk.org}}} or \texttt{R}\footnote{Open Source; \href{http://r-project.org}{\texttt{r-project.org}}} for statistics, are either not available at all or only accessible through complex client-server systems.
 %
@@
 \subsubsection{Data Preprocessing} \label{application:Data Preprocessing}
-The data are preprocessed as described in Section~\ref{Data Preprocessing}.
+The data processing follows the description in Section~\ref{Data Preprocessing}.
 %
@@
 %
-According to epidemiological publications, we set the number of groups to five, to allow for contingency assessment.
+Following epidemiological publications, we set the number of groups to five, to allow for contingency assessment.
 
@@
 %
-It uses a weighted sum of T1- and T2-weighted MR images to detect the lumbar spine shape.
-%
-The registered models capture resilient information about the shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae \cite{Klemm2013VMV}.
+The model uses a weighted sum of T1- and T2-weighted MR images to detect the lumbar spine shape.
+%
+Once registered, it captures information about the shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae \cite{Klemm2013VMV}.
 %
@@
 %
-For clustering, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature \cite{Klemm2013VMV}).
+For clustering, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature)  \cite{Klemm2013VMV}.
 
@@
 %The experts accessed the prototype by simply entering the website link into their browser of choice, no exchange of software was required.
-The experts accessed the prototype by entering the website link into their browser. %of choice, no exchange of software was required.
+The experts accessed the prototype by entering the weblink into their browser. %of choice, no exchange of software was required.
 %Since our prototype ran on a local server, simply needed the homepage link to access the prototype.
@@
 %
-Performing a \emph{local investigation} by clustering all subjects or using the contingency matrix for a \emph{multivariate analysis}.
-%
-The latter was chosen first by both experts.
+Performing a \emph{multivariate analysis} by viewing the contingency matrix or a shape grouping step using shape-based clustering.
+%Performing a \emph{local investigation} by clustering all subjects or using the contingency matrix for a \emph{multivariate analysis}.
+%
+The first was chosen by both experts.
 % The latter was chosen first by the domain expert with radiological background.
@@
 %
-To cite one expert, the contingency matrix ``illumunates the data black box'', making it possible to look at the data unbiased from assumptions.
+To cite one expert, the contingency matrix ``illumunates the data black box'', making it possible to look at the data unbiased of assumptions.
 
@@
 %
-The radiologist (KH) was looking for correlations with shape-related variables in the data, finding that they correlate with \emph{leg pain}, \emph{age}, \emph{body height} and \emph{hormone replacement therapy status}.
-%
-Due to the dense mapping of information in the contingency matrix it was suggested to make this visualization full screen.
+The radiologist (KH) was looking for correlations with shape-related variables in the data, finding that \emph{spine curvature} correlates with \emph{leg pain}, \emph{age}, \emph{body height} and \emph{hormone replacement therapy status}.
+%The radiologist (KH) was looking for correlations with the image derived \emph{lumbar spine curvature} variable, finding a correlation with \emph{leg pain}, \emph{age}, \emph{body height} and \emph{hormone replacement therapy status}.
+%
+Due to the dense mapping of information in the contingency matrix, KH suggested to make this visualization full screen.
 %Many correlations observed in this view
 
-After this initial overview, KH performed a \emph{multivariate analysis} by introducing variables, such as \emph{age}, \emph{waist circumference}, \emph{weight} or \emph{lumbar spine canal curvature} as bar charts views into the canvas area and selected subgroups to see how they are distributed and if they could observe unusual behavior in the mean shapes.
+After this initial overview, KH performed a \emph{multivariate analysis} by introducing variables, such as \emph{age}, \emph{waist circumference}, \emph{weight} or \emph{lumbar spine canal curvature} as bar chart views into the canvas area and selected subgroups to see how they are distributed and if they could observe unusual behavior in the mean shapes.
 %
@@
 %To perform an automatic feature suggestion using a \emph{multivariate analysis} by looking at \emph{Cram\'{e}r's V} contingency values of in the contingency pane.
-To get an overview of the suggested variables, the user opened the contingency pane to perform a \emph{multivariate analysis} by looking at \emph{Cram\'{e}r's V} contingency values of all clusters, revealing a strong correlation with \emph{gender} and \emph{body size}.
-%
-Therefore, another \emph{multivariate analysis} was carried out by the variable \emph{gender} to the canvas from the contingency pane and by selecting all female subjects (Fig.~\ref{fig:application}~b).
+To get an overview of the suggested variables, the user opened the contingency pane (not shown here) to perform a \emph{multivariate analysis} by looking at \emph{Cram\'{e}r's V} contingency values of all clusters, revealing a strong correlation with \emph{gender} and \emph{body size}.
+%
+Therefore, another \emph{multivariate analysis} was carried out by dragging the \emph{gender} variable to the canvas and selecting all female subjects (Fig.~\ref{fig:application}~b).
 %
@@
 
-\subsection{Case 2: Hypothesis-based Analysis}
-If the user proposes a hypothesis about a relation between a non-image variable regarding shape, the workflow slightly differs from the hypothesis-free analysis.
+\subsection{Case 2: Hypothesis-driven Analysis}
+If the user proposes a hypothesis about a relation between a non-image variable and shape, the workflow slightly differs from the hypothesis-free analysis.
 %
@@
 %
-To validate this hypothesis, a \emph{feature localization} was performed by introducing the dichotomous variable ``\emph{Did you experience back pain in the last three months?}'' together with the age as mosaic plot by dropping both variables on the canvas area.
+To validate this hypothesis, a \emph{feature localization} was performed by combining the dichotomous variable ``\emph{Did you experience back pain in the last three months?}'' with age in a mosaic plot by dropping both variables on the canvas area (Fig.~\ref{fig:application}~c).
 %
@@
 %
-Age influences the lumbar spine shape, while the differences between subjects with and without back pain are small (Fig.~\ref{fig:application}~c).
+Age influences the lumbar spine shape, while the differences between subjects with and without back pain are small.
 %
@@
 %
-A \emph{multivariate analysis} using the contingency table showed a strong association between \emph{back pain} with \emph{gender} and \emph{body height}.
+A \emph{multivariate analysis} using the contingency table showed a strong association between \emph{back pain} with both, \emph{gender} and \emph{body height}.
 %
@@
 %
-Therefore, the variable ``\emph{Did you experience back pain in the past three months?}'' is dropped into the canvas area. %, as seen in previous Fig.~\ref{ToDoA}.
-%
-Then, the variables derived from the clustering results for both subject groups are dropped into the canvas area as well (Fig.~\ref{fig:application}~d).
-%
-The clustering algorithm finds only three homogenous clusters close to the mean shape for subjects reporting no back pain.
-%
-The cluster analysis for back pain yields diverse clusters with various pathological shape classes.
-%
-Cluster~5 represents most of the subjects and is very similar to the global mean shape.
+Therefore, the dichotomous variable ``\emph{Did you experience back pain in the past three months?}'' is dropped into the canvas area. %, as seen in previous Fig.~\ref{ToDoA}.
+%
+%The result oh the shape-based clustering for both subject groups is visualized
+Figure~\ref{fig:application}~(d) shows the results of the automatically triggered shape-based clustering for subject with and without back pain.
+%Both subject groups (back pain and no back pain) are automatically clustered. The results are dropped into the canvas area as well (Fig.~\ref{fig:application}~d).
+%A shape-based clustering of both subject groups is performed and visualized as well (Fig.~\ref{fig:application}~d).
+%Both subject groups are clustered regarding lumbar spine shape and visualized as well (Fig.~\ref{fig:application}~d).
+%Then, the variables derived from the clustering results for both subject groups (back pain and no back pain) are dropped into the canvas area as well (Fig.~\ref{fig:application}~d).
+%
+The clustering algorithm finds only three homogenous clusters close to the global mean shape for subjects reporting no back pain.
+%
+The cluster analysis for back pain yields diverse clusters with pathological shape classes.
+%
+Cluster~5 represents most of the subjects and is similar to the global mean shape.
 %

### 2014-06-27 draft.tex
Final changes to revised version. It is done

@@
 %In this section we describe the epidemiological workflow and requirements.
-This is about the epidemiological workflow and requirements.
+%This is about the epidemiological workflow and requirements.
+This section covers the epidemiological workflow and requirements.
 
@@
 %
-Due to the short paper length, detail is missing on the data types and their algorithms of identifying similar subjects' works or whether they employ statistical measures.
+Due to the short paper length, detail is missing on the data types and their algorithms of identifying similar subjects or whether they employ statistical measures.
 %
@@
 %
-This requires a geographic colocation of associated variables.
+It requires a spatial colocation of associated variables.
 %
@@
 \paragraph{Prior Work.}
-We \cite{Klemm2013VMV} visualized lumbar spine variabilities based on a semi-automatic shape detection algorithm of 490 participants of the \texttt{SHIP-2} cohort.
+We visualized lumbar spine variabilities based on a semi-automatic shape detection algorithm of 490 participants of the \texttt{SHIP-2} cohort \cite{Klemm2013VMV}.
 %
@@
  %
- These subjects are shown as shade in the visualization, denoting their share.
+ These subjects are shown as shade in the visualizations, denoting their share.
  %Their shares are denoted in the bar chart as blue and the mosaic plot as dark shade.
@@
 %
-By running the prototypes on server machines, software exchange became as easy as sharing a website link, giving us the opportunity to include the clinical experts in the development process with little effort.
+By running the prototypes on server machines, software exchange became as easy as sharing a weblink, giving us the opportunity to include the clinical experts in the development process with little effort.
 %
@@
 %The result oh the shape-based clustering for both subject groups is visualized
-Figure~\ref{fig:application}~(d) shows the results of the automatically triggered shape-based clustering for subject with and without back pain.
+Figure~\ref{fig:application}~(d) shows the results of the automatically triggered shape-based clustering for subjects with and without back pain.
 %Both subject groups (back pain and no back pain) are automatically clustered. The results are dropped into the canvas area as well (Fig.~\ref{fig:application}~d).
@@
 \textbf{Segmentation quality is crucial.}
-The radiologist pointed out the unusual strong similarity of the L3 vertebrae throughout the population.
+KH pointed out the unusual strong similarity of the L3 vertebrae throughout the population.
 %
@@
 %
-The visualization of multivariate data using connected views and different views allows to get fast visual feedback about subject groups.
+The visualization of multivariate data using multiple connected views allows to get fast visual feedback about subject groups.
 %
